% ============================================================
%  Harald Høffding, Søren Kierkegaard som Filosof
%  TRANSLATION (English) — complete book
%
%  Translated from the Danish of the revised edition
%  (Copyright (c) 1919 Lindhardt og Ringhof; SAGA e-book, 2020),
%  ISBN 9788726363111. First edition: P. G. Philipsen, Copenhagen, 1892.
%
%  The Danish source text is the publisher's copyrighted modern
%  edition and is therefore NOT reproduced here; only this English
%  translation is provided. Page ranges cited in the section markers
%  are those of the 1892 first edition (per Project Runeberg) and are
%  approximate navigational aids.
%
%  Structure mirrors the book:
%    Introduction
%    I.   The Romantic-Speculative Philosophy of Religion
%    II.  Søren Kierkegaard's Older Contemporaries in Denmark
%    III. Søren Kierkegaard's Personality
%    IV.  Søren Kierkegaard's Philosophy
%           A. Epistemology            (DONE)
%           B. Ethics
%                a. The Leap
%                b. The Stages  (alpha aesthetic / beta ethical / gamma religious)
%                c. The Standard
%    V.   Søren Kierkegaard and Christianity
%           A. Personal Breakthrough
%           B. The Last Word
%    Conclusion
% ============================================================
\documentclass[12pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{libertinus}
\usepackage{libertinust1math}
\usepackage{textalpha}
\usepackage[english]{babel}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Søren Kierkegaard as Philosopher (English translation)},
  pdfauthor={Harald Høffding},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\rhead{Høffding, \emph{Søren Kierkegaard as Philosopher}}
\lhead{\leftmark}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\setstretch{1.3}

% Unnumbered chapter that still appears in the running header + ToC
\newcommand{\bookchapter}[1]{%
  \chapter*{#1}%
  \addcontentsline{toc}{chapter}{#1}%
  \markboth{#1}{#1}%
}

\title{\textbf{\Large Søren Kierkegaard as Philosopher}\\[10pt]
  \large Harald Høffding}
\author{}
\date{\small Translated from the Danish\\
  (revised edition, 1919; first published 1892)}

\begin{document}

\frontmatter
\maketitle
\thispagestyle{empty}

% ---- Epigraph -------------------------------------------------
\clearpage
\thispagestyle{empty}
\vspace*{0.28\textheight}
\begin{quotation}
\noindent ``What is the significance of a primitive genius? It is not so
much to bring forth something absolutely new; for there is really
nothing new under the sun; but it is to revise the universally human,
the fundamental questions. This, in the deepest sense, is honesty. And
to lack primitivity altogether---that is, the revising element---to take
everything without further ado as given custom and usage, and to let it
suffice that it is custom and usage, that is, to withdraw from
responsibility for doing likewise: that is dishonesty.''
\end{quotation}
\hfill\emph{Søren Kierkegaard's Papers}, 1847.

% ---- Translator's note ---------------------------------------
\clearpage
\thispagestyle{fancy}
\section*{Translator's note}
This is an English translation of Harald Høffding's \emph{Søren
Kierkegaard som Filosof}, first published by P.~G.~Philipsen
(Copenhagen, 1892) and here rendered from the revised edition
(\copyright~1919 Lindhardt og Ringhof; e-book, 2020). Because that
Danish edition remains under the publisher's copyright, the Danish text
is not reproduced alongside the translation; the Danish may be obtained
from the publisher (see the catalog entry for a link). Page ranges given
in the source correspond to the 1892 first edition and serve only as
navigational aids. Høffding's own notes, printed as endnotes in the
modern edition, are restored here to footnotes at their anchors. German
titles and verse are left in the original; Greek is reproduced as
printed.

\tableofcontents

\mainmatter

% ============================================================
%  INTRODUCTION  (pp. 1--4)
% ============================================================
\bookchapter{Introduction}
\label{ch:indledning}

It was a melancholy thought for Søren Kierkegaard that the time would
come when he too would fall into ``the Lecturer's, the Professor's''
hand and be made an object of exposition. Indeed, he foresaw that this
very melancholy thought would be lectured on as well. What he found
repugnant in this was surely not the prospect of being made an object of
criticism. Rather, it was something bound up very closely with his whole
cast of thought. In the first place, he detested the ``treacherous''
admiration that comes afterwards, when the struggle is over, when renown
has been won, and that whitewashes the prophet's tomb---forgetting that
it is in the relation of contemporaneity to what is great that it must
show itself what sense and will one has to acknowledge it. His life's
work was to expose the self-deception in which we so easily entangle
ourselves with respect to what he called ``the historically corked-up
Ideality.'' But in the second place it was also one of his central
thoughts that there is no continuous psychological and historical
development. To exhibit a spiritual phenomenon---a thinker's or an
author's activity---as a link in a development, as borne by definite
inner and outer conditions and as bearing and conditioning the
subsequent development: that, for him, was to drag it down, to rob it of
its value. He would not be ``set down in a Paragraph.''

If my attempt to give a characterization and an appraisal of this
greatest thinker of our people should stand in need of a defense against
his own protest, then I must first observe that I proceed from an
entirely different conception of the relation between psychology and
ethics than his. I hold the conviction that, just as those figures who
stand before us with an ideal stamp arise under quite definite
psychological and historical conditions, so their ideal
significance---insofar as it is really present---is not weakened by their
being seen in their definite limitation and conditionedness. On the
contrary, the demonstration of this is absolutely necessary if the
abiding value is to be distinguished from what is due only to transient
and purely individual circumstances. And this holds not least of a
thinker of so pronounced and singular a peculiarity as Søren Kierkegaard.
If those who are to learn from him are themselves, or wish to be,
personalities, they must first have \emph{his} personal peculiarity
determined, so that they can decide what, according to their nature and
under their circumstances, they can carry over from him. Precisely in
consequence, then, of the great weight that Søren Kierkegaard has taught
us to place on personal appropriation, on personal truth, we must
undertake a psychological-historical and a critically examining treatment
of his personality and activity.

And as for the fact that we easily come into a self-deceiving relation to
the figures of the past, I must be permitted to make the personal
confession that when I now (1890), after an interval of some twenty-odd
years, have again taken up the study of Søren Kierkegaard's writings, I
have once more had to experience the power he has to make us turn our
gaze upon ourselves, to wound us by awakening a self-knowledge we are
otherwise so inclined to avoid. There is a sting in these writings that
no one who earnestly immerses himself in them will fail to feel, whatever
presuppositions he may otherwise bring. To them applies a word from the
book of Ecclesiastes that Søren Kierkegaard often uses in his journals
and discourses: ``Keep thy foot when thou goest to the house of the
Lord!'' His interpretation of this word was: You might easily come to
learn that the ideal was higher, the demand stricter, than you had
imagined; beware, therefore, lest you come to hear the ideal!---The
presuppositions with which I now [1890] resumed the reading of these
writings were, to be sure, very different from those with which, nearly
30 years ago, as a young theological student, I began my acquaintance
with them. At that time they led me into inner struggles and conflicts,
partly of a practical-personal, partly of a theoretical kind. After long
reluctance I had to acknowledge his strict conception of the Christian as
far more justified than the usual, idyllic and harmonious one. If
Christianity is really to be one and all for us, in life as in faith,
Søren Kierkegaard has drawn the consequences of it. For a time I tried to
keep going by distinguishing between faith and knowledge and between
ideal and reality. But I learned, more through personal life-experience
than through study, that it was an impossibility to slip through in this
way. The main decision had, after all, to be whether in my personal
life, in my spiritual household, I could really say myself to be borne by
supernatural help, and whether in my way of conducting my life I could be
said to let myself be guided by the ideals and commands of religious
ethics. Self-knowledge gradually gave me, after a long time of unrest and
torment, a clear answer to this, which for me has been decisive for my
whole life. Confining myself here to what concerns the public, I will
remark that I was now led, through studies in the history of philosophy
and investigations in the domains of psychology and ethics, to trace the
conditions, forms, and laws of the human life of the spirit. I was led to
see that ``the Humane'' is by no means, as Søren Kierkegaard says in one
of his journals, merely evaporated Christianity, but that it denotes a
view of life and a conduct of life which---founded by the Greeks,
deepened and inwardized by Christianity, expanded and clarified by the
theoretical and practical labor of the modern age---cannot, to be sure,
easily be summed up in brief formulas, but which nonetheless exercises
its power even over those who combat it, just as its steady development
is not hindered by transient excesses and caricatures. It is from this
standpoint of humanism (if we must, after all, have a word ending in
``-ism'') that I now look back on Kierkegaard's activity. But for me it
has been significant to see how valuable his appearance must be said to
be even from this point of view. It is by no means with polemical intent
that I have returned to him, although several views with which I have
also otherwise collided appear in him with the clarity and consistency of
genius, which makes it especially instructive to bring them forward. But
in that, by the homage I can still bring the great thinker, I render my
thanks for what he has given me for my personal development, I will at
the same time bear witness that he can be much to many who stand far from
him in faith. In that respect he does not cease to be our
contemporary.---I hope it will be excused that I have here spoken so much
about myself. My concern was to give a single example, the one nearest to
me, of Søren Kierkegaard's influence.

% ============================================================
%  I.  THE ROMANTIC-SPECULATIVE PHILOSOPHY OF RELIGION  (pp. 5--15)
% ============================================================
\bookchapter{I.\quad The Romantic-Speculative Philosophy of Religion}
\label{ch:1}

\noindent\textbf{1.} The nineteenth century began with a union of
religion, philosophy, and art as the watchword in the world of spirit.
People cherished an enthusiastic faith that truth was one, and that
everything of value, in whatever domain and in whatever form it might
appear, was comprised in this one truth, if only one immersed oneself in
it with open sense. No barriers to thought---and yet preservation of
harmony with religious feeling and with artistic imagination: that was
the great idea that filled the leading thinkers in the first decades.
They believed that in the strict form of thought they were only
expressing what the newly awakening religious sense and the so mighty
poetic urge, under other forms, lived and breathed in. The faith of the
philosophers of religion was, in form and content, not very different
from that which Faust expresses when Gretchen is anxious about how things
stand with his religion. Faust expresses a feeling of infinity,
inwardness, and fullness, awakened by life in nature and among human
beings, and which cannot be summed up in any finite expression, nor
exhausted in any name---»Name ist Schall und Rauch!« And Gretchen thinks:

\begin{verse}
So ungefähr sagt das Pfarrer auch,\\
Nur mit ein bischen anderen Worten.
\end{verse}

\noindent These words might serve as a motto for the age's conception of
the relation between faith and knowledge, as it appears---with
characteristic and significant differences---in Schleiermacher and in
Hegel.

\medskip

\noindent\textbf{2.} Right at the century's beginning (1799)
\emph{Friedrich Schleiermacher} (b.\ 1768, d.\ 1834) published his »Reden
über die Religion an die Gebildeten unter ihren Verächtern«. He here
conceived of religion as an immediate consciousness that everything
finite has its subsistence in and through an infinite, everything
temporal in an eternal---not in such a way that the eternal and infinite
falls outside the temporal and finite, but in such a way that it is the
latter's inner essence and soul. While in our knowing and our acting we
direct ourselves toward the finite, the determinate and the limited, in
order to grasp it or to change it, in our feeling we live a whole and
undivided life, a universal life, in which opposition and limitation are
annulled, and in which the one-sidedness that clings to every theoretical
and practical striving falls away. Feeling is the domain of religion,
because it is a life in the whole and the universal---because there
announces itself in it that ``instinct for the Universe'' which is
precisely one with religion. Every feeling, according to Schleiermacher,
has a religious character when it is not morbid and corrupted. And as the
religious feeling thus expresses itself not in opposition to, but as a
deepening of, life in the world of experience, so the deity too stands
not as a being separated from this world, but as its inner unity and its
animating power. And immortality is not an existence that is to replace
the present one, but consists in this: to feel, in the midst of time, the
eternal within oneself.

Later Schleiermacher developed, in more theological form, his conception
of religion in »Der christliche Glaube« (1821). He here determines the
religious feeling as unconditioned feeling of dependence, and thereby
establishes a moment of the greatest importance for religious psychology.
It is, on the whole, Schleiermacher's fine psychological sense that gives
his works their abiding significance. Both his earlier description of the
religious feeling as universal feeling of unity, as a centralization of
the life of feeling in relation to existence, and the later determination
of it as feeling of dependence are valuable contributions to the
psychology of religion, even if more precise determinations are necessary.

Dogma is for Schleiermacher always something derived, something
secondary. Dogmatic propositions originally arose as expressions of the
religious feeling. First this feeling gives itself air in poetic and
rhetorical fashion; the forms of expression thus arising are then
developed in the form of doctrine, as confessions. And since the poetic
expressions often had to come into apparent conflict with one another,
even where the feeling that lay at their basis was in reality the same, a
reflective and dialectical treatment became necessary in order to bring
about full clarity. Through these intermediate stages the dogmatic
propositions came into being, and their validity and significance are
therefore different from such propositions as are won by a purely
theoretical or philosophical route. Schleiermacher wishes to carry
through a definite separation between the doctrine of faith and
philosophy, a separation he supports by his theory of knowledge (which we
know from the unfinished »Dialektik«, published after his death), which
critically demonstrates the limits of knowledge and holds a place open
for an immediate feeling-relation to the deity, of whose essence
theoretical thinking, in his view, can form no concluded concept.

By this fine and spirited conception Schleiermacher placed himself in
definite opposition to the two dominant tendencies of the age, orthodoxy
and rationalism. He rejected the former's external literalism and the
latter's superficial explaining-away. His work has an abiding
significance for the philosophy of religion, and he stands as the
greatest thinker theology can show beside Augustine and Zwingli.

\medskip

\noindent\textbf{3.} But Schleiermacher did not want to be merely a free
religious thinker; he wanted to present ecclesiastical theology and
believed he could do so from his standpoint. Here he was caught in a
great illusion and sacrifices at the altar of the age; for of his
personal love of truth there is, despite all that orthodox and radical
fanatics have thought, no doubt. He was the most Socratic personality in
the philosophy of that time. A rare sense for individualities and a
strong feeling of the right and significance of the individual mark his
life, as we know it through a rich and interesting collection of letters,
and animate his thinking. But he did not see that on such a standpoint an
ecclesiastical theologian cannot place himself. The Church as a
historical community rests on the assumption of revelation as objective
fact, as the absolute and adequate expression of the eternal truth, and
therefore comes forward with a claim to unconditional authority over the
feeling-life of individuals. For Schleiermacher the dogmas in reality
became only symbols, the standard subjective, and from his standpoint
there is only one step to Ludwig Feuerbach's, for whom theology is only
psychology. Symbols cannot be imposed; they must be freely chosen or
produced, and they can be changed without the underlying feeling
therefore needing to change. But the Church cannot concede this in the
case of its dogmas; for then the very concept of revelation would fall
away. For Schleiermacher even God's personality was only a symbolic
representation! But if, with respect to all religious representations, a
fundamental distinction is to be made between image and reality, then
thereby positive religion is abolished, which in any case must, on
certain points, maintain that truth has entered the world in its whole
fullness and in its adequate form.

And on closer inspection it appears at the same time that the very
determination of the religious feeling (both in its older form, as
feeling of unity, and in its later form, as feeling of dependence)
presupposes several theoretical presuppositions---more philosophy, that
is---than Schleiermacher is aware of. The religious feeling has, in his
conception, a harmonious character. It is precisely the feeling through
which all disharmony, one-sidedness, imperfection, and sin is overcome.
The feeling of dependence takes on, in the end, the character of trust
and enthusiasm. But thereby an optimistic and harmonious view of the
world is presupposed. It is presupposed that there are no experiences
that could bring about a wholly different total-mood toward existence, or
in any case give the religious feeling a less harmonious stamp. In
Schopenhauer and Søren Kierkegaard the opposition to the romantic,
harmonizing conception appears on this point as energetically as can be
conceived.

It is, on the whole, characteristic of the development in the domain of
spiritual, and especially religious, life in our century that the
oppositions grow ever sharper as time advances. The century begins with
harmony and reconciliation, but the splitting and division is consummated
with increasing emphasis. Schleiermacher did not live to see the radical
sharpening that came with Strauss and Feuerbach, whose forerunner he
himself is---partly through his critical investigations of the New
Testament writings (which led him, among other things, to reject on
historical grounds the authenticity of the first letter to Timothy and of
the accounts of Jesus' birth), partly through his psychological method.
He did, however, live to feel the sharpening on the orthodox wing. When
in the year 1821 he published for the third time his ``Discourses on
Religion to the Educated among its Despisers,'' he wrote in the preface
``that the times had changed so strikingly that the persons to whom these
discourses are addressed no longer seem to exist at all, but that when
one looked about among the educated, one would rather find it necessary
to write discourses to mopers and slaves of the letter, to superstitious
people who show both their ignorance and their unkindness by the manner
in which they condemn others.''---The times had changed, and they were to
change still more.

\medskip

\noindent\textbf{4.} Alongside Schleiermacher, and in a polemical posture
toward him, there worked at the University of Berlin in the twenties
another philosopher of religion who expresses the age's craving for
harmony between faith and knowledge to an even higher degree than he.
This was \emph{Hegel} (b.\ 1770, d.\ 1831). What distinguished this
thinker---whose writings'\footnote{Hegel's philosophy of religion is
most fully presented in the work published after his death, »Philosophie
der Religion«; but in his »Encyclopädie der philosophischen
Wissenschaften« (1817) it is already stated in more concise form. Most
easily accessible, perhaps, is his conception of Christianity in
»Philosophie der Geschichte«.} abstract form makes it difficult for most
present-day readers to penetrate to what he has of really significant in
his ideas---was his great sense for the unity of spiritual life under all
its different forms and at all its different stages. Since he now saw in
human spiritual life a form of the revelation of an eternal
world-spirit, it became possible for him to sketch an idealistic doctrine
of development, within whose framework he ranged---often with great
arbitrariness and with more or less spirited reinterpretation---everything
that his rich learning placed at his disposal from the various domains.
Since the divine expresses itself at all stages of existence and in all
forms, Hegel acknowledges no absolute oppositions. All differences and
oppositions are moments in the development, which serve to give it
fullness of content and energy, but which are themselves continually
annulled and united in a higher unity. Through the continual positing and
annulling of oppositions the development of the world proceeds. It is in
a certain sense the effect of contrast that is for Hegel the fundamental
phenomenon illuminating existence. It shows us, indeed, how the
oppositions call one another forth and cannot be separated. A higher
stage or a higher form of existence is, on the Hegelian conception,
characterized by the fact that oppositions have been run through and have
deposited their result in a new form of unity. Mediation, the
equalization and annulling of oppositions (or, as Hegel incorrectly calls
them, contradictions), became the watchword for Hegel's adherents. To
remain standing in a relation of opposition was a sign that one was
caught in the finite and external; the point was to reach the higher
unity, to unite the oppositions of existence in harmony.---An ingenious
idea and a great goal! But from the general idea to the special
theoretical and practical execution the way is long, and Hegel and his
adherents all too often confused the program with the execution.

The relation between philosophy and religion was, according to Hegel,
this: that they have the same content, only that philosophy expresses
this content in the language of abstract thought, while religion
expresses it in the language of feeling and imagination. The difference
is thus only a difference of form, which is not to touch the essential.
What the believer conceives as historical events---e.g.\ the creation of
the world, God's becoming man in Christ, and so on---are for philosophy
eternal truths and laws that hold for all times. The Christian doctrine
of the suffering god is thus for philosophy not the expression of a
single series of historical events; rather its essential content---which
only philosophy grasps in its clear, adequate form---is the truth that
the infinite and divine stirs in the finite world, while yet every finite
form is imperfect and must be shattered. It is, then, only through the
continual perishing of finite forms that the infinite unfolds itself in
the world. Only in the infinite process of ever new and ever again
perishing forms of life does the eternal life of the deity consist.---

Already from this one example it becomes clear that Hegel, just as much
as Schleiermacher, in reality makes the dogmas into symbols, only that
with him they become symbols for general ideas, not for subjective
feelings. It rested, then, with Hegel as with Schleiermacher on an
illusion when he believed he had reconciled faith and knowledge. But it
was Hegel's definite conviction that nothing of essential significance
was lost by the content of faith being transposed into philosophical
form; on the contrary, he believed that only then did it really become
the property of the human spirit. He did not see what Strauss and
Feuerbach first sharply demonstrated---that the transposition really
meant a transition to a wholly different view of life and the world than
that which is characteristic of positive religion, especially of
Christianity, of which of course there is here first and foremost
question. The whole relation between Christianity and humanity Hegel
conceived, like Schleiermacher, as a harmonious relation. Romantic
optimism is characteristic of both.

\medskip

\noindent\textbf{5.} But even purely philosophically and humanly there
was a misgiving that had to come forward in the face of the Hegelian
philosophy's fundamental thought of the unity of oppositions. Even if
there should be a higher unity in which all the oppositions of existence
find their reconciliation, we are, as existing beings, in the midst of
this world of oppositions, far from this unity. The surge of the
oppositions raises us and lowers us; we are dependent on experience, we
run up against barriers both in our thinking and in our acting, our
horizon is limited, and we often end in problematic results or run our
head against the wall; \emph{for us}, therefore, that unity is an
unattainable ideal, and the reconciliation it contains is only for the
imagination, or for faith. This essential point---the difference between
metaphysics on the one hand, psychology and ethics on the other---Hegel
overlooked, and therein lay his greatest philosophical error. The
consequence was that he, and still more his adherents, often confused the
purely abstract and logical establishing of the connection of oppositions
---e.g.\ of opposed standpoints---with an overcoming of the oppositions
through real work, or, in the ethical domain, through really having
\emph{lived} them, having practically attained a reconciliation of the
opposed tendencies between which life moves. Add to this that, so long as
life lasts, ever new oppositions, and thereby new problems, come forward;
``the higher unity'' cannot therefore be won once and for all. The theory
of the unity of oppositions thus sets up an ideal, but does not give us a
real possession. It was Hegel's great error that his endeavor, to use his
own words, aimed at ``rewriting the ideal into a system.'' Thereby he
came (as not only his philosophy of religion, but also his ethics and
theory of the state show) to violate the right both of reality and of the
ideal.

Søren Kierkegaard's opposition to Hegel acquires no small significance
through the sharp and penetrating manner in which he attacked this main
point in the Hegelian philosophy. And it must be well noted that we are
here not confronted with the ideas of a single philosopher. On the
contrary, the Hegelian doctrine on this point is really of a typical
kind. It represents---in its best meaning---a natural striving, which
human spiritual life can at no time give up, after finding connection and
harmony, after winning by its own powers mastery and survey over life.
And---in its more superficial application---it represents a
fantasy-relation to existence, an aesthetic-contemplative attitude toward
problems, crises, and oppositions, which one lets unfold as an
interesting spectacle, by whose conflicts one feels oneself untouched,
regarding oneself as the realized ``higher unity.''---Even in its best
meaning, however, this system---in which the romantic movement from the
century's beginning was clarified in abstract form---stood in need of a
serious reaction, one that in practice and theory laid an accent on
reality in its definite concrete and individual character. And in that
Søren Kierkegaard brings this reaction, as energetically as could be
wished, into the ethico-religious domain, he brings it at the same time
in a manner that can act as a standing resistance against similar
tendencies in human nature, such as later times too present.

Goethe's Faust, moreover, had already here expressed what was at stake.
When in Nostradamus' book he sees the symbol of the universe, he first
exclaims enthusiastically:

\begin{verse}
Wie Alles sich zum Ganzen webt,\\
Eins in dem Andern wirkt und lebt!
\end{verse}

\noindent---but he soon becomes melancholy that this unity of oppositions
is not given in the form of reality:

\begin{verse}
Welch Schauspiel! Aber ach! ein Schauspiel nur!
\end{verse}

\noindent Søren Kierkegaard conceived it precisely as his life's task to
impress what the preceding romantic-speculative generation had forgotten:
that life is something other and more than a spectacle for us.

% ============================================================
%  II.  SØREN KIERKEGAARD'S OLDER CONTEMPORARIES IN DENMARK  (pp. 16--27)
% ============================================================
\bookchapter{II.\quad Søren Kierkegaard's Older Contemporaries in Denmark}
\label{ch:2}

\noindent\textbf{1.} Johan Ludvig Heiberg (1791--1860) was the apostle of
Hegelian philosophy in Denmark.\footnote{A more detailed treatment of
this chapter's subject I have given later in my book \emph{Danske
Filosofer} (1909).} He first spoke up for it in a treatise from 1825,
which was a contribution to the deterministic controversy raised by
Howitz. Later he applied it especially to aesthetic and theological
subjects. He found in Hegel's philosophy an expression of the same thing
that Goethe strove for in his view of life and in his poetry. ``Hegel's
system,'' he says (\emph{Om Filosofiens Betydning for den nærværende
Tid}. Copenhagen 1833, p.~48), ``is the same as Goethe's. That two such
great minds should at the same time have arrived at the same result
cannot but awaken a favorable impression of the result\dots\ To
characterize the Hegelian philosophy in a few words, one may say that
it, like Goethean poetry, reconciles the ideal with reality, our demands
with what we possess, our wishes with what has been attained.'' As for
the relation between religion and philosophy in particular, Heiberg
expressed himself about it not only in his treatises and in his journal
\emph{Perseus}, but also in his great philosophical poems, which are
surely the finest of what our literature possesses in this direction. It
is Hegel's fundamental thought that he expresses when it is said, in the
Reformation cantata, that ``Thought ascended to the heights at the very
moment it descended into itself.'' For it is by thought's own path, by
thinking's clear elucidation of its own nature, that (according to Hegel)
the same ideas are to be found which constitute the essential content of
religion. To be sure, there is a great difference in the form: religion,
at its immediate stage, commands images and symbols, beside whose splendor
of color the forms of thought might seem poor and paltry. But Heiberg
consoles us in ``Protestantism in Nature,'' the poem in which he has
perhaps reached highest:

\begin{verse}
Turn your gaze toward your inner self!\\
What perished in your world\\
you will find within your thought.\\
All that vanished, in there,\\
gained life and existence.
\end{verse}

\noindent What has vanished is the outer, perceptible form, which the
childlike imagination misses, and with which it believes all is lost. But
for Heiberg, just as little as for Hegel, does anything go lost. He saw it
as peculiar to ``the present age'' that it wished to \emph{know} what
former ages had felt and believed; but this was, in his opinion, ``so far
from wanting to make an end of feeling and faith that, on the contrary,
it wants to consolidate them; for only when we know that they contain the
truth can we surrender ourselves to them undisturbed.''---This conception,
however, aroused misgivings among the theologians, and it was not in this
clear and resolute form that speculative philosophy of religion managed
to win wider assent here at home. It first had to undergo a dogmatic
modification of a peculiar kind.

\medskip

\noindent\textbf{2.} \emph{Hans Martensen} (1808--1884) felt himself,
already as a young man, strongly drawn to Hegel's forms of thought, which
seemed to him to open the possibility of a knowledge of the divine, while
Schleiermacher, who at the same time influenced him strongly, in the end
let the speculative problems lie. When Schleiermacher, during his visit to
Copenhagen the year before his death, was one day walking with the young
promising theological graduate, the latter asked him ``quite naively''
(see \emph{Martensens Levned} I, p.~69) whether he believed that a
knowledge of ``God's essence in itself, of the inner eternal life-process
in God'' was possible. Schleiermacher answered quite calmly: ``Ich halte
es für eine Täuschung.'' But the young theologian's speculative
enthusiasm would not be damped by the old master's warning. Hegel's forms
of thought enticed him, though he could not remain standing at Hegel, who
was too abstractly logical for him. He no doubt also sensed that, carried
through consistently, the dogmas could for Hegel, as for Schleiermacher,
become only symbols. He credited himself with the power to reach something
even greater than these thinkers had considered attainable: he wanted to
reconcile faith and knowledge in such a way that he retained the forms of
both! Characteristic of Martensen was a union of mysticism and
imaginative vision with thinking---that is, a thinking that strove to
bring about a certain coherence among the images that thrust themselves
forward before the imagination as moved by the temperament. His ideal was
the old mystics and theosophists; it was in the tracks of Meister Eckhart
and Jakob Böhme that he wished to walk. Of contemporary thinkers he most
admired the Catholic philosopher Franz Baader, who likewise wished
precisely to renew Jakob Böhme's philosophy. His distance from Hegel he
expressed early, in a critique of one of Heiberg's treatises
(\emph{Maanedsskrift for Litteratur} 1836) and in his dissertation on
``The Autonomy of Self-Consciousness in Christian Dogmatics'' (1837).
Against Hegel he objected above all to the impossibility of thinking
without presuppositions. The human being has not, after all, created
himself, but is posited by God---and thus has God as his presupposition,
his authority. In virtue of the dogma of creation, not independence
(autonomy), but dependence on God (theonomy), must become the watchword.
The task must become: speculation on the basis of faith. Earnest must be
made of the old saying that the fear of God is the beginning of wisdom.
Later (in his dogmatics) he expressed this by saying that the doctrine of
faith as a science was possible only for the consciousness reborn by faith
and grace.

One cannot deny Martensen an eye for the fact that his higher science
required a good many presuppositions; whereas it may well cause surprise
that he did not ask himself of what value a ``science'' is that rests on
so many and such enormous presuppositions. But the young theologian was
enthusiastic, and enthusiastic became those who heard him. People reckoned
``a new era'' from his appearance and expected a golden age of theology. In
his lectures on the history of recent philosophy he first led his hearers
to Hegel's philosophy as the necessary result of the development of
thought; and when they were impressed by this, there came astonishment and
admiration to the second power when he showed them the necessity and
possibility of a higher standpoint---beyond the greatest thinker of the
age! In his dogmatics he led his hearers, and later his readers, from
dogma to dogma without any decisive difficulty halting them. Trinity,
incarnation, atonement are presented as a series of images of great events
in the world of spirit, often accompanied by spirited and profound
remarks, but in such a way that every sting for thought is taken away.
That must come, presumably, from the rebirth. Only at one point does the
dogmatic thinker halt, without even his reborn consciousness being able to
clear the matter up for him: it is at the question of eternal damnation.
Here he finds---to the honor of his feeling of humanity, though not to the
honor of his orthodoxy---a stumbling-block for thought.

Philosophically, Martensen has, in the first place, offended against the
truth that thinking always needs presuppositions. He has not seen that
these presuppositions are always sought to be reduced to the least
possible, and that they are then (in the theory of knowledge) sought to be
demonstrated in their origin from human nature. We construct only those
presuppositions that are necessary in order to understand the reality
given in experience, and whose basis in human nature can be established.---
And in the second place, he has never shed any light on how it actually
comes about that the difficulties the dogmas present to the natural
consciousness fall away for the reborn consciousness. Does one then get a
different logic when one is reborn? And in that case, would it not be of
importance to have this higher logic set forth? Or should the whole
difference perhaps be that one no longer thinks? And is it, then, so easy
to cease thinking, to take reason captive? At any rate there came one who,
on this point, felt the most painful contradiction, the most passionate
collision between different forces, and who felt himself deeply outraged by
this attempt at a ``higher'' scientificity.

Seen from the religious point of view, the misgiving had to be this:
whether, after having obtained the presuppositions in question, one now
really also has the time, the desire, and the strength for speculation? For
it might be that those presuppositions were of such a kind that they had to
be acquired ever anew, that the appropriation could not take place once and
for all, and that the ever-renewed problem of appropriation made ``the high
unity'' impossible---even in the form in which Martensen, in his arrogant
attempt to outbid Hegel, had conceived it.

\medskip

\noindent\textbf{3.} It was German speculative philosophy that, with
Heiberg and Martensen, made its entry among us. But already before Søren
Kierkegaard made his mighty attack on this tendency, it had met able and
well-grounded resistance among Danish thinkers. Just as Treschow, at the
century's beginning, had by his sober critical thinking and his sound
psychology counteracted the obscurities of Schellingian philosophy, so
already in the thirties \emph{Frederik Christian Sibbern} (1785--1872) came
forward with a penetrating critique of Hegelian philosophy. Sibbern was,
by his innermost nature and by his whole development, unreceptive to a
philosophy that wished to spin everything out of itself, and that by its
speculative abstractions came into a relation of remoteness from life.
Characteristic of him in this last respect is a trait he himself relates.
``I remember that Søren Kierkegaard once, in his Hegelian period, met me on
Gammeltorv and asked me: what relation there was between philosophy and
life in reality? The question startled me, since all my philosophy aimed
at studying life and reality; but afterwards I certainly had to become
aware that for a Hegelianized thinker the question was natural, since the
Hegelians did not study philosophy existentially.''

For Sibbern, thinking always had to have a given, a factual basis.
Philosophy must be \emph{explicative}---that is, must make clear and
distinct the content presented in given experience---before it can become
\emph{speculative}, that is: set up its further-reaching hypotheses on this
basis. And to this was added that Sibbern, in the interesting philosophy of
development on which he early entered, conceived existence as a great
process, or a system of processes, proceeding from many different starting
points. It is the \emph{sporadic} character of development that he
especially emphasizes. Since the knower is himself one of these many
starting points, and is himself placed in one of these sporadic processes
of development, he cannot possibly survey the whole. Although we are links
in the universal existence, and its innermost essence must therefore stir
in us as well, still we cannot from our vantage point form a closed image
of the whole life of the world.

It was therefore firmly grounded in Sibbern's whole philosophical outlook
when he came forward with a \emph{Critique of Hegelian Philosophy} (1838),
directed especially against Heiberg---a work that contains many sound and
able thoughts and does Danish philosophy honor.---That he here called
Heiberg a philosophical dilettante caused Heiberg, some years later in
``A Soul after Death,'' to have Sibbern's writings belong to the library in
hell.

Sibbern's philosophy of religion held that the religious content should be
taken like any other given content of experience. The point was to live
oneself into it and find its value and validity. He was equally opposed to
the rationalistic or speculative transposition of living, personal faith
into abstract concept, and to the orthodox assertion of the dogmas as
outer, objective truths. It was, at first, the opposition to the aridity of
rationalism's understanding that determined Sibbern's position here. But as
the years went by, he felt himself more and more repelled by the orthodox
tendency, and he began to doubt how far the religious content could really
be called \emph{given} in the same sense as the experiences that science
takes as its basis. A definite critical attitude toward positive religious
faith comes forward in Sibbern in a series of remarkable university
programs from 1846--1849. It was not so much historical criticism that
brought him to doubt how much, in the religious domain, was really
\emph{given}, as it was a psychological conviction that what one took to be
original expressions of religious emotion could not be immediate
outpourings of personal experience, but had to be due to a more composite
development, in which manifold traditions and reflections had intervened.
He was, on the whole, led more and more away from the objective and
historical side of religion to the inner personal life, which is its
source of life. And in that he now connected with this his significant
thought that all development proceeds sporadically, from individual
starting points, he demanded that the right of subjectivity be
acknowledged. The aging thinker consistently carried through here what his
younger friend, Søren Kierkegaard, had already some years before
developed: that subjectivity is the truth---only that he does not, with
Kierkegaard, ground it merely on the necessity of personal appropriation,
but lays the weight especially on this, that it is in the individual
personalities, each for itself, that the real spiritual life stirs, and
that one would stop up these springs of life by imposing on all a common
form of view of life.

That Sibbern did not act more strongly on his contemporaries was connected
especially with his lacking the ability to shape his thoughts in an easily
accessible and condensed form. His expositions suffer from prolixity and
from various peculiarities, which stood hinderingly in the way of the
proper appreciation of the significance of this remarkable
personality.\footnote{A detailed characterization of Sibbern's personality
and of his cast of thought I have given in \emph{Mindre Arbejder}. First
Series, pp.~61--112.}

\medskip

\noindent\textbf{4.} A younger friend and colleague of Sibbern, the poet
and philosopher \emph{Poul Møller} (1794--1838), must be especially
remembered when the talk is of Søren Kierkegaard's older contemporaries,
since he was the one who stood in the closest personal relation to
Kierkegaard. Poul Møller had for a time been strongly occupied with Hegel's
philosophy, but later placed himself (as is seen from his treatise on the
possibility of proofs for immortality) critically toward it. As a
university teacher he worked most through his lectures on the history of
philosophy. He had, as is seen from his ``Stray Thoughts'' (\emph{Strøtanker}),
his eye open to the personal side of philosophizing; he kept not merely to
the objective content of the philosophical theories, but went back from
this content to the philosophizing personality. In his draft of a treatise
on affectation, Poul Møller's sense for the significance of personal truth
expresses itself especially in a characteristic way. It seems to me that
here is something that gives him an interesting position as Søren
Kierkegaard's forerunner. In reading these drafts one comes to sense what
the continual conversations between Poul Møller and Søren Kierkegaard had
turned upon.

Affectation is, according to Poul Møller, a combination of falseness and
self-deception. It arises through one's wanting to be what by one's nature
one cannot be, and therefore imagining to oneself and to others that one is
other than one really is. ``The affected person does on purpose what has no
significance except where it is called forth by the power and prompting of
nature or by a higher necessity.'' But Poul Møller has a clear eye for the
fact that it is something which, by the nature of circumstances, is not to
be avoided at certain stages of development. For when the personality is to
be developed and expanded, it must necessarily overstep its limits and take
foreign matter into itself, without at once being able to make this foreign
matter its own real property. Thereby a momentary affectation arises as a
point of transition. Affectation becomes fixed when one consistently
carries through the foreign principle, although it cannot take root in one's
personality. Moral feeling, self-regard, and immorality alike can in this
way be shells in which one hides oneself from oneself and others, taking
pleasure in them without, however, really filling them out. The use of
terminologies without insight into the corresponding thoughts, and
exaggerated symmetry in a scientific system, are more theoretical forms of
the same thing.

Poul Møller goes so far in his assertion of the importance of agreement
between the inner and the outer that he declares that ``no expression of
life has truth unless there lies in it a creative self-activity.'' He
advances this proposition with full awareness that it stands as a paradox.
For how much truth, indeed, will there be in the world if we are true only
in what we ourselves bring forth? But he holds the conviction that in the
falsely assumed, and in the objective common stamp, there is a great danger
to the originality and richness of life. ``Affectation most often comes
from one's not having the strength to fall out with the world in order to
show one's personality true\dots\ Everyone has by nature his deep stamp,
but lets it be effaced in order to please others.---He will not come
forward with his own person, does not believe in his person's infinite
depth. If each individual man, without fearing the reproach of simplicity,
judged of things just as they presented themselves to him, then glorious
phenomena would come forth.''---

On his deathbed Poul Møller sent, through Sibbern, a greeting to Søren
Kierkegaard. The three men belong together by kinship of thought. The idea
of personal truth and its significance they have all three grasped, though
only one of them was able to develop it in its whole force and rigor. And a
great sympathy did Søren Kierkegaard feel for his two teachers.---In the
introduction to the ``Posthumous Papers'' (\emph{Efterladte Papirer}) the
following is told. ``One of his somewhat younger schoolfellows has related
how he once, weighed down by heavy thoughts, met Søren Kierkegaard in
Frederiksberg Garden. `Without, moreover, any words being exchanged about
the mood that oppressed me,' the account runs, `Søren Kierkegaard suddenly
stepped up to me and asked, with a look and voice in which sympathy
trembled, whether I knew Professor Sibbern? \emph{Him} I should turn to;
\emph{that} was a \emph{whole} man; he was amiability itself, and in
intercourse with him there was something calming.' ''---As for Poul Møller,
Kierkegaard gave vent to his enthusiasm for him in the dedication to his
memory in \emph{Begrebet Angest} [The Concept of Anxiety]: ``The late
Professor Poul Martin Møller, the happy lover of Greek culture, admirer of
Homer, confidant of Socrates, interpreter of Aristotle---Denmark's joy in
the joy over Denmark, though `far traveled' always `remembered in the
Danish summer'---my admiration, my loss---to him this writing is
consecrated.''

This dedication indicates at the same time that Poul Møller introduced and
guided Kierkegaard in the study of the Greeks, which became so significant
for him. It means, on the whole, much in Kierkegaard's mouth when he calls
his late teacher the confidant of Socrates. There lies in it a hint of a
very essential spiritual connection between the two men, of which people
have not always been aware.

% ============================================================
%  III.  SØREN KIERKEGAARD'S PERSONALITY  (pp. 28--53)
% ============================================================
\bookchapter{III.\quad Søren Kierkegaard's Personality}
\label{ch:3}

\noindent\textbf{1.} There are thoughts that can grow up only in a
certain definite kind of soil and in a certain kind of air. And there is
no doubt that Søren Kierkegaard's thoughts belong among them. They are
like the wine that can grow only on volcanic ground. This ground, then,
we must first and foremost learn to know. And the ground lies deep. We
are directed back beyond himself to his family and his race, and to the
traditions that formed the spiritual atmosphere of his childhood.---For
the validity or invalidity of the thoughts, such a demonstration is not
decisive. Whatever peculiar conditions their development may be due to,
they can in themselves be equally valuable. It might well be that
especially valuable thoughts can grow up only under \emph{such}
conditions! But these conditions must be brought out and known when the
question of the thoughts' vital force under other conditions is to be
decided.---

\medskip

\noindent\textbf{2.} His family stems from West Jutland. In the West
Jutland race there is found, alongside tenacity, shrewdness, and humor,
also no small propensity to melancholy and weariness of life.

On the heath in Sæding parish near Ringkøbing, in the year 1768 a
twelve-year-old ``herd-boy'' walked with his sheep. He endured hunger and
cold and was oppressed by loneliness and forsakenness. Then in his despair
he went up onto a hill and cursed the God who had given him this wretched
life.---This boy was Søren Kierkegaard's father. In a journal entry from
1847 (\emph{Søren Kierkegaards Papirer}, edited by P.~A.~Heiberg and
V.~Kuhr, seventh volume) the following passage is found:

\begin{quote}
``The terrible thing about the man who once, as a little boy tending sheep
on the Jutland heath, suffered much evil, hungered and was famished, stood
up on a hill and cursed God---and that man was not able to forget it when
he was 82 years old.''
\end{quote}

When the first editor of Kierkegaard's papers showed Søren Kierkegaard's
brother, Bishop P.~C.~Kierkegaard, this passage, the old bishop burst into
tears and said: ``That is our father's history, and ours too!''
(H.~P.~Barfod's biography of Bishop Kierkegaard in \emph{Illustreret
Tidende} 1888). In the history of the Kierkegaard family that event plays
a great role, both as effect and as cause. It stands as an outflow of the
heavy temperament that felt everything more strongly and deeply than
others, and of the passion that drove every mood and every thought out to
its utmost consequence. But it also comes to stand as a threatening
symbol, which set the dark temperament in ever new motion and awakened the
thought of the God who visits the sins of the fathers upon the children.

The boy from the heath, Michael Pedersen Kierkegaard, was sent to
Copenhagen and apprenticed to a hosier. He worked his way up to becoming a
rich man, but never forgot that moment on the heath in which he believed
he had committed a sin against the Holy Spirit. His melancholy dread often
gave itself vent in despairing words, and it then stamped the life of the
home with its darkness, although he was otherwise a mentally lively man who
in wit and acumen could hold his own with his highly gifted sons. Already
in his fortieth year he had withdrawn from trade and lived thereafter
occupied mostly with philosophical reading; he especially studied the
German philosopher Wolff eagerly.---For the understanding of the following
characterization of Søren Kierkegaard, it must be noted that other
hereditary influences made themselves felt than those stemming from the
melancholy, brooding father. The mother is described as a nice little woman
with an even and contented temperament. A contrast that is not rare, and
that recalls the one between Goethe's parents, where likewise ``des Lebens
ernstes Führen'' derived from the father, ``die Frohnatur'' from the
mother. To be sure, the opposite elements appeared in Søren Kierkegaard in
a far more contentious and contrasting way than in that grand master of
life's harmony.

The younger son, \emph{Søren Aabye Kierkegaard}, was born in Copenhagen on
5 May 1813. Of his outward life there is not much to relate, and yet, when
one day all the posthumous papers are available, an interesting biography
of him will be able to be written---so true is it that the essential in
life, for a nature like his, is not the outward events. In Georg Brandes'
book on Søren Kierkegaard a brilliant characterization of his literary
personality is given. Something has here been accomplished that cannot be
done over. But I believe that Brandes, precisely because his eye was
directed especially toward S.~Kierkegaard's literary activity, has come to
lay \emph{too} great weight on individual events in his life, which
certainly became motives and themes in his production, but were not wholly
decisive for the direction of his spirit. And neither can I agree with
Brandes that piety and contempt were S.~Kierkegaard's two fundamental
passions. In my view they occupy a more subordinate place in his
psychology.---It must, however, be noted, especially with regard to this
last point, that when Brandes wrote his book, only the first two volumes of
the ``posthumous papers'' had appeared.\footnote{The first edition of
Kierkegaard's papers was undertaken by H.~P.~Barfod and the Swiss
theologian Gottsched. A new, critical edition is now being undertaken by
P.~A.~Heiberg and V.~Kuhr. I cite according to this edition, so far as it
is available.}

\medskip

\noindent\textbf{3.} The most decisive element for Søren Kierkegaard's
whole personality must be sought in his melancholy, as it lay in his race
and in his temperament and was nourished by his upbringing. ``I am not a
human being; I am melancholy to the very border of insanity,'' he says
(1849) in his notes, where he gives an account of his circumstances of
life. It was melancholy that lay at the basis of his great piety and of his
reverence for all authority. It brought with it a need for absolute points
of support and a shyness before the confusion of the new. He needed
something that could hold him up when everything seemed to dissolve for him
into a bottomless darkness. Only when such an absolute ground had been laid
could other sides of his nature, especially his great dialectical urge,
unfold freely and powerfully. In religious respects this led to his sharp
reflection and criticism being applied only to the question of the
appropriation of the religious content, not to the testing of the inner
coherence and validity of this content itself. And the feeling of contempt
that could stir in him is also easily understood from the consequences of
melancholy for his inner life. Accustomed as he was to walking between
abysses and terrors, ordinary, ``lightly living'' people had to make a
strange impression on him, just as melancholy itself had to lead him to
picture people's conduct and their motives in more unfavorable colors than
real experience warranted. But contempt was not an original element in his
nature. Rather there was in him a heartfelt compassion---understandable
also from the melancholy---with people, especially with those who suffer,
or whom he believed to be suffering. From several sides it is attested that
he had a remarkable eye for whether something pained those he was together
with, and that he knew how to soften and console in a manner as
considerate and delicate as no one else. He not only found comfort and
relief in going about and talking with common folk in the street, but felt
a heartfelt compassion with them and knew how to enter into their affairs.
And even at the last, when his conception of Christianity sharpened, and
when he felt the hour drawing near when the ideal was to be asserted in all
its rigor, he was for a while held back by compassion with the happiness he
was to disturb, with the unrest and suffering he was to cause where
hitherto a mild and harmonious conception of life had prevailed.

It is naturally impossible to derive the various qualities and tendencies
of a human being, especially of a richly endowed one, from a single cause.
Such people, whose significance rests precisely on their coming to make
spiritual experiences that most people are spared or miss, will have in
their nature several different starting points, several elements between
which they must struggle to bring harmony. The tasks they have to solve are
first and foremost set them by their own nature.

Alongside melancholy as the dominant element there has already been named a
power of thought, an ability for dialectical work, for dismembering
reflection. As he himself expresses it (1849): ``Bound in agonizing misery,
like a bird whose wings have been clipped, so am I, while I have yet
retained all the power of my spirit, that no doubt extraordinary power.''
Not only in reflection and thinking did this power express itself, but also
in an imagination that at every vibration, every flash of mood, could throw
out images in full splendor and with striking life, and in an art of
language such as had scarcely yet been seen in Danish literature. Favorably
situated in outward respects, he could develop these mental powers
according to his own pleasure. The use of these powers has, like the use of
every natural force, been bound up with pleasure; it was for him an
independent need to employ them, even if they were mainly employed in the
service of the prevailing moods.

In his work ``On My Work as an Author'' (\emph{Om Synspunktet for min
Forfatter\-virksomhed}) Kierkegaard attributes to himself, alongside ``the
enormous melancholy,'' an ``equally enormous facility for concealing it
under apparent cheerfulness and joy of life.'' By this remark he certainly
brings---as, on the whole, in the retrospect on his authorial activity---
more system and deliberateness into his conduct than there really was.
Alongside melancholy there lay an element of a wholly opposite nature, an
urge to give oneself over to all kinds of high-spirited play, to let
oneself go in wit and other trials of intellectual strength---an urge that
naturally connected itself with the urge to use his mental power. According
to the characterization his headmaster gave of him, after the old custom,
at his transition to the university, the childlikeness is said to have
persisted long in him; he had, it appears, a great love of freedom and
independence, and his character was lively and cheerful. The same
impression others got of him later in life. It cannot, at first, have been
dissimulation. There was in him an involuntary tendency in this direction, a
tendency that under a different upbringing might have been developed so as
to have a happy influence on his character. As it now expressed itself,
when melancholy withdrew, or as a contrast-effect of it, it could in fact
become a cover for the closed-off-ness of melancholy; but deliberately
produced it cannot have been. It was surely the same with this as with his
eager and constant roaming about the streets, which he later also presented
as part of a thought-out plan, but which at first too was an outflow of an
urge for movement, for people, and for changing impressions.

``I am a Janus bifrons---with the one face I laugh, with the other I
weep,'' he writes at 24 years of age. Neither of these faces was assumed,
even though the one came out most before others, the other in solitude. For
it is the nature of melancholy to withdraw when others are present, just as
the dwarfs vanish when the sun rises. This property of melancholy was
precisely what made it so fateful an element in Søren Kierkegaard's
personality. It was its working to make solitary, to produce isolation and
render openness toward others impossible, that became decisive for his
fate. The dull grief made him mute---but is it not a sin not to be able to
open oneself to those nearest one, not to be able to give oneself
inwardly? And do not, at any rate, the most intimate relations of communal
life become impossible where such an isolating melancholy prevails? When it
is told that Jesus drove out a devil, and it was mute---then Søren
Kierkegaard explained this as grief, melancholy, which in its egoism, in
order not to lose its dominion over the person, makes him mute, draws him
out of connection with others. And herewith there was now set for him an
enormous problem in his own inner life: was it insanity or sin that thus
possessed him? ``It is and remains the heaviest tribulation when a person
does not know whether the ground of his suffering is insanity or sin.''
Here was something that could well set the brooding dialectic in motion.
Streams of thought and series of images could develop from the one of the
possible points of view, only suddenly to vanish before new series of
representations from the other point of view---especially when there was
now added to this the emphasis produced by the strictly orthodox upbringing
and by the memory of how the father's melancholy had even led him to curse
God.

In \emph{Stages on Life's Way} the relation between father and son is
depicted: ``There was once a father and a son. A son is like a mirror in
which the father sees himself, and for the son again the father is like a
mirror in which he sees himself in the time to come. Yet they seldom
regarded each other thus, for the cheerfulness of a lively, animated
conversation was their daily intercourse. Only a few times did it happen
that the father stopped, stood before the son with a sorrowful
countenance, regarded him and said: Poor child, you are going about in a
quiet despair. Nothing further was ever said about how this was to be
understood, how true it nevertheless was.''---In the posthumous papers
Kierkegaard expresses himself far more definitely and sharply about how the
father ``unloaded all his melancholy onto a poor child,'' how his father
made him unhappy and robbed him of his childhood. ``The joy of being a
child I have indeed never had.'' And it is surely also grounded in his own
experience when he says: ``It is a rape, however well-meant it may be, to
force the child's existence into the decisive Christian categories''
(\emph{Concluding Unscientific Postscript}).

Kierkegaard's youthful years (from 22 to 25) have been illuminated by the
complete publication of his papers and by the investigations an energetic
Kierkegaard-scholar has undertaken on their basis.\footnote{\emph{P.~A.
Heiberg: En Episode i Søren Kierkegaards Ungdomsliv} (Kierkegaard Studier.
I) 1912.---\emph{Et Segment af Søren Kierkegaards religiøse Udvikling}
(Kierkegaard Studier III). 1918. (See my review of this book in
\emph{Tilskueren} 1919.)} It appears that there was a period in
Kierkegaard's youth that may be called the period of dissipation. He was
carried beyond the boundary of what he considered right and pure, and the
memory of this has ever since stood as a black spot in his life. When he
tried to explain to himself how such a fall was possible, he emphasizes the
demonically alluring power of foreboding and dread. The human being stands
before evil as before an abyss; a dizziness can seize him at the mere
thought of the possibility of casting himself out. Already in 1837 the view
of dread is expressed in the notes which he later developed in detail in
\emph{The Concept of Anxiety} (1844). It appears thus that there lies
personal experience behind the whole series of reflections on this concept
that are found in his works. And when in a note from 1837 he warns against
awakening in children, by allusions, suspicion, and untimely warnings,
representations of the possibility of evil, and thereby ``kindling the
tinder that is in every soul,'' he is surely thinking of his own upbringing
and of stray remarks of his father. When even the venerable figure he
looked up to with admiration for piety and firmness of character turned out
to suffer from inner unrest, perhaps to brood over secrets and in any case
to know terrible possibilities, this had to work on the son's mind like an
``earthquake.'' And the background of melancholy that in the father lay
behind the unrest and dread the son knew sufficiently from his own
experience; it was an inheritance that was increased by the strict
upbringing and by the father's remarks that the young man was going about
``in a quiet despair.'' It seems to have been in the year 1837 that Søren
Kierkegaard received confirmation of his foreboding as to what lay behind
the father's unrest. In his Old Testament conception the old man believed
that he had committed the sin that cannot be forgiven, because as a child
he had cursed God. Of the impression Søren Kierkegaard received of what he
had first only suspected, and then knew about the father, he has expressed
himself in a note from 1838.

\begin{quote}
``Then it was that the great earthquake occurred, the frightful upheaval
that suddenly forced upon me a new infallible law of interpretation of all
the phenomena. Then I suspected that my father's great age was not a divine
blessing, but rather a curse; that our family's excellent mental gifts
existed only to lacerate one another; then I felt the stillness of death
around me\dots\ A guilt must rest on the whole family, a punishment of God
must be over it.''
\end{quote}

Characteristic for Kierkegaard it surely is that the foreboding had for
him ``worked far more frightfully than the fact by which the truth of the
foreboding is confirmed,'' as he expresses himself in a note from January
1837. The foreboding and its confirmation put an end for him to the life in
speculation and in the mere spectator-relation in which he had an
inclination to place himself toward the world. ``The quiet despair'' could
no longer be kept down by such means.

His religious development had thereby, as by his whole temperament and his
whole upbringing, received an initial velocity that let his course run high
above the most traveled level roads. And the inner brooding to which he was
disposed now got a symbol to hold to. He had been led practically into
Old Testament religiosity in a way that falls to the lot of the fewest. The
patriarchal God-relation did not stand for him in the usual idyllic light.
And he was not among those who could make a great distinction between the
Old and the New Testament.

Søren Kierkegaard himself found his melancholy akin to the remarkable state
of feeling that has not been unusual in the monasteries of the Middle Ages,
and that was called \emph{Acedia}.\footnote{Cf.\ \emph{Den store Humor},
p.~96~f.} It expressed itself as spiritual slackening, disinclination for
religious exercises, longing for the world, and aversion to confession
(\emph{odium professionis}). He believed he found therein what his father
called the quiet despair. He interpreted the aversion to confession in a
wider sense: of a shyness about expressing oneself to others at all, and
believed he could then support the description with his own experience. He
also found that it betrayed a deep insight into human nature when the
medieval ethicists reckoned Acedia among the chief vices. It was a
monastery-sickness, and he also says occasionally: ``Had I lived in the
Middle Ages, I would probably have gone into a monastery.''---His
melancholy was in this different from the medieval Acedia, that it was far
more grounded in nature than the latter. Acedia arose, as it seems,
especially by contrast-effect, as a slackening after the strong mystical
exertion, a reaction after forced devotion---a «désenchantement de dieu,»
as a French philosopher has designated it. Yet Kierkegaard's state of
feeling surely also had this character in part. There was, in consequence
of the ``rape'' that early, by his own account, was practiced against him
(especially against those elements in his nature that pointed in a
direction other than melancholy), rich occasion for contrast-effect of the
kind mentioned.

\medskip

\noindent\textbf{4.} On the basis of melancholy---as the from the first
strongest, and by upbringing and the father's influence further potentiated
or inflamed, element of his nature---his engagement story now finds its
explanation, when we attend to the utterances in the posthumous papers.---
After he had led an aestheticizing and philosophizing youthful life, broken
several times by violent inner agitations, as at that ``earthquake'' and at
the father's death, he became engaged to a quite young girl, but broke off
the engagement the following year, to the great sorrow and offense of the
whole circle. He has depicted this episode in the third part of the
\emph{Stages}, and in the posthumous papers he returns to it again. It
seems to me that all necessary explanation lies in the often-varied
exclamation: ``Alas, she could not break the silence of my melancholy!'' He
had no happiness to share with her---on the contrary, it was first by
coming to know her bright, light temperament and immediate happiness that
he became rightly conscious of how unhappy a temperament his own was:
``Precisely her immediate youthful happiness, set beside my dreadful
melancholy and in such a relation, had to teach me to understand myself;
for how melancholy I was I had never suspected before; I really had no
standard for how happy a human being can be.''---As his high spirits and
power of thought could grow by the contrast to melancholy, so this again
grew when it got happiness as its counterpart. A web of actions and
reactions between the elements in his inner life and between his inner life
and his outer experiences, which with the power of consistency formed his
disposition in an ever more sharply pronounced style!

From what he calls his misery---the fact that ``the simplest conditions for
being a human being were denied him,'' the isolating, demonically inhibiting
melancholy---he derives not only the impossibility for him of entering into
a relation which, like marriage, rests on openness and inwardness toward
another, but also the impossibility of taking on an office. ``It has always
come easy to me to get on with people, and, accustomed as I am to being
bound, it never occurred to me that it should become so hard for me. But
here again comes my misery: I cannot, because I am not a human being,
because I am melancholy to the border of insanity---something I can indeed
gloss over so long as I am independent, but which makes me useless for a
position where I do not myself determine everything.'' Yet it was, according
to what the complete notes show, especially the black spot mentioned above
that stood for him as a hindrance, both toward the thought of marriage and
toward the thought of becoming a priest.

With admirable clarity and honesty Søren Kierkegaard gives an account of how
he came to be placed differently from other people. Had it been up to him,
he would have chosen marriage and social position, like so many others. It
was his sick temperament that made this impossible for him, while his
pecuniary circumstances on the other hand made it possible for him to live
(for a time---but it became his lifetime) without seeking paid work. It was
by no means the case that he kept from entering into any relation of
communal life on strictly ideal and principled grounds. That he became the
individual was due to the inner problems that, by his psychological
presuppositions, he got to struggle with. Only when, on
psychological-individual grounds, he was placed outside the ordinary way of
life of people, on a solitary post, did he there discover something that he
otherwise probably would not have caught sight of. What in his eyes stood as
the highest ideal showed itself to him with a clarity and a consistency
harder to reach for one who lives as a link in the manifold and conditioned
life of humanity. Thus from a lighthouse out at sea much is seen that cannot
be seen from the mainland; and yet it is not said that those who see it had
been drawn out there in order to see it. Melancholy became, so to speak, his
lookout station. Already in antiquity it was thought (it is so stated in the
so-called Aristotelian ``Problems'') that the melancholic temperament was
peculiar to geniuses. Even if by melancholy they did not then understand
quite the same as nowadays, this opinion may fit many cases. There are
experiences and discoveries made only by one who, with melancholy as a
weight beneath his feet, dives down into the sea of life.

The great personalities whose life and thought acquire decisive and typical
significance (whether now by the attraction or by the repulsion they
exert) are such as have great inner problems to solve, and who solve them
in a way that acquires value for others than themselves. When they come to
step forth actively and aggressively, the chief cause will lie in the fact
that the inner self-work needs an outer continuation. The sharp eye both for
the ideal and for what stands opposed to the ideal they have acquired in
their inner life. An inner drama has taken place before the outer one comes
to performance. What to contemporaries often looks like, or is by them
construed as, vanity, presumption, love of criticism, misanthropy---this is
in the really great the outflow of dearly bought experiences that they
cannot---without denying themselves---keep to themselves. Søren Kierkegaard
belongs decidedly to this class of personalities. That his inner life should
acquire significance for others through the clarity he could bring to some
of life's most important questions was his own conviction. In that
consisted, really, the faith he had in his mission. ``Very far back in my
recollection,'' he says (\emph{On My Work as an Author}), ``goes the thought
that in every generation there are a couple or three who are sacrificed for
the others, used to discover in frightful sufferings what comes to benefit
the others; thus I melancholically understood myself, that I was singled out
for this.''

\medskip

\noindent\textbf{5.} When Kierkegaard's melancholy, with its effects, stood
for him now as sickness, now as guilt, this surely came in part from the
fact that melancholy has a peculiar power to let the mind dwell on
possibilities it would otherwise not bring forward, and to give these
possibilities not merely a dark stamp, but a stamp of reality. This will
naturally happen all the more where there is a lively imagination and a
subtle dialectic to unfold the possibilities and work with them. Every
strong feeling has a tendency to spread in the mind and to give all emerging
representations partly an absolute character, partly a stamp of reality they
otherwise lack, just as it will also call forth recollections and
anticipations for its own explanation and clarification. (These phenomena I
have described, under the name of the expansion of feeling, in my
psychology.) Melancholy disposes to a life in possibilities---partly in
anticipation, partly in recollection. That Kierkegaard was well versed in
both is seen not only from his writings, where he both inculcates the great
importance of living in possibilities and testing oneself in them, and of
not confusing possibility and reality, but also from the posthumous papers,
which here too show how what he developed in his books was drawn from his
personal experience. With melancholy there worked here, naturally, the
poetic urge to image-formation, to the concrete development of the
representations that could serve as an expression of the moods, now a direct
expression of these, now, by contrast-effect, an expression of their
opposite. The following utterances show what power anticipation and
recollection had over him, and what danger he saw therein.

In the year 1839 (25 July) he writes in his diary: ``This is why I find so
little joy in existence: that when the thought of something awakens in my
soul, it awakens with such an energy, in so supernatural a magnitude, that I
really \emph{overstrain} myself upon it, and the ideal anticipation is for
me so far from being explanatory of existence that I rather proceed from it
too powerless to find what corresponds to the idea, too restless and, so to
speak, nerve-shaken to rest in it.''---One easily understands how melancholy
had to be increased by such constant re-living of the disproportion between
the anticipated and the real.

``For me,'' it runs in a note from about the same time, ``there is nothing
more dangerous than to recollect. Once I have recollected a life-relation,
it is impossible for me to get any interest in it again\dots\ Once I get time
to make the experience I have all too often enough made, that recollection
satiates more richly than all reality---then it is over.''---Here again it is
the life in representation that gets the precedence over the life in reality.
In the same direction points a note occasioned by a brother's death: ``I
have noticed that my grief is not one that apprehends momentarily, but one
that increases in the long run, and I am sure that when I some day grow old
I shall really come to think of the departed\dots\ As for my now deceased
brother, I am sure that the grief will first really awaken after a long
lapse of time.''

If melancholy, by its propensity to call forth and dwell in possibilities
and representations, gave the poetic urge and the dialectical drive wind in
their sails, there was thereby attained, on the other hand, a diverting
effect on melancholy; the latter was kept down by the work of thought and
imagination, which laid claim to the whole force and attention of the mind.
The whole first grand period of Kierkegaard's production (1843--46), in
which his most brilliant works fall, he himself has explained from this
point of view: ``Like that princess in the \emph{Thousand and One Nights} I
saved my life by telling stories, i.e.\ by producing. To produce was my
life. An enormous melancholy, inner sufferings of a sympathetic kind---all
this I could master when I was allowed to produce.''---Had there not been
this deep urge to work of thought and imagination, he would have gone under
spiritually.\footnote{The constantly repeated rhythmic play of melancholy
and thought-movement is depicted in the following note from 1843: ``It is
strange how strictly, in a certain sense, I am brought up. From time to time
I am set down in the dark hole; there I crawl about in torment and pain, see
nothing, no way out. Then suddenly a thought awakens in my soul as vividly
as if I had never had it before, though it is not unknown to me; but before
I have, as it were, only been wedded to it with the left hand, now I am
wedded to it with the right. When it has now settled fast in me, then I am
caressed a little, I am taken up in the arms, and I, who had shriveled up
like a grasshopper, now grow up again, healthy, thriving, glad, warm-blooded,
supple as a newborn. Thereupon I must, as it were, give my word that I will
pursue this thought to the utmost; I set my life in pawn, and now I am
harnessed in the traces. Stop I cannot, and my powers hold out. Then I am
finished, and now everything begins again.''} It was an independent element
in his nature, which under other circumstances could have been developed in
other forms than it was. It sought an outlet, and the direction of the
outlet was determined by what else stirred in his mind. But the whole
authorial activity was hardly, from the first, laid out so clearly and
definitely in one direction or with one purpose as he later believed. There
was from the first far more that was involuntary in him than he would admit
when he looked back on his activity.\footnote{The relation between the
involuntary and the deliberate in the design of Kierkegaard's production in
those three years I have discussed in \emph{Mindre Arbejder}. Second
Series, pp.~181--191.}

\medskip

\noindent\textbf{6.} If we stopped at the elements of his being hitherto
brought out, we could not expect to find in him more than---what is
naturally no small thing---a poet-nature, and perhaps also a thinker of a
psychological cast. Søren Kierkegaard has also often enough warned against
himself in this respect. When it was his conviction that people, especially
in his time, all too often confused a thought- and imagination-relation to
life and to life's great exemplars with a relation of reality, he knew, as
has been shown, this danger from himself. But he also knew the
counterweight. For just as characteristic of him as the strongly tensed life
of mood, the tireless dialectic, and the imagination obedient to every
nuance of feeling and every transition of thought, was the energetic will
with which he sought to preserve himself in concentration, to hold together
the elements of his being and lead all the streams of the rich inner life
toward one goal. He has with full right maintained that at no stage was he
merely a poet or a thinker. Already as a young man of 22 he wrote (at
Gilleleje, on a walking tour in the summer of 1835) the following:

\begin{quote}
``What I really lack is to get clear in my own mind about \emph{what I am to
do,} not about what I am to know, except insofar as a knowing must precede
every action. The point is to understand my destiny, to see what the Deity
really wants \emph{me} to do; the thing is to find a truth that is truth
\emph{for me,} to find \emph{the idea for which I am willing to live and
die.} And what use would it be to me if I discovered a so-called objective
truth\dots\ if it had \emph{no} deeper significance for \emph{myself} and for
\emph{my} life?\dots\ That was what I lacked: to lead \emph{a completely
human life} and not merely a life of knowledge, so that I should thereby
come not to base my developments of thought upon---yes, upon something one
calls objective---something that is at any rate not my own, but upon
something that is bound up with the deepest root of my existence, through
which I am, so to speak, grown into the Divine, cleave fast to it even if
the whole world should collapse. See, \emph{this is what I lack,} and
\emph{this is what I strive toward.}'' (The emphases are Kierkegaard's own.)
\end{quote}

The strong urge to think is thus from the first an urge \emph{to think in
existence,} to think in such a way that the innermost personality is
involved and is determined thereby in its core and its direction. He wants
to think in the same manner as, in one place, he writes that he has read a
book (Görres's \emph{Athanasius}): ``with my whole body---with the pit of my
heart.'' But this urge to bend everything in himself toward one purpose, to
join the separated, conflicting elements of his being together into unity,
had strong resistance to overcome. And yet the thinking of the age seemed to
have paved the way. The watchword, after all, was the higher unity, the
harmony of oppositions, and all around him he heard young speculative
virtuosos repeat these slogans in full assurance of being able to make good
on them. The thing was to mediate, to find the mediation; to remain stuck in
an either--or, in an extreme or an unreconciled relation of opposition, was a
sign of a poor head or of a narrow-minded holding-fast to a ``superseded
standpoint.'' These phrases had to grate on Søren Kierkegaard's ears when he
came from his solitary meditations or from his inner struggles, which taught
him well enough that here there was no question of a method, a knack, that
could be learned once and for all, but that it was something which, when
attempted in the reality of life---not merely in fantastic abstraction or on
paper---could demand the most strenuous willing, the most painful endurance,
and that it was a task that always had to be solved anew. But---as he wrote
in his diary in 1837---``there are many people who arrive at a life-result
the way schoolboys do; they cheat their teacher by copying the answer out of
the arithmetic book without having worked the problem themselves.''
Kierkegaard was here the strict auditor---first and foremost toward himself.
He lived through the elements of life and learned from his own experience
how hard a work it can be to construct a totality of them. A superficial
polishing-up so often covers over a botched job. He therefore declared
``the higher unity,'' with its both-and, a struggle for life and death. The
distinction, the disjunction, the either--or was given a place of honor, and
``repetition'' was set in the place of mediation. ``All talk of a higher
unity that is to unite absolute oppositions is a metaphysical assault on
ethics.'' One presupposes without more ado that everything is in order, and
that the elements join together as of themselves, so that no special act of
exertion, no new effort is necessary or must always be repeated. But often
this comes precisely from one's having no real natural ground in oneself.
For his own part he says: ``I sit and listen to the tones within me, the
glad hint of the music and the deep earnestness of the organ; to work them
together, that is a task not for a composer, but for a \emph{human being}
who, in default of greater demands on life, restricts himself to the simple
one of wanting to understand \emph{himself}.---To mediate is no art when one
has no moments in oneself.''

As life went on, the oppositions became for him stronger and stronger, the
moments more and more contentious, although his power grew at the same
time---or perhaps rather because his power grew; for had it not grown (even
to sickly overstrain), he would not have been able to hold fast such great
disharmonies and operate with them. All rational and natural coherence in
life was in the end burst asunder for him, and he passed a sentence of
condemnation on the natural life of man as corrupted and depraved in its
innermost root.

For it must indeed be well noted, and will in the following be more closely
demonstrated, that Søren Kierkegaard's indignation and polemic were most
immediately called forth by the romantic-speculative blurring of life's
oppositions---a blurring that, especially in the mouth of epigones and
parrots, became insipid and superficial---but that it was not merely such a
passing tendency of the age or of fashion that he wanted to do away with. He
aimed higher. What he wanted to do away with was really all human and
natural view of life and conduct of life, faith in life's unbroken power, in
its ability by its own natural ways to solve the problems its own unfolding
has brought with it. In itself the idea of ``the higher unity'' is also only
a modernized expression for the harmony that Greek thought in its way sought
and found, and by which no human thinking can refrain from steering, even
though it is of great importance for the richness and depth of life that one
not bargain away the oppositions, and that one not too early and on loose
ground celebrate the festivals of ``mediation.''

Kierkegaard also became ever more clearly conscious of what kind of power it
really was that he was fighting against. When the time of ``the system'' was
past, his struggle was not past. He saw clearly that the scientific view of
the world, by its naturalistic path, asserted the same continuity and inner
coherence in life that the speculative view had, in its idealistic manner,
maintained. It was the same enemy in a new shape.---Yet it was a pity that
Kierkegaard did not have before him greater representatives of humanism than
the speculative epigones. He did not have enough resistance to overcome. His
work therefore did not become \emph{as} significant as it otherwise might
have become.---

The higher unity was, according to Kierkegaard, so difficult to reach not
merely for the reason that there are sharp oppositions, conflicting moments
to overcome, but also because all human existence is in a state of
\emph{becoming}. Existence means, according to his definition expressly: to
be \emph{in time}, and on this last he lays (as we shall see more fully in
his theory of knowledge) a very great weight. So long as we are in time, we
stand ever before new possibilities and new tasks, whose solution is
problematic. All our knowledge and harmony is grounded on hindsight. ``Life
must be \emph{understood backwards}. But\dots\ it must be \emph{lived
forwards}.'' Which proposition, the more it is thought through, ends
precisely in this: that life in temporality never becomes rightly
intelligible, precisely because at no moment can I get complete rest to take
up the position backwards. Every won result must be ventured anew as a
stake, put anew up for debate, in order if possible to be won anew under the
constant development and the constant work. Here the concept of
\emph{repetition} shows itself in its significance for Kierkegaard's
thinking. It is closely connected with his emphasis on reality, on existence
as being in time. What has been won in the past has only a \emph{possible}
value in the future. The point is a conversion of this possibility into
reality; but every such conversion is a repetition, since what is to be
realized existed beforehand as a possibility, for representation. (To the
concept of repetition we shall return in connection with Søren Kierkegaard's
ethics.)---

\medskip

\noindent\textbf{7.} The relation between Søren Kierkegaard's poetic gift
and his gift as a thinker must still be touched on. The two sides in his
nature work constantly together. This places him in a position of his own in
literature, outside the usual rubrics. In a certain sense he might perhaps
have achieved a greater effect if he had been able to give poetic works that
did not require an understanding of philosophical thoughts, and works of
thought that did not present so many parentheses, elaborations, and images.
Many who would enjoy poetry by itself, or follow pure and rigorous thinking,
now pass him by. But the close cooperation of his poetizing and his thinking
sprang from the close connection of both with his own innermost nature. They
both work in the service of his personality. His urge to think was an urge
to think in connection with life-experiences and moods; therefore he
constantly needed concrete situations and examples. And the mood, which
constantly accompanies, then undertakes its little excursions, has its side
effects, to whose unfolding and clarification the ever-willing and
lightning-quick imagination lends its aid. Hence the richness and
poetry---in short turns and expressions, after parenthetical outbursts, or in
elaborating images---that is scattered about in his works.

``My thinking,'' he himself says, ``is \emph{presentic}; I have as much
imagination as dialectic.'' That he did not lose himself in abstractions
remote from life and the present he here aptly connects with the fact that
imagination and thinking work together in him. Since real experience seldom
or never presents us with a character, a passion, or a state of soul in full
and consistent development, he takes imagination to his aid and
constructs---by a thought-experiment that strictly holds to the decisive
conceptual determinations that are to be practiced---figures and situations
for the visualization and contemporization of the spiritual conditions and
problems he is occupied with. His most famous books---\emph{Either--Or},
\emph{Stages on Life's Way}, \emph{Repetition}---are such ``Attempts in
Experimenting Psychology.''---Yet I believe that in this respect too there is
more involuntariness and less systematic design in his authorial activity
than he often himself gives the appearance of. Images have surely thrust
themselves forward before their use, and he has---at any rate most
often---only afterward seen what they could be used for. And where the
constructing deliberateness has intervened more strongly, it has not
seldom been to the detriment of the poetic value, and also (as will be
sought to be shown in a later section) of the philosophical value of his
expositions.

Kierkegaard lays much weight on the need to test oneself through
possibilities, since the reality of life is not sufficient to educate the
character. But it is characteristic of him that he emphasizes only one
drawback of this method of possibilities, namely that one can easily shirk
one's way to a greater proficiency than one really possesses. ``It is
difficult to test oneself in possibility; it is as if someone, without
daring to use his voice, were to test whether he had a strong voice.
Hitherto in vain I have many times speculated on finding a means of being
able to check myself in possibility'' (\emph{Stages on Life's Way}). He does
not always bring out the other conceivable drawback: that the possibilities
constructed could be so dark and abnormal that strength and courage are used
up in the struggle with shadows, without any proficiency for taking up the
struggle of reality being fetched from this dwelling in the world of
possibilities. And moreover: with reality one can at times get finished; but
possibility is like a shadow or a fog that forms again as often as it is cut
apart; the struggle with possibilities therefore easily becomes a bailing
into the vessel of the Danaids. Kierkegaard has here worked far more as a
romantic than he wished and willed. He has for many conjured forth an
imaginary world that sucked all the strength and sap out of the real one.
There are several for whom his writings have been like the lion's den, where
all tracks led in, none back out again. Others have had to free themselves
violently from his world of possibility by one of those resolute ``jerks of
decision'' that he himself in another connection recommends.

So great was his zeal for personal truth that he expressly wanted it
established that these thought-experiments did not cover his own view.
Therefore he poetized not only characters and situations, but also authors.
The pseudonymous author-names that stand on most of his writings make up a
whole circle of individualities at different standpoints. It is his striving
after ``thinking in existence'' that leads him to these pseudonyms. What
virtuosity he possessed in thus psychologically and poetically entering into
different possible author-individualities can be seen from the little work
\emph{Prefaces} (\emph{Forord}), a collection of prefaces to various imagined
writings by imagined authors.---

His thinking is not only \emph{present,} bound to perceptible conditions and
situations; it is also, what is moreover closely connected therewith,
\emph{subjective}, that is: a thinking that proceeds from, and is consumed
in, the personal life, in the view of life and the conduct of life of the
individual, in the practical relation to life's circumstances and life's
possibilities. Such a subjective thinking is bound up with passion, for to
exist is an enormous contradiction. The subjective thinker must not swing
away from existence into abstraction; his task is rather to understand the
abstract and the universal concretely, in his existence as this individual
human being. To be the ideal in imagination or on the lecturer's platform is
not difficult; but to have to exist as an idealist is an extremely
strenuous life-task. The subjective thinker is an artist, not a man of
science, for his thinking goes together with a willing.---As such subjective
thinkers Kierkegaard especially admired Socrates and Lessing, also Jacobi
and Hamann. His doctoral dissertation dealt with Socrates (\emph{Om Begrebet
Ironi med stadigt Hensyn til Sokrates}, 1841). Both Socrates and Lessing he
has characterized in a brilliant, but also highly one-sided way; in
particular he has surely made them more ``subjective'' than they in reality
were.

It is an important idea that Kierkegaard has set up in his demand for
subjective thinking. It contains not merely the demand to leave the paths of
speculative abstraction and go back to critical reflection on the
presuppositions and barriers within which our thought has to work. But it
leads him to give a psychological and ethical introduction to a view of
life, or a theory of the art of the view of life. Søren Kierkegaard's
philosophy is such a theory. It is, to be sure, first and foremost designed
to illuminate the Christian view of life; but it is designed and carried
through in such a way that it can acquire significance for every view of
life.

Søren Kierkegaard's authorial activity falls essentially into two periods.
In the first (1843--46) he is occupied with the struggle against the
speculative-aesthetic evaporation of existence, with demonstrating
oppositions, transitional relations, and conditions in the domain of the
views of life. He here gives a kind of comparative philosophy of life. In
the second period (1847--1855) he carries on his great struggle against the
Church's mitigating conception and treatment of Christianity. This last
period too acquires significance in philosophical respects, since
Kierkegaard here draws his last consequences with regard to the relation
between religion and humanity.---In the characterization of the first period,
however, utterances from a later time may of course very well be used from
time to time, provided only that they are expressions of thoughts that were
already developed in the earlier time.

% ============================================================
%  IV.  SØREN KIERKEGAARD'S PHILOSOPHY  (pp. 54--126)
% ============================================================
\bookchapter{IV.\quad Søren Kierkegaard's Philosophy}
\label{ch:4}

The task Kierkegaard set himself as a thinker he has himself designated
as this: to make difficulties. Everything was being made easier nowadays,
he thought. In the outer world, traffic and intercourse were made easier
by railways and telegraphs; and in the world of spirit there was promised
a reconciliation of all oppositions, a harmonious unity of all conflicting
elements. So he set himself, then, to make something difficult;---for
otherwise those who wanted to make everything easy would come to lack
material in the end.

Of the men here at home by whom he was influenced during his development,
there has been talk in the foregoing. Here it shall merely be mentioned
that it was especially Martensen's lectures on speculative theology that
provoked his criticism. Already in the dissertation on Socrates there
appeared hints of a more critical, subjective, and existential conception
of thinking, and a doubt whether it should be possible to give our
knowledge as smooth and continuous a coherence as speculative philosophy
believed it had attained. When Schelling in the year 1841 began the
sensational lectures with which he wanted to set a new, higher and more
``positive'' philosophy in place of Hegel's, Kierkegaard was among the many
who hurried from all around to Berlin to become acquainted with the new
wisdom. In the beginning it seemed as if the expectations with which he
came would be fulfilled. In his diary he wrote: ``I am glad to have heard
Schelling's second hour---indescribable. I have sighed long enough, and
the thoughts have sighed within me; when he mentioned the word
`\emph{reality},' concerning philosophy's relation to reality, then the
child of thought leapt within me for joy as in Elizabeth. I remember almost
every word he said from that moment. Here perhaps clarity may come. This
one word reminded me of all my philosophical sufferings and torments.'' The
expected greater clarity failed to come; for some months later Kierkegaard
wrote to his brother: ``Dear Peter, Schelling talks the most utter
nonsense,'' and he traveled home before the lectures were at an end. Yet
that leap of the child of thought was the beginning of his subsequent life
of thought. For Schelling emphasized very strongly that speculative
philosophy could not get beyond the possible and the abstract, the
universal, and that the relation to the absolute reality---especially as
religious faith conceives it---could be posited only by an act of will,
motivated by longing and by practical, personal need: thus, as Kierkegaard
later expressed it, by a leap. The opposition between thought and
existence, between the universal and the individual, and the impossibility
of a continuous transition between them, Kierkegaard here got established
so firmly that he did not forget it. He was rightly repelled by the
arbitrary and fantastic way in which Schelling philosophized further after
having established that break in the continuity of thought, although amid
the fantastry there emerge flashes of genius. Schelling's concluding result
would hardly have appealed to him either; it aimed at healing the break
again by demonstrating a new coherence and by announcing a religiosity of
the future, ``a third kingdom'' (like Lessing earlier, and in our day
Henrik Ibsen), to which Christianity is only an introduction, and which in
its higher form is to lead beyond the opposition between mythology and
revelation through free religious self-understanding. Kierkegaard held to
the break that Schelling had established. In his philosophical writings
(especially in \emph{The Concept of Anxiety}) one can clearly notice
Schelling's influence, in part also his terminology.

A younger German philosopher to whom Kierkegaard owes much is Adolf
Trendelenburg. During his Berlin stay mentioned above he was so taken up
with Schelling and the Hegelians that he did not get to hear Trendelenburg,
which he later regretted. Trendelenburg was the first to give an
epistemological critique of the speculative method and to develop a more
critical theory of knowledge (\emph{Logische Untersuchungen}, 1840), which
has had no small influence on Kierkegaard's cast of thought. Kierkegaard
speaks of Trendelenburg as ``a man who thinks soundly and is happily formed
by the Greeks,'' and in his posthumous papers he even says: ``There is,
after all, no modern philosopher from whom I have had the benefit that I
have had from Trendelenburg.''

But whatever he owed to others, the essential ground was laid in his own
personality and in the original nature and direction of his thought.
Through solitary brooding and reflection his thoughts soon developed to
full maturity. After his return home from Berlin he lived in his studies
and his immensely productive authorial activity, from which he diverted
himself by walks, by excursions to the wooded regions of North Zealand, and
on a single occasion by a new journey to Berlin. By his roaming about the
streets and lanes he soon became a well-known Copenhagen figure, just as
Socrates was in Athens, and he too found his Aristophanes. He needed
movement for the sake of his health. But he also felt an urge to seek
relief for his melancholy and his sorrow by talking with common folk and
entering into their circumstances, helping and consoling them. He learned,
moreover, much both as a psychologist and as an author by hearing plain
people talk together: ``What one has vainly sought enlightenment about in
books, a light suddenly dawns on it by hearing one servant-girl talk with
another; an expression one has vainly tried to squeeze out of one's own
brain, sought in vain in dictionaries, even the Academy's, one hears in
passing: a common soldier says it, and he does not know what a rich man he
is. And as one who walks in the great forest, wondering at the whole,
sometimes tears off a branch, sometimes a leaf, now listens for a bird's
cry: so one goes among the crowd of people, sees now an utterance of a
state of soul, now another, learns and learns and only becomes more eager
to learn. Then one does not let oneself be deceived by books, in which the
human occurred so seldom; then one does not read about it in newspapers;
the best in the utterance, the loveliest, the little psychological trait, is
nevertheless sometimes not preserved'' (\emph{Stages on Life's Way}).---As
the deliberate everywhere comes after the involuntary, it was only later
that he conceived this roaming about the streets and lanes as a tactical
means: in order that the religious authorial activity, which was his real
task, might come to count by its own force, he wanted to weaken the
impression of his person. The technique of indirect communication required
this! We need not here enter more closely into this artifice. I have merely
wanted to sketch his circumstances of life in the time (1843--46) in which
his properly philosophical authorial activity falls.

% ------------------------------------------------------------
\section*{A.\quad Epistemology}
\label{sec:4A}
\markright{IV. Søren Kierkegaard's Philosophy --- A. Epistemology}

\noindent\textbf{1.} When Kierkegaard considers the question of what
knowledge an existing human being can attain, he is not thinking of
every possible kind of knowledge, but only of what he calls
\emph{essential} knowledge, by which he means the knowledge that bears
specifically on the knowing individual itself in its conditions of
existence---ethico-religious knowledge.\footnote{The main work here is
the \emph{Concluding Unscientific Postscript} (1846).} All other
knowledge---empirical,
mathematical, and historical---is a matter of indifference to him,
because it has nothing to do with existence. Already here he makes a
very important distinction, introduces an Either--Or. For it is surely a
great question whether much of that knowledge he regards as indifferent
is not---and has not been---of far-reaching significance for
ethico-religious knowledge. When the latter is not the same at all
times, this is due in part to that apparently ``indifferent'' knowledge,
which consciously and unconsciously comes to exert influence even on
those regions of spiritual life that one seeks to close off from it. But
Kierkegaard simply takes the correctness of that distinction as given,
and thereby marks off the boundary of his investigation.

\medskip

\noindent\textbf{2.} The train of thought in Kierkegaard's epistemology
is directed polemically both against Hegel and against those who (like
Martensen) held that once one had placed oneself on the ground of faith
and had one's consciousness reborn, one had thereby obtained the
conditions for a higher, hitherto unattainable knowledge. Against Hegel
it is observed that he has not really answered Kant's objections to a
knowledge of absolute reality. Thought cannot reach reality: for in the
very moment it would have reached it, it has converted it into
\emph{thought} reality, or into possibility; thought cannot get beyond
itself. Against speculative theology it is observed that faith has no
interest in understanding itself in any way other than by continually
remaining in the passion of faith. Of the highest things in the world
one can only be knowing as a happy or as an unhappy lover; by
approaching with speculative curiosity one in fact learns nothing. Take
away the passion that is ever renewed by uncertainty as a salutary
disciplinarian, and faith vanishes at once. The existing individual is
always in the process of becoming, and time is never concluded;
therefore full certainty can never be reached, and our knowledge can
never get beyond approximations and hypotheses, with which in the
domains of inessential knowledge we can also perfectly well content
ourselves. But if the Eternal and Unconditioned is to be grasped and
held fast, this can only happen in passion, because closure cannot be
attained. The craving for something finished and concluded is of the
Evil One, owing to indolence or impatience. We do not get beyond the
eternal striving after truth.---Kierkegaard joins with enthusiasm
Lessing's famous words: that if God held in his right hand all truth,
and in his left the ever-living striving after truth, he would grasp the
left hand, because eternal truth is not for a finite being. Lessing's
justification for this choice is, however, somewhat different from
Kierkegaard's. Lessing proceeds from the premise that it is not
possession, but only acquisition, that gives a human being worth,
inasmuch as one's powers are enlarged by seeking and one thereby grows
in perfection. This is a more positive and psychological justification
than the one we shall find in what follows from Kierkegaard, who is too
much the ascetic and too preoccupied with preparing his doctrine of the
Paradox to follow Lessing on his natural path.

More precisely, Kierkegaard's epistemology develops through the
grounding of two propositions: a logical system can be given; but a
system of existence cannot be given.

\medskip

\noindent\textbf{3.} A logical system can be given. We can develop a
coherent series of general determinations that hold for our thought. But
here one must be careful not to smuggle in concepts that are drawn from
reality. The logical system is a hypothesis whose validity for reality
is problematic. And it must be investigated in particular whether we
form this hypothesis after having come to know reality through
experience, that is, as a kind of abbreviated expression
(\emph{Abbreviatur}) for it, or whether we form it independently of
experience.---Kierkegaard expresses the wish to treat this last question
more fully on another occasion; but it never happened. It is, as one
easily sees, the question of whether the empiricist or the rationalist
conception of the fundamental presuppositions of thought is the correct
one.

A peculiar difficulty in the construction of the logical system lies,
for Kierkegaard, in how one can come to \emph{begin} it, i.e.\ get hold
of a starting point. For the simplest presuppositions are discovered
by reflection, by going back from derived assumptions to the assumptions
on which they rest---but when we have gone back far enough, how can we
halt the reflection? It has just as little reason to halt at one point
as at another, and will therefore, left to itself, continue
indefinitely. Or, as it can also be expressed: doubt cannot halt
itself. There must be a break, an arbitrary interruption. A decision
must be made to halt the regressive chain of thought and to fix the
point one has reached as the principle and starting point of the system.
But what then becomes of presuppositionlessness? For the beginning of
the logical system is then conditioned by something other than the
logical, namely by that which causes me to make the decision!

``What if, instead of talking or dreaming about an absolute beginning,
we were to speak of a \emph{leap}?''

---Against Hegel, who wanted to give an absolute system of thought that
closed upon itself like a circle, this line of reasoning by Kierkegaard
is telling. If one compares it with the superficial catechism-talk with
which Martensen ``went beyond Hegel,'' one has the difference between
them as thinkers given in a single example.---Yet it is a question
whether there is as great a difficulty about the ``beginning'' as
Kierkegaard thinks. We do indeed find our ultimate presuppositions by
reflection, or as Descartes expressed it, by doubt, or as one can
perhaps best express it, by analysis. But in our search for
presuppositions we always have a definite task in view, which is to be
solved with the help of the presuppositions we seek. The task is to
understand existence as given in experience, or some part of it. We
therefore have a definite standard for how far we need to go---namely
until we have found what is necessary for solving that task. If we can
go still further, we discover more remotely situated presuppositions
that present themselves with the same logical necessity as those first
found; but for the time being there will be no use for these
presuppositions. Perhaps advancing experience may lead us to make use of
them.---It is therefore no arbitrary act that forms the beginning of
logic, and its first principles have more than merely arbitrary validity.

\medskip

\noindent\textbf{4.} A system of existence cannot be given. This is due,
first, to the fact that all existence is in time, in the process of
becoming. The latter point especially must not be forgotten. Speculative
philosophy, which has claimed to explain everything, has forgotten that
the one who constructs the speculative system, however insignificant he
may be (``such a tiny little thingamajig as the existing Mr.\
Professor who writes the system''), still himself belongs to the
existence that is to be explained, and he is never finished with his own
existence. The system would only become possible if one could look back
on existence as concluded---but that would presuppose that one no longer
existed! As we have already heard: life is lived forwards, but
understood backwards.---And even that backward understanding, the
insight into the necessity of past development, may rest on an illusion:
for how can that which, so long as it is future, is only possible,
become necessary by belonging to the past? Should the past be more
necessary than the future? We understand the past properly only by
making ourselves contemporaneous with it---but then we see it as
becoming, not as concluded, and thus still for the most part merely as
possible! (See on this point \emph{Philosophical Fragments}.)

For the existing individual, moreover, thought and being are not one but
separate. We think either back on what has been or forward to what will
come. Simultaneity, absolute unity of thought and being, is impossible
for an existing being, and exists only as a fantasy. There can be no
continuity in our thinking, since it is continually interrupted by
absorption in existence. All unity and continuity is only abstraction.

And finally: only the individual, the concrete, exists; to exist means
to be individual. But only the universal and the abstract is the object
of thought. Herein too lies a disproportion between thought and
existence.\footnote{Before Kierkegaard, Sibbern (\emph{Hegels Filosofi
betragtet i Forhold til vor Tid}, 1838) had already criticized Hegel's
speculation from similar points of view. (See \emph{Danske Filosofer},
pp.~109--112.)}

Because a system of existence cannot be formed by our thought, it does
not follow, according to Kierkegaard, that such a system does not exist.
A system of existence is given---but only for ``one who is himself
outside existence and yet in existence---who in his eternity is
eternally concluded and yet includes existence within himself; that is,
God.'' (Since Kierkegaard defines existing as becoming, he consistently
says: God does not exist, since He is eternal.)---But for human beings,
existence is irrational, is a Paradox. The Paradox, the absurd and
contradictory, is according to Kierkegaard no concession but is, as he
puts it, an ``ontological determination,'' designating the relation
between an existing knowing spirit and the eternal truth.

It must be admitted that a Paradox has been established through the sharp
contradiction between the eternal truth (the eternal completedness of
existence in God) and the existence that is constantly in the process of
becoming. For how can, from any point of view whatsoever and for any
thought of which we can form a concept, that which is constantly
becoming stand as complete? Kierkegaard maintains firmly that ``motion
cannot be thought \emph{sub specie aeterni}'' [i.e.\ from the point of
view of eternity, apart from time]; but how then can God be knowing
about existence, about the whole of existence in its process of
becoming, since indeed ``existence cannot be thought without motion''?
---Here two thoughts have been violently yoked together, both of which
Kierkegaard demands be held fast. Epistemology must surely here turn his
favourite determination, Either--Or, against him and say: either the
concept of the absolute system in God must be an illusion, or else the
continual becoming must be a mere appearance.---Kierkegaard would not
need, if he chose the first alternative, to give up his theistic
presuppositions. Like other philosophical theists he could draw the
consequence that God himself must be in the process of becoming, that
God's being is not concluded. This idea was developed simultaneously
with Kierkegaard's philosophical production by the true founder of
modern philosophical theism, C.~H.~Weisse. Thereby the sharp antithesis
would be resolved---but the absolutely concluded element in the idea of
God would admittedly be sacrificed at the same time.---If one holds to
the view that the existence we know unfolds in time, and that we cannot
form any conception whatsoever of any other form of existence, then
there is no question of a concluded system. But existence then becomes
no Paradox either. It becomes irrational in the mathematical sense,
inasmuch as only an approximate knowledge of it becomes possible. It
stands before us like $\sqrt{2}$, which we can continually determine
with more decimal places without exhausting it. Existence could be so
rich and many-sided that there would be material enough in it for the
most energetic and comprehensive thought we can conceive. But that is
something entirely different from its being paradoxical, that is to say,
from its presenting us with insurmountable contradictions. The
inexhaustibility of existence for thought corresponds exactly to
Lessing's idea of the eternal striving after truth. If we hold fast to
this idea, we must as its objective counterpart assume a continually
becoming, and in its becoming unsurveillable, existence.\footnote{What is
here hinted at, I have developed in detail in \emph{Den menneskelige
Tanke}, pp.~306--375.}

Kierkegaard proceeds from the premise that we cannot, even with respect
to ``essential'' knowledge, content ourselves with approximations and
hypotheses. But why ever not? Provided they are genuine approximations,
and provided they are hypotheses that are increasingly verified by
experience? Then we have reason enough not to abandon our hope of
gaining clearer understanding and of being able practically to hold our
own in the struggle with the world. And it is such a hope that sustains
all our drive for knowledge.---But it will become apparent in what
follows that Kierkegaard treats the concept of approximation in a
remarkable way.

\medskip

\noindent\textbf{5.} That the ``essential'' truth is paradoxical is
grounded by Kierkegaard in the fact that the thought of God as the
absolutely eternal must be held fast in the midst of the unceasing
becoming of existence. On this depends also the fact that essential truth
can be held fast only with the passion of faith.

If one asks how the idea of God arises for Kierkegaard, his answer is
that God is a postulate, a last resort, without which we cannot endure
to live. An image that would aptly express his thought would be the
following: God is an anchor of distress that we need on the wild sea
of existence, and at which, amid the surging waves, there is a constant
tugging. We need an absolutely fixed point of support, a hook on which
the chain of existence can be fastened, a point where the infinite
reflection can find rest. But what is postulated is qualitatively
different from the postulating being: not even the concept of existence
is supposed to apply to God! God does not, according to Kierkegaard,
stand in relation to the human being as a superlative ideal. ``Between
God and man there is an absolute difference''; ``the qualities are
absolutely different.''---But then the situation is such that the last
resort to which the human being in existence anxiously flies itself
creates new difficulties and increases the suffering: for now the task
is to unite the absolutely different qualities, to think them together!
---Faith in God arises when the human being ``with infinite interest
inquires after a reality other than its own''---but thereby new problems
are immediately posed.

One understands that Kierkegaard, given his conception, could not have
it as easy as Martensen in building speculative theology on the ground
of faith. This ground was of a volcanic nature, was in constant motion,
just as it itself owed its origin to a revolution. Here there was
neither time nor place to build, but only to struggle anxiously to
preserve what already was.

An objective certainty, a confirmation by experience, can never be
obtained. Not only is the future, as has often been remarked, ever
equally uncertain, but when one contemplates nature in order to find
confirmation of God's omnipotence and wisdom, one finds alongside
traces in that direction also traces in the opposite direction. From
objective uncertainty, then, we never escape---and since it is
nevertheless a matter of life and death, this objective uncertainty is
embraced with infinite passion. Truth can be held fast only in the form
of subjectivity, as the object of personal feeling and passion.

\medskip

\noindent\textbf{6.} After Kierkegaard has thus developed the claim that
truth can be grasped and held fast only in a subjective, personal
manner---that truth is subjectivity---he reverses the proposition and
maintains: subjectivity is the truth. For only that which is grasped
with subjective energy and passion can be the truth. If there are two
people, one of whom prays in personal untruth to the true God, while
the other ``with the whole passion of infinity'' prays to an idol, then
the former in reality prays to an idol, while the latter truly prays to
God. What matters first and foremost is a \emph{how}, not a \emph{what};
personal inwardness and energy, not the content to which this inwardness
attaches itself. And with objective uncertainty, with the contradiction
that the content presents to the one who must hold fast to it in
personal existence, inwardness and passion increase. Truth is a venture.
To the proposition that subjectivity is the truth corresponds the other
proposition that truth (objectively considered) is the Paradox.

With this famous proposition of his, that subjectivity is the truth,
Kierkegaard has sharpened Lessing's proposition that only in eternal
striving can one relate oneself to truth. The entire value is shifted
from the objective (dogmatic) to the subjective (psychological) pole.
The meaning is really this: only when truth is appropriated and held
fast in personal feeling and striving does it have value. It is the
personal nutritive value of truth that is decisive. The standard lies in
the movement that is awakened in the inner life of the personality.

In fact, Søren Kierkegaard has in his own way stated the same thing that
Ludwig Feuerbach had maintained a few years earlier in \emph{The Essence
of Christianity}: that theology is psychology. (\emph{The Essence
of Christianity} appeared in 1841, the \emph{Concluding Unscientific
Postscript} in 1846. In this book, and already in \emph{Stages on
Life's Way}, Kierkegaard expressly refers to Feuerbach's interpretation
of Christianity.) Only through an inconsistency can Kierkegaard demand
a definite content, a definite objective faith, of subjectivity: the
main thing is that it be in a state of truth-striving; what it relates
itself to must be a matter of indifference. It must be the fullness and
height of the personal life that matters, and it must provisionally
remain an open question whether a certain definite dogmatic content of
faith is necessary for the development of such a life.

But he does not draw the full consequence of his proposition. That would
have led him to proclaim the principle of the free personality. He does
not abandon the assumption of a standard for truth different from the
personality itself, which for him is given in ecclesiastical dogmatics.
And yet---on closer inspection (which we shall undertake when we have
come to know his ethics) it will become apparent that this dogmatic
guideline is not the ultimate one. He is too much a thinker to be able
to rest content with it. His ultimate standard is of a formal nature:
the relation to the dogmatic-ecclesiastical content of faith is for him
the highest form of personal life, because in this relation the
strongest contradiction, the greatest suffering and passion, comes forth
in subjectivity. The last and decisive objection against him will then
prove to be that he has torn the personality loose from the real, natural
and social context in which alone its life can acquire a positive
content. For only in the free subjectivity, struggling with reality and
learning from reality, can truth dwell---not in the subjectivity that
through formal, ascetic exercises torments itself into obedience to a
content that is contrary to nature and hostile to life.---But only
through an examination of Kierkegaard's ethics can this point be more
clearly illuminated.

\bigskip

What is great about Søren Kierkegaard's epistemology is the seriousness
with which he grasps the connection of thought with personal life. There
is something Greek in him. He wants to return from abstract thinking and
contemplation to personal, existential thinking.

Our ideas all too easily become soldiers in peacetime. Søren Kierkegaard
has led ideas forth from the comfort of contemplation to the strife and
unrest of the struggle of life, and has pointed out the arena where the
validity of ``essential'' knowledge must stand its test. Had the
dogmatic impediment not been there, his thoughts would have attained a
more directly universal significance; as it stands, a translation, a
process of sifting, must take place before they can come to play the
role for humanity that is rightfully theirs.

% ------------------------------------------------------------
\section*{B.\quad Ethics}
\label{sec:4B}
\markright{IV. Søren Kierkegaard's Philosophy --- B. Ethics}

\subsection*{a.\quad The Leap}
\label{sec:4Ba}

\noindent\textbf{1.} There appears in the history of thought an important
opposition between two types of thinkers. For the one type the concern is
to find unity, coherence, continuity, to demonstrate successive
transitions and intermediate links between the apparently most opposite
forms of existence, to show how the apparently qualitatively different
phenomena consist of the same elements, only in different proportions of
mixture and forms of composition. For the other type the chief interest,
by contrast, is to establish the definite differences, to demonstrate how
at certain points in existence something qualitatively new emerges, to
maintain the abrupt transitions, the sudden swings, the sharp boundaries
between different domains. The chief concept of the former thinkers becomes
the continuous, successive transition, the difference of degree, the
\emph{relation of quantity}; that of the latter, distinction and
disjunction, the \emph{difference of quality}. The two types or tendencies
collide over a series of great problems in the domains of natural science
and the science of spirit. Thus over the question of the unity or original
heterogeneity of the forces of nature, of the unity of matter or the
qualitative difference of the elements, of the origin of life, of the
origin of species, of the relation of conscious life to the material, of
the mutual relation between higher and lower phenomena of consciousness, of
the relation of the will's decision to the law of causality, and so on.
Where it is impossible to determine quantitatively how great the
approximation is, how far the continuity reaches, or how great the distance
is between the oppositions appearing in experience---there the difference
between the two types of thought will especially assert itself; more
strongly, therefore, in the science of the organic than in the science of
the inorganic, more strongly in the science of spirit than in natural
science. And in philosophy it will become of special importance, since here
the concern is to fix the fundamental presuppositions and the points of
view of totality themselves.

Both types have their scientific justification. They represent the two
fundamental activities of our knowledge: synthesis and analysis, which, as
Goethe remarked, are related to each other as inhalation and
exhalation.\footnote{In my exposition of the doctrine of categories
(\emph{Den menneskelige Tanke}, pp.~154--162) it is shown how synthesis and
relation, continuity and discontinuity, likeness and difference are
constantly at strife in fundamental human thinking.} One might perhaps
think that the first had to be the properly scientific one, since synthesis
is the fundamental form of our spirit and our thought, and since all
science aims at finding coherence and unity in existence. But it is no less
essential that we become conscious of the whole qualitative richness that
existence contains. And synthesis itself first acquires its proper
significance when the content of experience is spread out before us in its
whole heterogeneity, with all its qualitative nuances and oppositions; only
then does synthesis have something to work with. Søren Kierkegaard has
taught us: it is no art to proclaim a unity when there are no opposing
elements to overcome.

That Søren Kierkegaard belongs decidedly to the latter class of thinkers,
of that there is no doubt. He is one of the first in the row of thinkers
who introduced that return to critical philosophy which is characteristic
of the last generation's philosophizing. He himself designated his
thinking, in opposition to Hegel's, as \emph{qualitative dialectic}, by
which he wanted to express it as his chief endeavor to make the qualitative
distinctions count, to maintain the fixed boundaries between the different
domains, whereas the preceding, romantic-speculative period wallowed to
excess in unity and continuity. His watchword was \emph{Either--Or}. It was
the title of his first significant work; it was the category he maintained
as the decisive one; it was the name by which he was known to the boys in
the street. He declares war on continuity and successive transition; they
are for him the categories of superficiality and softness. Life and
reality move, by his assertion, forward over constant crossroads, through
constant leaps. Something decisive is reached only by a jerk, a sudden
swing, that cannot be foreseen from, and is not determined by, what
precedes.

Spiritual life is in this respect, according to Kierkegaard, subject to
wholly different conditions than natural life. He denies most definitely
that there is any analogy between spiritual development and organic
development. Qualitative metamorphosis, in which the new stage, the higher
state, comes into force by a jerk of decision or by a leap, is what is
peculiar to the development of spiritual life, whereas organic development,
e.g.\ a plant's growth, consists in a successive unfolding of what
essentially was already in the germ. In that the transition is a leap,
there comes a break in the inner, natural coherence (in ``immanence''); an
absolute New is posited. The continuity that thought can produce is purely
abstract and in reality illusory: the real coherence must, in spiritual
life, be produced by ever-new exertion, by the work of repetition, which in
the new reality holds fast what was won from the past, and which by no means
continues its subsistence of itself.

\medskip

\noindent\textbf{2.} This peculiarity of Søren Kierkegaard's thinking
displays itself especially in the manner in which he conceives \emph{the
relation between psychology and ethics}---a relation in which, on the whole,
a thinker's peculiarity will as a rule most clearly display itself.

He was a psychologist of the first rank in his ability to grasp the
greatest oppositions and the finest nuances, the most violent storms and
the smallest ripples within the life of the soul. He was a well-traveled
man on the waters of the inner life, and he had poetic power to present what
he had seen.---But he did not have merely the observer's interest. He also
felt a passionate urge to understand. His whole strong emphasis on the
paradox, on the irrational, on the sudden jerk that abolishes coherence,
would also be unintelligible if he had not himself felt the urge to hold
fast to continuity. The doctrine of truth as the paradox is an expression
of the fact that this urge has been disappointed. Those in whom the urge to
knowledge is not alive are spared this disappointment, do not discover the
irrational, and will wholly lack the presuppositions for understanding the
passion an intellect like Kierkegaard's feels in striking against the limit
of thought. The pain of the halt is felt only by one who comes at full
speed.---``The paradox,'' says Kierkegaard (\emph{Philosophical Fragments}),
``is the passion of thought, and the thinker who is without the paradox is
like the lover who is without passion: a mediocre fellow. But the highest
potentiation of every passion is always to will its own downfall, and so it
is also the understanding's highest passion to will the collision, although
in one way or another the collision must become its downfall. This, then,
is thinking's highest paradox: to want to discover something that thought
itself cannot think.''

From this it can be seen that, even though one can divide thinkers into two
classes, only \emph{that one} can be great within his own class in whom
there stirs a strong urge in the direction that characterizes the other
class. Only through a strong urge to coherence is the break, the jerk,
discovered and felt.

As a psychologist Kierkegaard holds to coherence as far as he can; but as
an ethicist he posits the leap, the jerk of decision. He finds the task of
psychology in seeking out the possibilities of the new states, the
successive and quantitative approximations. It has to do with the
approaches, the motives, the preparations. But however far it comes in its
``calculation of the angles of possibilities,'' when the point comes at
which the properly, ethically significant decision takes place, there comes
a qualitative jerk by which the new is posited; a radical change occurs
that finds no explanation whatsoever in what went before.

\medskip

\noindent\textbf{3.} In the brilliant treatise \emph{The Concept of
Anxiety}, whose theme, as we have seen above, played a great role in his
youthful life, Kierkegaard develops this relation between the psychological
and the ethical way of regarding things.---He here treats the will's
decision at ``the Fall,'' following the narrative in the first book of
Moses. It might seem that it would have been more illuminating had he used
one of his usual psychological experiments. But the difference is not great,
for although he protested against that old narrative being a myth, he in
fact treats it as a myth and with all desirable freedom.---The concern is to
depict how the Fall could occur, since man is presupposed to be in the state
of innocence.

First psychology sets to work to find possibilities and approximations. One
such is offered by anxiety. While fear relates to a definite possibility,
anxiety relates to an indefinite possibility, a dream that emerges in the
imagination of innocence, in itself a nothing---which nevertheless comes to
stand as though it were something. And besides this interplay of coming and
vanishing, there asserts itself in anxiety yet another interplay: of
attraction and repulsion, sympathy and antipathy, approach and recoil. The
object of anxiety exerts a power over the mind, draws it to itself---and yet
repels it. Thus it is in bashfulness. What is the bashful person ashamed
of? Nothing, after all, has been said and nothing done; only a possibility
went by, without even being grasped by a definite thought---and this
possibility was rejected, though scarcely noticed. What is there to be
ashamed of, to blush over? And thus it is with dizziness. The possibility of
letting oneself fall, of giving up one's equilibrium, offers itself when the
gaze measures the abyss; but it is a possibility that gets no time to fix
itself, but is at once rejected, again and again rejected, and yet it
causes dread, because it draws.

Here is a possibility, an approximation---but which, according to
Kierkegaard, explains nothing when the decision comes, when the Fall really
occurs. This comes with a jerk, or as a leap---incomprehensible and
ungrounded by all that went before.---Just as suddenly as all possibilities
and approximations are thus swept away, ethics steps into the place of
psychology. It asks about differences of quality, about the opposition
between good and evil, and cares nothing at all about all that psychology
has laboriously calculated concerning the ``angles of possibilities.''
Reality stands in no relation to possibility.---This conception is extended
to hold for the relation of the individual action to the acquired
character, and for the relation of the individual to the nature inherited
from the race: what is given from the past (both the race's and the
individual's own) are possibilities, approximations, that in the moment of
real decision signify nothing. And this although Kierkegaard admits that by
habit and practice the possibilities and dispositions in the individual's
character are summed up, just as, in the matter of race- and
family-qualities, happens in the whole race. By the quantity the new
quality that is posited by the decision is to be \emph{wholly incapable} of
being explained! The continuous growth of the possibilities is broken off
by the jerk of reality.---

I cannot see it otherwise than that Kierkegaard, by this conception,
entangles himself in a very great contradiction. His recognition of
psychology in its independence of ethics must rest on the significance that
can be attributed to the possibilities and approximations that psychology
can demonstrate. But now these are, however far they may go, in the end not
to signify the very least! We then get possibilities that are not
possibilities of anything, approximations to nothing! Psychological
understanding---and with it not merely theoretical psychology, but also
practical knowledge of human beings---then becomes useless, indeed illusory.

Kierkegaard has in \emph{The Concept of Anxiety} disavowed his watchword
``Either--Or'' and wanted to try to carry through a ``Both--And.'' Indeed,
strictly speaking, he has wanted to have \emph{both an} ``Either--Or''
\emph{and a} ``Both--And.'' He has not wanted to choose between leap and
continuity, but has wanted to combine both! But he has thereby only
succeeded in falling into Scylla without avoiding Charybdis. He has offended
the interest both of ethics and of psychology by tearing apart the bond
between possibility and reality. For just as little as psychology can put up
with the possibilities it is able to demonstrate being wholly unable to
ground the real decision---even though it be willingly admitted that a
complete grounding is an ideal (and that for the reason that not \emph{all}
possibilities can be discovered)---just as little can ethics put up with the
ethical evaluation concerning only the finished result, the sharp turn, and
not also the approaches, the dispositions, the motives. For the practical
ethical way of regarding things the concern is precisely to get hold of the
possibilities, in order to arrest the pernicious reality in its growth, and
in order to shield and foster the good germs.

On closer inspection Kierkegaard, too, does not avoid extending the
evaluation to what precedes the jerk of decision, to the approaches before
the leap. He indeed maintains definitely that anxiety is \emph{not} any
\emph{imperfection} in the human being; but later he nonetheless declares
anxiety to be ``the most \emph{selfish}'' thing, and speaks of how the
indefinite possibility ``\emph{burdeningly} causes dread by its sweet
anxiousness,'' and how anxiety ``maintains a \emph{sly} communication with
its object.'' Here the evaluation is antedated, or a decision is assumed
before---the decision---that is to say: the possibility is taken for (what it
also is) a partial reality. Kierkegaard has not been able to carry through
the sovereign contempt for the approximative and the hypothetical that lay
at the basis of his distinction between ``essential'' and ``inessential''
knowledge. He protests against ``laying out the circle of the leap into a
straight line''; but he has not considered that the greater the radius of
the circle, the more the arc approaches a straight line, or that a
circumference can be conceived as consisting of an infinite number of
infinitely small straight lines. Continuity in thought and in nature does
not let itself be chased away so easily as Kierkegaard thought.

\medskip

\noindent\textbf{4.} Hitherto we have spoken of the leap or jerk only from
the negative side, conceived the decision as an interruption. But can there
not be given a positive description or determination of it? It is
characteristic of Kierkegaard as a thinker that, although by the category of
the leap he wants to break the bridge behind him, his thought nevertheless
cannot refrain from circling about the place where the interruption occurs.
The incomprehensible leap he wants to approach as closely as possible. He
is, so far as I know, the only indeterminist thinker who has attempted to
describe the leap. It is, philosophically regarded, one of the most
remarkable developments of thought to be found in him.

``The history of the individual life,'' he says (in \emph{The Concept of
Anxiety}), ``proceeds in a movement from state to state. Every state is
posited by a leap\dots\ Prior to every such leap there goes a state as the
nearest psychological approximation. This state is the object of
psychology.''---Thus the leap lies \emph{between two states}. Or, more
precisely, as it is said in another place, \emph{between two moments}.
``Anxiety may be compared with dizziness. He whose eye happens to look down
into a yawning deep becomes dizzy\dots\ In this dizziness freedom sinks
down. Further psychology cannot go and will not. In that same moment
everything is changed, and when freedom rises up again, it sees that it is
guilty. Between these two moments lies the leap, which no science has
explained or can explain.''

It seems to be clear that if the leap lies between two states or between two
moments, no observation can catch sight of it. The description of it then
becomes no description, since it can never become a phenomenon. What
Kierkegaard wants to grasp by his description slips away between his fingers.
He describes a succession of two moments or states (and what else, indeed,
can here be observed or described?), but what lies in between he does not
get hold of, and cannot get hold of. And yet he speaks so dogmatically about
it as if it were quite possible. I have devoted much reflection to the
passages cited from \emph{The Concept of Anxiety}, and was at first inclined
to take it as Kierkegaard's meaning that the leap could not be observed.
Then his development would be a counterpart to Hume's famous demonstration
that causality cannot be observed, since what we observe is always only
succession. Just as little as causality or continuity can be observed, just
as little can a break in continuity be. The moments succeed one another---but
what is the relation between two moments? Here is the great problem that
divides the two groups of thinkers characterized above, of which the one
maintains continuity (through the causal relation), the other discontinuity
(through the indeterminate appearance of new qualities). If observation
could decide the matter, the dispute could easily be settled. But even by
the most exact, most microscopic observation we succeed only in dividing the
moments into smaller moments, and the problem, the fundamental problem in the
theory of knowledge, returns.---But on closer consideration I cannot
attribute to Kierkegaard a clear eye for the point of principle in the
question. His dogmatic presuppositions and his propensity for poetic images
have here veiled his eye for the problem in its sharpness. Otherwise he must
also have become aware that he really lets the whole decision (the
\emph{will's} decision!) take place unconsciously. For what lies between two
states of consciousness or moments of consciousness happens unconsciously,
does not raise itself above the threshold of consciousness. A decision of
the will of the kind Kierkegaard is dealing with must surely be a willing of
the highest form, a choice in the proper sense, a resolution, a deliberate
act. But how can a resolution lie between two moments? Only that kind of
``resolution'' can, which one arrives at in one's sleep. Only involuntary
acts can be unconscious. But Kierkegaard has surely hardly thought of ``the
leap'' as a reflex effect!---

Looking more closely, Kierkegaard, too, involuntarily comes to place the
decision at the first moment, when ``the individual sinks down in the
impotence of anxiety.'' We then get a causal series: anxiety---dizziness---
sinking---fall; and where, then, has the leap gone? He who sinks does not
leap; sinking is (even if to sink is \emph{to let oneself sink}) nothing but
a beginning fall, a fall that could be prevented---if one were not standing
precisely at an abyss.---Still more clearly does this antedating of the
decision appear in a passage already cited: ``He whose eye happens to look
down into a yawning deep becomes dizzy. But \emph{what is the cause? It is
just as much his eye as the abyss; for suppose he had not stared down
there!}'' The guilt, then, lies only in part with the eye; here too
Kierkegaard gives up his Either--Or. But the main thing is: has the eye any
guilt here? There is an involuntary urge to let the eye roam about, an urge
that cannot be said to be of the evil. That the eye, in this its involuntary
roaming---which is the way to all significant experiences and thereby to all
higher development---now also, among other things, strikes upon the abyss and
regards it with involuntary attention, cannot be its own guilt; for it did
not know beforehand that the abyss was there, did not even know what an abyss
was. And when the dizziness is awakened by the representation of the possible
fall arising and working on the mind, there occurs indeed a transition from
the involuntary to the deliberate (insofar as by deliberateness one
understands a conduct determined by preceding representations), but this
transition itself takes place \emph{involuntarily}, in virtue of the laws of
the association of representations. Even if the fall occurs with intent, no
intent is possible (as Shakespeare and Spinoza have taught us) without
memory, and memory can present itself just as involuntarily as the abyss
presents itself to the roaming eye. It is thus possible to carry through the
psychological continuity even further than Kierkegaard has done; and by his
antedating of the point at which he places the decisive guilt, he in effect
admits that ethics does not stand or fall with ``the leap.''

Kierkegaard has, then, in \emph{The Concept of Anxiety} given psychology far
more than he can consistently. But these were only passing concessions. His
chief grip is and remains Either--Or, the jerk or leap of qualitative
decision; with it he operates everywhere in his ethics. And in his strictly
religious writings, especially later on, when his conception of the
Christian sharpened, there is no trace of the decision lying \emph{between}
two moments of consciousness. There he pushes all psychology away from under
his feet by an utterance such as this: ``To assert that a person
\emph{knowingly does} what is wrong---does what is wrong with his knowledge of
the right---Greek antiquity has not the courage'' (\emph{The Sickness unto
Death}). What happens \emph{knowingly} must fill a moment or a state, which
must be capable of being observed. Kierkegaard has, to be sure, not given us
this observation. He has not wanted to. But---he has also not been able to.

\subsection*{b.\quad The Stages}
\label{sec:4Bb}

The most important contribution to ethical thinking from Søren
Kierkegaard's hand is his presentation of what he calls ``the stages of
life.'' Each stage is a life-state, a view of life, which Kierkegaard has
presented partly in poetic form, partly with conceptual rigor,
without---as will be seen in the following---the poetic and the
philosophical presentation always agreeing with each other. At no point,
however, does his brilliant poetic-psychological endowment appear so
clearly as here. To give a comparative philosophy of life he was equipped
as few are, and what he in his way and from his vantage point has given in
this direction will surely be continued under other forms and from other
points of view. Ever more, it seems, the need for ``subjective thinking,''
for thinking about life in close connection with life, asserts itself, and
there is here a task that cannot be treated satisfactorily by mere
novelists. One must dig deeper than that.

Between the different stages the transition occurs, according to
Kierkegaard, by one of those leaps we have just discussed. Within each
stage the consequences of the standpoint posited by the leap unfold.---A
distinction is made between three main stages: the aesthetic, the ethical,
and the religious. Between the aesthetic and the ethical stage irony forms
a transitional form, just as humor forms an intermediate form between the
ethical and the religious stage.

\subsubsection*{$\alpha$.\quad The Aesthetic View of Life}
\label{sec:4Bb-alpha}

\noindent\textbf{1.} The first stage is the aesthetic. It is depicted in
the first part of \emph{Either--Or} and in the first section of
\emph{Stages on Life's Way}. As far as the general conceptual
determinations are concerned, it appears most decisively and clearly for
this stage in the essay ``The Rotation of Crops'' (\emph{Vexeldriften. An
Attempt at a Theory of Social Prudence}, in the first part of
\emph{Either--Or}).

This way of taking life seeks to transform it into mere possibilities,
with which one lets the imagination have its free play. Everything is made
an object of enjoyment, but by this one must think not so much of
immediate, sensuous enjoyment---though this enters in---as of the
arbitrariness with which everything is apprehended and treated: ``The whole
secret lies in arbitrariness. People think it is no art to be arbitrary in
such a way that one does not oneself get lost in it, that one oneself has
pleasure from it.'' What matters is to transform the whole of existence for
oneself into a kaleidoscope, which one constantly shakes in order to
delight in the various accidental combinations that in this way can
arbitrarily be produced.

In order that this free play may stand in one's power, one must not
personally bind oneself in any relation of life, not settle down at any
point in existence in such a way that one is held fast, so that
continuation and repetition exclude the production of new kaleidoscopic
situations. One must keep oneself in ``complete hovering.'' Therefore, to
be sure, contact with people, often also for a while accompanying them on
the way---but always securing for oneself such a speed that one can break
off the fellowship when one wishes.\footnote{The aesthetic view of life,
according to Kierkegaard's depiction, especially recalls in this trait
Aristippus of Cyrene. What he maintained as the highest was indeed the
enjoyment of the moment, but the condition for reaching this was for him
precisely the complete freedom that does not let itself be bound and
imprisoned. Therefore he would (as he says to Socrates in the conversation
in Xenophon: \emph{Memorabilia} II, 1) not enter into any civil relation,
but be a stranger or guest everywhere (οὐδ' εἰς πολιτείαν ἐμαυτὸν
κατακλείω, ἀλλὰ ξένος πανταχοῦ εἰμι). He wants to be free, neither ruler nor
slave. According to the other traits reported of him, it was especially the
preservation of freedom in the face of the ensnaring element in enjoyment
itself that was his gift. In this respect he was a true virtuoso. The
contradiction in his view of life was that so much art and preparation was
nevertheless required, with the use of the past's experiences and of
prospects for the future, that many moments had to pass without immediate
enjoyment. Really, the highest for Aristippus had to become virtuosity
itself---and so it is for the modern hedonism Kierkegaard describes.
Characteristic of the difference between the modern and the ancient is
precisely that in modern hedonism a wholly decisive weight is laid on
arbitrariness, on active experimenting, the conscious production of
situations. In Aristippus arbitrariness expresses itself merely in the
ability to choose without bias among what offers itself; it lacks the
somewhat romantic arranging that the modern age's greater self-consciousness,
and likewise its less naive and trustful relation to life, brings with it.}
Friendship is too fixed and solid a relation. And even if a needful rupture
should leave behind unpleasant memories, the unpleasant can indeed give a
piquant aftertaste that it may be worth taking along. In a similar way it
stands with marriage. Love has its time; but if one promises that it shall
last eternally, one promises more than one can keep. And besides, there is
this danger in marriage, that one loses one's freedom: ``one cannot order
travel-boots when one wishes, cannot rove about restlessly.'' (``There has
indeed been an example of a gypsy woman carrying her husband on her back
through life, but partly this is a rarity, partly also, in the long run,
tiring---for the husband.'') Finally, by getting involved with marriage one
is compelled to come under the dominion that custom and usage exercise; a
whole circle of considerations and duties excludes arbitrary roving. Thus,
to be sure, eroticism, but not marriage. And just as little any occupation
of a calling: thereby one would make oneself a cog in the great social
machine, no longer be the master of the enterprise oneself. One should of
course not be inactive, but pursue all sorts of breadless arts, take care
that the occupation constantly bears the stamp of idleness.

\medskip

\noindent\textbf{2.} This stage shows itself most characteristically in the
relation between man and woman. There is, indeed, in this relation too
application enough for the rotation-method or the kaleidoscope, since it can
run through a whole scale, vibrate in very different directions, branch out
into manifold derived relations. There is possibility here for a series of
the most opposite moods and passions: from high-spirited jest to deep
earnestness, from frivolous flirtation to heartfelt devotion, from flaring
enthusiasm to steady fidelity, from the brutal stirring of a pure natural
instinct to the sublime feeling of spiritual belonging-together. And through
its consequences the relation of the sexes leads, moreover, beyond the
mutual relation of the individuals into connection with the life of society,
of the race, intervenes in this and is determined by it in its fate and its
character. Here, then, is a great range of play. If one of the many strings
is struck with the requisite art, one can get more or fewer of the others to
vibrate along, without their yet taking the mastery. No wonder that one of
the poets' chief themes has at all times lain here. And it is for the same
reason that Kierkegaard especially characterizes his ``aesthetic stage''
from this point of view. It is one of the points where he has with lavish
hand strewn his most brilliant images and presentations, and there are many
who know him as an author only from this part of his production. (\emph{The
Seducer's Diary} in the first part of \emph{Either--Or}. \emph{The Banquet}
in the first section of \emph{Stages on Life's Way}.) Here, where the
concern is not an investigation of his literary art and greatness (which
would lie beyond my ability and has, besides, been done so brilliantly in
Georg Brandes' book that I have merely to refer to it), but the
philosophical significance of his presentation, I can content myself with
bringing out a couple of the most characteristic passages.

``The Young Man'' who is present at ``The Banquet'' has come to a halt at
the contradictory and comic element that for him lies in the erotic
relation, in that it seems to concern the individuals' highest desire---and
yet it is the species whose propagation is to be secured through the
awakened instincts.---``The lovers want to belong to each other for all
eternities. This they express in that strange way by enclosing each other
in the inwardness of the moment, and all the bliss and desire of love is to
lie therein\dots\ And yet they are deceived; for in that same moment the
species triumphs over the individuals, the species conquers while the
individuals are reduced to being in its service\dots\ When the lovers have
found each other, one should think they were a whole, and therein should
lie the truth of their wanting to live for each other in all eternities. But
lo, they begin, instead of living for each other, to live in the species,
and this they do not suspect.'' In the face of this irrational element in
love the Young Man has come to a halt. He understands well the introduction
to it, the being-seized in it---but he does not understand the
consequence.\footnote{The thought developed in the young man's speech was
at the same time developed, in more natural-historical form, by Schopenhauer
in his »Metaphysik der Geschlechtsliebe« (chapter 44 in the fourth book of
the second part of »Welt als Wille und Vorstellung«, which appeared in 1844,
the year after \emph{Either--Or}).}

Johannes the Seducer, the last in the row of the Banquet's brilliant
speakers, states the ``aesthetic view of life's'' position of principle
toward the problem the Young Man has brought out. He tells how the gods
feared that man, the original human being, would show himself unwilling to
bear their yoke, that he might even bring heaven to totter. By force he
could not be compelled; then they produced woman to enchant him and ``catch
him in all the diffuseness of finitude.'' And the bait worked---yet not
always. ``There were at all times some men, a few, who became aware of the
deception. These indeed saw her loveliness, more than anyone else, but they
suspected the connection. These I call Erotici, and count myself in their
number; the men call them Seducers, woman has no name for them, such a one
is for her unnameable. These Erotici are the happy ones. They live more
sumptuously than the gods; for they eat constantly only what is more
delicious than ambrosia and drink what is more delicious than nectar: they
eat the most seductive fancy of the gods' shrewdest thought, they eat
constantly only bait---o voluptuousness without equal, o blessed way of
life---they eat constantly only the bait---they are never caught. The other
men grab and eat bait as peasants eat cucumber salad, and are caught.''

\medskip

\noindent\textbf{3.} The aesthetic view of life places the highest in being
a tangent to the circle of life, in seizing the momentary contact, but then
at the right time going away by virtue of the ``centrifugal''
(midpoint-fleeing) force. It does not let itself be made a planet. At most
it engages---for the sake of the ``rotation-method''---in tangenting the
circle a second time, like a comet, at another point. The erotic relation is
precisely a relation where the momentary contact can quite rightly, as the
Young Man feared, become constitutive for wholly new, untried life-relations
and tasks. But if the hovering of arbitrariness is to be preserved, one must
break off in time. Whether greater fullness of life, more inward
understanding of life and feeling for life, might be attained if one ``let
oneself be caught''? The ``aesthete'' does not know it, will not and cannot
know it. If any inkling of this were possible, then the transition from the
aesthetic stage to the following one would not be a leap, but could come
about by a motivating growth. It is characteristic what Kierkegaard wrote in
his own copy of \emph{Either--Or}: ``A real falling-in-love could not be used
in the first part; for it always grips a person so deeply that he enters the
ethical. What I could use was a multiplicity of erotic moods.'' Under ``the
aesthetic stage'' Kierkegaard presents figures who have come far beyond the
immediate, the involuntary, the nature-forceful, and have practiced an art
of arbitrariness, an ability to point up, not merely to seize the piquant,
but also to produce the piquant where it does not offer itself of its own
accord. It is not a standpoint that people easily reach. It lies, as already
remarked, a good stretch further out in reflection than ancient hedonism.

Here there arises, from the standpoint of a comparative philosophy of life,
the objection that we are not told how this stage comes to be. As Kierkegaard
himself has seen, there can lie in the natural instinct a drawing power that
leads beyond the aesthetic stage, and by whose working the ``aesthetic'' is
only a co-operating element. How, then, did there enter the halt, the
deflecting movement, by which the individual went away along the tangent
instead of continuing the movement around the center under whose influence it
had momentarily come? The childlike forgetting of one moment for the sake of
another is something quite different from this artistic or artificial
rotation-method. It is the genetic or evolutionary way of viewing that is
wholly lacking in the Kierkegaardian presentation of the stages. He even
deliberately keeps away the forces that at the one stage could contain an
announcement of the next. Thereby each single stage comes (in quite good
agreement, to be sure, with the theory of the leap) to stand as something
fully finished, which points neither forward nor backward. It is finished
figures that he depicts. They have ``run themselves fast''---not in any
relation of life, but in their way of taking life. They have done away with
all possibilities. Even the last possibility---the one that rests on the
relation of contrast, on the tendency a strongly tensed state of soul has to
strike over into its opposite, a possibility of very great significance
during the development of the life of the soul---even that seems to be no
longer present for these ``aesthetes,'' in whom enjoyment and reflection have
blunted or burned out the force that is presupposed if even the mere
contrast-effect (which is really a phenomenon of fatigue) is to be able to
take place. The psychological thought-experiment has here given such extreme
forms as life probably does not present. It is, then, a little apt name that
Kierkegaard uses when he calls what he depicts stages. By a stage one thinks
of a link in a development; but about the Kierkegaardian stages one is not
told whence they come, and at bottom (for ``the leap'' is no answer) neither
whither they go. His depiction recalls rather Dante's \emph{Divina Commedia}
with its three divisions---and the aesthetic stage then corresponds to hell,
over whose entrance stands written: You who enter, abandon hope!

Kierkegaard has undeniably---the in other respects masterly quality of his
depiction left unmentioned---thereby made the task easier for himself than it
is.---How all forms of thought contain temptations! There was a temptation in
the quantitative dialectic (the unity of oppositions) in the direction of
regarding things as simpler than they are; but the qualitative dialectic, the
philosophy of the leap, shows itself also to have a temptation in the same
direction.

Everywhere that a development is possible, it is not from the fully finished
forms, in which life has as it were run itself fast, but from the more
indeterminate, vague, little-differentiated forms. It is only the popular
misunderstanding of the doctrine of development that lets one of the species
we now have descend from another (man, e.g., from the ape). According to the
doctrine of development the now-existing species descend from forms that had
only ``a general likeness'' with those that have arisen from them in the
course of development. This lies simply in the fact that all development is a
differentiation, an arising of differences. Such a process is impossible
where everything is beforehand pointed up, formed, and driven out to the
utmost tip. The tip can be broken off, but no new form comes out of that.

If one wants to understand how a new stage can develop out of a given stage,
one must therefore not hold to this in its fully finished form, but must go
back to an earlier stage of development, where the now-finished stage was in
its becoming and still had a more indeterminate shape. Only where there is
becoming and indeterminacy are there possibilities. That Kierkegaard himself
has taught us; but he will, once and for all, when it is a matter of
decision, have nothing at all to do with the possibilities. One must go back
to the point where the way divided---like a railway train that has come onto
a wrong track, one must go back to the switch-point. At the switch-point the
real decision takes place---but it does not always step up into
consciousness; it can happen half-consciously or unconsciously. We are, by a
natural psychological illusion, inclined to \emph{post}date our decisions and,
on the whole, to attribute to our conscious choice greater significance than
it has. It is often merely the conscious keystone.---Here the psychological
and the ethical interest show themselves in their harmony. In the face of the
unmotivated and unprepared leap we would stand wholly helpless, and the
ethical judgments would be mere outbursts of feeling without practical
significance, since they would come afterward and could get no influence at
all on the next leap. Only by going back to the possibilities of development,
that is, to the unfinished stages, can one do something (however much or
however little it may be) to determine the future.

\subsubsection*{$\beta$.\quad The Ethical View of Life}
\label{sec:4Bb-beta}

\noindent\textbf{1.} In the conception of the ethical view of life there
appears a discrepancy between the poetic presentations in \emph{Either--Or}
(second part), in \emph{Stages on Life's Way} (second section), and
\emph{Fear and Trembling} on the one hand, and the philosophical
characterization in the \emph{Concluding Unscientific Postscript} on the
other---a discrepancy that shows itself to be a contradiction pointing to a
decisive change in Kierkegaard's ideas.---We hold first to the poetic
depictions (and especially to the two first named, since \emph{Fear and
Trembling} can better be discussed in connection with the religious view of
life).

While the aesthetic stage dissolves all relations of reality into mere
possibilities, with which imagination and mood can drive their arbitrary
play, reality, earnestness, and responsibility are the chief determinations
at the ethical stage. And in that here too the relation between man and
woman is used for the poetic illumination of the view of life, marriage
appears as the expression of the ethical. Marriage distinguishes itself
from the merely erotic relation in that to falling-in-love there is added a
resolution. In the resolution lies, ethically regarded, the whole ideality
of the human being. But the question now is: how can deliberation and
resolution find application there where immediate feeling and passion alone
could seem to have the word? Kierkegaard's answer to this is fine and
beautiful. Deliberation and resolution, he says, have nothing to do with
falling-in-love itself: ``The holy place of falling-in-love, the
consecrated ground of immediacy, they dare not tread.'' What they have to do
with is ``the relation of falling-in-love to reality.'' What matters is
whether one dares take up the relation posited by falling-in-love and carry
it through in reality, in the face of danger and temptation, and with the
consciousness of the universally human in the relation of communal life that
is founded.---Thus falling-in-love is presupposed as given, as the
constituting element, and the resolution appears as the protecting power
that guards against inner and outer dangers. While the development given of
these thoughts is beautiful and rich (most successful in the \emph{Stages},
rather drawn-out in the second part of \emph{Either--Or}), there are, it
seems to me, several psychological determinations lacking. Kierkegaard has
too dogmatic a concept of ``the resolution.'' It appears as a \emph{deus ex
machina} and stands in a rather external relation to ``falling-in-love,''
with which it is nonetheless to enter into the most inward connection, in
order that the marriage may come to be---and, what must especially be noted,
by which it must from the first have been called forth and motivated. The
resolution here in question must have sprung from the individual's
personality, its real self, and have been produced by the manner in which
this central element in the individual is determined by the newly awakened
passion. A multitude of different relations can here be possible:
falling-in-love can, in manifold different ways, affect the innermost I. It
can enclose and cover it, but also merely touch it at the periphery; it can
bring wholly new elements to life in it, supplement and expand it, but it
can also work as a demonic power that narrows and isolates. And these
different ways of affecting can appear successively in the same individual,
in the same erotic relation. The ground on which the resolution in question
rests---namely the relation between the new feeling and the central element
in the I---can therefore be indeterminate and shifting; there is possibility
for several illusions and anticipations in the motivating presuppositions of
the resolution. And to this is added, further, that in the period of life in
which marriages are most frequently founded, the characters are not yet
fully developed. Even if there is now present the agreement or the
supplementing opposition that makes the communal life possible and happy, it
is not thereby said that the continued development of the two characters
will remain in the same tracks: they can each for itself develop ably and
soundly---and yet the tracks can part. The problems that thereby arise, and
the tragedies that thereby become possible, Kierkegaard does not go into.
This is connected with the psychological dogmatism in his concept of the
resolution; but it is also connected with a dogmatism in his concept of
marriage. He sees it merely as fully finished in its ideal shape and then
declares: all or nothing. Finally, it is also connected with his regarding
marriage as a religious institution. That resolution becomes in his eyes
impossible without religious assurance. And---characteristic of his
conception of woman---in the case of women there is no question of
resolution: ``A feminine soul has not, and shall not have, the reflection
that man has. Thus she shall therefore not reach the resolution. But swift
as the bird she reaches from aesthetic immediacy to the religious.'' In the
case of woman, then, the ethical stage falls away! And in the case of man
too, really, since ``the resolution'' is not to be thinkable without
religious presuppositions.---It will also appear in the following, from
several sides, how the ethical view of life comes up short in Kierkegaard's
presentations.

\medskip

\noindent\textbf{2.} In the transition from the aesthetic to the ethical
stage the concept of \emph{Repetition} plays an important role. It is akin
to the resolution, but does not coincide with it. It forms, like the
resolution, a boundary between the aesthetic and the ethical. Aesthetic it
is to rock oneself in moods and impressions; but these had then best be new,
shifting impressions. Ethically one returns with inwardness and fidelity to
the same; repetition is in a certain sense the ethical art: ``the watchword
in every ethical view.'' But repetition plays a role not only in the coming
to be of the resolution; it comes forward again in the resolution's
transition to action: in the resolution I have, after all, anticipated the
action; the resolution is an ideal, a possible action; but now it is a
question of whether I can repeat, that is, convert the possible into
reality, although reality here is only a repetition of what was contained in
the possibility. And yet once more repetition comes forward, when I have
victoriously carried through my resolution, won my cause. For in the
intoxication of victory the concern is not to forget the task of maintaining
what has been won, of continuing what has been carried out: ``Character is
not so much to conquer as to endure after having conquered, to preserve
oneself in character'' (cf.\ \emph{Repetition. An Attempt in Experimenting
Psychology}).

The concept of repetition places Kierkegaard's philosophy in a characteristic
contrast to Hegel's doctrine of the reconciliation of oppositions
(mediation) in a higher unity. ``Repetition,'' says Kierkegaard, ``is really
what, by a mistake, has been called mediation.'' The mistake consisted in
regarding the reconciliation of oppositions as self-evident, as determined
by the very nature of the terms of the opposition. In the domain of the will
it is, according to Kierkegaard, otherwise. There a psychologically
inexplicable act of will is required to connect the conflicting elements; a
new factor, then, is added besides these. It is this that makes him, as
already mentioned, deny the analogy between spiritual and organic
development. According to Hegel's conception the higher unity arises by one's
returning from the second term of the opposition to the first: but precisely
this return is the great problem for Kierkegaard, the one that marks the
boundary between the aesthetic and the ethical way of taking life. It marks
the point where the will's intervention---thus, according to Kierkegaard: the
jerk, the leap---has its place. Aesthetically one goes away along the tangent;
a peculiar force must be postulated that can hold the sphere fast in an orbit
around the same center.

In the writing \emph{Repetition} Kierkegaard has developed the significance
of this concept with a lavishness of poetic ornament that makes it a pearl in
the rich treasury of his works, although it cannot be denied that the poetic
apparatus spreads out at the expense of the concept and the thought.
Kierkegaard complains loudly of Heiberg's misunderstanding of his meaning;
but he had not done enough to secure a clear understanding.---I cannot
refrain from citing a piece of the said writing, which in the most beautiful
way expresses the thinker's idea.

\begin{quote}
``Hope is a new garment, stiff and taut and glittering, yet one has never had
it on, and therefore does not know how it will suit one, or how it sits.
Recollection is a discarded garment which, however beautiful it may be, does
not fit, since one has outgrown it. Repetition is an indestructible garment
that fits closely and tenderly, neither pinches nor flaps. Hope is a lovely
maiden who slips away between one's hands; recollection is a beautiful old
woman with whom one is nevertheless never served in the moment; repetition is
a beloved wife of whom one never grows weary\dots\ It takes youth to hope,
youth to recollect, but it takes courage to will repetition. He who will
merely hope is cowardly; he who will merely recollect is voluptuous; but he
who wills repetition is a man, and the more emphatically he has known how to
make it clear to himself, the deeper a human being he is. But he who does not
grasp that life is a repetition, and that this is the beauty of life, he has
condemned himself and deserves nothing better than what will befall him: to
perish; for hope is a beckoning fruit that does not satisfy; recollection is
a paltry pittance that does not satisfy; but repetition is the daily bread
that satisfies with blessing. When one has sailed around existence, then it
will show itself whether one has courage to understand that life is a
repetition, and desire to rejoice in this. Yea, if there were no repetition,
what then would life be? Who could wish to be a tablet on which time wrote a
new script each moment, or a memorial of the past? Who could wish to let
himself be moved by all the fleeting, the new, that always softly delights
the soul?''
\end{quote}

It shows itself, however, in the concept of repetition too, that the ethical
stage is not kept apart from the religious. For repetition---which in
Kierkegaard really means willing, acting, active intervention, and which he
therefore calls the watchword of the ethical view---is nonetheless for him
really a religious category. The young man depicted in \emph{Repetition}
only attains to ``repeating,'' to living life, through a religious movement.
Repetition means, on the whole, concentration, return and immersion in
oneself---and since now the deepest self-understanding, according to
Kierkegaard, occurs in the religious, it is explicable how repetition, which
first marks the boundary between the aesthetic and the ethical, can also
appear as a religious concept.

But whether repetition is conceived ethically or religiously, it is,
according to Kierkegaard, not to be grounded psychologically. It is
``transcendent,'' presupposes on the one side a ``jerk of decision,'' on the
other a relation to a supernatural principle. Here too he has surely himself
leapt too early from psychology. For the feelings that are attached to a
richer content of thought---and of this kind are especially those due to the
relation to other personalities---repetition is very well intelligible in
virtue of general psychological laws. But since I have elsewhere (in my
\emph{Psychology}) expressed myself more fully on this, and with special
regard to Kierkegaard, I shall not here go further into it.

\medskip

\noindent\textbf{3.} While Kierkegaard in his poetic presentations posited
marriage as the characteristic of the ethical stage, it is later (in the
\emph{Concluding Unscientific Postscript}) a wholly different concept that
becomes for him the ethical chief concept: \emph{the single individual}.
Ethically regarded, the individual has to do only with himself, he maintains.
``The individual's own ethical reality is the only reality.'' ``Ethically
there is no direct relation between subject and subject.'' The other's inner
life I know only as possibility; immediately real to me is only my own inner
life: it is there that the domain for my action lies.

There is between the poetic and the philosophical presentation a
contradiction that Kierkegaard has not been aware of. When an enormous
distance is placed between individual and individual, and each has to do only
with himself: how then can a relation of communal life without more ado be
made into a type for the ethical?---This transition from emphasizing a
relation of communal life to emphasizing the single individual by himself
signifies the emergence of one of Kierkegaard's most significant thoughts,
and it agrees with his fundamental view; but it signifies at the same time,
by the further developments attached to it, a transition to a
religious-ascetic ethics. He leaves the relation of communal life in such a
way that he later cannot find his way back to it.

The first time, so far as I know, ``the single individual'' appears is in the
preface to \emph{Two Edifying Discourses}, which appeared in the spring of
1843 (some months after \emph{Either--Or}). Kierkegaard here relates how he
let his eye follow the little book on its wandering. ``I saw then how it went
its way along solitary paths, or solitary along the frequented ones. After
some little misunderstanding it at last met that single individual whom I with
joy and gratitude call \emph{my} reader, that single individual whom it seeks,
after whom it as it were stretches out its arms.''---Kierkegaard tells that he
had decided to strike out this preface and therefore went up to the printing
house, where it was already set; but the typesetter, who had been moved by
it, begged that it might after all remain standing, and Kierkegaard then
thought for a moment of having only two copies printed and presenting the
typesetter with one: then \emph{he} could indeed be the single individual!---So
literally he did not take it with his ``single'' reader; but in his ethics he
made all the stricter earnest of this concept. If one takes account of what is
found here and there in his writings and posthumous papers to illuminate his
grounding of this concept's significance, then the following points of view
show themselves to be the most important.

a) In the single individual there occur the events that give spiritual life
its value; there lie the tasks, and there the decisions are to take place.
The race is a phantom, a fantastic medium; only the single individuals are
realities, and each only for itself an immediate reality. If the race has a
goal, then it in any case does not concern the single individual.

b) What others can do for me can only be to render awakening influences, and
this is more indirect than direct, and is a difficult art, the one Socrates
practiced, but which few can practice as he did. The single individual is
``the category of the awakening of spirit,'' the narrow pass through which the
race must go in order to reach the highest. The one human being cannot judge
the other, because he cannot know him in his reality; but on the other hand it
is also very little that the one human being can do and be for the other, and
Kierkegaard says of himself: ``As far back as I can remember, I was agreed
with myself on one thing, that for me there was no comfort or help to be sought
from others.''

c) And not only, ethically regarded, does everything on the whole depend on
self-activity; but also in the special determination of the degree and height
to which the single individual is to attain in his willing and acting, he is
referred to himself. ``The \emph{quantum satis} of manly will'' (to use Henrik
Ibsen's expression in \emph{Brand}, whose work is here, as at several points, a
poetic commentary on Søren Kierkegaard)---this only the single individual
himself can find. No other can tell him where the boundary lies between lack of
will and lack of ability, between sluggishness and selfishness and the
limitation to which every finite being is subject. Others will rather hinder
him from venturing too far out, say to him: ``Spare yourself! This shall by no
means happen to you!'' ``A human being can demand of himself exertions that
the most well-meaning friend, if he knew of them, would advise against\dots\
Everyone who has in truth ventured his life has had the standard of silence.''
And this is all the more important, since it is only by venturing that one
becomes rightly aware of oneself, gets one's innermost being forth!

d) It is of importance that one act \emph{at once}, as soon as the right
knowledge is won. If an interval is inserted, then ``the knowledge boils
away.'' And this will then easily be the consequence when I wait to apply it
until I can transform the world, instead of beginning with what I have right at
hand: myself. Hence it comes that our actions relate to our understanding---not
as the faithful imprint of a picture, but---``as the blotting-paper to the
writing by which it has lain!''

e) ``Every human being is gloriously constituted, but what ruins so many is,
among other things, also this unhappy talkativeness between man and man about
what is to be suffered, but also matured, in silence.'' The single
individual---``not a single distinguished one, but every single one''---stands
above the race. Here is the properly human, that by which the human being
distinguishes himself from the animal. About the highest in me I therefore have
no permission to speak with other human beings.---Hereby is understood, what has
already been touched on, that irony is a transitional form between the aesthetic
and the ethical. The ironist can (like Socrates) be seized by the ethical
demand, but he is also seized by the opposition between what he thus has in his
inner life and the outer reality; the incommensurability between the inner and
the outer he expresses by treating himself as a finite and indifferent thing
among existence's many finitudes and indifferences. An ethicist can use irony
as an incognito, in order not to become comic through the disproportion between
what occupies him and what occupies the world. He thereby distinguishes himself
from ``an immediately enthusiastic person who bawls out into the world early and
late, and always on the buskin plagues people with his enthusiasm, without
noticing that it does not enthuse them.'' But the ironist need not be an
ethicist; he can have practiced the mere form. Kierkegaard hereby corrects his
own earlier, more negative conception of Socratic irony in his doctoral
dissertation.

The great ethical significance of the concept of ``the single individual,'' in
the direction both of realization and of inwardization, does not need to be
demonstrated more closely. By it there is pointed, with sure energy, to what is
and remains the ethical's first and most important domain, and at the same time
its source and its constant abode: the single personality and the relation
between the ideal that has here formed itself and the willing that here stirs.
It is the first ethical idea, and it is emphasized in Kierkegaard partly over
against the evaporation to which the contemplative and aesthetic tendency
easily leads, in that the will is left out of consideration and the ideal
becomes an object of mere beholding---partly over against the tendency to lose
oneself in the social and historical way of regarding things, which does not
worry about what goes on in the single inner life, provided only that the race
on the whole or on the average shows changes in a certain direction. It seemed
to Kierkegaard that existence would become meaningless if one did not see the
highest value in the single individual: for how else is one to explain ``the
divine prodigality that uses the infinite host of individuals in one generation
after another to get the world-historical development going''? How comfortless,
if the highest were to be ``the play of colors of the generation, like that of
the herring-shoal in the sea,'' while ``the single herrings are not worth
much''! Otherwise if the highest task lies in and for the single subjectivity:
though the individuals be numberless as the sand of the sea, still the task of
becoming a personality is set to each of them. World-history one must then leave
to God, ``the royal poet,'' who alone can survey it.---``To the category of the
single individual,'' says Kierkegaard rightly in his ``Report to History'' (\emph{The
Point of View for My Work as an Author}), ``my possible ethical significance is
unconditionally attached.''

\medskip

\noindent\textbf{4.} The concept of the single individual as the chief concept
in ethics corresponds to the proposition ``Subjectivity is the truth'' as the
chief proposition in the theory of knowledge. In analogy with this proposition
one could set up the one: Subjectivity is the good. It does not depend on the
content, the object, one relates oneself to in one's acting, or in one's
knowing, but it depends on the inner, subjective movement. And Kierkegaard has
really also set up this proposition, in that he expressly declares that the good
cannot be defined, since it is nothing different from freedom (the will,
subjectivity) (\emph{The Concept of Anxiety}). But the same formal and inward
willing could indeed be thought to have the most different content! If Kierkegaard
had wanted to make earnest of his principle of subjectivity in ethics, then he
must not have isolated the single individual from all relations of reality, all
relations to other human beings: for it is only here that the ethical subjectivity
can acquire a real content. An exclusive self-occupation is only possible in a
monastic life, where there are no real tasks, but where the labors undertaken
serve only for formal exercises. Here again it shows itself that the ethical stage
gets no independence in Kierkegaard. All the ethics he acknowledges is in reality
religious and ascetic. The only content is the single individual's obedience to
God. The ethical stands for him in absolute opposition to everything that can step
forth and acquire significance for other human beings or in history. He could
imagine ``that God, without doing wrong, and without denying his essence, which is
love, could create a human being equipped with abilities like no other, and set
him in a remote place and say to him: now live through, in exertion like no other,
the human; work, so that the half might be enough to transform a contemporary age;
but you and I, we are agreed on this, \emph{all your striving shall have no
significance whatsoever for any other human being}, and yet you shall, you
understand, you shall will the ethical, and you shall, you understand, you shall be
enthusiastic, because this is the highest.'' When Kierkegaard accuses the social and
historical way of regarding things of establishing an incomprehensible prodigality
in the world, then it seems to me that he here squanders quite as much of human
powers in his zeal to maintain the formal obedience, the subjective tension in the
inner life, which for him is really the highest. To use an image I have already used
about this (\emph{On the Foundation of Humane Ethics}): Kierkegaard's ethics is here
realized by the Egyptian monks who spent their lives watering sticks they had planted
in the sand. It was a work that had absolutely nothing ``social'' or ``world-historical''
in it, and where there was truly need for ``repetition'' and for subjective exertion, if
it was to be continued through a life. If obedience is the highest, no matter what it
concerns, then here the highest is accomplished. But prodigality it is, for no real good,
no purpose is attained thereby.

The purely subjective and formal ethics for which Kierkegaard pleads overlooks that the
single personality is precisely first developed by life in the real human relations and
through the work on tasks that have real significance, apart from the exercise attained
by the work itself. The social ethical tasks are all such as aim at furthering the
personal life in as many as possible; it is not externality, not compliance with ``what
the age demands'' or world-historical self-importance, but a sound human feeling that
wills that our work shall have more significance than merely being an exercise-lesson for
us. When Kierkegaard himself thought that his ``\emph{ethical} significance'' was attached
to the category of the single individual, did he then merely mean that it had been
beneficial for himself to occupy himself with this thought---or could he not here have
written ``\emph{historical} significance'' instead of ``\emph{ethical}'' significance?---Only
the intention of the action matters, according to Kierkegaard; the effects of the action
are indifferent; ethics has only to do with the intention, world-history with the effects,
and ethics has not to worry about world-history. But this separation between intention and
effect is untenable. It must be the task to foresee as many of my action's effects as
possible, and if they do not occur, or if wholly different ones occur, then it can be my
fault. All ethics rests on its being possible, to a certain degree, within a certain range,
to foresee the effects of our actions. And we give up (when we are not ascetics) such actions
whose effects are worthless.

Kierkegaard's forced one-sidedness on this point displays itself in that he can speak of human
relations of communal life only in a mocking and sneering way. ``It may indeed,'' he says thus
in one place, ``be very praiseworthy for the single individual that he is a councillor, a capable
worker in the office, first lover in the Permanent, next to a virtuoso on the flute, king of the
marksmen, director of an asylum, a noble father with dignity, in short a devil of a fellow who can
do \emph{both}--\emph{and} and find time for everything.'' However apt his mockery is of the
finitude and externality in human bustle, it is nonetheless strange that an author who wants to
depict the ethical stage as a special form of life has no other tone to speak of human communal
life in than this. It is misunderstood idealism. It is a romantic lack of ability to find the
ideality in the seemingly prosaic and everyday. It is also a lack of ability to find the significance
that relations and endeavors which seem low and comic in the eyes of a distinguished romantic can
nevertheless have for the development of the personality. At the same time as Kierkegaard, under the
depiction of the ethical stage, hastens forward toward the religious stage, he has retained not a
little of the aesthetic in himself. In this respect the following journal entry is especially
characteristic: ``The ages one can learn from with respect to ideality are: the child, the youth, the
young girl, the old man;---from the active man, from the busy housewife one does not learn in that
respect, and why not? because they are essentially occupied with the purposes of finitude.'' Here it
shows itself that Kierkegaard has forgotten his own book on repetition; he now praises hope and
recollection as the only forms of ideality; reality itself becomes lost to him. For if he is right
that there must always be striving and working \emph{by} single individuals, it is no less true that
there must always be working and laboring \emph{in} and \emph{on} single, definite, and therefore
finite tasks. Here the concern is for the ethicist to have the psychological and historical eye that
can see the great in the small, the significant in the insignificant. But both psychology and history
are regarded by Kierkegaard as the enemies of ethics.

Finally, I cannot hold back the remark that Kierkegaard's absolute rejection of all social ethics is
also connected with his conservative disposition and absolute respect for all authorities. He who at
an audience with Christian VIII condescended to say to the king: ``Your Majesty's only misfortune is
that your wisdom and shrewdness are too great and the country too small; it is a misfortune to be a
genius in a market town'' [alas, the little country could very well have had use for the whole royal
wisdom---and more besides],---he who in the policeman's baton saw a reflection of the absolute principle
of authority,---he thought at bottom that everything concerning society was the government's business to
take care of, and that it was only self-importance and externality when single individuals thought about
it. The single individual had to leave the public to itself and attend to his own private exercises in
ethics.---When 1848 came, the only point of view he had was that it was an absolute dissolution, and that
Denmark's downfall was near, not through the danger from Germany, which he thought imaginary, but through
political freedom.

\medskip

\noindent\textbf{5.} Really, in Søren Kierkegaard the ethical only comes to be in order in the same
moment to be recalled. It goes with it, to use one of his own images, as with the child who stuck its
head into the world but at once drew it back again when it saw how evil the world was. The ethical sets up
an unconditioned ideal, which is to be brought into reality, whereas it is not its task to bring reality
up to the ideal. The ethical does not haggle, holds fast to its demand (which it is \emph{impossible for}
the human being to fulfill) without ``letting itself be disturbed by the chatter that it does not help to
demand the impossible.'' It then runs aground---and through repentance the transition to the religious then
occurs. ``The ethical sphere is a sphere of transition, and therefore its highest expression is repentance.''

Ethics stands, then, from the first in a purely negative relation to the human being's power and drive.
The Greek ethics, which willed a positive development of the human powers and drives, Kierkegaard does not
acknowledge as a real ethics. And consistently: the possibilities he has referred to psychology, the effects
to world-history; ethics retains only the impossibilities. There is therefore in Kierkegaard also a tendency
to push out the ethical as a special stage. This holds from the first in the case of woman; for since she
lacks the ability for deliberation and resolution [!!], she must at once pass over from aesthetic immediacy to
the religious. But it really comes to hold in the case of man too. ``Between poetry and religiosity,'' it is
said, ``worldly life-wisdom performs its vaudeville. Every individual who does not live either poetically or
religiously is stupid.'' Here the ethical stage is eliminated \emph{in optima forma}. Kierkegaard's lack of
ability to discover the ideality in real life shows itself here again. It is certain that the ethical is
doomed to death when it excludes the poetic. But there is also a poetry of life which is only experienced and
felt during the work itself, only breaks out at the will's collision with life's hard flint, and which the
hermit oscillating between aesthetics and asceticism does not know. He speaks here like the blind man about
colors.---For the rest, it must still be noted that the opposition between the poetic and the religious does
not correspond to the opposition between Kierkegaard's aesthetic and religious stage. By living poetically he
must understand something other than living like the aesthetes or hedonists he has depicted. This points to the
previously demonstrated defect in the depiction of the aesthetic view of life.

\subsubsection*{$\gamma$.\quad The Religious View of Life}
\label{sec:4Bb-gamma}

\noindent\textbf{1.} It is only through religious presuppositions that
Kierkegaard's ethics acquires a content, or perhaps rather an object. It is
the relation to God, the eternal being qualitatively different from the
human, and to the eternal blessedness that can be won by the single
individual through the relation to God---this is what lies at the basis of
the ethical striving. Kierkegaard therefore does not really know the
ethical; he knows only the ethico-religious. This is very closely connected
with the fact that the religious for him is absorbed in religious ethics. The
dogmatic he indeed presupposes, but it does not condition any special
relation; the will is at once taken into service, and there is no time for
contemplation or mysticism. Neither does he acknowledge the religious as
comfort---or rather, he acknowledges the comfort in religion only in such a
way that he at once adds: this comfort brings with it a hitherto unknown
suffering.

Here too there can---as in the ethical view of life---be traced a change in
Kierkegaard's conception. Earlier he had, in the writing \emph{Fear and
Trembling}, depicted the relation between the ethical and the properly
religious as a relation of opposition, arising in that, in virtue of the
human being's personal relation to God, a break occurred in the universal
ethical law. This is shown in a presentation borne by mighty pathos, by
Abraham's example: he is willing to sacrifice Isaac at God's command,
although the ethical law says: thou shalt not kill. The religious faith is
the single individual's affair, because the suspension of the ethical that
it can demand cannot be made intelligible to others, since it is conditioned
and explained only by the individual's own personal God-relation. The
ethical, on the other hand, is the universal, that about which there can be
common understanding among human beings. Kierkegaard stands, in this
conception of the ethical as the universal, here still---as in \emph{Either--Or}
and in \emph{Stages}---closer to Hegel than in the purely philosophical
account in the \emph{Concluding Unscientific Postscript}, where, as
previously emphasized, ``the single individual'' and his personal, isolated
existence become a chief concept for the ethical view of life too. According
to Hegel, ethics was essentially social; the spheres of ethical life were
family, civil society, and state, where universal rules hold, and where the
single individual is only a moment. With this conception Kierkegaard breaks
first on religious grounds, in order to maintain the personal God-relation in
its absolute validity; but the break, which is one with the bringing-out of
the concept of the single individual, acquires a retroactive significance for
the conception of the ethical. The conception he develops in the
\emph{Concluding Unscientific Postscript} is that the ethical first acquires
its proper object, its absolute purpose, through the religious, and that, on
the other hand, the religious is essentially of an ethical kind, is a life in
striving, willing, acting. That there must occur a transition from the
ethical stage to the religious stage then really comes to mean that all
ethics, more closely regarded, must be religious ethics if it is to be
anything other than externality and loss in the relative. It shows itself
that the single individual can maintain his own personal reality only by means
of the relation to an absolute personality.

By the relation to God and eternal blessedness there is set the single
individual an absolute purpose, an absolute object. The religious is different
from the ethical in that, while the single individual ethically has to do only
with himself, his own reality, he is religiously infinitely occupied with a
reality other than his own. But besides the absolute object and the absolute
purpose, the single individual stands in relation to a surrounding world of
finitude and a circle of relative purposes that belong there. What matters is
to relate oneself absolutely, with unconditioned respect, to the absolute
purpose, and to relate oneself relatively to the relative purposes; for it is a
self-contradiction to relate oneself unconditionally to, to set everything on,
a relative, finite purpose. But the task thus set is very difficult, since
there is a yawning deep between the absolute purpose and the relative ones, and
since there is at the same time an immediate urge in the human being to be
absorbed in (relate himself absolutely to) the relative purposes. And to this is
added, further, that the human being feels pleasure and satisfaction in using his
own power: but over against God and the highest purpose his work is nothing; here
the concern is precisely to see oneself in one's nothingness. The tearing-loose of
the will from the finite purposes and relations given in the world of experience,
the learning to die away from immediacy, demands a painful exertion that cannot be
applied once and for all, but must again and again be repeated; and when to this
is now added that one---in spite of all exertion---is to learn to be as though one
were nothing, then it is seen that the religious life is an initiation into
suffering and self-annihilation. ``What wonder that the Jew assumed that the sight
of God was death, and the heathen that the God-relation was the beginning of
madness!''---

The religious suffering distinguishes itself from other suffering in that it is
not accidental, not called forth by outer causes that could also be thought absent,
but is necessarily effected by the very nature of the religious relation. The
relation to God is the relation to a being absolutely different from the human, who
cannot stand over against the human being as superlative or ideal---and yet is to
rule in his inner life! Here is a doubleness that cannot be dispensed with, but gives
birth to ever new pain when it is to be held fast. It holds, then, of the religious as
of the ethical: the task is not to develop, heighten, and ennoble human nature, not
``to bring reality up into ideality,'' but ``to bring ideality into reality,'' to
bring an absolutely new element into nature---an element as hostile to nature as the
absolute (which demands \emph{all} power) is to the relative (which nonetheless demands
\emph{some} power)! The absolute is cruel, because it demands \emph{everything}.

It is a wholly new medium, opposed to its nature, that the human being is to live in
when he enters into the religious relation: though a finite being, he is nonetheless to
live in the infinite and absolute. He finds himself here in the world like a fish on dry
land: he has come into a world where wholly other instincts are required than those that
lie in his nature. It is a nature-opposed life that he is to lead.---Henrik Ibsen's
\emph{Brand} has here an image that in high degree recalls the one Kierkegaard uses to
depict religious existence:

\begin{verse}
I pictured to myself a dark-frighted owl,\\
a fish with water-dread. Loudly I laughed;\\
I would have smuggled them out of my mind,\\
but they hung on with tooth and claw.\\
What came of that fit of laughter?\\
Why, of the dimly felt discord\\
between the thing as it is,\\
and the thing as it ought to be,---\\
between the having to bear\\
and finding the burden all too heavy.
\end{verse}

\noindent A fish with water-dread, that must nonetheless live in the water, or a fish
(with its nature suited only to life in the water) on dry land---it comes to the
same.\footnote{On the religio-psychological type that under various historical conditions
has found expression in images of this kind, see under \emph{Religionsfilosofi}, § 36.}
When Brand tells that as a boy he laughed at this thought, then Kierkegaard too has been
aware that in the disproportion, the contradiction peculiar to the religious, there could
lie something comic; but he holds nonetheless that the religious is raised above this
comedy by the conscious taking-over of the contradiction and of the suffering.---

Just as irony was a transitional form to the ethical and could serve as an incognito for
it, so humor is a transitional form to the religious and can be its incognito; humor sees
the finite in its nothingness and indifference in the face of the infinite; but the
humorist need not himself positively have a God-relation. The religious person can use humor
in order to hide the inward movement in his inner life, which can never lay itself directly
and completely open in any positive utterance. He appears as ``the knight of hidden
inwardness''---until the religious relation assumes for him such an energy that it demands
outward appearance and a break with the world.\footnote{On Kierkegaard's conception of irony
and humor see more closely in my book \emph{Den store Humor}, §§ 16, 18, 30, 39.}

\medskip

\noindent\textbf{2.} By this conception of the religious Kierkegaard has from the very first
cut the bond between it and human nature. It is a foreign thing that comes into the world and
is to be forced together with nature without ever being able to become one with it. He proceeds
from an idea of perfection that is formed without any regard whatsoever to real human nature.
For the historical conception, on the other hand, the religious ideas are anticipations and
idealizations on the basis of what is given in experience and moves in the human spirit. Under
the influence of inwardly and powerfully moved feeling there form themselves, for imagination
and thought, ideas and exemplars that are able to react back on the soil from which they have
grown up. But if one cuts away the possibility that the highest ideals of human life can have
developed out of life itself, out of the mind moved by life-experiences, then one really cuts
away at the same time the possibility that the ideals can work on nature. The qualitative or
absolute difference abolishes the possibility of a positive relation. The mighty power of the
religions over mind and will rests on their being the great treasuries in which the human spirit
has laid down some of its most inward experiences. But the religious striving contradicts itself
when, in its zeal to set religion's object as high as possible, it places a gaping gulf between it
and the life whose exemplar it is to be. A God who is not ideal and exemplar is no God. There have
therefore also constantly asserted themselves ethico-religious misgivings about going along with
the assertion that the divine being should be qualitatively different from the human. It is on
this point, where the religious passion establishes the absolute difference between the divine and
the human, that the break occurs between the religious and the humane ethics; for an ethics grounded
from a principle absolutely different from human nature must necessarily---when it is
consistent---become a life- and man-hostile doctrine. But it is at the same time an ethics
contradicting itself: for when the absolute relation lays claim to \emph{all} power, how can there
then be \emph{any} power left over to live in the relative relations?\footnote{Cf.\ what
\emph{Brøchner} has already remarked on this in his acute critique of Kierkegaard's conception,
\emph{Problemet om Tro og Viden}, p.~221: ``The absolute purpose must, conceived as Kierkegaard
conceives it, take up the whole power of the human spirit, and let no [power] be left over for the
relative relations.''} Here even the great qualitative dialectician enters into a compromise, a
both-and, and only thereby does he halt the complete self-annihilation.

Really Kierkegaard was on the same road as Schopenhauer: the road to Nirvana. Just as Schopenhauer,
in the midst of the nineteenth century's culture, declared Buddhism's life-denial to be the highest
truth, so Kierkegaard, who in several respects was his kindred spirit, proclaimed strict, ascetic
Christianity to be the highest---in the last instance the only thing. It is, in ethical and
cultural-historical respects, of the very greatest interest to see two of the most superbly gifted
thinkers of our time pass such a sentence of death over all our vaunted cultural development. It
indicates that there must be defects and injuries in this which we, with frivolous optimism, are
accustomed to overlook. In Leo Tolstoy there comes forth in the most recent time a partly kindred
train of thought; yet he stands closer to the natural and human.

From Schopenhauer, Kierkegaard distinguishes himself by his ethical-practical tendency. Schopenhauer
bows to the ascetic ideal, but himself prefers the intellectual and aesthetic liberation of the will.
Kierkegaard works his way constantly forward in order, if possible, to become one of those on whom the
highest determinations could find application. He stakes his personality quite differently, seeks the
practical struggle and suffering as an honor---and finds them too.---In his last years Kierkegaard
occupied himself with the study of Schopenhauer's writings and was much taken up with them. But his
views were then fully developed, and he had merely to note the agreements and oppositions between
himself and the German thinker.

\medskip

\noindent\textbf{3.} The religious view of life appears in two chief forms that are so different in
their character that the transition from the one to the other is possible only by a leap, a qualitative
change. The difference between them consists especially in what is the object of faith. The first form
of religiosity (Kierkegaard calls it Religiousness A) relates to the eternal as to the general
background of life and existence. The contradiction that here calls forth suffering is then only the
one connected with the fact that the consciousness of eternity and the relation to the absolute purpose
are to be held fast in the midst of the world of finitude and time. The single individual feels at the
same time, the more deeply he takes the relation, ever more the distance at which he is from the highest,
and notices the resistance he himself makes to it. For the nearer one comes to the highest, the more one
notices one's distance from it! The advance becomes, in so far, here a retreat. There arises a
consciousness of guilt that increases the pathos peculiar to the religious standpoint. But even so, this
form of religiosity keeps itself within what is accessible to the natural human being. This form of
religiosity lies within ``immanence,'' presupposes no decisive break in the natural world-order, and is
possible within paganism.---In \emph{Philosophical Fragments} Kierkegaard had referred to Socrates as
the representative of this form of religiosity. This can scarcely be historically defended. Such a view
of life as the one Kierkegaard depicts under the rubric Religiousness A does not occur within the Greek
world before toward its close: in Neoplatonism. Only there is such a decisive weight laid on the relation
to the eternal in definite opposition to all outer and human purposes. The Greek world-view was a calmed
harmony; life's natural elements were united into an ethical work of art, but the beyond was not set as
the highest or only purpose, however deep a feeling of life's dark sides and of sensuousness's resistance
to clarity of thought and purity of will may often be expressed by poets and philosophers.

From this humane or immanent religiosity is distinguished the paradoxical, transcendent religiosity
(Religiousness B). Here all determinations are sharpened by the fact that the contradictions become greater.

The single individual relates himself not merely to existence's omnipresent eternal ground. The eternal
and divine has appeared in time, in historical shape, as a single human being who has suffered and died.
And precisely the fact that the essential revelation does not spread out in existence, but is restricted
to a definite time and place, expresses that existence for the rest is forsaken of spirit: nothing has
worth, nothing has decisive significance except this one figure---and on the relation to it, to ``the God
in time,'' everything depends. The object of faith is the paradox in the highest sense, the
self-contradictory, the absurd. Only by one way is it possible to enter into faith's relation to the
highest paradox: when the consciousness of guilt is potentiated into consciousness of sin, that is, when
the single individual discovers in himself not merely a resistance to the eternal, but a complete and
self-willed change in his nature. The consciousness of this is called forth and conditioned precisely by
the appearance of the divine paradox: only by a revelation can the human being discover the total change
of his nature and of human nature on the whole. By the paradox as the object of faith and by the
consciousness of sin as its presupposition, the greatest immersion in existence, the greatest inwardness,
and the greatest suffering are reached. It is the standpoint that appears in Christianity.---

As a moment that sharpens the suffering at the highest religious standpoint, Kierkegaard names the pain of
sympathy that arises in that the single individual can no longer sympathize with every human being as a
human being, but essentially only with the Christians. His nearest ones he must perhaps even hate: ``for is
it not as it were to hate them, when he has his blessedness attached to a condition which he knows they do
not accept?''---Not always has Kierkegaard given sympathy the right to speak a word of pain. In his journals
he cites with assent a Mohammedan legend, according to which Adam exclaims: ``O Lord, only save my soul; I
care neither about Eve nor about Abel.'' The saying ``Take heed to yourselves'' (Mark XIII, 9) he interprets
in a similar direction. And even where he lets the pain of sympathy come forth, he nonetheless does not let
it acquire influence on faith; he does not draw, like Schleiermacher, the consequence that with this pain of
sympathy no eternal blessedness is possible. Psychologically and ethically regarded, here lies perhaps the
greatest and most crying, not to say revolting, of the contradictions in the paradoxical religiosity.
Kierkegaard has named this contradiction---but he nonetheless places it last in the row. It is a consequence
of the principle of the single individual: what does it concern me \emph{when all is said and done} whether
the others also become blessed or not!---

The romantic-speculative doctrine of harmony, which had already received several fundamental shocks by the
qualitative jerks between the stages, now finally runs completely aground on the sharp and steep cliff of the
paradox.---Of Kierkegaard's conception of Christianity on the whole we shall have occasion to speak when we
come to his last appearance.

\subsection*{c.\quad The Standard}
\label{sec:4Bc}

\noindent\textbf{1.} In his presentation and evaluation of ``the Stages''
Kierkegaard applies a definite standard. In a single place he himself
expressly brings out this standard, in that he seeks to show that ``all
conceptions of existence rank in relation to the determination of the
individual's dialectical inwardization.'' But the inwardization grows again
with the qualitative and quantitative oppositions or contradictions that
assert themselves in the individual's inner life. It is the degree of
tension that determines the height of the standpoint. It is the increasing
tension that---without, however, the higher degrees of tension being
understood as successively arisen out of the lower---leads from stage to
stage, and at last (when there is no bargaining-down or sluggishness) to the
paradoxical religiosity, in that life without supernatural help becomes
despair. Kierkegaard constructs, by means of this formal standard, a series
of life-stages, interrupted at definite nodal points. The relation of
opposition thereby appears in two forms: as successive, by the leap, the
crisis, that effects the transition from one stage to another, and as
simultaneous, within one and the same stage, by the tension between the
conflicting elements that are to be held together.

One will scarcely go wrong in supposing that in this formal standard there is
an after-effect of Hegel's method, which consisted in driving thought forward
by the constant positing of oppositions, which it then became the task to
reconcile. The difference is that Kierkegaard does not believe that the
oppositions unfold out of one another, nor that their reconciliation occurs of
itself. He therefore sets the concept of repetition in the place of mediation.
But he has nonetheless preserved a remnant of the method, even if he applies it
far more cautiously. He constructs by its help possible standpoints and then
looks to see how they are found in reality. Against this procedure nothing in
itself can be objected; it marks, on the contrary, a great advance in relation
to the speculative method, which did not distinguish between the constructed
possibilities and the given realities; but it is the very justification of the
standard that must be asked about. On the answer to this question depends the
final significance that can be attributed to Kierkegaard's ethico-religious
standpoint.

\medskip

\noindent\textbf{2.} Of the leap there has been spoken earlier. I here return
to the successive relation of opposition in order to investigate whether
Kierkegaard is right in attributing to it so decisive a significance for the
personal life as he does.

A transition is for him always a break. Of crises there can, according to his
conception, really be no question ``within immanence,'' that is, within a view
of life that assumes a natural and continuous development in the spiritual
domain. Here Kierkegaard's polemic against speculative philosophy's smooth
transitions, which were consummated only in imagination or on paper, has
decidedly led him too far. The significant thing in his appearance is that he
maintains the significance of the \emph{real} transitions in opposition to the
\emph{thought} ones, in which people wallowed, in that, over the historical
knowledge they had won about various \emph{possible} standpoints, they forgot
to investigate on which they now themselves \emph{in reality} stood. But the
difference between the thought and the real transitions need not consist in the
former being continuous, the latter not. A transition by which a goal
significant for the personal life is reached can very well occur continuously.
The water moves forward in the streaming river just as well as in the plunging
waterfall, and it is only for the outer, sensuous apprehension that the latter
movement has more reality than the former. However eagerly Kierkegaard fights
for life, he nonetheless does not here let life come to its right; he does not
see the small things, the small increments, that quietly join together and
imperceptibly often produce what is prized as great and high. Only he gives life
its right who acknowledges that it has many ways and forms for its unfolding.
Why is it to be narrowed down according to one method? A jerk, a plunge, a
crisis can be necessary where a resistance is to be overcome, a decisive turn
undertaken. But even then continuity is not broken when one looks more closely.
The stillness before the catastrophe (as when---to use one of Kierkegaard's most
splendid images---the beast of prey lies quite still before the decisive spring)
signifies precisely that power is being gathered, which is released in the moment
of decision. Thus the water is dammed up, so that, when the sluice-gate is taken
away, it can plunge with gathered power. It must be the task of knowledge in the
spiritual domain, as everywhere, to investigate whether the apparent leap could
not be explained as a release of energy long gathered in stillness. There then
remains the great and significant difference between such a spiritual
development, which resembles more the evenly streaming river, and such a one as
resembles more the waterfall's plunge. Here one must acknowledge, if one will
stand on the very ground of life, that the inner and outer conditions are not
the same for all, wherefore development for different people occurs in different
ways. Such an acknowledgment is really contained in the proposition:
Subjectivity is the truth. But Kierkegaard has, as already demonstrated, not been
able and not willing to draw this proposition's great consequences.

\medskip

\noindent\textbf{3.} Still more does he sin against life by making tension and
suffering into criteria of a standpoint's height. Here something similar holds as
with the crisis or the leap. The relation of tension can be necessary for life's
truth and power, a sign that it does not push away tasks that it can and must
solve. Each time a new element, a new relation comes into force and is to be taken
up into life's coherence, there can come a relation of tension before the taking-up
has succeeded. But in itself the relation of tension is nonetheless a sign of a
state of transition, and the standard cannot be sought in it. There can be richness
and fullness in life even when it masters its elements and is able to maintain its
unity; even when oppositions assert themselves, they can enrich without splitting
and paining. In the general nature of consciousness as synthesis, comprehending
unity, it lies that not only opposition but also concentration must assert itself
so long as the life of consciousness exists. When one fetches life's standard from
life itself, one cannot one-sidedly see the decisive thing in the relation of
tension.\footnote{A certain spiritual kinship between Kierkegaard and Nietzsche
shows itself especially in the standard for the work of spiritual life that they
both apply. Yet a kinship appears also at other points, thus in their weight on the
concentrated personal life over against absorption in art and science, and in their
circling about the problem of repetition. In a treatise on Kierkegaard and Nietzsche
(\emph{Mindre Arbejder}, Third Series, pp.~189--195) I have demonstrated this.}

For the humane ethics, which fetches the standard from life itself, and for which the
ethical development can consist only in a higher, harmonious unfolding of the life
that is, it is not to drag the ideal and the demand down to acknowledge that they must
be determined according to the single individual's nature. That for which, in spite of
all seeking and awakening, no germs or possibilities are found in the individual's
nature, cannot be an ethical law for him. And even if tension and unrest \emph{could}
be awakened, this would only be defensible if there were powers present to overcome the
tension. The responsibility lies with the one who abolishes the secure harmony without
being able to lead forward to a new harmony. The Greek ethics, which Kierkegaard would
not acknowledge, in reality contains the fundamental thoughts with which all ethics that
is not life-hostile stands and falls. The ethical good is the harmonious unfolding of
life's powers in the single personalities, which again demands that harmony be brought
about between the single personalities' development in relation to one another.

But Kierkegaard will give no definition at all of the good. The only standard he can
get therefore becomes formal and is determined by his religious conception. His
construction of the stages is (like his theory of knowledge) determined by the purpose:
to prepare the paradoxical religiosity, to bring the psychological conditions to hand for
the passionate surrender to the absurd. Therefore he constantly operates with relations of
tension and lets the whip-lashes of the increasing suffering fall ever more violently on
``the single individual's'' back. What would be to admire in such an ethics, when it was
realized, would be the passionate energy, the strength of will, with which one holds out in
spite of the tension. But such qualities can also be displayed in the service of wholly
other purposes, as when the gold-maker in his way seeks ``the absolute,'' which here too is
cruel, early and late, with increasing passion and pain, ``hating'' and sacrificing wife and
children for its sake. (Balzac: \emph{La recherche de l'absolu}.) The humane ethicist stands
over against the ascetic as the chemist over against the alchemist. He admires the energy, the
endurance, the passion, and the many new and deep thoughts produced during the solitary work;
but he wishes all these noble powers applied in the service of other purposes.

The purely formal way of regarding things that Kierkegaard applies in the presentation and
evaluation of the stages leads him to consequences he himself shudders back from. If
contradiction and suffering were the marks of the highest, then it would have to be---it seems
to be necessary---the absolute paradox where God not merely became man, but became it in such a
way that he was unrecognizable, because he lived quite like other human beings and in the usual
human relations, without either way of life or supernatural events and actions drawing attention
to him! And must he not at the same time be tried in what appeared to Kierkegaard, from his own
experience, to be the most crushing suffering, melancholy, the sickness of the soul? These thoughts
have gone through Kierkegaard's mind and have received expression both in \emph{Philosophical
Fragments} and in several places in the journals. ``God help the poor head that has to do with that
kind of doubt!'' he exclaims.---To this exclamation it is answered that only great and strong heads,
who from the presuppositions they have once assumed think consistently and do not halve their
thoughts, only they get that kind of doubt to do with.

For Kierkegaard the solution lay in a distinction between the \emph{absolute} paradox and the
\emph{divine} paradox. Since God is love, he would not be able to reveal himself in the form of the
absolute paradox, which would not be a real revelation, since the connection between God and man
could not be brought about. This solution here, at the conclusion of Kierkegaard's trains of thought,
makes the same effect as, at the conclusion of \emph{Brand}, the cry that is heard through the crash
of thunder: ``He is \emph{deus caritatis}!''---This appeal to love takes the ground away from under the
whole preceding formal construction and evaluation of life's stages and points out toward a law of
life that does not begin by demanding life's own power and drive broken. If the love here appealed to
were not ``qualitatively, absolutely different'' from what human beings can know and understand as love,
then life's inwardization and deepening would have to be attainable by another way than the one
Kierkegaard prescribed. But he does not, at any rate, give that appeal retroactive force. On the
contrary, suffering and disharmony as life's laws acquire an ever sharper and more prominent place in his
consideration. Suffering is like a fatalistic necessity, under which even the divine will must bow, and
which---precisely because the deity is nonetheless conceived as love---casts its dark shadows into
eternity's world of light. It is said in a journal entry from Kierkegaard's last years: ``In my childhood
I heard much about there being great joy, sheer joy in heaven; I believed it too, and I imagined God
blessed in sheer joy. Alas, but the more I think about it, I come rather to imagine God sitting in sorrow,
as the one who most of all knows what sorrow is.'' Here Kierkegaard draws the consequence that in another
connection he would not draw. The human sympathy that he would not let shake the dogma of the eternal
un-blessedness for human beings has here grown to shake the dogma of the very blessedness of the deity.
The voice of nature makes itself heard even in the highest theological circle of ideas: a confirmation of
the proposition that theology is psychology. After the result of the inner life's experiences the content
of the idea of God is determined.

% ============================================================
%  V.  SØREN KIERKEGAARD AND CHRISTIANITY  (pp. 127--149)
% ============================================================
\bookchapter{V.\quad Søren Kierkegaard and Christianity}
\label{ch:5}

\section*{A.\quad Personal Breakthrough}
\label{sec:5A}
\markright{V. Søren Kierkegaard and Christianity --- A. Personal Breakthrough}

\noindent\textbf{1.} After the enormous production that filled the years
1843--46, Søren Kierkegaard had a feeling that he had said what for the time
being lay on his mind. ``Now I am finished with the books,'' he wrote in his
diary in the summer of 1847. In the year 1847 itself \emph{Works of Love}
appeared, which belongs among his finest purely religious writings; but there
are in it no new thoughts of decisive significance for the understanding of
his view. On the other hand, there is noticeable in the notes from this year
intimations of an important and far-reaching change in his personal life.
Hitherto Christianity had indeed been for him the highest, and in his
religious discourses he had developed Christian thoughts. But in his
presenting and inculcating of the Christian view of life he had nonetheless
predominantly chosen the indirect method. He had let it be maintained by a
pseudonym (Johannes Climacus) who himself declared that he was not a
Christian. He could not yet, in the strictest sense, find that his personal
life and the Christian faith covered each other. What he especially had to
struggle with, melancholy, he had hitherto kept down---not by directly
letting the good cheer of faith combat it, but by diverting it, in that the
exertion of the authorial activity laid claim to all his power. So long as he
could produce, melancholy was checked. We have above brought out the rhythmic
relation between these two elements in his nature. But now he felt an urge for
a more inward, more direct settling with the dark power in his inner life: ``I
now feel,'' he writes in August 1847, ``an urge in a deeper sense to come to
myself by coming nearer to God in the understanding of myself\dots\ My
melancholy I must see to get hold of more closely. It has hitherto rested
deepest at the bottom, and the enormous exertion of spirit has helped to keep
it down\dots\ Now God wills it otherwise. Something stirs in me that points to
a metamorphosis\dots\ I shall therefore now keep still, \dots\ see to come to
myself, rightly to think my melancholy's thought together with God on the
spot. In that way my melancholy must be lifted, and the Christian come nearer
to me.''

The metamorphosis whose approach he noticed seems to have set in for him at a
definite time, namely at Easter 1848. The diaries from 1848 and 1849 point
back frequently to this time as a decisive turning point.---``I now believe
that Christ will help me to conquer my melancholy, and then I shall become a
priest,'' he writes on 19 April 1848. Hitherto he had held himself in
resignation in the face of ``the fundamental misery of his being,'' and the
production had, by its diverting influence, helped him to this. But shortly
after that breakthrough he writes: ``Now I am in faith in the deepest
sense\dots\ For God all things are possible; this thought is now in the
deepest sense my watchword, has won a significance for me that I had never
imagined.'' ``Only now, now in my 34th year, have I perhaps learned so much
dying-away from the world that for me there can be question of finding my
whole life and my whole blessedness in faith in the forgiveness of sins.''

What he had several times let his pseudonyms develop about the difference
between resignation and Christian faith (see for example \emph{Repetition} and
the depiction of the difference between ``Religiousness A'' and ``Religiousness
B'' in the \emph{Concluding Unscientific Postscript}), this he could only now
find again and recognize in his own personal life, as inner, real experience.
How the man's great honesty toward himself here comes strikingly forth! And
what a great ability has he not had, in the form of imagination and reflection,
to develop standpoints and views of life in full, concrete illumination,
without, however, succumbing to the temptation to confuse what he could in this
way construct with personal experience! He is in truth a \emph{critical}
philosopher, if the critical consists first and foremost in a sharp distinction
between what is hypothesis and speculation and what is given reality. Scarcely
anyone has been his equal in such criticism within the subjective world, which
was his world. And how one understands his program: to find difficulties! For
what so many think to reach easily and cheerfully, the simplest Christian
determinations, that required in his case a long development and strenuous work
to reach. He could rightly say that it was \emph{honesty} he willed. He demanded
it of himself before he demanded it of others.

\medskip

\noindent\textbf{2.} It could now seem as if his public activity could be
concluded, and he could withdraw to a parsonage in the country. His economic
circumstances must also have urged him to obtain a salaried position, since his
fortune would in a few years be used up. But the personal breakthrough led to a
new, religious production of a more positive and at the same time polemical
nature than the earlier one. Besides, he saw that he could not live without
producing. ``I keep the sorcery of productivity in order to forget all life's
shabby pettinesses'' (5 November 1846). He thought he had a talent that must
not be buried. ``Should the whole wealth of thoughts that still lies in my soul
be conjured back---that becomes a torment, an agony'' (1847). And there were
also other circumstances that made him think he ought not to halt his authorial
activity.

Offended that the satirical paper \emph{Corsaren} (The Corsair), published by
Goldschmidt, mocked and caricatured so much and so many who in his eyes did not
deserve it, while it frequently spoke of \emph{him} with admiration and
reverence, he demanded (in an article in \emph{Fædrelandet}) henceforth not to
be praised but to be ridiculed in this paper. From that time he became a
standing figure in \emph{Corsaren}, with his thin legs and short trousers.
Since he moved about much in the street and was well known, the public
attention to him was thereby increased in a perhaps not very pleasant way,
which he in his melancholy manner interpreted in the very worst sense. He was
indignant in his innermost at what he regarded as a mockery and attributed to
it far greater dimensions than it in reality had. It pained him deeply that the
plain people whom he inwardly loved and among whom he so gladly moved now---as
he believed---regarded him as a comic figure or perhaps even suspected him. But
most he was indignant that the esteemed and distinguished men in the literary
circles who had long, like him, been offended at the abuses of the press, now
left him in the lurch, or, as he thought, even got nourishment for their envy
from the treatment he suffered. It was, he thought, a downright national crime
that was being committed against him, and he had by this whole affair got a
proof of the wretchedness below and the weakness of character above. To the
disgust at ``the crowd'' there now came the aversion to the authorities.
Something was shaken that had hitherto stood fast for him.---While Kierkegaard
rightly saw, in the way in which town-gossip busied itself with his person, a
proof of the small-market-town character of our conditions, it can surely also
rightly be said that he himself took this affair in a small-market-town way. It
is with a strange feeling that one sees in his diaries numerous pages devoted to
this affair. In his hermit-stillness he expresses to himself the indignation
that people have seen that his one trouser-leg is longer than the other, and
this most damaging insinuation he then also at once denies---in the diary---as a
``lie.''---Were it not for the melancholy and the subjective expansion by which
his life's small events are potentiated into decisions of principle, one would
have to call his way of taking this affair petty and unworthy. It now acquired
no small significance for him; it contributed to holding him back at his post,
since he did not want it to look as if he fled from the popular talk.

To this was added the strong political movement, which in his eyes brought a
complete dissolution, and in the face of which the authorities, especially
Bishop Mynster, to whom he had hitherto looked up, appeared without due firmness
and authority, in his opinion. It is strange to see how untouched he is by all
that stirred in 1848. At the beginning of this year he writes in the diary:
``And so I sit here. Out there everything is in movement, nationality surges
through all\dots\ And now I sit in a quiet room (soon, no doubt, I shall be cried
down for indifference to the nation's cause); I know only one danger, that of
religiosity.'' How he regarded the outer political conditions is seen from the
following remark from the same time: ``The whole fear of Germany is an
imagination, a game, a new attempt to flatter national vanity.''

What most mightily moved him in these years was not what went on in the world,
but what went on in his inner life. And that was then also the essential reason
that he did not halt his authorial activity and withdraw to a quiet life in the
country. The breakthrough that had occurred in him in religious respects had let
Christianity acquire a personal reality-significance for him that it had hitherto
not had. The more strongly awakened religious consciousness in turn awakened a
sharper criticism of the existing Christendom, which found expression in the
religious writings he composed in 1848, in a period when the unrest in outer
conditions and personal difficulties in economic respects in many ways tormented
him. This ``new production'' he himself set above the earlier. The chief writings
were \emph{Practice in Christianity} and \emph{The Sickness unto Death}, to which
was added \emph{For Self-Examination} (of which last writing the second part was
first published long after his death). In these writings the Christian ideal was
expressed in its whole rigor, with a sharp judgment on the existing religious
condition and its representatives.

The question was for him now, in the first place: should these writings be
published before he sought ecclesiastical appointment, or later? Would he not
make himself impossible as a candidate for office if he published them before,
and would it not be dishonest to publish them later?---But this question soon
drowned in a question far greater for him: should they be published under his own
name or pseudonymously? The first would indeed mean that he himself personally
took it upon himself to make good the conception of being a Christian that was
maintained in these writings. And was it not his duty? Must he not now take all
the consequences? How far his deliberations in this respect went can be seen from
the fact that he even imagined that he could come to suffer the usual lot of
witnesses to the truth, to be slain by the mob under the secret co-operation of
the envy of the distinguished! He then investigated, in a separate essay (the
first of \emph{Two Ethical-Religious Essays. By H.H.}), whether a human being has
the right to let other human beings slay him for the sake of truth, and came to
the result that such a thing only the God-man and the apostle could have the
right to; an ordinary human being must not drive the matter out to this point.
Later he regarded the whole thought, in his own case, as hypochondria. But it
shows how strong the surge in his mind was in these years. Especially the diary
from 1849 is extraordinarily interesting in this respect. Now he scolds himself
for his weakness, that he will not venture the utmost, now for his pride or his
hypochondria, that he thinks himself placed in the utmost trial. It was the
latter conception that gained the upper hand. He indeed gave up seeking
ecclesiastical appointment (after he had nonetheless sought a conversation with
the minister and the bishop---without finding them at home). But he published the
said writings pseudonymously, this time choosing a name (Anti-Climacus) that was
to express that here was presented a conception of the Christian higher than the
one he personally could vouch for, just as his earlier chief pseudonym (Johannes
Climacus) had designated a standpoint that did not yet reach the one he himself
acknowledged. ``In contrast to Climacus, who said of himself that he was not a
Christian, Anti-Climacus is the opposite extreme: to be a Christian in an
extraordinary degree---if only I myself manage to be quite simply a Christian.''
---It was a ``battle of ideas'' that had been delivered in his inner life. It was
in reality a \emph{witness-to-truth struggle} that he here fought out with
himself, long before the great public witness-to-truth struggle broke loose. The
result was: ``\emph{I confess that I am not, in the strictest sense, a witness to
the truth.}'' Here we have even the very word used in his inner debate---the word
that then also grated so hideously in his ears when he later heard it used of the
representatives of the official Christianity.---The outer struggle is here clearly
seen as the continuation of the inner. So it always is where a struggle is
conducted with authority and earnestness.---As a deep humiliation he felt what had
happened: ``In one sense I would so gladly venture; my imagination beckons and
goads me; but I am precisely to learn to be so good as to venture in a lower form.
It is indeed the most finished and truest thing I have written; but the relation
is not to be such that I should be the one who almost judgingly rushes in upon
others---no, I am myself first to be educated by the same; there is perhaps no one
who is allowed to humble himself so deeply under it as I, before I am allowed to
publish it.''

\medskip

\noindent\textbf{3.} The fundamental thought in the ``new production'' is that
Christianity has really been abolished by being conceived merely as the religion
of mildness and comfort. People have accustomed themselves not to understand that
the compassion and love that express themselves in Christianity are of a wholly
different kind than what people usually understand by these words. This is
inculcated by a masterly development of how Christ must have looked in the eyes
of contemporaries, and how questionable a matter, humanly speaking, it could be
to let oneself be helped by him. In the strictest, most cutting way it is
maintained that there is in Christianity a wholly different conception of misery
and of help than the usual, human one. But now people think afterward to have
learned that Christ was God, and by the help of this smuggled-in result they do
not come into the relation of contemporaneity, while they nonetheless take the
comforting and mild words to their own credit.

Only rigor can help here. The true authority has left the Church: ``Those who were
to command became cowardly; those who were to obey became insolent. Thus was
Christianity abolished in Christendom---by mildness. And now there lives in the
existing Christendom, where there is indeed never any question of rigor, there
lives a pampered, proud, and yet cowardly, defiant, and yet soft generation, which
occasionally hears these mild grounds of comfort delivered, but which scarcely
knows whether it will make use of them, even when life smiles most beautifully,
and which in the hour of need, when it is seen that they are really not so mild
after all, takes offense'' (\emph{Practice in Christianity}).

What must in any case be demanded of every single individual as well as of the
existing Church is sincerity, by admission of how great a distance one finds
oneself from the ideal. It is, after all, the least that can be demanded, that one
not lie oneself into an ideality one is very far from possessing (cf.\ ``The Moral''
in \emph{Practice in Christianity}).

``The strictest thing shall be heard; one shall not have permission to draw a line
straight through it and ignore it; it shall be heard, in order that one may humble
oneself under it; but it shall not be proclaimed in such a way that one would cruelly
force people to want to be spirit according to so terrible a standard. Here is my
difference from Mynster: he wants to have it kept quite silent. I want to have it
said, and am then, for the rest, willing to declare that, when I say it, it is only
poetic, since my life is far from being so spiritual. On the whole I think it is
ethical respect that is to be introduced.''

If only the admission of the distance from the ideal came from the side of the
existing Church, Kierkegaard would raise no strife. It was the personal truth he
demanded acknowledged; for it he could rise to battle, even though he did not dare
stand as the champion of the ideal itself.---But the admission did not come; instead
Bishop Mynster let Kierkegaard know, through an intermediary, that he in the highest
degree disapproved of \emph{Practice in Christianity}. A word from Mynster could have
prevented the strife. ``If I collide with the existing order,'' it is said in the
diaries some years after, ``it is Mynster's fault.'' Out of piety toward the old
bishop he nonetheless held himself back---until the challenge came.

\medskip

\noindent\textbf{4.} Yet, for a right understanding of Søren Kierkegaard's
personality, it must not be forgotten that there was something other than the
distrust of his own authority and the piety toward Mynster that held him back from
appearing in the sharpest way against the usual conception of Christianity. There
worked in him also a heartfelt compassion with those who would come to suffer by his
appearance---the unhappy, in whose comfort there would come a breach; the happy,
whose joy he would disturb.

``Existence occasions one to become attentive to the many, many less gifted, weak,
simple people, women, children, the sick and the sorrowful, and so on, among whom one
lives. And then existence says to the religious person: can you, against these, find
it in your heart to screw up the religious, the price of blessedness, as high as you
do, you cruel one? And if the religious person is in truth the religious person, and
thus has love, then this objection makes a deep impression on him, on him who so
gladly wants to be with the suffering, whose only comfort and joy it really is to
comfort the suffering.'' This objection is indeed rejected in what follows, on the
ground that it speaks the language of human compassion---that Christ is the absolute,
and that the absolute is cruel: does not even permit one to bury one's father.---But
the objection has again and again risen in Kierkegaard's mind. Some time after, it
runs thus: ``But now these people, the great multitude of people, who must spend most
of their time earning their livelihood, in subordinate work: in the relation to them
it would indeed be a cruelty to screw up the price for them. Here, surely, humanly
speaking, a comforting and a mildness ought to be provided, precisely because it can
in such people be their concern, and this true in them, that it precisely pains them
not to be able to live for something higher\dots\ No, at the educated and well-to-do
class, the if not distinguished, then still the distinguished bourgeoisie: there it
is to be aimed, there the price is to be screwed up in the salon.''---Yet also in the
face of the happy the compassion rises: ``When I was to proclaim Christianity in
truth, I should thus disturb this whole happy existence, which can take place where
one does not come into contact with spirit\dots\ At times it is to me as though I went
about bearing a crime by this my knowledge of Christianity.'' And he felt that by the
happy natures the rigor must be felt even more strongly than by himself, who had
indeed all his life through been placed in inner suffering: ``I now live, once and for
all, in melancholy's separate cabin---but I dare rejoice to see others' joy\dots\ To be
healthy and strong, a complete human being, for whom there hopefully lies a long
life---now, that was never granted me. But when I then step out from my dreadful pain
among the joyful---I thought I dared have the melancholy joy of strengthening them in
thus rejoicing in life. Oh, but if it is to be proclaimed: to die away, that to be loved
by God is to suffer, and to love God to suffer---then I must, as it were, disturb all
the others' happiness\dots\ \emph{At this difficulty I have halted.}''

These utterances I have cited so fully because I do not think one understands Søren
Kierkegaard when one does not get as full an impression as possible of the whole side
of his nature that displays itself through them: his mild and loving mind. In his
writings it does not come forth, it seems to me, with such inwardness as in his
settling with himself in his solitude (in notes from the years 1849--1853).

\medskip

\noindent\textbf{5.} That the halt which compassion could effect became for the time
being only provisional came from the fact that Kierkegaard's whole conception of
Christianity and Church underwent a considerable sharpening after the personal
breakthrough and under the changed relation to the authorities that the events of the
time had brought about. He got a sharper eye for the weaknesses in the existing order,
at the same time as he felt more secure in his conception of the ideal.

His conviction was now that a great betrayal had been committed on the part of human
beings against Christianity. It was for them too high and too strict;---and they would
indeed feel akin to the high and take its help and comfort, but for as cheap a price as
possible. The first bargaining-down occurred when, in the Middle Ages, people began to
regard those who made earnest of following Christ in their lives as extraordinary
Christians. ``There the meaning burst in Christianity!'' Luther continued what had thus
been begun. He combated the Middle Ages too much, broke out of the monastery too early.
If people had earlier one-sidedly had an eye for Christ as exemplar, Luther now forgot
the exemplar for the reconciler. He bargained down the Christian---without making it
sufficiently recognizable that he bargained down. ``Luther, Luther, Luther, you have a
great responsibility!'' The Reformation became not a return to the original Christianity,
but a \emph{modification of the Christian}. Luther's appearance marks the human's reaction
against the Christian; he has invented that Christianity essentially exists in order to
calm and comfort. And then he was a giddy head, busy casting off burdens, attacking
authorities---such as now this unhappy political thing, that he wanted to overthrow the
pope!---And yet Luther can still be respected: he came from twenty years of penitential
exercises and anguish of soul. But how many Protestants come thus? And nonetheless
Protestantism levels everything; it is a plebeian tendency that does not acknowledge the
difference between the great and the humble, but---despite all opposition in
existence---feels itself on the same line as apostles and witnesses to truth!---It was
Kierkegaard's view that Catholicism will never be able to sink so deep as Protestantism,
since it in any case constantly has representatives of the strict, ideal conception of
Christianity to show. Protestantism lives off the Christian heroic age, the Church's first
300 years, without producing new heroes; but: ``the 300 years' laboriously acquired working
capital, it is used up, ladies and gentlemen; no more can be squeezed out of it by new
frauds.''

Yes, perhaps one must go still further back. The abuse began with the very beginning of the
Church. For it could not possibly have been quite right when the apostles on one day made
3000 into Christians. In their joy and enthusiasm something human must have befallen them;
they forgot what Christianity really was.---And even in the exemplar's life there can be found
traits where he gave the human too much. He went, after all, to the wedding feast at Cana, was
here in the joy of life and thereby strengthened people in it! But that he did only at the
beginning of his public career, before he had seen how evil the world was. Later he did not go
to weddings, and his apostle, who knew nothing but Jesus Christ and him crucified, he did not
go to weddings.

People have made Christianity too mild and coddling, into an expedient, a comfort for the worst
cases. One loses oneself in the ``coddling drivel'' about how Christianity satisfies the deepest
longings, and so on. God has become a good old grandfather, not a father who with rigor chastens
his children. In the place of the Christmas festival in the church Kierkegaard finds especially a
symbol for the changed character of Christendom. It is pure paganism that has there been taken up.
Christianity has become Christmas-fuss!

(A multitude of utterances in the diaries 1849--55 lie at the basis of this summary. See also
\emph{Judge for Yourselves!}, Second Series.)

\medskip

\noindent\textbf{6.} The mitigation and coddling that had corrupted Christianity, Kierkegaard
derived from woman---woman, whom Protestantism had especially brought forward by its praise of
marriage and family life. Earlier he had praised Luther's marriage as a heroic deed that in the
eyes of contemporaries posited the paradox: the monk's marriage with the nun! Now he thought that
Luther had here too been rather too quick. It is woman who makes it impossible for man to live for
spirit, for an unconditioned. Once man is bound with a woman who, through loving her own and hers,
loves herself, then the idea can whistle for him: mother-wit egoism holds him fast! ``Christianity
knows very well that with woman and love and so on, all this weaker and coddling comes up in a human
being, and that, in so far as the man does not hit upon it himself, the wife regularly represents it
with an unconstraint that is extremely dangerous for the man, especially for the one who is to serve
Christianity in a stricter sense'' (\emph{Practice in Christianity}). Woman represents the desire for
life; man is far more disposed to be spirit. Woman is bait---as Johannes the Seducer has said!
(Papers 1854).

How characteristic, that here from the strictest religious standpoint there is a pointing back to the
aesthetic: they meet in the same view from wholly opposite sides! The whole ethical view is pushed out
or treated as though it were not there.---One will recall, from the section on the ethical view of life,
that this was really not possible for woman, but that she was to pass immediately from the poetic to the
religious. Now (1855) the determination is added that the religious she gets access to only by an indirect
way; for the existence placed in contradictions, which the Christian is, she really cannot bear: she relates
herself to the religious through man, by being witness to how he lifts at the great task. Kierkegaard rejoices
to find support for his view of woman in Schopenhauer; he thinks he can also lean on the New Testament.

But from woman and the family the consequence goes to the natural instinct that leads to the formation of the
family. Reproduction of the race itself is of the evil. The New Testament presupposes that Christians do not
marry. And how, really, should they be able to, how should they be able to set children in the world, when they
know that they are born in sin and to misery? Christianity wants to bar reproduction of the race, does not want
history's personnel renewed. And if this is so, how thoroughly has not Protestantism, by its praise of family
life, corrupted Christianity! (1855).

Here, then, the consequence is drawn. Life is declared to be evil in its very urge and striving after subsistence
and renewal, the fundamental urge which directly and indirectly is the source of all human activity. It offers
resistance to the implanting of the supernatural element---so it must be rotten, and the branding-iron then stamps
its innermost marrow as corrupted. Already in the section on the religious view of life we saw self-annihilation set
as the highest. What has just been developed is really only a closer expression of the same thought.

\medskip

\noindent\textbf{7.} When now this whole view of Christianity and of the world was to be made to count, it is not
difficult to say how the judgment must come to sound over the existing Church. Already in the diaries Kierkegaard
has formulated the two chief propositions about which his later public polemic came to turn. Already here occurs the
thought: ``Luther had 95 theses; I have only one: Christianity does not exist,''---and the other: ``The whole worship
of God is a grandiose attempt in the direction of making a fool of God, although one is not conscious of this\dots\
Can one defend taking part in such a thing?''

We have heard what for a long time held him back. There came the situation in which nothing could hold him back from
loudly uttering the thoughts that had matured in him in solitude.

\section*{B.\quad The Last Word}
\label{sec:5B}
\markright{V. Søren Kierkegaard and Christianity --- B. The Last Word}

\noindent\textbf{1.} The occasion of the strife was a sermon that Martensen, a
few days after Bishop Mynster's death, preached in the palace church, and in
which he presented the deceased as a link of ``the holy chain of witnesses to
the truth that stretches through the ages from the days of the Apostles.'' This
utterance offended Søren Kierkegaard in the deepest way. He at once wrote down a
powerful protest, which he published three quarters of a year later (in
\emph{Fædrelandet} of 18 December 1854), when Martensen had safely and well
become Mynster's successor on the episcopal throne of Zealand. The strife thereby
awakened, and to which our literature knows no equal, when regard is had to the
greatness of the subject and to the passion with which it was conducted from the
side of the one who had raised it, I shall not attempt to depict in detail. It
deserved its own historian, who would have to be able, with dramatic life, to
set the acting personalities over against one another, and who could give an
account of the degree to which the great accusation Søren Kierkegaard has raised
against Christendom particularly strikes the Danish state church and its
representatives in his time. It is an affair of a similar kind to that between
Plato and the Sophists, and between Pascal and the Jesuits. Only historical
knowledge and historical art can settle it. Here I have to do with the strife
only from its universally human side. And quite apart from the strife's result,
Kierkegaard's polemical writings have an abiding interest, just as well as
Plato's \emph{Dialogues} and Pascal's \emph{Lettres à un provincial}. Scarcely
will anyone be able to immerse himself in the violent contributions from
Kierkegaard's side without their working on him as ``the single individual.'' It
would be highly mistaken if one merely read them with aesthetic interest in the
powerful polemic; weighty words are spoken about the relation between ideal and
self-deception, which have an errand to everyone, wherever he may otherwise
stand.---As for Kierkegaard himself, he showed himself as a violent agitator who
shunned no means to give emphasis to his words. He was concerned that no
immersion in inessential circumstances and no long-winded zeal to find excuses and
exceptions should divert attention from the great proposition he wanted to
establish. He thought at the same time that his opponents had indeed acknowledged
an absolute standard, and that their motives and their character, seen in its
light, must acquire a questionable earthly character. Nevertheless it is strange to
see this champion of ideality, this knight of thought, right at the beginning of
his polemic come forth with a turn such as this, that Martensen's sermon about
Mynster could also be called a memorial address in the sense that it brought
Martensen to remembrance for the vacant episcopal throne. And counterparts further
on in the polemic are not lacking. Where there is such superiority of spirit and
such an ability to use the weapons of satire and mockery as here, such insinuations
must in a double sense be said to be out of place. Kierkegaard's shield would have
been brighter without them.---

\medskip

\noindent\textbf{2.} The witness-to-truth struggle really turned on one of the chief
thoughts about which Kierkegaard had all his life circled, on the question:
continuity or break? Both-And or Either--Or? From the side of the existing Church it
was maintained, and it lay especially at the basis of Martensen's appraisal of
Mynster, that there was an unbroken coherence between the present-day ecclesiastical
representatives and the apostles, the representatives of New Testament Christianity.
A holy chain connected them. And there was maintained at the same time a conception of
Christianity according to which the Christian had partly entered, partly could enter,
union with the purely humane culture, so that a higher unity of both could be
reached---as people thought, on Christianity's own ground. Christian state, Christian
family life, Christian art and Christian science---these were a series of concepts whose
mere setting-up is characteristic of the ecclesiastical doctrine of harmony. For
Kierkegaard these concepts mark just as many sense-deceptions and self-deceptions, just
as many lies. He maintains a break in the coherence between the now-existing Christendom
and the New Testament's Christianity, because he maintains that the Christian in the New
Testament's sense is ``the deepest, most incurable break with this world'' (S.
Kierkegaard's newspaper articles). As little as Mynster's sermon expressed the Christian
in this meaning or brought about such a break with the world, so little was there a trace
of it in Mynster's personality and conduct of life. When one has nonetheless reckoned him
among the true witnesses to the truth in the Christian sense, this betrayed, in
Kierkegaard's eyes, an impermissible confusion of concepts, and, when one would not admit
it, an insolent corruption of the ideal. The ecclesiastically existing order could, he
thought, only be defended in the way he had attempted in \emph{Practice in Christianity},
by admission of the great distance between ideal and reality and by fleeing to the divine
grace when this admission had been made. But when from the ecclesiastical side one will not
make this admission, the ecclesiastically existing order becomes ``an insolent indecency, an
attempt in the direction of making a fool of God,'' and it becomes necessary to set up the
thesis, the only one: New Testament Christianity does not exist (newspaper articles.
\emph{Fædrelandet} of 28 March 1855).---The coherence with the New Testament is broken,
because the coherence with ``the world'' is not broken: that was Kierkegaard's standpoint in
the strife.

He would not let himself be put off with the notion that he maintained a stricter conception
of Christianity, Mynster and Martensen (as well as Grundtvig, whom he occasionally also takes
along) a milder. For here there is no question of rigor or mildness. It is a matter of simple
\emph{human honesty} in the answering of the question whether one now lives on the conditions
that New Testament Christianity presupposes, or not: ``For this honesty I will venture. On the
other hand, I do not say that it is for Christianity I venture. Suppose it, suppose I became
quite literally a sacrifice: I should nonetheless not be a sacrifice for Christianity, but
because I willed honesty\dots\ I dare not call myself a Christian; but honesty I will, and to
that end I will venture'' (newspaper articles).

\medskip

\noindent\textbf{3.} After the strife had for half a year (from December 1854 to May 1855) been
conducted as a newspaper debate, it assumed from Kierkegaard's side an even more passionate
character, in that he now, through small pamphlets, addressed larger circles with an open and
definite summons to break off connection with the Church. This summons was formulated in a
pamphlet that appeared in May 1855 (with the title: \emph{This Must Be Said; So Let It Then Be
Said}) thus:

``Whoever you are, whatever your life otherwise, my friend---by (if you otherwise take part in it)
refraining from taking part in the public worship of God as it now is (with the claim of being New
Testament Christianity), you have constantly one, and a great, sin the less; you do not take part in
making a fool of God by calling New Testament Christianity what is not New Testament Christianity.''
Thereafter followed, in the course of the summer, the nine issues of \emph{The Moment} (May--September
1855), with increasing violence in attack and agitation.---From my boyhood years I still remember the
little white pamphlets and the movement they called forth in the minds and talk of the grown-ups. Only
later was I myself to experience what a message of earnestness they brought.---Kierkegaard's language
here at times attains a force and at the same time a cutting pithiness that make some of the
``Moments'' unique in the history of our language.

The Church must go---there is nothing else for it; for it slays Christianity. It has abolished the
relation of opposition to the world, the world's goods and the world's tasks, and thereby it has
abolished Christianity. By a revolt undertaken not in open defiance, but quietly, gradually, in
hypocrisy, people have got something quite different out of Christianity than it really is. The
corruption began early: it began perhaps already with the apostles. Yes, has Christianity really come
into the world? By the help of the dogmas (Kierkegaard thinks especially of the dogma of the atonement)
one has pushed the exemplar aside; in the original Christianity's great funeral march one has found a
few mildly sounding rhythms and has from them taken occasion to compose the whole of Christianity over
into an idyll (\emph{The Moment} no.\ 5).

The matter is: ``To be a Christian in the sense of the New Testament is, in the upward direction, just
as different from being a human being as being a human being, in the downward direction, is different
from being an animal.'' But are people nowadays capable of such a spiritual existence, can they still
feel the proper religious passion? There has occurred a weakening in character, which is essentially due
to marriage and family life (\emph{The Moment} no.\ 7).

So much is certain: if the present condition in the Church is Christian, then the New Testament can no
longer be a guide for Christians, for the presupposition it rests on---the relation of opposition to the
world---has fallen away (\emph{The Moment} nos.\ 2 and 4).

The whole question has a far-reaching significance: it is not restricted to the moment: ``It is nothing
ephemeral I will, just as little as it is anything ephemeral I have willed; no, it was and is something
eternal: by the ideals against the sense-deceptions'' (\emph{The Moment} no.\ 1). For the ideals shall be
proclaimed, however it may go; otherwise mediocrity triumphs (no.\ 9). There is, after all, a way that has
at all times been for people an expedient, by which they have got troublesome problems off their necks---the
simple one: ``Be twaddle---and you shall see, all difficulties vanish'' (no.\ 9). (This last piece everyone
ought to read for himself who wants to know what Søren Kierkegaard willed, and how he knew how to express what
he willed.)

\medskip

\noindent\textbf{4.} A short time after the last number of \emph{The Moment} had appeared, Søren Kierkegaard
became ill in the street, fell down, and was brought to Frederik's Hospital. He had long been sickly; he
suffered from a spinal ailment that he derived from a fall in his childhood. ``I come here to die,'' he said
when he came to the hospital. According to the testimony of those who saw him during his last sickbed, there
was an extraordinary clarity and peace over him, and his eyes shone with even greater brilliance than before,
when the pains did not quite overwhelm him.

How strangely everything here joined together! When his last word was said---the one toward which he had, as
it were, drawn breath all his life through, and the last word that he could say---then his bodily strength too
was broken. And at the same time his outer means of existence too were exhausted; a short while before he had
drawn out the last money he had. The outer conclusion was like a symbol that he had exhausted the consequence
of his spirit and his thought.

And now he felt himself to be one whose strife was ended. Life had indeed been one long strife for him. He
fought honestly with himself before he fought with others, and he demanded of them only the same honesty that
he had first demanded of himself.---Now this inner and outer strife was at its close. As one will remember,
there was, according to his faith, an absolute opposition between the eternal and the human; only in passing
inconsistencies have we seen the shadows from the latter stretch over the former. And thus it was also now his
faith that the strife would for him now be relieved by peace, as it is said in some lines of Brorson, which he
wished to stand on his grave:

\begin{verse}
It is a little while,\\
and I have conquered,\\
then is the whole strife\\
at once vanished.
\end{verse}

\noindent From some interesting conversations he had in his last days with his youth-friend Pastor Boesen, one
sees that he did not receive the sacrament of the Lord's Supper, because he would not receive it from any
priest, but only from a layman. He died on 11 November 1855.

% ============================================================
%  CONCLUSION  (pp. 150--159)
% ============================================================
\bookchapter{Conclusion}
\label{ch:slutning}

\noindent\textbf{1.} ``The rest is silence!''---this last word of Hamlet's I was
most inclined to end with after having brought out Søren Kierkegaard's last word.
For to take the word thereafter can seem presumptuous. A great spirit's last word
about one of history's and life's very greatest questions may well make one silent.
But silence has worth only when it means that one is thinking, and I break the
silence in order to say what thoughts have---since I first came under the influence
of Søren Kierkegaard's ``only thesis''---fixed themselves in me.

Not because a definite position toward his last word should be necessary in order to
appraise his significance. His great merit lies first and foremost in his wonderful
ability to ask, to find difficulties, to wound in such a way that not merely the life
of thought, but the personal life on the whole, comes into inward and lasting movement
in those who immerse themselves in his writings. And in his proposition that
subjectivity is the truth lies this: that it is the personal inwardness, the creative
self-activity, which according to his and his teacher Poul Møller's conviction is
possible in every human being, on which everything depends. No handed-down,
externally-arriving knowledge is sufficient. Only that is truth for the human being
which he himself produces or reproduces. This proposition is so great and rich in its
application that it encloses consequences that---as previously intimated---go further than
Kierkegaard would acknowledge.

But for our own sake we must discuss whether he is right that the human race has shirked
Christianity, has out of softness and in hypocrisy turned its back on the ideal. If it
were so, it would have not merely religious significance. Whatever one may think about
Christianity's abiding significance---one thing is nonetheless clear: it would be sad if
the race had in \emph{such} a way forsaken the ideals of its past. Ideals one can forsake
in a worthy way only either by having outgrown them, or by the circumstances of life
having, for other reasons, changed and demanding new exemplars. Thus for purely humane
reasons we must seek, as well as we can, to make clear to ourselves the great question,
raised by Kierkegaard, of humanity's relation to the New Testament's view of life.

\medskip

\noindent\textbf{2.} Kierkegaard remarks in one place that in the New Testament there is
no guidance with respect to the problems that can arise from one's being brought up from
childhood in Christianity. And in ``the Moments'' he comes, by another way, back to the
thought that the New Testament in any case cannot be a guide for the present-day
Christendom. If one pursues the thought he has here touched on, one naturally comes to
investigate what the most important difference is between the presuppositions that
condition the New Testament view of life and those that prevail nowadays in people whose
minds are disposed to take up as much as possible of this view of life. When one asks thus,
there will be especially one circumstance that, the more one investigates the New Testament's
circle of representations, will show itself to be of decisive significance, namely this: that
the New Testament throughout proceeds from the assumption that history will soon---already
within the lifetime of the then-living generation---be concluded by Christ's return. If one
does not take that circumstance into consideration, one does not understand the New Testament's
ethics. It forms the constant background. The kingdom of heaven is near!---The end of all things
approaches!---that is the constant refrain, which for the attentive sounds along even where it is
not expressly repeated. Since one thus worked with a near horizon, had no outlook to unsurveyable
periods of development, to a long wandering for the race, all more comprehensive endeavors, all
proper cultural work, all social life had to become insignificant, just as everything had to be
avoided that could divert attention from the one, great and near goal. ``The end of all things
approaches; be therefore sober and watch,'' writes the apostle Peter (1 Pet. IV, 8), and it is with
the same motivation that Paul advises the Christians against marrying and against exchanging the
state of slavery for the free state (1 Cor. VII). What had that kind of goods and joys to say in the
face of the glory that was to be revealed? They could only divert mind and thought, which now, when
the great decision was imminent, it was precisely the concern to keep gathered.

When this presupposition falls away, Christian ethics must be placed quite otherwise. Its commands
acquire a wholly different significance when one works with a far horizon than when one worked with a
near horizon. The obscurities in the later Church's position toward the original Christianity find
their simple explanation in the fact that one has not sufficiently attended to this great difference.
When the long historical wandering began, tasks and purposes pressed forward to which one had before
attributed no value. On the long journey one has other needs than when one has only a couple of steps
to go. Relations asserted themselves with respect to which the New Testament could contain no real
guidance; there developed an independent human life in family and state, science and art---or rather,
the life that was already developed in these domains now had to unfold in the Christianized world. But
there came then a need for other exemplars and for an ethics that could attribute to this life a
positive significance, could regard life as a development and not merely as a time of exercise until the
speedy conclusion came. Early, then, one begins to connect Greek ethics with the Christian, most often
in a purely external way.

If Kierkegaard is right in his proposition that New Testament Christianity does not exist, then the
explanation lies in the fact that the presupposition of the New Testament ethics no longer in the same way
fills the mind. For the fact that one still speaks of the last day in all indeterminacy has in this respect
not much to say; as one indeed also knows, for the time being, how to arrange oneself quite comfortably.

It has interested me greatly to see that Kierkegaard in a single place touches on this point. It is said in
a note from 1849:

\begin{quote}
``The whole difficulty with Christ's return being predicted as near at hand, and yet not having occurred, is
explained by drawing attention to the fact that it is a subjectively true remark. And thus not only Christ
must say it, and next, which is indeed also a fact, the Apostles, but thus every true Christian must say it.
That is to say, it is to such a degree tormenting to be the true Christian that it would not be endurable if
one did not constantly expect Christ's return as imminent now. \emph{The torment, the suffering gives birth to
a necessary illusion}\dots\ One is no true Christian except in the pain and torment that it is to be a true
Christian in this world; and if one is in that pain and torment, then this illusion is a necessity.''
\end{quote}

This interpretation is completely in consequence of Kierkegaard's whole conception, and it is at the same time
the only one by which the presupposition of the New Testament ethics can be kept up. Who knows whether there are
not those who really do it? If there are such, then they have---what Kierkegaard has not emphasized---one
contradiction more to bear: the contradiction between the prophesied speedy conclusion and the long historical
time with all that it has brought. It is a sharpening of the paradox that Kierkegaard, had he become aware of it,
would have been able to develop with force and vigor.---But I think that he reverses the true relation. It is not
the tension that from the first has called forth the expectation; it is the promise that has called forth the
tension. \emph{The nearness of ``the Kingdom'' is a part of the original Christian proclamation and lies at the
basis of the evaluation of all life's relations and tasks in the New Testament.} In connection with this an
essential correction must be made to Kierkegaard's conception of the original Christianity. Christianity was not,
as he says, a funeral march (\emph{The Moment} no.\ 5). It was a march of victory. The joy and jubilation over the
imminent glory had far the preponderance over the feeling of suffering. The original Christianity was not so
consumptive as it often seems in Kierkegaard's presentation.\footnote{It is in this respect characteristic that
Kierkegaard will give no closer determination of what ``blessedness,'' the ``absolute purpose,'' really is. Every
description of the goal will in his eyes divert attention from the way---and ``the way is the decisive thing,
otherwise we have aesthetics.'' Of the eternal blessedness he will therefore say nothing other than that it is the
good that is reached by venturing everything. The New Testament has not this misgiving. The enthusiastic expectation
of the goal's glory and fullness is here precisely also the condition for one's coming forward on the way. See
especially the Revelation of John, which in its full depiction of ``the goal'' differs only in degree, not in kind,
from what in the other New Testament writings lies as presupposition.} It lived in great visions, in great and lively
expectation, and thereby got power and courage for sacrifices and exertions that later on, when they were to be
imitated under wholly other presuppositions, acquired the character of spiritualistic asceticism. The original
Christianity is by no means spiritualistic. The time of spiritualism first came when the expectation had ceased, or
at least been calmed.

Those who can go along with the conception I have here maintained will be able to agree with Kierkegaard in his
proposition that New Testament Christianity does not exist; but they will not explain it in the same way as he. They
will see that the changed position toward Christianity is simply due to the fact that a long historical development has
taken place. When one now does not assume that the human race's ideals can be prescribed to it once and for all---which
is especially impossible when such decisive presuppositions as the one just mentioned can change completely---then one
comes to the conviction, not that Christianity's time is past, for that it will in a certain sense never become, but
that Christianity's significance, like the significance of every great spiritual movement, is this: under certain
definite historical conditions to have awakened feelings and ideas which, though they cannot keep themselves unchanged
when the conditions become other, can nonetheless enter as essential elements into the ideal world that humanity can
never do without. Christianity has made the ideals of love, of inwardness, and of purity count with a force of feeling
such as has not happened before or since. It has, if not from the first awakened, then at least sharpened, the feeling
for the suffering and the weak, and maintained the invalidity of the outer differences in the face of the common human
(although here the confessional difference early hampered). These elements the race takes with it on its way forward. A
noble type of life is not lost, even if the conditions for its complete realization fall away. It is continued as an
element in the continued life, which constantly seeks new forms.

The humane view of life stands over against Christianity as it stands over against Greek culture. The Greek manliness,
clarity of thought, and harmony we can just as little do without as the love and inwardness that were the greatest and
will be the abiding thing in Christianity. Out of the human spirit these ideals and qualities have developed under
definite historical conditions; and the same human spirit takes them with it on its continued wandering, in so far as
they can be freed from their original presuppositions and worked together with the new qualities and exemplars that the
new conditions of life make necessary. We take the good where we can find it.

\medskip

\noindent\textbf{3.} This continued development of life, and its right and ability under the new conditions to form new
ideals, Kierkegaard denied in the most definite way. A whole arsenal of scornful words he had ready against such a way
of regarding things, which for him became one with directing one's conviction about the highest ``according as the
time, the public, the profit demands.'' No one is able, as he is, to caricature what he rejects or---does not understand.
And yet there is an evident self-contradiction in his conception, which is closely connected with his at one and the
same time pleading the cause of subjectivity, of personality, and holding fast to a foundation of fixed dogmas as an
absolute condition.

In his theory of knowledge Kierkegaard maintains that existence cannot be exhausted or comprised by knowledge, but that
there is constantly new work to do---that in the world of thought things must constantly be ventured anew. And in his
conception of religious faith it is a chief thing that the decisive element is the relation of contemporaneity to the
struggling and suffering ideal. He designated it in his last days as the thought of his life: ``what you do in
contemporaneity is the decisive thing.'' And earlier he had inculcated that world-historical figures are not suited to be
ideals, because they have a concludedness that makes the contemplation secure; ``ideality cannot be got historically
corked-up.''

If earnest is made of this, every dogmatic religion becomes impossible. For in the dogma lies precisely the conclusion:
the dogma is a declaration once and for all about the significance of a historical person or event, an establishing of a
won result. All positive or dogmatic religion in reality gives ideality historically corked-up---and therefore there was
needed a Søren Kierkegaard's enormous exertion of thought and great art of presentation in order again to bring out the
great significance of the relation of contemporaneity. All instruction in dogmatic religion is an instruction in
answer-keys; the sum has been worked out, and it is in high degree difficult to carry out the re-checking in such a way
that one does not make use of one's knowledge of the result. Bad novel-readers peep at the back of the book about the
outcome---but thereby miss the tension that alone gives the mood during the reading a likeness to the mood at an
experience.---Kierkegaard saw at last the misfortune in the dogmas; it was by their help that one had pushed the exemplar
aside.

Contemporaneous in earnest one can be only with that which is not concluded, and whose result does not lie guaranteed. And
that holds only of the real human life itself, with its purposes and tasks, its struggle, its desire and its pain.
Kierkegaard's doctrine of the paradox contains the great truth, which also---or: precisely---the humane view of life fully
has use for, that the great and significant so often develops under humble and unpretentious surroundings. Its germs are
often difficult to discover and easy to misunderstand. The concern is then for an open eye and a firm faith in the value of
what will perhaps slowly and late be able to work its way through to recognition. The judgment over us rests especially on
our relation to the ideal powers that have lain in the bud or struggled to be able to unfold in our time. A dogmatic guideline
cannot be given; the ideal cannot be formed once and for all; it must form itself anew, assume new shapes, and then makes its
demands on us under each of these shapes.

Thus, approximately, must Kierkegaard's doctrine of the paradox and of the relation of contemporaneity be adjusted, if it is
to have abiding and general significance. As sharply as he, as thinker and ethicist, stepped forth against the harmful effects
of the dogmas, he nonetheless did not break with the dogmas; his noble wine he put into the old leathern bottles. If the wine
is to be preserved, it must be poured into new vessels.

\medskip

\noindent\textbf{4.} It would be idle to brood over how Søren Kierkegaard would have placed himself if his life had been
continued. His sympathy with Catholicism on the one hand, his acknowledgment of free researchers' (Feuerbach's) conception of
Christianity on the other, open opposite possibilities. And a new possibility is added when we think of a development like Leo
Tolstoy's, which offers several similarities with Kierkegaard's. It could be imagined that he had continued to work as the
single individual, in imitation of ``the Exemplar,'' but with increasing contempt for ``the Dogmas.''---But who can exhaust the
possibilities for a great spirit's future course, especially one for whom ``the leap'' is a favorite thought?---And just as little
can the possibilities for his influence be easily calculated. Mainly he has worked as a divider of streams, has made turbid
mixtures separate, has brought clarity into the spiritual world. Some he has led to seek more inward immersion in Christianity
(as they conceive it). Others he has, in virtue of the principle of personal truth, led beyond it. And the indirect effects and
consequences of his thoughts branch out into all the most important movements of spiritual life here in the North---and beyond
the North.

\end{document}
