% ============================================================
%  H. Høffding, "Forholdet mellem Tro og Viden i dets
%  historiske Udvikling. Et indledende Foredrag."
%  Først trykt 1885; optrykt i: Mindre Arbejder,
%  Kjøbenhavn: Gyldendal, 1899.
%  TRANSLATION (English) --- Hans Halvorson
%  [Sections I--II translated; remainder forthcoming]
% ============================================================
\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{setspace}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\usepackage{hyperref}

\hypersetup{
  pdftitle={The Relation between Faith and Knowledge in its Historical Development},
  pdfauthor={H. Høffding},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\rhead{Høffding, \emph{Faith and Knowledge in its Historical Development} (1885)}
\cfoot{\thepage}

\setstretch{1.3}

\title{\textbf{The Relation between Faith and Knowledge\\[4pt]
       in its Historical Development}\\[6pt]
  \large An Introductory Lecture}
\author{Harald Høffding\\[4pt]
  \small Translated from the Danish by Hans Halvorson}
\date{Printed in: \emph{Mindre Arbejder}, Copenhagen: Gyldendal, 1899\\
  (Lecture delivered 1885)}

\begin{document}
\maketitle
\thispagestyle{fancy}

\bigskip

\noindent The problem of faith and knowledge arises of necessity at a
certain stage of cultural development. We encounter it in the Indian,
the Greek, and the Muhammadan worlds as soon as intellectual and
spiritual development has reached a certain point. Here, however, we
shall confine ourselves to tracing the main forms in which it has
appeared within the European spiritual development shaped by
Christianity.

\section*{I.}

No instruction on this question can be drawn from the source writings
of Christianity. On the standpoint from which they were composed, no
knowledge distinct from faith exists. The gaze is fixed on one single
object, one single goal, which lays claim to all striving and all
spiritual force. The endeavor to know the truth is here immediately one
with the urge to win the salvation of the soul, and does not go its own
way by its own method. There is no distinction between theory and
practice, since theory and theoretical interest simply do not exist.
When the Deity itself has revealed itself visibly to all, --- when faith
makes it possible to attain understanding of the Mystery that contains
all the treasures of wisdom and knowledge, --- every motive for seeking
knowledge by other paths is lacking.

It was natural that faith should from the outset have to be all in all,
leaving no room for other spiritual needs. The first generation still
stood so close to the mighty spiritual movement that led to the founding
of the Christian community that it lived and breathed entirely in the
great memories and visions. Added to this was the fact that this first
generation lived in the belief in the imminent return of Christ. Then
everything was to be concluded, the last judgment rendered, and all human
and earthly arrangements give way to a higher order of things. Who, with
such a catastrophe in view, had the time and composure of mind for
philosophical thinking or scientific inquiry? Science could no more than
any other human endeavor acquire any significance. There was to be no
historical development at all; the goal had been reached, and the end
stood at the door. One does not understand the New Testament and its
relation to cultural endeavor in its various forms unless one holds fast
to this. What happened here with scientific striving was the same as with
the striving for personal freedom and with the entering into marriage
(see 1~Corinthians, ch.~7). How could all such things acquire
significance --- or even become the object of a drive --- when all
thoughts and energies had to be strained to be prepared for the great,
the absolute decision!

% -------------------------------------------------------
% Section II onwards --- translation forthcoming
% -------------------------------------------------------

\bigskip
\noindent\textit{[Sections II--VI: translation forthcoming.]}

\end{document}
