% ============================================================
%  H. Høffding, Religionsfilosofi
%  Copenhagen and Christiania: Gyldendalske Boghandel
%  Nordisk Forlag, 1901. Second edition 1906.
%  TRANSLATION (English)
%  Translated from the Danish transcription (transcription.tex).
%  Cf. B. E. Meyer's 1906 translation (The Philosophy of
%  Religion, London: Macmillan), which is based on the German
%  edition and differs in places from the Danish original.
% ============================================================
\documentclass[12pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{setspace}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\setlength{\headheight}{14.5pt}
\addtolength{\topmargin}{-2.5pt}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Philosophy of Religion},
  pdfauthor={Harald Høffding},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\textit{Philosophy of Religion}}
\fancyhead[LO]{\textit{\leftmark}}

\setstretch{1.3}

\title{\textbf{Philosophy of Religion}\\[6pt]
  \large\textit{Religionsfilosofi}}
\author{Harald Høffding\\[4pt]
  \small Translated from the Danish}
\date{Copenhagen and Christiania:\\
  Gyldendalske Boghandel, Nordisk Forlag\\
  1901. Second edition 1906.}

\begin{document}
\maketitle
\thispagestyle{empty}

\vspace*{\fill}
\begin{center}
\textit{Alte dubitat, qui altius credit.}\\[6pt]
(He doubts more deeply who believes more deeply.)
\end{center}
\vspace*{\fill}
\newpage

\tableofcontents
\newpage

%% ============================================================
%%  I. PROBLEM AND PROCEDURE
%% ============================================================
\chapter{Problem and Procedure}
\label{ch:problem}

\begin{flushright}
\textit{ou lyseis agnoōn ton desmon.}\\
\textsc{Aristotle}.\\[4pt]
\textit{(One cannot untie a knot without knowing where it is bound).}
\end{flushright}

\bigskip

\noindent\textbf{1.} By \emph{philosophy of religion} one could mean
either a thinking that springs from religion and builds on religion, or a
thinking that takes religion as its object. It is in this latter sense
that the term shall be taken here. The thinking that springs from and
builds on religion itself belongs to the subject matter of the philosophy
of religion and should not be called philosophy of religion.

By religion we think first of all of a state of mind in which feeling
and need, fear and hope, enthusiasm and devotion play a larger role than
reflection and inquiry, and in which intuition and image hold more sway
than investigation and reasoning. To be sure, again and again in the
religious life there will arise an involuntary need to give an account
of one's condition and of its content---of the worth of one's motives
and the validity of one's thoughts. But the religious thinking that thus
arises does not pose the religious problem in a decisive way. Its
problems arise within religion, but religion itself does not become a
problem. No other starting point is yet recognized than religion
itself; in any case other starting points are so subordinate in relation
to religion that they do not acquire decisive significance. Of this kind
is religious thinking in those times in which religion is everything on
the spiritual plane. These are the classical times of religion: partly
the great times of origins, in which religion with all the power of
originality draws all strength and interest to itself; partly the great
times of organization, in which all existing culture is shaped or bent
into obedience under the highest religious ideas. In such classical
times there is a great unity, or at least a great harmony, in the domain
of spiritual life. Christianity had such a classical time in the age of
the primitive Christian community, and later in the great organizational
period at the height of the Middle Ages. A religious problem forms only
when other sides of spiritual life---science and art, morality and
social life---begin to free themselves and to appear with the claim of
having independent value. They then judge religion from their points of
view and by their standards, whereas religion in its classical times
either simply ignored them---as in the times of origins---or---as in the
times of organization---valued them from its own point of view and by
its own standard, so that the question became whether these two
different evaluations could be brought into mutual harmony.---There must
be definite historical conditions present before the religious problem
can be posed in a decisive way. It is, if one will, only in unhappy
times that a religious problem exists. For this problem is the
expression of a spiritual disharmony: the different elements of
spiritual life no longer work together in so unified a way as before;
they lead in different directions, among which one must perhaps
ultimately choose. But then both the necessity and the possibility of an
investigation are also present.

That religion is made into a problem will be offensive to many. But
thought must have the right to investigate everything, once it has
awakened, and only thought itself can mark out its boundaries. Who else
should mark out its boundaries? He who perceives no problem has of
course no reason to think; but neither has he any reason to prevent
others from thinking. He who fears losing his spiritual support must
keep away. No one wishes to rob the poor man of his only lamb---but then
the poor man must not unnecessarily come dragging it along the highway
and demand that traffic be stopped on his account. Experience shows,
moreover, that it is rather the oxen than the lambs that loudly announce
themselves with their indignation in season and out of season
(especially out of season). It is not so much the genuinely poor in
spirit as the stiff and blustering dogmatists who take offense when free
inquiry claims the right to stir on the religious plane as on any other.

The investigation I wish to undertake here addresses itself neither to
those who are finished nor to those who are fearful. The finished are
found in every camp, not least among the so-called ``freethinkers''---a
class of people who, like the ``Worms'' in Linnaeus's system, can be
characterized only by negative predicates, because the class is supposed
to encompass so many different forms. The finished are in possession of
a definite negative or positive solution to the religious problem and
have thereby lost the sense for further thinking about it. The fearful
dare not think about it.---Rather, the investigation addresses itself to
those who are \emph{becoming}. ``One who is becoming will always be
grateful''---no matter in which direction his becoming goes.

Philosophy is, as is well known, richer in ideas, perspectives, and
discussions than in definitive results, and the philosopher would often
be in a sorry state if he did not believe that the constant striving
after truth belongs to the highest spiritual values, which it is the
task of all true religion to uphold. Even if we could learn nothing else
from a philosophy-of-religion investigation, we could perhaps learn
something about the causes of the strife that rages over religious
questions, and about this strife's significance for the development of
spiritual life. And should the religious problem prove insoluble, we
might perhaps learn why it is insoluble.

\medskip

\noindent\textbf{2.} The philosophy of religion should not proceed from
a finished philosophical system. In a certain sense all thinking must be
systematic, for ``system'' literally means ``that which stands
together,'' and it is an inexorable demand placed on our thoughts that
they must be able to ``stand together'' with one another. There is an
inner coherence and consistency which in every domain is the ultimate
mark of truth, and which must therefore also be required in the domain
of the philosophy of religion. But the philosophy-of-religion enterprise
will be most fruitful where religion and its phenomena are not placed in
relation to an already completed philosophical conception, but are
illuminated by a philosophizing that is itself in the process of
investigating whether a completion is possible at all. The task is to
discuss religion's relation to spiritual life. Religion is itself a kind
of or form of spiritual life, and it is judged rightly only when seen in
its relation to other kinds or forms. It is no alien standard that is
thereby applied to religion. For there is a service from which no
spiritual power---even if it appears under the greatest names---can
exempt itself, namely to serve the deep and rich development of
spiritual life. The philosophy-of-religion task will therefore above all
be to find the means for investigating whether religion, under the
conditions of our spiritual culture, can continue to render such
service. In solving, or better: in treating this task, the philosophy of
religion cannot come equipped with other presuppositions than those
which every other science brings to the treatment of its subject. Very
often it will surely be compelled to declare that there are questions
that cannot be answered, because the conditions in the spiritual domain
are more complex and more diverse than positive and negative dogmatists
suppose. Yet it is not without significance that the problems are
discussed. The conscious insight into the impossibility of a solution
(\textit{docta ignorantia}) naturally invites one to investigate whether
it is then so absolutely necessary that a solution be found. In
addition, one's eye will be sharpened for the particular conditions
under which our spiritual culture lives, conditions which would probably
not continue to persist, just as they even now do not exert equal
influence on all personalities.

A philosophy-of-religion investigation can suitably be arranged so as to
consist of an epistemological, a psychological, and an ethical part.

In its classical times, religion satisfies all of humankind's spiritual
needs, including its drive for knowledge. Religious ideas then contain
for human beings the explanation of existence in its entirety and in its
individual parts. But when an independent science has developed, another
kind of understanding or explanation appears alongside the religious
one, and the question becomes whether these two kinds are compatible.
Possibly the situation might be that the new, scientific mode of
explanation increasingly replaces the religious explanation where
particulars are concerned, but that it cannot be applied to existence
considered as a whole---to what one customarily calls the ``first'' or
the ``last'' questions. Then one must discuss whether religion is able
to solve the riddles that science cannot solve. If it should then turn
out that religious conceptions, whether with respect to the whole or to
the particular, cannot help with anything that could rightly be called
understanding or explanation, then one must raise the question what
significance these conceptions can still have.

This last question leads from the epistemological philosophy of religion
to the psychological. For if religious conceptions no longer have
cognitive value, the value they might possess must consist in being
expressions of sides of the life of the soul other than the
intellectual. An attempt must then be made to describe the religious life
of the soul, especially the relation of conceptions to religious
experience and religious feeling. Through this it will become apparent
that religion in its innermost nature has to do not with the
understanding of existence but with the \emph{evaluation} of it, and
that religious conceptions express the relation in which factual
existence, as we know it, stands to that which for us gives life its
highest value. The core of religion is---according to the hypothesis to
which we are led---a faith in the persistence of value in existence.
This faith appears in great, vivid images in the popular religions,
especially in the highest among them. Even in those who stand outside
every popular religion, such a faith may stir, even though it does not
here take definite shape.

The transition from psychological to ethical philosophy of religion is
naturally motivated by the question: what ethical value does it have
that one believes in the persistence of value? Ethics concerns the
production of values, and the possibility cannot be dismissed that the
work of producing values leaves no time and no strength for living in
the thought of their persistence. The ethical philosophy of religion
forms a parallel to the epistemological. In its classical times,
religion directly and indirectly determines the evaluation of all
actions and all conditions of life. The religious problem arises on this
point when an independent ethics has developed that seeks, by its own
path, the grounding and motivation for the evaluation of human actions.
And on this point the problem is posed in the sharpest and most pressing
way, for the ultimate standard for judging religion must belong to the
ethical domain, where the final reckoning of the spiritual account takes
place.---

Within the framework here indicated, we should be able to find room for
all essential aspects of the religious problem. I wish to treat this
problem not only with the intellectual interest that is aroused by a
great and comprehensive subject, but also in the mood that comes from
the consciousness of standing before a spiritual form of life in which
the human race through millennia has deposited some of its deepest and
most inward experiences.

\medskip

\noindent\textbf{3.} Before we enter upon the individual parts of the
philosophy of religion, it will be useful to give a somewhat more
detailed characterization of the content of the religious problem.

In the Protestant world it is fairly generally acknowledged that one must
distinguish between knowledge and faith. The Catholic Church does not
acknowledge this distinction in its full extent and therefore does not
acknowledge any religious problem in the intellectual domain. It
maintains that God's existence can be proved by the paths that Thomas
Aquinas, supported by Aristotle, pointed out, and it has even declared
it a matter of faith to accept that these proofs suffice. The tradition
from the great time of the Middle Ages, when ``one did not doubt and was
therefore nearer to the truth'' (to use a sentence from one of Thomas
Aquinas's modern biographers), is continually upheld, and it is not
acknowledged that a disharmony has really entered into spiritual life.
The distinction between knowledge and faith concedes that such a
disharmony is present---that the unity of spiritual life is broken. Is
this distinction itself then only a station on the way, in that
development will lead to an elimination of all religion from spiritual
life, because only that can endure which can enter into a harmonious
relation with the other elements of life?---Whether one now assumes that
it is only the classical time of religion that is past, or that its time
is altogether past, the question cannot be set aside whether spiritual
value is lost in the process of dissolution that has set in, or whether
a replacement can be found for the spiritual unity and self-containment
that was lost. This consideration alone shows that the question of the
persistence of value is an essential aspect of the religious problem,
whatever solution one inclines toward. It is not least from this side
that philosophy in the last century and a half (from the time of
Rousseau, Lessing, and Kant onward) has treated the religious problem.
Both Romantics and Positivists have been attentive to this aspect of
it.---It must of course be remembered that even if one could show that
no spiritual energy was lost in the dissolution of religion, it does not
follow that the energy would under the new conditions receive as
valuable an application as in the classical times of religion. In the
spiritual domain no less than in the material, the conservation of
energy is not without further ado the same as the persistence of value.
Energy can persist without value persisting, while the reverse cannot
well be imagined, since energy in and of itself will be a condition for
the persistence and development of what is valuable. Many freethinkers
proceed without further ado from the assumption that human life will
become more powerful and richer when religion falls away---an assumption
that is not self-evident and that rests on the presupposition that
psychic equivalents are present---equivalents both in terms of value and
in terms of energy. The task will then be to demonstrate these
equivalents and thereby establish the persistence of value. But it is a
great question and an essential aspect of the religious problem whether
such equivalents can be demonstrated.---Moreover, the problem of the
persistence of value will also arise within the Church, for religion and
the religious life undergo constant changes; and the question becomes
whether the values from the classical times of religion persist through
these changes---e.g., whether primitive Christianity, both in terms of
value and in terms of energy, is preserved in the Christianity of our
day. In its sharpest form this problem has appeared in the assertion:
``the Christianity of the New Testament no longer exists.''---

But the religious problem concerns not only the persistence of value
within the human world. It raises not only a psychological but also a
cosmological question, namely about the relation between that which
stands as the highest value for human beings and the whole of existence.
Is there any connection between values and the laws and forces of
existence? Are these laws and forces perhaps even ultimately determined
by the highest values? Or can we perhaps not ascribe any validity to the
concept of value beyond the domain of human life?

The hypothesis that will later (in the psychological philosophy of
religion) be sought to be grounded holds that the proposition of the
persistence of value is the actual religious axiom, which is expressed
in different ways in the different religions and at the different
religious standpoints. The religious problem thus concerns the question
whether this axiom can really be taken to hold. And at the same time
this axiom, insofar as it expresses the fundamental thought of all
religion, can be used as a standard for the consistency and significance
of the individual religion or the individual religious standpoint.
Finally---as already noted---the relation between ethics and religion
can, with the help of this axiom, be expressed in a very simple way: how
does faith in the persistence of value relate to the work of finding,
producing, and preserving values?---

If the proposition of the persistence of value were to be discussed in
its full scope, the discussion would become interminable, and clear
starting points for conducting it would soon be lacking. I therefore
emphasize at once the limits within which my investigation will move,
and which must be kept in mind throughout all that follows.---When I
attempt to show that the said proposition is the fundamental thought of
all religion, this attempt will not have failed because it turns out
that no religion expresses the proposition with clarity and consistency.
It will be sufficient if an explicit need or tendency to hold fast to
this fundamental thought can be demonstrated, so that the standard for
the evaluation of religions in their mutual relations becomes the degree
to which they are able to articulate and carry through the
proposition.---When the persistence of value is spoken of, this could be
understood to mean that value does not vanish, but that there is
constantly value in existence, whether it decreases or increases or
alternates between decrease and increase. But I understand the
expression \emph{the persistence of value} in analogy with the
expression \emph{the conservation of energy}, such that the proposition
asserts a constant persistence of value through all changes of form. It
will perhaps be necessary here to distinguish between potential and
actual value, just as one distinguishes between potential and actual
energy, and the one distinction is just as clear and just as justified
as the other.---But the persistence of value through all changes of form
and despite the difference between possibility and actuality is
nonetheless only a minimum. It may be maintained that the mere
persistence of value is not satisfying and actually contains a
contradiction, since value can persist only by growing, since mere
repetition dulls, while on the other hand change is valuable only when
it brings enhancement. That an increase of value does not conflict with
the conservation of energy needs no further demonstration; the increase
of value presupposes only a new and more perfect application of energy,
not an increase of energy as a whole. It is for the sake of simplicity
that I limit the investigation to the persistence of value. And it will
also turn out that religious consciousness has at bottom limited itself
to this as well. The persistence of value will present us with
difficulties and tasks enough, even if we do not enter into everything
that lies in the unavoidable consequence that value can persist only if
it increases. Yet we must not forget that this consequence is
present.---Finally, when the persistence of value is under discussion,
one might ask: which value and how much value? I enter into these
questions only insofar as the difference between the different religions
and religious standpoints essentially depends on the different answers
they give to them. It is clear that the proposition of the persistence
of value can be accepted whatever answers one gives. The hypothesis that
the persistence of value is the fundamental religious proposition is not
shaken by the fact that the different religions and religious
standpoints diverge greatly from one another with respect to the
character and extent of the valuable in whose persistence they believe.

Value is the property of something that it brings about a real
satisfaction or can become the means to such. Value can therefore be
immediate or mediate. Immediate value we seek to hold fast when it is
given, and to produce when it is not given. We thus make it our end.
That we make something an end presupposes, conversely, that we have an
experience or a supposition of its immediate value. Mediate value
belongs to that which serves the attainment of an end---that is, of an
immediate value. Mediate value need not be potential value. For what
serves as a means for attaining something of immediate value itself
acquires value in our eyes, even though its value stands as derived.
There are all possible transitions between mediate and immediate value,
and through the shifting of motives the former passes over into the
latter, so that what at first had value only as a means now acquires
value as an end. Potential value, on the other hand, often means only
the possibility that something will acquire value as a means---that is,
the possibility of mediate value.---Which values are experienced and
therefore acknowledged at the different standpoints will depend on the
different character of those standpoints. It is a being's nature that
determines its need, and it is its need that determines what acquires
value for it. The religious axiom thus shows the necessity of a
religion's character being determined by the nature and need of the
human beings who profess it. For one cannot seriously believe in the
persistence of values that one does not know at least approximately from
one's own experience.---

It is to Kant's philosophy that we owe the independence of the problem
of value from the problem of knowledge. He taught us to distinguish
between evaluation and explanation. That this distinction cannot,
however, be a complete opposition is clear from the fact that it must
depend on the laws and forces of existence whether something can acquire
value for us and whether we can make the valuable into an end that can
be realized. The religious problem can therefore not be separated from
the other philosophical problems, even though it is most advantageous to
treat it by itself. It presents a clear analogy with the other problems.
The fundamental philosophical problem concerns the continuity of
existence in its relation to the special and individual forms of
existence. And if the hypothesis that the proposition of the persistence
of value is the religious axiom turns out to be well-founded, then the
religious problem concerns the continuity of existence from a
distinctive point of view and thus falls under the fundamental problem
as a special form of it.


%% ============================================================
%%  II. EPISTEMOLOGICAL PHILOSOPHY OF RELIGION
%% ============================================================
\chapter{Epistemological Philosophy of Religion}
\label{ch:epistemology}

\section{Understanding}
\label{sec:understanding}

\subsection{Causal Explanation}
\label{ssec:causal}

% [text to be added]

\subsection{The World of Space}
\label{ssec:space}

% [text to be added]

\subsection{The Course of Time}
\label{ssec:time}

% [text to be added]

\section{Concluding Thoughts}
\label{sec:concluding}

% [text to be added]

\section{Thought and Image}
\label{sec:thought-image}

% [text to be added]

%% ============================================================
%%  III. PSYCHOLOGICAL PHILOSOPHY OF RELIGION
%% ============================================================
\chapter{Psychological Philosophy of Religion}
\label{ch:psychology}

\section{Religious Experience and Religious Faith}
\label{sec:experience-faith}

\subsection{Religious Experience}
\label{ssec:experience}

% [text to be added]

\subsection{Religious Faith}
\label{ssec:faith}

% [text to be added]

\section{The Development of Religious Ideas}
\label{sec:development}

\subsection{Religion as Drive}
\label{ssec:drive}

% [text to be added]

\subsection{Polytheism and Monotheism}
\label{ssec:polytheism}

% [text to be added]

\subsection{Religious Experience and Tradition}
\label{ssec:tradition}

% [text to be added]

\subsection{The Scientific Conclusion of the Psychology of Religion}
\label{ssec:conclusion-psych}

% [text to be added]

\section{Dogmas and Symbols}
\label{sec:dogmas}

% [text to be added]

\section{The Proposition of the Persistence of Value}
\label{sec:persistence}

\subsection{More Precise Determination of the Proposition of the
  Persistence of Value and of its Relation to Experience}
\label{ssec:value-determination}

% [text to be added]

\subsection{Psychological-Historical Discussion of the Proposition
  of the Persistence of Value}
\label{ssec:value-psychological}

% [text to be added]

\subsection{General Philosophical Discussion of the Proposition
  of the Persistence of Value}
\label{ssec:value-philosophical}

% [text to be added]

\section{The Principle of Personality}
\label{sec:personality}

\subsection{The Significance and Justification of the
  Principle of Personality}
\label{ssec:pers-significance}

% [text to be added]

\subsection{Main Groups of Personal Differences}
\label{ssec:pers-groups}

% [text to be added]

\subsection{Buddha and Jesus}
\label{ssec:buddha-jesus}

% [text to be added]

\subsection{Is the Principle of Personality a Principle of
  Growth or of Dissolution?}
\label{ssec:pers-growth}

% [text to be added]

\subsection{Layman and Scholar}
\label{ssec:layman-scholar}

% [text to be added]

%% ============================================================
%%  IV. ETHICAL PHILOSOPHY OF RELIGION
%% ============================================================
\chapter{Ethical Philosophy of Religion}
\label{ch:ethics}

\section{Religion as the Foundation of Ethics}
\label{sec:foundation-ethics}

% [text to be added]

\section{Religion as a Form of Spiritual Culture}
\label{sec:spiritual-culture}

\subsection{Psychological Considerations}
\label{ssec:psych-considerations}

% [text to be added]

\subsection{Sociological Considerations}
\label{ssec:sociol-considerations}

% [text to be added]

\section{Primitive Christianity and Modern Christianity}
\label{sec:primitive-christianity}

% [text to be added]

\section{We Live on Realities}
\label{sec:realities}

% [text to be added]

%% ============================================================
%%  NOTES
%% ============================================================
\chapter*{Notes}
\addcontentsline{toc}{chapter}{Notes}
\label{ch:notes}

% [notes to be added]

\end{document}
