% ============================================================
%  F. C. Sibbern, On Cognition and Inquiry.
%    Toward an Introduction to Academic Study.
%  Copenhagen: At Fr. Brummer's Press (Printed by C. Graebe), 1822.
%  TRANSLATION (English) — SCAFFOLD.
%  Source of truth for translation: transcription.tex (Danish, COMPLETE + proofed).
%  Method: TRANSLATION-PLAYBOOK.md. Fill each "% [text to be added: ...]"
%    marker batch by batch (~10 printed pp.), mirroring transcription 1:1.
%  Register: lightly modernized (faithful meaning, syntax untangled).
%  Key terms: on first use in a section, gloss the Danish in brackets, e.g.
%    cognition [Erkjendelse]. Maintain GLOSSARY.md for consistency.
%  Page-offset (Danish scan, for reference): body printed = PDF - 22.
%  See TRANSLATION-NOTES.md for state + the section->page map.
% ============================================================
\documentclass[12pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{libertinus}
\usepackage{libertinust1math}
\usepackage{textalpha}   % real Greek glyphs, copied verbatim from the Danish
\usepackage[english]{babel}
\usepackage{enumitem}    % for \arabic*)-style lists (playbook: easy to forget)
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{setspace}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\setlength{\headheight}{28pt}
\addtolength{\topmargin}{-13.5pt}
\usepackage{hyperref}

\hypersetup{
  pdftitle={On Cognition and Inquiry},
  pdfauthor={F. C. Sibbern},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[CE]{\textit{\leftmark}}
\fancyhead[CO]{\textit{\rightmark}}

\setstretch{1.3}

% TITLE — rendering to confirm with Hans (see TRANSLATION-NOTES): "Granskning"
% could also be Investigation / Scrutiny / Examination.
\title{\textbf{On}\\[6pt]
  \Huge\textbf{Cognition}\\[6pt]
  \large and\\[6pt]
  \Large\textbf{Inquiry.}\\[12pt]
  \normalsize Toward an Introduction to Academic Study.\\[16pt]
  \normalsize By\\[6pt]
  \large Dr.\ Frederik Christian Sibbern,\\[4pt]
  \normalsize Professor of Philosophy.}
\author{}
\date{Copenhagen 1822.\\
  At Fr.\ Brummer's Press.\\
  (Printed by C.\ Gr\ae be.)}

\begin{document}
\maketitle

\frontmatter

\chapter*{Preface.}
\markboth{Preface}{Preface}

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

There are many writings on the study of the sciences [Videnskaber] in
general, and on academic study and the student's vocation [Kald] in
particular. The present one is a new attempt of this kind, following a plan of
its own. If giving oneself a full account of what one does and accomplishes
can do nothing but advance a person in his striving and activity---so that he
may come to act with complete clarity and composure---then this must be
especially important for the one whose whole endeavor is precisely to examine,
everywhere, the innermost nature and essence [Væsen] of things. And though
there may admittedly be, in other spheres of activity, people who, instead of
being advanced, would be held back and arrested---indeed led astray in their
striving---were they to strain to penetrate into their own selves and search
out all that underlies their active power and the natural gift they have to
thank for it, this cannot hold of the student as such; for he, called to
penetrate to the ultimate grounds of all cognition [Erkjendelse] and of all
things, and above all to ponder the nature of spirit [Aand] and of the
spiritual life, cannot exclude from that inquiry the form of life that emerges
in his own inner being.

This is why, in our own time, reflections on academic study have been made the
subject of separate lectures at the universities.

But if one considers more closely what belongs to such a treatment and what
does not, difficulties arise---difficulties that the author of the present
attempt, too, could not but feel. To clarify the nature and essence of
scientific study, one must enter into considerations that other scientific
disciplines nonetheless claim as their own, and the question then becomes how a
boundary is to be drawn between the two. Let me be permitted a few words on
this matter here, so as to clarify the relation in which this presentation is
meant to stand to the other scientific developments it meets at several points.

To show what a person who strives after genuine cognition and science
[Videnskab] ought to keep in view, the first and most important thing is to
consider what cognition and science are. But underlying all cognizing lies
thinking [Tænkning]. The ordinary developments of logic, however, cannot belong
here. Just as little does it belong here---though the life that is to live and
take shape in the scientific inquirer [Gransker] is indeed the object of our
consideration---to set out the various life-activities that are to be
distinguished in our spiritual life, together with their relations and
connections, or to enter into psychological investigations. Nor should an
introduction to academic study be confused with that general introduction to
philosophy which has, with good reason, lately been established as a discipline
of its own. Yet this latter cannot but come into manifold contact with all
these investigations. And so the question arises what an introduction to
academic study really is to be and to accomplish; indeed, doubt [Tvivl] may
well arise whether it can even claim and fill a place alongside those other
scientific presentations.

We bring this matter to clarity most easily by asking how we came---how others
before us have come---to make academic study a distinct object of distinct
investigation. It was because it had to matter to the student, who as such has
his own particular vocation, to reflect on this vocation of his, on his
striving and activity. Among the forms in which the general spiritual life that
psychology considers can come forward in a particular individual existence,
hypostatizing itself in a distinctive way, the life that is to live and dwell
in the student is one; in other individuals, in other spheres of activity, it
takes shape in other ways. What now serves to illuminate this form of life from
every side, and to remind us what it really is that we as students strive for
and ought to strive for---and what we, in the midst of a life in which so much
can distract and confuse, or draw us toward one-sidedness [Eensidighed], hold
us back and detain us, ought constantly to keep in view---this is what belongs
here. And thus there appears a distinctive center around which the manifold
[Mangfoldighed] that presents itself to reflection must gather into unity
[Eenhed].

A presentation of this kind may be regarded as a sort of monograph. Of such
self-standing scientific treatments of the whole nature and character of
certain particular objects---or of objects belonging to larger scientific
systems---one can conceive two kinds. The usual natural-historical ones, and
those resembling them, treat their object exactly as it is otherwise treated in
its place within the doctrinal edifice to which it belongs, and so are to be
regarded merely as excerpted sections or pieces of the larger scientific
whole---perhaps only elaborated more fully than usual. Of a different character,
by contrast, is a monograph such as the present one. Items of knowledge that
otherwise belong in the most diverse places in other sciences are here to
gather about a single point, in order---together with, and in one with, the
reflections on what is peculiar to the object---to illuminate it fully and from
every side. Such a monograph, more difficult and intricate than one of the
first kind, also has a better claim to stand forth as a self-subsisting
scientific inquiry, or as a smaller whole within the great whole of the
sciences; for here there is a distinctive center for a scientific organization
of its own, whereby a body of interconnected cognitions grows together into
unity in a peculiar way. And this organization---the inner ground-form of the
presentation, the inner mode of connection and sequence of its parts, the
course of its development---is determined by the object's own distinctive
fundamental nature, not by a norm and manner of presentation borrowed from
another science and fixed by that science's course and method [Methode]. A
monograph of this kind stands to the sciences to whose fundamental
considerations it attaches itself---and to which much of what it presents and
works out is to be traced back, though always in a distinctive way determined
by the object, and in a light of its own---as a talent stands to the general
activities of the mind. Take away from a talent all that belongs to
understanding [Forstand] and judgment, sense and imagination, reason [Fornuft]
and feeling, and nothing will remain that could be called a talent; and yet a
talent is something other than those more general activities of the mind. What,
then, is a talent? It is the distinctive combination of them all in certain
definite directions---the distinctive organization of the mind with respect to
their united cooperation, aimed at a certain kind of productivity. Just as a
thorough psychological portrait of a given talent does not consist in a
portrayal of those more general mental activities in general, but must
presuppose them as known---while it is precisely they that it must everywhere
attend to and keep in view---so a monograph like the present one may have to
look toward many particular points which, taken in and for themselves, belong
to definite places in definite sciences, and may have to present much that,
piece by piece, other sciences could claim and draw to themselves, and can
nonetheless be, and assert itself as, a distinctive scientific whole in its own
right.

It was, moreover, my intention from the outset---the reader will perhaps notice
this easily in several places---to lay the content of the present work before
the public in a form other than the present one: in the same form in which I
have several times imparted it to my listeners, the form of lectures. The
lecture is, among us, without doubt the mode of presentation that could most
easily take the place the dialogue holds in Plato; like the dialogue in his
day, it is taken directly from life, and the development of general concepts can
hardly, in any way natural to us, be better combined in a connected
presentation with speech's immediate address to individuals---and take on the
liveliness that must follow from such direct attention to the thinking beings
who are to be called upon to meet what is presented with their own independent
thinking [Selvtænkning]---than in the form of the lecture. But the fear that I
might not come even remotely close, that my attempt might fall all too far short
of the ideal of such a presentation which hovered before me, brought me to
abandon the execution I had already begun.

I had also intended from the outset, in the Remarks to \S\S~13--17, to treat
the individual principal fields of study, both with respect to the philosophical
and in the other principal respects. But what I might have said about this I
later decided to set aside and reserve for the oral lectures I hope still to
give on academic study, for those students who are still at the beginning of
their course. I dared not credit myself with such an acquaintance with so many
sciences that I should be able, even in general terms, to offer reflections on
their study fit to be laid before the public in print, in a word fixed for
good---or at least for a long time.

I must further note that if, in the present work, I have departed from the
practice I observed in the two scholarly works I have previously
produced---namely, of emphasizing no word---this has happened because I believed
I ought to yield to the urging of one of those among my friends to whose company
I owe much, especially with respect to my philosophizing endeavor. But since I
am not accustomed to underline as I write---because it almost never occurs to me
that it might be necessary---and since this time I had nearly finished the whole
book before the decision to do so was prompted in me, so that I carried out the
underlining only at the work's final revision, whereas the moment of writing
something down is the right moment to decide which word should be
emphasized---I dare not hope that I have carried out the decision with the best
success, and I therefore ask the reader, who will in any case find enough else
to overlook, to extend his forbearance to this too.

This is what I had to send ahead of the work I herewith deliver into the hands
of the public.

\begin{flushleft}
Copenhagen, 13 November 1821.
\end{flushleft}
\begin{flushright}
Sibbern.
\end{flushright}

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

\chapter*{Synopsis.}
\markboth{Synopsis}{Synopsis}

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

All cognizing [Erkjenden] is essentially a finding, a receiving, and a
beholding. Truth, presenting itself out there of its own accord, refers itself
to an apprehending individuality and carries with it, for the latter, a
necessity that produces conviction [Overbeviisning], to which a personal
affection is joined. (\S~1.)

Yet we cognize only insofar as we ourselves bring forth. \textit{Scimus, qvia
facimus.} A living personality cannot receive except in and through its own
forth-working. (\S~2.)

Receiving and producing here become one. Precisely by ourselves shaping truth's
expression or presentation within us do we receive it and are made alive by it.
(\S~3.)

In pure beholding, however, productivity is lost and does not come into view.
But in the striving to cognize perfectly, reflection makes it become free and
show itself as such. If we ask why we come to truth only slowly, through
reflection---why its fullness is not given at once---the answer is: because a
realm of freedom is to come forth, so that a realm of love may become actual.
(\S~4.)

The thinking called forth by reflection now leads to a cognizing by force of
grounds [Grunde]. The highest cognizing would indeed be a pure beholding of
perfect truth, raised above all inquiry [Gransken] and grounding; but such an
immediate beholding of the highest is not granted to the human being. Rather,
all his assurance rests, in the end, on an immediate faith [Tro]. (\S~5.)

The endeavor to cognize by force of grounds is an endeavor to cognize truth in
its truth. This is what calls forth genuine scientific study, which first truly
begins with academic study. (\S~6.)

Cognition now becomes an intelligent cognizing (as opposed to the empirical); a
cognition of truth in its truth is necessarily a cognizing by force of the
constitutive First, hence a cognizing \textit{a priori}, which leads to a
construction \textit{a priori}. (\S~7.)

Such a cognizing then aims not merely at grounds for our assuming something, but
at things' own grounds of existence [Tilværelsesgrunde]. All philosophizing
strives to find these. (In mathematics, by contrast, we have a great example of
a cognizing---even a wholly intelligent one---that does not aim at this.)
(\S~8.)

The real grounds of things we find, most immediately, in the rational
orderliness [Forstandighed] that, within a coherent order of things, requires or
entails them---indeed, in the concepts [Begreber] of things. (\S~9.)

But as the ultimate ground there appears an all-real active First which, in
bringing a determinate order of things into actuality, sets the individual
things into existence in accordance with a rational lawfulness
[Forstandsmæssighed]---indeed, in accordance with concept. (\S~10.)

This all-real presents itself to us in the Idea [Idee]. Cognition is in itself
the Idea's own self-presentation in consciousness, even if it does not always
show itself---or is not seen and possessed by us---as such. That truth is not a
mere correspondence between an assumption and its object, but is that which
livingly produces this correspondence, is clear from this. (\S~11.)

The Idea presents itself to cognition and comes to outward life and subsistence
in science. (\S~12.)

The first thing in all genuine science is the Idea itself, as science's center
and all-determining soul. In the cognition of it, and in the tracing-back of
all cognitions to it, the speculative and philosophical element in science shows
itself. Yet philosophy can also show itself already in the immediately
judicious---in a certain scientific way of thinking and judging. (\S~13.)

In speculation's transition to the manifold---which is thereby to be beheld in
its proper light---and in the determination of that manifold, dialectic enters,
partly as a philosophical art of annihilation, partly as a scientific art of
construction essentially and immediately bound up with it. (\S~14.)

Genuine cognition is, further, necessarily and essentially systematic. Yet the
immediately free judicious element, just as it must prepare the way for the
system, must also not be displaced by it. (\S~15.)

Furthermore, in all science the exact study of the particular [Detail] is of
importance. (\S~16.)

Finally, in all science method [Methode] is still to be noted. It is best
developed out of the science itself and its historical course of
development---just as the study of a science's history is of essential
importance for the study of the science itself. (\S~17.)

Not only in science, but also, through science, in a personal individuality does
the Idea take shape, and thereby come to a complete outward existence or
subsistence in real actuality. (\S~18.)

To attain this subsistence and life in the individual, the Idea then calls forth
its own opposite, setting the individual's sense of independence in motion and
summoning a recognition and avowal [Er- og Vedkjendelse] that is to follow on
the individual's part. (\S~19.)

Independent thinking [Selvtænkning] is called forth by a questioning. The
possibility of assuming, but also of not assuming, then appears. Denial and
affirmation stand opposed to one another. But the assurance that arises out of
doubt [Tvivl] beaten down---the Yes that comes forth from the denial of a
negation---is of an entirely different nature than what has not passed through
this trial. (\S~20.)

The Idea's coming to the individual then appears as a kind of testing, and
everything depends on whether that now follows in the individual which is to
follow. There is indeed something here that lies within the individual's power;
but that on which everything depends does not lie within his power, and must,
when it is given, be received as pure grace. The conflict that is here to be
brought to unity is seen above all in theological study. Yet in several respects
there is a trial to be undergone here, since, on the one hand, egoism easily
comes to draw nourishment from the consciousness of the independent thinking to
which the summons has been made, and, on the other, the very endeavor after
insight can easily lead into uncertainty, confusion, and error. (\S~21.)

Finally, the Idea will live and take shape not only in individuals, but in a
whole of humanity, in which each is to find himself as a member and to intervene
practically in the exercise of a definite vocation [Kald]. In the choice of that
vocation, science and Idea must themselves guide us. We must, further, take
science with us into life, and the exercise of one's vocation must at the same
time be a constantly continued study. (\S~22.)

But the highest, the one thing needful---however lofty and significant genuine
cognizing may be---is nonetheless not yet cognition itself. (Conclusion.)

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

\mainmatter


\begin{center}\large\S.~1.\end{center}
\markboth{\S~1}{\S~1}
All cognizing [Erkjenden] is essentially a finding, a receiving, and a
beholding. (\textit{Scimus, qvia accepimus.})

For all cognizing aims at the true, and in and through it we are to come to
truth [Sandhed]---to come to have and possess it. But truth is not something
dependent on us; it is and remains, without any regard to us, what it is in
itself. Ours, then, can only be to seek quite directly to become conscious of
it as it is in itself and presents itself to us; but such a coming-to-consciousness
is a receiving, and---when it is determinate and clear and at the same time
immediate, or certain and self-evident by itself alone---a beholding. But this
immediate, self-certain element is present at every point in all cognizing,
even when we reach it in a mediated way; for cognition then still always rests
on the insight that, from such prior cognitions, the present one follows---and
this following, or consequence, must be self-evident directly, or must be
beheld. With this, what thus follows likewise shows itself as that which, on
the presupposition of the truth of the prior cognitions, is self-evident,
though only on that same presupposition is it also cognized. So that even
relative and mediated cognition, at the point where it first enters by virtue
of its mediateness, proceeds from a direct insight, and so far is itself an
insight and a beholding. It is a beholding that gradually develops. In every
case it is a receiving.

As such, it is always and essentially bound up with a necessity to represent
[forestille] just as one does represent, and not otherwise. The immediate
feeling of this necessity is conviction [Overbeviisning], which can arise only
in that we are immediately seized by it and feel it. The object of cognition
presses itself upon us, and precisely as it is now cognized to be.

The true that is cognized thereby shows itself as that which exerts a
determining and effective influence in us---an influence in which it expresses
and presents itself, and thereby lives again, in us.

\begin{center}Remark and Supplement.\end{center}
1.~That all cognizing is a receiving and a beholding is soon taught, too, by a
glance at the various domains of cognition. Not only what strikes the eye at
once: that the \emph{beginning} of all cognition---in the higher spheres just
as much as in the lowest sensory one---is a direct becoming-aware and
apprehending, a noticing and sensing. In its advance toward the highest
completion, too, it remains an immediate taking and beholding. In the domain of
lower sensation this is at once, at first glance, entirely obvious; for although
here a certain combining and synthesizing activity must indeed be added to the
apprehending one---so that sensation comes about only by means of a reproducing
construction---this construction too aims only at presenting the thing exactly
as it is found and as it presents itself to us. We feel ourselves affected, and
just as the affection brings it about, so sensation turns out: precisely as we
are determined to it by this affection, so must we represent. And though a
difference not seldom arises here between what seems to be and what is found to
be in itself, still this latter too is found, and must precisely be cognized,
only as it is found through a scrutiny---if not by sense, then by deliberating
reflection.

As it here shows itself in the sphere of the senses and of the experience they
occasion, so it stands with every higher cognizing of the understanding
[Forstand] or of reason [Fornuft]---indeed, it is especially striking in the
highest cognizing. Whatever is cognized without appeal to experience, by an
insight springing from pure thinking---whether by force of mathematical
demonstration or of a priori reflection or of philosophical speculation---is
always cognized as something that is seen to have to be so, and a necessity of
understanding or of reason, determining our whole representing, thereby enters:
as this necessity declares it to be, so it must be cognized to be. But when, in
this way, all reflection---all comparing, connecting, and inferring---aims at
investigating or fathoming what must necessarily be assumed, as it will itself
turn out to be, so that it is cognized as true only in being cognized as the
one thing that can be assumed, then reflection aims at letting the thing
\emph{give} itself, by letting it be forced upon one through a necessity of
thinking. It is cognized only in being discerned---that is, seen and beheld
(with the higher eye of the spirit [Aand])---as it is in itself. The triangle I
construct; but that its angles together equal two right angles I discern as
necessary in accordance with the triangle's own nature, once given to me.
Indeed, in the very act of constructing it, I do so in a certain determinate way
and under certain conditions, and this manner of construction and these
conditions I must take as they force themselves upon me. All sorts of human
circumstances and relations I can think out for myself (though even here I must
proceed according to certain given conditions and determinations); but what, in
a given relation, is right or wrong, good or bad, the fundamental Idea
[Grundidee] of the good and the right must teach me, so that by it I must let
myself be guided in my judgment, and in every case let stand as right and good
what, in accordance with that Idea, I see to be right and good. Mine it can only
be to inquire into the essence [Væsen] of the right and the good, and then to
cognize it as it presents itself to my spiritual eye. So that, even if the
cognizing which goes beyond mere experience is called forth by questions---which,
as inner spurs to seek more than the experienced, show that we do not rest
content with what is given in experience alone---still the whole thinking
thereby set in motion leads only to a new receiving and apprehending of a
higher order, under which we must again seek directly to observe and behold what
presents itself to us.

But now, in the receiving and beholding in which all cognizing consists, we also
become aware of a \emph{principal distinction}. What we apprehend and come to
know through the senses---indeed also what the ideal or intellectual
becoming-aware and beholding, which enters already in this sphere by way of
thinking, can teach us, when the inner ground-forms of the things revealed to
the senses, their relations and their activities, are cognized exactly as they
are directly found---all this is merely received, as something given to us from
without in the reality that presents itself in time and space. This cognizing is
the merely \emph{empirical} kind; the necessity by which it enters, and by which
it acquires the character of a cognizing, is only a necessity of the sort that
proceeds from a certain \emph{affection} bound to time and space---a wholly
sensory affection of a lower nature. What is cognized in this way is still
cognized only as \emph{contingent}. It is given to us as the actual, and for us
there is a necessity to take it so; but that it should itself necessarily be as
it is, is not yet seen in this merely empirical cognizing.

Of a wholly different kind, by contrast, is mathematical and philosophical
cognizing; here there is an intellectual beholding that has nothing of that
stamp of contingency. What is here cognized is not merely taken as given, as
though that were enough, but is apprehended and represented as something that is
so because it must be so, in accordance with an inner \emph{necessity of
understanding and reason} of a wholly different origin than that by which
certain sensory affections force themselves upon us. It is indeed given to us,
but given from within---or at least given as something that accords with an
inner demand, and not at all given to us by way of sensory affections. It is
cognized as something that cannot be otherwise; it is cognized as
\emph{essential}, and therefore valid for all times and everywhere---which is
why only in this sphere can one speak of eternal truths. We will call this
cognizing the \emph{intelligent} kind, both because this word seems more apt
than the word ``intellectual,'' and because the latter can also cover the
just-mentioned merely empirical---yet still ideal---cognizing, in which the
inner ground-forms and relations of things are apprehended entirely as they are
directly found in experience.

2.~Necessity enters when truth shows itself in relation to (refers itself to)
\emph{the individuality} that takes it up; and in the feeling of conviction, the
life and working of the true in the individual is met by something purely
individual, in which a person shows himself and comes forward as such, in that
he attaches himself wholly to, and gives himself over to, that working. In pure
receiving and beholding, the true was as yet seen only as presenting itself, so
that only its own (universal) appearance showed itself therein---no personal
life. The reception was merely like a reflection or mirroring, like that by
which water gives back light and color---not a personal taking-up, like that by
which the gemstone has, in its color, taken a power and effect of light into
itself, into unity with its own nature and existence. But in conviction it
becomes an inward appropriation, whereby the true that presents itself in and
for an individuality now also comes to dwell powerfully in this individuality,
as in a being in which it has taken its seat, and which itself now is and
becomes something through the reception. In this binding-together of the true
and the individual into unity, the necessity that is indeed present is so lost
that it is not particularly felt or brought into view. Here there is an
\emph{undivided} free life in the true, raised above the feeling of necessity.
Only when the individual, tearing itself loose from the universal, moves in a
particular personal striving does the opposition---and with it that same
necessity---come to light.

3.~In conviction, a personal \emph{affection} and emotion, πάθος, is added to
the reception---with a personal attachment and devotion to the power of
truth---so that the individual not only lets truth gain room and power within
him, but does so with delight, feeling that in this he possesses something and
that his personal life is thereby expanded and heightened. The devotion now
shows itself as \emph{love}. This brings it about that, together with the power
of the true, personality and freedom too come fully forward; for in love an
independent being goes willingly and gladly to meet the true. Whereby the
endeavor to cognize shows itself as a cultus or \emph{worship} of truth. The
inquirer's [Gransker] relation to truth and to his science [Videnskab] now
becomes a \emph{religious} relation---in the not ordinary, but literal and
far-reaching sense in which the word denotes a relation, bound by sacred ties,
to something holy, in whose power we feel ourselves, with willing devotion, to
be. (Indeed, in the word's more ordinary and proper sense too the relation can
show itself as religious; for just as a human being can and should do all that
he does with God in view and out of love for God---so that this highest motive
can and should be the everywhere-operative one---so too not only can that same
motive, as such, embrace the scientific endeavor as well, but everything can and
should in the end be cognized and beheld in its relation to the highest divine
in such a way that the Idea of God becomes, scientifically, the all-illuminating,
just as, practically, the all-animating. So that in the end every relation to
truth shows itself directly as a relation to God. But in this we look toward an
ideal that the scientific culture [Videnskabelighed] of humanity is still far
from having approached.) With reverence for the voice of truth, and a heartfelt
desire to bring it---and it alone---to speak and live in his inquiry, the votary
of science must give himself over to being its organ, its servant and herald,
who wills only what it wills. His personality must be as a holy temple for it.
But at the same time the speaking and living of the true within him must be felt
and held fast as his own innermost and best life, and the relation must show
itself as a true relation of love. What has sometimes been said---that a person
is consecrated to his science and his art---we may recall here. To be kindled by
truth, and then to take it up and faithfully preserve and tend it within us,
until it grows and develops of itself and takes shape in us, so that it may
truly come to fulfillment in us---this is ours. Let love of truth
[Sandhedskjerlighed] be, like all genuine love, careful and conscientious.

Truth thereby shows itself as that which, \emph{by its own might and power},
seeks to take shape in the individual, while the individual gives himself over
entirely to letting it work in him whatever it will. And just as this holds of
everything that, in any science or body of knowledge, shows itself as the true
whose cognition study aims at, so everything set forth here holds of every
scientific endeavor---even of one that still stands so far from the higher Ideas
that it only rarely feels itself inspired by them. In \emph{every} study,
faithful devotion to the cognition that forms itself in us will bring the
fullness that love gives. If, in the lower domains of inquiry, what we cognize
seems not to bring us nearer to the eternal, still the fact that it produces
complete cognition and living conviction in us will make us feel as though
filled by a divinity. (Just as, even in the lowest spheres, cognition in itself
really does come from the higher source, even if it does not come to
consciousness as such. See \S~11, Remark~2, below.) So that every study in which
faithful and clear inquiry prevails has its \emph{Muse}. In it the individual
stands over against something he would make his own, needing it as something
without which he himself cannot come to full life. But what he would thus make
his own is not something he can treat as an object for the satisfaction of a
sensory need---to consume, or to take as a mere vehicle or means. It always
remains standing over against the individual as something that subsists in and
for itself in its significance and worth, whether or not it comes to the
individual---indeed, as something that has worth for the individual precisely
only insofar as it has worth and significance (that is, in this case, has truth)
in and for itself. And thus, everywhere that truth is sought, in whatever
sphere, the relation between the individual and the true he seeks is such a
relation of attachment as can---and, under a genuine striving, also will---become
a relation of love.

We now also see here what it is to \emph{inquire}, to \emph{fathom}, to
\emph{reflect}---or, as one less felicitously puts it, to carry out reflection,
investigation, consideration upon something. Namely, that it can be nothing
other than, under this or that arrangement and handling of certain
representations, to try to bring it about that the truth we desire to know may
come, of itself, by its own might and influence, to present itself to us---so
that we now directly become aware of it or, as one says, find it out.

We must therefore also \emph{immerse} ourselves in every study. Only thereby
does it become genuine and powerful. If the true that produces cognition in us
is to gain us into its power, so that it may form and shape itself in us, then
we must forget everything else over it and, as it were, lose ourselves in it.
Even the sensory forms of things are only then observed and apprehended fully in
their distinctiveness and manifoldness when attention is thereby wholly
captivated, as though we were entirely absorbed in the contemplation, or lost in
it; for only then does the living \textit{nisus formativus} that here
enters---under which the true would take shape in us---work whole and undivided
in its full strength. A book is best studied and works most fully when, for a
time, it so captivates us and, as it were, takes our whole mind into possession,
that we are wholly immersed in it. In this way---that is, with this vigorous
attention and devotion---should every \emph{classical} work be studied.

But only \emph{classical works}, too, deserve and can be wholly studied thus.
For a scientific work is classical not by the collection of items of knowledge
it may contain, and which can now be drawn and learned from it piece by piece,
but by its spirit as a whole, its inner genius, which speaks out of it in its
totality, and by the total form that accompanies this---the way in which
everything in it is presented and composed and, as it were, grown together into
unity, so that everywhere in the work we meet the same vigorous and profound
spirit, and all the manifold in it is as though of one piece. Such a work
stands, by its spirit and by its form, valid and significant for all times,
which the word ``classical'' especially seems to indicate; for it works not
merely by a manifold of material that can be transplanted and reworked, but
works as a whole by its distinctive original stamp, whereby spirit speaks to
spirit. But then this spirit too must work in its totality and, as it were,
enter into the reader or beholder---and it does so all the more, the more he
immerses himself in the contemplation, so that he treats the work not merely as
a means to the acquisition of knowledge, but as a work that has full worth by
and in itself alone, and therefore also works toward the very formation
[Dannelse] (\textit{Bildung}) of the spirit as a whole.

We must read in such a way that we enter into a \emph{harmonious relation of
reciprocal action} with what we read, and are so set in motion that, through the
words and sentences, we come to feel ourselves addressed by the general
spiritual element in the work, which constitutes the inner source and first
mover (\textit{primus motor}) of the words and sentences---the soul that lives
and speaks in them. Only such reading is a truly soulful reading. The study of a
work thus becomes an inner communion with a glorious spirit, which in several
respects can have a beneficial influence on us and advance us in our whole
striving. It becomes, throughout, all the more genuine nourishment for us the
more it is a \emph{disinterested} study; for the more we inquire and probe not
for the sake of this or that particular aim, but solely out of pure love of
truth, in order to take it up entirely into ourselves, the more it will work in
us with its full power. The more our devotion is undivided, free, and
wholehearted, the more easily and fully can the total immersion, or
being-gripped, come about. For a time, then, we let the work carry us away, and
let it do with us, and make of us, what it will. Thereupon truth's own power
itself begins to speak in us and to bear witness to what is expressed in the
work; whereby an immersion in our own immediate contemplation of the
intrinsically true is added to the immersion in the work's utterance, and now
determines what the work is to make of us.

In general, the spirit that lives and addresses us in a work can, of course,
come to carry us away and captivate us, and become for us a source of new
effects and formations of life within us, only in that it comes from truth
itself---indeed, from that which can present itself immediately to every
inquirer, enter into everyone, and in everyone bear witness to itself. Only in
that we find the same thing again in ourselves, so that we sympathize with it,
can what is handed down to us speak to us through the forms of transmission in
which it is imparted. (This holds even of the study of the works of nature;
nature too is a book that must be studied like any other book, but in which a
spirit akin to us also speaks to us: the same truth that bears witness to itself
within us as well. Yet this will only become fully evident in what follows. See
\S~7, in the Remark.) To grasp, cognize, and enjoy a work handed down to us we
come, in the proper sense of the word, only in that it \emph{reproduces itself
in us}, and we, in the course of this its reproduction, are moved and enlivened
by the spirit which---just as it now lives again in us---first lived in the
activity by which the work came into being. But by this we are led directly to
the next principal consideration, as we take into closer examination this
reproduction, which shows itself as one whereby we ourselves are productive.

\begin{center}\large\S.~2.\end{center}
\markboth{\S~2}{\S~2}
All cognizing comes to actuality and subsistence by means of a determining, a
doing, a producing---and only by means of such; so that the latter is
essentially bound up with the former. \emph{We cognize only insofar as we
produce.} (\textit{Scimus, qvia facimus.})

Cognition takes place when the true shows itself as such to a personality and is
taken up by it. But a personality must reveal itself in a personal life. Now a
life cannot be conceived without a Something, a real [Reale], in which it
expresses itself in uninterrupted activity, continually forming and determining
this real, or outward, of its own; it gives itself a definite shape in giving
this a definite shape. Personality, on the other hand, shows itself in finding
and positing itself as the immediate center of its own existence [Tilvær], and
in having this existence as its own. A personal life, then, must express itself
in an activity and productivity in which the personal being finds and moves
itself as the producing agent. If, then, the true is to be taken up into the
personality, it too must be embraced by this personal productivity. The
individual can therefore cognize only in that it itself works and produces a
real---a (psychic) outward---corresponding to the cognition, in which it holds
the cognition fast by determining and developing this outward.

Consciousness is the personality's first expression. In it is taken up
everything that is to be taken up into the personality. Thus we too can cognize
only by becoming conscious of what is to be cognized. But this, according to the
fundamental nature of consciousness, we can do only by presenting it to
ourselves, or representing it to ourselves [forestille] (objectifying it for
ourselves). But this presenting or representing is our own act of fixing,
whereby we posit (or \textit{ponere}) the given and received before us and, as
it were, make ourselves an inner image of it, in which we express it for
ourselves and hold it fast. In this representing consists, here, that
productivity. The representation [Forestilling] is our own product and, as an
expression of our own life which we keep as ours, cannot be anything else. But
only in this constant representing, and under its constant determination,
development, and completion, do we come to cognize, and do we have what we
cognize.

The possession of cognition therefore also shows itself as an \emph{ability}, a
\emph{capacity}. We know only (sufficiently and fully) what we are able to bring
forth in one form or another and thereby give to others. This productivity of
ours, by which the received is now embraced, is also in itself a free one, which
can arbitrarily present something other than the true. In cognition, however,
this free element is bound and determined to a single way of representing, by the
necessity that accompanies cognition and now comes forward in this respect too.

\begin{center}Remark and Supplement.\end{center}
1.~Here we see the essential connection of \emph{language and speech} with
cognition. All cognizing that does not aim at immediate or reproduced
sense-intuitions, or at constructions carried out directly in time and space in
the same way as these (such as, for example, the mathematical ones), takes place
by means of a constant judging, which in turn takes place through an inner
speaking, by means of words (or possibly other signs), and cannot come about in
any other way. But that this is so---which experience teaches directly as a
fact---also shows how immediately a positing, determining, and producing
activity is added to our apprehending and receiving. In word and speech we set
what is cognized before us, thereby binding it and taking possession of it,
making it something of our own, which we now, in our own way---in our
designating and naming---hold fast (cf. Hegel's \textit{Encyklopädie}, § 379 and
the following §§).

2.~Yet not only by words: in all representing generally we do the same. The
preservation of representations after the immediate beholding is past, as well
as everything we can do with them, testifies to this generally. The productivity
essentially belonging to all cognizing also soon shows itself, to the observing
eye, in every one of cognition's domains. As for simple \emph{lower sensation},
we refer to the psychological observations that lead directly to the insight
that spatial forms and relations, as well as temporal determinations and the
qualities ascribed to things in respect of these, are cognized only by means of
a reproducing construction of the objects that present themselves---observations
which, already merely as such, lead to conclusions that justify assuming the same
as regards the merely visible, audible, and other merely sensory qualities. (See
the author's \textit{Psychologie}, § 55.) How \emph{thinking} takes place, and
the cognizing springing from it comes about, by means of a constant determining,
positing, and constructing, the logical investigations teach throughout and
everywhere. Whether concrete objects are conceived and their nature thereby
cognized by force of concepts [Begreber], or concepts themselves are regarded as
the objects to be cognized, in either case construction takes place. For the
former happens only when a connection is discerned; but this connection is seen
only when the manifold whose connection it is is set apart from one another and
then joined again into a whole, by means of a reproducing construction or a
genetic procedure, under which we let the thing arise before us---that is,
ourselves form it for us in an image---and now also see everything in its place,
in that we see the necessity of setting it thus into the whole. Connection, and
the real significance of objects within a certain coherent whole, can be
cognized only by projecting and completing the intuition [Anskuelse]
corresponding to it; to cognize it is to produce, or at least to be able to
produce, its concrete expression. The concepts themselves, next---which are
thereby abstracted, as the connections thus beheld in things and the
ground-forms and real meanings thereby determined, or as the fundamental
thoughts expressing themselves in things themselves---are cognized only when
they are lifted out on their own, by cognizing their so-called content: that is,
the sum of determinations that, according to the concepts, must be found in the
object and can be asserted of it as predicates. Whereby the judging mentioned
above enters, as a determining, a fixing, and a pronouncing. Deductions of
concepts---whereby the fundamental thought that determines the concept's whole
validity and content is derived, or seen in its origin from a higher
cognition---take place, like inferences and proofs, through trains of thought in
which thoughts are linked to thoughts by determinations, whereby a whole of
thought is developed. The deduction of concepts takes place only through a
construction of the scientific whole, in which the concept now shows itself as an
essential member; and proofs likewise come about only through combinations. An
Idea [Idee], finally, is likewise cognized only in that it is either beheld as
the expression of a higher Idea---in which respect, on the side of form, it shows
itself as a concept---or else beheld as the essence, the determining being, and
moreover as the real principle, giving essentiality and existence, in real
objects and ground-forms; which objects and ground-forms are then not only, like
all others, cognized solely by means of constructions, but can also be seen as
the expression of the Idea only in being seen as those which through it become
what they are---something that can be seen only under a reproducing, genetic
presentation.

\begin{center}\large\S.~3.\end{center}
\markboth{\S~3}{\S~3}
Reception and production, beholding and one's own determining and doing, here
become entirely one, so that cognition, precisely in entering under the one
form, essentially also takes on the other. (\textit{Scientia et potentia in idem
coincidunt.})

In cognition, the production and presentation is not arbitrary but one
determined with necessity, and what determines it is precisely our receiving and
beholding, and the true that presents itself in and for it. As thereby the whole
manner and character of the representation [Forestilling] is determined, the true
expresses itself---but through our productivity, which it governs and determines,
and in our product, which now nonetheless becomes the expression of that true,
and, as it were, becomes the material in which the true realizes and
individualizes itself in us, while at the same time our own life too realizes and
substantializes itself therein. We are determined in beholding, but determined to
produce something ourselves---namely, to form representations that become ours,
yet such that we produce them precisely so as to correspond to our
having-been-determined. And thus we are determining and determined in one. What
is given to us or presents itself to us comes to clarity in us---indeed, comes to
be ours at all---only in what we ourselves do. And yet here too cognition takes
place only by means of a constant receiving and beholding; for that by which we
are determined we have only in this its determining influence, and the
representation likewise, as necessary in accordance with that having-been-determined,
takes on the same character. \emph{We produce as we see that we must
produce.}

The personality, insofar as it finds itself as a consciousness and
self-determination---or we, who call ourselves ``we''---find ourselves here
between \emph{two regions}. On the one side, the influence by which we are
determined, and to which we turn and look as we receive and are determined; on
the other side, our sphere of representations, which we ourselves work over and
make into what it becomes, in that we seek therein to hold fast the received and
make it our own. The former, as determining, above us, before us; as that which
is to be expressed and to come to shape in us---the more inward and active, that
by which we are governed. The latter, as our product, below us; as that in which
we express our life---the more outward, the psychic material of our life, by its
nature an object of our governing. Here there is always, on the one side,
something that is obtained and possessed merely through the mastery it exercises
in us, or the determining influence it has on our working and determining; on the
other side, something that is merely expression, presentation, or result, which
is had and kept only through a capacity or skill to reproduce it.

This latter may now remain behind as merely bare, lifeless material, like a
residuum of the activity---as something now merely had and assumed because it was
once assumed, a \emph{standing result}, without its being any longer enlivened by
the living activity by which it was \emph{formed}. But it should not be
mechanized or crystallize itself in this way; rather, it should be kept
constantly ready, in the course of life's advance, to enter again under the
living influence of that determining First, and, under a constantly living and
organic rebirth, to receive transformations and new shapes. Only then is the
relation preserved as it is in itself and therefore also ought to be. As the
influences from above, from within, from that which is before us, direct and
command, so ought we to work in our lower and outer sphere.

\begin{center}Remark and Supplement.\end{center}
1.~In all cognizing by means of thinking (concept and Idea) this opposition and
doubleness is clear. Concepts, as such, are had and held fast only either by
means of the constructions whose norms they are, or by means of the judgments by
which their so-called content is determined. But the norm, or determining
fundamental thought, of a construction indeed enters and comes forward in it as
governing and ruling, yet is not itself constructed; it is therefore had only in
the living activity that proceeds from it. In judgments a concept is expressed,
in that its content is determined; but since this content is a sum of
determinations ascribed not to the concept itself but to its objects---yet in
accordance with the concept---the concept here too is had only as the determining
fundamental thought of a doing and determining, and is not embraced by it, but
hovers over it, or constitutes the inner first moving ruler within it, its inner
\textit{primus motor}. Which holds still more of every Idea. The concept can, to
be sure, in turn be regarded as an object that is itself determined; but this can
then happen only in accordance with another concept, which then constitutes the
ruling fundamental thought in the determinations. Such ruling fundamental
thoughts or concepts, which are had only in what is posited and carried out by
force of them, also underlie all inferences. In lower sensation, too, a
determining first enters into opposition to the representation thereby
determined. In striving to see and hear rightly and precisely, and to obtain a
complete representation of the object in which it can show itself in its full
distinctiveness, then---precisely because the need for one's own doing now comes
forward---that which is to be beheld will also show itself as that toward which
one must constantly look: that is, as that by which that doing is governed and
mastered, and which is obtained only by developing, in accordance with its own
determining influence, a complete image of it.

2.~The apprehending and cognizing individuality here shows itself as one which,
just as in its representations it forms a material in which it expresses its
cognizing and life, so too is itself formed into being that in which truth
expresses itself---quite as if the individuality were itself, in turn, the
substrate or material of the self-presenting truth, in which the latter seeks to
come to appearance and to take shape (cf. § 4, Remark 3). \emph{Thereby we grow},
on the one side, \emph{into that} whole in which a universal life expresses and
realizes itself, in that we are embraced, filled, and made alive by it as members
or organs; while on the other side a whole also grows and develops in us, in
which we ourselves realize ourselves as personal beings subsisting for ourselves.
We are ourselves assimilated by that more general life, in the very act of
seeking to assimilate to ourselves the true that presents itself to us, by
assimilating to ourselves its expression within us.

I refer here to the ingenious Franz Baader's treatise in Schelling and Markus's
\textit{Jahrbücher der Medicin}, third volume, first part, as well as to several
of his shorter philosophical-biological writings.

\begin{center}\large\S.~4.\end{center}
\markboth{\S~4}{\S~4}
In simple, pure, wholly immediate beholding, productivity is \emph{bound} and
\emph{lost}, or is so absorbed therein that it does not come into view on its
own, or come to consciousness, and stir according to its particular nature---just
as neither is the influence, with its determining and compelling power, felt and
brought to consciousness as such. The opposition between the cognized true and
the individuality in and for which it comes forward does not here show itself as
an opposition. This cognizing is also, as regards the manner of cognizing, the
most perfect; for the fact that the production comes entirely of itself proves
that truth works with complete power, and that it is also taken up fully.

But in a striving to cognize quite fully what presents itself immediately in this
beholding, and at the same time to cognize still other and more---because what is
thus cognized points to other and more---the \emph{object of cognition and the
cognizing personality}, which in pure beholding were undivided, step, by means of
a reflection, \emph{apart from one another} and against one another. Productivity
becomes unbound and shows itself as free, while the determining influence, on the
other side, shows itself as such, with its necessity. The individual finds itself
as one that must itself work, by way of deliberation and thinking, to reach the
desired cognition; and just for that reason productivity must show itself as a
free one, capable of presenting something other than what is found. But now,
therefore, the attachment too will be embraced by self-determination and will
occur in accordance with a willing. Truth now shows itself as an object that we
can come to possess only gradually, by effort, through a free activity that must
accord with truth's declarations---to which we must now attend closely, as they
gradually sound to us in the course of our inquiry. Whereby productivity, while
showing itself as free, is bound and mastered anew, so that it presents only in
the one true way---yet in such a way that it now occurs with self-determination,
and as one that could act otherwise, and therefore also by way of deliberation,
and such that a free choice is possible. But pure, wholly immediate
beholding---through a new, complete self-loss of the productivity that has come
to clarity and self-consciousness---is nonetheless that toward which one strives,
so that truth in its full fullness may enter into the cognizing personality.

\begin{center}Remark and Supplement.\end{center}
1.~The productivity that has become free can now also \emph{express itself in its
own} productions, independent of the given. But even therein it will always be
determined---since it must take place in definite ways---by certain norms or
concepts, which of themselves govern within it and are had directly in and
through the power with which they determine the self-activity. And if these
productions are to be more than a play with representations, then this freer
productivity---although its products are now not to be mere images of what is
found, but its own creations---must nonetheless, in the same way as in the
endeavor to cognize, be bound and mastered anew by a higher cognition, under
which the true shows itself to us in new forms and takes shape in us. So it is in
mathematics and in art.

In general, life comes forth just as fully in action and self-activity as in a
striving to cognize. In the former, free forth-working prevails, but also a
certain cognition, which underlies it as governing and moving First. But here too
the two opposites can, in the same way, be lost or merge into one, as in mere
beholding, so that neither is the self-activity particularly felt as free, nor
does the fundamental thought guiding it come to consciousness as such, since the
action happens entirely of itself: a self-activity that, as regards its manner of
working, is the most perfect---since therein the true and the being, the
essentiality that is to take shape in us in our acting too, shows itself as that
which works with complete, undivided, and unarrested power.

2.~So far as the free productivity comes to consciousness as such, cognition's
determining First too comes to consciousness as such; and vice versa. In a free
working at cognizing, as in an acting undertaken with self-conscious
self-determination, there can therefore, while \emph{something} comes
particularly \emph{into view}, be \emph{something else} deeper or more inward
which, with its determining influence, is lost or merges into the cognition or
the self-activity---in which respect the productivity, in another respect become
free, does not show itself as free. But a reflection can then afterward, in a
striving for deeper cognition, draw forth what was thus present hitherto but was
lost in the whole. ---Thus, for example, the principle of causality may at one
time underlie, as a determining fundamental thought, a striving to fathom certain
phenomena, without this principle being thought of; at another time reflection is
turned upon it; and the standpoint at which the cognizing striving stands in the
former case is quite other than in the latter.

3.~But why does the true, full essence of things not show itself at once to the
individual striving after cognition in such a way that all cognizing becomes a
pure, immediate beholding? It is natural that this question should arise here.
But its complete answer would require a general reflection on the ground of there
being a time and a temporality at all, and of life's coming forth at all in a
becoming [Vorden], under which it only gradually attains a definite
subsistence---something for whose full elucidation a fathoming of the innermost
meaning and essence of all human existence [Tilvær] would be needed. ---Only the
first basic features of such a consideration can find place here.

In the finite world the eternal and divine is to come forth, rule, and hold sway,
so as therein to reveal its infinite power and fullness: this is the meaning of
everything finite---that it is to be \emph{the Other} than the eternal, in which
the eternal is to come forth and reveal itself. But only in a consciousness and
spiritual life does the eternal and divine reveal itself \emph{as such}---in that
the being in which it lives comes to know of it, and moreover not only to bear it
within itself and be moved by it, but also itself to be able to behold and have
it as what it in its truth is. But the kingdom that is thus to come to the human
being is to be a \emph{kingdom of love}; for only in such a one does the true
fullness of the eternal come to show itself, and to show itself in its pure
spirituality, uniting a community of beings who are themselves spirits into a
whole that can now in truth be called the kingdom of God: the kingdom in which
God himself reveals himself in his divinity, in that he reveals himself in
eternal love. But a kingdom of spirit and love \emph{necessarily}
presupposes---and therefore necessarily calls forth---as the other in which it is
to be the ruling element, \emph{an intelligence and freedom}. If, on the part of
the beings to whom this kingdom is to come, a free and fully conscious devotion
did not come to take place---namely, a devotion of one's own mind, one's own
desire and impulse---then it would not be a kingdom of love that came forth. But
a personal life that is to be a free one---which now, out of pure love and
moreover with clear consciousness, is to give itself over to being enlivened and
filled by the divinity, and to move only in God---must itself make itself into
such a free life, in coming to clarity about the true meaning and essence of its
own existence, and likewise of the whole of existence. But then the opposition
must come forth fully, so that the reconciliation may become the true
reconciliation and unity in love and freedom. But if consciousness is thus to
come to self-consciousness and intelligence, telling itself what it is---if the
personality is to come to freedom under a self-determination in which it takes
itself as such and makes itself free---then it must come to a productivity in and
under which it objectifies itself for itself. It must come to express itself in a
sphere (that lower region mentioned above), which is its own, which it forms for
itself, and in which it becomes conscious of itself and of its freedom in a free
productivity. And thus life descends into a time and a temporality.

The personal being can, to be sure, no more make itself free (although we came to
use this expression) than it can give itself truth. It finds itself free as
regards possibility, but to have and enjoy freedom in deed (\textit{actu}) it
comes only through the liberating power of truth and love. But now it cannot be
liberated either, except \emph{in and through its own self-determination}---and
that a self-determination which it itself wills and feels as its own. That which
brings and gives the individual all reality, essentiality, and fullness can,
since it is a kingdom of love that is to form itself, give it the fullness only by
referring itself to \emph{something} that must now \emph{follow} in the individual
itself, as the individual's own self-determining: namely, its own self-surrender
and the acting that follows from it. The decision (that in which the whole working
concludes or ends) must always be the individual's own. Therefore, too---so that
we may still note this---the inner effects, in which the life that is to come to
reality operates or presses forward into consciousness, have the character of
incitement and summons: namely, a summons to seek what the individual can obtain
only when it desires and craves and seeks it, but which it can doubtless likewise
obtain only in that it is given to it outright. But of this again further on; see
§§ 19--21. In order that freedom may fully come to consciousness, and a complete
self-consciousness may set in---yet so that this self-consciousness also clearly
beholds what it is to strive to make itself into, so that now spirit and truth
may be had in spirit and truth---for this reason the latter opens itself to the
individual only gradually, under the individual's own striving, inquiry, and
labor.

\begin{center}\large\S.~5.\end{center}
\markboth{\S~5}{\S~5}
Insofar as the opposition between the object of cognition and the cognizing
personality comes forward in such a way that productivity, or representation,
shows itself as free---yet is now brought, by an incoming self-determination, to
present in one single way---cognition becomes a cognizing by force of grounds
[Grunde].

All cognizing is to be such a \emph{cognizing by force of grounds}.

For all cognizing is to come to complete clarity in and for the cognizing
personality and consciousness. But if it is to do so, then the productivity
required for it, and lying within it, must also come to clear consciousness; and
this it can do only in being carried out or undertaken by the individual with
self-consciousness and self-determination. Thereby it shows itself as free; but
if it is then, as that by which a cognition is to come about, bound to a single
determinate presenting, it must be bound to it in accordance with a certain
corresponding determining Something which, when it comes to consciousness as
such, shows itself as a ground.

Cognition now comes about as the result of an investigation and a deliberation.
Reflection and consideration call forth a \emph{study} that steps in between a
first immediate receiving and a final or concluding cognition. The mind, which
at first opened itself merely to receive and gave itself over purely to
contemplation and reception, is now set in motion in such a way that it is
itself, by its own self-conscious labor, to bring about the production in which
it is to possess truth; and it now seeks, by inquiry and grounding, a principle
or first mover for this production of its own. But the cognition that thereby
enters is nonetheless, when it now comes about, and as it now turns out, a new
receiving and beholding; for in the end we assume what, in accordance with the
determining grounds, we see that we must assume. This cognizing by force of
grounds is also to be a passage to a complete beholding, in which the full power
of truth works immediately and undividedly. But for the free personality, which
is to come to complete freedom and self-clarity, it shows itself as a necessary
passage.

\begin{center}Remark and Supplement.\end{center}
1.~The ground may, for that matter, well be merely a necessity accompanying pure
immediate apprehension and beholding, so that cognition takes place as one that
grounds itself; but the question of a ground has now been set in motion, and even
if what is assumed is assumed merely because it is immediately certain by itself,
this very immediate certainty [Vished] now shows itself as a ground, and
cognition now takes on a different character than it had before, when this
reflection had not yet entered.

This is especially to be remembered of the mathematical axioms. Sense-intuition,
too, no doubt presents itself as immediately grounding itself, and may also
remain with this stamp. But even then our representing is not seldom led to find
itself as a free reproducing---especially if the sensory shape is so manifold, or
the relations of its parts so intricate, that it can come to clarity only by
deliberation and consideration; and then what leads to fixing the representation
will also come, in opposition to the representation itself, to show itself as its
ground. This happens still more when it is noticed that things often cannot, or
should not, be represented as being just as the impression would immediately
suggest. But once such so-called sense-deceptions have shown themselves as
possible, doubt can now arise about the correctness of every immediate,
self-forming sense-representation, so that, even if it is directly assumed, this
now happens no longer merely on account of its own immediacy, but also on account
of a cognition that no so-called sense-deception is present---and the
consideration that teaches this now enters as a ground. Indeed, in the advance of
the endeavor and of cognition, the whole of sense-intuition finally shows itself
(to the philosophical eye) as one whose truth and validity can be doubted at all,
and must, moreover, first be warranted and established. And thus the representing
that at first formed itself entirely of itself, and seemed reliable enough in
itself, will show itself as still needing grounding---in general and in each and
every respect.

The intrinsically certain, from which this grounding is to proceed, or by which
all other cognition is to be warranted, is now sought on a wholly different side,
and cognition rises to a highest Idea, by force of which all else is now to be
beheld in its true and full light, in its meaning and truth. But even if this
highest Idea is, so far, immediate and resting in itself, we still come to it only
by a thinking that must, of course, throughout be guided and determined by
grounds---just as the very fact that all else can be clarified by that highest
Idea may be the ground on which its cognition rests. And thus we find, at the
highest point just as well, a cognizing by force of grounds, as we found the
lowest to be in need of grounding; and this because all cognizing comes to
actuality only through a being-determined to a certain representing, a
being-determined that, in the advance of the endeavor, must come to consciousness
as such. Everywhere the question is asked why we let ourselves be thus
determined, by what right this determining power is exercised over us.

2.~The perfect cognizing would be one that, independent of all deliberation and
grounding with all their unrest and striving movement, would rest wholly in
itself alone and be immediately clear and certain in itself, without any question
of its grounds arising, or any thought that something other than just what is
beheld could be the true. But such certainty and security---a cognition come to
rest and fullness in such a beholding---can rest only in the certainty of, indeed
in the pure immediate sense or feeling of, the fact that what is thus beheld
comes to us directly from the source of truth, is given and told to us by him
from whom all truth, like every good and perfect gift, comes. This is the
beholding that has been meant by the expression: to behold all things in God. It
is the cognizing by which we are to cognize even as we ourselves are
cognized---to know what we know because the spirit speaks it thus in us and gives
it to us. Then truth no longer stands over against us as the thing we seek;
rather, it dwells in us, in that the same that gives all things their
essentiality now also gives us cognition---pure, immediate, absolute, or in and
for itself, without its passing through any deliberation and reflection and
emerging as the result of such. A consciousness and assurance that everything we
cognize is given us by God---that our words are not ours, but, as we find them in
us, are worked in us by the divine spirit---can alone be that in which this
cognition can rest. Christ's cognizing we can conceive only in this way, never as
a cognizing by way of reflection and inference.

As a highest goal, as an ideal, such a cognition stands before us. Yet this
highest beholding cannot even be regarded as the goal that mere inquiry and
grounding, once it reached its completion, could attain. A completed thinking can
lead to a completed cognition by way of thinking---to a completed, wholly
certain, clear, and self-reposing knowledge according to, and by force of, the
highest grounds. But that highest beholding's pure immediacy can be reached only
in that the highest true---the eternal source of truth, the spirit---becomes
living in us and enlivens us in a wholly different (a differently real) way than
the way in which truth, as highest Idea, animates and enlivens a thinking and a
grounding.

For the forward-striving human being, however, \emph{the way of thinking} is the
one along which he is to advance, in order to press more and more toward the full
cognition of truth. It is a more earthly way, on which one attains piece by
piece. It belongs to the sphere in which, within a finitude, under its own
striving, a personality is to develop and be raised more and more to clarity and
independence---in a life in which it more and more objectifies itself for itself
and establishes itself in an existence of its own, yet only in order to ripen to
the point where the spirit can now, in its fullness, enter into it as into an
individuality now fully developed as its organ. Whereby then, when this point is
reached, fullness will enter in place of the piecework; and now thinking---which
surely is not to cease---and the fathoming of things---which is not to fall
away---will themselves be illuminated by the higher light, enter into the higher
peace, and, while an infinite fullness stands before them for contemplation and
fathoming, will nonetheless themselves walk the secure course that no longer has
constantly to ask itself about the grounds why it assumes what it beholds, why it
cognizes what it finds in its grounding, but has in this respect cast out ``every
witness of human neediness.''

Until it comes to that, all cognition is burdened with \emph{uncertainty}. The
immediately given, which is directly beheld, cannot satisfy us, because it is not
the true essence on whose cognition everything depends, and because we have of
this true essence a fundamental consciousness that drives us to seek something
other than what is directly beheld. It itself now points us to something else and
higher; but this else and higher we come to only precisely by being thus shown
and led to it---that is, in a mediated way---and this mediateness remains, so long
as our cognizing is a cognizing by way of thinking, always attached to it for us.

But what can here already lift us above this mediateness is the firm
\emph{conviction} [Overbeviisning], forming itself, of its reality. This springs
up in the course of our inquiry and grounding, as we are seized by the truth of
what is cognized, feel ourselves carried away by it. In the course of our
comparing and relating, deliberating and considering, it lifts itself more and
more forward and becomes---in the subjective, individual, and personal respect
(what concept and insight are in the objective respect)---the moving soul of all
thinking and determining, from which it proceeds and to which it finally returns.
And thus a \emph{faith} [Tro]---overcoming and dissolving all deficiency in
insight and in the developments and fixings of thinking---namely, a \emph{faith in
the reality of the fundamental cognition [Grunderkjendelse]}---takes for us the
place of that perfect cognition. As faith it is---though awakened and nourished by
a grounding, and constantly giving birth to a closer deliberation and grounding of
itself---nonetheless not a cognizing by, and by force of, the development of
grounds alone, but a more immediate appearance of the power of truth, which,
bearing witness to itself, seizes and irresistibly compels to cognition---yet not
because there is immediate beholding, but because everything so points and presses
us to this cognition that it shows itself as the center on which everything
depends and in which everything rests, and without which we cannot subsist or
manage at all. As our life in general, so too our life in cognition must seek its
last support and rest in an infinite trust and confidence, and, as the life of a
finite being, can in the end rest in nothing else. Yet that here too, as
everywhere, divine grace is in the end the one thing on which everything depends
and in which everything rests---this we shall come to set before ourselves more
closely in what follows. But because we receive only by grace, that too is why we
can have only by faith. Everything is in the end given, and must be taken as it is
given, in faith toward the Giver. But precisely that which makes cognition not
become a total beholding now also opens for this infinite trust the room in which
it can show itself as infinite trust---so that the very fact that we do not have
the fullness brings it about that attachment, devotion, and love come all the
more into view.

And that, then, is what all the life in us aims at, and moreover also all the life
that labors to reach a perfect cognition: that we, the more we exert ourselves to
have something of our own and to possess something in ourselves, should the more
come to feel that we can do nothing and have nothing except what is directly and
outright given us; and that, the more we rise in independence and intelligence and
command---as regards the manifold and its relative determination---the more we
should recognize that, in respect of the One thing that is needful and on which
everything depends, we must give ourselves entirely into God's keeping, who alone
works in us both to cognize and to carry out.

3.~But it is only of the fundamental cognition (that on which everything depends,
and which so far is the One thing needful) that this holds---that it rests, in the
end, in a faith and trust. Everything else, which is illuminated by this
fundamental cognition and is to be determined in accordance with it, must be
beheld as given with it---all in its connection therewith and its consequence
therefrom. So that, even if on the one side a faith and trust becomes that in
which all conviction rests and cannot but rest, cognition is nonetheless not on
that account to cease, on the other side, being a cognition by force of grounds,
and of an \emph{insight} arisen through the cognition of grounds. And just as
everything cognized in accordance with the fundamental cognition must be cognized
so, because we clearly see that the fundamental cognition necessarily entails it,
and can follow the whole connection from the single point under consideration up
to the highest, so too the fundamental cognition itself must precisely be clearly
seen as that which bears the whole connection, and to which this connection
necessarily leads. Under which the course will be this: that what was perhaps at
first cognized by a certain feeling of truth---indeed, by a relation to the
subjectively human---comes more and more to be evident to thinking in and for
itself and in an objective way, until at last a point will be reached where the
highest, to which all else points, is evident not merely because it is clearly
seen that all else points to it, but because it seems immediately \emph{evident in
itself alone}. But this consideration---in which we meet the profound thought
underlying the famous so-called ontological proof of God's existence---leads us
over to our next principal proposition. ---God bears witness to himself in
consciousnesses, and no cognition that God is would be possible if God did not
thus bear witness to himself within us, that is, in our gradually forming
spiritual life, beholding, and contemplation; indeed, this fundamental
consciousness of God (and of one's own innermost communion with God) is what
first makes the human being human. Herein, then, is immediate faith. But should
not the same cognition also be able to become just as immediately evident for
speculation, so that this evidence too, as evidence, rested only in itself? By
which, however---since this evidence too would then have to be given to us, in
that it was evident precisely by itself---faith would not cease to be that in
which all cognition would in the end rest. Insight and self-evidence, or evidence,
are themselves \emph{worked} in us as well; but where there is question of a life
working itself forward in consciousness in a real way, and what comes forth is
seen as this life's inner operations, there is grace and governance---and there,
moreover, faith, attachment, and trust must show themselves. The purely
intelligent is itself something that likewise comes only through real influences
from above, and must therefore be received as a gift---and that with trust toward
the Giver. In the cognizing that we conceive as Christ's cognizing, an immediate
self-evidence becomes one with the assurance that trust gives; it is faith and
beholding in one.

\begin{center}\large\S.~6.\end{center}
\markboth{\S~6}{\S~6}
Insofar as inquiry [Granskning] aims at a cognizing by force of grounds, it
strives not merely to cognize the true, but also to cognize it as the true and in
accordance with its truth. All cognizing is to be such a cognition of truth in
\emph{its truth}. Only then is it genuine and complete. It is the endeavor to
cognize thus that sets the whole search after grounds in motion; it is what
really expresses itself in that search.

In the first immediate apprehension and beholding, we receive what presents
itself to us immediately and let that suffice. Of its validity there is as yet no
question; even if it is the true that has therein shown itself to us, it has still
not stepped forth as the true; it is not yet that for us. (Which is why Plato,
too, in the \textit{Theaetetus}, seeks a quite different definition of cognition
than that it should be correct representation.) But if truth is to come to us
fully and take shape in us, it must give itself to us by itself, in that it
becomes evident to us by its truth; only then do we behold and have it as it is in
itself. Only thereby does true cognizing set in---which is not merely a taking-up
and beholding, but a taking-up with assurance. But this already happens, though
not fully and in the highest degree, insofar as cognition becomes a cognizing by
force of sufficient grounds, and as far as this reaches; for so far and so long we
assume what we assume because, in accordance with these grounds, it is evident to
us as the only thing assumable on the presupposition of these grounds---hence as
that which alone can be cognized as the true.

But if cognition is to become the complete cognizing of truth in its truth, in
which truth shines by its own light, then at every point the grounds from which it
proceeds, and on which its validity is so far conditioned, must be pursued up to
the \emph{highest and last grounds}---so that what is cognized is cognized by
force of these, which, regarded from another side, may also be called the
\emph{first grounds}.

These, which then cannot in turn be traced back to higher grounds, must now be
evident immediately by their own truth.

\begin{center}Remark and Supplement.\end{center}
1.~The free self-determination and productivity, with its own deliberation and
consideration---which was called forth in the individual personality and came to
meet the reception, testing it---was thus set in motion only so that truth might
be more perfectly cognized and taken up. By this the cognizing personality is to
rise to a higher level of cognition, and freedom is raised up so as to belong all
the more fully to truth. So that truth may shine for us in its own light and
thereby fully take shape in us, we enter into constant testing; we demand
everywhere ``a full account and correctness,'' as the phrase goes---demand grounds
and proofs in science, even where, for example in certain of the simplest
mathematical propositions, the certainty is otherwise unshakable enough and no
doubt arises or can arise. It is not enough that the true be given us; it is to be
given us by the complete self-evidence of its truth alone. From no other hand will
we receive it.

If, then, we see a higher \emph{tendency toward independence} come into motion,
and the personality assert itself as one that now, as far as possible, will rely
only on itself and its own scrutiny in all its cognition, we nonetheless see that
this happens only so that the power of the true may become truly complete in us.
Even though we would test everything ourselves---indeed, it may even take on the
appearance that we were calling the true itself to account, as though we ourselves
would decide what is to be acceptable to us and would let truth hold only when we
ourselves had confirmed it---still it is not ourselves or our own that we seek
therein, but only that truth itself, in its truth, may fully come to us. If we
seek confirmation, it is that the true must confirm itself before our eyes as
well. If we call to account, then what we call to account is only forms and
representations that offer themselves to us but that we will not at once accept on
trust---and we do so in order to let truth itself give us what we are to accept. It
is here, where there is as yet only question of \textit{rationes cognoscendi},
that we call ourselves to account. If we tear ourselves loose from every other
authority by which we were perhaps formerly led, it is in order all the more to
\emph{subordinate ourselves to the highest authority}---that is, truth's own. Just
as it is said elsewhere that only he shall find his life who knows rightly how to
lose his life for the Lord's sake, so we may say here that we seek ourselves, or
seek to win ourselves, precisely only in order to be able all the more completely
to give ourselves up and lose ourselves.

2.~Here we also see the point at which genuine scientific \emph{study}, which
first truly begins with \emph{academic study}, comes to the fore, and where it
becomes evident what it signifies. Precisely at the age at which, bodily and
organically, a higher drive toward independence or self-subsistence stirs, and at
which the transition from the child's life to the matured and completed human
existence is to take place, the drive toward spiritual self-subsistence awakens
and works, in the scientific respect too, in fresh, vigorous movements. For the
student it is the stage in life at which he is to begin to strive to cognize truth
in its truth. Therefore he is now to go to the sources from which his teachers
themselves drew, and to know everything by force of the grounds and considerations
by which the inquirers before him---from whom he learned, and naturally continues
to learn---came to their cognitions. It is the stage of scientific coming-of-age
[Fuldmyndighed] that he is now to attain. He is to raise himself to the height of
his age in his science, in order from there---if the aptitude for it is given
him---to carry it further, whether by a theoretical or a practical path. Therefore
he is now not merely to learn, receiving what is handed down to him as the result
of inquiries, but \emph{independent thinking} [Selvtænkning] is to set in---inquiry
and grounding of his own: he is to study. In school, mere \emph{learning}---though
a thoughtful learning, to be sure---was the main thing. To satisfy his teachers,
and only so far also himself, is what the pupil is to and will do. Now, by
contrast, what chiefly matters is to satisfy the immediately felt demands of the
feeling for truth. The endeavor is now everywhere to press toward the \emph{last
and highest grounds}; toward these the whole direction of study is to be aimed;
and even if inquiry cannot yet, from a given point and within a certain sphere,
reach with definiteness and clarity so far as to see the present thing in definite
connection with the last grounds of all things---so far as to follow the
connection of objects and their chain as far as possible---still these grounds are
always to be kept in view as the goal, however distant, toward which one is to
strive, and a feeling for the importance and essentiality of one day reaching them
is to be preserved and kept alive. This is what characterizes the academic
citizen, and the man of science or scholar formed by academic study, as such. This
is also why philosophical study---as a fathoming of the last grounds of all
things, and a scientific endeavor called forth and moved by such a fathoming, with
the distinctive stamp that such a fathoming must give it, a study first begun at
this stage of life---now becomes here, too, of such decisive essentiality. ---It
is not, however, required, for a study to be called genuinely scientific, that its
tendency be to cognize the given in its connection with the highest grounds of all
things, or that its aim be strictly philosophical. Even if the inquirer keeps
within certain limits and traces connection and relation only within a certain
sphere, inquiry can still be genuinely scientific, provided only he aims to go,
within this sphere, as far as he can---both by penetrating here as far into truth
as possible and letting it shine as far as possible in its own light, and, partly
to win it, partly to test what has been found and corroborate it if it seems to
withstand testing, by extending his re-examination to the scrutiny of all the data
that lie within the once-fixed sphere. The sense for all the more distant
considerations to which one may have to go back must also be kept open and awake,
together with a readiness to enter into them, so far as the present work will
thereby not be interrupted but rather advanced. A constant general regard for them
must show that, in the lowest, nearest, and most particular, a contribution to the
highest, most distant, and most general is nonetheless intended. ---A definite
\emph{limitation in the endeavor} may, moreover, be led to not only by personal
considerations and grounds---insofar as every individual must more or less limit
himself in order to accomplish anything with force and effect---but also demanded
by objective grounds, insofar as every object ought to be investigated not only in
its connection with the whole, and its significance thereby determined, but also in
its particular appearance on its own, how it shows itself in and for itself;
whereby it is then first regarded as a member of a certain smaller whole, which is
investigated on its own, before this whole, and the object along with it, are
considered in their larger relations. The philosophical spirit shows itself here
in the direction the study takes toward the utmost and last that offers itself
within this sphere---moreover, in the working toward the point where a higher
consideration can now take hold.

3.~How, again, the endeavor considered here---to cognize truth in its own truth
and to let it be given by this alone---leads to the thought at work in the
ontological proof of God's existence will perhaps be easily noticed; but to enter
into all the considerations thereby called forth is not the place here.

\begin{center}\large\S.~7.\end{center}
\markboth{\S~7}{\S~7}
Insofar as a cognition of truth in its truth is attained, there arises an
\emph{intelligent cognizing}, or a cognizing with insight and by force of a
discernment. This cognizing---in which what is cognized is evident by its
agreement with what understanding [Forstand] or reason [Fornuft] demands, or at
least allows or admits, and is now assumed in accordance with a necessity of
understanding or reason to cognize it---now stands in opposition to the cognizing
of the given and the found, in which the latter is assumed merely on account of
its bare givenness or foundness alone: a cognizing that as such has an empirical
character.

Even a ground can be cognized as a ground, and a consequence as a consequence,
only in that the former is seen as that which necessarily leads to its
consequence, and the latter as that to which its ground necessarily leads, so
that the consequence shows itself as one that, on the presupposition of the
ground, cannot be thought otherwise---so that the necessity of assuming it shows
itself as a necessity for thinking. What is assumed by force of grounds is always
assumed in accordance with something the understanding can test and has tested.
But just as the nexus between ground and consequence---the consequence's being
grounded in the ground---can be beheld only by an intelligent beholding, so too,
when the true is not merely to be received but to be received as the true and
cognized because it shows itself in its truth, mere foundness cannot be that on
which cognition can rest; rather, it must rest on the true's self-evidence. The
true is seen in its truth only when it is seen as that to which we cannot avoid
ascribing validity---when, that is, our thinking is so bound by it that we cannot
think it otherwise; but this intelligent necessity enters only when an agreement
with an understanding's own demand or expectation shows itself.

A genuine and complete cognizing is therefore an \emph{intelligent cognizing}.
Only by such a cognizing is truth cognized in its truth. The empirical
apprehending and intuiting must therefore, if it is to become a genuine
cognizing, be so embraced by the intelligent endeavor that everything found as the
actual is now also beheld in its---at least relative---conformity to understanding
or reason. Only then does a genuine science come about. Whereas the knowledge we
reach by merely empirical cognizing (so called only in a looser sense)---even if
it leads to a doubtless needful and important, ordered and complete survey of the
empirically given objects, and to a clear and exact presentation of their
appearances or shapes and found qualities---still offers only an aggregate of
phenomena, into which thought must first bring light and clarity by explaining
them: that is, by showing their emergence and existence as one that the
understanding itself, even if only under certain quite special, however narrow,
conditions, would have to demand, or at least expect. Whereby they then take on
the stamp of essentiality and are lifted above the contingency with which they
showed themselves, so long as they were as yet taken up only as merely found.

In intelligent cognizing, assurance is produced by a quite different and higher
necessity than the empirical necessity to assume, which determinate affections
bring with them. Although the apprehending individual must, in accordance with
these, represent as it does represent---so that its productivity is so far found
entirely bound and determined---it can nonetheless, so long as it is not discerned
that the found is necessarily as it is, think it, that is, represent it,
otherwise; and productivity is thus so far not completely bound. Empirical
apprehension merely receives; but the consideration that raises itself above it
can now also let it stand at its own worth. In the inquiry after the true in its
truth, by contrast, thought is called forth, which will not merely receive
plainly, but discern. But when it now, moving according to its own fundamental
nature and its fundamental concepts, nonetheless arrives at what presents itself
in the object itself, it becomes---precisely because it moved in its freedom and
yet arrived only at this same one thing---bound with a higher power; so that
complete boundness and masteredness enters by means of intelligent cognizing.
Precisely because free testing and consideration has been called forth and
asserts itself, what is cognized in accordance with it now comes to rule with the
most complete power. Truth cannot shine in its truth, and be evident by this its
truth itself, without seizing and carrying away with a higher necessity.

So far as intelligent cognizing goes and comes forth clearly as such, it now
expresses itself in a \emph{construction a priori}, under which what is
cognized---being derived from the fundamental thought to which intelligent
beholding has pressed forward---is presented in its necessity for the
understanding. The thought that brought light into the cognized whole is now set
forth as the first mover of cognition, by which everything is determined as it
entails it must be.

\begin{center}Remark and Supplement.\end{center}
We here follow cognition as it ascends, through a series of levels, to the
highest. Only when we have made our way to this can all the foregoing be seen in
its true meaning and its full light. But we will nonetheless, right here on the
way, enter into the questions we gradually meet, and consider what they lead us
to.

That an intelligence opens itself up in cognition, and that cognition everywhere
strives toward \emph{intelligence-character} [Intelligenthed], is soon seen once
the endeavor after complete cognition comes into motion and becomes clear to
itself. But why is it so? \emph{Why must all genuine cognizing be intelligent?}

Further: \emph{how can the nature and constitution of an object given from
without be cognized by a cognition whose living fundamental norms, whose
determining first mover and determining manner, even at every point, are of inner
origin}---the very own of the understanding and reason speaking in us?

1.~That this last---we will take it into consideration first---is so, is clear.
Every example of a cognition that proceeds from a deliberation, or in which
connection, relation, ground, and consequence come into consideration, can serve
to show it. Already when, in the simplest lower sensation, the representation is
fixed in a certain determinate way---even entirely in agreement with the immediate
momentary appearance, indeed even when it is only a matter of fixing this, and
this happens with consciousness, so that a deliberation arises, or at least a
question of whether a deliberation is needed or not---something intelligent will
at once show itself, even if no difference between a semblance and the real is
assumed but the possibility of a sense-deception is denied, insofar as something
is thereby always determined in accordance with an inference whose validity and
consequence---in accordance with grounds whose grounding power---can be seen only
with the eye of the understanding. It is, in general, only a cognition by way of
thinking that determines the reality of sense-intuition---which holds both when,
within the sphere of the senses, it is asked whether and how far something is
semblance or reality, and when, occasioned by considerations of a quite different
kind, the question arises whether sense-intuition as a whole has any reality at
all; so far, we receive the whole of sense-cognition itself, as regards the
significance it acquires for us, from the hand of intelligent cognition.

The intelligent and the merely empirical cognizing thus already interlock in the
formation of lower sense-intuition, and the former extends all the way through
the cognition that, as regards its main character---namely, as regards what is
assumed---still rests only on a simple apprehending or receiving of what is given
in experience in its givenness. The same is seen, too, when now, by means of an
intellectual intuition, the ground-forms discovered in objects are cognized.
Although these too must be taken as they are found, intelligent cognizing enters
here as well: not only in that, when the connection is not manifest, we must be
led by grounds to determine the representation of it in a certain determinate way;
we can also, in general, see a ground-form as a ground-form only by discerning how,
in accordance with it, the outer shape or appearance whose ground-form it is
actually comes about. The intelligent comes forward still more when a Why is
answered---when, that is, the explanation of phenomena is found; for a phenomenon
is explained only when one goes back to certain fundamental determinations and
fundamental relations, and the phenomena are now seen as necessary consequences of
these---consequences that the understanding itself derives and cognizes as
consequences thereof, because it is discerned that from such fundamental
determinations only such phenomena, and no others, can be thought as consequences;
as when the well-known laws of gravity, which might perhaps have been found in a
purely empirical way through the well-known experiments, are discerned to be a
necessary consequence of the nature and constitution of such an uninterruptedly
attracting force as gravity is, which cannot be thought to act according to other
laws.

Even if, in doing so, one often proceeds from first determinations that are merely
apprehended in a purely empirical way and assumed on account of their foundness,
without having yet been traced back to such first determinations as are evident to
the understanding itself---indeed, even if the agreement with what the
understanding itself entails, which shows itself when something is made
comprehensible, is not seldom highly relative and conditioned (as when even the
contrary-to-reason element in human actions is made comprehensible by being seen
in its consequence from certain presuppositions, of which the understanding itself
must recognize them to be a natural consequence: that is, a consequence whose
following, or way of following, is conformable to understanding)---still the
explanation and comprehension, and the cognition proceeding from it, is, so far as
it goes as such, always entirely intelligent; within the given sphere of
conditions, what is assumed now no longer stands as merely found and contingent.
Finding ourselves in the midst of the empirical, we seek to come to light and
clarity by starting from what is nearest and searching out its connection with
what is nearest; and we can now also behold the members and parts most closely
linked to one another in this their connection---indeed, behold them in their
significance and essentiality for one another, and in respect of certain first
members---without having yet pursued this discernment and comprehension to its
utmost limits. Thus a certain animal form, or a certain mineral, that one takes
from the series as an example, can be seen as an integral member---indeed, as an
essentially belonging member---in the whole series; indeed, the whole system of
animal forms or of species of minerals can be projected in an a priori way, in
that the whole course of development of these forms is presented as though it were
nature's course of ideas in the production of these forms, even if the
determinations from which one thereby proceeds are still taken merely as those
abstracted from experience itself by a broad survey of nature. But science
naturally strives everywhere to widen the limits within which intelligent
cognizing has for a time remained standing, and thus to overcome this relativity
of its. As far as the true, genuine cognition of the object, joined with insight,
goes, so far too goes an intelligent cognizing; and where the latter gives out, so
does the former.

Herein the understanding, in making its consequences, constructions, and
deductions, everywhere proceeds from its \emph{own} fundamental norms and
principles. A logic, mathematics, metaphysics (metaphysical fundamental concepts
and principles) not received or abstracted from experience, but springing in an
independent way from thinking, is hereby everywhere laid at the foundation. And
its determining influence extends into the smallest detail, whose particular
constitution, in being explained and comprehended, is precisely also derived and
cognized as that which, under the given conditions, is dictated by the
understanding just so and not otherwise, and is grasped and comprehended only in
that it, precisely in this its particular constitution, shows itself as that to
which that logic, mathematics, metaphysics---and the construction guided by their
fundamental determinations, or the execution and development of a certain
intuition---necessarily lead.

And yet this cognition, although its determining fundamental determinations are of
inner a priori origin, comes forward with a claim to objective validity, and
cannot do otherwise. What comes forward as that which the understanding
necessarily entails comes forward necessarily with this demand for reality. It is
utterly unthinkable to us that it should not have it. Could a real conflict arise
between experience and the necessary demands of the understanding, there could be
no relation at all between the thinking being and reality. Yet such a conflict we
could never find or establish. If something comes out as the result of a thinking
or calculation that does not agree with experience, we always seek the ground of
this disagreement in the fact that either our presuppositions were incorrect or an
error has crept into our inferences. It can never occur to us that nature's
consequences might perhaps be other than those of thinking---that nature's logic,
mathematics, or metaphysical principles could be other than those of the
understanding and reason that work and speak in us.

But how can such a cognition, springing from the understanding alone, be
everywhere validly laid at the foundation in experience, and a whole empirical
cognizing then built upon it?

Manifestly only because the intelligence that shows itself as such in the
contemplating human being runs through the whole of reality and underlies it;
because \emph{the intelligence itself} is precisely an expression of nature's
\emph{last laws}. And how should something other be, in the human being---as one
member of this reality---the first and innermost mover, than what is such in the
whole of reality in general? So that here one is not to think of a double
understanding, as though our understanding met with, and agreed with, another
understanding, nature's. It is the very same intelligence that, in its undivided
unity, reveals itself in nature and in our thinking and contemplating, presenting
itself, self-identical, here and there. What ties the connection and agreement
between us, or our thinking, and reality, is here, as everywhere, something third
which, higher than both the connected terms, equally underlies them both. To
this, too---so far as the imperfect and piecemeal can bear witness to the perfect
and total, in that it points to it as the source of its own possibility---to
this, we say, it bears witness, that in the endeavor, guided by independent
thinking, to penetrate into nature, the further it advances, the more a light
really dawns for us over nature. What the understanding speaks, entails, demands
is also given to us just as fully as what we reach in a wholly empirical way, even
if it is given to us from a quite different side, in a quite different manner, and
with a different stamp than the empirical; so that in the intelligent we have to do
no less with something objective, and no more with ourselves or our own, than in
the empirical. If here at every point there is an intelligent deriving and
determining, cognition is nonetheless at every point a receiving and beholding,
since we derive and determine, in conformity with the understanding, as we see
that we must.

2.~But in the human being the intelligence comes forward as such, or in \emph{its
intelligence-character}. Therefore the selfsame is here no longer the selfsame.
Therefore that \emph{a priori} enters here, which now shows itself to us in its
real meaning and essentiality. If the intelligence is to come forward in its
intelligence-character, if the rational orderliness [Forstandighed] that is in
nature is to reveal itself in the human being as it is---and if in nature it is
that which \emph{underlies}---then in the human being too it must show itself as
the underlying, indeed show itself pure in and for itself in its original
independence, as the \emph{constitutive and determining}, and moreover as the
\emph{a priori} in opposition to the manifold presenting itself in reality, and as
the principle for an independent construction. And thus we see how the nature and
constitution of an object given from without not only can, but also, if it is to
be cognized fully, must be cognized by a cognizing whose determining principle and
determining manner are of inner origin---the very own of understanding and reason.

It now also becomes clear why, although it is the selfsame intelligence that comes
forward in us and in nature, nonetheless---and precisely because it is an
intelligence and in us comes forward in its intelligence-character---something
inner, or presenting itself from within (subjective), comes forth in opposition to
what presents itself from without (objective): both insofar as, in general,
original demands proceeding from thinking itself show themselves, with which the
given must be found to agree if it is to be comprehended, and insofar as, in the
particular too, every point is cognized fully only when it is seen as that to which
an a priori derivation or deduction could also lead. In the a priori construction
both the inner and the outer show themselves in their greatest opposition---how
they stand over against one another---but also, when the given now shows itself
precisely as the necessary result of this construction, in their fullest agreement
into unity: an agreement without which no genuine cognizing comes about. ---Insofar
as this a priori construction aims at setting a given reality in its full, true
light, it shows itself as a merely reproducing construction, whereby we, as witness
and assurance that we have fully penetrated into the cognition of the given in its
truth, now by force of this cognition carry it out for ourselves in an image,
which must then agree entirely with what an exact observation lets us find and
behold in the given reality. But insofar as the fundamental cognition attained
constitutes the moving, all-determining First in this construction, the
construction comes forward as an independent one with the a priori character.

3.~But just as an exact empirical observation must furnish the test of this
construction's correctness and validity, so too only such an observation---an exact
and careful study of the given reality itself, that is, of the rational or
intelligent in it---can lead to that construction. Thought will penetrate the
found; but for this it is required that the rational orderliness which reveals
itself therein be apprehended---that is, understood---as it is, and wholly and
entirely as it is. Here, therefore, it always takes on the appearance that our
understanding met with an understanding revealing itself in existence. Only
insofar as we find sense and understanding in nature and reality can we grasp and
understand it---that is, cognize it fully; but this we do find therein, and
therefore we can cognize it thus. The study of nature is like the study of a book,
into whose meaning and whole spirit we seek to penetrate through its letter, but
can also penetrate only insofar as we find meaning and spirit therein. But how do
we penetrate into it? Only in that what we find expressed therein becomes evident
to ourselves---and in such a way that we see why it has expressed itself as we
find it expressed, and that we now ourselves agree with it therein. If an
excellent work presenting deep truth is to be understood by us fully, the light
that shone for its author must dawn for us, and we must come into possession of the
moving first principle from which all his determinations necessarily proceeded, so
that we now, with him, take truth at first hand. But precisely so it stands with
the study of nature; what in nature itself determines the significance of things
must be grasped by us in such a way that a light dawns for us over nature from it,
and what nature expresses, too, we must---penetrating through its letter to its
spirit---seek to take at first hand. From which it is sufficiently clear that the
aim cannot be, by an a priori constructing that tears itself loose from experience
and proceeds entirely on its own, to make empirical observation and re-examination
superfluous; but rather, precisely through the latter---as through an exact study
of the words and sentences in nature's book---to penetrate to its meaning and
spirit.

The relation between \emph{empiricism and rationalism}---an opposition that,
especially in certain sciences, for example in medicine, has stirred up so much
strife and struggle---must now surely be clear from this. Namely, that---just as,
with respect to another distinction, we may say: only reason makes one properly
understanding, but only the understanding, too, makes one properly
reasonable---so too, surely, only a careful and exact study of the empirically
given can lead to a completely rational cognizing, yet only a genuinely rational
endeavor leads to a solid empiricism. If it is empiricism or experience to know
that this or that applied means generally has these or those effects, without
discerning how and why precisely such an effect comes to be the consequence of
such a means, then empiricism or experience is no true cognizing. It is then to be
classed with the knowledge that is merely an aggregate of certain results and
learned propositions, without one's being in possession of the grounds and the
thinking that led to these results---without one's seeing what lies in the learned
propositions and is really asserted in them.

4.~By means of free thinking there takes place a higher and more inward
appropriation of the sphere of experience that offers itself to be cognized. The
apprehending personality, which thereby, as it were, assimilates it and takes it
up into inward unity with its own (rational) life, thereby becomes, as it were, at
home in it and master over it, so that it moves and goes about in it as in its own
property, in which nothing is any longer foreign to it. But in and by the same
self-appropriation this personality becomes more and more taken into possession by
the cognized truth, so that the personal life is absorbed into this truth's own
self-presentation, and more and more stirs and moves in truth and according to
truth.

\begin{center}\large\S.~8.\end{center}
\markboth{\S~8}{\S~8}
The endeavor to cognize truth in its truth leads---since it turns out that only
that cognizing in which the given is beheld in its agreement with the demands of
the understanding is a genuine cognizing, and since the question now everywhere
becomes whether the found must also be assumed necessarily to be as it is---to
transferring the search after grounds to the objects themselves, and to asking
why these themselves are as they are found to be. If formerly grounds were sought
that were to determine our assuming---grounds of cognition (\textit{rationes
cognoscendi})---now an inquiry of a quite different kind arises, as the
\emph{grounds of existence} of things (\textit{rationes essendi}) are sought out.

Only with this does intelligent cognizing set in. With the most complete insight
into the necessity of assuming what is assumed, this very thing assumed can
nonetheless still stand before us as something merely found and contingent, of
whose essentiality there is as yet only a question. Not this thing assumed
itself, but only its assumption---whose grounds of existence the grounds of
cognition now show themselves to be---is embraced by intelligent cognizing. This
is always a cognizing of a \emph{real necessity}. But if this real necessity is
not merely to be evident, but to be evident in its truth and fully, then it must
not merely be discerned, but discerned by force of the cognition of the real
grounds.

Only then, when we behold things in their own real necessity, can we not refuse
to ascribe essentiality and validity to them; and then we also see how far and in
what relations it is to be ascribed to them---namely, so far as this necessity
shows itself, and in the relations in which it shows itself. Only then,
therefore, do we behold things in their truth. But in a real actuality, where
things stand as members and parts of the actual whole, the real necessity---the
necessity in and for the things themselves---must show itself as that in
accordance with which these things actually come, or are present, and maintain
their place. And thus their necessity must show itself as a necessity springing
from certain grounds of existence, which grounds of existence must therefore be
investigated. If cognition is to correspond completely to what is cognized---if
the productivity by which, through the completion of the representation, cognition
itself first comes about is to present what is cognized entirely as it is---then
what is first and determining for the object must also stand as such in the
cognition and representation. If things are what they are in accordance with a
Something thus determining them, which as their real ground constitutes them, then
cognition would be imperfect if they were not also beheld as those which are what
they are precisely in accordance with this Something, or if this Something were
not also, together with the things themselves, cognized as that by which they are
so constituted.

\begin{center}Remark and Supplement.\end{center}
1.~The cognition of the \textit{rationes essendi} of things now also lets us---only
in that we see these to determine the modes of existence of things---behold them
in their more or less \emph{relative validity}, or in their greater or lesser
essentiality. They, as it were, warrant their existence and their descent to be
granted a certain significance, as their real grounds show themselves; but now it
is also seen that this significance and essentiality of theirs is in general only
a conditioned one. Further than that on which their own reality rests, this
reality itself cannot go. And just as it is only the thinking consideration for
which the question of reality arises at all, and it is an intelligent fundamental
concept that is thereby set in motion, so too it is the intelligence that
determines the difference both between semblance and reality and between more and
less valid reality, and now accordingly assigns everything given its place in
this respect. There is nothing that does not have its reality; but this is seen
only when it is seen how---indeed, how far---the thing cannot but exist, that is,
when it is traced back to its real ground. Even semblance has its reality---namely,
as semblance, as the apparent---and has a greater or lesser reality, the more or
less such a semblance shows itself as necessary, the more extended are the
relations that must be surmounted in order to escape it; but it is seen as such a
semblance, and now also as a semblance necessary so far, only when it is discerned
why and how it necessarily arises in accordance with that which, as its ground,
brings it with it.

2.~One easily sees that it is something quite different to investigate what we are
to assume takes place---and to set forth the grounds of cognition that must
determine us to it---than it is to investigate the real grounds why it is as we
find it to be. The one investigation and insight often lies in a quite different
sphere than the other. By the first, the phenomenon, or the constitution and
relations, whose Why is to be answered by the second, is first fixed. But just as
it is important first to know what must be assumed as given, before its
\textit{rationes essendi} are sought out, so the cognition of the latter often in
turn casts such a light on the object of that first consideration that the
\emph{assumption's} own certainty and reliability is corroborated and completed by
\emph{insight into the real grounds}. If we see that a certain member or point in
a given whole, in accordance with the connection cognized or assumed in this
whole, cannot but be thus and so, and if what this conformity to understanding
demands agrees with what observations and inferences from observation, or other
grounds of cognition, lead us to assume about this member or particular point,
then confidence in the correctness of the latter cannot but gain thereby---indeed,
rise to irrefutable certainty.

3.~Proofs aim at the presentation of sufficient grounds of cognition, and are
carried out by \emph{inferences}. The inference is, in general, the form for this
kind of cognizing. By its form it leads, in general, only to seeing the necessity
of assuming this or that. This is evident not only in the many cases in which the
consequences or effects of an object's fundamental form, or fundamental
constitution and fundamental manifestations, become that from which we infer the
object itself, and from which we thus derive the proof of the truth of the
representation formed about it---as when it is established, from the well-known
grounds, that the Earth is spheroidal and that it turns on its axis and goes
around the Sun, whereby the question why it has such a figure and motion is as yet
not at all answered, indeed not even touched, since it lies in a quite different
sphere of investigation. But also where one does not thus infer back, from the
constitution of certain phenomena, to a certain factual fundamental state or
fundamental form, or a certain fundamental fact, but where a train of thought,
striding straight forward from what is first in itself to what follows from it,
leads to a series of cognitions by way of strict inferences, and to a whole
scientific doctrinal edifice---even in an entirely intelligent way---the kind of
cognition may still be such that it merely shows, in accordance with valid grounds
of cognition, how the objects are constituted and what their relations are,
without the grounds of existence for this being fathomed and brought forth.

\emph{Mathematics} is, at least according to the usual manner of presentation, a
great and telling example of this. The whole conduct of proof here gives, in
general, only grounds of cognition, and no others are sought or demanded. The
necessity with which and in accordance with which cognition takes place is indeed
intelligent, and it concerns---which must not be overlooked---the objects
themselves, the mathematical forms or intuitions themselves; it is really
discerned that these objects themselves necessarily have the cognized qualities
and relations, so that the necessity is not a merely outer one, determining us,
but an inner necessity in and for the objects (the triangles, the squares, and so
forth) themselves. But that whose intelligent necessity is discerned is
nonetheless always only that qualities and relations are as they are; as for the
grounds, only why we must \emph{cognize} that it is so is shown---not why it is so
\emph{in itself}. In the auxiliary lines that are so often constructed---in
general, in the accessory constructions that are so often made for the sake of the
proof, and in other detours that are used---the actual grounds of existence of the
qualities and relations considered are surely not to be sought. Mathematics in
general aims only at the determination of forms and relations, not at their inner
essence and connection with a higher and highest being. Mathematical
\emph{demonstration} is not yet a philosophical \emph{deduction}. The proofs by
which cognition takes place, along with the way in which cognition takes place by
force of them, are, with respect to what is cognized itself in itself, nothing
truly essential. The proof gives a more certain cognizing and leads to a more
coherent one, but gives no higher and more essential cognizing. The source of the
relations and qualities is not exhibited; they are not derived from their origin
or descent; (their warrant is not a true legal ground, but rather a kind of
\textit{causa in facto posita}). Cognition therefore also takes place piecemeal,
not in an all-embracing connection in which everything would be beheld in mutual
essential connection. This science has as little the form of a genuine, complete
science as it is itself a genuine and complete knowledge from the ground up. That
the mathematical method ought not to be adopted in philosophy---the strongest
ground for this (to note it in passing) is that this method is by no means perfect
enough for it, by no means truly thorough (going to the ground) and genuinely
scientific enough. To take up the philosophical content requires a quite
differently systematic and organic form, with a quite differently
closely-connecting method than at least the Euclidean, and any other that proceeds
in his spirit---however much its rigor and exactness are otherwise to be praised.
As little as it would be fitting---despite all recognition of the beauty and
significance of so many crystallization-forms---to force a tree into such a form,
just as little is it fitting, even if one fully recognizes what is in itself
finished and elegant in a genuine mathematical method, to want to force
philosophical endeavor, which strives after a freer, more vital, and more organic
form, to enter into the mathematical, when it is to express itself in definite
presentation. In the organic manner of appearance toward which it strives, the
last must bear the first just as much as the first grounds the last, since the
whole cognition rests in that which, as ruling soul, runs through the
whole---from which it follows that no single determinate proposition can be
completely evident in itself, without any regard to the following ones and to the
whole scientific aggregate of propositions; whereas in mathematics, which proceeds
more in straight lines than in a cycle, each earlier proposition is certain enough
without needing the subsequent one, and the evidence of the first propositions
does not rest in the execution of the whole system---which, by contrast, is the
case in philosophy, as will become clear in the following (for here we anticipate
it). It can even be a question whether what mathematics, according to its usual
procedure, gives us can be called a scientific system in the proper sense; at
least, the many propositions, so often evident independently of and without regard
to one another, are not seen in any such essential connection with one another
that they are traced back to a unity and show themselves as those which together
form a self-rounded whole.

[Mathematics as a whole is, from the standpoint of philosophy, to be regarded in
general---like the empirical---as a great phenomenon (like a certain particular
world of the understanding), whose possibility and last grounds become a task for
philosophy. A doctrine of mere forms and relations, then, cannot yet be a complete
science either, and cannot therefore be withdrawn from philosophical endeavor, as
though it were sufficiently complete in itself and enough for itself. The
genuinely scientific mathematician will also strive to bring a philosophical
spirit to move and stir in this sphere of cognition. But if we ask how
philosophical endeavor can gain room in this science, and in particular how a
complete cognition of real grounds, and by force of the cognition of real grounds,
can be attained here too, that is surely not so easy to say: no science is at
present further from what could be called its philosophy than mathematics. Two
principal points, however, can be indicated in this respect. We have
already---this is the \emph{first}---noted that, although the proofs give only
grounds of cognition, the proved qualities and relations are nonetheless cognized
as such as cannot be otherwise in their objects, so that their real necessity is
discerned. And this is discerned because it is discerned that in every possible
case the same could be proved under such presuppositions, since the object's (e.g.
the triangle's) fundamental form will always allow it; but in that this
fundamental form of the object shows itself as the constant real ground of
possibility of the same manner of proof, it is seen that it is really the ground
of existence of the proved qualities and relations. It therefore depends only on
this: that this actual ground of existence too be seen as such, and that what is
to be proved be seen as \emph{its} consequence---and that as its direct
consequence, not one moreover resting on certain auxiliary constructions---or
rather, that it be seen as belonging to this fundamental form itself and to its
essential expression. Such a manner of cognition and proof was the aim of the
sketch of a geometry according to a genetic method which Ørsted, a good many years
ago, began to work out and communicate to several persons, and which he---as we
wish and hope---will surely one day find occasion and inclination to bring to
light. The objection that such an attempt will probably meet soonest and
first---namely, that the proofs in this way could not attain Euclidean rigor, but
that, when figures are now to be considered in their arising through the movements
of lines, even the \textit{crux mathematicorum} which in the higher mathematics
gives the mathematician striving everywhere after rigor so much to contend
with---the concepts of the infinitely small and infinitely large---might now gain
admittance into the lower mathematics as well: this objection will mean so little
in the philosopher's eyes that what is feared will for him be precisely the
gratifying thing about it; for that precisely where the so-called irrational in
mathematics presents itself, the rational comes into view, is a remark that is
now, and has already long been, no longer new. (See in this connection the both
deeply and sharply seeing Hegel's \textit{Logik}, first volume, the Remark, p.
206.) But even if in this way the fundamental norm of a mathematical construction,
which is at the same time the fundamental form in the mathematical intuition
thereby produced, is everywhere brought forth so that it is everywhere seen as the
ground of the qualities and relations cognized to take place in the constructed
intuition, and further, a certain necessary course develops from it in the
sequence of the propositions after one another---there still remains the question
of the real grounds of existence of the various constructions themselves (of the
triangles and squares and other forms themselves) and their real significance in
the whole of cognition. Real concepts (that is, proper and true concepts) of them
mathematics does not yet give. Should such not, however, also be found? But where
and how are we to seek them? In order to see finite forms in their real
significance and essentiality, we must---which is \emph{the second} thing we would
point to---see them as integral members in real actuality; only as members and
parts of this, or as forms that each in their place necessarily arise in this
actuality, are they seen in such a connection with the whole that this can now
also be pursued far enough that they are seen in their connection with the last
grounds of all things. Thus they will already be regarded, in the first place,
when they are seen as constructions to which the real intuiting and productive
activity by which they are produced must necessarily come in its necessary course
of development. The order and sequence in which the propositions are to be
presented, and the light in which they are thereby to be set, will then especially
become an object of consideration; but then something philosophical will thereby
already show itself in mathematics, which will now also be able to become an
object of debate, as well as of dialectical consideration---something that
mathematics, in accordance with its apodictic but not philosophical manner of
cognition, otherwise does not know. But more important than this merely subjective
consideration will be to seek them out in \emph{nature}, and, when one finds
mathematical forms there---whether as fixed shapes, for example in the crystals,
or as forms of motion, for example in the course of the planets---to investigate
why in nature precisely \emph{such forms arise at such points}. The mathematical
forms are found here \emph{as given objects in reality}, hence in a certain
necessary connection with the whole of given nature, and now the endeavor to
penetrate everywhere in nature to the last grounds cannot but embrace them as
well. Precisely because in pure mathematics they were not regarded as given, no
question arose about their \textit{rationes essendi}. In such a philosophizing over
the mathematical forms it would now probably also turn out that the real
significance of a mathematical construction can be different in different respects;
from which a certain uncertainty as to its correct determination, or at any rate a
difference as to the course and manner of its determination, could result, and the
philosophical difference in mode of intuition, and the strife following from it,
might again come forth in this sphere of cognition otherwise freed from such
contentions. But this would nonetheless bear witness that the cognition had become
organic.]

3.~Although the grounds by force of which cognition takes place in mathematics are
merely grounds for the assumption, still---since that to whose cognition these
grounds lead shows itself as something that, in the constructed forms or
intuitions, indeed in the objects themselves, is as it is in accordance with an
objective intelligent necessity---we have in mathematics an example of a purely
intelligent, and indeed an absolutely intelligent, cognizing. Nothing here is
taken as merely found, but everything is found and cognized as something the
understanding demands. The understanding finds itself here entirely in its own
sphere, in which it is completely at home; therefore the certainty here is a
complete self-evidence.

Philosophy too strives after such a certainty, but its far higher nature and aim
make it here far more difficult to attain. While mathematics aims at investigating
the qualities and relations of objects that are products of its own
constructions---constructions of whose essential necessity there is not even any
question, since their mathematical possibility is here enough for their reality,
so that mathematics has therefore only to find out what is an essential
consequence of certain determinations thus possible---philosophy, by contrast, has
not only to find out and determine what in and for itself is the true with respect
to objects (physical or moral) that are actually given, and which must be taken
with the determinations with which they are found, but it even aims at---indeed,
is to start from---that which is the absolutely true in general, which underlies
all the conditions and determinations with which any objects whatsoever can show
themselves. Since here all things show themselves in infinitely many connections
and relations, and these relations precisely determine things' own modes of
existence or forms of appearance---since all things in general have only a
relative being and essentiality---there is here an infinite series to live
through, an infinite aggregate of objects to behold in their connection, before
absolute certainty can be attained with respect to any of them. Many points now
soon shine forth as so evident in themselves that no doubt seems possible about
their certainty; but yet precisely that from which complete certainty should
spring can become what shakes their certainty, because they, when an accounting is
now demanded of their validity, do not seem capable of being established with a
self-evidence against which a skepticism could not, after all, arise. Here, then,
everything finally depends on whether \emph{a highest, absolutely fixed point} can
be found, to which everything can be so linked that by it its more or less
relative validity can be determined with firmness. This highest point will now
soon show itself to be not merely the highest, but also the center around which
everything gathers into unity and validity, and by whose light every point is to
be illuminated, so that it is beheld as the all-valid that runs through
everything. (See further on, § 10.) But now the question again arises wherein the
certainty of this pure Absolute's reality and truth rests. And then the matter
takes on, on the one side, the appearance that this certainty could rest only on
the fact that everything \emph{else} pointed precisely thus \emph{to that}, as the
first and all-conditioning---that, as surely as so much else, which could now be
gone through in its principal points, was true and undeniable, so too that must be
true and undeniable. But on the other side it is clear that everything else can
point thus to it only on its own presupposition, and that it must have shown
itself in its reality in order for us to see everything else point to it. But then
it must be cognized as that whose reality must not merely rest in itself alone, but
whose reality must also show \emph{itself by itself alone}.

And then we see how natural is the thought already taken into consideration
earlier, which underlies the ontological proof of God's existence---a thought to
which an endeavor to cognize truth in its truth, and by its own truth, could not
but lead, and to which all the reflections in which we here find ourselves point
at almost every point. Just as faith in God necessarily takes on---as indeed it
already has from the outset---the character of a firm faith resting in itself, so
the intelligence too strives to bring this cognition to an equally immediate
self-evidence resting in itself. But this is not the place to carry out these
considerations.

\begin{center}\large\S.~9.\end{center}
\markboth{\S~9}{\S~9}
The inquiry after the grounds of existence of things aims most immediately at
finding out their connection and relations, and the real significations thereby
determined, from which their existence and constitution are comprehended and
explained. But such a cognition is a cognition of \emph{a concept} and a cognition
\emph{by force of a concept}. The proper, true concepts of things
(\textit{sensu eminenti})---namely, their \emph{real concepts}---now show
themselves as those in which the true grounds of things are most immediately to be
sought.

For to ask after the ground of a thing is to ask after something \emph{other} than
the thing itself, from which it follows, and with which, therefore, its existence
is given as necessary. An inquiry after the ground of an object therefore goes
beyond the thing---unless the thing is found to have its ground entirely in
itself, or to be itself the Other in question. Another object is then sought, of
which the given one can be regarded as a consequence. But then (not to mention
that this other object, if it is of the same kind and sphere as the first, will
show itself as just as contingent, just as far from being enough in itself, as the
first) what most immediately brings the object with it---indeed, its nearest
ground---is really the connection, the nexus, between the one and the other, which
brings it about that the former can and must carry the latter with it. Which holds
equally, whether the one follows as effect from the other as cause, or several
objects in conjunction show themselves as those that essentially go together with
one another. Always, when any other object whatever is adduced as the ground of one
in question, it will and must be asked why the former has the latter as
consequence---which is a question after a certain mode of connection between them,
in accordance with which the one carries the other with it. The connection then
shows itself as what most immediately grounds; the conformity to understanding
itself---so far as it is beheld as the conformity to understanding and necessity of
understanding that, in the thing itself, with the one requires the other---is here
most immediately the actual real ground; only by discerning this is the thing
comprehended and explained.

But connection is cognized fully only when it is seen as that by means of which an
aggregate of parts or other particulars is joined into full unity and wholeness.
Such a connection is the expression of a concept, and a cognition of a thing's
essentiality and significance in accordance with such a connection is both a
cognition of a concept and a cognition by force of a concept. A thing's concept,
as the fundamental thought determining the mode of connection, thus shows itself
as the ground in accordance with which the individual members belonging to the
thing are as they are. Because the thing is this or that---that is, corresponds to
this or that concept---it has the qualities and parts it has, and these stand in
the relations in which they stand. But the thing is, in turn, the thing it is, and
corresponds to its concept, because as member and part it essentially belongs to
the larger real whole in which it is found, and whose concept carries it with it.
Its own concept here shows itself as a single thought necessarily following from
the concept of the larger whole; whereby the concept that before stood as ground
with respect to its individual members is now itself grounded, or seen in its
essential following from a conformity to understanding.

The cognition of a great connection embracing all given objects, in accordance
with which they would all be seen as essentially needful members in an
all-embracing great whole---so that it would be clear how a necessity of
understanding entailed or demanded them all, with all their qualities---is
inquiry's ultimate aim in this direction of its.

\begin{center}Remark and Supplement.\end{center}
1.~Where phenomena show themselves as facts---but in nature everything can
everywhere be regarded either as facts or as results of facts---inquiry aims at the
\emph{laws} according to which effects are produced, and to fathom the grounds of
things becomes, so far, to fathom their laws (though naturally as such as are
discerned to be necessary in accordance with an intelligent necessity); but these
laws are then, as the fundamental forms of activities, a kind of concept.

2.~We see here how the Platonic Socrates (in the \textit{Phaedo}, pp. 97c and 99e)
could feel himself driven to reach for Anaxagoras's νοῦς, and, in order to come to
a true cognition of the grounds of things, to take his refuge εἰς τοὺς λόγους.
This cognition is indeed not yet the highest; it itself points to a higher one;
but this higher cognition, to which it leads, is not to displace it, but to
corroborate it---indeed, it will first show it in its fullest validity. Meanwhile,
we can and will, right here where it enters, establish ourselves entirely in the
cognition that the conformity to understanding itself is that in which the sought
reality rests---that the concepts of things are their proper, nearest grounds of
existence.

The general ground for this cognition is that everywhere the actual Because can
most immediately not be an object, but the nexus that brings it about that this
object has the first as its consequence. If, for example, heat is adduced as the
ground of a body's being expanded, cold as the ground of water's solid state when
it has become ice, these objects are still not thereby cognized as grounds of these
phenomena; rather, this cognition---that the adduced object really is the ground of
the phenomenon in question---arises only when it is discerned how heat expands, cold
solidifies, and why the one cannot but expand, the other solidify. But if this is to
be discerned, the whole nature of heat and of cold must be investigated, together
with the whole aggregate of effects that, in accordance with this nature, it is
discerned as unable not to have. But the question then becomes what heat, what cold,
really are. And this question is a question after the concept of heat and of cold,
from which the laws of their effects are then discerned. For this it will now be
required to investigate what significance heat and cold have in the whole of nature,
and also under what conditions they come forth. But when it is now discerned that
heat, for example, must come forth under these or those conditions, this can still
only really be discerned when it is discerned how it follows from heat's whole
concept, from the real significance it has in the whole, that under such conditions
it must set in---whereby heat's own inner nature shows itself as the properly causal
element (αἴτιον, \textit{Phaedo} 99b), which under the given conditions comes forth
because its nature entails that it must then come forth. But heat's nature is its
distinctive constitution, or manner of being, determined by its real significance;
and moreover the cognition of its nature is one with the cognition of its concept.
And so it stands not only when the ground of a certain kind of phenomena in general
is inquiry's aim, but also when in a particular individual case the ground of a
certain fact is sought; there too it can be found only by entering into the
investigation of a certain connection, but then the insight into the ground will
rest on the insight that an object such as the one in question, in accordance with
the nature of such an object in general---hence in accordance with its concept---had
to step into reality under such circumstances. Let it be well noted that a ground is
not cognized and discerned as the ground of a fact merely because the thing is
assumed and cognized as that on which such a fact always follows, and according to a
definite rule, but only when it is at the same time discerned that this fact must
follow upon it because it necessarily carries such a fact with it; and this cannot be
discerned without discerning the connection, or nexus, that leads from the one to
the other---so that this nexus is that on which everything depends. And this holds
not only with respect to our cognizing, but just as fully with respect to the thing
itself; for what is it, too, in the thing itself, that brings it about that after, or
with, the one thing the other always and necessarily follows, other than precisely
the real connection between them, which entails that with the one the other must set
in? Once it is discerned as entirely necessary that from a certain object another
follows---but without this insight the cognition of groundedness is not
complete---this too must be cognized as something that is so in the object itself
because in it it could not be otherwise; and the necessity of understanding then
shows itself as that which, in the object itself too, brings it about that the
ground has the consequence it has, so that the conformity to understanding shows
itself as what properly grounds. When---to go back to an example that occurred
earlier (§ 7, Remark 1)---the laws of gravity are explained and discerned by a
mathematical speculation, what else can be named as their real ground in nature than
precisely this: that a force like gravity cannot be thought to act, and therefore
cannot actually act either, in any other way.

What has here been elucidated about the kind of ground called cause is even more
evident when several things, found together at once without temporal succession,
show themselves as those which carry, the one the other, with them. That the ground
of a certain animal form's existence must be sought in the whole developmental
course of the animal series---indeed, in the connection within the whole series that
requires such a form as an integral member in the whole series---is easy to see;
likewise, that the heart has the form it has because the whole organization---hence
the fundamental form of the organism, or its concept---demands such an organ among
the rest; and so forth.

3.~But if the grounds of things are thus sought in their concepts, will not the
inquiry after grounds end, in the end, at something quite different from what it
sought---come out at a place that seemed least of all to be the one where it should
come out? The question after the ground of something requires something \emph{other}
than the thing itself, of which other the thing is to be a consequence, or with
which other it is at the same time to be given; and it is manifestly no answer to
such a question when the very thing whose ground is asked for is given back, under
another shape, as the ground. But does this not happen quite manifestly when, with
Plato (the \textit{Phaedo}, at the passage cited), seeking the grounds of things in
their concepts, one says that the ground of, for example, two things' being two is
not the adding by which the one is joined to the other, or the dividing by which one
thing is separated into two parts, but is nothing other than twoness; or when one
would say, for example, that the cause of a human being's dying is not the poison or
the disease or the like, but is death; or that the ground of its freezing is not the
north wind, which is said to have brought such cold with it, but is nothing other
than frost? Is such not the manifest tautology?

And yet it is clear that---to keep to the examples---it is not the poison alone
that, without more, makes the human being die; not the north wind alone that,
without more, makes the water frozen. Upon the introduction of the poison into the
body, there comes forth, by means of the relation between it and the organic
life-process, such a state that now life itself, in its corresponding activity,
causes death---that is, that under a concurrence of these conditions (partly the
outer, the poison, partly the inner, the living activity working in accordance with
its nature) there sets in the state that is now precisely death itself, which sets
in because under these conditions precisely it, and nothing else, must set in,
because the nature of such a state, of such a fact, is that of such a one---and, as
it were, adds this new thought, and demands its realization, to the preceding ones.

And thus, as regards first \emph{facts and states}, their real grounds are to be
sought both in the things themselves in which they are found---or these things' own
nature, that is, in their concepts---and outside them, in what is usually named as
their grounds and causes. Everything sets in, and has its thus-and-so constituted
subsistence, only under certain determinate relations, which are therefore
conditions for it; and when they are all present, it too must necessarily come to be
present, and thus these conditions too take on the character of grounds. But when it
now sets in and comes to be present, it comes forth with its distinctive
constitution, as something new that comes to the rest and whose place can be taken by
none of these; it comes forth thus because it is itself as it is, and with this its
distinctive nature and constitution belongs at this place; it sets in, therefore, in
accordance with the distinctive thought that requires such a Something at such a
place---that is, in accordance with its own concept. The point at which the thought
of such conditions coincides with the thought of such a fact---so that it is
discerned that the latter, on the presupposition of the former, must necessarily be
thought in this determinate way---is the point where the actual ground lies; but
this is precisely the thing's own concept. In the special case, everything depends
on seeing how a certain general thought, a certain concept, under these given
circumstances necessarily came to reality; how, for example, under these conditions
frost or death, in accordance with the concept of frost or death, came to actuality;
only then are frost or death seen as the cause of the water's freezing or the human
being's dying.

And if we then, secondly, ask how these very thoughts that demand the setting-in of
such facts---\emph{these very concepts} in which we seek the grounds of these
facts---come forth, we see that they are, in accordance with a necessary train of
thought, derived or deduced from certain preceding concepts, which necessarily also
lead to these in a whole system of thought. A certain principal thought demands,
when it is to be realized, this and that determinate thing, which in accordance with
it must be; but in this there are now many particulars, each of which must be
constituted thus and so, to which, therefore, just as many concepts correspond,
which dictate not only why they must necessarily be thought along with it, but also
how they must, on that account, be. As the thing, the fact, or the factual state, is
explained or seen in its groundedness---because it is discerned that under these
relations a necessary train of thought necessarily requires precisely the thing's
existence in this way---there is now also seen, at the same time, the thought
determining it to this, the thing's concept, in its necessary emergence from such a
train of thought. Meanwhile it nonetheless comes forth as a particular thought, a
particular concept, on its own, which in the whole train of thought is indeed
demanded thus by the pressing necessity of the rest, yet is a new thought on its own
and a member subsisting in itself just as fully as all the rest, with which together
it forms this whole of thought. So that the concept itself rests equally in itself
and in the whole connection in which it comes forth and from which it is derived. The
thought shows itself as what determines the thing, and therefore the ground of the
thing's existence and constitution must be sought in its concept; but if one asks
after the ground of the concept itself, this is to be sought in its deduction, in the
necessity of understanding that, in a certain necessary train of thought, leads to
such a concept---so that not precisely the fundamental concept from which one
proceeds, but the connection of thought and necessity of thought as a whole, which
leads from the fundamental concept to the concept in question, is to be named as the
latter's ground.

4.~The question of grounds here still everywhere comes forward only as \emph{a Why}.
This leads most immediately from phenomenon to phenomenon, or from a given object to
another given in the same way. But because this other cannot, without more, be
adduced as the answer, but is only just so much: it is not always easy to determine
what, in a fact, is the actual fact. But once it is found out, it only now becomes
the object of a grounding---in order now to find out its \textit{ratio essendi}. The
question why the planets go around the Sun, and so forth---or the inquiry into an
explanation of the whole planetary system, regarded as a great found
phenomenon---already leads over into a quite different sphere. It can now lead to
seeing this fundamental fact in its essential connection with others, and thus the
concept of a larger connection may open itself up, whereby a \emph{more deeply lying
fundamental fact} would come into view, as a task for a new grounding; but the
explanation may also perhaps lead back to \emph{intelligent determinations} that are
evident by their own self-evidence, and from which that fact is seen as a necessary
consequence---whereby, however, a certain mode of connection always comes into view,
as the foundation of the whole explanation. In general, this course of explanation
leads all questions of grounds back to certain fundamental questions, on whose
answering everything will finally depend, but with whose answering light will also
finally be brought into the whole. Partly, several coordinate \emph{isolated facts}
will gradually more and more be seen as \emph{the same, or alike}, so that one and
the same fundamental fact is recognized again in them all; as when in combustion, in
the rusting of metals, and so on, one finds again the same process of oxidation.
Partly, several coordinate \emph{facts interlocking with one another}, or states or
fundamental forms connected in a real way, will be able to be beheld as consequences
of a \emph{real connection} embracing them all, so that only this finally becomes
the remaining problem; as when all the activities and manifestations of the mind
show themselves in so essential a connection with one another that finally only the
existence of such a self-conscious and self-determining personality in the midst of
a finitude becomes the problem that remains; or when all individual animals seem so
to be essential members in a whole self-developing series that now the fathoming
endeavor turns to this whole formation expressed in such a series, in general, with
its fundamental determinations, in order to ask why it exists at all. But thus,
finally, the necessity of assuming still something \emph{quite other} than \emph{the
whole chain} with all its members---something in which the whole, as it were, hangs,
or by which it is in general borne and held---will press itself upon us more and
more; which leads over to the next consideration.

\begin{center}\large\S.~10.\end{center}
\markboth{\S~10}{\S~10}
The necessity of understanding, or conformity to understanding, that demands the
existence of things and determines their qualities---the concepts of things, on
whose thorough cognition the cognition of things themselves depends, and which in
things themselves show themselves as that in accordance with which they exist and
are as they are, hence as a foundation for their subsistence and whole
distinctiveness---may indeed be posited as the \emph{nearer} \textit{rationes
essendi} of things; but in these, taken in and for themselves alone, their
\emph{first and highest ground} cannot yet be sought. Even if in this intelligent
grounding one were to proceed from determinations---and from something one must
proceed---that (like the first ones of mathematics) were self-evident and had to be
assumed in accordance with an immediate intelligent necessity, still only an
assumption would thereby be grounded, not the existence of the corresponding things
in a reality; and the whole real existence would so far still stand as contingent.
Intelligence and concept can indeed \emph{entail} that such and such objects are
there; it can indeed \emph{demand} them; but it does not set them into existence or
make them actual. The whole total connection, from which things' essentiality and
the necessity of their existence follow and become evident, will still---so long as
it alone shows itself as cognition's last foundation or first origin---stand as a
great problem, of which it must equally be asked why it, in general and in its
wholeness, is as it is, as it was asked how the actual whole, in which it is found
realized, comes to be actually present. We must therefore still seek the last
grounds of existence in something other than the whole rational orderliness---in a
Third that connects the two, the given reality on the one side and the rational
orderliness on the other, and that everywhere brings the concept into the
immediately actual (into objectivity), and brings it about that the former is
expressed in the latter. Which Third must then be just as embracing as the whole
connection---must show itself, just as fully as this, as One and all.

This, then, cannot be merely a rational orderliness, but must at the same time be a
power [Kraftighed]: an \emph{active First}, which shows itself as a principle
\emph{by which} things are as they are, so that everywhere, in its working, it is
present in them, and they are only its outer expression and effects. A principle
that not only \emph{determines} (in an ideal way), but realizes, or \emph{makes
actual} (in a real way). Such a principle grounds and gives essentiality by itself,
working, being present in things and, as it were, filling them with
essentiality---quite as the soul is everywhere present in every part of the
organism. It is the true \emph{being} and essence, the true real or substantial;
and only when it is beheld in them, or they are beheld as belonging to its full
manifestation, are they themselves beheld so far down to the ground that they are
thereby illuminated, as by an inner light. It is beheld only as it should be beheld
when it is seen as \emph{a life} that enlivens, or at least makes actual, all the
things grounded by it; and all these things, which now show themselves as a mere
substrate or material for its life and working, are seen only as they should be seen
when they are seen as enlivened or made actual by it---so that these, in their whole
aggregate, are seen only as its expression, and it itself is seen as that which
dwells in them and shines toward us, or speaks to us, out of them.

In place of that Why there shows itself here a \emph{By-which}, and in such a way
that that by which things are shows itself as that which, as the inner essence, is
\emph{present} in these things. But that \emph{Why} does not thereby fall away. It
can and must still constantly be asked: why does that first Being and Active One
work precisely in \emph{this} determinate way, why does it step forth just into
\emph{these} effects? A question that, when one starts from its consideration as
fundamental consideration alone, and does not look toward an already given reality,
will run: in \emph{which} effects and forms of expression will that true being come
forth? In general, it cannot be thought as the ground of existence of any individual
object except in accordance with a certain all-embracing connection, a certain
conformity to plan, a definite \emph{order of things}, according to which precisely
these or those definite things are brought forth, as it carries them with it. But
this whole mode of connection, or order in things, must itself be grounded in that
active First, that true real. But if it is asked: why precisely this order and no
other? or, better asked: which order of things, then, must that First set into
existence? then no other answer can be given than the same that must be given when
the question is: why is there anything else at all besides that true, essential one
alone? why does it set anything else into existence? And to this the answer can only
be: in order therein completely \emph{to reveal itself}; hence in order that this
other may be (or become) its complete expression, so that it can itself completely
fill it and be in it. That to which we must look in order to find the \emph{first
origin} and determining ground of all things is also that to which we must point when
it is asked: to \emph{what end} everything exists. That \emph{by which} it is, is
also that \emph{for which}, or for whose sake, it is. But then that mode of
connection, or order of things, must also be such as this requires it to be, in order
fully to reveal its essentiality. We see, then, how the inner, all-grounding being
entails and demands, while at the same time it works and makes actual. While it is
itself One and all, as that which is everywhere present in all as the sole thing that
in truth is, it nonetheless at the same time demands this all as something
\emph{other} than itself, in which it is now to be as the all-filling, and thus
itself also stands in relation to this other, as though it were merely a member in
the chain---albeit the first moving member. Which becomes especially evident when the
whole connection is now to be discerned and cognized as that which is expressed by it
in reality, and in accordance with which every individual thing is determined and
brought forth; for then one must proceed from certain first determinations, as those
which that all-real being first demands, to which, therefore, cognition is first
led---and since in these a connection must at once show itself, the all-real being
itself stands as the very first in the series or chain of this connection, hence as a
beginning first, as a member. Its significance with respect to the other, its
relation to it, must now be determined, and indeed a \emph{concept} be formed of it.
But although it thus stands as the first determining member, it is nonetheless not
merely the first, but also the all-moving, which, animating and determining, runs
through the whole, and everywhere shows itself therein at every point, so that the
whole that now comes forth is, in its wholeness, to become its total expression.
---This true Being, Absolute and Eternal, so far as it is beheld in one with the form
under which it comes forth---or the concept through which it comes to its real
appearance---is what we call \emph{Idea}. In the Idea, then, all cognizing finds its
last support and rest, in that it finds therein its absolutely first point of
departure.

\begin{center}Remark and Supplement.\end{center}
1.~We have only a few strokes to add in order to carry the train of thought up to
the very highest point---to the point where the highest Idea comes. If things'
first and only true \textit{ratio essendi} is to be sought in an all-real, single
Being and principle, but this is such as at the same time entails and posits an
infinite, coherent whole and an infinite rational orderliness or conformity to
plan in this whole, then it shows itself not merely as a being and a life, but as
an all-real, existing and living being, which posits an infinity in order to reveal
its infinity therein, yet which in itself is independent of what it thus
posits---determining it, but not determined by it.

But why, then, does it posit or bring forth this whole, when it subsists in itself
without it? \emph{Why is it not merely, but also reveals itself?} We answer: so that
it may come forth in infinite \emph{fullness}. Hence, after all, so that it may
still more fully \emph{be}. There must be something in it itself that requires its
revelation, so that it cannot and will not rest in itself alone. We cannot conceive
it as stepping out and revealing itself without conceiving it as though it
\emph{had to} reveal itself and could not do otherwise; and yet there can exist no
compelling thing, no commanding necessity, for it: if it must reveal itself, it is
because it \emph{must will} to reveal itself. But how is a Must to be thought
together with complete freedom---how is that which does not at all \emph{need}
anything else outside itself for its subsistence nonetheless to be thought as that
which \emph{requires} something else outside itself, as though it were, after all,
not enough for itself? It can be thought only as a necessity and a demand of
\emph{love}. For love has this in itself: that it cannot be without the other,
without the object of its love, and that it nonetheless requires this object for no
other reason than solely in order to love---which here is not in order to receive,
but only in order to give. That which, besides the highest being---which now, as a
loving being, must necessarily be thought as highest \emph{personality}---must
further exist, is \emph{its kingdom}, the kingdom of God, as a kingdom of holiness
and love, and moreover of beauty and fullness. But therefore a \emph{freedom} too
must be, and a \emph{kingdom of freedom}; and in order that this may be, freedom
must itself make itself into such---whereby we come to the consideration already
called forth earlier (§ 4, Remark 3). (Let it only be further noted how the Idea of
the Son of God---as the first and at the same time all-embracing object of eternal
self-revelation and love---here presents itself quite directly; but this
consideration lies too far from the present investigations for us to enter into it
here.)

2.~The highest aim in this direction of inquiry, and for all striving after
cognition in general, now becomes to behold everything so in relation to God that
this highest Idea---just as it is \emph{practically} to be the
\emph{all-animating}---now also \emph{scientifically} becomes the
\emph{all-illuminating}, as we have already noted earlier, in passing, in Remark 3
to § 1. But a philosophical cognizing (a cognition by force of Idea) can, at many
points, already fully set in as such, without the object's yet being beheld in the
highest light. What makes a cognizing philosophical is, in general, not what is
cognized alone, but the way in which it is cognized: namely, that it is cognized by
force of a fundamental cognition of something beheld as the productive
principle---but thereby also as the sole real, or existing and essential, that gives
all reality---and with respect to that, whatever it may be, which is to be cognized
by force of it.

\begin{center}\large\S.~11.\end{center}
\markboth{\S~11}{\S~11}
What determines for all things their more or less relative essentiality, and with
respect to which alone they are and mean anything, must naturally also be that which
gives cognition itself its essentiality and makes it what it is---so that \emph{the
last ground of existence of cognition} itself is likewise to be sought therein.
Cognition too, then, is a form under which the Eternal reveals itself or realizes
itself, and the complete cognition of truth is, in its essence, the very \emph{life
and self-presentation of the Idea in consciousness}. But if it is to be a complete
cognizing, it must also show itself to consciousness and be had and possessed by the
cognizing personality as what it thus is, so that it is cognized as being the Idea's
own life in us---to which we should therefore, with willing mind, give ourselves
over, to be its pure vessels or organs.

\begin{center}Remark and Supplement.\end{center}
1.~The point at which this form for the Eternal's revelation---belonging to the
highest forms---enters and is beheld in its essential connection with, and its
significance within, the whole (which whole exists in general only to serve the
Eternal, so that the Eternal may come to a kingdom and dominion therein), is evident
from the foregoing, where we have seen that a kingdom of freedom must come forth in
order that a kingdom of love may come, but that the former in turn necessarily leads
to an intelligence presenting itself as such---that is, coming to consciousness.

2.~What the \emph{truth} is whose votary the inquirer is to be now becomes entirely
clear: namely, that here there is something quite other than a mere agreement between
our representing and its objects, higher or lower. The truth, or the true, is that
which \emph{livingly produces this agreement}, in that it, itself working, determines
the whole manner and formation of the representation. Precisely what gives things,
forms, and relations themselves reality, or essentiality and truth, now shows itself
to us (and has already earlier announced itself to us, in the Remark to § 7) as the
very same that also, living again in us and in our conscious cognizing, gives the
latter its reality. Truth itself, which now shows itself also to be, like virtue for
example, the expression of an Idea, is to be the enlivening principle in our
cognizing, just as virtue is to be it in our acting; and just as the latter is it
under one form, so truth is, under another form, the Infinite and Eternal
living---indeed, present---in us, whose organs we are to become by completely bringing
its inner declaration to consciousness, clarity, and a powerful dominion in us, in
inner utterances and productions. So that we too, in relation to nature and existence
itself, are to take---and to take with clear and fully conscious cognition---everything
that we have and possess in our cognizing at first hand. If, then, a reproduced image
of the whole is to live again in us, and we, precisely thereby, to grow completely
into, and together with, this whole, then in this we are nonetheless to have more than
a mere image; we are therein at the same time to have the mighty influence and
presence of the Eternal itself, which in that actual whole too is the true real and
existing.

It goes without saying---though we will not omit to note it expressly---that what is
here said must not be interpreted as though the meaning could be that, for example,
the principle of gravity, or the effect manifesting itself in nature under the form
of gravity, should now as such be assumed to be that which called forth our concept
of it or our cognition of it. Whereas indeed the same One that entails---or
demandingly and effectively posits---a gravity in the system of nature now also posits
the concept of it in our system of understanding.

3.~But what the (essential and genuine) true---what the very power of the Idea---thus
works in us is not, for that reason, at once cognized by force of the origin from
which it in itself really comes. Because a cognition in itself proceeds from a life of
the Idea in us, cognition does not for that reason yet take place by force of the
cognition of the Idea; the objects are not for that reason yet seen in the Idea's
light, come to the clarity of consciousness. Thus mathematics, for example, is
indeed---as a science, and as the thought calling and bringing this science to
consciousness---the expression of an Idea and a life of the Idea in us. But
mathematical cognizing is nonetheless not a cognizing by force of Idea.

Even mere lower sensation, and wholly empirical apprehension and cognition, is to be
regarded in the same way: as that which in itself comes about through the Infinite's
own life and working in us---though it is so far from coming to consciousness as an
expression of a higher life in us that it does not even seem to be worked from within,
which, however, as an expression of our life, as a manifestation of life and a result
of a life-activity, it cannot but be, as it already shows itself to be when the
productivity by which it comes about is noticed. But the actual mover of this
productivity, its determining First, can likewise now only be a life, or a
proper---indeed inner---principle. And now all considerations lead to seeing a
universal working---equally present in the outer as in the inner, and in both the one
and the other all-determining---as the last and true ground of all lower sensation.
See the author's \textit{Psychologie}, § 58, p. 212.

\begin{center}\large\S.~12.\end{center}
\markboth{\S~12}{\S~12}
If the Idea is to come to live as such in the individual by presenting itself as
Idea in his cognizing, it must come forward as the governing and determining First,
as the forming and ordering principle, in an aggregate of cognitions that come to
constitute a coherent whole of cognition, in which the Idea---as it calls forth and
determiningly develops it---presents itself. This presentation of the Idea by and in
accordance with its own self-presentation---in which the Idea itself (as the more
inward) is had and beheld in one with its (more outward) presentation---is science,
genuine, complete science. The individual cognitions are \emph{scientific} insofar as
they are had and beheld as essentially belonging members in such a whole, and now
therefore proceed from the inquiry and discernment by which such a whole is gradually
sought to be brought about.

In science, then, not only \emph{the One}---which is everywhere to be regarded as the
essence, to which all is to be traced back and on which all depends---is cognized, but
also the \emph{manifold} thereby determined, in its infinite diversity; yet in such a
way that everything in it is beheld both in its nearer or more distant relation to
that One---indeed, in its greater or lesser reality---and in its definite relation to
all the rest, with which together it makes up the full presentation of the Idea, and
moreover at its definite place in the coherent whole. So that in complete science
these three---\emph{Idea}, \emph{system}, and \emph{detail}---are inseparable.

To such science \emph{all} cognition is to be developed; \emph{every} cognition is to
be a scientific cognition in this way.

For---this is the first point---if the Idea is to be beheld as Idea in its truth, it
must necessarily be beheld as that which gives essentiality; indeed, it must be beheld
in one with that to which it gives this essentiality, or beheld as the determining and
all-illuminating element in that other, which has its reality by it; otherwise the
Idea would not be beheld in its full truth, for it belongs to its essence to give and
determine essentiality. Next: if the whole aggregate of individual objects that
present themselves to us, and at whose investigation our striving aims, is to be
cognized in its truth---or cognized fully, as it is---and if it is now something only
with respect to the Eternal, and has significance only through its relation to this,
then it must also be beheld in its relations to this, and in this relativity of its;
but if all else is thus to be determined and illuminated with respect to that Eternal
and by the Idea of this Eternal, then this Idea too must now show itself as the
all-illuminating and all-determining element in this whole manifold sphere of
cognition---and then this sphere must also, by the power of that Idea, be formed into a
whole of cognition, in which the Idea will then show itself as the all-real and
all-filling. To behold the individual members or parts in their truth belongs not only
that the real essence which is their true ground be beheld in them, or that they be
beheld in relation to this, but also that the necessity of understanding---indeed, the
connection of thought---be seen, in accordance with which that highest being comes,
besides all else, also to lead to precisely such a particular's acquiring reality; for
only thereby does this particular come to its reality at all. But the individual member
cannot be seen in its place and in its relativity except in that a whole systematic
cognizing, embracing at the same time all members connected with it, comes about.

\begin{center}Remark and Supplement.\end{center}
In science the Idea comes forward in an \emph{Other}, in which Other it now shows
itself to be the truth, or to be the true and the essence. Indeed, it first shows
itself truly as Idea when it shines forth as the true in the Other whose truth it is.

It is thus, to be sure, subjected to an \emph{opposition}, and regarded as one that is
seen in its truth and significance only in such an opposition. Even when we speak of
the Idea in and for itself, in its purity, we think of it as over against the Other,
above which it is to raise itself in order to be seen in its purity. But it cannot be
otherwise at all; for already, in presenting itself to consciousness, there is here
something besides itself---namely, the consciousness to which it comes---so that it
already so far at once comes forward under relation, and this all the more because our
beholding necessarily goes in one with a productivity, and this productivity, in which
our own life expresses itself, at once enters into all the relations to which our life
in general is subjected. Because we ourselves belong to that Other in which the Idea
comes forward, we cannot have it entirely without any accession of this Other. So that
the expression ``the Idea in its purity'' can have only relative validity. But this
relative validity it does have. As the Idea shines forth through the Other, we must
nonetheless behold it itself as something other than this Other through which we behold
it; and in that we, as far as possible, see both in their opposition, we see them each
on their own---indeed, we see the Idea too in its purity, as it were separated out from
the Other from which it shines forth.

\begin{center}\large\S.~13.\end{center}
\markboth{\S~13}{\S~13}
The first thing in genuine and complete science, the true real and substantial in it,
by which alone all its manifold first has any significance, is the \emph{Idea}. As the
self-constituting and self-realizing living principle, it is to run through science and
live in it, and thereby bear the whole; therefore to meet us everywhere in science as a
true center, or to be that to which every point in it points. As the spirit in the
whole of science, as that by which everything in it first finds its truth and reality,
it is not to be sought in these or those principal concepts or principal propositions
alone; only the whole science in its infinite completion would be its adequate
expression---so far, that is, as what has determinate form can express or utter the
true infinite, which moreover lives above all form in pure spirituality and freedom.
Nor is the Idea present in any uttered whole, as in a finished work, otherwise than in
the active life by which this work, this expression of it, is constantly brought about
anew and brought forth or created, or constantly reproduces itself. \emph{Not in
standing results}, but in that which brings it about that such results \emph{come
forth}, do we have the Idea and the true rational element in cognition.

In referring everything determinate---everything that shows itself with a determinate
form and as a determinate Something---to the Infinite and Eternal, as to the Absolute,
from which all consideration must proceed and to which it must return, the
\emph{philosophical} element in cognition shows itself; and since everything is thus to
be referred to the Absolute, all cognizing must strive to become a philosophical one.
Insofar as here an immediate glimpsing and beholding of this Absolute everywhere lies
at the foundation, there everywhere enters here a \emph{speculation}. The speculative
shows itself in this: that, in the course of this referring and determining, the same
single Absolute, as the all-embracing and all-being, everywhere stands before the eye
of the spirit in its wholeness and infinite significance---as that which, while it
determines and, as it were, assigns to all things their relative essentiality, is
itself above all form and constitutes the formless and invisible center toward which
all directions point, but which comes to visibility only by being that to which all
else is referred. Speculation is here a pure beholding; for what, as the utterly first
and immediate, is had only in being beheld as the enlivening principle and essence in
all else can be cognized only in that it is now directly posited and seen as the
immediate being, as the essence in the forms which by it become its expression---just
as the life in the organism must be directly beheld as the first real being in it.

From the Idea there indeed necessarily proceeds a discernment and comprehension. But
the cognition of the Idea is nonetheless no discernment and comprehension. For the
latter already presupposes that what is considered is beheld in a certain connection
and determined in a certain way in accordance with a connection---so that it is already
seen in relations and as a member in a chain. Meanwhile, concept is essentially bound
up with every Idea, and as the Idea comes forward in science, it comes forward in a
system in which everything is concept at every point, so that now the Absolute itself
(as was noted earlier) comes to stand as a member---as the first and foremost, to be
sure, but still always as a member. But in and under the discernment thus proceeding,
and the conceptually correct deriving and determining, the Idea is then had not in this
foremost member, but had as that which---just as, in the manifold considered, it is the
single real essence that comes forth from it in its unity---thus becomes for us the
soul in our whole deriving and determining. Cognition is an ideal beholding, which is
had as the foundation in a living infinite thinking, in which the Absolute
constantly---while it mediates itself in a manifold---shows itself in its pure,
immediate absoluteness.

\begin{center}Remark and Supplement.\end{center}
1.~By the cognition of the Idea, and the illumination it gives, the manifold
\emph{detail} of individual cognitions---apprehended empirically or found out through
reflection, more or less deep-reaching---first \emph{truly benefits us}, because
thereby we first see what we really have in them. So far, the Idea is that on which
everything first depends, that which is first to be striven for. But on the other side,
in turn, the exact study of the manifold detail is that by which the Idea first truly
benefits us. A certain immersion in the philosophical---under which the individual and
particular is regarded only as a means and vehicle for the consideration of the Idea,
and certain principal truths or principal propositions, with their all-sided
consideration, are chiefly brought forward---is, on the one side, just as important in
academic study as, on the other side, the exact re-examination of the material and the
detail in its whole extent.

2.~Just as wisdom consists in knowing what has unconditional worth, what has only
conditional worth but still has worth, and what has no worth at all or only a slight
one, and in now keeping this in view in all judgments and actions and determining these
in accordance with it---so philosophy shows itself in cognizing and knowing what has
absolute essentiality, what has only a relative being and essentiality, and what,
finally, has none, or has it only as semblance, or at least in the lowest sense---and
in now, accordingly, resolving all scientific questions, so that it is discerned
\emph{what} each thing, with respect to the highest being, can properly be said \emph{to
be}. Although this can indeed not happen without seeing every thing in its place in a
scientific whole---and all philosophical grounding and determining therefore
necessarily has regard to such a whole---still genuine and deep philosophy shows itself
not only in the execution or possession of a scientific system, but can also lay itself
open and step forth with living power and clarity in the way in which singly given
questions are treated and individual cases are considered and judged. Indeed, this
immediately \emph{judicious} element---in which philosophy comes to view in a more
practical way than in the system, though as yet only in the domain of cognition---belongs
essentially to the appearance of a genuine philosophy. Philosophy shows itself so far in
a certain \emph{scientific way of thinking and judging}; indeed, even when it comes
forward in a certain systematic aggregate of determinate cognitions, it is had not in
these, but in the living thinking and beholding that has called them forth and can
constantly call them forth anew in other expressions and under new turns---so that here
too the same must show itself as in that immediate, free judicious element.

3.~What speculation, and the beholding for which the Idea constantly stands in its
totality, as the One and all from which everything proceeds and to which everything
returns, is in the scientific respect---that, in the merely personal and individual
respect, is this: to have an \emph{ideal} in view, to bear an ideal within oneself, and
to be inspired by and for it. The ideal is here the utterly completed science, as the
full, no-longer-deficient presentation of the Idea, in which everything would be beheld
in its infinite connections and significances. To such an ideal we look in the midst of
the deficiency of what has been attained, in order to strengthen ourselves, under our
piecemeal working and determining, by the glance toward the perfect---to which, after
all, the imperfect we already have points, and whose reality we already behold and, as
it were, possess in advance, in that we find ourselves on the way to it.

4.~But how, then, is concept different \emph{from Idea}, as regards the way in which
both come forward and are had? For a concept too is had in a genuine way only as the
moving and determining First and Ideal in the judgments in which it is expressed. In
these judgments it is to be sought neither in the subject nor in the predicate, but in
the fundamental thought in accordance with which such a subject is joined with such a
predicate; indeed, a concept is expressed not in one but in many judgments---from one
alone, but first from many, it shines forth as that from which these judgments proceed,
yet which, as it were, hovers freely over them. How, then, is Idea still something other
than concept?

The concept aims at the particular member, which is thought as it is thought because the
concept itself---as the fundamental thought that determines the member's real
significance and its mode of existence---determines how it is to be thought; and since
its real significance rests on relations and is determined with respect to relations,
which can be innumerable and vary in many ways, the concept therefore now becomes the
ruling and determining fundamental thought for a multitude---indeed, an
innumerability---of determinations; and we also see why it so often turns out that the
selfsame thing can be subsumed, and from another side again cannot be subsumed, under
one and the same concept, why stringent definitions are often so difficult, if not
unattainable. But insofar as here only \emph{one member}, a certain particular, is
considered in its real significance and now determined according to its relations
(though it is precisely this relativity of its real significance that brings it about
that the determination runs out into an innumerability of determinations, which can
vary incessantly, and in which the only constant is that the judgment in every
individual case turns out as it turns out)---insofar there is here, although there is
something ideal as determining First, still only a concept. If, by contrast, in the
given particular there is also beheld the not merely determining but also really
\emph{active First}, which brings it about that such a thought everywhere, in its
relations, comes to real actuality---or which sets the concept into reality---then we
behold the Idea therein. But now the concept is realized not only in the things we
consider; in ourselves too these determining fundamental thoughts arise. In this, then,
the Idea too is present. That which brings it about that such a concept comes to life in
us, that which enlivens the concept now moving in us in its particular sphere---this is
the Idea. That, too, is why only those concepts which aim at objects beheld as members
in the whole and considered in this their conjunction with everything else show
themselves as organic centers for an innumerability of determinations and judgments;
mathematical concepts do not have this character; in mathematics each form is seen
entirely isolated on its own, and there is no question there of real significances.
Where these are kept in view, one at once looks toward a whole; but in the cognition of
the relativity that thereby everywhere shows itself, there stirs of itself a sense of
the Absolute. If it comes to cognition what it is that brings it about that we determine
and judge and discern as we do by force of the concept, then we are already led into a
system and are now also brought to glimpse that from which all system proceeds, in which
all system hangs; and then we already see the Idea, as that which reveals itself at the
individual point through the concept---and the cognizing that was in itself already an
expression of the Idea's life in us now becomes a cognizing by force of the cognition of
the Idea.

\begin{center}\large\S.~14.\end{center}
\markboth{\S~14}{\S~14}
As the purely speculative would express itself in a scientific whole, the
\emph{dialectical} enters, which is the expression of speculation itself as it sets
itself in relation to the manifold, determining this and forming it after itself.
Dialectic enters in the transition of the Idea and of speculation to the development
of a systematic whole. It thereby comes forward at once and in one: first, in a
\emph{negative} way, as a dissolution and annihilation of all the modes of
representation that fix themselves at lower standpoints, insofar as these---perhaps
even naturally, indeed necessarily---have taken on the stamp of a significance,
indeed an absoluteness, that did not belong to them; but then, secondly, at the same
time in a \emph{positive} way, as that which precisely, in showing things together
with their concepts in their relative invalidity or nothingness, \emph{eo ipso} shows
them in their relative significance or somethingness, and thus leads to assigning them
their place and to carrying out, or \emph{constructing the scientific whole}, in which
everything is seen in its relations, and moreover seen as it should be seen. So that
the dialectical---as that which proceeds from the Idea, beheld and had as the pure
substantial in cognition---precisely because it exposes the nothingness with which
everything named as something determinate in reality (and not only the objects but the
concepts themselves) is afflicted, and thus, as it were, shakes the fixed assumptions
in their foundation, at the same time also leads to assigning them their relative
validity and establishing them in it.

\begin{center}Remark and Supplement.\end{center}
1.~Dialectic shows itself here as an \emph{art}: first, insofar as it consists in the
movement of thought, or advance in thinking, by which the Idea is in fact made to hold
with respect to a Something which---though it is first to be determined and illuminated
by the Idea---has nonetheless, independently of it, presented itself in its
manifoldness and fixed itself, each on its own, with a significance, indeed an
absoluteness, which, when it is now to be warranted before the philosophical
consideration that is to appraise and determine its title, cannot hold up. But since
nothing can be thoroughly shown in its invalidity without its also being seen whether
and how far it is, in its sphere or in its relations, nonetheless something that has not
without reason entered, or come to be named along with the rest as something particular
on its own, this \emph{philosophical art of annihilation}---which, unlike skepticism,
does not have a merely negative aim, but sets out to annihilate all the lower
understanding's modes of consideration in order to put the highest cognition of reason
as the sole valid one in their place---will at the same time necessarily and
essentially go in one with the philosophical, scientific art of construction.

Just as in the organic world the incessant change is a necessary consequence of life's
uninterrupted living in its material (see the author's \textit{Psychologie}, § 5), but
now also points to, and bears witness to, life's inner power, bound by no form, so
\emph{the relative}---and what is therefore incessantly determinable otherwise, never
exhaustible in determinate expressions and propositions---is, in all philosophical
cognition, a necessary consequence of the Absolute's absoluteness, and of the fact that
the Other in which the Absolute comes forth is never the Absolute itself; but this
relativity without end then also points to this Absolute, and from the dialectical
movements of thought (in the usual negative sense) the cognition of the Absolute must
therefore always emerge strengthened. Even if the selfsame can be both affirmed and
denied of the selfsame (if, to take simple examples, matter can be said both to be
something on its own and not to be something on its own; if it can be said both that the
concept of several virtues has reality and has no reality), we nonetheless have the
positively valid in that which brings it about that we must thus both affirm and deny;
and when we rightly see how, precisely in affirming, we must deny, and vice versa---or
how we see the affirmation in its validity only when we at the same time see the
validity of the denial---then we have, precisely in this discernment, the Idea, or the
immediate Absolute, under this its appearance. And a direct consequence is now to seek
the union in a deduction of the concept, in which the concept is beheld at its
\emph{relative} place with its relative validity; and the dialectical movement of
thought thus directly points to, and leads to, the scientific construction.

2.~In dialectic the \emph{logical} and the \emph{metaphysical} go in one; and it has
often been noted that in the ancients' division into dialectic, physics, and ethics, the
first stands in place of both the logic of later times and its metaphysics, or at least
its ontology. A speculative logic (not a merely observing, reflecting, and describing
doctrine of logical forms) is seen in the dialectical in its practical appearance in
investigations in which given concepts are thought through and determined with respect
to the highest true, which highest true is the object of metaphysics; so that both in
this respect, and insofar as a speculative logic itself has the cognition of this
highest true for its basis, the metaphysical goes in one with the logical. If that which
is in living movement in the dialectical is made the object of a special attention,
logical cognitions could spring from it.

3.~The dialectical, in the ordinary sense---namely, in its negative direction---sets out
to contest and abolish, or at least to refer back within the proper relations and
limitations. First it turns against such \emph{modes of representation} as may have
formed and fixed themselves, in humanity's endeavor to fathom and cognize, with a
validity that cannot belong to them. Next it is directed at the very \emph{objects}
given for intuition, or at the representations necessary to the human being in the
sphere of immediate real intuition. The highest, finally, is that it turns against
\emph{the concepts themselves}---namely, the necessary concepts, especially those
obtained a priori---and shakes even these through in such a way that their validity
comes to confirm itself only through an insight into their relative invalidity. Which
last cannot but be the aim of a dialectical endeavor; but this is what Plato (see his
\textit{Parmenides}, p. 129), while laying particular weight on it, also finds
especially difficult.

\begin{center}\large\S.~15.\end{center}
\markboth{\S~15}{\S~15}
The Idea's execution in a complete presentation, in which everything is beheld in its
place and in its significance---indeed, in its truth---in accordance with the Idea and
by force of the cognition of the Idea, leads to a \emph{system}, and scientific
cognizing now becomes a \emph{systematic} one. In the execution of the system the Idea
must stand its \emph{test}, whether it can now show itself everywhere as the real and
can, in the particular, furnish and fulfill what it, as Idea, should be able to furnish
and accomplish. That all cognizing ought to be systematic is clear enough, since only in
a system can everything be seen in its relative place and in its relations, whereas
everything is precisely only what it is, and has its significance, in its relations and
at its place.

Such a system is now necessarily carried out in and by \emph{concepts} and thereby
itself becomes a system of concepts; for since in it everything is to be seen in its
connection and determined with respect to this connection, as it shows itself in
accordance with this connection, cognition becomes throughout a cognition of connection
and of the objects' real significations and forms of appearance, or fundamental forms,
thereby determined---so that everywhere every point is determined as it is determined in
accordance with a discernment and comprehension; indeed, even the object of fundamental
cognition, the Idea, must now (as already noted earlier) enter as a member---namely, the
foremost member---into the whole, quite as if it, though the highest object, were
nonetheless only a certain particular object as well, like the rest and alongside them.
But although the Idea must thus itself come forward as a concept, it must nonetheless at
the same time be seen as the all-real that runs through the whole system, and presents
itself in, and shines forth from, every point, as we have earlier sought to show. If,
then, the Eternal is seen as the real essence in and under the many fundamental
forms---if the Idea is beheld and had as the all-determining and everywhere really
working element that, through that system of concepts, presents itself, in its unity yet
under many shapes---then it becomes evident how, in science (as in real actuality), the
Idea, its unity unimpaired, can come forth in a \emph{plurality of fundamental forms}, in
each of which it shows itself in a new shape, and that one can now speak of a
\emph{plurality of Ideas}. How the Idea is only one and is not itself several, and yet
again so steps out into several that several Ideas can be named, becomes clear when the
Idea is seen as that which externalizes itself in a system which, so far as it is derived
from the Idea, is its expression, and in which, therefore, in every form thereby
determined the Idea itself must be beheld again---but beheld again as in many effects and
principles hypostatizing or constituting themselves on their own, with just as many
corresponding distinctive individualizations, hence as in many Ideas. These then, each in
its sphere, show themselves as the all-real and absolute, on which everything depends,
from which everything proceeds and into which everything returns, and now also step out
into a \emph{plurality of sciences}. But this all-reality and absoluteness of a given
Idea is nonetheless only such as holds with respect to everything belonging under this
Idea's sphere---so that it is only a relative absoluteness.

The Idea's life then shows itself, under the form of cognition and science as well as
under that of real actuality, as an \emph{organizing} one which---in becoming the
determining principle for a manifold aggregate of individual cognitions that are
distinct, yet in their distinctness and relative opposition closely connected---now
also, in thus giving this aggregate its whole shape, at the same time itself takes shape
therein. The stringent, sharp, and rigorous element that is in all life which organizes
itself and, moreover, fixes itself in a distinctive form in a determinate way, then also
comes to view here.

\begin{center}Remark and Supplement.\end{center}
1.~Although the Idea can indeed come forward livingly in the free \emph{judicious}
element, still even therein a system---at least certain basic features of one---must be
had \emph{in mind and in view}; and if the particular is to be determined in an exact way
and fully seen in its significance, then the reflection will necessarily enter into an
investigation of a comprehensive connection on all sides, and it will soon be felt that
only a total working-out of all the many considerations set in motion, into a
well-connected whole, can bring one to the exact determinations that alone can satisfy.

To genuine scientific endeavor there belongs, therefore, not only interest and respect
for the matter itself that is to be expressed, or for the content of cognition, but also
for the form in which it is to express itself, and a careful, indeed scrupulous,
reflection on at what place and in what connection each individual point is to be
presented, so that there may be definiteness in the particular, order and sequence or
consistency in the whole. A certain \emph{logical enthusiasm} must proceed from, and
accompany, the enthusiasm for the Idea. The former is of the same source as the latter.
In order that the Idea may enjoy its full right and show itself in its full power and
totality in the presentation, form too becomes so important.

2.~But only as the Idea's essential and necessary expression and becoming-real is the
system what it is. Therefore the free judicious element, as well as the dialectical, must
go before the system and prepare its way. Otherwise the systematic tendency easily leads
to the forming of \emph{opinions}---in order, after all, to bring the manifold that has
been set in motion, the many isolated cognitions that have been attained, to a certain
unity and wholeness---which, though hypothetical at first, gradually captivate and
ensnare. The development of a system must therefore not be \emph{hurried}.

The system must, further, also be kept perpetually alive in all its parts, or in
\emph{living movement}, by the fact that these are constantly beheld, as they are beheld,
or posited, as they are posited, by force of the Idea, which then, in incessant
reproduction, calls them forth anew. Hereby the many individual cognitions are first held
fast in such a way that, preserving themselves in constant freshness, they are now also
each time seen in their truth---which they do not have or show at all in their isolation,
or in their merely systematic conjunction with the other members, but only in their
connection with the Idea, as those which belong to its living total expression.

3.~Two things are here to be feared and forestalled. First, the \emph{nebulism} (to use a
word Goethe adopted in judging artists) that merely keeps the form, or the system, in a
certain haze, out of which it only glimmers forth in indeterminate outlines, in a certain
generality, without the particular being exactly determined, so that there comes to be no
firm bearing in the whole. The desire to behold the Idea at once in its great totality
sometimes even leads to a certain rapturous enthusiasm, in which only certain cognitions
take on a more or less settled shape, while the greater part is had only quite loosely,
and large gaps are filled out by a dim fantasizing, so that the eye may at least be
gratified by the semblance of a whole. Next, one must watch that the systematizing does
not end in a \emph{crystallization}, whereby in the end only a dead substrate remains,
which once bore an Idea and was the expression of its living movements, but is no longer
filled by it. If the system is to continue to be what it, in its truth, is---and moreover
ought to be---then it must be preserved under constantly new formations, in which it is
constantly reborn, in that it then also constantly transforms itself and comes forth with
new determinations, until, at certain stages of development, it even ripens into a
metamorphosis, in which it passes over, out of a form developed up to a certain point and
so long, with good reason, held fast, into another.

Yet even without a system's thus crystallizing---even if, under a constantly living
influence of, and reciprocal action with, the fundamental thought from which it
proceeded, it keeps itself in the most living movement---it can easily lead to
\emph{one-sidedness}, and so ensnare and bind that, since everything is finally seen only
in the light of this one system, it does not even enter into the new considerations that
might lead it to be reborn in a new shape, by taking up something new into itself with
respect to the first fundamental thoughts. And for this, room, or freedom of spirit and
willingness, should always be kept open. For since we behold \emph{the One} on which
everything depends only \emph{so far as}, or in the measure in which, we see it as one and
all in the corresponding whole, cognition must not only be able constantly to expand and
become more inward and powerful with respect to the manifold that is more and more
penetrated by the Idea, but the Idea itself must, in the course of this, be able
constantly to show itself not merely in a clearer, but also in a different light; and thus
cognition, precisely by expanding in the one direction, must be driven constantly to seek
a new---namely, a more primitive and more all-sidedly referring---\emph{first
starting-point}; so that, therefore, along with the manifold, the fundamental cognition
and the fundamental principle itself---that is, the way in which the fundamental principle
is seen and grasped---must be able to undergo a constant \emph{reformation}. And therefore
that free, immediate \emph{judicious} element---just as it must precede the system and
prepare for it---must also \emph{follow upon it}, when one now goes from the system to the
manifold detail. As important as a definite system is, so necessary is it to be able, to a
certain degree, to free oneself again from the system's bonds and let the Idea move
freely, independently of the form in which it has sought to come to a definite shape. To
this belongs that we, in living reciprocal action with the phenomenon or the problem
itself, let it exercise its power over us to lead us wherever it will, and let it itself
call forth in us the thought by which light is to be brought into it---which, however,
doubtless presupposes that there is a fund of thought from which the new thought, that is
now to correspond to the thing as the thing itself points to it, can spring. True interest
and love for the matter, and a desire to fathom it as it is, will lead one to let the
relation of reciprocal action between what we have and are, and the new particular we are
to consider, become as it should be, so that a free glance learns whether the new datum
must go straight into, and, as it were, be assimilated by, the system---so that the system
cannot thereby be brought to any transformation of significance---or whether its influence
must be such that the system thereby comes to move back into itself and, to a certain
degree, to loosen itself out of its fixed state, in order, in assimilating the new, to
advance in its own developmental process and, in consequence of this reception, itself to
undergo a transformation, be it only in one single respect. In particular, the endeavor to
bring what has been cognized to others---when it becomes really important to us to convince
them, and we are therefore compelled to enter into their trains of thought---can work
against the one-sidedness to which the development of a system easily leads, against the
bonds it can lay on free thinking. In general, reciprocal action with others---and with
several different thinking beings, whether under a mutually living communication in
intercourse, or under the study of the works of different authors, or of the history and
historical development of a science---cannot but be of importance in this respect.

4.~A complete system (that is, one come to a certain degree of completion) lays itself
open \emph{practically} by the fact that every concept and every object of investigation
is seen exactly in its connections with all the other concepts and objects set in motion
along with its consideration, and that the course of the Idea can be carried from every
point straight up to the highest, from which, as the fixed starting-point, everything
proceeds, and can be carried on all sides, so that the individual point considered can
thereby be set entirely in its proper light and fixed in its relations and with its
significance. Insofar, now, as all the principal points in a science are seen in this
clarity and connection, a whole system can already shine forth---in sure and manifold
features all referring to one fundamental form---from the treatment of the individual
questions that arise, and from the way of judging and determining the concepts on the
whole, without its yet being had expressed in the systematic form. To deliver the system
as system in an outward scientific \emph{presentation} can itself, even if it is already
thus practically possessed, have its great difficulties. In particular, the order and
sequence in which everything is to be presented can cause many scruples and doubts. The
illuminating and grounding element ought indeed to stand in front; but since everything
individual, as in an organic whole, rests in the Idea---while this in turn is had only in
the execution of the whole system, and becomes evident in its full truth only when it
shines forth from the whole scientific aggregate of cognitions as the true in
them---every individual cognition really rests in the whole system itself, which
nonetheless, when it is to be presented, can be built up only by the individual cognitions
gradually joined to one another; and just as the thinker who would express the system
already has the whole system in view at every point and sees the individual in the whole,
so too, in the presentation, at the individual points---indeed, already at those with
which it begins---the whole total cognition should come forward livingly and be set before
the beholder's eye. But this cannot happen in a presentation that can only be given member
by member. Something must be taken first and placed in front. Indeed, this cannot even be
the fundamental cognition, which must first arise before our eye and ground itself as
fundamental cognition before it can be set up as such; nor can it, as that without which
the whole is not seen in its true light, come last. One must be led up to it, and then it
must enter as that from which the whole system is now to proceed. As regards, too, the
many individual cognitions springing from the fundamental cognition, which become the
manifold members of the system, it will not seldom turn out that the one must already be
had and presented merely in order to be able \emph{to arrive at} the other, but that this
other in turn must be \emph{presented} before the first can be seen in its \emph{full truth
and significance}. Indeed, not only individual cognitions and propositions, but \emph{whole
courses of Ideas} (even whole sciences) can stand in such a relation to one another that,
while each of them demands to be worked out and presented on its own in its direction and
in its sphere, the one nonetheless can neither be fully grasped nor become entirely evident
without presupposing the other---and that mutually. In such cases the systematic
arrangement brings great difficulties with it.

It then becomes necessary \emph{to take from a certain side} what in itself, however, does
not proceed from this one side, or raise this side up in itself more than the others. But
just as from this is seen \emph{the possibility of several procedures or methods} that can
make an equal claim to groundedness and worth, so from it too it becomes evident how, in
the same science, \emph{several scientific systems} can be set up alongside one another,
without any of them being able to claim sole validity or a decided superiority in all
respects---and that even if, as regards the scientific content on the whole, agreement
prevails.

5.~Genuine system is only that to which we come from the Idea and by the Idea, in that the
Idea is considered as it expresses itself in a certain aggregate of detail, the extent and
kind of which detail is then determined by the Idea itself. But also when an aggregate of
particulars is given to us on its own (in an empirical way) and becomes the object of an
observation aimed at beholding it as it once for all offers itself, without its yet being
cognized in the light of the Idea, the endeavor will always be to order this aggregate of
observations, or observed forms, and the (merely formal) common concepts formed about
them---or genus- and species-concepts---into a system, into a comprehensive survey. But
here too the philosophical endeavor can and should show itself \emph{partly} in the way in
which a certain ideal of such a classification, with respect to what it is by its purpose
to furnish, is constantly kept in view, \emph{partly} in the way in which, in its
execution, a higher consideration is worked toward. (Cf. § 6, Remark 2, at the end.)

6.~The necessity of a system in science follows so directly from the fact that nothing
individual can be seen in its proper light without being exactly seen in its relations and
its connection with everything connected with it, that no exception can find place there.
One science, however, seems to make an exception---a science that, when its
\emph{presentation} is in question, is regarded rather as an art: \emph{history}. But
events too are seen, like everything else, in their proper full light only when they are
seen in their often many-sided connection with others---partly contemporaneous, partly
preceding and following---and characters too must be seen in their times and at their
places. So that the same grounds that elsewhere demand the systematic in cognition also
clearly show themselves here; but what elsewhere, in other sciences, has another form and
is called system, is here the \emph{historical composition}.

\begin{center}\large\S.~16.\end{center}
\markboth{\S~16}{\S~16}
% [text to be added: pp. 146--152 (\S 16, main text)]

\begin{center}Remark and Supplement.\end{center}
% [text to be added: pp. 146--152 (\S 16, Remark and Supplement)]

\begin{center}\large\S.~17.\end{center}
\markboth{\S~17}{\S~17}
% [text to be added: pp. 153--158 (\S 17, main text)]

\begin{center}Remark and Supplement.\end{center}
% [text to be added: pp. 153--158 (\S 17, Remark and Supplement)]

\begin{center}\large\S.~18.\end{center}
\markboth{\S~18}{\S~18}
% [text to be added: pp. 159--159 (\S 18, main text)]

\begin{center}\large\S.~19.\end{center}
\markboth{\S~19}{\S~19}
% [text to be added: pp. 160--161 (\S 19, main text)]

\begin{center}\large\S.~20.\end{center}
\markboth{\S~20}{\S~20}
% [text to be added: pp. 162--164 (\S 20, main text)]

\begin{center}Remark and Supplement.\end{center}
% [text to be added: pp. 162--164 (\S 20, Remark and Supplement)]

\begin{center}\large\S.~21.\end{center}
\markboth{\S~21}{\S~21}
% [text to be added: pp. 165--193 (\S 21, main text)]

\begin{center}Remark and Supplement.\end{center}
% [text to be added: pp. 165--193 (\S 21, Remark and Supplement)]

\begin{center}\large\S.~22.\end{center}
\markboth{\S~22}{\S~22}
% [text to be added: pp. 194--204 (\S 22, main text)]

\begin{center}Remark and Supplement.\end{center}
% [text to be added: pp. 194--204 (\S 22, Remark and Supplement)]

\begin{center}\large Conclusion.\end{center}
% [text to be added: Slutning (Conclusion), pp. 203--204]

% [Rettelser / errata leaf, printed p.205: corrections to the DANISH text
%  (page/line orthography fixes). Meaningless in English translation.
%  DECISION DEFERRED (see TRANSLATION-NOTES): omit, or keep as a short
%  translator's note recording that the Danish had an errata leaf.]

\end{document}
