% ============================================================
%  Harald Høffding, Human Thought: Its Forms and Its Tasks
%  (Den menneskelige Tanke, dens Former og dens Opgaver)
%  København og Kristiania: Gyldendalske Boghandel, Nordisk Forlag, 1910.
%  ENGLISH TRANSLATION
%  Source of truth: transcription.tex (Danish), COMPLETE (392 pp.).
%  Page-offset (scan only): PDF = printed + 15.
%
%  HOW TO USE THIS FILE
%  - This is a SKELETON. Every heading is in place; the prose is not.
%  - `grep -n "text to be added" translation.tex` lists the markers in order.
%  - For each marker: open transcription.tex, find the SAME heading/label,
%    read from that heading to the next, translate the span, replace the marker.
%  - Conventions: ../../../TRANSLATION-PLAYBOOK.md + ./TRANSLATION-RESUME-NOTES.md
%  - Heading glosses below are PROVISIONAL — verify against the Danish on first
%    use (see TRANSLATION-RESUME-NOTES.md "Heading glosses" for flagged choices).
%  - Markers under a heading that has NO intro text before its first sub-heading
%    (B., C. in Ch. I; A., B. in Ch. IV; a. Det etiske) cover an empty span —
%    just delete that marker.
%  - The user commits and pushes. You never do.
% ============================================================
\documentclass[12pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{libertinus}
\usepackage{libertinust1math}
\usepackage{textalpha}   % Greek glyphs carried verbatim from the Danish
\usepackage[english]{babel}
\usepackage{amsmath}     % \underset in the difference-series formulas (§§62ff.)
\usepackage{amssymb}     % \gtrless etc. (Poincaré's continuity note, §59)
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{setspace}
\usepackage[margin=1.2in]{geometry}
\usepackage{enumitem}
\usepackage{fancyhdr}
\setlength{\headheight}{28pt}
\addtolength{\topmargin}{-13.5pt}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Human Thought: Its Forms and Its Tasks},
  pdfauthor={Harald Høffding},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\textit{Human Thought}}
\fancyhead[LO]{\textit{\leftmark}}

\setstretch{1.3}

\title{\textbf{Human Thought}\\[6pt]
  \large Its Forms and Its Tasks}
\author{Harald Høffding\\[6pt]
  \normalsize translated from \textit{Den menneskelige Tanke} (1910)}
\date{København og Kristiania:\\ Gyldendalske Boghandel, Nordisk Forlag, 1910}

\begin{document}
\maketitle
\thispagestyle{empty}
\newpage

\tableofcontents
\newpage

%% ============================================================
%%  FOREWORD (Forord, August 1910)
%% ============================================================
\chapter*{Foreword}
\addcontentsline{toc}{chapter}{Foreword}
\label{ch:forord}

\noindent The life of thought, which it is the task of my book to
describe, is intimately interwoven with the other sides of the life of
the spirit. I have sought to show that thought cannot serve the life of
the spirit as a whole unless it first and foremost becomes master in
its own domain and unfolds itself according to its own laws as an
independent side of the life of the spirit. That this unfolding, when it
is rightly carried through, will bring with it great and far-reaching
changes in personal life as a whole, is something of which I become more
and more convinced. There are here oppositions and problems that will
take on ever sharper forms, and I feel no admiration for those who here
sense no sting, although they have sufficient opportunity to come to
know the life of thought.

Just as I have on the whole stressed the personal factor in all
thinking, the more this approaches the border-regions, so too I have not
concealed that my own stance at thought's final station has a personal
character, and I have not been able to refrain from stating how, in my
experience, one feels at this station.

I still have much to learn. My work will need filling out and
correction on many points. But beyond the framework I have given here, I
cannot go. And it affords me a good place to work. May there be one or
another who could also find work here, even if this work should lead him
to alter both the framework and its content.

\bigskip
\noindent August 1910.\hfill \emph{Harald Høffding.}

%% ============================================================
%%  I. THE PSYCHOLOGY OF THOUGHT (FUNCTIONS)
%% ============================================================
\chapter*{I. The Psychology of Thought (Functions)}
\addcontentsline{toc}{chapter}{I. The Psychology of Thought (Functions)}
\markboth{I. The Psychology of Thought}{I. The Psychology of Thought}
\label{ch:psykologi}

\begin{center}---\end{center}

\section*{A. Psychic Energy}
\addcontentsline{toc}{section}{A. Psychic Energy}
\label{sec:psykisk-energi}

\noindent 1. The task I have set myself in this book is to investigate
human thought---a task that all philosophers have set themselves,
whether they have been conscious of it or not. For all questions that
can be raised concerning life and existence depend---both with respect
to the way in which they are posed and with respect to the way in which
they are answered---on the conception one has of the nature and
operation of thought. It is therefore also an outline of my philosophy
that I shall give here. That is to say: I shall give an account of the
thoughts to which I have come back again and again, when I have occupied
myself with such questions as: the presuppositions of science, the
significance of the results of science for one's view of life, and the
relation between scientific cognition and other sides of the human life
of the spirit.

2. We know thought best as reflection. The word \emph{Eftertanke}---``after-thought''---I
use not only in its strict sense, ``in one's thinking to \emph{seek
after} something or \emph{pursue} something,'' but also in the sense
that thought \emph{comes after} an immediate experience. Reflection thus
presupposes that we have seen, felt, experienced something. During the
experience we were perhaps wholly absorbed in it; but now we seek to
bring it forth anew, either to dwell once more on what was experienced,
or to go through its various parts or properties or its relations to
other experiences. In mere dwelling, reflection is one with a beholding,
a mere representing. But such dwelling can scarcely persist for long in a
wholly quiescent form. Involuntarily---perhaps precisely because of the
interest that what is experienced holds---a movement of thought will be
released, one that serves to bring out its inner and outer relations.
Thus when the poet involuntarily reaches for similes and images in order
really to let his subject come to expression. Or when perhaps it is
precisely the contrast with other experiences that serves to illuminate
the peculiar character of the single experience. Both likeness and
difference can thus serve to cast a clearer light on what was given in
the original experience, and reflection therefore does not necessarily
stand in opposition to it, but can work in its service and in its
spirit, when it has itself vanished.

By the word experience we think chiefly of states in which we were
essentially receptive. But reflection can also serve the involuntary
life of the spirit, where this appears as urge or striving. In an
instinctive or inspired manner, involuntary striving is often led to its
goal without any reflection being needed, and perhaps better than could
happen with the help of any reflection. When reflection comes to operate
in our action, it will be because the involuntary striving met
resistance. There may then be needed a deliberation upon what it really
is that is being striven for, upon this goal's relation to other
possible or actual endeavors, and upon the ways and means that can lead
to the goal. Here too it is through the bringing-out of likenesses and
differences that reflection renders its service. And here too a
difference can appear between dwelling and movement. It may be that the
representation of the goal is held fast without great oscillations, and
that the relation to ways and means or to other possibilities plays a
subordinate role, as when the helmsman stares at the compass or the
seamark, and his hand involuntarily moves the rudder according to what
he sees. But it may also be that the strongest movement back and forth
among various possibilities takes place and must take place before the
decisive concentration that is called decision sets in.

The arising of reflection is like a kind of awakening, not from sleep or
from unconsciousness, but from the involuntary mental life and its
experiences, from states that may provisionally be characterized as
sensation, intuiting, representing, mood, emotion, urge, and striving.
In its dwelling form, reflection stands closest to the involuntary
mental life. In its striving to bring out likenesses and differences, it
moves further away from the involuntary. Linguistic usage marks this
difference rather aptly by distinguishing between \emph{thinking of}
something and \emph{thinking over} something. It is the latter kind of
reflection, the discursive, mobile kind, that we shall have most use for
in what follows, since it is the one we can best observe, and likewise
the one that plays the decisive role in the tasks of thought. Only for
this kind of reflection do problems exist. In what follows I therefore
understand by thinking this kind of reflection. And yet there is only a
difference of degree between the two kinds. Only for a brief moment is a
dwelling representing, a holding-fast of a memory-image, possible;
precise observation shows that in greater or smaller rhythms we move
toward and away from the object we are thinking of. And every return can
bring with it a recognition, so that the relation of likeness comes to
play a role just as in discursive reflection. Perhaps the representation
was new to us at the first moment; but through repeated dwelling it
becomes familiar. Apart from this too, dwelling reflection by no means
excludes that manifold other representations and experiences can
contribute and perhaps precisely condition the interest that makes
continued holding-fast of a single representation possible. What holds
for discursive reflection will therefore also hold, to a certain degree,
for the dwelling kind.

It emerges from this description that the work of thought consists in
comparison, in distinguishing or finding likenesses. When I think over
something, I seek to determine its relation to something else, and the
more relations of likeness and difference I find, the better I have
thought it through. This holds of course also when it is one and the
same object that I examine in different connections and under different
conditions; I then compare the different ways in which the object
appears in the different cases.

But comparison in turn presupposes a holding-together or combining,
synthesis. I can compare two objects only when I either have them given
to me simultaneously in my sensation, my memory, or my imagination, or
else retain a representation of the one while the other is present to
sensory observation. If the one object constantly vanished entirely when
the other emerged, no comparison could take place. The work of thought
consists in a combining by which likenesses and differences become
determinative for the relations in which the objects are placed to one
another in the connection that results from the work. Synthesis is an
expression of energy, of the capacity to perform work.

The resistance that is overcome through this work is due to the objects
against which reflection is directed and which are given in the
involuntary mental life---in sense-sensation, memory or imagination,
feelings of pleasure and displeasure, urge and striving.

The combining work must become the more difficult, the greater the
multiplicity of objects that are to be brought into relation with one
another. The fullness of content can be so great that the psychic energy
available in the particular case cannot master it. There then threatens
a dissolution of the mental life, a transition to chaos.

Yet it depends not only on the multiplicity, but also on the degree of
difference that the objects present. The more they resemble one another,
so that they can fuse, or so that at least the work done in the case of
the one need not be repeated in the case of the other, the less work is
required. Objects that stand in strong mutual opposition, or perhaps
even conflict, require a great energy if they are all to be taken up into
the conscious connection of the mental life. This holds both for the
inquirer, who must find a point of view from which the conflicting can
stand as one, and for the practically striving person, who must
subordinate the conflicting impulses and interests to the thought of a
great and guiding purpose.

It will, thirdly, depend on how independent a place the individual
objects are to occupy in the connection that is being worked on. A
schematic whole is easier to form than a whole that lets the
peculiarities of the individual objects each come into their own. Here
it is precisely likeness that can create a difficulty, since it easily
leads to fusion and thereby to the abolition of the independence of
individual objects. On the other hand, from this point of view a
pronounced difference or opposition can be a relief.

Furthermore, a greater work is required the further apart the objects
lie from one another in time and space. The simultaneous and the near
join together involuntarily, and the connection perhaps even forces
itself upon us. And if we leave the objects before us in order to
immerse ourselves in the world of memory and imagination, here too the
connection can often form itself with ease; ``with great ease thoughts
take up their abode with one another,'' but when it is a matter of
holding together a present object with distant memories, or with ideals
that have been properly to the fore in the soul only in isolated moments
of enthusiasm, then special demands are made on psychic energy. To have,
in the situation of the moment, the treasures of one's experiences,
one's reflection, and one's ideality fully at one's disposal is perhaps
the greatest testimony to mental energy.

Finally, it plays a role how close and intimate a connection the
objects, despite their difference, are to enter into. There are here
many degrees, from wholly external juxtaposition and grouping to the
inner connection that a scientific concept, an artistically formed
image, or a great decision can express.

Besides these five conditions---which might be said to form an analogy
to what mass is in the case of physical energy---account must also be
taken of the speed with which the synthesis proceeds. The genius and the
man of will can in an instant expend a capital of energy that other
people must distribute over a longer span of time. In the domain of the
life of feeling, this opposition appears as an opposition between
vehemence and inwardness, emotion and disposition\footnote{See on this
my \emph{Psykologi} IV, 7; VI E, 4--5.}, and analogous oppositions hold
for the other sides of the mental life. In research, discovery often
forms an analogy to emotion, the conduct of proof to disposition.

The psychic energy that expresses itself in synthesis we measure by
taking all these circumstances into account. Of exact measurement there
is of course no question here, only of an evaluative estimate. But as
soon as we seek to give grounds for why one work of thought, one
intellectual orientation, one view of life can be called lower or higher
than another, we must take all these circumstances into account, and we
in fact do so, more or less consciously, and more or less precisely.
There is less psychic energy available in the dream-state, where the
changing objects each release their own train of representations, than
in the waking state, where dominant purposive thoughts bring about a
main direction. It is a purely psychological measure that we apply here.
And even though, in forming the concept of psychic energy, we have had
guidance in the analogy with the concept of physical energy, that
concept is nevertheless built upon its own peculiar experiences, in
which physical or physiological observations and investigations need
play no decisive role. For this reason we need not fail to appreciate
the significance of the attempts that have been made, from the
experimental-psychological and physiological side, to measure the
brain-work presupposed in certain elementary mental activities---e.g.
calculating or remembering---through its relation to the muscular work
that can be performed simultaneously with it. One can detect such
thought-work by the diminution it produces in the muscular work
performed at the same time. But this very measurement presupposes that
one can, by purely psychological means, distinguish between more
difficult and easier mental work; otherwise one moves in a circle. The
brain-work corresponding to the mental work described above is, after
all, not known directly, but only through the very mental work to which
it is assumed to correspond.

3. Reflection presupposes objects on which it can exercise its combining
and comparing activity. But now these objects? Does the law of synthesis
hold for them as well? We stand here before a question of a peculiar
difficulty, closely bound up with the whole nature of the mental life.
It is only reflection that acquaints us with the objects. It is the
medium through which they are accessible to us. The involuntary mental
life in its immediate unfolding must not be described and judged
straightway according to the way in which it shows itself to reflection.
It is a life within which we are not permitted to presuppose properties
and determinations that first come into being through the work of
reflection. When, in isolated moments, we are able to immerse ourselves
in this involuntary---but not for that reason unconscious---life in such
a way that we can afterwards retain a memory of it, it stands before us
as a stream, a gliding-along of manifold objects, which we designate as
sensations, memories, moods, endeavors, and so on. It is not a chaos of
independent objects; already in the involuntary mental life there are
connections and wholes as well as differences and oppositions. Every
word we use about what we can retain from the involuntary states bears
witness to this. Most striking, no doubt, is the continuity; this is why
the word ``stream'' offers itself so readily for what we call the
immediately given. But it is not a uniform stream. Differences of
quality make themselves felt; otherwise we could not distinguish between
the kinds of objects called above sensations, memories, and so on. And
there can be eddies in the stream, or windings, so that the direction is
not kept in a straight line. That is to say, strivings of different
tendency, of more or less opposite direction, can make themselves felt.
With this is bound up, again, that inhibitions can set in, while in
other cases, where the different tendencies go in the same direction, a
facilitation or an acceleration can appear.

Through the indefinite and figurative expressions that alone stand at
our disposal for indicating the basis of objects that reflection
presupposes, it shines through that this basis forms no absolute
opposition to what emerges as a consequence of the work of reflection.
There too we find both unity and multiplicity, both continuity and
discontinuity, and we find many degrees and nuances of these relations.
It is therefore not justified, with the older English school
(\emph{Locke}, \emph{Hume}, \emph{Mill}), to conceive the immediately
given in the mental life in an atomistic way, as if it consisted in a
multiplicity of mutually independent elements. Both in the psychic and
in the physical domain, it is only reflection that leads to the
formation of atomic concepts, guided by the urge to make the differences
as distinct as possible, so that the connection of them can become as
clear and perspicuous as possible. The unity to which reflection bears
witness, and which it itself seeks to bring about, thereby becomes
enigmatic. But it is also unjustified, as in our day especially
\emph{William James}\footnote{\emph{The Stream of Thought}. Mind 1884
(reprinted as Chap.\ 9 of \emph{Principles of Psychology}). Later
especially \emph{A pluralistic universe} (1909), Sect.\ 7.} and
\emph{Henri Bergson}\footnote{\emph{Les données immédiates de la
conscience} (1888). Later especially in \emph{Introduction à la
Métaphysique}. (Revue de Métaphysique et de Morale 1903).} have
attempted, to present that involuntary life as standing in the sharpest
opposition to reflection and its work. It then becomes, in the last
analysis, the indescribable and inconceivable, and it would become an
insoluble enigma how there can be any knowledge at all of that
``immediately given'' from which reflection is nonetheless constantly to
draw. James and Bergson represent a significant reaction against
intellectualism's misunderstanding and overvaluation of the significance
of reflection, a reaction that in several respects recalls the stand
taken by \emph{Hamann}, \emph{Herder}, and \emph{Jacobi} against
\emph{Kant}. Not pure reason, but the immediate stream of the life of
the spirit is, they maintain, the highest, and brings truth forth. The
distinctions and definitions of thought are necessary in order that we
may practically find our way in life. And analyzing science has carried
the opposition to the immediately given still further---an opposition
that language itself already favors, by forming independent words for
the individual objects and by taking spatial relations as the basis for
our intuitive images. It is as though a Fall had taken place when
reflection awoke and turned, inquiringly, toward the involuntary life
from which it had itself sprung.

Without a kinship between the involuntary mental life and reflection, it
would be impossible to understand how the latter could arise. But in the
very descriptions that James and Bergson give of what, for them, in
sharp opposition to all elaboration by thought, is the immediately
given---descriptions that are masterpieces of descriptive
psychology---it is nonetheless expressly stated that in this ``given''
there are differences and connections, qualities and continuities. James
emphasizes especially the constant transitions (continuous transitions)
and the difficulty of distinguishing between relation and term (relation
and matter related); Bergson stresses especially the undivided and yet
heterogeneous nature of the inner stream (durée hétérogène et
indistincte; durée dont les moments se pénètrent). But with this an
inner possibility for the arising of reflection has been indicated. The
whole weight must not be placed on the external causes (practical need,
the mechanics of language, and the technique of science). The motive for
the awakening of reflection is given with the opposition between
continuity and qualitative multiplicity, between the connection and the
individual elements. Herein lies the possibility of an inhibition of the
involuntary stream, an inhibition that only discursive reflection can
remedy, through the clarity and definiteness it is able to bring. As a
combining that is accomplished through comparison, reflection can bring
about clarity concerning the relation between continuity and
discontinuity, unity and multiplicity. And it is not anything wholly
foreign that it brings in by its work; for already in the immediately
given there appear, after all, likenesses and differences, connection
and multiplicity---and likewise, what is especially significant, an
involuntary striving to bring about harmony among the different
tendencies. If there were not such an original striving, it would be
unjustified to speak of a stream, of a continuité mouvante, a unité de
direction\footnote{One of my young listeners wrote to me some years ago:
``I can declare myself free of every purpose, as of every interest or
belief in anything outside or inside me. I do indeed feel something
willing, something pressing forward, but it lives, as it were, its own
life, pushes aside everything that gets in its way, and surely never
commits any act that is not to its own advantage.''---Here reflection
has brought no result, only dissolution; but the involuntary life has
maintained itself in its unity and in its definite direction, which was
all the more striking to the young psychologist the less his reflection
could attain to any firm standpoint.}! There shows itself here, then, an
analogy to what, in the case of reflection, we called synthesis---a
tendency to gather and unite. And reflection awakens when there is a
special urge for a fuller carrying-through of this tendency than is
possible within the involuntary mental life. Already within this---not
only after reflection has awakened---crises and catastrophes can arise,
and chaos can threaten. A work can be required even before reflection
can awaken; but the difficulties arise, are worked on, and are resolved
on different terms in the involuntary mental life than where reflection
has taken the helm. Perhaps some of the most significant things in our
mental life take place in the regions below the domain of reflection,
and it is often only the after-effects of these deep-lying processes
that reflection has to deal with. It will always be impossible to
demonstrate that at any point we stand before an immediately given in
the absolute sense, that is, one in which absolutely no psychic work has
been done. What we in each particular case call immediately given
deserves this name only in relation to a certain definite psychic work,
while in comparison with other objects it would perhaps have to be
regarded as the result of a work.

4. By starting from reflection and seeking to become clear about its way
of operating, we arrived at the concept of synthesis, and we then found
that this concept could also be applied to the involuntary mental life,
which forms the constant basis from which reflection springs, in order
thereupon to make that basis itself an object of its activity. We must
make ourselves familiar with the fact that the involuntary mental life,
as it unfolds prior to the awakening of reflection, can be illuminated
and designated only by way of analogy. We apprehend it only in analogy
with clear reflection, and the expressions of language are throughout
drawn from the world of reflection. So far is that involuntary mental
life from being the thing easiest to apprehend, that it takes precisely
great art to get a proper view of it in its difference from the mental
life shaped by reflection. What one calls sound common sense operates,
in its psychology, with concepts drawn from the world of reflection. A
childlike tendency makes itself felt to regard everything in the mental
life as the fruit of clearly conscious thinking, of deliberation and
calculation. One personifies, so to speak, not only outer but also inner
nature, placing a person behind the person. The childlike psychology
explains personal life itself, in its involuntary unfolding, by
supposing that behind it there is a person who consciously thinks
whatever stirs in it---just as in older works on optics one can find
depicted a man inside the head, looking at the image that the light-rays
form on the retina!

Reflection is required in order to discover and apprehend the
involuntary mental life, and yet the peculiarity of the latter consists
in the fact that reflection plays no role in it. This difficulty can be
resolved only by our apprehending and expressing the
involuntary---insofar as any memory of it persists after the awakening
of reflection---in analogy with what we know clearly from the world of
reflection itself. This is what we do when we say that synthesis is the
fundamental form, not only for comparing reflection, but also for the
involuntary mental life, and is thus the most comprehensive
psychological concept.---

But now a new difficulty arises. Synthesis means combining---but there
must then be something that is combined, and this something must be a
multiplicity. The objects toward which reflection directs itself in
order to hold them together and compare them form a coherent
multiplicity; each individual object comes into being for us through an
act of distinguishing. And in a single object we can again distinguish
individual sides, properties, or parts, which we may call elements. The
result seems then to be that we end in an opposition between synthesis
and elements, and that the multiplicity of elements is the constant
presupposition for synthesis being able to operate! From this one might
then draw the conclusion that what is original in the mental life was a
multiplicity of elements, and that only at a later point of development
did the combining activity intervene. This conception was also asserted
in its day by John Locke, and it still often haunts psychology. It
relies on the analogy with the formation of a mechanical whole out of
elements that could be pointed out beforehand each by itself, as when a
heap of sand is formed from grains of sand that previously lay each in
its own place and are only now brought together. The question becomes
whether this analogy is justified.

When reflection awakens, the mental life does not, however, begin with
it; for then it would have no objects. The elements---that is, the
parts, sides, or properties that reflection can distinguish in the
object or objects, and that it can determine through their mutual
likenesses and differences---are, after all, precisely the fruit of a
work of thought. They are drawn out of the connection in which they
involuntarily occurred, and set in a certain opposition to one
another---an opposition without which they could not become objects of
clear consciousness. And the characterization that each individual
element, and thereby each individual object, receives is conditioned by
its relation to other elements or objects. At no point is it justified
to assume isolated elements in the mental life, each absolutely
independent by itself---a kind of absolute psychic atoms. That is to
confuse the effect of reflection with reflection's own basis---at
bottom, again an example of the naive psychology that tends, after all,
to confuse differences that thought first makes with originally given
differences.

And insofar as we succeed in retaining a memory of the state we are in
when involuntariness prevails, this state appears---and here the masters
of descriptive psychology are undoubtedly right---not as a chaos of
independent parts, but as a stream, a pulsation, yet not without
differences, but such that there is a connection even between the lowest
troughs and the highest crests of the waves. We find, as far as can be
judged, unity and multiplicity so interwoven with each other that they
cannot be separated in an external manner. It was, after all, precisely
in the urge to gain greater clarity about their relation that we found
the possibility and the motive for the awakening of reflection.

At no element can we, in the domain of the mental life (any more than in
that of matter), come to a stop as absolutely isolated or absolutely
simple. There constantly poses itself, for reflection, the task of
finding a connection between the elements or objects that might seem to
be isolated, and of finding an inner multiplicity in the elements or
objects that appear as simple. It is a sign of the dullness of
reflection when it comes to a stop too early in either of these
respects. In other words: the law of synthesis must always be carried
through anew with respect to the objects or elements themselves.

On the other hand, neither can there be any absolutely conclusive
synthesis. As long as the mental life lasts, new objects are taken up
into the involuntary stream and are involuntarily worked into it. They
are new drops, or perhaps new brooks, which are partly reshaped by the
stream and partly can also gain influence on its character and its
course. To this is added the character of the new riverbeds that the
stream must form for itself when the terrain changes. And as long as
reflection is awake, it will constantly be able to set itself the task
of bringing about a greater, more comprehensive, and more intimate
synthesis than before. There is always a psychic work to be done. The
synthesis that has hitherto enclosed the multiplicity of the mental life
can again come to stand as object or element for a later synthesis. As
soon as we become conscious of our preceding mental states and
activities, such a transition to a new synthesis takes place, whether or
not this signifies a progress, a more valuable state. It is a new I that
becomes conscious of our earlier I's. No synthesis is definitive. Even
the philosophy that spoke most of higher unity always fell into
embarrassment when it had to speak of the highest unity.

It can of course be entirely justified, in a special connection, with a
view to a particular investigation, to disregard these difficulties of
principle and to regard elements or objects as if they were absolute and
simple. Every special scientific investigation presupposes a certain
abstraction. If, e.g., experimental psychology really needs to operate
with a kind of psychic atoms or flashes as independent and simple, this
can be just as justified as when, in natural science, one operates with
material atoms. But one must avoid the dogmatism that confuses concepts
we have arranged for the needs of an investigation with ultimate
realities. And likewise it can be justified that in everyday speech, or
in a practical---e.g. ethical or juridical---respect, we regard the
individual as an absolute unity with a completed synthesis, without
regard to the fact that synthesis at bottom signifies a work, an
activity, not something given once and for all. There is an irrational
relation between the mental life and what we call its elements. The
point now is to find as many elements as possible; but it can be
justified and necessary to confine oneself to a few decimal places,
provided only that one does not forget that the question is then not
exhausted. The irrational is here, as everywhere, both a barrier and a
task.

There expresses itself here an antinomy that is closely bound up with
the essence of the mental life and is peculiar to the concept of
personality. Mental life and personality cannot be conceived as products
of previously given elements, since the elements we know always have
their properties precisely by being members of the mental life. We here
come to move in a circle, unless we are clear that an inexhaustible
reciprocal action prevails: the parts are determined by the whole, and
the whole by the parts. In each particular case, in each particular
psychological consideration, we must investigate how far the synthesis,
the formation of the whole, has advanced, and on this basis observe how
the emergence of a new object operates. But the very distinction between
synthesis and object is always more or less artificial in relation to
the constant interplay of meeting and parting, of continuity and
discontinuity, of wholeness and dissolution. There is here a mental
rhythmics that forms an analogy to---or rather a prototype for---the
rhythms that characterize the course of organic life, a rhythmics that
by no means begins with reflection, but, by all we can judge, already
makes itself felt in the involuntary mental life. Indeed, this rhythmics
recurs in the very relation between reflection and the involuntary
course of the mental life. The difficulty of understanding the origin of
the mental life rests precisely on the fact that it cannot be explained
as a product of previously given elements, but from its very first
moment moves in a rhythm---and that in a rhythm of the peculiar kind
that none of the members in the rhythm can have absolute priority over
the other.

5. In the history of the concept of synthesis this antinomy has been a
stumbling-block. The concept of synthesis itself is an old idea,
traceable in Plato and Aristotle, and especially in Plotinus in
antiquity, and which from Leibniz's time on has always, with more or
less energy, been asserted in psychology and in philosophy
generally\footnote{On the history of the concept of synthesis in more
recent times, see \emph{Georges Dwelshauvers}: \emph{La synthèse
mentale} (Paris 1908), pp. 179--223 (an account of the concept of
synthesis in Leibniz, Kant, Wundt, Høffding, and Pierre Janet).---In
\emph{S. Kierkegaard} this concept appears in an energetic and profound
manner, especially in the measure he applies in his doctrine of the
stages. See my book \emph{Danske Filosofer}, pp. 153 f.; p. 165 f.}.

\emph{Hume} cut the knot by dissolving the mental life into absolutely
independent elements, and consistently he therefore found an insoluble
enigma in all unity and connection in consciousness. The principle of
unity (the uniting principle) was for him the unintelligible. \emph{Kant}
started from the opposite point of view, that it is precisely the
unconnected, the isolated and mutually independent, that is
unintelligible to us, both in psychology and wherever else we meet it.
But he still held fast to Locke's presupposition that, as the basis for
thought's combining and ordering activity, there was given a
multiplicity that was only now to be connected. Thereby he arrived at
his fateful distinction between the form of cognition and its matter.
This too was to cut the knot. And yet Kant did not lack the insight that
already in the involuntary mental life there operate forms analogous to
those that clear reflection employs. Synthesis is, he says, the effect of
a blind but indispensable function of the soul, without which we would
have no cognition at all, but of which we only rarely become conscious.
It is the task of thought to form the synthesis into a concept. And this
same blind function of the soul (Kant calls it imagination) is likewise
what conditions the connection between the two extremes of our cognition,
``sensation'' and ``understanding.'' It was surely only Kant's
reluctance to give psychological considerations too large a place in his
investigation that held him back from giving a fuller characterization
of the involuntary course of the mental life. J. F. \emph{Fries}
developed Kant's doctrine further from this side and is even regarded by
his disciples as the real discoverer of the involuntary
basis\footnote{Against this, see my review in \emph{Göttinger Gelehrte
Anzeiger} (1907) of ``Abhandlungen der Frieseschen Schule. Neue
Folge.''}. But he then believed that in this basis---the spontaneous
striving, the immediate cognition---he possessed a rule of truth exalted
above all error. Unfortunately this rule is ineffable, since it can come
to consciousness only in an imperfect and fragmentary way through
reflection! A dogmatism comes to the fore in the firm belief in the
validity of the involuntary mental life---a dogmatism that signifies a
decided step backward in relation to Kant. Already when Fries assumes
that ``spontaneity'' is not only continuous but also constant, he says
more than any reflection could demonstrate. The irrational relation
between immediacy and reflection is not strictly maintained.

This relation is, as already mentioned, very sharply emphasized by
\emph{Henri Bergson}. This thinker is not satisfied with the concept of
synthesis, because it always presupposes an opposition between unity and
multiplicity. Is it not an opposition---he asks---that recurs, whatever
one thinks of or speaks about? What matters is to say \emph{which} kind
of unity and \emph{which} kind of multiplicity we have in the mental
life. The streaming life of the personality cannot be defined by means
of concepts that always stand in a certain relation of opposition to one
another. One cannot, through thesis and antithesis, arrive at a real
unity. One must, conversely, immerse oneself in the immediate stream of
the mental life in order to see the opposition emerge from it. Immediate
beholding (intuition) must take the place of
reflection\footnote{\emph{Introduction à la Métaphysique}. (Revue de
Métaphysique et de Morale 1903, pp. 15--17.)}.

Bergson's standpoint recalls, viewed from one side, that of Fries,
except that he gives us, with far greater mastery, an understanding of
what is meant by ``spontaneity.'' But whereas Fries and his disciples
held that true cognition is attained by translating, as well as we can,
that spontaneous basis into the forms of reflection, Bergson demands that
we subdue reflection and, by an act of will, dive down again into the
immediate, feeling ourselves one with it. It stands here as an enigma
how the validity of a cognition can be demonstrated without reflection.
One of the special problems of thought will later give us occasion to
bring out this difficult point in the philosophy of the brilliant French
thinker.

In response to the remark that the concepts we use in our reflection to
characterize the mental life---synthesis, unity, multiplicity, and the
like---find application to all objects, I will maintain that the reason
why that group of concepts constantly finds, and must find, application
to all objects is to be sought precisely in the fact that they express
the forms in which the mental life, and cognition in particular, unfolds
itself. Only through the application of these forms does the
appropriation, the taking-up into the connection of our own being, in
which all understanding, all cognition consists, become possible. No
wonder that all products of cognition bear their stamp! Our cognition
consists in a conscious or unconscious analogy with our own being.---

The relation of opposition that has been discussed in the foregoing
under various expressions (object and reflection, matter and form,
spontaneity and reflection, intuition and analysis), \emph{Stumpf}
designates as a relation of opposition between phenomena and psychic
functions\footnote{\emph{Erscheinungen und psychische Funktionen}.
(Abhdl.\ der preussischen Gesellschaft der Wiss.\ 1907), pp. 15--21.}.
With respect to this relation he advances, with all possible
reservation, the hypothesis that psychic phenomena are not altered
because they are analyzed. When we become aware of something in an
observation that we did not notice at the first moment, this is supposed
to bring about no change in our observation itself, but only in the
degree of explicit consciousness we apply to it. A sound rings the same
whether or not I notice individual tones in it; a dish makes the same
impression whether I can distinguish something sour or sweet in its
taste, or whether the taste acts as a unified whole. In my judgment it
is, from the outset, not probable that this is how matters stand. The
differentiated and differentiating attention in which analysis consists
must alter the object for us. The fact that we have discovered a spot in
an otherwise beautiful picture---a spot we did not notice at the first
moment---can give the whole picture a different character. The timbre
changes; an element has, quite simply, been added that was not present
in the first glance. And the relation between the individual parts is in
each case a different one after the analysis than before. It can cost an
exertion of will to return to the immediate beholding in which the
particulars have not yet become the object of special apprehension.
There is here a difference between experience and reflection that can
scarcely be effaced. There are of course many degrees, as we shall come
to see in what follows, in the investigation of the relation between
intuiting and judging. And individual differences certainly also make
themselves felt. In elastic natures the transition from experience to
reflection, and vice versa, will take place relatively easily; and yet I
should think that even in them a certain contrast-effect will make
itself felt between the two states.

\emph{Stumpf}, for that matter, will only maintain the possibility of a
differentiated subsistence, during the experience, of what is first
differentiated with consciousness during reflection. And he concedes
that there is probably no absolute independence in the relation between
phenomenon and function ``under the usual, complicated circumstances of
the psychic process.'' Whether experimental investigations can, as he
surmises, lead to any result here is doubtful. It is not easy to see how
the difference between immediate experience and reflection could fall
away. And in a psychological experiment it is difficult to obtain a
wholly immediate experience.

Chemists often say that the atoms of the elements are present in
compound substances---so that, e.g., carbon and oxygen are really
present in carbonic acid. And Stumpf holds that the psychologist must
have the same right in his domain. I think, however, that there is a
difference. For the chemist can demonstrate that there are just as many
parts by weight in carbonic acid as in carbon and oxygen each by itself;
and at bottom that is all he means by that manner of speaking. Such an
equivalence the psychologist is, for good reasons, debarred from
demonstrating.---For that matter, a chemist such as \emph{Sir William
Ramsay} has expressly denied the right to take that manner of speaking
literally, and \emph{Ernst Mach} has expressed himself in the same
direction. The analogy is therefore doubtful.---

We do not get beyond an antinomy between experience and reflection, and
we may not, without further ado, transfer to the one what holds for the
other. Experience itself we may think of neither as absolute continuity
nor as a chaos of differences. The continuous and the discontinuous
occur in it in all possible degrees, and it is only reflection that
takes the opposition between continuity and discontinuity sharply into
view, in that it---as will later be shown---receives the double task of
finding differences where continuity was immediately given, and
intermediate links where discontinuity was immediately given.

6. In the concept of synthesis there lies, as we have seen, an
opposition between object and form, which gives rise to great
difficulties that become insuperable when by object one understands an
absolutely unformed matter. As already indicated, every object will pose
a double task for combining thought: to determine its relation to other
objects, and to find out differences and relations within the object
itself. Hereby arises the constant striving to make the discontinuous
continuous and the continuous discontinuous.

But when we do not separate thought from the other sides of the mental
life, there is yet a third element, besides form and object, that we
must bring out. When a task is posed, there is presupposed an urge to
have the task solved. The work on the solution then acquires value, since
we designate as valuable everything that satisfies an urge, whether this
urge expressly announces itself, or whether it is only from the
satisfaction we feel at the work or its result that we infer that there
has been a lack that has now been filled.

Thought has biological significance, in that attentiveness to changes in
the surrounding world makes it possible for dangers to be averted and
advantages gained. But here too the significance of synthesis shows
itself, since it is only under very elementary conditions that
attentiveness to a single change is sufficient; it will be of importance
that the single experience be brought into connection and reciprocal
action with other, present or earlier experiences; only then can the
reaction to what is experienced become entirely appropriate. Only
instinct reacts to an elementary stimulus; when reflection is awakened,
a more serious orientation, and thereby a richer relation to surrounding
existence, is possible. An interesting example of a synthesis that is
perhaps of a purely physiological kind is afforded by what
\emph{Pavlov}\footnote{\emph{The Lancet}, 6 Oct. 1906.} has called
conditioned reflexes. Salivation can be produced not only by direct
stimulation with a substance brought into the mouth, but also by
stimulation of sight, hearing, smell, or the skin-sense, on the
condition that these sense-stimulations have not occurred too long
beforehand together with the action of that substance. Here an
associative synthesis is at work, which makes it possible for one kind
of stimulation to take the place of another with which it originally had
nothing to do. Already in so simple a case as this there takes place a
combination of present and earlier states. What shows itself in this
simple example repeats itself at the higher stages of psychic
development.

Immediate value belongs from the first to the result---in the given
example, salivation (which of course, as we may here disregard, itself
again has its value as a member of the digestive process). The
sense-sensations that indirectly produce it acquire at first only
mediate value. And at more developed stages it will, similarly, be the
result of reflection that has value, while reflection itself is only a
means. But it is quite possible that a value can pass over from being
mediate to becoming immediate. There then occurs what we in psychology
call a displacement of motive.

When it is bitter need that leads to thinking, the value or motive
stands in a purely external relation to thought. The value is here
itself an object, since, as it cannot be immediately attained, it poses
the task of finding ways and means, and thereby comes to stand as a
purpose. In the definite and compelled relation between purpose and
means we have man's first encounter with the necessity of thought. If he
wills A (the purpose), he must also will B (the means). By this path man
enters the world of reflection even where there are no more direct
motives for entering it. But it can acquire an independent value, can
become an object of direct interest, to investigate that necessary
relation between purpose and means, which is, after all, the same as the
relation between effect and cause; and the value is then bound to
reflection itself and its work. It comes to stand as an activity that it
is attractive to exercise, whatever result it may lead to.

When a stage has been reached where the value or motive attaches itself
closely to the activity of thought itself, so that this is both end and
means---when, in other words, an intellectual feeling has developed, a
joy in thought itself, as it unfolds according to its own laws and
according to the demands of the objects---then we need not specially
recall motives and values in the case of cognition. We then desire no
other satisfaction than that which is bound to the use of our faculty of
thought.

It is such a stage that we presuppose in an investigation of human
thought. All purposes other than that of seeing where consistent
thinking leads then do not exist, and we need not constantly recall the
standpoint of value; this is given once and for all together with the
very urge to---observe and to infer. Even where other motives than the
intellectual interest itself actually make themselves felt in a movement
of thought, it will nevertheless be fortunate if such motives---in
relation to which the work of thought stands only as a means---pass
over, through displacement of motive, for a while into purely
intellectual motives. Thereby the work is best done. And it will in any
case always be, in the end, an examination undertaken from purely
intellectual motives that must revise the work performed from other
motives, whatever these may have been.

7. The concept of energy in the sense in which we have here taken it has
its origin in natural science, and it is by way of analogy that we have
attempted to transfer it to the domain of the mental life. All
psychological expressions are, after all, transferred from the material
to the spiritual, from the world of space to the world of spirit and
thought. But there are difficulties in carrying the analogy through. An
exact measurement is possible in the psychic domain only as regards
relatively elementary mental phenomena, and never in so perfect a manner
as in the case of material energy. Modern physiology stands, as regards
method, on the standpoint that it requires physiological states to be
explained by physiological causes, insofar as physical causes from the
surrounding world do not intervene. Every state of the brain or
brain-activity is therefore also to be explained on the basis of the
earlier states of the brain and of the organism. According to strict
natural-scientific method, every material state within and outside the
organism must find its explanation in being seen as having arisen
through the conversion into new form of the physical energy that
expressed itself in preceding material states. When two brain-states
thus succeed one another, the explanation will therefore not be found
until it is demonstrated that all the energy that expressed itself in
the first state has, without remainder, been converted into the second
state, insofar as it has not radiated out into other physiological
processes. It is perhaps impossible to carry out experiments in this
direction; but such experiments would mark the completion of the
physiological method, which can set as its motto \emph{Spinoza}'s words,
that an explanation of a bodily state by the intervention of the soul is
one with an admission that one cannot explain it. Even if now one of the
experiments one might make seemed to show that, in the arising of an
organic state, physical energy arose or vanished without a physical
equivalent, one would, from a purely natural-scientific standpoint,
sooner believe that there was an error in the experiment than rest
content with having discovered a limit to the conservation of physical
energy. ``The principle of the conservation of energy,'' says
\emph{Maxwell}\footnote{\emph{Scientific Papers}. II, p. 759.}, ``has
acquired so great a scientific weight that no physiologist would place
any confidence in an experiment that showed a considerable difference
between the work performed by a living being and the sum of the energy
received and expended.'' Since Maxwell wrote down this proposition,
experiments have been carried out which---within certain limits of
error---show that the exchange of matter and force during the
performance of mental and bodily work presents an equivalence between
what the organism took in and what it expended\footnote{\emph{Atwater:
Neue Versuche über Stoff- und Kraftwechsel des menschlichen Körpers}
(Ergebnisse der Physiologie III, 1. 1904).}. It will hereafter become
even more difficult than before to assume that in any part of the
organism something should take place that conflicts with the energy
principle.

But how, under these conditions, will it be possible to maintain psychic
energy as a peculiar form of energy? Is it not, then, more reasonable to
assume that all energy is physical, and that the psychic energy we have
spoken of in the foregoing is only a symptom of a physical energy to
which we have no direct access?---For we cannot carry through the
principle of equivalence within the psychic domain itself, as we can in
the physical domain. Interruptions occur within one and the same
individual conscious life (through sleep, faint, and so on), and between
different individual consciousnesses we cannot demonstrate psychic
continuity, as we can demonstrate physical continuity between different
bodies---e.g. between two organisms.

Accordingly, the demand has been raised that the psychological
definitions should be replaced by physiological ones. Among us,
\emph{Carl Lange} (in his ``Nydelsernes Fysiologi'') has asserted this
demand. In strict epistemological and psychological form,
\emph{Richard Avenarius}, in his acute work \emph{Kritik der reinen
Erfahrung}, has maintained a similar view. Full scientific understanding
is attained, he holds, only when the psychic states are traced back to
the corresponding physiological states: hence he calls the series of
psychic states the dependent vital series, the series of physiological
states the independent vital series. And his justification is that only
in this way can we reduce the qualitative differences that the psychic
phenomena present to quantitative ones, so that an exact formulation of
the relation between the changing states becomes possible. The
completion of our cognition will thus consist in a reduction of the
psychic to the physical\footnote{Cf. my characterization of
\emph{Avenarius}'s philosophy in \emph{Moderne Filosofer}, pp.
98--106.}.

In the face of the views that thus demand a reduction of psychology to
physiology in order for a scientific psychology to become possible---views
that thus would really abolish psychology in order to make it a
science---it must be acknowledged that the actual discontinuity and the
qualitative difference among the psychic phenomena will always set
limits to a strict carrying-through of psychology as a science. But the
question is whether the path they point out leads to the goal---whether
the method they praise does not at bottom contain a palpable
self-contradiction.

When it is demanded that the psychological definitions should be
replaced by physiological definitions, the presupposition for this must
surely be that the psychological definitions are complete and exact. But
to provide these definitions is psychology's own affair, and if it
cannot provide such definitions, physiology cannot know what it is that
it should seek to explain in the brain. When that which is to be
replaced is vague and wavering, the replacement must turn out
accordingly. There will be no certainty that one has really found the
brain-states corresponding to the definite psychic phenomena one has in
view. The independence of psychology must therefore in any case be
acknowledged, since it is to give us the symptomatology that sets
physiology its tasks. Now, for every empirical science it is a long and
difficult work to arrive at conclusive definitions; such are possible
only when the empirical science is essentially finished; they come at
the end, not at the beginning, of the investigations. All too often one
operates with unfinished psychological definitions that are regarded as
reliable starting-points for brain-physiological interpretation. Thus
\emph{Descartes}, and in more recent times \emph{Lotze}, regarded the
very concept ``soul'' as so clear and fixed that one could confidently
call upon physiologists and anatomists to search for ``the seat of the
soul.'' And in the most recent times \emph{Flechsig}\footnote{\emph{Gehirn
und Seele}. Leipzig 1896, p. 24 f.} has regarded the concept of
association of representations as so simple and clear that he could
suppose he had found special tracts in the brain for processes that are
supposed to bring about the connection between the physiological
correlates of the individual representations. It is in reality a great
abstraction that Flechsig works with when he sets the connection itself
(the association) in opposition to the elements (representations) that
are to be connected, so that the former could be thought of as localized
in places other than the latter. He thereby quite overlooks the
psychological antinomy brought out above, which warns against confusing
without further ado the differences and distinctions that the
investigation necessitates with oppositions in the unfolding of the
mental life itself. There is no ``element'' in a mental life that does
not stand in, and is not determined by, some connection or other, some
relation or other. When a representation resists entering into a
certain, seemingly natural association, the reason is to be sought in
the fact that it is a member of another, more intimate association and
cannot be detached from it. The conditions here are very complicated,
and with a purely mechanical schema neither physiology nor psychology
comes into its own.

But let us suppose that the psychological definitions are complete and
perfect. Even then great, unsolved problems will remain. How can
physiological states have psychological symptoms? How can qualitative
differences correspond to quantitative ones, and how can discontinuous
phenomena be bound to continuous processes? Far from such questions
falling away through the ``reduction'' of psychology to physiology, they
are precisely sharpened by it and appear in a more glaring form. And in
any case it stands as an unsolved enigma how the qualitative differences
and the discontinuities come into being. Even if one should succeed in
showing that two phenomena, A and B, which one has hitherto regarded as
different, are in reality identical, so that one can set A = B, one has
not thereby explained how A and B could in the first place stand as
different. To call this difference ``subjective'' does not help; it is
not thereby gotten rid of out of existence. It is, after all, precisely
the fact that there is something ``subjective'' in the world that makes
psychology both possible and necessary. It is, moreover, within this
``subjective'' that all cognition of material nature---physiology too,
and with it the knowledge of our own body, especially of our own
brain---comes into being. Cognition is itself a psychic function, both
where it appears as sensation and where it appears as thinking. By
regarding psychic states and activities as accidental appendages
(``epiphenomena''), the natural scientist would saw off the branch on
which he sits. To study this branch itself is, to be sure, not his task;
but he does well, nonetheless, not to forget where he is sitting.

When \emph{Avenarius} attempts to carry through the reduction of the
doctrine of ``the dependent vital series'' (psychology) to the doctrine
of ``the independent vital series'' (physiology), it is nonetheless
apparent through his exposition that he constantly infers from the
``dependent'' to the ``independent'' series, not the reverse. It is on
the basis of the psychological observations he has collected that he
constructs the physiological schema which for him is the proper object
of science, and one cannot understand what is said about the
``independent'' series except by means of the facts that the
``dependent'' series contains. In general it turns out that
brain-physiology is still far from developed enough to give a thorough
elucidation of the material processes to which the great, richly nuanced
world of psychic processes is in one way or another bound. What one
finds in this respect among physiological psychologists and experimental
psychologists are mere schemata, which are supposed to ``explain'' what
psychological experience presents. No one has more occasion to
experience this than a psychiatrist who is at the same time an
experienced and critical psychologist. Thus \emph{Pierre Janet} has
declared that, apart from individual observations and histological
investigations, the physiological interpretations are only translations
of the psychological concepts into another language, and that if one
will think only anatomically, one must wholly give up thinking
psychiatrically\footnote{\emph{Les Obsessions et la Psychasthénie}. I,
p. 496; 605. Similar statements by \emph{Wundt} (\emph{Philos. Studien}
XIII, p. 363) and by \emph{Ramon y Cajal} (cited in \emph{Herrlin}:
\emph{Minnet}, p. 21).}.

8. It is the discontinuity and the qualitative differences that cause
psychology's difficulties. But greater weight is often placed on them
than experience warrants. Precise reflection often leads to the
discovery that even where there seemed to be a psychically empty space,
there was in reality a psychic content given, although it was not grasped
by attention or by recognition, or was forgotten immediately after the
experience. And as regards the qualitative differences, observation will
often find more and finer nuances than those one first stopped at and
perhaps confidently declared to be utterly disparate. The continuity in
conscious life is surely greater than is often assumed, and there can
lie a scientific danger in too strongly proclaiming discontinuity and
qualitative difference as the distinctive mark of the phenomena of
consciousness. The inner connection in consciousness---the fact that the
more we immerse ourselves in, and make ourselves one with, the
involuntary course of the mental life, the more this appears to us as a
whole, a streaming---is and remains the primal phenomenon in the
psychological domain. We overlook it so easily only because in everyday
speech, and on the whole in practical life, we have most use for certain
single drops or waves in the gliding stream. We must perform an
integration when we go back from our more or less mechanical practice to
the immediately given. And even if we confine ourselves to functions
that can be exercised with clear consciousness---e.g. memory and
comparison---these very functions bear witness to an inner connection
behind the particulars. They are forms of synthesis, of the tendency to
combining and holding-together that attests the mental life's constant
striving to be a whole. The unconnected is the psychologically
enigmatic, that which sets psychology its problems. The task of
psychology is to find connection between the elements of consciousness
as well as between its different states.

\emph{Leibniz} has the merit of having energetically maintained the
principle of continuity in the psychic as well as in the physical
domain, especially by pointing to the elements that are vanishingly
small in relation to the clearly conscious states, and that only by
being summed up and combined produce a result accessible to untrained
self-observation. Already a comparison between modern and ancient or
medieval poetry shows how many more mental nuances self-observation has
now come to perceive. It is scarcely only because the nuancing in the
psychic domain is now in itself greater than in earlier times that this
greater richness is found. It is partly precisely the knowledge and
understanding of the immediate and involuntary in the mental life itself
that has grown. The childlike and primitive psychology is inclined to
conceive everything that takes place in the soul as clearly conscious
and all actions as undertaken with deliberation and intent. Now an
explicit demonstration is required that reflection and deliberation have
been present in the particular case. The transitions themselves between
the unconscious and the conscious, and between the involuntary and the
voluntary, can take place through many intermediate forms, many degrees.
The more practiced self-observation and criticism are, the more
difficult it is found to be to put one's finger on a definite place
where the boundary should lie.

Leibniz was inclined to regard all psychic differences as differences of
degree with respect to clarity and obscurity. This attempt was
characteristic of the century of the Enlightenment, which Leibniz
inaugurated. But there are other paths by which psychology can find a
continuous connection despite the qualitative differences. There
constantly take place fusions and compositions in the domain of
conscious life---of sensations, representations, feelings, and
endeavors. And displacements take place by which what at first stood in
the foreground is supplanted by what was at first background---where what
at first had only mediate value comes to stand as immediately valuable.
And in integral characters, the highest and noblest objects of
psychology, there appears to contemplation so intimate a dependence of
everything that stirs in the soul on one purpose, one guiding thought,
that we have here a causal connection that is every bit as firm and
definite as any in the physical domain. Every single representation,
every single mood, and every single impulse is, in such a character,
clearly determined by its place within the whole to which it belongs and
which it itself helps to shape.

Where, despite everything, a lack of connection shows itself in the
psychic domain, we need not, however, give up the principle of
continuity, any more than the physicist gives it up when energy seems to
have vanished in a state of rest or equilibrium. Just as the physicist
introduces the concept of potential physical energy to express the fact
that what seemed to have vanished can be released again, so the
psychologist will be justified in speaking of potential psychic energy,
supporting himself especially on the fact that psychic elements can be
reproduced after an interval in which they have not been in
consciousness. Words such as ``predisposition,'' ``possibility,''
``disposition,'' ``capacity,'' ``trace,'' ``tendency'' express the same
as the word potential energy. The psychologist is, however, worse placed
in applying this concept than the physicist in applying his
corresponding concept. In the first place, the physicist can directly
point to the phenomenon in which the potential energy has its seat---e.g.
the tensed spring or the stone resting on the roof. In the psychic
domain, by contrast, complete equilibrium (rest) means the same as
unconsciousness---hence something inaccessible to observation (whatever
it might otherwise be in itself). And in the second place, in the
physical domain one can carry out conversions that show that a quantum of
energy can be released corresponding to the quantum that apparently
vanished on the transition to the ``potential'' state. But in the
psychic domain it cannot be demonstrated in this way that predisposition
and development correspond to each other. If the concept of potential
psychic energy nevertheless has its justification, it is because it
stands as the expression of an unsolved but undeniable task, and because
it contains a summons to go ever further in observation and analysis in
order to find as much continuity as possible between the different states
of the mental life. There is a long way to absolute rest---and it might
be that it occurs as little in the psychic as in the physical domain.
Possibility everywhere subsists in an actuality, whether or not this is
accessible to us.

9. The only way in which psychology and physiology can work together
without either of them coming to break with its guiding thoughts is
provided by the psychic phenomena being regarded as symptoms of
physiological states---and, conversely, physiological phenomena being
regarded as symptoms of psychic states. We can then \emph{infer} from
the one series of phenomena to the other, although we cannot
\emph{derive} the one from the other.

It is the natural-scientific method that step by step leads to such a
conception. The law of inertia requires that every change in the state
of a material point (of rest or of rectilinear motion with a certain
velocity) be derived from the influence of another material point.
Thereby it is excluded that a non-material (that is, non-spatial) cause
can acquire natural-scientific significance. Thereby the principle is
established that a material phenomenon must have a material cause.
\emph{Maxwell} finds the experiential proof of the law of inertia in the
fact that, every time we encounter a change in a body's state of motion,
we can trace this change back to an action between this body and another
body. This experiential proof natural science is constantly in the
process of furnishing.---To the law of inertia is joined the proposition
of the conservation of physical energy, which, as already mentioned,
seems to have to be maintained also in the case of organic nature.

There are in the physics of our time intimations that the said
propositions, despite their great importance for the whole conception of
nature, are nevertheless derived from laws and relations that hold for
the ether-vibrations in which one sees more and more the fundamental
thing. Philosophically regarded, this would change nothing in the
conception of the problem of the relation between the spiritual and the
material. The vibrations of the imponderable ether are just as different
from the psychic phenomena as the movements of the ponderable atoms. The
problem remains the same, even if electrons take the place of atoms. It
is the spatial character that, for philosophy, is the decisive property
of matter; in this \emph{Descartes} saw rightly. It must not here
mislead us that physicists sometimes use the expression the immaterial
in the sense of the imponderable.---

If now this is so, there can be found no place for the mental life in the
series of physical, or respectively physiological, phenomena. The only
possibility then becomes the conception that in the mental life we have
the expression that self-observation gives us for what appears to our
senses as organic processes. Soul and body are then not two different
things or beings, but two languages in which existence speaks to us. We
can perhaps translate from the one language into the other; but we cannot
derive the one language from the other.

Since the mental life cannot be derived from the laws and forces of
material nature, we must assume that in all material activity there
cooperate, besides the properties whose presence natural science
demonstrates, other properties that are not accessible to sensation, and
in which the mental life would be able to find its explanation.
Philosophically regarded, natural science's concept of matter cannot
exhaust existence. But by the ``predispositions'' or ``dispositions''
that we presuppose in order to understand the possibility of the mental
life's arising in nature, we do not in the least shake the application
that natural science makes of what it has learned and seen.---

The conception to which we here adhere really returns to the childlike
and practical view that regards the body as the outer expression of
personality, accessible to the senses. And we thereby obtain at the same
time the best and most durable working hypothesis, in that we are
summoned to follow at once the series of psychological and of
physiological phenomena, each by itself, as far as is at all possible,
and to investigate the relation between them, the analogies and
proportionalities, as exactly as possible. It is a functional relation,
in the mathematical sense of the word, that this conception assumes
between the mental and the material phenomena; we seek, and often we
also find, that a variation of the mental states corresponds to a
variation of the bodily states, especially the brain-states, and vice
versa. And to this is in reality confined the knowledge we have of these
things. Our hypothesis (for which the most fitting name is the identity
hypothesis, since it does not assume that soul and body are two
different beings) we regard provisionally only as a working hypothesis.
Whether it gives a conclusive explanation of the relation between the
spiritual and the material, between psychic and physical energy, is
another question, which will be discussed in a later connection. Even
thinkers who reject our hypothesis as a definitive solution nevertheless
concede that it is a useful working hypothesis. Thus \emph{H. Bergson}
and \emph{C. Stumpf}. Whether this concession is consistent from their
standpoint is a question in itself. A hypothesis that has no ground at
all in the nature of things will not in the long run be suited to serve
as a working hypothesis.---

It has sometimes been thought that the enigma could be solved by
assuming (or saying) that psychic energy is nothing other than
nerve-energy. But when one does not assume that all nerve-energy is
connected with psychic phenomena, a distinction must be made between
conscious and unconscious nerve-energy---and the problem is then posed
anew. Besides, we know as good as nothing about this so-called
nerve-energy and about its relation to other forms of physical energy,
and it is of no use to give the nerve-energy to which psychic phenomena
are bound a special name. Thereby one only introduces a qualitas occulta
and rests content with it. The natural-scientific concept of energy is,
wherever one meets it, always formed on the basis of phenomena with
spatial, geometrical properties. Natural science knows energy only as
the expression of relations between phenomena in space. And we cannot,
unfortunately, form a concept of energy from which the physical and the
psychic concepts of energy could be derived as special forms. In any
case, every attempt to form such a concept would fall outside the limits
of psychology. In a later connection we shall return to the problem we
here leave\footnote{With respect to many details concerning this problem
I refer to my \emph{Psykologi}, chaps. II and III.}.

10. We have in the foregoing determined thought as psychic energy and
sought to show how, provisionally and purely methodically, we must
regard it in relation to physical energy. But there will perhaps arise
in many the question, \emph{who} it is that \emph{has} or \emph{exercises}
this energy? One reasons somewhat thus: ``Every activity must be
exercised by someone or something. We must distinguish between that which
acts and the activity itself. In particular, in the psychological domain
we must distinguish between that which thinks, feels, and wills, and
thinking, feeling, and willing themselves.''

Such a conception, which occurs not only in everyday speech but also
among philosophers, spiritualists and materialists in brotherly union,
is due partly to the influence of language's mode of expression, partly
to the influence of material analogies.

Language distinguishes between subject and predicate and as a rule
designates each of them with its own word. There is therefore formed one
word for the acting being, another for its activity. Even where it is
difficult for intuition to make this distinction between the acting thing
and the activity, language nevertheless often employs a formal subject,
as in ``impersonal'' sentences of the form ``it is raining.'' It is a
grammatical schema that prevails here, a schema one has no right to give
philosophical significance, either psychological or metaphysical. Long
ago the acute philologist \emph{Højsgaard}\footnote{\emph{Methodisk
Forsøg til en fuldstændig dansk Syntax}. Copenhagen 1752. (The section
``Register og Forklaringer.'')} impressed upon us the difference between
grammar and philosophy. ``Grammatically,'' he says, ``a word or a
discourse is considered when one looks not so much at the things that are
understood by the words, as at the words themselves, their order and
case, and at their mutual dependence or relations to one another. Whereas
a philosopher considers only the things that the words call to mind.''
This warning is constantly needed.

Because language often needs two words to express what happens when
something acts, it does not follow that we have two ``things'' before
us. And how, really, should the relation between these two things be
thought---the relation between consciousness and that which ``has'' it,
between energy and that which ``exercises'' it?

In material phenomena, which have exerted precisely the greatest
influence on language's mode of expression, we distinguish with ease
between the subject of the activity (the movement) and the activity (the
movement) itself. We have often seen the body that is now in motion
lying apparently still in its place, and we thus know it both as resting
and as in motion. When we say that the body moves, it is really these
two states that we compare. We therefore fall into embarrassment when we
discover that no absolute rest can rightly be assumed. And language
especially falls into embarrassment when one has to speak not of
``ponderable'' but of ``imponderable'' matter, since this latter is
presupposed to be in constant wave-motion. ``Ether'' can therefore be
defined only as that which undulates. It is now the analogy with
ponderable matter, and with the apparent difference between equilibrium
and motion, that operates when one distinguishes between the soul (the
I) and its activity. But the soul we know (like the ether) only as
active. What we know here is precisely psychic energy, not any resting
``being.'' The soul's being or nature is activity. It is an intuiting,
often a lazy, imagination, not a real thinking, that has the word when
one makes that distinction between the thinking thing and the thinking.
Spatial images or symbols are confused with psychological concepts. The
urge for fixed, resting images, such as material nature seems to present,
prevents one from holding fast the concept of activity or energy in its
peculiarity and its consistency. Or one fears that the mental life will
evaporate if it is not hung upon a ``something,'' a ``substance,'' as
upon a secure hook---although one is not so consistent as to ask upon
what this hook is fastened. The ``substance'' could just as well need a
``bearer'' as the activity. Just as one earlier asked who ``has'' the
energy, so one might here ask who ``has'' the soul or the I.---Thinkers
such as \emph{Fichte}, \emph{Sibbern}, and \emph{Wundt} have the merit
of having seen clearly on this point.

We can very well say what we mean by ``ourselves'' or ``our I,'' without
entering into any spiritualistic or materialistic hypothesis. Keeping to
experience, we see the mental life unfold itself through different stages
and under different forms, and we seek to describe these and to explain
the transitions between them. At every given stage of development we then
have the mental life before us with certain properties and dispositions,
and the task is then to find out what, under certain conditions, can
develop from them. (Cf. 8.) ``We ourselves'' or ``our I'' has at every
stage its content in the properties and dispositions that the phenomena
of consciousness make known, and its form in the degree and manner in
which these are combined and ordered.

Now the more the continued mental development is conditioned by the given
properties and dispositions in their mutual relation, the more energy we
ascribe to the mental life and the more active it is said to be. The
more, on the other hand, the continued mental development is conditioned
by influences that make themselves felt independently of the individual
mental life's past, the more passive it is said to be.

Just as the planets in our world-system tore themselves loose, with a
certain initial velocity and initial direction, from the great nebular
mass, so the mental life---at the point where it awakens from the night
of unconsciousness---begins by acting according to the same laws that it
later follows. It begins with peculiar dispositions that give it a
certain independence in the face of outer influences. With this agrees
the fact that the nervous system is one of the parts of the organism
that first appear with distinctness, and that from the first it occupies
a central position in relation to other parts of the organism. The
functions of the nervous system, to which the mental activities
correspond, are not mere effects of outer influence, but determined in
form and direction from the individual's first coming into being.---

In what follows we investigate, for the closer elucidation of the nature
of human thought, first the more immediate and involuntary way in which
it expresses itself in intuiting, association, and comparison (B), then
the more conscious and exact form it takes on in judgment (C).

\begin{center}---\end{center}

\section*{B. Intuition, Association, and Comparison}
\addcontentsline{toc}{section}{B. Intuition, Association, and Comparison}
\label{sec:anskuen}

\subsection*{a. Sensation}
\addcontentsline{toc}{subsection}{a. Sensation}
\label{ssec:sansning}

11. When one often distinguishes sharply between sensation and thinking,
it is because one conceives sensation as a purely passive coming-into-being
of new psychic content, while in thinking one sees a working-over of
such content. But one cannot at any point wholly separate passivity and
activity---that is, wholly separate states whose causes lie entirely
outside an individual's preceding states from such as should be wholly
due to these. There always show themselves traces of psychic work, even
in the simplest elements of consciousness that can be distinguished. We
can be more or less active---that is, the causes of the changes that take
place in us can to a greater or lesser degree lie in ourselves, in our
character and our preceding experiences; but wholly passive we never are.

If we bring out the conditions for the arising of a single
sense-sensation, e.g. a light-sensation, it turns out that the condition
for this arising is a change that has a certain degree of strength, a
certain degree of suddenness, and a certain degree of qualitative
difference---all in relation to the given state of consciousness. A
sensation arises only when the new, in the three respects mentioned,
stands in a certain definite relation of opposition to what was
previously given. Perhaps, of the three conditions, suddenness is the
most important, which from a practical biological point of view is easily
understandable, since it is of importance in the struggle for existence
to be awakened to attention for changes in the surrounding world, even
apart from the degree and character of such changes. But all three
conditions point, psychologically regarded, in the same direction. The
weaker the change is, the more slowly it sets in, and the more alike the
new and the earlier state are, the less possibility there is for the
arising of an independent sensation; hence one can rightly designate
every sensation as a distinguishing.

It is, to be sure, a distinguishing of a purely elementary kind. The
single sensation steps forth in consciousness with its definite degree
of independence and with its definite quality, and it is only indirectly,
especially by way of physical and physiological investigation, that we
can here demonstrate conditions similar to---in any case analogous
to---those that condition a distinguishing undertaken with reflection and
intent. When we set ourselves to apprehend the difference between two
things distinctly, we can do nothing further toward it than to bring
these two things into immediate proximity to each other---then the
difference announces itself involuntarily, springs forth as it were,
since each of the two things stands involuntarily with its own
peculiarity. We here produce, by express holding-together, the opposition
necessary for a distinguishing; but the apprehension of the difference we
call forth only indirectly. And thus too, in the arising of the simplest
sensations, the given changes or oppositions are only releasing
conditions. The sensation arises as if it were a result of comparison;
but the two opposed terms are here not known each by itself; it is only
the change in the state, released by the opposition, that appears as an
element in consciousness.

It is, then, a justified analogy to call the sensation a
distinguishing---and so far a thinking---and already in it to find an
expression of psychic energy. Language, which has formed itself under the
influence of reflection, has no expressions for functions so simple as
those we here confront. Linguistic usage is inclined now to set a very
great difference between the simplest psychic functions and the more
developed ones, now to read ``inferences'' and other forms of reflection
without further ado into the simplest utterances of consciousness. But by
the concept of analogy both the kinship and the distance between
sensation and thinking can come into their own.---How one conceives the
unconscious basis from which the simplest utterances of conscious life
are thus released by the power of the oppositions---whether one declares
it to be merely purely physiological processes, or whether in the
unconscious one will see a potential mental life---is in this connection
a matter of indifference. The main thing is that we already here trace
the chief form of the mental life. For that matter, it shows itself
already on a purely physiological consideration how little the
sense-sensation can be conceived as a passive effect of stimulations from
without. The physical impression traverses, on its way from the outer
sense-organ to the great brain, so many layers of nerve-cells that it can
only most improperly be said to be the same impression when it reaches
the brain as when it comes to the sense-organ. And in the brain itself
there is released a reciprocal action of a mass of cells as a consequence
of the stimulation. What emerges in consciousness as a sensation
corresponds, therefore, physiologically regarded, to a multiplicity of
processes; there is gathered into unity in it what physiologically stands
as a plurality.

12. That the sensations are thus not simple, passive effects of
stimulations from without, but can be regarded as elementary forms of
psychic energy, becomes of essential importance with respect to the
validity that can be ascribed to them as means to cognition. We build up
our world-picture on the basis of them, and yet one became very early
aware that they could not be used without further ado to form a coherent
and intelligible world-picture. Reflection must at once use them and
disregard them if it is to reach its goal. The reason for this is that
they are not only the expression of something different from our
consciousness, but also of the very nature and mode of operation of our
consciousness.

In a characteristic way, reflection's two-sided relation to sensation
appears already in one of \emph{Democritus}'s fragments, in which the
senses speak thus to the understanding: ``You poor understanding, you
fetch your grounds from us and yet want to overturn us! Thereby you
prepare your own downfall.''---At the founding of modern natural science
this paradox appears again, in that \emph{Galileo}, \emph{Descartes}, and
\emph{Hobbes}, at about the same time and on the basis of the same
considerations, set up the doctrine of the purely subjective validity of
the sense-qualities, all while the new science was nonetheless grounded
upon sense-observations. In \emph{Kant} we meet it again in his
distinction between matter and form, in that he counts all sense-sensations
to the matter and regards only the form---which, however, never occurs
separated from the matter---as that with which science has to do.

If sensation were purely passive, it could be assumed simply to reproduce
what calls it forth from without. The subjectivity of the sense-qualities
is grounded precisely on the fact that they come about as a consequence
of an activity on the part of consciousness. That there exists a problem
of cognition rests---as will later show itself---precisely on the fact
that we are self-active in our cognition. How can we, by our own activity,
cognize something different from ourselves? By our very activity we
transform it, after all, according to the laws of our consciousness.
Cognition is our work. And in this respect there is, according to the
foregoing, no difference of principle between sensation and thinking. The
paradox then disappears when one gives up the absolute difference between
sensation and thinking, matter and form. The work of reflection then
stands as a continuation of the psychic work that sensation has already
begun.

Insofar as reflection does the work of sensation over again, it is thus
not because with it a wholly new principle comes into force. It is only
psychic energy operating in another manner, more appropriate for the
solution of the task of cognition. The understanding can thus quite well
at once use and overturn sensation. And the ``overturning'' takes place
partly with sensation's own help, in that the one sense controls the
other. To the relation between the qualities within the one sense there
can correspond relations between qualities within the other sense, and it
is through the analogies that can here be demonstrated that reflection
reaches its result. The senses by which the forms of synthesis most
clearly come to light, and by which the highest degree of continuity can
be attained, are taken as basis. And it is especially the senses that
permit clear apprehension of movement. One might call them the rational
senses. To these belong sight, hearing, the sense of touch, and
especially sensations of movement. The sensation of movement takes place
distinctly by a synthesis, since it presupposes the holding-together of
the successive positions and stages of the moving thing; only by such a
holding-together does movement become an object of consciousness. The
special qualities---color, hardness or softness, sound---that correspond
to the different stages of the movement could in themselves be thought of
as exchanged for others, provided only that the relation between the
qualities among themselves continued to be the same. One can then
disregard them, and the sensations then become, when they stand in the
service of cognition, only symbols for certain stages or places in
continuous processes of movement. In a sense-sensation, with its degree
of independence and its peculiar quality, a concentration has taken
place, a combining of what, physically and physiologically regarded, is a
multiplicity. Modern physical theories interpret what for us is only the
object of a single act of sense as an aggregate of small parts in
constant movement. Our sensation appears here as analogous to a
statistical result of a mass of particulars that are each by itself
inaccessible to us and that stand in the most complicated mutual
connection. The kinetic theory of gases is especially an example of this.
According to it, every gas is an elastic fluid whose elements are in
constant movement, with numerous collisions against one another and
against the barriers, and consequent changes of direction and velocity.
For our sensation it nonetheless appears as a whole, within which we
distinguish nothing. And the same holds, as already remarked earlier, of
the relation of sensations to the brain-processes. What in our
consciousness appears as a single sensation corresponds to a very
composite reciprocal action between the millions of cells of the great
brain and between these and the other parts of the brain.

What takes place in the nervous system and in the brain does not resemble
what stimulates the sense-organ from without. And what takes place in the
brain probably does not resemble what takes place in the sense-organ, the
ingoing nerve, and the lower centers that the impressions pass through on
their way. The sense-sensation resembles neither any of these
physiological processes nor the outer stimulations that release them.

Innumerable enigmas conceal themselves here. And yet it is a fact that
sense-sensations can guide living beings in their struggle for existence.
Our states can very well be decisively determined by our own nature,
provided only that they lead to the release of activities by which we can
maintain ourselves against the surrounding world. In instinct, which is
released by definite sense-stimulations, this appears especially clearly.
The answer is given in a quite different way than the question is put. In
instinct, urge and capacity are so united that they are both released and
operate together when a certain definite change sets in in the state of
the living being. Thus when the pregnant bird builds a nest, or when the
newly hatched chick scratches if it feels gravel under its feet.

Already the first thinkers who became clear about the subjective nature
of the sense-qualities emphasized, in connection therewith, the practical
significance of sensations as the original, the theoretical significance
as the derived. Thus especially \emph{Descartes}, \emph{Malebranche},
and \emph{Berkeley}. Modern evolutionary theory has again brought out
this point of view. Sensation is first and foremost a means in the
struggle for life, only secondarily a means to come to know the nature of
existence. The connection between its practical and its theoretical
significance lies in the concept of sign or symbol. The sign can be a
signal for action; but it can also become a signal that announces the
setting-in of a certain change, apart from the action necessitated by
this change. There are many intermediate degrees between these two
significations. But in neither of the two significations is there needed
any likeness between the sign and what is to be signified. And the sign
perhaps operates best by being as simple as possible---by presenting as a
unity what is in itself a multiplicity. The main thing is that it can
serve as an introduction either to an action or to a train of thought
that can attest their validity---the action by leading to the
satisfaction of an urge, the train of thought by leading to understanding
and to agreement with new sensations. In this last case too it is a
matter of the satisfaction of an urge---namely of the very urge to
understanding, to the agreement of our reflection with as many sensations
as possible. Motive (value), form, and object thus always operate
together. (Cf. 6.)

The special qualities with which the different sense-sensations arise
nevertheless contain a great, perhaps insoluble, problem. For from a
biological as well as from an epistemological point of view, whatever
quality a sensation appears with, sensation can perform its function in
the service of life and of reflection, provided only that it is able to
release activities or trains of thought that can satisfy the practical or
theoretical need. Why precisely these definite colors arise for us when
certain electromagnetic stimulations have propagated themselves through
space to our eye, or why we hear precisely these definite tones when
certain air-waves have propagated themselves to our ear, stands as a
constant enigma. We stand here at mere facts and must for the present
remain standing at them.

Each single one of our senses has its series of qualities, more or less
continuous and systematic. Even where the continuity is greatest, as in
the domain of the sense of sight, it is nonetheless interrupted at
definite places, and the series of qualities is always limited. To
infra-red rays the retina is not receptive, and ultra-violet rays affect
it only at a particularly great degree of strength. This limitation,
which probably has its ground in certain physical or physiological
conditions, and perhaps has its appropriateness, entails a limitation of
the possibilities of knowledge.

Each sense-organ has its own peculiar advantage. The ear possesses the
fine power of discrimination by which partial tones can be recognized
within the sound in which they occur. The eye's power of discrimination
shows itself especially as the capacity to localize the different visual
impressions. In the different living beings, according to their
conditions of life, different senses show themselves especially
developed. All such special degrees of development must, just like the
limitation, be assumed to have arisen through reciprocal action between
the inner and the outer conditions of life.

When we here speak of living beings and of a surrounding world, we make
use of a knowledge that, like all other knowledge, is itself won on the
basis of sensations and by the work of reflection on this basis. It is
then presupposed that the psychic energy that already expresses itself in
sensation has developed in the form of reflection. The biological
knowledge we make use of for understanding the development of the psychic
activities is itself a product of these activities.

13. It has already shown itself above that a single sensation is an
abstraction. The character of every sensation rests on the relation
between the conditions under which it arises and the conditions under
which simultaneous or preceding sensations arise. And there showed itself
here an analogy between its arising and the discovery of a difference
effected by reflection. It was for this reason that we could call it a
distinguishing.

The psychic equilibrium must be abolished for a sensation to arise. But
there always expresses itself a tendency to equalize this disturbance.
And this tendency appears in two forms, a more elementary one, which
consists in fusion, and a higher one, which consists in the formation of
a sense-intuition.

Fusion sets in, as already indicated, when the change produced by the
impression is slight in relation to the state we are already in, or when
changes follow very rapidly upon one another. Examples of fusion we have
in the timbre of tones and in the cooperation of colored and uncolored
light-impressions in the coming-into-being of light-sensations. In a
visual image due to seeing with one eye, it is not noticed that the blind
spot on the retina gives no contribution to the filling-out of the image.
When a regular alternation takes place of light-stimulation in the time a
and complete non-stimulation in the time b, there is nonetheless
obtained, when the time a + b is very short, a continuous sensation. Even
changes that take place simultaneously in different sense-domains fuse
insofar as they mutually inhibit or reinforce one another.

It would scarcely be right to see, in such a fusion---in which the
individual elements cannot make themselves independently felt---the
original form of conscious life, so that an ``undifferentiated
continuum'' should be the starting-point for mental
development\footnote{\emph{Einar Buch}: \emph{Om Fornemmelsers
Sammensmeltning, særlig ved Klangindtryk}. Copenhagen 1898, pp. 31--35;
60 f.}. All conscious life is scarcely ever fused; there will in any case
probably stand different fused complexes over against one another. A
complete state of fusedness would be a state of rest, which would be one
with a state of unconsciousness. A certain breaking of oppositions, or at
least of differences, is necessary for the subsistence of conscious life.

When fusion is not possible, and there nonetheless expresses itself a
tendency to equalization, a sense-intuition can form itself, in that the
sensations are ordered in such a way that each single one can make itself
felt without swallowing up or supplanting the others. With the least
possible change, the new content that a sensation brings is then
incorporated into the whole connection of consciousness. The
sense-intuition solves a double task, in that it at once lets the novelty
and peculiarity of the single sensation and the connection of conscious
life come into their own. It presents at once multiplicity and unity,
while fusion abolished the multiplicity. The sense-intuition is clearest
when the differences have previously appeared each by itself and only
thereupon are connected in such a way that a single act of attention can
survey them. The clearest form of such a surveying is space-intuition,
which we therefore have a pronounced tendency to use as a schema for
everything we wish to make clear to ourselves. In particular, a
time-intuition first becomes possible through the symbolic use of
space-intuition, although we can of course quite well apprehend change
and succession---e.g. of tones or of our own inner states---without using
a spatial schema.

Sense-intuition denotes a higher stage of psychic energy than fusion. The
multiplicity is preserved through an involuntary ordering. There will
always be one or more single elements that determine the ordering, and
around which the others group themselves---namely those that correspond
to the dominant urge or interest. Therefore every involuntary
sense-intuition will be articulated, pointed in a definite direction,
that in which attention is concentrated. This determinacy of direction
forms an analogy to what later stands as purpose for striving and object
for reflection. Out of this concentration the other members in the
sense-intuition fade off into the indefinite and obscure.---

In this description we must constantly remember what was above called the
psychic antinomy. When it is said that elements fuse or become members in
a sense-intuition, the meaning is not that the elements, apart from these
kinds of synthesis, have the same character as they have within the
syntheses. It is only in our abstract language that we can let the
elements go in and out of the different connections like a kind of atoms.

\subsection*{b. Recognition}
\addcontentsline{toc}{subsection}{b. Recognition}
\label{ssec:genkendelse}

14. A pure sense-intuition would be one in which a multiplicity is
involuntarily ordered without earlier lived experiences or experiences
contributing. Apart from the most primitive states, such a pure sensation
scarcely occurs. As soon as, in the multiplicity that is ordered, there
are found elements that have previously been present, there is the
possibility of a new form of psychic energy: recognition. The simplest
example of recognition we have where a sensation, by reason of
repetition, immediately appears with a peculiar quality, different from
that with which it appeared when it first appeared. I use the expression
quality to designate the wholly immediate difference that can obtain
between the known and the new---a difference that can be as simple and
indescribable as the difference between yellow and blue. The quality that
a sensation can acquire through repetition I call the familiarity-quality.

While in the simple sense-intuition it is essentially the difference
between the elements that operates, in recognition the relation of
likeness makes itself felt. What conditions the transition from new to
known is not itself something new, but is, on the contrary, precisely a
repetition. How the familiarity-quality is to be explained physiologically
is not decisive for its psychological and epistemological character. It
is, however, most natural here to point to the influence that repetition
and practice have in altering the course of processes and functions.
Through repetition one and the same function, probably also the brain's
function, can take place with greater ease, and recognition might then be
the psychic correlate of the altered character of the functions. However
this may be, the circumstance that the relation of likeness now makes
itself felt marks a significant advance. One can already here speak of an
understanding: the present here stands immediately as one with something
earlier experienced. There is a foothold, a home for it in consciousness.
\emph{Plato} called the dog a philosophical animal, because its behavior
toward men is determined by whether it knows them or not. In this sense
newly hatched chicks too are philosophical animals, since they scratch
only when they feel gravel under their feet, not on a carpet. The
remarkable thing about instinctive actions is precisely that they are
released by stimulations that are at first new to the individual and yet
operate as if they were known.

In everyday speech we often use the word ``understand'' in the sense of
recognizing. It is an understanding of the most elementary kind. I
understand a sound when I know what kind of sound it is, that is,
recognize it. I can in this respect understand a word when I know its
sound, even if I do not know its meaning. Understanding has many degrees.
I understand what kind of figure it is that approaches when I recognize
the human figure in it; but there can still be much to understand, namely
which person it is, and so on\footnote{It is therefore not an apt
objection that \emph{Husserl} (\emph{Logische Untersuchungen}, II, p. 73
f.) makes against me, that unintelligible words too can sound familiar. In
the wholly immediate ``understanding'' of a word, it is only the sound or
the sign, not the meaning, that is known.}---

Just as in the simplest apprehension of difference, so in the simplest
recognition language lacks apt designations. But there is a significant
analogy between such recognition and the comparison undertaken with
reflection. In genuine, free comparison two independently given
sensations or representations are presupposed, whereas immediate
recognition takes place without the present and the earlier experienced
needing to appear each by itself in consciousness. Just as a sensation
can arise immediately when there is a certain opposition between the
influence of a stimulation and the state at hand, so immediate
recognition arises on the presupposition of a relation of likeness
between the influence of a stimulation and the dispositions or
after-effects that earlier stimulations have left behind. In both cases
it is only the results, not the presuppositions, that appear in
consciousness. The elementary distinguishing and the immediate
identifying are the first intimations of what at higher stages appears as
thinking in the narrower sense of the word.

15. Distinguishing and recognizing denote two poles in our cognition,
which under different forms are found again at all its stages of
development, and which correspond to two necessities of life: receptivity
for the new, the changing, and the capacity to take it up into oneself in
as intimate a way as possible. The two capacities can stand in sharp
opposition to each other. They are two different kinds of psychic energy,
a centrifugal and a centripetal one. But there is a natural reciprocal
action between them. Recognition makes possible a saving of force. If, in
the face of changes that repeat themselves, we had to go through anew
each time the experiences that the first change called forth, it would
lay claim to all our energy, and we would not get anywhere. By reason of
recognition, on the other hand, we can turn with fresh force toward the
new, and the very opposition between the known and the new lets the new,
by a contrast-effect, step forth more vividly and characteristically.

Yet recognition itself presupposes energy to perform the synthesis of the
present and the earlier. In that weakening of psychic energy which
\emph{Pierre Janet} calls psychasthenia, and which essentially expresses
itself as a lack of will (aboulia), the capacity to recognize
(understand) the present situation is often lacking. The bond that should
connect the present with the earlier is broken. On the occasion of one of
his aboulic patients, Janet says: ``The attention to the present, the
connection (synthèse) of the new sensations with the old images, which
constitutes recognition, has here, like the will, vanished on account of
the mental weakening''\footnote{\emph{Les Obsessions et la
Psychasthénie}. II, p. 29.}. Yet such weakening can also lead to false
recognition or memory-illusion; for energy is also required to
distinguish the new from the known.---

It is only an analogy that obtains between immediate recognition and
reflection. One has no right to identify it without further ado with acts
of thought performed with clear consciousness. It would thus be incorrect
to call it a judgment, as several authors have
done\footnote{\emph{Meinong}: \emph{Beiträge zur Theorie der
psychologischen Analyse}. p. 374.---\emph{Hans Cornelius}: \emph{Versuch
einer Theorie der Existenzurteile}. p. 56 f.}. For a judgment there is
required a transition of thought from a subject to a predicate; but in
immediate recognition no transition takes place between two terms; the
familiarity-quality appears as if at a single stroke. Immediate
recognition can form the basis and presupposition for a judgment, but the
judgment first comes into being when a distinction is made between the
given, present thing and the earlier thing with which it is identified.
The so-called exclamatory judgments, which we shall later come to
discuss, are the first judgments that become possible on the basis of
recognition; in them there is a predicate, even if there is no subject.
Thus when the gypsy boy in \emph{Blicher} cries: ``A dog! A hunter!''
These are immediate expressions of recognitions.

16. When it is a whole, not a single property or a composite phenomenon,
that is recognized, it will not always be necessary to recognize
expressly all the members (the parts or properties). Recognition of a
single member can be sufficient, and the familiarity-quality is then
involuntarily transferred to the other members, so that the whole stands
as familiar, or an expectation is awakened that the other members too are
present---an expectation that can be so secure that one does not seek
confirmation of it. The recognition is then partial, supporting itself
merely on a part of the whole that serves as a mark of recognition.

If the expectation leads to a continued going-through of the members of
the whole, and these, when they are sensed, correspond to the expectation,
a mediate recognition of them takes place. A person's gait and bearing
seem familiar to us, and we then expect that other properties will
correspond to them; we address him, and his expression and speech are as
we expected; they are then recognized mediately, in that the
representation contained in the expectation covers what we see and hear.

In immediate, in partial, and in mediate recognition alike, an
involuntary interpretation is woven into apparently simple observations.
It is a surplus of energy that makes us at once interpret the given and,
in virtue of the fundamental law of synthesis, bring it into connection
with previously given content. Later, renewed observation may perhaps
correct our first interpretations and expectations. But it is only by the
contribution we ourselves give that we really come to know things. If we
sat still and waited, the circle of experiences would become narrow. Now,
on the contrary, the urge to use our energy leads us to an involuntary
experimenting, which can then confirm or correct our expectations and
representations. We have here, in purely elementary form, already an
intimation of the method that developed science applies: an observation
(recognition) leads to a conjecture (expectation) or a hypothesis, which
can be confirmed or corrected by new observations. In the simplest
mediate recognition there is already described that arc from observation
through reflection to new observation in which the peculiarity of all
scientific method consists. This kind of recognition testifies to greater
psychic energy than immediate recognition: the latter is bound to the
given and present, whereas the former goes out beyond this in expectation
and representation. It is from the first the urge that drives out beyond
the given---partly the urge for an outlet for energy, partly the urge for
the attainment of something that does not involuntarily fall to our lot.---

Some psychologists have held that there is no immediate recognition, but
that all recognition takes place mediately, that is, through previously
awakened expectations and representations, and perhaps even through
express comparison. After the discussions conducted on this, it can
nonetheless be regarded as settled that there is an immediate
recognition, and that it is an interesting transitional form between
sensation and reflection\footnote{Cf. my \emph{Psykologiske
Undersøgelser} (Vid. Selsk. Skr. 6. Række), pp. 8--26. \emph{Katz} in
\emph{Zeitschr. für Psychologie und Physiologie der Sinnesorgane} XL., 6.
1906. p. 432. \emph{Herrlin}: \emph{Minnet}. pp. 128--218.---\emph{H.
Bergson} nonetheless presupposes, in his critique of the theory of an
immediate recognition (\emph{Matière et Mémoire}. p. 91 f.), that all
recognition presupposes a free representation.---Besides the kinds of
recognition discussed above (immediate, partial, and mediate) there are
several kinds. In \emph{Psykologiske Undersøgelser}, p. 35 f., I have
attempted a systematic survey.}.

\subsection*{c. Memory- and Imagination-Intuition}
\addcontentsline{toc}{subsection}{c. Memory- and Imagination-Intuition}
\label{ssec:erindring}

17. The longer the interval becomes between the arising of the
expectation and its confirmation or correction, the more independently
the representations appear---the more liberated they become, or can
become, from the impressions that first awakened them. There can
then---on the presupposition of sufficient energy---form itself a content
of consciousness with a certain independence of the supply from without, a
circle of free representations.

Just as little as the sensations do the representations subsist
independently of one another. They are members in a train of
representations. They call one another forth and connect themselves with
one another, and by reason of their mutual connection (association) a new
kind of intuition becomes possible, \emph{memory- and imagination-intuition}.
In many cases one can fairly distinctly point out the process of
association by which such an intuition forms itself; but the formation can
also take place so rapidly that a total image appears in consciousness at
a single stroke. This holds especially in inspired imagining, where nature
operates ``as if it were art,'' performs deeds that in other cases would
require conscious deliberation, if they could be performed at all. Only by
roundabout ways can one, in inspired production, gain an insight into how
imagination-intuition has come about. And the matter becomes still more
complicated by the fact that urge and interest cooperate in all memory and
imagination, and that this cooperation here takes place more hidden than in
sensation and recognition. The whole past of conscious life operates along
with it. The psychic energy is therefore of a higher degree than in the
more elementary processes.

Memory- and imagination-intuition is, like sense-intuition, a form in
which the energy of consciousness solves the task of combining a
multiplicity of elements. Both the unity and the multiplicity are to come
into their own. In memory and imagination a more or less rich content of
experience and representation is worked together into a whole, often in
such a way that a quite different ordering comes about than the original
one. Such total images have their definite directions, determined by the
interests or purposes that cooperate in their formation. Every time we lay
a plan, we make use of the past to prepare the future and transform,
often without ourselves knowing it, the given representations in such a
way that they can enter as members into the new whole. When it is
intellectual or aesthetic interest that determines the ordering, it is not
an outer end that is aimed at, but the involuntary endeavor is directed at
the intuition's becoming as clear or as characteristic as possible.

Once an intuition has been formed, there will be a striving to hold it
fast as long as possible. Under the influence of new experiences,
incitements, or associations of representations, individual members are
perhaps shed or taken up, or else the mutual relation of the members is
altered. The direction of the intuition can also be changed. Just as in
quite simple organic beings, which have no proper limbs, displacements of
the organic mass in certain directions can take place, so too the
intuition can be articulated functionally without being replaced by
analyzing and judgment. A simple example from the domain of sense-intuition
is offered by the so-called Schröder staircase figure, which by changing
accessory sensations and associations of representations, or by altered
attention, is seen now as a staircase, now as an overhanging wall. When we
stare long down into the water from a bridge, it can come to look as if we
ourselves and the bridge were moving in the direction opposite to the
water's real movement. According to changing moods or outer occasions, an
image that stands in our memory or our imagination can thus shift without
our noticing it. It can be one of our earlier experiences that is thus
altered, or the image we have formed of a poetic figure. Our memories and
imagination-images have their own history, whose individual stages it is
not always easy to trace, any more than it is easy to find the grounds for
their shifts. Their history takes place in the involuntary mental life,
right down in the ground of nature within us, and reflection can here often
only ascertain that changes have taken place, when it holds together two
mutually divergent intuitive images of the same object from different
points of time in our life. In other cases reflection can in time become
aware of the change that threatens to set in, and work to prevent it. The
astronomical world-pictures, which from the first formed themselves
involuntarily, are an example of this. Scientific hypotheses often have
their advantage in their intuitability and in the support they have in
sensation's apparently immediate testimony, and there is therefore often a
hard struggle before they are let go. To a still higher degree this holds
of religious intuitions, in which a whole host of life-experiences---experiences
of the course and value of existence---have formed themselves into great
images. Here too the most important thing---both as regards the intuition's
original coming-into-being and its alterations---takes place right down in
the involuntary mental life, although reflection can enter the lists for
the right to hold fast, despite everything, the images that have been laid
down. In the relation between intuition and reflection many individual
differences make themselves felt. In some it is feeling that holds back,
and reflection is then forced to fight for the images in which it has found
rest. In others it is the urge for intuitability that opposes reflection's
intervention\footnote{Cf. on the opposition between intuitive and
reflective natures my \emph{Religionsfilosofi} § 95.}. In others again
there is such spiritual elasticity that emotional interest and the urge for
intuiting relatively easily give way to reflection.

In animals the capacity to help oneself with functionally articulated
sense-intuition probably has preponderance over the capacity for
reflection, if the latter occurs at all except in the most highly
developed animals. The dog that runs toward a man whom it ``believes'' to
be its master will, during the running, constantly alter its intuiting,
according as the man's features become more distinct, and when these
features finally show themselves to be quite other than the master's, a
disappointment will arise; but whether the free memory-image of the master
thereby appears in its opposition to the sense-image of the stranger is
impossible to decide. There is no reason to assume a proper judgment or
inference.

What I call articulated intuiting, or functional articulation of the
intuition, is nonetheless regarded by several authors as a judgment. But
it is then in any case a judgment of a quite different kind than the
proper judgments that rest on reflection. When \emph{Gassendi} thus
ascribes to the dog that ``believes'' it meets its master a capacity to
judge, he adds: ``An affirmative judgment is, after all, nothing other
than the apprehension of a thing with an addition or a property.'' The dog
sees, Gassendi holds, the man immediately as its master, and this seeing
or representing (imaginari hominem herum) is---according to
Gassendi---the same as forming the judgment that the man is the master
(enunciare hominem esse herum); yet this judgment is passed tacitly or
potentially (tacite vel virtute)\footnote{\emph{Gassendi}: \emph{Opera}.
Lugd. 1658. II. p. 411.}. In a similar way several psychologists express
themselves. \emph{Lloyd Morgan} thus distinguishes between perceptive and
conceptive judgments. The perceptive judgment (perceptual judgment) is
what I call articulated intuiting, in that it consists in a single
feature's becoming predominant in a previously formed image. A conceptive
judgment, on the other hand, presupposes a conscious distinguishing
between individual features; only this, therefore, is a proper judgment.
What Morgan calls a perceptive judgment, \emph{Ribot} calls the logic of
images (logique des images). This expression is more fortunate, since the
image-forming, intuiting process is decidedly the essential thing in the
phenomena here in question. They belong under intuition, not under
judgment\footnote{\emph{Lloyd Morgan}: \emph{Animal Life and
Intelligence}. p. 328 ff.---\emph{Th. Ribot}: \emph{L'évolution des idées
générales}. p. 32 ff.---\emph{Hobhouse} (\emph{Mind in Evolution}. p.
200) ascribes to higher animals a capacity for ``practical judging.'' But
he adds cautiously that if animals have no proper [free] representations,
they must have something that can perform a corresponding function.}. The
expression articulated intuition seems to fit best, since the intuitive
image need not be dissolved because a single element is strongly brought
out. If a new articulation within the intuition is not possible, a new
intuition can form itself, within which the changes come into their own.
Possibly the earlier intuition then vanishes entirely from consciousness.
If, on the other hand, it is held fast or reproduced, a comparison and an
analysis can get going, which may perhaps lead to a judgment.

The articulation stands as a preparation for, or a presage of, an
analyzing and judging. Thus what at later stages of development becomes a
special organ is, in embryonic development, intimated in bulging parts of
the still uniform and continuous tissue.

18. In itself one might suppose that the intuition-formations, with their
shifts and articulations, could be continued to infinity, without any
judging conditioned by reflection being necessary. It must be taken as a
simple fact that this is not so, and that---at least in human beings---a
necessity for reflection and judgment sets in. The equalization and
equilibrium that can be attained through intuiting does not always show
itself to be sufficient. Conditions can arise that make the combining have
to take place with clearer consciousness than happens in the involuntary
and half or wholly unconscious process by which the intuition comes into
being. In sense-intuition, recognition, memory- and imagination-intuition,
it is really only the result that appears clearly in consciousness, while
the formation-process itself takes place predominantly below the threshold
of consciousness.

Reflection is in a certain sense not productive. It only brings to clear
consciousness what is already contained in the involuntarily formed
intuitions. It compares and tests, chooses and rejects, connects and
separates, fills out deficiencies and determines exactly the mutual
relation of the elements. The new that reflection brings and expresses in
concepts, judgments, and inferences is the clear consciousness of the
individual parts of the content of consciousness and of their mutual
relation. Through the purgatory of reflection it can soon show itself that
we know more than we thought we knew, in that there can lie in the content
of consciousness unsuspected presuppositions for orderings and
inferences---soon that we know less than we thought we knew, in that
incompatible elements or invalid connections are discovered.

Judgment is the form in which the content of consciousness can come into
equilibrium anew, when reflection has caused a dissolution of the
intuition. In a judgment different elements of consciousness are brought
into definite and conscious connection with one another. Insofar as the
appearance of judgment is a sign that the wholeness and equilibrium of the
intuition no longer subsists, it can be regarded as a sign of an
imperfection, especially since the single judgment cannot express the
whole fullness that can lie in an intuition. Perhaps an infinity of
judgments is required for the whole content of consciousness, with all its
nuances, to come into its own. \emph{Goethe} thus sets ``das
zersplitternde Urteil'' in opposition to ``die stille Fruchtbarkeit'' in
immediate intuiting, and \emph{Wundt} even defines judgment as the
dissolution of a composite representation. But judgment in reality signifies
a new formation of a whole, which has become necessary because the tension
within an involuntarily formed intuitive whole has become so strong that a
dissolution must set in. It has a similar advantage over intuition as
intuition has over fusion: both the differences and the likenesses in the
content step forth more distinctly than before. The formation of judgment
is therefore a higher expression of psychic energy.

There must of course be a close connection between the judgment and the
intuition it builds on. One often expresses this by saying that the
judgment ``lies in'' the intuition, or that the intuition is a potential
judgment, as Gassendi said. The difference between intuiting and judgment
is that the latter lets the elements step forth independently, so that a
distinction can be made between ``subject'' and ``predicate,'' while in the
intuition a more continuous connection prevails. When I believe I know a
person who approaches, I see him immediately before me as the one I take
him to be; but if I feel an urge for a closer testing, I institute a
comparison that distinguishes between the figure that approaches and the
memory-image of my acquaintance, and thereafter there is the possibility of
a positive or a negative judgment.

But the way from intuition to judgment can be long and toilsome, and it is
not said that the ``possible'' judgments can become actual. There can be a
logical possibility for a judgment without a psychological possibility
being present. Thus it is said to Peer Gynt: ``We are thoughts; you should
have thought us!''---The logical possibility is present when the intuition
contains elements that can be separated out, and whose mutual relation can
receive a closer determination. But this separating-out and determination
is undertaken only when there are sufficient conditions for an abolition of
the whole equilibrium of consciousness that the intuition denotes. Here, as
at the first arising of consciousness, there must take place an
interruption of the uniform state, in order for a further-reaching mental
work to get going. A special interest must attach itself to individual
elements, so that an urge arises to become clearly conscious of their
mutual relation. Thus, e.g., if an intuitive image contains something that
could become a means to the attainment of a purpose, or serve for a better
understanding of another intuitive image, or form a contrast to it.

The judgment must correspond to the intuition, since it is to express its
content in another way, more appropriate under the given conditions. But
this does not exclude that it is an essential change that takes place in
the transition from intuition to judgment. When one does not embrace the
psychological atomism that I have combated above, it must be clear that the
elements of consciousness undergo essential changes by passing into new
connections, and the connections of judgment can be quite other than those
of intuition. The elements are not indifferent to the connection in which
they stand. In the intuition an element stands in a single special
connection; the judgment brings it into connection partly with elements
that lie within the same intuition, but perhaps at great distance, partly
with elements from other intuitions. In the intuition I dwell, e.g., on a
man's action as a total image and am absorbed in the mood this awakens in
me; in the judgment ``he has acted nobly!'' I make use of a standard with
which I more or less consciously compare the action. The validity of the
judgment presupposes only that from the relation established between subject
and predicate one can infer back to the intuition that lies at the basis,
so that the judgment can again lead to a construction of a similar
intuition in those who hear the judgment without having had the intuition
at first hand. There must be analogy, but not necessarily identity, between
judgment and intuition.

\subsection*{d. Association of Ideas}
\addcontentsline{toc}{subsection}{d. Association of Ideas}
\label{ssec:association}

19. The content to which judgment corresponds does not always appear in
the resting form of intuition, and as a surveyable whole. Judgment can be
formed out of a series of associations of representations, in which
intuitive images of greater or lesser extent replace one another more or
less rapidly. The difference between association and intuiting is only a
difference of degree. The more the successive gains the upper hand over
the simultaneous, the more association---which in its higher degrees
becomes a flight of thought---steps in place of intuition. What intuition
is in static form, association is in dynamic form. Already in the
formation of memory- and thought-intuitions associations cooperate, and
no less is this the case in shifts and articulations, as well as in the
dissolution of the intuition.---But we now regard association as an
independent form of the involuntary conscious life.

Just as involuntarily as an intuitive image can emerge in consciousness,
just as involuntarily can one representation draw another after it, this a
third, and so on. In both cases laws operate that do not themselves come
to consciousness. The articulating intuition stands closest to
association; but in it too the dynamic is still subordinate to the static,
since the shift quickly leads to a resting image.

The transition from association to judgment takes place through the
relation between the members of the association-series becoming the object
of distinct consciousness. The urge for this will already arise from
practical interest, as when the preceding members stand as means in
relation to the following ones. Where merely practical interest prevails,
one will attend to the preceding members only insofar as they indicate
means to the attainment of an end, and the whole order in which the
members are brought will be determined no longer by their involuntary
sequence, but by their relation to the end. It is, as \emph{Hobbes}
remarks, the thought of a purpose (cogitatio finis) that makes us, in the
waking state, order our representations in a definite and coherent way,
while in dreams, where the firm thought of purpose is lacking, they can
unfold more chaotically. For a thoroughgoing ordering there will be
required holding-together and comparison with other association-series and
intuitions than those involuntarily setting in, and the more such series
and images thus interact, the greater the psychic energy that comes to
light.

Every attempt to order, by way of reflection, the members in the
involuntarily arisen association-series consists in a testing of the
members' relations of likeness and difference. Even if the objects a and b
have occurred together ever so often in my experience, so that a has an
irresistible tendency to call forth the representation of b, the
justification of expecting b when a has set in will nonetheless depend on
whether the a that earlier occurred as predecessor of b is identical with
the a that has now set in, and further on whether it can be shown that a
cannot set in without being followed by b. The greater the continuity
between a and b, so that they really form a whole, the more justified the
expectation will be. If it is an association by likeness that we have
before us, the transition from the one member ($a_1$) to the other ($a_2$)
will be the more justified the more the likeness approaches identity; it is
then merely a transition from one example to another. Thus association
approaches judgment the more the transition from object to object is
determined by continuity and identity. In the genius, the formation of
judgment can proceed as if it were an involuntary association-process,
since the psychic energy is released with great ease and in great fullness.
Reflection is here limited to a minimum. In everyone, practice in a
definite train of thought can operate in the same direction. Where
reflection can be spared, it is spared.

20. Association of representations is, just as much as reflection, an
expression of psychic energy, and it is wrongly that one has regarded it as
a purely passive process and set it in absolute opposition to ``thinking.''
What happens in every association is really that a whole, a larger
connection, is built up, constructed out of a single given object that
occurs in a certain isolation. There are indeed always several objects or
elements that cooperate at the springing-forth of an association-process;
but a single one will be the decisive one, and for simplicity's sake we can
keep to it. Consciousness has both an urge and a capacity to go out beyond
the single object or element and connect it with others. In this instinctive
tendency the psychic energy expresses itself. All association can be traced
back to a transition from part to whole; what we call association by
likeness and by contiguity are only limiting cases. In every association
relations of likeness and of contiguity operate together, in such a way
that now the one, now the other has preponderance\footnote{See more
closely my \emph{Psykologi} V B., 8.}. The urge for the release of energy
is, in the domain of the life of representation, so strong that a minimal
occasion can operate as a release. There is here an analogy to the urge for
movement. Now the incompletely given is filled out (by predominant
association by contiguity), now one circles around it through dwelling with
kindred representations (by predominant association by likeness). In both
ways the single element can be incorporated into the whole psychic world,
and its relative isolation can be abolished. Which of the different paths is
taken in the particular case---and which special relations of likeness or
contiguity come to operate---nonetheless does not depend on purely
intellectual conditions. Like a hidden weight, interest can press the
scale-pan down for a single definite relation of likeness or contiguity
that in itself would not be decisive. It is ultimately man's striving that
determines his associations of representations, even though these, once
they have got going, can in turn alter the strength and direction of the
original striving.

When now association is replaced by judgment, the relation is reversed. It
is no longer a transition from part to whole that takes place, but a
transition from whole to part, that is, an analysis. Association is, like
intuition, an integration, but judgment presupposes a differentiation. In
the association-process it holds that a little tussock can upset a great
load; in the formation of judgment a needle is found in the great load. The
associations can bring forth the whole content of consciousness, but the
formation of judgment draws out a single thread from the rich web.
Association is productive, the formation of judgment is critical.

Yet neither is this opposition complete. The associations are---as already
remarked---not determined exclusively by a single element. This is seen
already from the fact that one and the same representation does not, under
all conditions, awaken the same associations in one and the same
individual. And on the other hand the formation of judgment presupposes
that attention is directed toward a single element's relation to other
elements (in or outside one and the same association-series), and so far it
too can be said to begin at a single point, with the little tussock.

Yet another opposition has been found between association and judgment, in
that the former is supposed always to be able to be continued, since there
is no inner limitation for it, while the latter is supposed to denote a
definite conclusion. ``Association,'' says \emph{W.
Jerusalem}\footnote{\emph{Die Urteilsfunktion}. p. 79; 82.}, ``is never
concluded in itself. Ever new members attach themselves, ever new
deviations from the original direction take place, and nowhere is there
found a natural conclusion lying in the very nature of association ... In
judgment a process is released from its connection with the rest of the
content of consciousness; it is isolated and, as it were, concluded for
consciousness. We are for the present done with it, and precisely this
concluding moment in the act of judgment forbids regarding it as an
association.'' Against this it must be objected that an association-series
can very well find its natural conclusion, namely when the motive that has
called it forth is fully satisfied. One can also reach a representation that
perhaps lies far from the starting-point, but that offers an independent
interest and therefore takes attention captive. When one takes into account
the interest, which is never lacking in the association of representations,
an inner conclusion of this can very well be given, not to mention that the
relations of likeness and contiguity that are at one's disposal in the
particular case will always be limited. On the other hand, judgment does
indeed denote a settling of accounts, a decision conditioned by reflection,
but only a provisional decision, since new reflection can correct it or
else bring it into connection with other judgments and thereby unfold what
``lies in'' it.

\subsection*{e. Comparison}
\addcontentsline{toc}{subsection}{e. Comparison}
\label{ssec:sammenligning}

21. The transition from intuition or association to judgment takes place
through reflection, which consists in an analysis, a successive turning of
attention to the individual objects and elements in the intuition or in
the association-series. Thereby free comparison becomes possible, in
distinction from the elementary comparison in sensation and the bound
comparison in recognition.

There lie two objects before me on the table, which I can see at a single
glance. But although they are both within my field of vision, I can fix my
attention now on the one, now on the other, move back and forth between
them without necessarily moving my eyes. The one object does not fall out
of the field of vision because attention is directed toward the other. By
such alternation of attention an experiment is made, whose result is that
the two objects stand either as similar or as different. The likeness or
the difference springs forth for us as a peculiar quality, an element that
was not present as long as we regarded the objects each by itself without
holding them together. It is this holding-together that brings it about
that the new property we call likeness or difference comes forth. The
older English psychology (especially \emph{James Mill} and \emph{Stuart
Mill}), which laid the whole weight on the individual representations each
by itself, could not explain how comparison was possible. For them,
relation-representations differed from other representations only in that
they presupposed two ``things'' instead of one.

The active element in comparison---apart from the energy that holds
together the elements that are compared---is one with attention. This is a
successive concentration toward the different members of the intuition or
the association, a concentration that precisely by reason of its
alternation lets the content between the elements step forth distinctly.
Experience shows that we, all else being equal, apprehend a successive
relation more distinctly than a simultaneous one.

If the objects we compare are not given simultaneously, the comparison
comes about through the one member being reproduced while the other is
directly observed. Memory thus provides the simultaneity within which
attention can then swing.

If, at the passage of attention from the one object to the other, no
difference at all appears---apart from the exertion involved in the
alternation of attention itself---the relation between the two objects
stands as a \emph{relation of identity}. Identity in the strictest sense
is the likeness an object has with itself. When different objects are said
to be identical, the meaning is that it is as if attention returned to the
same object; and conversely, that an object is the same means that
attention can return to it without any difference appearing.

If it is emphasized that the two objects, despite their lack of
difference, nonetheless occur at different times or in different places or
both, their relation is called \emph{Sameness} (\emph{Congruence}).

If some of the two objects' parts or properties appear as identical or the
same, others as different, we get a composite \emph{Likeness}. It has
sometimes rightly been held that all likeness---apart from
identity---was of this kind.

If a likeness is present without there being identity or sameness at any
point, we can call the relation \emph{Qualitative Likeness}. It is a
peculiar kind of qualitative difference. The remotest degree of likeness
is \emph{Relational Likeness} (\emph{Analogy}), where the likeness does
not concern any single property or part in the two objects, but the
relation between properties or parts in the two objects.

22. Comparison can take place involuntarily or voluntarily. Comparison is
involuntary when no intention, no preceding representation, no wish to find
out likeness or difference makes itself felt. Voluntary comparison
presupposes that one has set oneself as a task, as a purpose, to find
whether there is likeness or difference between two objects. The transition
from involuntary to voluntary comparison is formed by such cases in which
the one object is reproduced, so that it is merely held together with the
other. It is a transition reminiscent of the transition from immediate to
mediate recognition. In mediate recognition a representation is released,
which thereafter covers a following sensation. An analogy in the domain of
will is formed by the transition from urge or instinct to drive (when by
drive we understand an urge connected with a representation of its goal).
And here there is really more than analogy; for there is an urge to compare
just as to act, and in both cases it can be of importance that one becomes
conscious of what one feels an urge for.

Now it is not always the case that both the objects to be compared are
given or can without further ado be called forth in memory. Often only the
one object is at one's disposal, and the task is to find another object
that stands in a definite relation of likeness or difference to it. The
more clearly and distinctly this relation can be represented, the greater
the possibility that the new object can be found. The most exact
indication of that relation we have where the task can be put into an
equation. We then have a definite standard to decide whether the sought
object appears or not, and by a methodical procedure we can perhaps force
out the object that satisfies the demand. In a more indefinite way
something similar happens when we wish to call forth something in our
memory and know of it only that it stands in a certain connection with a
representation already present in us; by concentration of attention on
this, or, as it has been called, by a boring, the sought thing can then
spring forth. The psychic energy appears quite especially in such cases.
The point then is not merely to connect given members, but to find a
member that is perhaps quite new, but that stands in a familiar relation to
a given member. Where we can indeed determine the relation to the given
member, but on account of the limitation of our experience cannot find the
new member, there we end in a problem. Thus we are placed, e.g., when we
assume that in the psychic domain there is something corresponding to the
difference between potential and actual energy in the physical domain; for
of potential psychic energy we can form no positive representation; we can
only say that it is related to actual psychic energy as potential energy is
related to actual energy in the physical domain. (Cf. above 8--9.)

In all comparison an urge makes itself felt, even if the finding-out of
relations of likeness and difference is not directly set up as a purpose.
It is an urge for clarity and distinctness, for the use of the combining
capacity that constitutes the innermost essence of consciousness. Often,
and especially under the most elementary spiritual conditions, it is a
downright urge for self-preservation that leads to thinking. When
something that has value for the subsistence of life does not immediately
fall to our lot, we must make it a purpose and strive to find the means for
its realization. In the relation between end and means, man meets for the
first time with necessity: if he wills the end, he must also will the
means. The urge for the end begets the urge for the means. Will and thought
here go together. This meeting has not merely practical significance.
Practically regarded, it is a roundabout way that man here comes to make;
but this roundabout way points to a definite unavoidable connection that
can become the object of an independent interest, a joy of cognition. In
this way the intellectual interest can, through displacement, develop out
of the practical one (Cf. 6).

When it is a matter of determining the relation between two objects or of
finding an object that stands in a definite relation to a given object,
several possibilities will often make themselves felt. Several different
possible relations or objects emerge, and a choice must then be made,
which is determined by which of them best satisfies the definite
conditions. The choice is determined by a comparison between what is
required and what offers itself. If the two objects are given, the point
is the fixing of the relation that takes place between them; if only the
one object and the relation to the sought object are given, the point is
the fixing of which object stands in the definite relation to the given
object.

As far as possible, comparison leads to an Either---Or, so that there is a
choice between two possibilities. The other possibilities are rejected by a
preliminary comparison, or they show themselves able to be subsumed under
one of the two main possibilities. We stand here before a fundamental
peculiarity of our thinking. This moves forward through two-membered
relations. A predicate must either be ascribed or not ascribed to a
subject. No concept can at once contain and exclude a certain mark. A
grounding of this peculiarity of our thinking cannot be given. It is
evidently bound up with the psychological fact that attention, during
reflection, can be concentrated only on one element at a time, and that
comparison takes place through the passage from one element to another.
\emph{George Boole} has called the logical principle springing from this
necessity \emph{the Law of Duality}. Before it even the most pronounced
practical interest must bow when it wishes to take reflection into its
service.

Boole shows in an interesting way the purely factual character of the said
feature of our thinking. Starting from the fact that the combination of a
concept with itself has no logical significance, he finds the nature of our
thinking expressed in the equation $x^2 = x$, which is satisfied only by
$x = 1$ or $x = 0$. Here 1 means All, 0 Nothing. The derived equation
$x(1-x) = 0$ expresses the Law of Duality. A combination that should at
once contain a concept and its negation is invalid. But if now the
fundamental equation had not been of the second, but of the third degree,
our thought-activity would be otherwise constituted than it in fact is. We
would then not work our way forward through two-membered relations,
holding together two elements at a time and affirming or rejecting their
likeness (or difference), but through three-membered relations,
trichotomies instead of dichotomies---a procedure of which we can form no
real representation. Boole by no means wishes, by this presentation, to
teach that the fundamental method of our thinking is accidental or
arbitrary, but only that it must be taken as a simple
fact\footnote{\emph{Boole}: \emph{An investigation of the laws of
thought}. London 1854. pp. 48--50.}.

Already earlier \emph{Salomon Maimon} had shown that, on account of the
unity of our consciousness, what is to be connected by our thought in a
definite way can immediately consist only of two members. Every immediate
comparison that leads to a definite connection concerns two immediately
consecutive members. A mediate connection can take place only after a
series of immediate connections; thus a mediate connection comes about
between a and d when there has successively taken place a connection of a
and b, b and c, c and d\footnote{S. \emph{Maimon}: \emph{Versuch einer
neuen Logik oder Theorie des Denkens}. Berlin 1794. p. 73 f.}.

By Maimon's and Boole's remarks, the law of the excluded middle between
two mutually negating assumptions---already set up by
\emph{Aristotle}---is illuminated with respect to its connection with the
whole nature of our reflection.

23. Besides the object there are, as already said several times, in all
reflection two moments that stand in a constant relation of reciprocal
action: the form of the thought-work and the motive that awakens
reflection and drives it forward.

A motive or an interest that could not be satisfied by the laws and forms
that human thinking must by its nature follow will in the long run be
without intellectual significance. Such a motive can perhaps become
fruitful for imagination, for the capacity to form images and symbols, but
it would be a misunderstanding to attempt to force thought onto paths it
cannot by its nature walk. Fortunately, ``reason'' cannot be taken captive
forever; it will with inner necessity burst its prison, and its captivity
will not be of benefit to the interests that were to be served.

The laws and forms of reflection are the same, whether the thought-work
takes place involuntarily or voluntarily, half unconsciously or with full
clarity about goal and ways. These laws and forms operate, as we have seen,
in analogous forms already in sensation, in memory and imagination, and in
association of representations. That the thought-work, the comparison,
takes place ``with will'' entails merely greater definiteness and firmness
in the synthesis and greater clarity in the transitions from member to
member. However significant the transition from the involuntary to the
voluntary can also be, it nonetheless does not mean that a wholly new
``capacity'' should come into force.

One of the older psychologists who has laid most weight on the
will-element in thinking, \emph{Charles Bonnet}, has stated the
proposition: L'attention est la mère du génie. But however decisive
concentration in a single direction is, it nonetheless gives only the
introduction, the impulse. No less important is it that, as a consequence
of this impulse, a series of intuitions and associations is involuntarily
released, which can then become objects for reflection. In the content of
consciousness thus emerging, new elements and new connections can come
forth that show themselves able to pass the purgatory of reflection, and
genius often consists precisely in the richness and depth of the
involuntary intuitions and associations. \emph{Kepler} and \emph{Fechner}
are examples of inquirers whose guiding thoughts first came forth in the
form of imaginative combinations, which through critical testing led to
significant results. Such minds unite the freedom and freshness of the
involuntary with the exactness of attention and comparison. Genius can
finally also express itself in the indefatigability and carefulness with
which the working-over of the objects at hand takes place. Regarded from
this side, \emph{Buffon} was right when he defined genius as great
patience.

We easily come, in our psychological theories, to emphasize the difference
between the involuntary and the voluntary more strongly than close
observation warrants. The transition itself from involuntary to voluntary
comparison of course takes place involuntarily: the representation of a
goal that is the condition for voluntary striving---the one that
\emph{Dante} called ``the will's first thought''---must itself first arise
by reason of involuntary association. Here can lie a significant
turning-point in the life of thought and in the whole development of the
mental life; but it does not justify assuming a break in the continuity.

\section*{C. Judging}
\addcontentsline{toc}{section}{C. Judging}
\label{sec:dommen}

\subsection*{a. The Formation of Judgment}
\addcontentsline{toc}{subsection}{a. The Formation of Judgment}
\label{ssec:domsdannelse}

24. It has already been mentioned several times, provisionally, how
intuition and association can be replaced by judgment. Judgment is a
higher solution of the task that is solved in more elementary forms
through fusion, sense-intuition, memory- and imagination-intuition, and
association. Such a higher form becomes necessary when neither the
likeness nor the difference between the objects fully comes into its own
in those forms.

Judgment is a synthesis, a connection of objects that, prior to the
judgment, have occurred in one or more of those simpler forms. The
connection can only be clear and definite when the objects are each by
itself clear and definite. When we become clearly conscious of an object
(or a group of objects), so that our representation of it does not vary
in the different cases in which it appears, we have formed a concept. So
far it might seem that concept-formation would have to precede the
formation of judgment. But a closer consideration will show that there is
a constant reciprocal relation between the formation of judgment and the
formation of concepts, or rather, that these two processes are at bottom
not different.

Concepts as well as judgments are formed by analysis of intuitions or
associations. When we become clearly conscious of an object that is given
in a single state, through attention's successive turning to its
different elements (all that can be distinguished in the object), we form
a concrete individual-concept; when we become clearly conscious of an
object that has appeared in a series of different states, we form a
typical individual-concept, in that a single state comes to stand as an
example of the whole series; when we hold together and compare a whole
series of objects and can regard a single object as an example of the
whole series, we have formed a general concept. Between the different
examples of one and the same typical or general concept there need not be
identity, only analogy. It is only in certain definite respects
(indicated in the concept's definition) that the examples can take the
place of one another. In place of the example there can step a schema, a
symbol, or a word. They are directions for finding examples, when one
wishes to save the time that the more detailed representation of such an
example takes. The main thing is that one keeps, in one's attention, to
the elements that can approximately recur in the different objects or in
the different states of the single object, even though such elements by
themselves alone would not be sufficient to constitute a content of
representation. But it is then really by a series of judgments that
concept-formation takes place. In order to form concepts we must connect
elements in one or more objects in a clear and definite way. It is by
judgments that we establish that an object presents certain elements---that
the same object occurs in different states---and that different objects
present the same (analogous) elements.

We seem then to move in a circle: judgment presupposes concepts, and
concepts are formed by judgments! But the matter is that concept-formation
and the formation of judgment are really one and the same
process---reflection's analysis of what is given in intuition or
association. The difference lies merely in the purpose and direction of
the analysis. Now it is a chief interest to find out the elements that
the objects contain and gather them into a whole, and then we speak of
concept-formation; now our interest is essentially directed at becoming
conscious of the relation between different objects or states of the same
object, and then we speak of the formation of judgment.

Insofar as a difference can be made between the two processes, they stand
in constant reciprocal action. The clearer the concepts can become, the
clearer too can their connection become in the judgment; and the clearer
a connection can be shown through judgments between a concept and other
concepts, the more perfect the concept itself becomes.

The difference that can nonetheless always be made between concept and
judgment is a new example of the antinomy that---as earlier shown---lies
in the nature of consciousness: synthesis presupposes elements, but these
are only what they are by being comprehended by the synthesis. Hereby is
explained the constant unrest in the life of thought, and in the mental
life as a whole; ever new tasks are set: now that of expanding the content
of consciousness, now that of bringing about a more intimate connection
within the content.

25. The connection that judgment brings about is due to a recognition.
Two concepts can be connected into a judgment only when the one can be
found in the other. The one concept then denotes a point of view from
which the other can be seen, a closer determination of it. Here again it
is seen that thinking is comparison. When I say ``A is B,'' I must have
found B in A. This re-finding can have different degrees or forms. It can
show itself that A and B have wholly the same content, so that it is only
due to limited or inexact apprehension when they earlier stood as
different concepts. Or B can be a single one of A's marks, hence be a
means of recognizing A (by a kind of partial recognition). Or the
connection of A and B presupposes that account is taken of a larger
connection within which A and B belong, so that only under certain
conditions, which cannot be derived from A or B alone, can one find a
relation of identity; it is then not A and B that are identified, but
$A_x$ and $B_x$\footnote{This tripartite division is due to \emph{Jevons},
\emph{Principles of Science} (1874), Chap. 3.}. The two first-named
relations might seem to belong to Kant's ``analytic'' judgments, the
third to his ``synthetic'' judgments. But the critique of this Kantian
division of judgments has shown that no real boundary subsists between
them. Even in absolute identity (A = B) there will always be certain
presuppositions and limits for the identity. The difference between the
three named kinds of relation will in particular cases depend on how far
back one goes in the series of presuppositions.

When, with \emph{Leibniz}, one conceives judgment as the expression of a
relation of identity, its kinship shows itself clearly with the forms and
processes that form its psychological basis. Like these, it is a form in
which consciousness seeks unity and harmony in its content, and it is,
like them, a stage in the constant breaking of likenesses and
differences. It lets both the relation of likeness and that of difference
come into their own to a greater degree than the more elementary forms
were able to.

Judgment itself can enter as a member in reflection's continued work.
Inference is related to judgment as judgment is to the concept. Judgments
can be connected into inferences when identical concepts can be inserted
from the one judgment into the other. The relation between premises and
conclusion is analogous to that between subject and predicate in the
judgment, and that between intuition and concept. At all stages reflection
can lead only to a higher degree of consciousness of what is contained in
its objects. In the concept it becomes conscious of the intuition's
content, in the predicate of the subject's, in the conclusion of the
premises'.

\subsection*{b. Obstacles to the Formation of Judgment}
\addcontentsline{toc}{subsection}{b. Obstacles to the Formation of Judgment}
\label{ssec:hindringer}

26. No judgment is formed when two intuitions or two members in an
association-series replace one another in such a way that the first object
has vanished when the second steps up. If the dog mentioned earlier,
which ``believed'' it was going to meet its master, had wholly forgotten
this ``belief'' when it turned out that it was a stranger who came, it
would feel no disappointment and would have no occasion for any judgment.
And even when the conflicting objects, the representation of the master
and the sight of the stranger, meet and suddenly replace one another, so
that a disappointment is felt, it is not said that this leads to a
judgment. (It would have been the negating judgment: ``it was not my
master.'') The contrast, the collision, can constitute the whole
experience. Experience shows that one must to a certain degree have freed
oneself from an experience in order to be able to convert it into a
judgment. It is more than rhetoric when \emph{Shakespeare}, both in a
comedy and in a tragedy, puts utterances in this direction into the
mouths of his characters. ``Silence is the perfectest herald of joy: I
were but little happy, if I could say how much,'' says Claudio (Much Ado
About Nothing II, 1). ``The worst is not so long as we can say, `This is
the worst,''' exclaims Edgar (King Lear IV, 1). What in memory, or in the
mind of the outside spectator, forms itself into a judgment need not, in
the experience itself, be formed so. It is a frequent psychological error
that one reads judgments into where the immediate experience laid claim
to the whole energy of the mind. We have met this error already in
immediate recognition, which one has also wished to conceive as a
judgment. Disappointment can, like every relation of contrast, be
formulated as a judgment, but need not entail one.

Far sooner than entailing a judgment that expresses what has happened,
disappointment will---when it does not lead to complete slackening, but
when there is energy still to spare---lead to a more or less impatient
demand, a claim that the expected shall come. The collision between
expectation's rich image and impoverished reality begets the demand. In
children the demand appears not only before the judgment, but also before
the question. Or rather, the question is at first a demand. The purely
theoretical question always presupposes a certain development---a
cooling-down.

The question can be a determination-question, in that a whole stands
before consciousness within which there is a part that is not filled in.
There is lacking, e.g., a time- or place-determination for an event. Here
the answer consists in a factual indication, the bringing forward of an
object by which the deficiency is filled. In the decision-question, which
presupposes a higher stage of development than the determination-question,
the opposition between two mutually exclusive total images makes itself
felt, and the answer must be either Yes or No. (Is he dead? Is it ten
o'clock?) Yet the difference between the two kinds of question is not
absolute; for in the decision-question too it is a matter of the
filling-out of a whole, namely a whole that comprises both of the two
mutually exclusive wholes. In the question ``Is he dead?'' it is a matter
of whether we are to think of the world of the living with him or without
him. Only, this larger horizon here stands more in the background, and
that comprehensive world can in turn have different extent.

Negation too appears first in practical form; it has its first germ in
aversion, anger, and fear, and appears as rejection. A dim urge lies at
the basis both of negation and of the demand. A transition between
practical and theoretical negation is formed by the triumphant or
despairing ascertainment of what is lacking.

Negation itself, like disappointment, presupposes a transition from an
object with a manifold content to another object in which something of
this content is lacking, and also a holding-fast of the first object after
the second has set in. AB is replaced, e.g., by A, and AB and A are held
together. The negating judgment ``B is not here'' ascertains the
difference. It springs immediately from the opposition between the two
objects. In simplest form the negative judgment, like the positive, can be
expressed by a mere exclamation: ``Gone!''

Probably positive judgments arise before negative ones. Consciousness
will at first be more occupied with an increase of its content than with a
diminution of it. The transition from A to AB brings a fullness that can
be enjoyed or become a motive for action. For a judgment to appear, it is
not necessary that the preceding object (A alone) be preserved in memory;
consciousness revels in B's entrance, the new positive thing. If, on the
other hand, the transition from AB to A is to beget a negating judgment,
AB must be held fast in memory, since it is B's absence that is to be
ascertained. All else being equal, a negative judgment therefore requires
greater energy than a positive one. Energy is required to miss and to
grieve, and to demand and to wish. There is lacking here the help that the
positive B provides when A is replaced by AB.---The relation becomes of
course more complicated when, after the transition from AB to A, A gets a
new companion, so that AX appears. What the consequence will be will
depend on what relation X stands in to the earlier content of
consciousness---whether AX can become an equivalent for AB or not.

We have here an example of a small series in which the order of the
members is not indifferent. The transition from A to AB is of another kind
than the transition from AB to A. The two processes that lead to positive
and to negative judgment can be equally simple, but the psychological
motives for positive judging can be present earlier than those that lead
to negative judging.

The state or experience that gets its expression in a negative judgment
can very well in itself have a positive character. Lack is a positive
state; it is only its cause that we express by a negation. There are no
negative states at all. The negative numbers denote, after all, merely a
quantifying in the opposite direction from the quantifying that the
positive numbers denote. The Hindus denoted what we call the positive
numbers by a word meaning fortune, the negative numbers by a word meaning
debt. The representation of fortune must, from the nature of the case,
precede the representation of debt. Life leads to positing before it leads
to negating. A logical parallel to this psychological fact is offered by
the circumstance that a negative judgment, as regards content, presupposes
a corresponding positive judgment, just as a negative concept presupposes
a corresponding positive concept. ``Alle Begriffe der Negation sind
abgeleitet,'' says \emph{Kant}. The person born blind does not know what
darkness is, the ignorant person does not know what ignorance
is\footnote{\emph{Kritik der reinen Vernunft}$^2$, p. 603.}. A negating
judgment gives us no new content, but rejects a positive content, which
is thus presupposed known. A negative judgment can therefore, logically
regarded, not be placed on a par with the corresponding positive judgment,
although it can, as we have seen, build on just as simple a relation of
opposition as the latter. It stands, on the other hand, on a level with
that kind of positive judgment which rejects a possible negative judgment,
and which is often expressed by means of more or less solemnly
asseverating words (``truly,'' ``really'') or by strong emphasis on the
copula. Such affirmative judgments are a peculiar kind of positive
judgments. \emph{Hegel}\footnote{\emph{Wissenschaft der Logik}. II. p. 76
f.} wrongly held that narrative sentences which do not reject a doubt
about the correctness of what is narrated do not deserve the name of
judgments. A narrative judgment will, when it is more than idle
repetition, be the fruit of an analysis of events or reports, hence of a
work of thought, even though it does not stand in opposition to a doubt.

Therefore \emph{Sigwart} and \emph{Lotze} (and in antiquity
\emph{Menedemus of Eretria}) have rightly denied the negative judgments a
place among the logically simple or primary judgments. Sigwart calls the
negative judgment a judgment about a judgment, namely a judgment whose
content is that another judgment does not hold\footnote{\emph{Sigwart}:
\emph{Logik}\textsuperscript{1} I. p. 119; 260.---\emph{Lotze}:
\emph{Logik}. p. 61.---Cf. also \emph{Bergson}: \emph{L'évolution
créatrice}. p. 312.---On Menedemus see \emph{Diog. Laërt}. II., 135.}.
When \emph{Hermann Cohen} does not agree with this\footnote{\emph{Logik
der reinen Erkenntnis}. Berlin 1902. p. 89.}, it is because he overlooks
the difference between positive judgments in their simplest form and the
special kind that can be called affirmative judgments; with these, not
with the former, the negative judgments must, logically regarded, be
paralleled.

27. The formation of judgment can, as we have seen, be hindered by the
memory of one of the objects falling away, or by not all the objects
necessary for the completion of the judgment being given, so that only
demand or question are possible. It can further happen that a judgment is
formed only to be rejected, so that negation prevails. There is yet
another way in which a judgment can be hindered from asserting itself in
consciousness, namely when there also appears another judgment with which
it shows itself incompatible. A problem then arises. The word problem
really means ``that which is laid before,'' ``task.'' In this general
sense a problem can already be said to be set when closer determinations
are lacking. The demand and the question thus already set problems, in
that they set the task of finding determination and decision. In a
narrower and sharper sense the problem arises when two judgments---either
two positive judgments, or a positive and a negative, or two
negative---stand as incompatible.

One might call the one kind of problems completion-problems, the other
kind liberation-problems. In the first kind the sting of the problem lies
in the incompleteness of our world of thought; in the second kind the
sting lies in a discord within our world of thought. Common to both kinds
is the relation of tension (vital difference) described by \emph{Avenarius}
in his natural history of problems\footnote{See my book \emph{Moderne
Filosofer}. p. 100--104.}, which arises through a psychic equilibrium or
harmony being abolished.

Psychic energy is required to have problems. Both lack and discord subsist
only when there is a capacity to hold oppositions together. If one lets go
the one opposed term, neither the incompleteness nor the contradiction is
felt. There are problems in all spiritual domains, and it is through their
purgatory that spirits are tested.

In this connection we keep to the second kind of problems, that due to the
collision between judgments. Where a simple positive and the corresponding
simple negative judgment stand over against each other, the rejection of
the one will, according to the Law of Duality, be one with the acceptance
of the other, and vice versa. B must either be a mark of A or not be one.
Where two positive or two negative judgments stand against each other, the
relation will not always be so simple. The relation can be many-membered,
so that both judgments can be false and it is a third that holds. The
rejection of a hypothesis thus does not necessarily lead to the acceptance
of a definite other hypothesis, when more than two hypotheses are
possible. The task must in such cases be to reduce the different
possibilities to two, between which the choice must then be made.

Often the choice is made on negative grounds. One becomes clear that one
cannot go the one way, and only therefore does one go the other. Necessity
is then expressed negatively. In the history of logic this shows itself in
that the negative logical principles, the principle of contradiction and
the principle of the excluded middle, were set up before the positive one
(the principle of identity). Our criterion of reality often gets a
negative expression: we cannot but believe this or that! The primitive
ethical commandments have a predominantly negative form: ``Thou shalt
not---.'' It has been the task to set up barriers within which life could
unfold. The commandments are the tribe's reaction against those who
deviate from the paths that have shown themselves salutary. Indignation
and disapproval express themselves before recognition and approval. It was
above all a matter of preventing encroachments. The prohibitions were of
course motivated by positive purposes that were to be promoted; but the
negative measures stood as most necessary.

Often one uses, out of politeness, considerateness, or melancholy,
negative expressions to cover over the crass and sharp character with
which a positive reality can announce itself. One then denies the opposite
of what one means. It is the so-called litotes. ``It is no joke.'' ``I
cannot praise him.'' ``It is not going well.''\footnote{\emph{Kr. Nyrop}:
\emph{Ordenes Liv}. p. 47.} When \emph{Luther} at Worms said: ``I can do
no other,'' he thereby not only indicated that he had tried all
possibilities, but at the same time expressed his resolve in the most
considerate form. His conviction received at once the strongest and the
gentlest expression.

28. Only those negative judgments are fruitful which are necessary
intermediate links for arriving at a positive judgment. But it follows
precisely from the conditions of our cognition that we can often reach a
positive determination only through a series of negative determinations.
``Empty'' space was at first a purely negative concept; one could ascribe
no physical properties to it. Later such properties, especially electrical
and magnetic ones, have been demonstrated. When one says that an organ
dies away through ``non-use,'' it is because one stops at a preliminary
observation. What in reality takes place is that when the function falls
away, the chemical processes in the tissues stop, and the organ then
becomes an inactive mass that is absorbed by the neighboring tissues.

The history of classification shows us interesting examples of how we
often have to content ourselves with staking off a domain, in that we can
say nothing else about it than that it is different from another,
definitely and positively characterized domain. It then becomes the task
of continued inquiry to find positive marks for the domain thus staked
off. \emph{Aristotle} divided animals into rational and irrational, the
irrational again into those that had (red) blood and those that did not
have (red) blood\footnote{Yet Aristotle did not base his classification on
a single mark. He precisely combats dichotomy. \emph{Gomperz}:
\emph{Griechische Denker}. III. p. 117.}. \emph{Linnaeus}'s classification
of animals seems to support itself on positive marks. He has six classes:
Mammals, Birds, Amphibians, Fish, Insects, and Worms. But \emph{Cuvier}
pointed out that the sixth class is characterized only negatively. ``It is
well known,'' he says, ``that Linnaeus united under the designation
\emph{Worms} extraordinarily numerous and differently formed animals, for
which it was difficult to indicate any common mark. While I was working on
my first treatises on comparative anatomy, I came to face the
impossibility of saying anything universally valid about the worms'
nervous system, or about their circulation, or about their organs of
respiration and reproduction, indeed not even about their organs of
digestion. Thereby it became clear to me that this class was not, like the
others, founded on positive marks.''\footnote{\emph{Cuvier}: \emph{Sur un
nouveau rapprochement à établir entre les classes qui composent le regne
animal} (1812).}

The inductive method often moves, like classification, forward through
negations, in that different possibilities or hypotheses are set up and
tested, and only those that are not rejected under the testing are held
fast. \emph{Kepler} arrived at the assumption of the elliptical form of
the planetary orbits first after he had rejected various other forms. It
was the ``method of exclusion'' that led to his result.

Between the significance of negation under induction and under
classification there is the difference that the negative paths under
induction mean paths that are barred, while under classification they mean
paths that are left for later inquiry to travel. Perhaps our cognition
ends finally at a crossroads where it can get no further, neither as
classification nor as induction. We would then stand at a conclusion of
the process that began with the first, primitive distinguishing. Through
the forward-driving force of differences and oppositions, lacks and
contradictions, ever higher forms of psychic energy have been developed,
until the limit is reached where perhaps pure speculation or poetic
analogizing still develops thoughts about the nature and worth of
existence, but where experiential confirmation of our judgments is no
longer possible.

That negation plays so great a role during this whole development comes
from the fact that we have to orient ourselves in existence from a
definite standpoint in time and space, and from definite presuppositions
that are given with the very nature of thought. The richness of existence
is too great for it to be surveyed at a glance or by a single progressing
series of intuitions. On the other hand, thought can see more
possibilities than experience shows us realized in the existence we know.
Now it is our all too limited points of view, now our all too
comprehensive ideas, that must be negated or at least corrected, before
true cognition can be attained.

\subsection*{c. Subject and Predicate}
\addcontentsline{toc}{subsection}{c. Subject and Predicate}
\label{ssec:subjekt-praedikat}

29. A wholly definite point where judgment replaces intuition or
association cannot be demonstrated. There is a multiplicity of transitions
in the degree of attention that is directed toward the relation between
the elements in an intuition or in an association-series. The boundary is
formed on the one side by the immediate and involuntary streaming-along,
in which neither objects nor processes are distinguished, and on the other
side by hypnotic or ecstatic absorption in a single element. Between these
two boundaries lie a series of states that contain more or less favorable
conditions for the arising of a judgment. The decisive thing is that,
however much attention is directed toward a single member, the connection
with the whole intuition or the whole series in which it occurs is
nonetheless preserved, so that its relation to the whole can stand clear.
Continuity is the constant presupposition for a fruitful application of
attention in a single definite direction. It does not go, as older
presentations of logic might mislead one to believe, that one first knows
the individual members each by itself and then knows the relation in which
they must stand to one another when they are to constitute a whole. If
there were not beforehand a whole that enclosed the members, it would be
inexplicable how thought could come to connect them. What thinking can do
is only to bring the relation between the members of a given whole to
clear consciousness, and this may perhaps require a rearrangement of the
immediately given relation between them.

One has often thought to find a definite mark of distinction between
judgment and the simpler mental processes we call intuition and
association in this, that judgment is expressed in words. But the
significance of the word for thinking consists essentially in the help it
provides toward holding fast the typical and general in its difference
from the concrete and the individual. Herein it is presupposed that a
comparison of the objects in question has beforehand been instituted. And
such a comparison is also necessary in order to decide the right to
transfer a word from one object to another. The judgment is expressed by
the word, but does not owe its coming-into-being to the word.

The word is, in its simplest form, an exclamation in which the state of
mind called forth by an object gets air. An exclamation common to several
individuals can become a means of communication. And this means of
communication can in turn become a means of thought, a means of holding
fast a thought in its limitation, in particular of holding fast the
difference between points of likeness and of difference within a circle of
objects. Hereby the word comes into so close a connection with thinking
that it is no wonder it has been difficult to keep the grammatical, the
psychological, and the logical side of the nature of judgment apart from
one another. The psychology of cognition and logic have developed under
the constant influence of grammar and its forms. Especially as regards
judgment, one has transferred relations and forms from the nature of the
grammatical sentence, determined by the laws of language, to hold
immediately for the thought-activity itself. This has especially had
influence on the conception of the relation between subject and predicate
in judgment, a relation I shall now attempt to illuminate.

30. The logical predicate is the most important member in the judgment. It
denotes the new that comes to be added and that is pointed out. The
subject is always presupposed. All words therefore begin as predicates,
and it is only later that they are used to express subjects. It is at first
the regard for the hearers that leads the speaker to express the subject
in words: for the speaker himself his object, the subject, the thing at
hand, the situation, stands as given; he sees in each case the object that
is the starting-point for his thought and speech before him in spirit; but
this is not always the case with the hearers, and therefore he must
indicate that about which he wishes to say something.\footnote{Cf. on this
\emph{Ph. Wegener}: \emph{Untersuchungen über die Grundfragen des
Sprachbaus}. Halle 1883. p. 54.} There are, however, sentences in which
the initial representation, the logical subject, is not expressed at all.
This does not mean that the subject is lacking in the speaker's
consciousness, but only that there is no urge or time to express it. The
interest rests on the predicate. The logical subject is often hidden, and
yet is that about which everything turns. It is the pivot on which the
door turns;---but it is the turning, not the pivot in itself, that is the
new thing.

Such sentences, in which the logical subject is not expressed, we can call
predicate-judgments. Different kinds of them can be distinguished. The
exclamation is the shortest and most forceful linguistic expression for
the fact that something new in relation to the thing at hand or preceding
has set in. No special initial representation (terminus a quo) is formed
at all, if consciousness is immediately occupied by the final
representation (terminus ad quem). Small children are especially occupied
with what moves. A little boy of a year and a half said, when he saw an
animal coming, ``Bow'' (from bow-wow), and the same word he used when he
saw something move, e.g. when the tall grass on the lawn where he played
moved in the wind. It was clearly enough the movement that laid claim to
his attention, in contrast to the preceding, relatively resting state. The
resting and unchanged called forth no outburst\footnote{W. \emph{Stern}
(\emph{Beiträge zur Psychologie der Aussage. Erste Folge}. Leipzig 1903.
p. 52 f; 124 ff), by showing children of seven years pictures, has found
that what first interests them are persons and things, later activities
and actions, last relations and properties. These experiments have, like
so many psychological experiments, the defect that they do not show us the
purely involuntary. Furthermore, they concern children who already have a
certain development and command many memories. And finally the striking
opposition between rest and movement does not come forth. ``Activity and
action'' cannot be presented as strikingly in a picture as they are in the
experience. ``Substance'' is not, as Stern holds, the child's first
``category.''}. The exclamation ``A star!'' when a star is suddenly
discovered in the bright evening sky is a predicate-judgment. Likewise the
exclamation ``A good idea!'' when a happy inspiration puts an end to an
embarrassment. In naive narration the initial representation is often
presupposed known, in that the narrator, absorbed in his subject, hastens
straight to the predicate and forgets to acquaint readers or hearers with
the situation. In \emph{Herodotus} several delightful examples of this are
found (I, 144; V, 35; VIII, 21).

Often what one feels an urge to express is only the value of what happens.
It can be done by a single word (``Splendid!'' ``Bravo!''). What is
splendid stands as given and presupposed, and one has neither desire nor
time to express it. In such exclamations, which contain the germ of the
proper value-judgments, is expressed an object's intervention in the whole
inner state of the speaker, and the logical predicate is here the property
ascribed to the object as a consequence of this intervention; the object
itself is not expressed.

Different from the mere exclamations are the so-called impersonal
sentences. Here too only the logical predicate is expressed. There are two
kinds of such.

There are impersonal, subjectless sentences in which the subject is either
named beforehand or emerges from the connection in such a way that one at
least knows of what kind it is, and can construct it in more or less
definite form. In \emph{Kaalund}'s poem ``Nissernes Vandring'' it is
described how the nisses leave a region and in great throngs swarm over a
bridge; ever new flocks come: ``it was as if it would never come to an
end.'' What seemed as if it would never come to an end was the swarming of
the nisses. The indefinite impression of the endless is emphasized by the
sentence's having no definite grammatical subject. In \emph{Schiller}'s
poem ``Die Erwartung'' the lover waits for his beloved; every sound and
every sight can announce her coming; it is only a question of whether the
interpretation of what is heard and seen is correct. ``Hör' ich nicht
Tritte erschallen? Rauscht's nicht den Laubgang daher? . . . Glänzt's
nicht wie seidnes Gewand?'' By the indefiniteness of the grammatical
subject, the phenomenon comes to step forth in all its immediacy. And yet
there is an initial representation; for only on the basis of the definite
expectation do the rustling of the leaves and the bright gleam have
interest. In other cases the initial representation is more indefinite.
Only its kind can be represented. E.g. ``There is dancing,'' ``Someone is
knocking.'' Often one here uses the indefinite grammatical subjects ``one''
or ``they.'' ``One says'' = ``it is said.''

In another class of subjectless sentences it is not grammatically possible
to indicate any definite subject. A wholly indefinite representation lies
at the basis. These are the impersonal or subjectless sentences in the
narrower sense. They are used to express natural phenomena, indefinite
states, the course of fate. E.g. ``It is cold.'' ``It is spring.''
``Things are going backward.'' ``It goes up and down.'' The chaotic and
total character of the initial representation shows itself in that one can
often, instead of ``it,'' put ``everything.'' E.g. ``It is so still'' =
``Everything is so still.'' The word ``it'' is, in the opinion of the
philologists, used in impersonal sentences only to fill an empty place in
analogy with complete sentences---a grammatical counterpart to the way in
which one often reads the forms and functions of developed consciousness
into the elementary psychic phenomena.

Psychologically regarded, an initial element is never lacking (even if it
does not become a distinct initial representation), and logically regarded
one will always be able to construct such a terminus a quo and thereby
translate the most primitive exclamations into logical judgments. But the
initial element can be vanishing in relation to the final element.
Hypnotic and ecstatic states denote the culmination in this respect.

If one defines a judgment as a conscious and definite connection of
concepts, this definition does not fit predicate-judgments, in which the
subject-element lies wholly in the background of consciousness and is not
represented distinctly according to its whole content, as we require of a
concept. But that definition can hold as an ideal. Logic can say that it
really has to do only with the fully developed judgment, in which both
subject and predicate are completely determined. The predicate-judgment
and other judgments in which this is not the case then become only
approximations to judgments. And of such approximations there can be
extraordinarily many, since there can be many degrees in distinctness and
definiteness.---

Predicate-judgments express acts of recognition. The application of the
predicate contains a referring of the object to other objects, a kind of
classification. The simplest cases of recognition permit no special
designation. The object can stand as ``known''---but this is also the only
predicate at one's disposal. When immediate recognition is translated into
the form of judgment, the classification consists in the object's being
reckoned to the known in contrast to the new. In contrast to this there
can be predicate-judgments where ``unknown'' or ``new'' becomes the only
predicate, in that a closer description would already presuppose
recognition of certain features in the new. In the very simplest cases one
can only point.---

The more many-sided a relation the predicate-content is brought into with
the subject-content, the greater the role played not only by acts of
attention and recognition, but also by further-reaching associations and
comparisons. If the predicate's relation to the subject is to be
determined exhaustively, the consideration must not be limited to the
immediately given, but this must be held together and compared with as
many other objects as possible. Greater and greater demands are then made
on the psychic energy.

31. Also in judgments in which both subject and predicate are expressed,
the logical predicate continues to be the most prominent element, while
the subject can stand in different degrees of obscurity. It is on this
relation between the obscure or indefinite and the clear or definite that
one must fix attention when it is to be decided what is logical subject and
what is logical predicate. One cannot always see it from the single
sentence, but must often infer it from the whole connection.

Often the logical predicate is placed first in a sentence for emphasis'
sake, while the subject follows toneless afterward: ``Great is Diana of
the Ephesians!'' ``Now the time is up!'' ``Felix qui potuit rerum
cognoscere causas!''---The logical predicate is often the grammatical
subject or an adjective belonging to it: ``\emph{Thou} art the man!''
``\emph{All} the guests have come.''---One will always be able to
recognize the logical predicate by the emphasis, whatever place it
otherwise occupies in a grammatical respect: ``The king is \emph{not}
coming!'' ``He \emph{has} gone.''---In questions---both determination- and
decision-questions---it is always a logical predicate that is lacking and
sought. One has the terminus a quo and seeks the terminus ad quem. When we
go back to the chief questions that lie at the basis of all other
questions, we come to the fundamental predicates, the so-called
categories. \emph{Kant} aptly called the categories, the fundamental
concepts of thinking, ``predicates of possible judgments.'' To think, he
says, is cognition by concepts, and concepts determine, as predicates for
possible judgments, something that hitherto stood
undetermined\footnote{\emph{Kritik der reinen
Vernunft}\textsuperscript{2} p. 94; 105 ff.}.

One and the same subject can be successively determined by a series of
predicates. Indeed, in sentences of a describing or summarizing kind
almost every single word can express a logical predicate. E.g. ``A poor
peasant went in to his neighbor to ask him to stand godfather to his
newborn son'' (the beginning of \emph{Tolstoy}'s tale ``Gudsønnen''). We
here get a whole series of new things to know at once.---

32. If now a predicate is attached to a subject as a closer determination
of it, then a new, more complete concept has been formed than the
subject-concept one began with. The earlier predicate has now become an
attribute. The attributes of existence (fundamental properties or
fundamental forms) are found by analysis of experience and of the objects
it presents. It is also this path that \emph{Spinoza} has gone in his
doctrine of attributes, despite his system's apparently deductive
character. He finds in experience objects of spiritual nature and objects
of material (spatial) nature, and since these two characteristic marks are
irreducible, he regards them both as attributes or predicates of that
which lies at the basis of all existence
(``Substance'')\footnote{\emph{Ethica}. II. Ax. 5.}. And so it goes in
each single case; we constantly make a transition from a predicate found
by analysis to an attribute. If I discover that a man is just, then I can
for the future operate with the concept ``this just man.'' The new concept
thus formed can then in turn become the initial representation for a new
formation of judgment, and so on.

Of special interest here are such sentences in which the grammatical
subject appears with a quantitative determination attached, whether this
determination is expressed by a number or in words (all, some, one). If
the emphasis lies on this determination, it is in reality the logical
predicate. In the sentence already used above, ``\emph{All} the guests
have come,'' the new thing, hence the logical predicate, is ``all''; the
formally logical formulation would thus be: ``the guests who have come are
all who are expected.'' So it goes with all quantitative judgments, where
the quantity is the main point. \emph{Maimon}, \emph{Herbart}, and
\emph{Sigwart} have therefore rightly maintained that there is no ground
for setting up such judgments as a separate kind of judgments.---Something
corresponding holds for the hypothetical judgments, just as we have
already seen it for the negating judgments.

\emph{Benno Erdmann} polemicizes against this and maintains that in a
judgment such as the one cited the quantitative attribute belongs to the
logical subject and is not predicate; one could otherwise, he holds, with
the same right make all attributes predicates, which would be a forced
reinterpretation of the linguistic mode of expression\footnote{\emph{Logik}
I. p. 325--329.}. But precisely the linguistic mode of expression
distinguishes between two different cases here. Where the tone lies on the
quantity-determination, this is terminus ad quem, the final
representation. And where the quantity-determination is not emphasized, it
must have been won by an earlier judgment, so that we in any case have a
secondary judgment before us.

\emph{Husserl} concedes that all attributive designations mediately or
immediately have sprung from judgments. But he holds that with these
earlier judgments we have nothing more to do\footnote{\emph{Archiv für
systematische Philosophie}. 1903. p. 114.---\emph{Logische
Untersuchungen}. II. p. 438.}. This is now in a certain sense correct. But
the main thing is whether a certain kind of attributes (e.g. the
quantitative) entail acts of judgment of a certain peculiar kind. And it
is this that must be denied. When Husserl praises Benno Erdmann because he
has ``the courage to follow the natural meaning of the words'' (Wortsinn),
---and when he adds: ``Ich halte es mit ihm für gewiss, dass 'alle S sind
P' ein rein affirmatives Urteil ist ... das 'alle' gehört zum
Subjekt,''---then it must be answered that one cannot at all pronounce on
the sentence ``all S are P'' without knowing the emphasis; for this too
belongs to the ``word-meaning'' (Wortsinn)---in any case it belongs to the
meaning (Sinn)!

It is remarkable that two such acute thinkers as Erdmann and Husserl can
still discuss the logical significance of the sentence without taking into
account either the emphasis or the whole connection in which the single
sentence occurs, and to which one must go back in order to see where the
tone is to be laid, that is, where the logical predicate lies. Every
skilled reader-aloud knows this. And it was long ago stated in the old
\emph{Art de penser} (Port-Royal's Logic) that only by attending to the
connection in which a sentence occurs can one decide what is the logical
subject and what is the logical predicate: L'unique et véritable règle est
de \emph{regarder par le sens} ce dont on affirme, et ce qu'on affirme.
Car le premier est toujours le sujet, et le dernier l'attribut [i.e.\ le
prédicat], en quelque ordre qu'ils se trouvent (II, 9).

It makes no difference in this connection that judgments with
quantitatively determined subjects acquire a quite special methodical
significance on account of the great importance that counting, measuring,
and weighing have for research. This has nothing to do with the judgments'
purely logical and psychological character. Here it is a matter of not
letting oneself be misled by differences in the linguistic expressions
into assuming differences of kind with respect to the logical significance
of acts of judgment as well as of judgments.

33. It will now be time to bring out the difference between the
psychological and the logical consideration of judgment. Logically
regarded, it is really only the relation between the judgment's two
members that matters, while psychologically regarded it is the process by
which thought reaches from the initial representation to the final
representation that matters. Logically regarded, it is ultimately
indifferent with which of the two members we begin, provided only that the
relation between them stands clear and definite. But psychologically
regarded, we in each particular case start from a certain representation
and pass from it to another; the former becomes in the judgment the
logical subject, the latter the logical predicate, however the judgment
may be expressed purely grammatically.

No object can become logical subject without an interest attaching to it.
If there are wholly interestless representations, they glide without
effect through consciousness. The interest attached to an object makes the
representation of it into an initial representation and operates as a spur
to finding a closer determination of it. A certain tension arises before
the final representation is reached, and this tension is replaced by the
interest in the predicate found, which is therefore emphasized in the
linguistic utterance. The interest in the subject begets the interest in
the predicate. The whole formation of judgment would be superfluous if the
interest in the subject were not inhibited by its indefiniteness. The
mystic, who revels in his one, constant thought, forms no judgment; and
his object would---on account of its unity and lack of difference---only
be violated by any predicate whatever. And he who is absorbed in intuiting
or in associative imagining will also seek to keep judgment at bay.
\emph{Goethe} thus says in his ``Italienische Reise'': ``Ich halte die
Augen nur immer offen und drücke mir die Gegenstände recht ein. Urteilen
möchte ich gar nicht, wenn es nur möglich wäre.''

In the completed logical judgment subject and predicate appear clearly and
definitely connected. Yet even under the psychological process in which the
formation of judgment consists, they do not at any point fall wholly
outside each other. If that were the case, it would be unintelligible how
they could be connected into a judgment. While the judgment successively
develops out of the intuition or the association, the whole from which the
judgment's members are fetched stands constantly before consciousness. It
is within this whole that attention's transition takes place from the
initial representation, which becomes the judgment's subject, to the final
representation, which becomes its predicate. And the movement will be
repeated several times back and forth. Two currents will form, a
forward-going and a backward-going. What was at first initial
representation becomes, by the backward current, final representation: in
the judgment the subject is therefore just as much determined by the
predicate as the latter by the former.

In logical theories one has now had more regard for the one
determination-relation, now more for the other. The most common conception
has been that the subject is, in the judgment, subordinated (subsumed)
under the predicate, whose extension was thus greater than the former's.
``The man is good'' will here mean: ``The man belongs to the good
beings.'' The predicate is continens, the subject contentum. Here only the
subject gets a determination, the predicate none---beyond this, that it
contains, among other things, the subject in itself. In contrast to this,
\emph{Benno Erdmann}\footnote{\emph{Logik} I. p. 251 f; 261 f.} has
maintained that it is the subject that determines the predicate, in that
the latter's content is subordinated to the former. ``The man is good''
will thus here mean: ``The property goodness is a part of the man's
being.'' These apparently conflicting conceptions are easily reconciled
when one remembers attention's double movement during the formation of
judgment. When Galileo, according to the legend, came to think of the
motion of falling by seeing the church lamps swing, his attention---before
the judgment ``to swing is to fall''---moved back and forth between the two
representations; he saw the swinging as a kind of falling-motion, but at
the same time saw a falling-motion contained in the swinging. The intuition
of the swinging lamps became clearer through the final representation
found, of falling. The judgment's clarity is attained only by a
going-back from terminus ad quem to terminus a quo.

In the impersonal sentences the initial representation still stands in
indefiniteness. One notices a rustling in the foliage and a gleam
(Schiller's ``Erwartung''); but it is still uncertain who causes these
sensations. If it turns out to be the awaited one, the impersonal sentence
can be replaced by a new judgment;---but here the subject then becomes what
in the impersonal sentence was predicate, and the predicate becomes the
subject that in the impersonal sentence stood indefinite: ``the one whose
steps make the foliage rustle, and whose white gown shines through the
leaves---it is \emph{she}!''---In an example like this the two currents are
clearly seen, before the proper judgment is formed.

Logically, the time it takes before the relation between subject and
predicate can be determined plays no role, and the stages the
judgment-process must run through have also more psychological than logical
interest. Discovery and its winding ways interest the psychologist and the
historian; but the logician as such asks for grounding and proof. In the
last instance, therefore, it really does not concern logic what, during
the formation of judgment, is initial representation and what is final
representation. The difference between subject and predicate loses its
significance when one formulates the judgment as exactly as possible,
which is attained by expressing it in the form of an equation, as the
designation of a relation of identity of a certain kind. This path Leibniz
entered upon in some treatises that were first printed in 1840. Later, by
various paths, William Hamilton, Morgan, Boole, Jevons, and their
successors have worked in the same direction.

If the concept A shows itself by analysis to contain the concept B, one
can with Leibniz express this by: $A = A + B$. For one can everywhere,
where A is found, put $A + B$, since B is given when A is given. B has
here, during the formation of judgment, been the final representation,
terminus ad quem, and in the judgment it is predicate (A is B). But when
that logical equation is formed, there is no ground to distinguish between
start and conclusion. We can now keep to the result won, which can be read
from right to left as well as from left to right. That distinction is only
a memory from the preceding psychological process, insofar as it is not
due to an influence from grammar.

Leibniz's mode of expression at the same time reminds us aptly that we do
not let go of the subject when we pass over to the predicate; A occurs,
after all, on both sides of the equals sign. Yet every mode of expression
will more or less distinctly show the mutual determination of subject and
predicate. Even according to the usual manner of presentation, according
to which the judgment means that the subject's extension belongs under the
predicate's, so that the judgment can be expressed by $A < B$---even there
not only is A determined by B, but also, conversely, B by A, insofar as it
can serve to characterize B that it comprises A in itself. When I, e.g.,
learn that Amphioxus belongs to the vertebrates, I learn not only something
about Amphioxus, but also about the vertebrates, e.g. that a brain is not a
necessary organ in them. In the languages in which the predicate-word
conforms, in gender, number, and case, to the subject, the mutual
connection between the two members is also seen; this mode of expression
reminds one of Leibniz's.

In a logical equation the relation between the judgment's members stands
before us at a single stroke, wholly clear and precise. Here has been
attained, in a perfect way, what intuition attained at earlier stages of
development: the abolition of equilibrium that took place at the
dissolution of the intuition is now replaced by a new form of unity. The
judgment expresses a higher form of psychic energy, insofar as the whole
is here attained through a more composite work, and insofar as the whole
that is attained is more exactly determined. But if the whole fullness of
objects that an intuition (or an association-series) can contain is to be
expressed in the form of judgment, a great mass of judgments will be needed
for it, and it will always be only a matter of an approximation.---

What holds of the relation between subject and predicate in the
judgment---namely that the time it takes to get from the former to the
latter plays no role when the judgment is finished---holds analogously of
the relation between premises and conclusion: there is no time-relation
between them. The conclusion ``lies in'' the premises, just as the
predicate ``lies in'' the subject, and just as the judgment as a whole
``lies in'' the intuition (or the association). Inference is really only a
continuation of judgment: we infer when we discover that the predicate we
have ascribed to a subject is itself subject to another predicate
(\emph{Aristotle}'s first figure), or that a subject has two different
predicates (his third figure). Thus inferences grow out of judgments. But
the more judgments that must be connected for an inference to be drawn, and
especially when they must be fetched from highly different connections, the
higher degree of psychic energy the inference-process presupposes.

Just as little as all intuitions can be exhausted in judgments, just as
little can all judgments enter as members in inferences. And even where
judgments and inferences can be formed, the original forms of
unity---intuiting and associating---do not lose their significance. They
form the involuntary basis that progressing reflection cannot do without.
And perhaps, with the help of the judgments and inferences that have been
formed, a higher intuition can be built, an intuition in \emph{Spinoza}'s
sense, a new comprehension of the whole won through reflection, in which
the single thing could be intuited in its connection with the larger whole
to which it has shown itself to belong. What value such an intuiting can
have will be discussed at a later place in our investigation.

34. Historically there appears, right from \emph{Aristotle} on, a tendency
to lay the chief weight on the logical subject, although this is always
given prior to the judgment, and although reflection's interest is
attached to the closer determination that can be won by the predicate.
According to Aristotle\footnote{\emph{Analyt. post}. 1. 22.---\emph{De
categ}. c. 5.} there are concepts that can be used only as subjects, never
as predicates: these are concepts of individual beings (``substances'').
In Aristotle subject and substance are denoted by the same name
(ὑποκείμενον); the predicate is in the judgment added to the subject, just
as in reality actions, states, and properties exist only in the beings
that perform the actions and live in the states. The judgment's subject
thereby gets a higher dignity than the predicate: the subject corresponds,
after all, to the one who performs the action, has the state, or is the
bearer of the property---to ``that which lies at the basis.'' This
conception has, under various forms, maintained itself to the most recent
times. One did not heed the lesson that language itself gave through
emphasis, perhaps because one studied the written more than the living
language.

The inclination to lay the chief weight on the judgment's subject
expressed itself in a characteristic way in the first attempts to form a
Danish terminology in logic. \emph{Eilschow}\footnote{\emph{Cogitationes
de scientiis vernacula lingva discendis}. Hafniæ 1747. p. 36.} translated
``Subject'' by ``Hoved-Sag'' (principal matter) and ``Predicate'' by
``Bi-Sag'' (subordinate matter), and \emph{Jens Kraft}\footnote{\emph{Logik}.
København 1761. p. 141.} used in his logic the expressions ``Sag'' (thing)
and ``Tillæg'' (addition). Even \emph{Højsgaard}\footnote{\emph{Methodisk
Forsøg til en fuldstændig dansk Syntax}. Kbhvn. 1752. p. 256.}, who
otherwise sharply distinguishes between the grammatical and the
philosophical significance of the sentence, does not let his fine
linguistic apprehension gain sufficient influence on the philosophical
conception. He aptly emphasizes the predicate's predominant significance
by translating ``Verbum'' by ``Hoved-Ord'' (head-word) and adds: ``There
is no kind of word that has so important a post in any discourse as the
verb has. Every verb is in its proposition (sentence) like a captain for
his company, and is not unlike the head, which enlivens all the limbs, and
by whose nod they must all direct themselves.'' But he nonetheless starts
from the assumption that, philosophically regarded, the judgment's subject
expresses ``the efficient cause,'' which quite conflicts with the fact that
this subject first gets its determination by the predicate, whether or not
this determination is to be a cause.

In more recent times \emph{Wundt}\footnote{\emph{Logik}\textsuperscript{3}
I. p. 149 f.} finds in the relation between subject and predicate an
expression for the relation between thought-activity and its special
content; \emph{Benno Erdmann}\footnote{\emph{Logik} I. p. 261.} finds in
that relation an analogy to the inherence of properties in the thing;
\emph{W. Jerusalem}\footnote{\emph{Die Urteilsfunktion}. Wien 1895. p.
92--96.} sees in every judgment an expression for the relation between will
and action. But all this is analogies, due to language's tendency to
personify---analogies that can have their significance with respect to
making things intuitable, but are not necessary for understanding the
logical nature of judgment. The analogy between the subject's relation to
the predicate and the thing's to its properties, or the will's to its
actions, can do a similar service to that which other logicians have found
in the use of geometrical figures, especially for making the doctrine of
inferences intuitable. The proof never lies in the analogies or the
schemata. One must first show one's logical right to connect two concepts
as subject and predicate before one can apply those analogies. Even though
we perhaps always involuntarily use certain analogies, schemata, or symbols
when we think, there are nonetheless great individual differences both with
respect to which intuitive forms one uses, and with respect to the more or
less predominant role they play in individuals. Logic must therefore
distinguish definitely between the thought-relation itself and its being
made intuitable. Signs, as well as words, are always only invitations or
directions to carry out a certain act of thought. We must therefore always
subject our images, analogies, and schemata, as well as our words, to a
thorough critique, by testing whether we can draw all the consequences of
them, carry them through in all details. Where this is not possible, there
is a limit to the validity of the sign in question. We can perhaps not
think at all without images or signs; but thought shows its independence of
them in the constant critique to which it subjects them. \emph{Leibniz} has
clearly illuminated this relation, especially in the following utterance:
``Even though the signs are arbitrary, there is nonetheless in the use of
them and in the manner in which they are connected something that is not
arbitrary''\footnote{\emph{Dialogus de connexione inter res et verba}.
(Opera philos. ed. Erdmann) p. 77.---Cf. my treatise on \emph{Analogien og
dens filosofiske Betydning}. (Mindre Arbejder. Anden Række). p. 35.}.

In the use of examples the relation between thought and image appears
clearly. We draw only those inferences from the example that we could just
as well draw from any other example. And if we do not keep to this rule, we
misuse the example, just as when we extend a simile beyond the tertium
comparationis.

35. Until the judgment is finished, the subject lies, as we have seen, in a
certain indefiniteness. And even when the judgment lies formed, it is a
question whether the whole content of the initial representation has been
taken up into the predicate (or the predicates). This content must always
undergo a limitation in order to enter into a judgment. So far every
determination is a negation. To define is to delimit. Beneath our clear
judgments there therefore always lies an obscure basis that is not
exhausted, and for which predicates may perhaps later be sought. The
limitation that is necessary for the formation of the judgment may perhaps
later be abolished. There is an irrational relation between judgment and
intuiting (or association). There is always the possibility of more
decimals.

A first subject can therefore not be found. But just as little can a last,
concluding predicate be found. Every predicate can in turn become subject,
or the subject determined by a predicate can in turn become the basis for a
new formation of judgment. Our thought moves constantly from a more or less
indefinite basis toward a complete determination, which nonetheless
constantly sets new tasks and stands as an ideal. Existence, which
announces itself in our experience, cannot be exhausted in any series of
predicates, or in ever so many series of predicates. New initial
representations can always emerge, and new final representations must be
sought.

\subsection*{d. The Validity of Judgment}
\addcontentsline{toc}{subsection}{d. The Validity of Judgment}
\label{ssec:gyldighed}

36. Since judgment denotes a new state of wholeness or equilibrium, it
will be connected with an assurance of its validity, even if this
assurance does not specially come to expression. But this is not to be
understood as if it were first the judgment that brought the assurance or
the presupposition of validity at all. Consciousness operates from the
first with an involuntary, instinctive trust in everything that emerges in
it and for it, both in its individual elements and in their connections.
We are all born in faith; doubt comes only afterward---if it comes. And
that faith is there is seen from the involuntary way in which it shows
itself in action. A sense-impression releases a movement just as quickly
as it releases a sensation, and perhaps quite as quickly, as often in
reflexes and instincts. The single impression operates like the spark that
falls in the powder; it releases an energy that can seem wholly
disproportionate to the stimulation. And this energy can get full outlet
only through movement and action. Faith shows itself in action before it
appears as a special state of consciousness. It expresses itself in the
immediate absorption in, or devotion to, that with which it enters into
relation. Drive and expectation precede the deliberating thought, the calm
contemplation.

English psychologists have strongly emphasized this original and
involuntary faith. \emph{Hume} laid the weight especially on the expansion
by which an arousal, a lively movement from a single point in
consciousness, can spread to the whole state; under the influence of a
strong sensation or feeling we are inclined to place trust in all
representations that hang together with them; they get a share in the
latter's life and strength. \emph{Bain} has laid weight especially on the
original motor tendency, the urge for movement, that makes us often act on
representations before we have tested them. The single representation's
connection with the urge for movement is from the first closer than its
connection with other representations that could determine and limit,
perhaps even inhibit, it. It is the urge for movement that makes us
involuntarily test our representations by applying them in action and thus
apply a kind of experimental method.

This original faith or expectation makes itself felt, as we have seen
above (16), already in mediate recognition. It operates in simplest form
already before proper, free representations have really developed. It is by
no means the case that we first develop our representations completely and
clearly, and then afterward ask about their validity.

But apart from the influence of expansion and the urge for movement, it
follows from what one might call psychic inertia that sensations and
representations win full trust and assent as long as no conflicting
elements appear. We have no ground to doubt their validity as long as
nothing inhibits them. Just as a movement changes speed and direction only
when an outer force intervenes, so the development of our consciousness in
a certain direction changes only when motives arise that lead in another
direction. An urge or a drive that has once been awakened can (when it
does not die from lack of nourishment) be inhibited only by another drive.
And thus men's first assurances and expectations, whether they are due to
their own observations or to traditions, change only when opposing
experiences or assertions exert their influence. Doubt is always
secondary.

Every intuiting, every association, and every judgment therefore has an
existential character. And the same holds for the inferences we draw from
our judgments. Dogma precedes critique. The history of thought bears
witness to this by the tendency that expresses itself to regard points of
view, concepts, and forms of intuition as absolute realities. The
Pythagoreans' numbers, Plato's Ideas, Spinoza's Substance, Newton's
absolute space, materialism's atoms, Kant's ``Ding an sich''---are
examples of this. There stirs here in the thinkers the same tendency that
gives the ordinary consciousness so secure an assurance of the reality of
the sensory surrounding world, despite all subjectivist objections and
despite all the disappointments produced by sense-illusions.

This primitive assurance, which in practice bears our whole life and stirs
in our cognition from its first germ, has its ground in the instinct of
self-preservation and is of great significance in the struggle for life.
But it does not have quite the same character as the conviction that can be
attained through the purgatory of doubt and contradictions (cf. above
26--27). It does not appear as an independent element different from the
intuition, the association, or the judgment itself, and does not itself
become the content of a predicate. Precisely because it is wholly
involuntary, it does not become the object of a seeking and does not come
to stand as final representation, terminus ad quem. It determines the
whole character of the state, but is not itself object or subject-matter.
This wholly immediate character of the assurance one could express by
ascribing to our intuitions, associations, and judgments an original
existence-quality, which first becomes the object of explicit
consciousness when an opposite quality, non-existence, can appear over
against it. This happens first through disappointment and lack. Then the
sharp difference between being and not-being is experienced. It goes with
the existence-quality as with innocence; it is first brought forth before
consciousness when it comes to stand over against its opposite---is first
experienced when it is threatened or abolished.

I here use the word quality, as where the talk was of the simplest kind of
recognition, to designate the wholly immediate and involuntary, ultimately
indescribable feature of a state or an object, and in order to exclude the
tendency that, both as regards recognition and existence-assurance, makes
itself felt to read relations that first come forth at more developed
stages of consciousness into the purely elementary phenomena of
consciousness. Just as one has held that all recognition is due to a
judgment, so one has also held that the difference that at later stages can
be found between existence-conviction and mere representation must be
present from the first. But what one can find out by analysis of the grown
European's consciousness one has no right to presuppose in the child or the
Patagonian. Analytic and genetic psychology here enter into decided
opposition to each other. When one, with \emph{F. Brentano} and his
disciples, starts from the assumption that a judgment must be presupposed
for there to be existence-assurance, one naturally comes to read judgments
into the most involuntary utterances of the mental life. It becomes
scholasticism, neither psychology nor epistemology\footnote{Cf. my
critique of \emph{Meinong}, one of Brentano's most eminent disciples, in
\emph{Göttinger Gelehrter Anzeiger} 1896. p. 299 f (where I have for the
first time used the expression existence-quality).---\emph{Stumpf}, who in
his \emph{Tonpsychologie} (1883) joined Brentano's conception and regarded
every observing and noticing as a judgment, now distinguishes definitely
between these more elementary functions and judgment. \emph{Erscheinungen
und psychische Funktionen}. (Abhdl. der preuss. Academie der
Wissenschaften. 1906). p. 16.}.

37. The involuntary experimenting that the original presence of the
existence-quality calls forth can lead to experiences by whose help
critique can be applied to the immediate existence-quality. If the urge to
act in a certain way or in a certain direction meets disproportionate
resistance or entails unbearable suffering, it will fall away by degrees,
if such an energy cannot be mustered---perhaps by the very resistance or
suffering---that the inhibitions are overcome. Or it can happen that the
original secure expectation, by reason of the opposition to what actual
experience shows, is formed into a more distinct intuition or into a
judgment, so that a clear comparison between the expected and the occurred
can take place. Here one of two things can then happen. Either the content
of the expectation can lose the existential character and pass over to
becoming ``mere'' representational content, a representation of something
``merely'' possible, indeed perhaps of something impossible, or at least of
something whose time is past, if it has ever been. Or the original
expectation can be altered in a similar way to that in which we have seen
the immediate intuition can be altered. A lesson is then drawn from the
disappointment met, and the expectation is limited or reshaped in such a
way that it is no longer shaken by conflicting experiences.

The chief task for the psychology of cognition becomes, after the whole
consideration we have here made, not to explain how our sensations and
representations can come to stand as the expression of something real; but
the task becomes, on the contrary, to show how representations of something
merely possible and of something unreal or impossible can arise. And only
when this has happened can there be a question of how we distinguish
between the real and the unreal---what criterion of reality we have---and
what reality really is.

The criterion of reality now shows itself to be merely a use, clarified
through reflection, of the involuntary experimenting to which the
existence-quality and the original expectation already lead. The
expectation must be formulated as a definite question. Definite inferences
must be drawn from the objects at hand and the involuntary assumptions they
have led to, and it must then be investigated how far these inferences are
confirmed by continued lived experiences and experiences. The expectation
is thus formulated as a hypothesis that seeks confirmation. The criterion
of reality consists in agreement and law-governed connection between as
many sensations and representations as possible. The testimonies of the
different senses, the different memories, the different inferences, and the
different individuals must agree and show themselves compatible, preferably
in such a way that a law-governed connection shows itself between them, so
that one can infer from the one to the other. All that exists must
ultimately constitute a great totality whose members join together in a
natural, or even necessary, way. Illusions and hallucinations, dreams and
fantasies, speculations and intimations must, even where they appear with
the most vivid existence-quality, give way to the great, consistent
connection that is our last refuge in cases of doubt, the firm fortress
from which we judge what is to be called real and what is to be called mere
representation. If the world of waking and sober life could just as well be
subsumed under and understood from the world of dreams as the latter from
the former, it would be impossible to draw any boundary between reality and
unreality.

Reality is real truth. Formal truth a circle of representations can be said
to have when it is won by consistent reflection from a definite point of
view. And since the firm, consistent connection is an essential element in
the concept of reality, the concept of reality presupposes the concept of
formal truth, which it is logic's affair to develop. The concept of
reality comprises, besides formal truth, also the law-governed connection
and agreement between as many different points of view as possible. It
requires a working-together of all the points of view from which we can
orient ourselves in existence\footnote{Cf. with respect to this whole
consideration my \emph{Psykologi} V D (The Apprehension of the Real).}.

That the criterion of reality cannot consist in anything wholly simple was
already impressed in antiquity by \emph{Carneades}\footnote{See \emph{Sextus
Empiricus}. ed. Bekker. p. 229--232.}, in that he pointed out that one
cannot rely on any single and momentary sensation or representation. It is
not the immediately persuasive force of a representation that is decisive;
what matters is whether it can be held fast and carried through by
exhaustive testing. (The representation must not only be πιθανή, but above
all ἀπερίστατος and διεξωδευμένη.) One cannot rely on a single mark, but on
the agreement (συνδρομή) of all marks.

The law-governed connection is the criterion of reality of the more recent
science. Thus \emph{Galileo} says: ``The mark of what is natural and true
(la proprietà e condizione delle cose naturale e vere) is not merely that
no self-contradiction is present; nor merely that an easy transition can be
made from the one to the other; but it consists in the transition's taking
place with necessity, and in its being impossible that it can proceed in
another way''\footnote{\emph{Dialogo sopra i duo massimi systemi}. Giornata
quarta, (Ed. Fiorenza 1710. p. 417.}. And the same thought recurs in all
the philosophers of the more recent time from Descartes \emph{(sixth
Meditation)} to \emph{Kant} (\emph{Kritik der reinen Vernunft}, p.
236--247). In the period of Romanticism it became a chief thought that
truth or reality must be a totality, outside of which nothing can be
understood or have reality. And the fundamental features of this totality
one held one could construct by way of pure speculation.---In a work that
on many points seeks to hold fast Hegel's fundamental thoughts, F. H.
\emph{Bradley}'s \emph{Appearance and Reality} (1893), the idea of reality
as a totality of all possible experience has taken on the character of a
standard: a reality is the higher the greater the multiplicity it encloses,
and the greater the unity that makes itself felt in this multiplicity. But
Bradley has given up the Romantics' hope of being able to construct the
absolute totality (the one absorbing experience)\footnote{Cf. on Bradley's
philosophy my book \emph{Moderne Filosofer}. p. 44--54.}.

38. How great the totality of experiences must be, into which the single
phenomenon must fit in order to be able to be regarded as real---that
cannot be decided once and for all. The world of experience expands
constantly and cannot be concluded. Therefore the concept of reality or
existence is, however strange it sounds, an ideal concept. For an
absolutely secured assumption of reality it would really be required that
the single object had been held together with all possible observations,
and this is of course impossible. We can only strive to work with as large
a horizon as possible. Only a provisional assurance can ever be won. The
concept of reality is constantly in its becoming. But this does not exclude
that this concept, having emerged from the application of the criterion of
reality that is possible at a given time, can now become a
predicate-concept, in that reflection investigates how far it can find
application to the single object at hand. Only now do proper
existence-judgments become possible. The concept existence is formed in a
way analogous to that in which the concept likeness is formed. Sensations
and representations must often have been placed together on account of
their likeness before the concept likeness could be formed and become
predicate in judgments. The association by likeness operates long before
the concept likeness is set up. Thus the existence-quality can long have
been present without reflection having turned against it and investigated
its conditions and its justification. Existential judgments, such judgments
in which the concept existence is predicate, cannot be primary, but
presuppose a whole development. Behind the immediate assurance they
express, e.g. ``There \emph{are} good people!'' ``There \emph{are} flying
mammals!''---there really lies the presupposition that the judgment's
subject belongs within the great connection of our experience as a single
member. The presupposition for the existential judgment is a combining and
comparing process, whose possibility is maintained, even if the process is
not carried through explicitly.

It is by way of reflection, not by way of mere sensation or the
involuntary train of representations, that we reach reality. It is so far
of ideal nature, is an object of thought. In many ways, as the history of
science shows, thought must reshape the sensuously and involuntarily given
in order to be able to ground a conviction of reality. Naive, secure
realism has another concept of reality than the constructive realism that
seeks to carry through a justified concept of reality, and that it is the
affair of the cognition of nature, of historical research, and of
philosophy to separate out. Slowly but surely human thought moves in this
direction.

Yet it will always show itself that there are certain definite objects
that form starting-points for reflection in its striving to apply the
criterion of reality. All objects do not stand on the same line; reflection
must begin by presupposing the reality of certain objects and testing the
others in relation to them. Our thought must have a place to stand, and
only at a later stage of reflection can this place itself be drawn into the
discussion, namely when another starting-point has been won. The
beginning-point is in a certain sense accidental; it is perhaps
historically and psychologically explicable; but history and psychology are
themselves part of the cognition whose validity is to be tested, become
themselves objects for epistemology. From this side too it shows itself
that the concept of reality is an ideal. It is not only the content of our
apprehension of reality, it is also the very standpoint from which we have
found this content, that must pass through the critical purgatory. And
without a standpoint no reflection can, as said, be exercised.

39. Value-judgments offer an analogy to existential judgments. In their
simplest, most involuntary form they appear as mere exclamatory judgments
(see 30). When doubt arises about the validity of the involuntary
valuation---when, e.g., the one cries ``Abominable!'' where the other cries
``Splendid!''---a criterion is needed here too. And here too the only
criterion will show itself to be agreement between the individual
valuations. The two conflicting parties must seek out something about whose
valuation they agree, and then investigate whether one of them did not
commit an inconsistency in the valuation that became the occasion of the
dispute. If they can find no such point at all, the dispute is hopeless.
Each one must then seek to preserve agreement with himself in his
valuations. But whether or not agreement can now be reached, it will show
itself that, just as the criterion of reality in the individual cases
presupposes the reality of certain objects as given, so there is in fact,
in the value-judgments that are passed, presupposed one or more values.
Their validity is not---at least provisionally---drawn into the discussion,
but the relation to them conditions the validity of all other values. Such
values we can call fundamental values. A fundamental value denotes, for
each individual personality and for each individual historically given
group of men, the presupposition on which valuation in the particular
case, the fixing of the individual values each by itself, rests. Only in
reflection's further development can the fundamental value itself be drawn
into the discussion---but this will presuppose that a new fundamental value
has developed and consolidated itself. Reflection can here, just as little
as in the purely intellectual domain, hover in the air; but it can
successively investigate its earlier standpoints, and it can become
conscious of the very standpoint from which it makes its concluding
valuation.

The two considerations we have here distinguished between, that which
concerns reality and that which concerns value, stand in a very close
relation to each other. For on the one hand the cognition of reality
itself has its value, a value that for many minds stands as the
fundamental value, and on the other hand the values, especially the
fundamental values, must be so constituted that they can be held fast and
developed within the world whose reality is shown by the greatest possible
connection between the greatest possible number of experiences. This
possibility belongs to the conditions of the value's validity. Finally,
the reality that a purely intellectual interest leads to discover gets a
peculiar value if it shows itself to afford possibilities for the
development of a world of values. In the special discussion of the tasks of
thought we shall return to what is here intimated.

\begin{center}---\end{center}

%% ============================================================
%%  II. THE HISTORY OF THOUGHT
%% ============================================================
\chapter*{II. The History of Thought}
\addcontentsline{toc}{chapter}{II. The History of Thought}
\markboth{II. The History of Thought}{II. The History of Thought}
\label{ch:historie}

\begin{center}---\end{center}

\section*{A. Animism, Platonism, and Positivism}
\addcontentsline{toc}{section}{A. Animism, Platonism, and Positivism}
\label{sec:animisme}

40. The psychology of thought presupposes its history. The functions
arising step by step, which have been described in the foregoing, have
developed under definite conditions of life, physical and social, and
their character is determined by a constant reciprocal action between inner
and outer conditions. It falls to sociology and comparative psychology to
unravel more closely this relation, which in the foregoing could only be
intimated. In \emph{Hobhouse}'s book \emph{Mind in Evolution} (1901) a
comparative-psychological treatment of the development of the life of
thought is given. And it is here shown at the same time how social
conditions enter into reciprocal action with intellectual development, in
that the life of thought is developed through the living-together of
individuals, while on the other hand social instincts and feelings
presuppose a certain intellectual development, without which recognition of
other individuals and understanding of their conditions would not be
possible. Already language, which is the work of the race, not of the
single individual, shows the life of thought's dependence on social
development. The individuals do indeed give their contribution to the
growth of language, more or less significant and more or less conscious
contributions. But essentially language is nonetheless a social
inheritance into which the single individual lives himself. In it many
representations and points of view are already arranged. It has, as has
already shown itself in the foregoing, its own metaphysics, which is a
result of development that now promotes, now inhibits the individual life
of thought. And language, as well as the thought-content deposited in it
and the thought-forms developed through it, has itself developed in
connection with the whole social life in family, tribe, religion, and
state. The biological-psychological consideration thus leads over into the
sociological, which shows the development of the life of thought under the
influence of social institutions. It has even been maintained from the
sociological side that, just as little as obligation can be understood from
the standpoint of the individual will, just as little can the necessity
with which certain fundamental concepts (categories) make themselves felt
be understood from the individual's intellectual development; in both
places the explanation lies in social and historical development. The
categories are of social origin, are tools of thought that human societies
have laboriously formed through centuries, and in which they have gathered
the best of their intellectual capital. \emph{Émile Durkheim}, who has
especially asserted this pronounced sociological point of view, does not,
however, thereby wish to dispute the significance of a direct psychological
investigation of the life of thought. To a complete investigation of the
human fundamental concepts belongs, according to him, also an investigation
of the germs of thought-life that can be found by direct analysis of the
individual consciousness\footnote{\emph{É. Durckheim}: \emph{Sociologie
religieuse et théorie de la connaissance}. (Revue de Métaphysique et de
Morale. 1909) p. 750.}---such an investigation, then, as I have attempted
in the foregoing. It must at the same time, in the face of the strong
emphasis on the sociological point of view, be definitely maintained that
the overwhelming social influence on the life of thought is easiest to
demonstrate at more primitive stages of human development. During the
development of culture the individual thought frees itself more and more.
Already comparative inquiry in the domain of psychology and sociology shows
this. Sociology itself would not be possible if such a liberation had not
taken place, which of course can never be absolute, any more than, for that
matter, the social influence can ever be absolute. When it is so easy to
demonstrate the social influence on spiritual development in earlier
periods, this has for no small part its simple explanation in the fact that
the individual influences are forgotten. One has the products left, but the
factors have vanished, and one then lets ``society'' or ``the group'' bring
forth itself in a mystical way. At every given stage ``society'' consists
of definite individuals, each with his dispositions, with his urge and his
circle of experiences. ``Society'' easily becomes an asylum ignorantiæ. It
is a convenient word to use where one does not know or cannot survey the
operative individual forces in their many differences and in their infinite
reciprocal action. The historical point of view must not be made without
further ado one with the social point of view. Every individual lives
himself into a historical connection, also as regards the life of thought;
but this historical connection is itself the result of manifold individual
forces' cooperation through the ages, of giving and receiving between
individuals in constant reciprocal action. The social is what makes the
reciprocal action possible; but the reciprocal action itself presupposes
the individual forces.

41. That the individual life of thought always unfolds on a historical
basis was strongly emphasized by both the two great main directions in the
domain of philosophy in the nineteenth century, Romanticism and Positivism.
Both directions had---in common opposition to the eighteenth
century---a clear understanding of the social and historical atmosphere
within which the spiritual development of individuals takes place. They
could think of no spiritual development on bare ground. By \emph{Hegel} and
\emph{Comte}, two eminent representatives of the two main directions, the
history of thought has therefore also been written in broad strokes. In his
\emph{Phänomenologie des Geistes} (1807) Hegel showed how cognition begins
with the single and isolated thing, but is from there driven forward step
by step through the contradictions in which it becomes entangled. Through
step-by-step development thought unfolds its essence and shows itself to be
an ideal totality within which everything single can find its place, at
once as a peculiar member and yet ``sublated'' in the ``higher unity'' that
encloses all oppositions in itself. The individual stages in this
development are illuminated by examples and types, partly of individual,
partly of social kind. But it is Hegel's view that only the thought that
can unfold on the basis of what the history of spirit has brought about
will have the conditions for exercising a significant work. The ``Spirit''
whose history of manifestation (Phenomenology) he describes in his work is
the spirit of the human race, in which the individual spirit is only a
single moment. It is from the ``higher unity''---won at the given
stage---of formerly struggling oppositions that the individual thought-work
is to be exercised.

\emph{Comte} too distinguishes (\emph{Plan des Travaux}. 1822) between
different stages in the development of the life of thought. But with him it
is not the spurring force of oppositions and contradictions that leads
forward. His chief thought is that at all stages a relation makes itself
felt between what is given in experience (the objects) and the
representations by which it is interpreted. The difference between the
stages rests partly on how large a circle of objects is at one's disposal,
partly on which representations prevail in man, so that they can more or
less involuntarily be taken as basis in the interpretation. The narrower
the circle of experience is, the greater the role the ruling
representations will play; the more comprehensive it is, the more the
thought-work will come merely to aim at bringing about connection within
the given, so that this interprets itself.

The points of view that the two philosophers take up supplement each
other. For when experience necessitates a new mode of interpretation, or
when the difficulty of carrying through an interpretation leads to seeking
out new experiences, it is evidently because it discovers contradictions or
oppositions in the hitherto given, or between the prevailing interpretation
and the given. Thought is shaken in its accustomed interpretation by new
experiences, or it hopes that its difficulties will fall away through new
experiences\footnote{These two sides were emphasized by Danish philosophers
in the eighteenth century, the first by \emph{Holberg}, the second by
\emph{J. S. Schneedorff}. (\emph{Danske Filosofer}. p. 5; 18.)}. Whether by
these paths ``higher unities,'' more comprehensive unified intuitions, can
always be reached, as Hegel held, is a matter in itself, which will be
discussed later. And whether Comte is right in the sharp difference he sets
up between three stages in cognition, the theological, the metaphysical,
and the positive, we will here also not discuss more closely\footnote{See
on this \emph{Den nyere Filosofis Historie}\textsuperscript{2} II. p.
334--338 and a treatise by \emph{Hobhouse} in \emph{Sociological Review}.
1908. p. 262--279.}. But with some change in the points of view one will
rightly be able to preserve Comte's tripartite division, as I shall now
attempt to show. The change entails a corresponding change of the
designation; one will suitably be able to distinguish between Animism,
Platonism, and Positivism as three main stages, in that these three words
are each taken in an extended sense.

42. In its simplest form the difference between object and interpretation
appears in the opposition between distinguishing and recognition, the two
poles in our cognition, which correspond to the two great conditions of
life: openness for the new and application of the old to the new (Cf. 14).
But only when judgments are formed, and especially when questions and
problems are set (26--27), is there a matter of interpretation proper. The
presupposition for this is that one does not immediately proceed to action.
For when the emerging cognition-elements immediately release reactions in a
reflex and instinctive manner, there is neither urge, time, nor force for
interpretation. Interpretation presupposes a certain degree of intellectual
interest and, in connection therewith, experiences that there are limits to
our capacity to intervene in the course of things, so that we must come to
know the essence and laws of things in order to be able to place them in
the service of our purposes.

\emph{Frazer}'s hypothesis about the relation between magic and
religion\footnote{\emph{The Golden Bough}\textsuperscript{2} I. p.
61--78; III. 458 f.---\emph{Adonis---Attis---Osiris}. p. 3--4.} is of
interest in this connection. Prior to religion, which always presupposes
experience of dependence, there goes, according to Frazer, a period in
which man has trust in his will's power to such a degree that he believes he
can bend nature under himself, not only where he really intervenes
actively, but also by mere wishing, made known by word or ceremonies. Or
rather, the difference between action, word, and ceremony does not exist
for man at this stage. It is experience of the will's limitation that leads
beyond this stage, and it is likewise this that leads into the world of
cognition. The magical art does not always lead to the goal. Events often
take place independently of or in defiance of the human will. Through
disappointment and the feeling of impotence, man learns that in existence
there prevail powers that are stronger than he. These powers he conceives
as personal beings, and this conception---however it has arisen---afforded
the first means of interpretation, the first ``categories.'' Man did not
give up magic; but it was now exercised toward the overmighty beings whose
will it was a matter of bending, or whose power it was a matter of
supporting by the performance of certain ritual actions. Already the
magician at the pre-religious stage performs his ceremonies in a certain
definite way, since he otherwise has no trust in their operation, and he so
far recognizes a certain order of things; but this recognition becomes more
penetrating and comprehensive when magic is replaced by, or at least
subordinated to, religion.

I here disregard the question whether the arising of religion can be
explained merely by the disappointments to which autonomous magic led.
There must, after all, have been many other paths by which man could
experience his dependence. What in this connection is of interest is the
indisputable opposition between a standpoint where man has no occasion to
recognize an order of things, and another standpoint where such a
recognition is alive. From the one standpoint to the other man can only
have come by way of experience, a way that has probably often led through
hard disappointments. Only when this way has been traversed is there a
matter of interpretation. And history teaches that the first attempts at
interpretation have the character of \emph{Animism}, in that the
explanation of natural phenomena, at least the most striking and
far-reaching ones, was sought in the intervention of personal, man-like
beings.

Animism appears in a series of forms---from momentary and special gods to
polytheism's individualized god-figures, and from these to a more or less
thoroughgoing monotheism\footnote{See my \emph{Religionsfilosofi} §
46--53.}. These transitions are not only of significance for the
development of the religious life, but also for the development of the life
of thought. The mythological imagination's individualization of a
god-figure---which corresponds to the psychological transition from
concrete to typical individual-representations (24)---leads naturally to a
more definite distinction than before between the natural phenomena that
depend on the god's intervention and the god himself. This distance-relation
makes a more thorough investigation of the natural phenomena themselves
possible; the god withdraws and hides himself more behind his work than
before. Human experience gets time and space to spread out, since it does
not everywhere or immediately strike upon the god. And the god-figure
itself now presents an inner connection, a more or less consistent
characterization, so that the representation of it denotes an advance in
psychological insight. A typical individual-representation presupposes that
one regards all the individual utterances and situations in which a being
lays itself bare as examples of its character. It indicates, where it is
distinct, the law for the development of the being in question, la loi du
changement (Leibniz). Both physically and psychologically, then, polytheism
denotes a great advance. Comte has even called its development the most
important advance within the ``theological stage.'' The transition from
polytheism to monotheism too has of course its great significance. An
absolute monotheism is scarcely ever carried through. But the fewer gods
there are, the more the divine intervention in the course of natural
phenomena can approach a minimum. On the other hand, the strictest
monotheism can get a likeness to the belief in momentary gods, when the one
god does all too many miracles.

During this whole process there is a reciprocal action between religious
and intellectual elements, or, if one will, between religious experiences
and other experiences. When by religious experience one understands the way
in which the human life of feeling is attuned through what is experienced
with respect to the fate of human purposes and ideals under the course of
things, then this experience will contribute much to the way in which the
powers of existence are conceived; religious experience has a spurring and
forward-driving influence on the whole world-view. One need only compare
Homer's world-view with Hesiod's, Plato's with Marcus Aurelius's, the Vedic
poems' with Buddha's, older Judaism's with that of the prophets. But by
degrees experience and understanding of the natural connection of things,
hence physical insight, is also gathered. And the more social life and the
personal life in individuals develops, the more connection the human life
of personality presents, the less can one think of the gods' operation as
guided by momentary and changing motives. Psychological and physical
experience thus meet with religious experience.

43. Animism (in the widest sense) has not yet disappeared as a principle of
interpretation because strict science develops. It is limited to ``the
creation of the world,'' or to certain extraordinary phenomena strongly
intervening in human destinies, or one finds in it the ultimate explanation
of the fact that there prevails law-governedness in nature at all. In some
of the greatest inquirers of the more recent time the animistic principle
of interpretation is found in one or more of these forms. \emph{Kepler} and
\emph{Descartes} ground the conviction of the universe's natural
law-governedness on God's perfection. According to \emph{Leibniz}, all
monads, existence's ultimate elements, ``radiate'' from a primal monad, by
whose choice the best of all possible worlds has come to be, a world whose
course is determined by strict natural laws, but in such a way that these
natural laws themselves are determined by regard for the greatest possible
``perfection.'' The solar system's harmonious structure and other purposive
natural arrangements are, according to \emph{Newton}, due to God's
intervention and are not subjected to natural-scientific explanation; the
infinite space itself that mathematics presupposes is God's sense-organ,
the organ for his omnipresence. \emph{Linnaeus} derived the different
species of organic beings from just as many different divine acts of
creation, and \emph{Kant} found in the concept of time the only possible
solution of the great problem set by the opposition between moral ideality
and nature's actual course.

Yet during the development of the life of thought there takes place a
constant elimination of animism. This is used less and less for the
interpretation of the objects at hand. It is by this indirect way, by
non-use, that animism disappears, not by direct disproof. It really cannot
be disproved. \emph{Kepler} saw a disproof of the assumption that the
planets were moved by souls in the fact that the moving force decreases with
distance\footnote{Cf. \emph{Den nyere Filosofis Historie}\textsuperscript{2}
I. 165.}. But how could he really know whether a soul's operation did not
also decrease with distance?

Animism has, despite everything, exercised a significant function during
the development of the life of thought. It was the first coherent train of
thought in which man could set his scattered experiences and lived
experiences. And if it had been possible to develop and carry through the
concept of personal beings in such a way that it could afford frame and
basis for all understanding of material and spiritual nature, and at the
same time guide toward setting research new tasks, then its time would not
be past. Theology would then be both the highest and the fundamental
science. The history of the life of thought shows that it has not been
possible.

To this is added that the personal life itself shows itself to be subject
to definite laws and is itself a part of the objects that are to be
interpreted. Animism consists really in a part of our experience, namely
our experience of the mental life, being taken as basis in the
understanding of all other parts of our experience. One cannot, with Comte,
distinguish absolutely between the given and its interpretation, since every
interpretation ultimately builds on something that is itself given.
Interpretation consists at all stages in the one object's coming to
illuminate the other. There is thus a constant reciprocal action between
experience and interpretation. In each single case there are some objects
or experiences that are to be interpreted, and others that serve for
interpretation. This was, in Comte's view, to be the characteristic of the
``positive'' stage, but it is in reality characteristic of all stages. The
difference is only the choice that is made of interpreting objects, and the
exactness with which the interpretation is undertaken; and both these
things hang closely together, when it shows itself that not all sides or
parts of our experience are equally well suited as basis for the
understanding of the others.

44. Animism always had to undertake a certain ordering, in any case have a
capacity for recognition, in order to be able to know under which divine
being's domain of power the single object belonged. In ``special gods''
(departmental gods) this shows itself especially clearly; but it is a
relation that is found again under one form or another in all religions.
But the ordering or classification that here took place was very external;
there was no inner relation between frame and content. And yet an ordering
survey of the objects that are to be interpreted is the first condition for
their understanding. The point is then to order them according to
likenesses and differences. Reflection's essence as comparison expresses
itself here. It soon showed itself that there are many degrees of likeness
(and difference), and it dawned on thought that it is possible to bring
about an ordering of the content of our experience from this point of view.
One had then to try to discover as many and as essential likenesses between
the objects as possible, and order them accordingly. The significance of
Greek philosophy lies for a great part in its having impressed the
importance of this task and made the first attempts at its treatment, and
\emph{Plato} can here be regarded as its typical representative. For Plato
thought's essence laid itself bare in its capacity to form concepts that
contained the points of likeness between large groups of objects and
thereby comprehended these in themselves. The discussion of objects that
led to sharp determination of their likenesses and differences Plato called
dialectic, and the dialectician had to have a capacity for comprehending
survey (ὁ μὲν γὰρ συνοπτικὸς διαλεκτικός, ὁ δὲ μὴ οὔ. Republic 536 C).
There could by this way be formed a world of Ideas, in which the nature of
the objects found its highest expression. The individual things and events
arose, changed, and vanished, but this changed nothing in the concepts in
which thought had comprehended their essence. All that is beautiful, e.g.,
is perishable, but the Idea of the beautiful is imperishable.

The ascent from the world of differences and changes to that of Ideas, in
which thinking for Plato essentially consisted, was now for him not merely
a formal systematic ordering, a classification, but in the Ideas or general
concepts that held for a group of objects he found at the same time the
ground that such objects existed at all. The doctrine of Ideas is not
merely a doctrine of method, an attempt at a conceptual ordering of all
thought-content. For Plato's aesthetic-religious conception the Ideas also
came to stand as plastic figures, as constituting the true reality, of
which the individual special objects are only imperfect copies. The value
of the copies rests on their nearer or remoter relation to the prototypes,
and of these prototypes again the Idea of the Good is the
highest\footnote{\emph{Natorp}, in his work \emph{Plato's Ideenlehre}
(1903), keeps only to the methodological significance of the doctrine of
Ideas, and interprets Plato rather too much from modern points of view.
Thus especially when he interprets the phenomena's ``participation'' in the
Idea by the relation of single events to the law (p. 151).---J. A.
\emph{Stewart} (\emph{Plato's doctrine of Ideas} 1908) distinguishes
sharply between the methodological and the contemplative significance, but
concedes that Plato constantly confuses these two significations.}.
Interesting here is Platonism's relation to Animism. They meet in the
presupposition that what lies at the basis of the phenomena must also be
something that can be the object of men's urge and longing, contain values
or be the cause of values. For Animism it was the god that was both the
ground of things and the source of values. Plato gives up the
personification, and the highest value he found in the very relation to the
Ideas, in the happiness of beholding them. And this holds for gods as for
men. The advantage the gods have is only that they stand nearer to the
world of Ideas. In the great procession that in the ``Phaedrus'' sets out
to admire the Ideas, the eternal stars in the heaven of thought, the gods
drive first. Herewith is denoted a decisive turning-point in the relation
of religious representations to scientific thoughts.

By the projection that let the Ideas stand not merely as thoughts, but as
things, there came forth in Platonism a great problem: how do the Ideas
relate to the phenomena? It was a reflection of Animism's problem, how the
gods relate to the phenomena. A critique that possibly derives from
\emph{Aristotle} emphasized that consistently there would have to be set up
an Idea under which both the individual Ideas and the phenomena belonging
under them were to be comprehended. The justification of this critique lies
in the fact that reflection cannot end in any principle that should ``have
its ground in itself,'' but must always anew discuss the relation of the
Ideas or principles won to experience---and that (what Aristotle did not
see) not merely to the experiences hitherto won, the objects hitherto at
hand, but also to the new ones that can emerge or perhaps even be called
forth.

At another point too Platonism shows its insufficiency. In that it makes
classification, tracing-back to Ideas, one with causal explanation, it
comes to assume just as many causes as there are kinds of objects to
explain. That the Ideas are again to form a systematic connection, and
finally be subordinated to the Idea of the Good, makes no difference in
this connection. Here too Platonism reminds one of Animism; just as the
latter has a god for each group of things or events it is able to
distinguish, so Platonism has an Idea for each such group. Thereby a way
was paved for the kind of pseudo-explanations such as that life is derived
from a life-force, gravity from a force of gravity, and so on. It was this
peculiarity that Comte especially laid weight on for the characterization
of the second of his three stages, the one he called the ``metaphysical.''

Yet it denotes a great advance that the ``Idea'' or the force stands as
something that operates constantly, always in the same way under the same
conditions, something that could not be said of Animism's gods. But the
advance lies especially in Platonism's weight on the ordering thought-work
that systematizes according to likenesses and differences. This will always
continue to be the first step toward gaining penetrating understanding.
From the mere catalogue with its external ordering to the ``natural''
systems resting on deeper-lying principles, such work appears in its great
significance. Platonism's time will therefore never be past, as long as
series-formations, type-arrangements, and overview (synopsis) continue to
be of necessity. And there are domains, such as the psychological and the
historical, where research perhaps never gets beyond this kind of work, at
least as regards the most significant questions.

45. \emph{Plato}'s Ideas express relations of likeness between different
objects. When it is asked why the individual objects appear precisely with
their definite properties and at a definite time and place, Platonism
gives---just as little as Animism---any answer to this. Its method can only
lead to an ordering that is grounded on comparison between the objects with
the properties they happen to have, but cannot say anything about the
origin of the objects and their properties. It sees existence in
equilibrium; the comparing and ordering reflection stops the stream of time
in order to fix the types that the objects present. But since the objects
and their properties arise each by itself at definite stages of the stream,
the task naturally comes up of investigating the relation between the
coming-into-being of the different objects and properties. Does the one
object arise independently of all others? Can the single property appear
without other properties appearing or vanishing? Answers to these questions
are not obtained by a merely resting ordering, but by finding out an
ordering of the succession in time according to definite rules, so that the
ground for the arising of the following objects and properties can be found
in the presence of the preceding ones.

We can use the word Positivism to denote the direction in the life of
thought that sets new tasks in the way just intimated. Here again the point
is to understand objects through comparison. The comparison concerns the
definite conditions under which an object comes to be or a property
appears, the conditions for its localization in time or in space or in both
respects, and the connection is to give the unavoidable historical
succession as determined by a law. Every event is thus attached to another
event, and it is at this that the endeavor aims, not merely at finding the
likeness and difference between the different objects and properties. And
the explanation is now sought not outside the series of events, in a being
or an Idea, but in the very law according to which the members of the series
follow one another. What is called cause is thus itself to be demonstrated
as object, as member in experience, just as well as the object that was to
be explained. It is the standpoint of the more recent empirical science, as
it was indicated by \emph{Kepler} and \emph{Galileo} through the demand for
a cause demonstrable in experience (vera causa).

The differences stood unexplained in Animism and Platonism. How could very
different, perhaps mutually conflicting, objects and properties have their
ground in the same god or in the same Idea? The more rich in content and
comprehensive experience became, the more the sting of this question had to
be sharpened, and attention had to turn toward the mutual connection of the
differences. The urge for recognition can nonetheless constantly find
satisfaction, if it succeeds in re-finding the same events when the same
conditions are present. The earlier experiences comprised in a certain
sense the new ones, if these are due to the same law as the former.

Yet it is not merely the urge to overcome in thought the multiplicity and
change of objects that operates at the transition to Positivism: it is also
the wish for power---the wish that in magic expressed itself in immediate
acting, in Animism more mediately and on the presupposition of a certain
resignation, and in Platonism faded into intellectual contemplation. Power
over nature is now won by understanding of the laws according to which
nature operates and by attempts, in the single case, to bring about the
law-determined conditions.---

Is now this form the definitive one, or will there be able to come new
stages in thought's great history of development? To this, from the nature
of the case, no answer can be given. But we can see---and it is an insight
of the very greatest significance---that the positivistic tendency itself,
the standpoint of strict empirical science itself, rests on certain
presuppositions concerning thought's relation to existence, so that, even
if this standpoint should become the definitive one, problems remain with
respect to the ultimate grounding of it. It is at this point that
philosophy in the more recent time has taken up its work. The seventeenth
century's great systems sought to find out what world-view resulted when
the new standpoint, which I have here for brevity's sake called the
positivistic, was strictly carried through. The eighteenth century's
philosophy then went back to the presuppositions for the whole standpoint
and subjected them to a critical testing. The nineteenth century's
philosophy made partly an attempt to return to Platonism, partly to stop
the thought-work there where the positivistic principle did not lead us
further. It will be a chief task for the continued thought-work to carry
further the treatment of the problems that through the purgatory of the
great discussion have shown themselves unavoidable for every waking
reflection. A contribution to this it will be attempted to give through a
closer investigation of thought's modes of operation and tasks. But first
there is yet another opposition to illuminate within the history of
thought, an opposition that indeed reminds one of the opposition between
Platonism and Positivism, but does not quite coincide with it and in any
case illuminates it from a new side.

\section*{B. Ancient and Modern Thinking}
\addcontentsline{toc}{section}{B. Ancient and Modern Thinking}
\label{sec:antik-moderne}

46. The opposition between the second and the third stage, as it has been
described in the foregoing, coincides in part with the opposition between
ancient and modern thinking. But it will serve for the closer illumination
of the nature of human thought if we give a more thorough characterization
of this opposition. It can be summed up under two points of view. Ancient
thinking has a predominantly formal character; on this rests in part its
great and lasting significance. And next, and in connection therewith, it
is characterized by the presupposition that the unchangeable stands, not
only in a logical but also in an ethical respect, in value above the
changeable.

47. The opposition between formal and real knowledge appears already in
\emph{Parmenides} of Elea, and so sharply that he recognizes only formal
knowledge as real knowledge. Only those propositions that can be formed as
pure identities can be the object of true conviction, and only such
propositions can express what truly is. Being must be unborn, imperishable,
unchangeable, indivisible, concluded in itself. Of the manifold,
changeable, and perishable no knowledge can be had, only ``opinions of
mortals.''

Here appears, with the thought-energy that distinguished the Greeks, the
principle of identity, the demand for thought's unity with itself, the
strict recognition, as the first principle of all thinking, as the formal
ideal to which thought seeks to bring all cognition as near as possible.
Parmenides would not even concede that differences and changes could set
serious problems, whether such problems can be solved or not. Only
conjectures, not true science, are here possible\footnote{Cf. my treatise
on Parmenides in \emph{Mindre Arbejder}. Anden Række, p. 141--149.}.

Two other Greek thinkers, the one Parmenides' predecessor, the other his
successor, made, on the contrary, attempts to bring identity, the highest
form and norm, into closer connection with the objects' difference and
change. \emph{Heraclitus} maintained that in all the constant streaming and
under all the conflicting oppositions there is nonetheless something that is
constant, namely the law, existence's eternal fundamental thought (λόγος),
which subsists under the constant change of all phenomena. And
\emph{Democritus} set up the proposition that one reaches a true
understanding only when one disregards all the qualitative differences that
the senses show us, and explains everything by the reciprocal action of
homogeneous parts. His atomic theory is Parmenides' doctrine of identity
applied to each single one of existence's smallest parts.

While Parmenides denied the reality of what falls outside pure identity,
Heraclitus and Democritus set themselves the task of tracing the given
differences and changes back to identity, partly the law's identity under
different conditions, partly the elements' identity in different
connections.

In \emph{Plato} appears the same conviction as in Parmenides of the world
of identity as the true world, only that he finds the identical in the
general concepts or Ideas that hold for the sense-world's multiplicity and
change, and that for him are not merely human thought-products, but the
expression and content of what truly is. The logical point of view, which
more hiddenly made itself felt in his predecessors, and which he had in
Socrates' school been trained to use, came in him to clear consciousness.
What filled his mind with ever new enthusiasm was the fact that concepts
can be formed which, in their unchangeableness, their identity with
themselves, hold for large groups of differences. The victory of likeness
over the differences in human thought was for him a testimony that the
world of Ideas was the true world. All cognition was recognition of the
Ideas in the sensory world, and the highest spiritual capacity was that by
which, from the points of likeness within the sense-world, a world of
thought is formed, so that an ascent to the true reality becomes possible.

The Greek concept of science is predominantly formal. True science was for
the Greeks logic and mathematics; their empirical science for the most part
did not get beyond the collection of more or less exact observations.
Democritus and Plato are the classical Greek thinkers, and against them
both there was in antiquity directed the objection that they set aside and
rejected what the senses teach (τὰ αἰσθητά), and kept merely to what pure
thought grasps (μόνοις ἑπόμενοι τοῖς νοητοῖς)\footnote{\emph{Sextus
Empiricus} ed. Bekker. p. 299.---Cf. my treatise on \emph{Forholdet mellem
Platon og Demokritos} in Nordisk Tidsskrift for Filologi. 1909.}. Both
Democritus' atoms and Plato's Ideas were pure objects of thought,
constructed from the demand for identity. Both for the founder of the
atomic theory and for the founder of the doctrine of Ideas the
sense-qualities conflicted with thought's pure identity.

\emph{Aristotle} maintains---in opposition to Plato's doctrine of Ideas and
Democritus' atomic theory---the reality of quality and of individual
existences. It was thereby established (without Aristotle's fault, for
millennia) that the science of empirical objects could consist only in
classification. Aristotle himself did not fail to appreciate formal
science; he is, after all, himself the organizer of formal logic, probably
influenced by the development that mathematics had gained; but logical
thinking and qualitative comparison continue in him to stand side by side.
He saw well that movement, change of place, lay at the basis of all other
change in nature---of qualitative change, of coming-to-be and
passing-away, and of increase and decrease; but he made no attempt to carry
through this thought, which would have led him to the point where modern
natural science begins.

It was two different, but both deep-lying, presuppositions that hindered
the Greeks from letting formal and real thinking cooperate in a fruitful
way.

From the Orient the Greeks had inherited the presupposition that the
star-world's movements took place with absolute regularity. It was this
unchangeableness, in opposition to the earthly (sublunary) world's
mutability, that first opened the view for the possibility of science. Here
one found the self-identical being realized. And one found then, in
thought's firm and constant forms, expressed in the general concepts, a
counterpart to it. Both the visible and the invisible unchangeableness were
embraced with religious awe, as a refuge under the change of things. A
peculiar religious type makes itself felt here\footnote{Cf. my
\emph{Religionsfilosofi}\textsuperscript{2} p. 51--55; 125--128.}---but in
remarkable connection with a physical and a logical ideal. The
presupposition that the unchangeable is better or more perfect than the
changeable the Greeks did not give up, any more than the Indians gave it
up.

In a peculiar opposition to this tendency in Greek spiritual life stands the
Greek sense for the concrete, the individual, and the intuitable, the
living urge for plastic formation of the spiritual content. Not only
mythology's gods and art's figures, but also thought's Ideas stood as clear
images, whose self-resting firmness neither thought's analysis nor
experience's increasing fullness could dissolve. The single observation and
the single individual stood as something apart, both in opposition to
conceptual thinking. In Aristotle this opposition culminates; for just as
convinced as he is that only the universal is object for thought and
science, just as convinced is he that only the individual has existence or
being. The individual is that which is the object of all assertions, but
which can never itself be used as predicate. (Cf. 34.)

48. The Renaissance's thinking led beyond the opposition at which Greek
science had stopped. The multiplicity and mutability that life and
experience present were now conceived precisely as means to come to know
nature rightly. \emph{Nicolaus Cusanus} and \emph{Giordano Bruno} showed
that thought itself works its way forward through oppositions, operates by
constantly setting limits and overstepping them in order to find new
limits, so that no opposition, no barrier, can be established once and for
all. And experience itself now led to the discovery that one had drawn in
far too external a way the boundary between the world of unchangeableness
and that of changeableness. \emph{Tycho Brahe}'s ``new star,'' which so
suddenly emerged in the sphere of unchangeableness, moreover presented
changes both in brightness and color, and his investigations of the comets
showed that these apparently so capricious beings did not move about in the
sublunary world, but on the contrary traveled through the ether-world
higher up. And Galileo stated outright that the earth appeared to him
precisely ``most noble and admirable'' (nobilissima e ammirabile) on
account of the many changes that take place on it. He cannot in itself find
any perfection in the unchangeable. The standard for value becomes another.
It is a matter not merely of ascending to the unchangeable world of eternal
Ideas, but of finding the laws according to which the changes in the world
of experience take place. As one had found changes in the world of
``unchangeableness,'' so one also sought to find unchangeable laws in the
world of change. In antiquity \emph{Archimedes} had made the doctrine of
equilibrium an exact science. Galileo began to subject the doctrine of
movement to an equally exact investigation. Ancient science had essentially
been statics and morphology; now dynamics gradually came into its own in
the different domains.

If one was to find an unchangeable law through the multiplicity of changes,
one had to keep to the properties of things and events that make
quantitative determination possible. Counting, measuring, and weighing had
now to become means of thought to a far higher degree than before. One had
then to keep to properties such as magnitude, figure, and movement, and try
to determine the appearance and changes of the qualitative properties
through their relation to them. The opposition between formal and real
thinking now became an opposition between quantitative and qualitative
knowledge, and in this form a more intimate cooperation between the two
kinds of thinking was possible. Speculation and observation, deduction and
induction, could now work into each other's hands, in that the inferences
drawn from general propositions could, by reason of the quantitative
formulation, be tested in an exact way by their relation to what experience
showed, while conversely, from experience, ever more exact general
propositions could be formulated. Thinking could now describe an arc from
experience through inference to new experience. Ancient thinking really
knew only the first part of this arc. When it had found a starting-point
for its abstraction, it ascended to the world of Ideas, the world of
logical and mathematical forms, and shrank from descending again to apply
and test the thought-results won in the world of experience. It believed in
apotheoses, but not in incarnations.

It was now no longer the ideal to order qualities under general concepts of
different degrees and thus form classificatory systems, but to find
relations of dependence between changes in the world and express them in
definite laws. Explanation steps in place of mere classification, and
quantitative description forms the transition from the latter to the
former. Logic and mathematics enter as tools into the work of real
cognition; they no longer form a world subsisting by itself, but their
principles and laws are woven into the world of thought that builds on the
ground of experience. Platonism and Positivism strive to enter into as
close a connection as possible.

49. Purely logically, the change that now takes place in thought's task
and mode of working shows itself in the fact that judgment, not the
concept, becomes the most important thought-form. As we have earlier seen,
the opposition between judgment and concept means only a different weight on
two sides of the same thought-process (24). But the opposition between
ancient and modern thinking can very well be expressed thus, that the
former uses the formation of judgment to win concepts, the latter uses
concept-formation to reach judgments. The necessity of forming concepts was
Socrates' great discovery, which gave Plato enough to think about
throughout his whole life, and which led to the formation of a world of
Ideas as the expression of true reality. But for modern thinking the point
is to find judgments that express definite relations between definite
changes. Through such judgments---by virtue of the relation of reciprocal
action between concept and judgment---concept-formation will in turn be able
to attain greater perfection.

In a characteristic way this tendency appears in the fact that two of the
Renaissance's thinkers interpret Plato's Ideas as the laws of things.
\emph{Bruno} teaches that the task is not merely to form general concepts
that express what is common to several phenomena, but to find expression
for the real connection between different phenomena (e.g. the parts of an
organism) among themselves, since only by such a connection can one
understand the nature of the individual phenomena. \emph{Bacon} expresses
himself in a similar direction when, for the understanding of a phenomenon
(e.g. heat or organic life), he demands that one must not remain standing at
abstract concepts, but shall try to find the continuous, often hidden
process (latens processus---per minima) by which things get their definite
constitution\footnote{\emph{Den nyere Filosofis
Historie}\textsuperscript{2}. p. 126; 194--196.}.

With this there now also follows a change in the conception of inference.
As long as one had merely to do with a system of super- and subordinate
concepts, it was natural to build the theory of inference on the principle
that what belongs under a concept also belongs under the concepts under
which this concept belongs. Inference then stands as a process of
subsumption, culminating in a highest concept. But when weight is laid on
the fact that concepts as well as judgments are to enter as members in the
ever-continued thought-movement, it will be natural---as has also (see 33)
shown itself to accord best with the essence of judgment---to conceive
judgment as the expression of a relation of identity, so that inference can
take place by different judgments being combined by means of substitution.
The purely logical thought-movement thereby approaches the mathematical,
which proceeds through equations and substitutions. \emph{Leibniz}'s
identity-logic is thus a characteristic expression of the more recent
time's thinking---even though a later time's logic has in several respects
had to strike out on other paths\footnote{Cf. on this \emph{Couturat}:
\emph{La Logique de Leibniz}. p. 386 f; 433.---Couturat overlooks, however,
that all relations can be expressed as predicates.}. It must especially be
emphasized that Leibniz did not bring his logic into connection with a
doctrine of categories, and thereby came to overlook that there are many
other relations than identity that can be of great significance for the
thought-work (even though they can all be purely formally expressed by
predicates in identical judgments). In a later section we shall have
occasion to discuss the relation of identity in comparison with other
relations. But it is extraordinarily characteristic that identity for
Leibniz does not stand as an object of mystical beholding, like the ``Idea
of Likeness'' for \emph{Plato} (\emph{Phaedo} p. 74), but as a tool of
thought. This shows itself already in his definition of identity as the
possibility of substitution: identical are the concepts that can be put in
place of one another without the validity being violated (salva veritate).
There lies in this a constant summons to test identity experimentally, to
see whether it can be maintained in the individual special connections in
which the concepts occur. It is in the world of multiplicity that the
thought-work is to be exercised, and the question is how far relations of
identity can there be demonstrated.---

While Platonism was inclined to confuse conceptual determination with
causal explanation and could thereby lead to assuming just as many
``forces'' as there were kinds of objects, in modern thinking the concept
law stepped more and more in place of the concept force, or rather this
concept is subordinated to the former. In the Aristotelian doctrine of
nature, nature formed a series of stages, within which every stage
contained the possibility or the force (δύναμις) for the following one. But
about the possibility or the force one then learned nothing other than
precisely this, that it was---possibility or force for the following one.
There were then just as many possibilities as there were realities. The
qualitative differences were explained by just as many qualitates occultæ.
In scholasticism, which in many domains maintained itself far beyond the
Middle Ages, one lived high on that kind of word-explanation. When it is
first and foremost a matter of finding a law, that is, a judgment that
states one phenomenon's dependence on another, one need not indeed give up
the concept force, but this gets the definite meaning of denoting that in
the preceding phenomenon which contains the ground for the following
phenomenon's setting-in---the peculiarity of a phenomenon that from it one
can infer another's setting-in. Leibniz is clear that the concept force
contains no explanation, but only a summons to seek why the transition to a
new phenomenon takes place precisely under these definite conditions. The
concept of force is constructed in each single case on the basis of the law
that has shown itself to hold for such a transition\footnote{\emph{Den
nyere Filosofis Historie}\textsuperscript{2}. I. p. 346.---In a fragment
brought forward by \emph{Couturat} (see Bulletin de la Société Française de
Philosophie. II. p. 88 f) Leibniz says: Vis activa nihil aliud dicit quam
id, quo sequitur transitio, sed non explicat in quo consistat.}.

50. The new concept of science could not immediately be applied in all
domains, and it became of decisive significance for the philosophical
treatment of problems in the centuries immediately following the
Renaissance that there were some domains that were especially suited to be
treated according to its demands. These domains then naturally came to
stand in the foreground in the scientific conception of the world. It was
the doctrine of movement and mechanics that determined the conception of
nature in the seventeenth century's great systems. In these domains the
differences of quality and individuality stepped back before the universal
laws. \emph{Leibniz} here forms the transition to a new period, in that he
so strongly brings out the concept of individuality, and demands not only a
continuous connection and law-governedness demonstrated between all
existing beings, but also within the little world that every single
individual being constitutes. Within the individual beings each by itself
too there take place changes, and if we could find the law for these
changes, we would thereby have determined the individual being's
peculiarity: La loi du changement fait l'individualité de chaque substance
particulière (Lettre à Basnage 1698).

By degrees, as we in the series of the sciences pass over from the formal
to the real, the more difficult the cognition of law becomes, the more we
stop at mere experience without being able to carry through the reciprocal
action of deduction and induction. The logical law of the inverse relation
between concepts' content and extension shows itself here in its great
significance: the simpler the objects we face, the easier it is to form
concepts and laws; the more concrete and composite the objects we have
before us, the more we stop at mere descriptions and at characterizations
of individual peculiarities. Biology, psychology, and sociology present in
increasing degree difficulties for the development of a cognition of law.

Modern thinking has not disowned Platonism. As the Greeks sought universal
Ideas, so modern science seeks universal laws. And herein shows itself the
unity of human thought under all forms. Yet it would not be right, with
\emph{Henri Bergson}, to assert that the modern concept of science is only
different in degree from the ancient. Bergson grounds this conception on
the fact that, just as the ancient ``Ideas'' meant general concepts about
things, so the modern ``laws'' mean general concepts about changes. Modern
thinking has, therein one finds the advance, taken up time as an element.
But the main thing for it is nonetheless always to be the universal and
abstract, not the single and individual\footnote{\emph{Bergson}:
\emph{L'évolution créatrice}. Paris 1907. p. 359 f.}. To this it must be
remarked that modern science, the more it becomes conscious of its proper
goal, sees it as its task to use the general concepts and laws as means and
ways to understand the individual and historical totalities. It is the
individual totalities whose understanding stands as science's last goal,
and the poorer-in-content, but at the same time more comprehensive,
thoughts are to serve to lead in the direction of this goal. The ``law of
change that constitutes the individuality of the single being'' is the
highest ideal of modern thinking. Therein lies a difference of principle
between the ancient and the modern conception of science, at the same time
that the world of problems becomes far greater than it could be for ancient
thinking. The difference thus lies not only in the way, but also in the
goal.---The doctrine of categories will give occasion to take up this
question again.

%% ============================================================
%%  III. THE FORMS OF THOUGHT (THE CATEGORIES)
%% ============================================================
\chapter*{III. The Forms of Thought (The Categories)}
\addcontentsline{toc}{chapter}{III. The Forms of Thought (The Categories)}
\markboth{III. The Forms of Thought}{III. The Forms of Thought}
\label{ch:former}

\begin{center}---\end{center}

\section*{A. The History and Method of the Doctrine of Categories}
\addcontentsline{toc}{section}{A. The History and Method of the Doctrine of Categories}
\label{sec:kategorilaere-historie}

51. Through the preceding psychological and historical presentation we
have come to know the different stages by which human thought works its
way forward. Through the work we here, as everywhere, come to know the
energy. Sensation, intuiting (with association), and judging stood,
psychologically regarded, as the most important stages, and the rise from
stage to stage meant an ever clearer consciousness of objects and elements
and their mutual relation. In a historical respect the relation between the
objects and their interpretation was the main thing, and the ideal here was
that interpretation takes place through the objects' mutual illumination of
one another, an ideal that, however, in turn pointed back to definite
presuppositions.

It will now be the task to give a survey of the fundamental forms with
which thought works, and which condition presuppositions to which it is by
its nature bound. Such fundamental forms have from old times been called
categories.

The word category denotes, according to a usage established by
\emph{Aristotle}, assertion, predicate. There are therefore, in the widest
sense of the word, infinitely many categories. But in philosophy one
understands by category a fundamental predicate, a fundamental concept, and
the doctrine of categories sets itself the task of gathering and ordering
the predicates that with greater or lesser degree of consciousness are
applied by thinking in the different domains where it works, and tracing
them back to certain mutually kindred and yet each individually peculiar
fundamental forms.

The doctrine of categories occupies a central position in philosophy. It
points on the one side back to the study of the cognition won, as the
psychology of cognition on the one hand and the history of science on the
other present that development, on the other side forward to the
philosophical treatment of problems, which investigates how far the
assertions we have at our disposal can lead us.

The categories come into being involuntarily during thinking's work at
satisfying the objects' demands. They arise sporadically, to meet
individual special demands or answer momentary questions, and only later
can the question be set whether their significance is limited to a single
domain or a single situation, or can extend further. The point is then to
gather them, to demonstrate their connection among themselves and with the
nature and tasks of human thinking, and to trace them back to as few and
simple forms as possible.

The most significant attempts in this direction have, at intervals of
millennia, been made by \emph{Aristotle} and \emph{Kant}. All other
attempts that exist, also the attempt to be made here, relate to these
models as modifications that strive to give closer grounding and more exact
connection. Both masters of the doctrine of categories started from an
investigation of judgments---quite naturally, since predicates appear only
in judgments. The fundamental predicates must be the determinations that
constantly recur in the judgments that objects give us occasion to pass. An
analysis of the different assertions a judgment can contain, or of the
different kinds of judgment that can be passed, was necessary in order to
find the fundamental predicates. Aristotle goes the one way, Kant the
other.

52. Judgment is, as we have seen, not the first form in which objects
appear to us. Judgment presupposes intuiting---sensation, memory, or
imagination. In intuition the objects are involuntarily ordered in us and
for us, and already at the coming-into-being of intuition there operates the
psychic energy that can later appear in the formation of judgment. In
intuition, whether it is sensation, memory, or imagination, a multiplicity
is involuntarily ordered; a form of wholeness is brought about. The arising
of every new sensation is an interruption of consciousness's given state,
and work is now done involuntarily at ordering the new object or element
(the new distinguishable thing) into the whole connection. As long as
possible consciousness dwells in intuiting---in the three named forms,
which, for that matter, can only at a later stage be distinctly
distinguished from one another. But when there appear changes that can no
longer be equalized in the form of resting intuition, there begins a work
at a new kind of connection. Reflection is awakened, attention moves from
object to object, from element to element, brings them out each by itself
in its peculiarity, and seeks through such analysis to find a new ordering
of the objects. By this way judgment can step in place of intuiting.

During the formation of judgment, as we have seen, the thought-movement
goes from a given and presupposed, more or less indefinite basis, the
subject, over to a more definite determination, the predicate. We think in
predicates, or rather by predicating (asserting). The formation of judgment
begins with the one leg of thought's compass standing in or being set on a
single point, and a place then being sought where the other leg can be set.
A doctrine of categories becomes possible through all special
determinations or predicates being able to be subordinated under a few
fundamental predicates that recur again and again. The categories are the
chief forms in which decision- and determination-questions (cf. 26 and 31)
are preserved.

When an object emerges in my sensation, in my memory, or in my imagination,
I will involuntarily be able to raise the following questions:

Is it one with something I know?

Does it resemble something I know?

If it is different from something I know, is it then a difference in kind,
in time, in number, in degree, or in place?

Does it contradict something I know?

Can it be derived from something I know, or can something I know be derived
from it?

Is it cause of something I know, or is something I know cause of it?

Is it part of something I know, or is something I know comprised in it as
part in whole?

Through what different intermediate forms has it become what it is?

What value does it have? Is it a means for something valuable that I know,
or does it perhaps denote a higher purpose than the valuable I have hitherto
known?

In the preceding psychological and historical investigations we have in
various places come upon objects that can give occasion to such questions.
But whether the list of questions gives us a direction to all categories
must remain open. It will on the whole be impossible to demonstrate that any
list of categories is complete, although both Aristotle and Kant and most
of their successors have considered it possible to set up concluded systems
of categories. We form our judgments on the basis of intuiting, and within
intuition new objects and elements can always appear, which can require new
judgments and give occasion to new predicates. The involuntary, often
obscure, basis for the analyzing thought-work can shift, without our being
able to say in advance how far this shifting can go---and wholly new
questions can be called forth thereby. And even if the basis does not shift,
proof can never be given that it is exhausted through the judgments we have
with full consciousness and exactness formulated. Reflection constantly
draws from a source whose spring is hidden\footnote{Cf. my treatise
\emph{Filosofien og Livet}. (Mindre Arbejder. Anden Række.)}. History shows
that categories can arise and pass away by reason of the shifting of the
basis, or because hitherto unheeded objects or elements draw attention to
themselves. A presentation of the doctrine of categories can then have
validity only for a certain stage in the development of cognition. It has
its essential significance in being a stocktaking that can lead to a clear
understanding of the problems.

There must (cf. 3--4) indeed be a distinction made between the basis and
the reflection that is directed toward it. Reflection can be compelled to
make distinctions and establish differences that cannot without further
ado, or in the same way as they lie before reflection, be said to be
present in or ``lie in'' the basis. Of course there must in the latter be
occasion and justification for setting up those distinctions; but they can
there make themselves felt in another form. There is analogy, not identity,
between intuition and judgment. They are both forms of the psychic energy
that at all stages strives for connection and equilibrium; but they are two
different forms of this energy.

Only in the clearer conscious life can there, e.g., be distinguished, as
\emph{Kant} does it, between matter and form, or, as \emph{Stumpf} in the
most recent time does it, between phenomena and functions. (Cf. 5--6.) In
the involuntary mental life this difference does not come forth. One will
perhaps say that the functions are present, even if they are not specially
noticed and not set in opposition to the phenomena. But it must in any case
make an essential difference whether a state appears as an undivided whole,
or whether analysis has brought out differences---whether it (to use
Stumpf's expression) appears ``unzergliedert'' or ``zergliedert.'' There is
a difference of quality between these two forms, even though we can
understand them as forms of one and the same energy at different stages.

It is this difference that makes it that the doctrine of categories,
however important its connection with psychology (as well as with the
history of science) is, cannot without further ado be called a part of
psychology. It stands between this and epistemology proper.---In sharp
opposition to the psychologism that effaces the difference between the two
disciplines, there appears in the history of thinking a constant tendency
to give the doctrine of categories without further ado a metaphysical
significance. One then sees in the fundamental concepts of the human spirit
an immediate revelation of existence's innermost essence. The Platonic
doctrine of Ideas is the greatest example of this assumption, which later
appears under various forms, not only in philosophers like Spinoza and
Hegel, but also in natural scientists like Newton (especially in his
concept of spatium absolutum as sensorium dei).---The doctrine of
categories can be called a descriptive epistemology, an introduction to the
special discussion of the problem of cognition, and it takes up as such
other points of view than psychology and metaphysics.

53. Some features of the history of the doctrine of categories will serve
for the illumination of its significance. I begin with the two great
masters of Greek thinking. The opposition between Plato and Aristotle
appears here in a characteristic and instructive way.

For \emph{Plato} it was the greatest fact of thought and life that it was
possible to form concepts that hold for multiplicities of objects. His
philosophizing was a constant ascent to the world of general concepts, of
Ideas. (Cf. 44.) Starting from the objects' obscure, chaotic basis, he
formed a world of predicates. In the predicates, not in the subjects, the
world of objects, his thought found its home. There expresses itself here
the correct insight that the predicates are thought's proper work and
property. Only, Plato overlooks that the objects are the constant
presuppositions, and that during the formation of judgment a constant
reciprocal action between subject and predicate takes place, so that in the
judgment passed they steadily determine each other. (See 33--35.)

Not all Ideas stand for Plato on the same line. Repeatedly he comes back,
in his dialogues, to a class of Ideas that show themselves to be
inseparable from thought's activity, whatever special objects this
concerns. He is, as we have earlier seen, one of the predecessors of the
synthesis-theory. The soul must be able to exercise a combining, as surely
as it can have sensations of different kinds and compare them among
themselves. And by this very combining and comparing certain predicates are
applied, which express the relations that are ascertained: Being and
Not-being, likeness and unlikeness, identity and difference,
number-concepts. It is by the application of the forms expressed in these
concepts that thought ascends from the objects to the Ideas, and the named
concepts are therefore themselves the most fundamental Ideas.
(\emph{Theaetetus} p. 185--187.) And he especially emphasizes judgment as a
connection (συμπλοκή) by which the Ideas are determined in their mutual
relation (\emph{Sophist} p. 259).

In his last, Pythagoreanizing period he occupied himself with ordering the
fundamental concepts in a series analogous to the number-series. But of
this attempt little certain is known.---A doctrine of categories proper is
first found in the thinker who introduced the very word category.

Here and there in \emph{Aristotle}'s different writings some concepts are
applied that denote the most general points of view that can be used in
more special investigations, no matter what these concern. Nowhere,
however, has he---as Plato did---brought them into inner connection with
thought-activity itself, and in the writing (\emph{Categories}) that
(whether it directly derives from Aristotle's hand, or only indirectly)
especially treats them, they are placed---ten in number---in a very
external way, in a purely catalogue-like manner side by side. Probably,
however, he came to them through analysis of a number of judgments.

First in the series of categories stands the concept thing (essence,
substance: οὐσία; τὸ ὑποκείμενον). And yet it is really no category---for
it cannot be used as predicate, cannot be asserted of anything. It denotes
the basis, whose closer determination is sought through the other nine
categories. In itself it can only be denoted negatively, namely as that
which is not asserted of anything---and then one must, for that matter,
point back to intuition. Aristotle has here aptly characterized thought's
constant basis, that which is first to be determined through the
thought-work itself and therefore prior to this cannot have any
determination. Every positive determination would already be a predicate,
and the ``essence'' was to be precisely that which was to be determined by
predicates. In opposition to Plato, Aristotle lays great weight on this
constant basis, while Plato forgets it in his enthusiasm for the Ideas,
which were yet really its predicates. Here shows itself the realistic
character of Aristotle's philosophy. He has a sharp eye for the fact that
thought always works on objects at hand.

The object, the given, is for Aristotle always an individual (ἄτομον καὶ ἓν
ἀριθμῷ. \emph{Categ.} c. 5). Only the individual exists. Thought has here
one of its limits; it needs a starting-point that cannot itself be used to
determine anything else! Further back we cannot come.---Analogies to this
concept appear in the absolute atomic theory, in Spinoza's Substance, in
Leibniz's Monads, in Kant's ``Ding an sich,'' as well as in the higher
religions' concepts of God. (Cf. 34--35.) Aristotle demands absolute
conclusion of thinking. The one limit thus lies in ``Substance''; the other
lies in the principles (ἀρχαί) that condition all judging and inferring and
therefore cannot themselves be proved (\emph{Analytica} post. I, 22).

The (other) nine categories are: Quantity, Quality, Relation (πρός τι),
Acting, Being-acted-upon, Place, Time, Condition, and Having (the
possession of a property). This list Aristotle considers complete, although
he does not demonstrate the necessity that precisely these, neither more
nor fewer, exist. He considers them irreducible, independent points of view
that cannot in turn be brought under a common point of view, but only stand
in a relation of analogy among themselves. More fully he discusses only the
first three (Quantity, Quality, and Relation).

Of special interest, especially in comparison with Plato's doctrine of
categories, is the category of Relation (relational determinacy). As
examples he names equality and inequality of magnitude, numerical
relations, cause and effect, cognition and the cognized. These are
relations that have their root precisely in the combining and comparing
thought-activity taken as basis by Plato\footnote{See more closely on the
concept of relation in Aristotle: \emph{Trendelenburg}: \emph{Geschichte
der Kategorienlehre}. p. 117--129.---G. \emph{Grote} (\emph{Aristotle}. I.
p. 115--122) shows that relation lies at the basis of all nine
categories.}. Both Quality, Quantity, and the other nine categories could,
on closer consideration, be brought under the category of relational
determinacy, and it will later show itself that in the more recent time
attempts have been made to set up Relation as the first, most comprehensive
of all categories. Such an attempt Aristotle did not make, however much
most of his examples point in that direction. And the ground for this must
surely be sought in the fact that he did not (at least not in the writings
we have) attempt to bring the individual categories into natural connection
with thought's mode of operation as a whole.

The changes that the doctrine of categories underwent in antiquity after
Aristotle it is not necessary for our purpose to enter into. For spiritual
development it was, however, a characteristic sign of the times that the
last eminent thinker of antiquity, \emph{Plotinus}, limited the
Aristotelian categories to holding only for the sensory world. In the
supersensory world the differences and oppositions that the categories
denote fell away, as well as the very difference between the essence and
its predicates. All is there in inner unity and harmonious fullness. Since
Plotinus nonetheless wishes to think about this supersensory world, he is
compelled to use some of the categories, but they then get a more or less
mystical significance\footnote{Cf. \emph{Arthur Drews}: \emph{Plotin und
der Untergang der antiken Weltanschauung}. (1907). p. 158--173.}. In a
similar way Christian theology placed itself. \emph{Augustine} reproaches
himself in his \emph{Confessiones} (IV, 16) for having in his youth been so
occupied with Aristotle's doctrine of the categories (decem prædicamenta),
since they cannot, after all, serve for thinking God, in whom no
distinction can be made between essence and properties. God cannot be
subordinate (subjectum) to his own properties; one cannot, e.g.,
distinguish between him and his goodness. The difficulty that all
theological thinking has to struggle with here came clearly forth. And it
was the Neoplatonists who had first felt it.---

Apart from this, the Aristotelian doctrine of categories prevailed for
millennia, and its after-effects can still be traced.

54. The next great and epoch-making attempt at a doctrine of categories was
made by \emph{Kant}. For details of his theory I must refer to my treatise
\emph{Om Kontinuiteten i Kant's filosofiske Udviklingsgang} (Vid. Selsk.
Skr. VI) and to my work \emph{Den nyere Filosofis Historie}. Here I shall
only bring out some chief points.

With full consciousness Kant derives the fundamental predicates or
categories from the judgment-function. And in that he conceives judgment as
arising from a connecting operation, a synthesis, and furthermore sees in
synthesis an operation that lies at the basis of all consciousness, whether
we notice it or not, he has brought the doctrine of categories into inner
connection with the whole nature of human consciousness.

Therefore Kant is also clear that the categories are provisionally only
forms of our thought, while Aristotle immediately conceived them as
predicates of existence (κατηγορίαι τοῦ ὄντος. \emph{Categ.} c. 2). After
having set up the categories, Kant then passes over to investigating how,
with what right, and to what extent they can be said to lead to valid
cognition. He thus distinguishes definitely between the fact that we work
with certain forms of thought, and the validity of the cognition that can
be won by their help.

Kant was convinced that there was a definite number of categories, which
follow with necessity from the nature of thought. He supports this
conviction on the fact that there is a definite number of mutually
kind-different classes of judgments, and that to each such class a category
must correspond. Thus, e.g., the negating judgments were to form a class of
their own, likewise the quantitative judgments. According to what we have
earlier developed (26; 32), there can here be no question of a difference of
kind. As regards the other kinds of judgment (e.g. the problematic and
hypothetical) too, it can be shown that it is linguistic, not purely
logical, grounds that lead to distinguishing between them. Kant's number
twelve is therefore just as little grounded as Aristotle's number
ten\footnote{Cf. already G. E. \emph{Schulze}'s critique in \emph{Kritik
der Theoretischen Philosophie} (1801). II. p. 291--333.}.

Of great interest, on the other hand, is that Kant himself demonstrates that
there is a common characteristic mark for all categories, namely the concept
of continuity, which in turn appears in two forms, in the concept of
magnitude and in the concept of cause. Under the concept of magnitude belong
again concepts concerning extension in time and space, and the concept of
intensive magnitude (degree), while the concept of cause, besides the
cause-relation itself, also comprises the concepts of possibility,
actuality, and necessity.

Kant's doctrine of categories can accordingly be schematized in the
following way:

\begin{center}
Synthesis (Continuity)

\medskip
\begin{tabular}{p{0.42\textwidth} p{0.42\textwidth}}
\multicolumn{1}{c}{Magnitude} & \multicolumn{1}{c}{Causality.}\\[2pt]
a) Unity, Plurality, Totality. (Extension). & a) Subsistence and Change.
Cause and Effect. Reciprocity.\\
b) Reality, Negation, Limitation. (Intensity). & b) Possibility, Actuality,
Necessity.\\
\end{tabular}
\end{center}

The concept of continuity hangs closely together with the concept of
synthesis. For the closer the connection between objects or elements, the
more intermediate links can be found between them, the more complete and
exact the synthesis becomes. From his doctrine of categories Kant therefore
derives the following principle: ``To allow nothing in the empirical
combining that could hinder the continuous connection of all phenomena; for
only thereby does that unity of experience become possible within which all
observations have their place.''\footnote{\emph{Kritik der reinen
Vernunft}\textsuperscript{2}. p. 282.} Experience is for Kant the connection
within which every single object is a member determined by all other
members. And experience in this sense becomes possible only by reason of
consciousness's original combining energy. The fruitful thing in his way of
considering lies in the summons it contains to ever-renewed inquiry in order
to find new members in the connection of experience and to determine this
connection in an ever more exact way.

Besides continuity there is yet a concept that consistently ought to have
had a place in Kant's doctrine of categories, namely the concept of Relation
(relational determinacy). It is seen from Kant's notes to the ``Kritik der
reinen Vernunft'' (the ``Reflexionen'' published by Benno Erdmann) that it
was by occupying himself with the concept of relation and its significance
for thinking that he became clear about synthesis as the fundamental form of
all consciousness and all thinking. He states thus not only: ``Our reason
contains nothing other than relations,'' but also: ``The category of
relation (of the unity of consciousness) is the most important of all. For
unity really concerns only relation.''\footnote{See \emph{Om Kontinuiteten i
Kants filosofiske Udvikling}. (Vid. Selsk. Skr. 6. Række. p. 53. kfr. p.
27).} He has thus assumed a very close connection between the concepts
synthesis and relation. When he has not entered them among his categories,
the ground has been that he regarded them as presuppositions for all the
special categories. Therefore the significant utterance cited has not gained
influence on his systematic presentation.

An essential objection to Kant's doctrine of categories is that he narrows
the concept of category too much.---In the first place, space and time are
supposed to be not categories, but ``forms of intuition.'' But in our
cognition we do not stop at the mere staring at what spreads out in space and
time; it is the definite relations of space and time that we have to do with,
relations of a peculiar kind that do not coincide with any of the relations
that Kant's categories denote, but that must nonetheless be reckoned among
the fundamental relations we work with in our thinking. Kant on this point
makes too sharp a distinction between ``intuition'' and
``understanding.''---In the second place, he does not reckon the purely
logical relations identity, difference, and likeness among the categories,
although they, as Plato already clearly saw, are the simplest and most
necessary of the forms with which our thought works, and although we must ask
about the validity of these forms just as much as about the validity of
concepts like magnitude and cause.---In the third place, he distinguishes
sharply between categories, which are always applied to finite, limited
relations, and ``Ideas'' (partly in another sense than the Platonic), which
demand absolute conclusion of the thought-work in a comprehension of the
whole. He overlooks that the categories too really make ideal demands on our
thought-work. When are the concepts of magnitude and causality completely
satisfied? Do they not also, just as much as, e.g., the ``Idea'' of the
world, require a comprehension of the whole, if they are to be carried
through completely? Continuity, indeed experience itself (in the sense in
which Kant takes this word), set up great ideals for our cognition. Already
above (37--38) it showed itself that the concept of reality is an ideal
concept.

55. While Plato and Kant started from an analysis of thought-activity, as it
expresses itself in judgments, \emph{Hegel} begins his logical system
directly with the simplest, most abstract concepts, relying on the fact that
every concept, when only it is thought through rightly, must lead over to
other concepts. For a contradiction will come forth when a single concept is
set as the sole, all-comprising expression of existence. And on the
forward-driving force of self-contradiction rests the train of thought in
Hegel's logic. Self-contradiction is only a testimony that no thought can
stand isolated---that the world of thought constitutes a whole against which
one sins by holding fast a single point of view. But precisely for this
reason one could, Hegel maintains, safely begin with a single thought, even
the poorest of all; it will with inner necessity lead over to the others, so
that ever richer thoughts are won.

Hegel has, however, in fact conceded that the doctrine of categories
presupposes a psychological-historical preparation. In his \emph{Phänomenologie
des Geistes} he has presented the stages by which the human spirit---and the
``world-spirit'' through it---raises itself to a standpoint on which the
thoughts can unfold according to their own law, constantly through
contradictions working their way forward to higher unities encompassing the
conflicting oppositions.

The category that is all-determining in Hegel is one that in Aristotle was
only one among many, but that will show itself to be the one that most
immediately hangs together with the nature of thought: relational
determinacy, relation. Contradiction comes forth in Hegel only through one's
treating a concept's content without regard to the relations that determine
and limit it. It is a contradiction to remain standing at the abstract
concept Being, since it can get no content at all and is so far a Being of
Nothing---a Not-being. It is a contradiction to remain standing at the
concept Identity, for it presupposes members that are identical, and these
members must be different, if there can be any distinguishing between them at
all; the concept Identity then points over to the concept Difference, and is
moreover itself different from it!---It is Relation's law, the law of
relation, or perhaps better the law of relational determinacy, that
Hegelian dialectic impresses.

What Hegel overlooks is that every one of the concepts that can thus be
taken up and tested as examples of our thoughts' ``dialectical'' character
have themselves grown up in a definite, given connection. That they, when
they have been tested and determined, can be ordered in series, proceeding
from the simplest and most abstract to the most concrete and individual, is
a matter in itself; the question in this respect will only be whether Hegel's
ordering by triads---position, opposition, and higher unity of both---is
really the natural ordering. The main thing is that what in Hegel stands as
thought's self-development, a kind of logical generatio æquivoca, is in
reality a fruit of a constant reciprocal action between thought's forms and
its objects. Our thought is a reflection---both in the sense that there is
always something it seeks after, and in the sense that at every stage it
presupposes objects and so far comes after these (cf. 2--4). Hegel constantly
draws from the memory of experiences, the race's and his own; from there he
has the objects that his reflection presupposes.

Reflection itself (and the concepts or Ideas in which it deposits its work)
comes in Hegel's world of thought not first but last---precisely because it
is the richest, most concrete category, the one that in its law of
relational determinacy contains all other categories in itself. His logical
system looks, by virtue of the law of the triad, thus:

\begin{quote}
I. Being. a. Quality. b. Quantity. c. Degree.\\
II. Essence. a. Ground. b. Phenomenon. c. Actuality.\\
III. Concept. a. Subjective Concept. b. Objective Concept. c. Idea.
\end{quote}

From the simplest thoughts (I) one proceeds to thoughts that are distinctly
relationally determined (II), and from these to thoughts that express wholes
(III). By Idea Hegel understands, like Kant, a concept of totality. The Idea
is liberated from all the limitations and contradictions under which all
preceding members in the system more or less suffer.

But Hegel definitely rejects Kant's limitation of the categories and Ideas
to being human fundamental thoughts whose applicability was first to be
tested. His doctrine of categories is, like Plato's and Aristotle's, a
metaphysics. The categories are laws of existence, not merely human laws of
thought, and what drives forward from stage to stage, what operates in the
dialectical method, is existence's own innermost essence. Hegel could say
with Goethe: ``Ist nicht der Kern der Natur Menschen im Herzen?'' It is
existence's essence as ideal and absolute whole that, because it cannot be
exhausted in any single thought, compels thought to proceed to ever higher
unities. Thought's law is one with existence's own law.

Even if one would accept this conception, there would be no ground to assume
that contradiction should be the sole motive that drives thought forward. New
objects can arise that demand their right, even if they do not stand in
contradiction with other objects, and that can bring about the downfall of
some categories and the transformation of others, or demand wholly new
categories. And reflection can also be driven forward to conjectures or new
seeking by a tendency to expansion, to extend won insight beyond the hitherto
given, a tendency that is of such great significance for the development of
the life of the spirit.---

The thinkers who in our day stand or think they stand nearest to Hegel in the
doctrine of categories have, however, modified his philosophy more or less in
the direction of Kant's conception. The dogmatic certainty is lacking. What
animates them is the conviction that no thought can have validity when it is
held fast isolated and absolute. Truth must constitute a totality, an infinite
connection. ``The Absolute'' would be the completed, all-comprising
experience, but such a one stands as a constant ideal. We work constantly at
supplementing and integrating the fragmentary in our cognition. While Hegel in
his logic lets the concepts develop out of one another with inner necessity
and prefers to dwell up in this ethereal world of thought, his modern
disciples are most occupied with ascending from the world of finite and
limited experiences to that standpoint of completion. Their works remind one
more of Hegel's ``Phänomenologie des Geistes'' than of his ``System der
Logik.'' Their thinking is really only an attempt to carry through the
criterion of reality, and it is characteristic that F. H. \emph{Bradley}, who
can be regarded as the type for this group of thinkers, uses the word total
experience (experience entire) for the ideal of cognition. (Cf. above
37--38.)

56. An interesting attempt at a doctrine of categories with the emphasis on
Relation (relational determinacy) as the chief concept was made, in essential
agreement with Kant, by \emph{William Hamilton}. Every act of thought
establishes a condition and a limitation: to think is to condition. All the
fundamental concepts our thought works with indicate relations and
conditions. Subject and object, thing and property, cause and effect,
relations of time, space, and degree are examples of this. And with the law
of condition or of relation it accords that comparison and judging cooperate
in every apprehension, in every function of cognition. What is peculiar to
Hamilton's conception, as to that of other Kantians, is that the law of
relation is conceived especially as entailing a limitation---that it is the
expression of our finitude. He does not emphasize that every positive
cognition and understanding, every advance in our knowledge, rests on more
relations being demonstrated and drawn into consideration, and on each single
relation being determined with greater exactness\footnote{See \emph{Den nyere
Filosofis Historie}\textsuperscript{2}. II, p.~388 f.}.

In \emph{Comte}, the founder of the so-called positivist school, we find a
doctrine of categories intimated that in its chief features reminds one of
Kant's. For Comte connection is thought's most general form: tout se réduit
toujours à lier. Cognition springs from an urge for unity. And connection
takes place in two ways, partly by reason of the objects' mutual likeness,
partly by reason of the order in which they follow one another in time.
Herewith relation is already given as an essential determination. Besides the
relations between the objects among themselves, Comte also emphasizes---and in
express agreement with Kant---the relation between subject and object.

The difference between likeness and succession lies at the basis of the
difference between relations of equilibrium and relations of movement,
statics and dynamics, which Comte seeks to demonstrate in the different
scientific domains. Where equilibrium and rest prevail, there are only
relations of likeness to demonstrate; where there is succession and movement,
the task becomes to demonstrate the dependence of the following on the
preceding. Where the relation of likeness prevails, general propositions can
more easily be set up and deduction applied; where the relation of succession
prevails, so that the objects get a historical character, we do not so easily
get beyond the special, and induction becomes predominant. According to this
principle (généralité décroissante---complication croissante) Comte has laid
out his ordering of the special sciences---from mathematics through astronomy,
physics, chemistry, and biology to sociology. In this last, most concrete
domain the relation between statics and dynamics appears as a relation
between order and progress\footnote{\emph{Den nyere Filosofis
Historie}\textsuperscript{2}. II. p.~333 f; 352--355.}.

In \emph{Charles Renouvier}, the most important representative of critical
philosophy in France, both Comte's and Kant's influence is traced in the
doctrine of categories. In his list of categories Relation stands first. It
makes itself known in thought's distinguishing, identifying, and determining
activity, an activity that is also decisive for all other categories, although
Renouvier sets them up alongside the first. A definite principle for the
ordering of the categories cannot be seen. The ordering is the following: 1)
relation; 2) nombre; 3) position; 4) succession; 5) qualité; 6) devenir; 7)
causalité; 8) finalité; 9) personnalité. Within each of them Renouvier seeks
to demonstrate a triad of thesis, antithesis, and synthesis, which Kant had
already intimated, but which cannot be carried through without artifice. A
closer critique of Renouvier's arrangement I shall not give; such a one will
indirectly emerge from the ordering I shall later attempt to demonstrate
among the categories I find. Renouvier is clear that a deduction of the
categories such as Kant had attempted is impossible, and that only experience
can decide whether our system of categories is the right one. Experience
itself is a synthesis, and the categories are the constant relations that make
themselves felt in experience---relations that presuppose just as many
special forms of synthesis\footnote{Cf. besides my book \emph{Moderne
Filosofer}. p.~72 f.: \emph{G. Séailles}: \emph{La philosophie de Charles
Renouvier}. (1905). p.~84--133.---\emph{L. Dauriac} in Revue de Métaphysique
et de Morale. 1909. p.~487--496.}.

In \emph{Eduard v. Hartmann} too Relation is a chief category (Urkategorie),
because our conscious thinking is a reflecting, a positing of relations; it
brings out the relations to which everything that becomes an object of
consciousness is subject. All Being is relational Being. And this holds also
for all sensation, although this has its special categories (quality,
quantity, extension in time and in space). Within thinking Hartmann
distinguishes between the reflecting thinking, which expresses itself in
comparison, distinguishing and connecting, measuring, inferring, and the
speculative thinking, whose categories are causality, finality, and
substantiality\footnote{Cf. \emph{Arthur Drews}: \emph{Hartmanns
philosophisches System}. (1902). p.~766--846.}.

Here too I cannot enter into a closer critique. It was in this survey
especially a matter for me of bringing out the chief points, at which there
also shows itself a certain convergence between the different catalogues. The
convergence shows itself especially in the weight that is laid on
consciousness's, especially thinking's, combining or connecting character
and, in close connection therewith, on relational determinacy, relation as
necessary consequence thereof and moreover as common characteristic feature
for the more special concepts. It is of no slight interest that in a chief
point of the doctrine of categories, this central part of philosophy, there
is found such an agreement. The more special categories do not hang together
so closely with the general nature of thought that they could be derived from
this without further ado. They get a more historical character, and the order
in which they are placed is conditioned in many ways by more special
philosophical problems and views, something that is especially clearly seen
in Renouvier and Hartmann.---

When some authors (Lotze, Sigwart, Wundt), following \emph{Herbart}, place
the categories object and property before the category relation, the
objection must be made that a thing's concept contains only relations within
itself and to other things, and that properties only express such relations.
And to this is added, as the most essential objection, that the object is
known only as thought's object; it cannot then itself be the first category.
Aristotle indeed placed it first; but he added that it was really no category,
since it was precisely that which was to have its determination through all
categories.---

57. As regards the method of the doctrine of categories, it has already been
said that the categories must be found by reflection's seeking to separate
out the forms in which the life of thought involuntarily moves in reciprocal
action with objects and tasks. The individual categories arise at different
times and in different connections. But the presentation of them can be laid
out in such a way that we begin with the simplest and proceed to the more
rich in content. The simplest will probably be those that hang together most
closely with the nature of thinking and therefore express the most necessary
presuppositions that the objects must satisfy; they must therefore also hold
in the greatest extent. They can merely indicate the demands that must be
made at all on every object that is to be able to be thought. The more rich
in content a category is, the less comprehensive a circle of phenomena it
will be able to be applied to. It is a rule that holds for all our concepts,
that content and extension---in concepts that lie in the same ascending
series---stand in inverse relation to each other. From this springs, as will
later show itself, one of the greatest problems for our cognition.

The transition from the one category to the other, that is, from the simpler
to the more rich in content, is necessitated by the appearance of objects or
elements that are not satisfied by the simpler, more comprehensive
categories. Existence is then richer than thought. There must then be sought
more special categories, forms in which the new objects or elements can come
into their own. The more special categories will not be able to be derived
from the more general; but it will perhaps be possible to show that they are
forms in which the more general categories must appear under definite given
conditions, or to find intermediate links between the old categories and the
new. At every point thought's own nature stands in reciprocal action with its
objects and its tasks. There will be a constant breaking between the two
sides, a breaking that in our day has called forth a dispute between
Criticism and Pragmatism. The doctrine of categories need not enter into this
dispute, which belongs under the discussion of the problem of cognition
itself. But it will perhaps be able to give a contribution to the
illumination of the disputed question.

\section*{B. Fundamental Categories}
\addcontentsline{toc}{section}{B. Fundamental Categories}
\label{sec:fundamentale-kategorier}

58. What is given as an object for reflection is always a whole, whether
this be loose and changing, or firm and lasting. If the object were an
absolute single, thought would become one with a staring and lead to
ecstasy. Even in the moment consciousness awakens from the unconscious by
the abolition of the equilibrium that the unconscious presupposes, the
state is determined by the opposition to the preceding state and constitutes
a whole with this. Phenomena of consciousness are always determined by a
relation between background and foreground, a relation that can be
successive or simultaneous, or both. In the simplest case it is only the
background that determines the foreground; but when the background does not
immediately vanish, but is preserved or recalled, it can itself in turn
undergo change, become foreground in relation to what was before
foreground. In all sensation, memory, and imagination this reciprocal
relation lays itself bare. Especially clearly, however, it comes forth, as
we have seen (33), in judgment, where the predicate not only determines the
subject, but is also itself determined by it. In all judging, as well as in
sensation, intuiting, and association, the psychic energy makes itself known
as a combining, and the law of synthesis therefore becomes the fundamental
law of cognition. \emph{Synthesis} becomes the first category. It is applied
involuntarily as long as consciousness subsists, and it is in analogy with
it that we apprehend everything. The type for all cognition, and therefore
also for everything cognized, is connection of a multiplicity into unity.
Neither a wholly connectionless multiplicity nor a differenceless unity
(singleness) can be cognized. Already purely psychologically, states of
these kinds are impossible, although there can be strong approximations to
them.

In this first category now lies at once contained the second. When a
multiplicity of objects or elements is connected or combined, they are
brought into relation with one another. This happens in every sensation,
every memory, and every imagination, as well as in every judgment. When a
definite connection is dissolved, as when intuiting is replaced by judging,
this dissolution is only a conversion into other relations. The wholly
relationless can be neither sensed, remembered, represented, nor judged. To
think is, as \emph{Fichte} has rightly said, to set something in relation to
another.

\emph{Relation}, relational determinacy, then becomes the second category.
It denotes both something that appears involuntarily and an ideal.
Involuntarily, in a consciousness all objects and elements are set in a
certain relation to one another. But our reflection seeks to determine this
relation as exactly as possible and to reshape it in such a way that all
objects, earlier emerged and possible new ones, can come into their own.
Wherein this ``own'' consists is decided by the more special categories.

The often misunderstood proposition that everything is relative does not
mean that everything is equally right or equally wrong, but precisely the
opposite, that everything that is cognized stands and must stand in definite
relations, and is known the better the more exactly these objects are
determined. Within the definite relation, cognition is absolute.

59. There are, as said, two limits for consciousness and therefore also for
thinking, absolute unity and connectionless multiplicity. At neither of
these limits does any relation hold, and therefore no synthesis either.
Between both limits lie relations that can be designated as \emph{Continuity
and Discontinuity}.

These two categories must be named together, since they mutually presuppose
each other. An abolition of equilibrium is required for consciousness to
replace unconsciousness---for an intuition to be able to be articulated
(17)---for judging to replace intuition and association. An interruption is
required, both for the simplest sensation to come to be, and for reflection
to be awakened. So far \emph{Søren Kierkegaard} is right that thought begins
with a leap, and that a special interest is required to make this leap; in
any case the presupposition is an involuntary interest in preserving or
restoring the equilibrium and the connection. This leap expresses a
discontinuity; but it reveals at the same time that there was previously a
continuity, and it may perhaps itself later show itself to be a member in a
deeper-lying continuity.

It might perhaps be disputed that continuity is in itself a relation. For
insofar as it subsists, no distinction can be made between members that
could stand in relation to one another. When it prevails, consciousness
glides imperceptibly forward, seems constantly identical with itself, and it
is only the reflection awakened by discontinuity that becomes clear that
changes have uninterruptedly taken place, even if these were so small that
they could be set as near to unchangeableness as one likes. One can say with
\emph{Poincaré} that in all continuity a contradiction is contained. If a
continuous series of the members A B C subsists, then $A = B$ and $B = C$,
but $A \gtrless C$\footnote{\emph{La Science et l'Hypothèse}. p.~35.}. But
this contradiction is first discovered when experience or memory by a leap
leads to an immediate juxtaposition of A and C. Then comes the awakening.
One can then ask: ``whither are we gliding?'' As long as one is absorbed in
the gliding, one does not ask; the contradiction does not exist as long as
continuity prevails. Continuity is a relation that is not discovered until
it is past. It is only when discontinuity has come into force that one can
apprehend continuity as a connection of objects or elements that lie
infinitely near one another.

We stand here again at the earlier-mentioned psychological antinomy. When we
say that in continuity, in the continuous states of consciousness, a
multiplicity of members is contained, this can hold only for reflection,
which looks back and makes distinctions. In the states themselves the
multiplicity did not come forth as such. Both in our apprehension of
sensation, of intuiting, and of thinking, this must indeed be noted, so that
one does not read something into the more immediate and involuntary states
that first comes forth in reflection. In any case one has no right to
identify without further ado the differences that reflection sets up with
those that can already be said to ``lie in'' the immediate processes. (Cf.
5.) Or, as one can also express it, the rational continuum, the continuum
formed by reflection, is different from the empirical continuum. When
reflection is awakened, a closer determination is sought of the empirical
continuum that the involuntary mental life constitutes, and this can only
take place by apprehending it as discontinuous multiplicity. Discontinuity
is thus not merely a condition for the awakening of reflection; but we seek,
when reflection is awakened, to apprehend continuity in analogy with
discontinuity. And only now can one define continuity positively; at first
it stands as a negative concept, as absence of multiplicity as well as of
unity. But one now assumes within continuity differences of a lower degree
than the clear oppositions that discontinuity presupposes. Often great
problems are hereby set. Thus when a new mental peculiarity, a new psychic
tendency, suddenly reveals itself: can there previously have been such a
continuum as one formerly assumed? In the material domain the discovery of
radioactivity, in the biological domain the demonstration of mutations,
leads to analogous questions.

There is, however, no ground to deny that discontinuity is just as original
a property of conscious life as continuity. It would be incorrect, with some
philosophers in our day (cf. 3), to conceive discontinuity as a more or less
artificial result of attention and reflection, set in motion by practical
motives. Consciousness subsists under a constant breaking of continuity and
discontinuity, and this holds especially for the life of thought. The
experience of difference, opposition, and interruption is just as elementary
as the experience of the gliding transitions. And the experience of
discontinuity is indeed, biologically-practically regarded, even a more
significant experience, since it is only by an awakening that energy can be
released. Experience shows us strife and tension between different tendencies
in conscious life, no less than harmony and continuity, and it is only
thereby that the psychic energy gets its tasks, its resistance to overcome.
The differences belong to life, not merely to reflection. And since they
present the clearest relations, reflection uses them to make the continuous
states and objects intelligible to itself. There expresses itself here a
feature that constantly recurs in our cognition---that a part or side in our
experience is used to illuminate other parts or sides by way of analogy.

Conversely, continuity too is then used to illuminate discontinuity. Given
differences, oppositions, and relations of tension are made distinct by the
finding-out or assumption of intermediate links and transitions in as great
a number as possible, so that the differences finally become infinitely
small. Objects or elements that stand in pronounced opposition to one
another are then sought to be apprehended as extreme points of connected
series. The very impulse or leap with which both sensation and thinking
begin is then sought to be brought into a relation of continuity to other
objects and elements in conscious life or outside it. The apparently
accidental thing in the starting-point thereby falls away. In this way the
principle of continuity has constantly shown itself, in the history of
science, as a great principle of discovery, in that one tries to give a
once-discovered relation as great an extension as possible. The scientific
imagination works according to this principle until reflection anew
demonstrates discontinuity. Thus Newton asked whether the force that lets
the stone fall to the earth should not also operate on the moon, and further
extended this question to hold for all bodies in relation to one another. In
all representations and inferences concerning the human past, life's past,
nature's past as a whole, we infer from the forces and laws we now know and
of whose validity we have convinced ourselves. The principle lying in this,
the principle of actuality, is only a peculiar form of the principle of
continuity. There makes itself known here an expansion that psychology
demonstrates not only for the life of thought, but also for the life of
feeling. Thus a mood called forth by a special experience will have a
tendency to spread over the whole feeling-state, as long as it is not met by
a mood called forth from another side.

Thinking thus gets a task, both when there are too many intermediate links,
and when there are too few.

It seeks to make continuity discontinuous and discontinuity continuous, and
both these apparently opposite processes stand in the closest connection with
each other. In both of them thought moves forward through combining in
definite relations.

60. If a closer determination of Relation is sought---besides its being able
to be continuous or discontinuous---it shows itself that in the individual
cases it is either a relation of likeness or a relation of difference. This
is already presupposed in the foregoing. For continuity is a relation of
likeness within a multiplicity that is not apprehended as multiplicity,
because the difference within it is not noticed, and discontinuity
presupposes a certain difference within a multiplicity.

That all thinking is a positing of relations can then be more closely
expressed thus, that all thinking is a comparison. (See already 2.) It can
also be said to hold for all sensation and all intuiting that it is relations
of difference and likeness that are determinative for the individual objects'
or elements' appearance and for the way in which they are connected. (See
11; 14; 17.) There is an analogy between these more elementary functions and
judgment. But we keep from now on especially to this latter, because the
relations are more easily accessible here, and because judgment is of
decisive significance for the further development of cognition. Every
judgment expresses both a relation of likeness and a relation of difference.
Despite the difference between subject and predicate, we nonetheless have in
the predicate the subject itself from a new side or in a new form. It is the
relation of difference that makes the judgment possible and gives it a
content; but the judgment is at the same time precisely a victory over the
difference, in that this is either wholly abolished or gets a limitation.

We have earlier had occasion to mention the different kinds or degrees of
likeness: identity, congruence (sameness), composite likeness, qualitative
likeness, and relational likeness (analogy). (21.) The different kinds or
degrees of difference we shall in what immediately follows have occasion to
mention, since it is they that give occasion to the appearance of the more
special categories.---

That every relation is either a relation of likeness or of difference cannot
be proved in any other way than by one's finding in each single case, by
analysis, that the said relations lie at the basis. It is not, however,
conceded from all sides that all combining and positing of relations is a
comparison. Thus Th. \emph{Lipps}\footnote{\emph{Grundzüge der Logik}.
(1893). p.~93--95.} maintains that a distinction must be made between
combination-judgments and comparison-judgments: in the former a multiplicity
is connected into a whole; in the latter different objects are compared among
themselves. A comparison-judgment we have when it is said that the one of two
tones is higher than the other. A combination-judgment is, e.g., the
judgment: ``this is a tree.'' In this judgment, says Lipps, ``the given
manifold---`this'---is by the name tree combined into an objectively
necessary unity: this naming-judgment includes in itself a consciousness
that, when this designation is to hold, I must think that manifold as
one.''---To this it must in the first place be remarked that only the
different can be called manifold and connected into a whole. If the elements
in the given object were absolutely identical in quality, time, and place,
they would not be manifold: the object would then be an absolute unity. The
relation of difference is thus presupposed. Next, there lies in the naming
itself a presupposition of likeness between the object (``this'') and other
objects, and it is this likeness that the predicate expresses. Finally, the
object, in order for its elements to be able to be gathered under a common
predicate, must already be distinguished from other objects as something
apart (cf. 30 ff.). Here again the relation of difference lies at the basis.

The breaking between unity and multiplicity will always show itself to be a
breaking between likenesses and differences. In this breaking there
constantly intervenes the interest or the purpose that guides thought. It
cooperates in the individual cases in determining which group is
formed---which object is separated out to be determined---hence which
difference is made; and it cooperates also in the point of view that is taken
toward the object. The separated-out object (the group) corresponds to the
judgment's subject, the point of view to its predicate. Instead of saying
``this is a tree,'' I could under other conditions, toward the same ``this,''
come to say ``this is hard,'' or ``this is something I must go around.'' And
likewise I could under other conditions have formed another group, so that
``this,'' e.g., became the tree with the earth it grows in, with the birds
that build a nest in it, the insects in its bark, and so on. The object then
became a whole little world. That already the judgment's subject presupposes
a distinguishing is easily overlooked, since it can stand in a certain
indefiniteness---since it is presupposed known and therefore not always
expressed---and since the distinguishing act is undertaken beforehand and
perhaps wholly involuntarily.

61. We involuntarily order given multiplicities according to likenesses and
differences, and this naturally becomes the first step in our cognition. The
question is what a complete ordering would be, and whether such is always
possible. Here, in the doctrine of categories as descriptive epistemology,
we have to do only with the possible orderings; the question of validity
belongs under the problem of cognition.

Instead of the word ordering I shall in what follows for simplicity's sake
use the word series, although this word is as a rule used in logic in a
narrower sense, namely of such orderings as make inference and calculation
possible (cf. 33). Epistemology must investigate whether there are not other
possibilities than those that make science possible. It is no matter of
course that everywhere, at all times, and in all domains, orderings can be
formed that make inference and thereby science possible. We have no right to
establish in advance the rationality of objects. It is, as we have seen (35;
52), not certain that thought can exhaust the whole content of sensation,
intuition, and association in the clear form of judgment. And there is also
no certainty that our sensations and our memories are exhaustive. There is
every ground to believe that they are always fragmentary. Our objects are
always limited. Already for this reason orderings or series of an imperfect
kind could be thought of, at which thought must nonetheless stop. The
doctrine of categories then needs an investigation, an orientation with
respect to the in-themselves-possible series of difference and likeness.

When I try to order a multiplicity of objects or elements according to
likeness and difference, it shows itself that the members in the series that
is formed can stand in very different relations to one another. There are
here especially three relations that become of significance. There can
possibly be formed a series of such a constitution that any member whatever
can be pushed out, and that the found relation of likeness or difference
nonetheless continues to hold between the remaining members. Such a series
\emph{Augustus de Morgan}\footnote{\emph{Cambridge Philosophical
Transactions} IX. (1860). p.~104.} called a \emph{transitive} series.
\emph{William James} has later called it a series that makes mediate
comparison possible\footnote{\emph{Principles of Psychology}. II. p.~646:
The principle of mediate comparison is only one form of a law which holds in
many series of homogeneously related terms, the law that skipping
intermediary terms leaves relation the same. Cf. Vol. I. p.~190.}. And there
can possibly be formed a series of such a constitution that any member
whatever in it can be exchanged with its neighbor-member without the
relation of likeness or difference thereby being changed. Such a series I
will call \emph{convertible}. In a transitive series ABCD, B can fall away,
and the relation between A and C is then the same as between A and B, and as
between B and C. In a convertible series ABCD \ldots\ B and C, e.g., can be
exchanged, and the relation between A and C, C and B, B and D is then the
same as between A and B, B and C, C and D. A convertible series (in the
indicated sense of the word) must at the same time be transitive. The
exchange operates in part as a pushing-out. Finally, a series can be formed
in which the relation continues to be the same, whether one goes through the
series from the front or from behind. Such a series can be called
\emph{symmetric}. The series ABCD\ldots\ is symmetric when the relation
between D and C is the same as between C and D, likewise the relation between
C and B the same as between B and C, and so on. Such a series can be
intransitive, namely when neighbor-members cannot be exchanged. Because the
relation D--C is the same as the relation C--D, it is not said that I can put
D in place of C, since it depends on how the relation between B and D is. In
an intransitive series every member is determined by its relation to other
members, and exchange is therefore not possible. While a symmetric series
can be both transitive and intransitive, a convertible series can only be
transitive.

De Morgan did not distinguish between convertible and symmetric series. And
neither later has that distinction, as far as I know, been made. One has as a
rule used the expressions synonymously and thought only of whether the whole
series could be reversed, not of whether the members in the series could be
reversed among themselves.

Finally, it will show itself that a series can often be read from behind,
when one constantly reverses the relation. It is an inversely symmetric
series.---

I now go through these possibilities in detail and seek to find examples of
them. We shall thereby come upon more special kinds of relations of likeness
and difference and thus be able to make the transition from the fundamental
categories to the more special ones.

62. It could be thought that absolute difference prevailed between the
objects among themselves, so that it is only an occasion accidental with
respect to their mutual relation that has led to placing them together. And
the differences could further be thought to be the same between any objects
whatever, so that the order is indifferent and members can be pushed out.
Such a series, which can be called a \emph{chaotic difference-series}, would
be at once transitive, convertible, and symmetric. It can be expressed thus
(the sign $\wedge$ expressing a wholly indefinite relation of difference):
\[ A \wedge B \wedge C \wedge D \ \ldots\ldots\ (1). \]
The difference between A and C is the same as between A and B and between A
and D, and so on.

Where such a series can be formed, we are at the limit of our
cognition---either at its beginning or at its end. If everything given, all
objects and elements, were of such a kind that such series could be formed,
cognition would be impossible. Of course there must be an occasion to form
the series. The series can indicate what shows itself to us at a certain
point of time or at a definite place, in our sensation, in our imagination,
or in our dreams. But if no other ground for the placing-together shows
itself, the members can be exchanged in any way whatever, or pushed out, or
read in any order whatever. The only bond that for a while connects them has
nothing to do with their mutual relation.

Immediately given such a series can really never be. Only by a mass of
thorough comparisons can one reach the conviction that all points of
likeness or definite relations of difference are lacking. An approximation to
a chaotic difference-series we have in the mere collection of observations
and pieces of knowledge (experientia vaga). That between our objects there
are no other definite relations than precisely this, that they are gathered
at the same time and place, can only be shown by thorough investigation.
Absolute unreason (irrationality) can only be shown by the use of reason
(ratio); only this can know its limit, and if it does not know it, there is
no other force or capacity that knows it. The limit of cognition conditions,
when it is demonstrated in this way, a docta ignorantia.

If a chaotic difference-series held for all objects in all possible
experiences, an absolute pluralism would follow from it. In the assumption
of absolute atoms, in Hume's doctrine of connectionless ``perceptions,'' in
Kant's ``matter'' as well as in his ``things in themselves,'' in Herbart's
``reals'' there are greater or lesser approximations to such a conception.
But pluralism is maintained by an inquiry that goes beyond the immediately
given. One cannot here appeal to immediate experience. Pluralism must always
be rationalistic. Only reflection would be able to demonstrate an absolute
chaos, and the demonstration would be difficult.

63. The chaotic begins to disappear when it is possible to form a series in
which the differences vary from member to member, so that the members can
neither be pushed out nor exchanged, and in which, moreover, no definite
direction in the variation can be demonstrated. Such a series can be called
an \emph{indefinitely varying difference-series}. It is inconvertible,
intransitive, and asymmetric, and can be expressed thus:
\[ A \underset{\mu}{\wedge} B \underset{\gamma}{\wedge} C \underset{\pi}{\wedge} D \ \ldots\ldots\ (2) \]
An example of such a series we have when the ordering takes place according
to a memory-schema, or according to our different interests for the
different objects, or purely externally chronologically. In distinction from
the chaotic series there is here a principle of ordering. It is not merely
the placing-together in general, but also the ordering, that takes place
from a definite point of view, only that this is just as inessential for the
objects' mutual relation as, in the former case, the principle of
placing-together.

64. A step further we come when the differences indeed constantly vary from
object to object, but the variation goes in a definite direction. The
ordering can then take place according to decreasing likeness and increasing
difference, or conversely. Such a series can be called a \emph{regularly
varying difference-series} and is, like the preceding, inconvertible,
intransitive, and asymmetric. It can be expressed thus:
\[ A \underset{\alpha}{\wedge} B \underset{\beta}{\wedge} C \underset{\gamma}{\wedge} D \ \ldots\ldots\ (3) \]
An example we have in the color-series. There is a qualitative difference in
the transition from color to neighboring color; but yellow is more different
from red than orange is, and the difference between orange and yellow is
another than that between red and orange. In all comparative science
(comparative biology, psychology, and sociology) the objects are ordered in
series according to as essential likenesses or differences as possible. If
one lays weight on the fact that even after such an ordering many
questions---especially about the causes of the likenesses and
differences---still stand unsolved, one comes back to the indefinitely
varying difference-series. It is the point of view that determines how the
individual series is characterized.

65. An \emph{identically varying difference-series} we have when the
difference between any two neighbor-members whatever always varies in the
same way, each time one goes from member to member. Of this kind is the
relation between the numbers in the number-series, when this is regarded as
a series of ordinal numbers (serial numbers), whose distance from the
beginning constantly grows in the same way. Between events among themselves
and between locations among themselves the same relation can obtain. Such an
ordering is more primitive than counting and measuring proper.

The identically varying difference-series is inconvertible, intransitive,
and asymmetric, and can be expressed in the following way:
\[ A \underset{=}{\wedge} B \underset{=}{\wedge} C \underset{=}{\wedge} D \ \ldots\ldots\ (4) \]
The series (2)--(4) denote an advance in relation to series (1) in that the
members lie fast and are determined by the place they occupy in the series.
The differences are now bound, and chaos is abolished. It is the first
condition for getting further.---A further advance is made when it succeeds
in forming a transitive series, even if it continues to be inconvertible and
asymmetric.

66. Such a series is \emph{the progressive difference-series}, in which
every single member relates to the preceding as the following relates to
it---e.g. as whole to part, as the later to the earlier, as right to left,
as up to down, as the greater to the smaller. Here members can be pushed
out, and inference can therefore take place: when B lies to the right of A
and C to the right of B, C lies to the right of A, and so on.---This series
can be expressed thus:
\[ A < B < C < D \ \ldots\ldots\ (5). \]
The sign $<$ then here does not merely mean ``less than,'' but denotes any
inconvertible, transitive, and asymmetric relation.

The relation of time and of place are here not (as in series (2) and (4))
mere outer points of view or principles of ordering; reflection is here
directed toward the very relation of time and of space as such.

67. As soon as a difference-series is to be formed, likeness of course plays
a part along with it, since decreasing difference is increasing likeness,
and conversely. Even in the chaotic (1) and the indefinitely varying (2)
difference-series, there is at each individual ordering a common point of
view, although it can shift, as in the chaotic, or be wholly external, as in
the indefinitely varying. In the regularly varying (3) and the identically
varying (4) likeness makes itself felt within the series itself, in the
former between the members among themselves (two and two), in the latter
between their relations. And in the progressive relation-series (5) it steps
forth still more definitely, in that here quite the same relation holds
between any members whatever in the series, which shows itself by
pushing-out.

Still more definitely, however, the relation of likeness appears when every
following member can by analysis be found in the preceding, hence be said to
be contained therein. Such is the relation when the following member is
predicate in relation to the preceding as subject, or consequence in
relation to the preceding as ground. Such a series can be called a
\emph{partial identity-series} and is expressed in the following way:
\[ A \to B \to C \to D \ \ldots\ldots\ (6). \]
The arrow means ``subject to'' or ``ground of.'' In this series the members
cannot be exchanged, since one cannot without further ado let predicate and
subject, consequence and ground, change places. But members can be pushed
out, so that inference is possible. The series is thus, like (5),
inconvertible, transitive, and asymmetric.

When one conceives judgment as an act by which the subject is subordinated
(subsumed) under the predicate's extension as part under whole (cf. 33), a
chain of judgments can be brought under (5) as well as under (6). And series
(5) can sometimes be formulated as a partial identity-series, since the part
is, after all, identical with a part of the whole, the smaller with a part
of the greater. But it then becomes a series that runs in the opposite
direction from (6), for in (6) it is conversely the following member that is
comprised in the preceding. In the transition from part to whole something
is added; in the transition from subject to predicate a closer determination
and thereby a limitation takes place.

68. There can be series that are symmetric, but neither convertible nor
transitive. Such a series one can just as well read from behind as from the
front. It can be called a \emph{reciprocity-series}. As examples de Morgan
names: A associates with B, B with C, and so on, or: A is in conflict with B,
B with C, and so on, or: A is cousin to B, B to C, and so on. Here one
naturally cannot push out members or exchange them among themselves; but the
series can be read in reverse order. It can be expressed thus:
\[ A \smile B \smile C \smile D \ \ldots\ldots\ (7). \]
The sign $\smile$ means that the relation is the same between any
neighbor-members whatever, but that no member can be pushed out. A is not in
conflict with C because he is in conflict with B, and because B is in
conflict with C. Rather he will become friends with C.---From (3) and (4)
this series is distinguished by the fact that the relation does not vary,
from (5) and (6) by the fact that it is indifferent whether the series is
read forward or backward.

The series (4)--(6) can also become symmetric, hence be read from behind,
when one reverses the relations, hence goes from the remotest location to
the nearer, from the greater to the smaller, from predicate to subject. But
in (6) the justification of this will require a special investigation in the
individual cases. One could call such series inversely symmetric series or
\emph{absolute correlate-series}.---A ``war of all against all'' would belong
under the chaotic difference-series.

69. If the differences in one respect or another, or in all respects, fall
away on closer investigation, we get an \emph{absolute identity-series}:
\[ A = B = C = D \ \ldots\ldots\ (8). \]
Where such a series can be formed, it shows itself that the given
differences were disguised identities. Such a series is at once symmetric,
transitive, and convertible. It can be read from behind as well as from the
front; the members can be exchanged; and members can be pushed out. These
properties this series has in common with its complete opposite, the chaotic
difference-series. Between these two limits our cognition moves. The two
limits have this in common, that they can only be demonstrated after careful
investigation. An absolute identity-series can only be set up when one can
everywhere put one member in place of any other whatever, without any change
taking place in the whole connection. This is aptly expressed in
\emph{Leibniz}'s definition of identity (49): ``Identical are the things
that can be put in each other's place without breach of validity.''
Difference, on the other hand, is present where the one cannot everywhere be
put in place of the other (in quibus substitutio aliquando non
succedit)\footnote{\emph{Non inelegans specimen demonstrandi in abstractis}.
Opera phil. ed. Erdmann. p.~94.}. It is clear that whether the substitution
succeeds everywhere can be very difficult to decide, especially in special
domains where many conditions operate together. One will, strictly speaking,
never get beyond the hypothetical---any more than at the opposite limit, the
chaotic difference-series.

Absolute identity often holds only in a single respect. All are, in the more
recent civil law, ``equal before the law,'' however different they otherwise
are in station and circumstances. Regarded as ``siblings,'' all children of
the same marriage stand alike. In general an absolute identity-series can be
set up when different objects fall under the same concept. It lies precisely
in the concept of example that different objects can be placed in place of
one another in a definite respect, however different they otherwise are.
Regarded as horses, a yellow horse is identical with a red one, a Norwegian
pony with a Norman cart-horse, and so on. Analogies can on the whole---in
the definite respect in which the analogy holds---be regarded as identities.

In common with (1) and (7), (8) has that it can be read from behind as well
as from the front. But it distinguishes itself from both by the absolute
equivalence of the members, e.g. of the examples of the same general
concept, or of ground and consequence, when the inference can be reversed.
From (7) it distinguishes itself at the same time by the fact that the
members can be exchanged or pushed out.---

Since the absolute identity-series, where it can be set up, denotes the
completion of the thought-work, it is made the standard of thinking, and one
gladly places it (since \emph{Leibniz}) at the head of the logical
principles in the form $A = A$. It then expresses the fundamental demand
that thought shall be identical with itself, and that therefore every
concept, in whatever connection it occurs, shall everywhere have the same
content. This principle is really presupposed in all series, since every
member in these, whatever its relation to the other members, must be
identical with itself, and since every member can be defined by its place in
the series (even if this place, as in the chaotic difference-series, can
shift and holds only momentarily).

If all series could be reshaped and traced back to a single absolute
identity-series, an absolute monism would be the result. Despite the great
weight \emph{Leibniz} laid on individuality and difference, as his doctrine
of monads shows, it was nonetheless for him the ideal to trace all factual
truths back to identical judgments, and it was in his view only human
limitation that hindered the carrying-through of this. One has even found in
this the original motive for \emph{Leibniz}'s system\footnote{\emph{Couturat}:
\emph{La Logique de Leibniz}. See also Bulletin de la Société française de
philosophie. II. p. 65--89.---Cf. already \emph{Kant}: \emph{Kritik der
reinen Vernunft}\textsuperscript{2}. p. 337.}. But it is certain that the
monads, which according to \emph{Leibniz}'s conception form what has above
been called a regularly varying difference-series, cannot be reduced to
disguised identities. \emph{Leibniz}'s logic and his doctrine of monads
stand in sharp opposition to each other. Still more pronounced and dogmatic
does identity appear as ideal in \emph{Parmenides}, \emph{Plato}, and
\emph{Spinoza}.

But even if all objects could conceptually be traced back to being members
in an absolute identity-series, the question would still stand unanswered
whence the differences derive with which the thought-work begins, and which
set its tasks, and how it is to be explained that identity did not
immediately reveal itself. It is of no use to declare the differences merely
``subjective,'' or ``empirical,'' or ``illusory,'' as one has done with
sense- and value-qualities. The qualitative and the historical will always
make themselves felt anew, however often, when an identity is reached, one
tries to cast off the bridge behind oneself. We must therefore constantly
work with the different series, ascending upward from chaos in the direction
of identity.

70. The purpose of setting up these series has only been to give a schematic
presentation of the thought-work, as it takes place under constant breaking
of relations of difference and likeness. There may perhaps be found still
more series. But the weight does not lie on completeness. The significance
that the categories of likeness and difference have for our thinking, when
synthesis and relation are to be applied in the particular, shines forth
sufficiently from the end-points (chaos---identity) and the intermediate
forms.

I shall only mention further that the series set up end indefinitely. It
remains open how far they can be concluded. In such cases where a conclusion
is given or lies in the nature of the matter, we can get series-forms that do
not occur in the list set up. If a series merely consists of two members, A
and B, and if the relation between A and B is the same as between B and A, we
have a little series that is both convertible and symmetric (like (1) and
(8)); while there can of course be no question of pushing-out; it is thus
neither transitive nor intransitive. As one of my auditors, \emph{Naphtali
Cohen}, has pointed out, one can, in closed series of several members, have
both convertibility and intransitivity. Every member is then determined by
its relation to \emph{all} other members; none can be pushed out, but all can
be exchanged. Thus, e.g., the series formed by the three vertices in an
equilateral triangle; each has the same distance from both the others. The
series of the Decemvirs is closed; none can be pushed out; but the members
can be exchanged arbitrarily, provided only that one says of each single
Decemvir what can also be said of the others.---Such series are at the same
time symmetric, since it is indifferent whether one begins the series from
the one or the other end. And such series can moreover also be formulated as
absolute identity-series, in that one regards the individual members as
examples of one and the same concept (vertex in an equilateral triangle;
Decemvir), only that in this case the concept's extension is definitely
delimited.

71. As already remarked (61), one has especially been interested in the
series ((5), (6), and (8)) that make inference possible. They are the
properly rational series. Inference is conditioned by pushing-out
(transitivity). But an inference always presupposes a limitation of the
series: one wishes to prove a definite judgment and then seeks the premises,
or one has a judgment given and then seeks what follows from it. There must
be a goal set, a task given. Purely objectively regarded, it is arbitrary
where the beginning and the end of the series are set (apart from the
just-mentioned conceptually closed series). A series within which
pushing-out can take place makes the formation of a logical totality
possible, in that \emph{only} such members can come into consideration as
can serve as beginning-, intermediate-, and end-members in an
inference-process. It is a thought of purpose, a wish, that is the driving
force for the formation of such a totality. Yet this wish can itself be an
effect of the relation between the series-members at hand; one wishes
clarity about this relation, wishes to see the consequence.

But we must not, in the doctrine of categories, overlook the fact that there
are also other series than the rational ones. It is precisely our task to
orient ourselves with respect to the question whether and how far it is
possible to form firm series of difference and likeness. We must then also
take into account other possibilities than those that make rational thinking
possible. A chaotic difference-series would even make it impossible.

The series that make inference possible we naturally try to build wherever
the objects permit it. And when we are now to attempt a survey of the
special categories, conceived as peculiar forms of likeness and difference,
we must, among the examples with which we have been able to support the
different series-formations, keep to those that offer the greatest
significance of principle, because they are suited to be guiding thoughts
for research. It is a matter of a struggle between likeness and difference;
the task is to reduce the differences to the least possible and yet fully to
let them come into their own. We begin then with the thought-forms that are
most intimately attached to the very thought-activity, as this is known
psychologically and historically. Here there is the greatest independence of
the objects; for of all objects it holds that thought must apply to them the
forms without which it cannot work. Already here---in \emph{the formal
categories}---there appears a breaking between likeness and difference. This
becomes still more pronounced when the difference between objects and
thought-forms is especially emphasized. It will show itself that this
especially happens when the relation of time between the objects' appearance
becomes of significance. Even if all other differences could in one way or
another be conceived as disguised identities, the relation of time
nonetheless always remains as irreducible, as that which constantly anew
sets thought in tension. In thought's attempt to overcome this tension it
applies \emph{the real categories}. When the interest that is always
attached to the thought-work is especially emphasized, the concept of value
presses forward. And since there now in fact are other interests than the
one that animates the thought-work itself, the different values naturally
become objects for thought. The question becomes whether there is also, for
such objects, a possibility of a rational thinking, that is, of a thinking
that makes use of such series as make inference possible. Since every value
can be said to contain a demand, an ideal, namely either attainment,
preservation, or development of the valuable, we can call the fundamental
concepts applicable to this class of objects \emph{the ideal categories}.

It is time that leads us from the formal to the real categories; it is value
that leads us from the formal and real categories to the ideal. Everywhere
thought's nature must be the same; but the thought-work takes place under
different conditions. It is the triad Form---Object---Interest that entails
the tripartite division of the special doctrine of categories.

\section*{C. Formal Categories}
\addcontentsline{toc}{section}{C. Formal Categories}
\label{sec:formale-kategorier}

72. Whatever it is that we think over, it is a comparison that we
institute. In every experience there is an involuntary comparison; the new
is referred to what was previously experienced, and if this is not possible,
the novelty appears, by reason of the opposition, so much the more strongly.
Every description takes place by the indication of points of likeness and
difference between an object and other objects; these points of likeness and
difference are what I call \emph{Elements}. And it is here indifferent
whether an object is apprehended as ``real'' or not. Memory and imagination
are in this respect treated just like sensation.

Psychologically regarded, the relation of difference appears first, since it
is the condition for an object's being able to come to be for us at all. To
sense is to distinguish. An object must raise itself up in relation to other
objects in order to be able to be apprehended. There must be relief. The
same holds in the world of memory and of imagination. When the thought-work
replaces the immediate intuiting, one strives to bring out the likenesses.
We have distinguished a series of stages by which thought can ascend from
the world of differences to that of identity (62--69). Identity is thought's
ideal and standard, the presupposition for all concept-formation, judging,
and inferring. Different objects must in one respect or another be able to be
regarded as identical, in order for us to be able to form a common concept
for them. And without a certain identity between initial representation and
final representation no judgment can be formed. Inference, finally, is
possible only by substitution, by identity of concepts that occur in
different judgments.

While psychologically and historically identity is an abolition of
difference, logically regarded difference is an abolition of the identity
that is thought's standard; the different is that which cannot be
substituted---the irrational (since all rationality rests on the possibility
of substitution). \emph{Identity} is a relation between $a$ and $a$,
ultimately an object's relation to itself. In opposition to it stand the
quality-relations, partly qualitative likeness ($a_1$ and $a_2$), partly
qualitative difference ($a$ and $b$), which both exclude substitution. The
thought-work, in the face of \emph{the qualities}, aims at making
substitution possible, or at least at reshaping qualitative likenesses and
qualitative differences in such a way that they approach identity to as high
a degree as possible. The more continuous the series become, the more such
an approximation is reached; the difference then steps forth only when
one---by a leap---compares more remotely lying members in the series. There
is thereby attained an overview and thereby already a certain mastery over
the objects, even if the purely rational power over them should not be
attainable. If no such series can be formed, we must content ourselves with
a separating-out of that which cannot be placed in a series with certain
objects. Thereby we apply \emph{the category of Negation} ($a$ and
non-$a$), leaving it to the future to find a positive relation.---It is in
the quality-category that the series (2)--(4) come to application, while
Negation stops at (1). If it succeeds in demonstrating relations
corresponding to the series (5), (6), and (8), which are transitive, in such
a way that either the relations or the members or both are always identical,
wherever in the series we are, then inference is possible, and
\emph{Rationality} comes into force.---As for series (7), it too gets its
significance, insofar as the relations between the members two and two can be
reversed and yet are identical.---Rationality's schema is $a
\overset{\angle}{\underset{=}{\to}} b \overset{\angle}{\underset{=}{\to}} a
\ldots$, corresponding to the progressive difference-series, the partial and
the absolute identity-series. Even the chaotic difference-series can be
interpreted as a kind of (5): if A, B, C, and D form a chaotic
difference-series, I know that since A is different from C, it is also
different from D. It is a ray of rationality down into the chaos of
differences.

\subsection*{a. Identity}
\addcontentsline{toc}{subsection}{a. Identity}

73. Identity is a degree or kind of likeness. It is through a series of
degrees of likeness that thought ascends to identity. (Cf. 21, besides the
series (1) to (8).)

It has been held that the relation is the reverse, so that the concept of
identity would lie at the basis of the concept of likeness. Identity, it has
been said, cannot be defined, but likeness can be defined by means of
identity, namely as the respect in which the identity takes place; likeness
is the relation that subsists between things that fall under the \emph{same}
species\footnote{\emph{Husserl}: \emph{Logische Untersuchungen}. II. p. 112
f.}. But it must then be asked how we come to form species-concepts. Does
the species-concept not express that there are certain points of view from
which the things in question can be regarded as identical, so that they can
be put in place of one another? And this is precisely the definition of
identity (see 49; 69), which can therefore very well be defined. The
possibility of substitution is now the highest degree of likeness. Likeness
is thus the most comprehensive concept, identity a special case of it. What
likeness is in general cannot be defined. That two things resemble each
other is something we apprehend immediately, just as that a thing is red or
blue. (See 21.) But if I presuppose likeness as immediately known, I can
define identity as the degree of likeness that makes substitution (in some or
all respects) possible. This definition has the great significance that it
reminds us to distinguish well between the different degrees or kinds of
likeness, since not all such can be used for the function so important for
the thought-work that substitution is. It also shows the way by which we
reach the concept of identity. It is an analytic definition, in that we go
from a larger whole (likeness in general) to a part or kind (identity), just
as when in geometry we define the point by the line, the line by the
surface, the surface by the solid\footnote{Cf. \emph{Zeuthen}:
\emph{Matematikens Historie} I. p. 102 f and \emph{E. Mach}: \emph{Erkenntnis
und Irrtum}. p. 354 on the advantages of analytic definitions.}.

One can therefore also not define likeness as identity $+$ difference. This
is a single kind of likeness (see 21), but not the only one. When two colors
resemble each other, one cannot ``divide'' this likeness into identity and
difference. Likeness here stands for us as an immediate quality that appears
involuntarily when we place two colors like yellow and orange side by side,
at least when we compare this relation, e.g., with the relation between
yellow and violet. There is, as said, a kind of likeness to which that
definition can fit, namely the composite likeness that is present where two
objects have in some of their parts wholly the same constitution, in others
a different constitution. We can often substitute a part of one thing for the
corresponding part of another thing, e.g. put a leg of one chair on another
chair whose seat and back are different from the first's. But this composite
likeness is always derived. It presupposes the separating-out of the concept
of identity from the general and indefinite concept of
likeness\footnote{On the fateful influence this definition has had in
psychology, cf. my \emph{Psykologi}\textsuperscript{5} p. 208, 221 f.}.

We define identity psychologically-genetically as the highest degree of
likeness, and logically, in agreement with \emph{Leibniz}, as possibility of
substitution. This logical definition is experimental. In each single case
differences will make themselves felt, and we must test our way forward, how
far substitution can be carried through. It depends in the individual cases
on the task and the point of view whether objects are to be regarded as
identical or not. It is the purpose that decides. If I have a purpose that
for me is one and all, everything that does not serve this purpose stands as
identical, as indifferent. For the fanatic, all belief that does not
coincide with his own (positive or negative) belief is indifferent.
Intellectually regarded, identity is a mighty weapon of thought against the
confusion of multiplicity.

In concept-formation whole groups of objects come to stand for us as
examples, both in the sense of equally valid representatives and in the
sense of prototypes for future ordering. \emph{Socrates} demanded of
Euthyphro that he should determine for him the concept of piety, so that he
could then use the definition as an example or prototype (the Greek
παράδειγμα means both), and then call what accords with the concept pious,
and not what does not (Plato's Euthyphro p. 6 E). But actions that accord
with the concept of piety can very well be different among themselves and
not everywhere be placed in place of one another.

By reason of the law of the conservation of energy, the different natural
forces can be regarded as equivalent and substituted for one another, where
it is merely a matter of the energy-standpoint itself, not of the special
form of the energy or its practical significance. Potential energy is thus,
from that point of view, identical with actual energy. In the psychic domain
we can within certain limits apply a corresponding point of view. We can here
too speak of equivalents. Thus self-mastery is only possible by reason of
the equivalence of different drives or interests. A drive can (when it does
not die a slow death from lack of nourishment) not be supplanted except by
another drive's drawing energy to itself, as when, e.g., the interest in art
supplants the drive to sensory enjoyment. The identity-standpoint is here:
something that fills the mind and lays claim to its energy.

A great example of the purification of the quality of likeness into identity
is the principle of the absolute uniformity of all parts of space and time.
This thought works its way forward in the thinkers of the Renaissance
(\emph{Telesio}, \emph{Bruno})\footnote{\emph{Den nyere Filosofis
Historie}\textsuperscript{2} I. p. 89; 119 f.} and lies at the basis of the
more recent time's exact natural science. It is formulated in the following
way by \emph{Maxwell}: ``The difference between two events does not depend
merely on the mere difference between the times and places where they occur,
but only on differences in the nature, form, or motion of the bodies
concerned''\footnote{\emph{Matter and Motion}. § 19.}. Antiquity, on the
other hand, spoke of ``natural places.''

To take a very external example: things of the same weight---a human being,
a block of ice, a lump of gold, and so on---can be regarded as identical
when weight is the only point of view.---The word ``thing'' itself expresses
an identity of highly different objects, when by ``thing'' one means
something distinguishable that is to a certain degree concluded as a whole in
itself. Everything is a thing.---Here belongs also the beautiful explanation
that Adam Homo's mother gives her little son of what it is ``to be.''---

Naturally there are different points of view from which identity is asserted,
not equally significant. In its struggle with multiplicity, thought ascends
from external and accidental points of view to essential points of view that
can be held fast under the change of objects. The world of points of view
thus built up is the true world of Ideas. It does not hover over experience,
but is precisely the condition for experiences being able to be made in the
strict sense of the word, not merely isolated observations that cannot even
be distinguished from illusions. The objects must, in order for true
experience to be possible, be so constituted that they permit the
application of firm points of view. It is this that \emph{Plato} in his
language expresses by saying that all phenomena ``strive after'' the Idea,
especially the Idea of Likeness (\emph{Phaedo} p. 74--75), pure identity. We
would rather say that it is we who strive to form thoughts that can lead us
through the multiplicity of objects. And while identity (``the One'') for
Plato denoted the absolute conclusion of the thought-work, it is for us
indeed the conclusion of a thought-work, but at the same time the beginning
of a new one, since it will be a matter of bringing the identity-standpoint
won into relation with other points of view.

\subsection*{b. Quality-Relations}
\addcontentsline{toc}{subsection}{b. Quality-Relations}

74. Where substitution is not possible, the multiplicity and the differences
remain standing. All differences are immediately qualitative differences,
since the word quality denotes the immediate peculiarity with which an
object (something distinguishable) appears in relation to other objects, a
peculiarity that can only be described through likenesses and differences
toward other objects. The very difference between likeness and difference is
a qualitative difference. Other examples are: the differences between our
different senses and between qualities or nuances within the same sense;
relations of magnitude such as fullness and lack, whole and part, as they
immediately act on us; the difference between the psychic and the material,
between the known and the unknown, between the real and the unreal, between
the valuable and the valueless, between values among themselves.

When thought stands before such qualitatively different objects, and
substitution is not possible, at least provisionally---its first task is to
order them according to decreasing likeness (or increasing difference).
Qualitative likeness and qualitative difference would then mutually bring out
each other. Herein lies the intellectual significance of ordering and
classification. It brings about a certain continuity, which in turn can lead
to the discovery of finer discontinuities than the originally given ones. And
conversely, many immediately appearing discontinuities can be diminished or
abolished by an ordering of the mentioned kind. This work succeeds in
different degrees. The inconvertible and transitive series (5) and (6) here
become of special significance. Not only does each member here have its
definite place, but transition can constantly take place to new members
according to the same law as that by which the preceding members were
reached, and intermediate members can be pushed out, so that new relations
can be formed between members that previously lay far from one another.

There are now some in-themselves indescribable and indefinable qualities
toward which attention has early been directed, because they are quite
especially suited for series-formations. These are time, number, degree,
and place. The series that these four qualities each form can then be used to
cast light on other qualities that do not so easily permit clear
series-formations. We then place these more imperfect quality-series together
with those more perfect ones and can thereby, by way of analogy, reach
greater clarity about them and perhaps even discover new members in them,
hence lengthen them or make them more continuous.

We stand here before the concept of analogy and its extraordinary
significance for thinking in all domains\footnote{This point I have already
discussed in my treatise \emph{Om Begrebet Analogi og dets filosofiske
Betydning}. (Mindre Arbejder. Anden Række.)}.

75. For \emph{Aristotle} and \emph{Kant} analogy was no category. It stood
for them as the last means of connecting what could not go under a common
concept. They did not see that it is already analogy that prevails between
different objects that go under the same concept. Apart from the
identity-standpoint (73), the differences prevail, e.g. differences in
magnitude, form, and color in different horses. It is really only the
relation between the different parts or properties that can be said to be the
same in different objects for which the same concept holds. It is thus
relational likeness, analogy. (Cf. 69; 73.)

The history of science has demonstrated the significance of analogy as a
means of thought in the most diverse domains.

Already quite involuntarily it plays a role in man's interpretation of his
observations, as well as in all mythology, religion, and metaphysical
speculation. It is by an involuntary analogizing that man applies the
representations that stand for him with special clarity and order to as many
other domains as possible. It is an expansion that here operates, a spreading
of a tendency that has provisionally shown itself appropriate to objects
other than the original ones. There can be dispute about which analogy is the
right one, or, if several analogies are applicable, how the relation between
them among themselves is to be conceived. Religious and metaphysical disputes
often concern such questions. In the domain of strict experience such
disputes can often be decided by exact carrying-through of the analogy in all
its consequences and by holding these consequences together with the
testimony of experience. Thus the question was decided whether the movement
of light was to be conceived in analogy with a wave-movement (undulation) or
in analogy with a hurling-out (emission). In the religious and the
metaphysical domain it is both difficult, if not impossible, to carry through
the analogies applicable there with full consistency, and to institute strict
comparison between the consequences and experience; the justification of the
analogies can therefore not be verified.

\emph{Leibniz} first discovered\footnote{\emph{Den nyere Filosofis
Historie}\textsuperscript{2} I. p. 350 f.} that all idealistic metaphysics,
and thereby also all religious representations, rests on analogy, in that man
conceives existence's innermost forces in likeness with what stirs in his own
soul.---A materialistic metaphysics too, for that matter, builds on analogy,
only that it fetches this from another side of experience.

An extraordinarily important use of analogy was made by \emph{Kant}\footnote{See
the section on ``The Analogies of Experience'' in ``Kritik der reinen
Vernunft.''}, in that he taught that the causal proposition only expresses an
analogy between the changes in nature and the transition from ground to
consequence in thinking. The series of changes and events are thus
interpreted in analogy with a series of premises and conclusions in logic.
This too was a discovery. Earlier one identified the concepts ground and
cause (thus Spinoza and Leibniz), and when this identity showed itself
untenable, it stood as the only way out to deny the causal proposition
(Hume).

\emph{Maxwell} and \emph{Cournot} conceived the application of the concept of
number to real objects as grounded on analogy between the number-series and
the series of objects. ``By physical analogy,'' says \emph{Maxwell}\footnote{\emph{Om
Faradays Kraftlinier}. 1855. (German trans. by Boltzmann in Ostwald's
``Klassiker der Naturwissenschaft,'' p. 4.) Cf. my book ``Moderne Filosofer''
p. 83 f.---E. Mach and H. Hertz have carried Maxwell's idea further. (Moderne
Filosofer. p. 90 f; 93.)}, ``I understand the partial likeness between the
laws of two domains of experience that brings it about that the one
illuminates the other. Thus all applications of mathematics in science are
grounded on relations between the laws of the physical magnitudes and the
laws of the whole numbers, so that the endeavor of the exact sciences aims at
tracing nature's problems back to the determination of magnitudes by
number-operations.'' And \emph{Cournot}\footnote{\emph{L'enchainement des
idées fondamentales}. (1861.) I. p. 13 f; 22.} has stated that the concept of
number and the concept of grouping and ordering hang closely together;
quantity is a kind of quality that can vary continuously and thereby serve
for regular determination in a way that no other quality is able to.

Within logic, now the analogy with geometry, now the analogy with algebra has
played a role. One has made the concepts' relations to one another intuitable
by the relation between smaller and larger circles or lines, or one has
formulated the judgments as logical equations. And sometimes one has found
the strict logical proof in the possibility of such formulations, although
the validity of the analogy itself is, after all, the constant presupposition.

Within physics the so-called mechanical conception of nature offers a great
example of a carried-through analogy, namely the analogy between movements in
space and qualitative changes. Light, sound, electricity, and magnetism were
conceived as expressions of wave-movements in media of different kinds. The
analogy brought one domain after another under its dominion. And during the
great work here undertaken, analogy between those different phenomena among
themselves has again played a great role. Thus the analogy between sound and
light in \emph{Young}, the analogy between electricity and magnetism in
\emph{Faraday}, and so on\footnote{\emph{Th. Merz}: \emph{History of European
Thought in the 19th. Century}. II. p. 16--89, has a whole series of
examples.}.---\emph{Leibniz} conceived rest in analogy with movement, while
Greek thinking was inclined to apply the reverse analogy and precisely
therefore found such great difficulties in thinking movement.

New analogies will be to develop, if the modern attempts to lay
electromagnetism at the basis of mechanics succeed. It has shown itself that
from the theory of electricity propositions can be derived that are analogous
to mechanics' propositions about work and energy\footnote{\emph{Chr.
Christiansen}: \emph{Om det elektromagnetiske Grundlag for Mekaniken}.
(Tidsskrift for Fysik. V.) p. 17.}.

Within biology the analogy between different organs, by reason of the
likeness in the significance they have in different species, contributes to a
clearer understanding of such organs' forms and changes. One and the same
function is performed by organs that differ to a high degree in structure and
degree of development. There is an analogy between lungs, gills, and tracheae,
and there is at the same time an analogy between such special organs'
function and the skin-respiration that also takes place in animals that have
no special breathing-apparatus, but that can breathe with the whole surface
of the body.

Between the nervous system's structure and function on the one side, and the
mental life's peculiarity and mode of operation on the other side, there
subsists a significant analogy. In both places an activity is exercised that
has the character of a combining, a concentration. However one conceives the
relation between mental life and nervous system, this analogy must play an
essential role in the formation of the hypothesis.

Analogous problems can arise in very different scientific domains. A problem
similar to the one that radioactivity sets physics is set in biology by the
mutations and in psychology by mental new-formations and sudden changes of
character. And thinking's stance toward such questions must also be
analogous.

Sometimes within a natural science there can arise a dispute similar to that
between \emph{Luther} and \emph{Zwingli}: between ``it is'' and ``it means.''
Thus when some chemists (in more recent times, e.g., \emph{Arrhenius}) hold
that the atoms of the elements continue to subsist as such in the compound
substances, while others (like \emph{Ostwald}, \emph{Mach}, \emph{Ramsay})
maintain that it is only figurative speech when one says that the atoms
subsist after the composition has taken place. What one can ascertain is only
that the weight of the compound substance is equal to the weight of the
elements of which it is formed. There is otherwise no qualitative likeness
present.---Of an analogous question in the psychological domain there has been
spoken at an earlier place (5).

However it may now be in the individual case with respect to the question
``identity or analogy?'', it is certain that it is a sign of dogmatism to
confuse them. In philosophy a rationalistic dogmatism came forth by the
identification of ground and cause, whereby substitution and succession were
confounded. One overlooked the great difference that the relation of time
entails toward all purely logical and mathematical determinations. In natural
science, and in the philosophy that believed it had in natural science's
analogies the absolute truth, a materialistic dogmatism came forth in that
the mechanical conception of nature is regarded as proved by the possibility
of its applicability to the different domains, in that one overlooked that
the mere analogy just as well as identity can make such applicability
possible.---

After having demonstrated by examples the significance of analogy for
thinking, where substitution is not possible, we now return to the
earlier-mentioned four qualities (time, degree, number, and place), which can
afford a basis for fruitful analogies and thereby cooperate in the
rationalization of other qualities.

\subsubsection*{1) Time}
\addcontentsline{toc}{subsubsection}{1) Time}

76. No thinking, no consciousness is possible without a change of
qualities. It is change that maintains consciousness and sets thought in
activity. But it is at the same time change that sets us all our problems.
Change, alteration---time in the most elementary sense of the word---is
itself a quality, an immediate experience that cannot be described. That B
replaces A is as simple an experience as is at all possible. It does not even
necessarily presuppose that we have a distinct consciousness of A and B each
by itself; the change can be felt as an indefinite impulse. The simplest
experience of time is the sensation of change. A representation of time, time
as a category, first becomes possible through expectation and memory. But
even where expectation and memory make themselves felt, qualitative
differences still prevail within the representation of time. It will at first
be especially practical motives that entail this representation. If every
urge, every wish, were immediately fulfilled, it would scarcely arise. But
when there is an interval between the appearance of the urge or purpose and
the aimed-at state, we get three members: the now, where the urge is
felt---the intervening time, where the fulfillment is awaited or aimed
at---and the attainment, or the disappointment. These members can be
qualitatively different to a high degree. Think of the difference between
urge---lack and longing---and enjoyment. Most pronounced the difference
between the members becomes when in the intervening time we are compelled to
hold ourselves passively waiting; there arises then a state of inhibition or
a subduedness that stands in sharp opposition both to the flaring-up of the
urge and to the dwelling of enjoyment. The more active, on the other hand, we
can be in the intervening time, the greater the continuity between the three
members. With the highest degree of activity, time can as good as vanish.

A fourth member can be added when there is such a surplus of energy that
thoughts can be spent on what lies before the now, hence before the urge or
the thought of purpose. A passionate urge will look only forward, not
backward. Memories get practical significance only by entering as members
into wishes and purposes. Dwelling on the past, the time-before-the-now, is a
sign of a certain liberation from the practical motives. Memory itself can
get immediate value, even if what is remembered belongs absolutely to the
past.

When the representation of this four-membered rhythm is formed, and when
experience shows that this rhythm recurs in highly different
experiences---when at the same time many such rhythms have been experienced
and in memory join together into a chain---then the representation of time as
a general form or frame for highly different qualitative content can form
itself. And it then shows itself that the frame can constantly be expanded by
virtue of the same rule according to which it is formed. The attainment can
beget new urge, and so on, and the past that was remembered can awaken memory
of a still more remote time, and so on. Step by step there can then take
place a purging of all qualitative and practical elements. This very purging
can be of practical significance through the greater overview and the more
definite measuring-out that thereby becomes possible. It can be practical to
disregard the practical.

The relation between the members of the time-series then becomes pure
succession, the pure relation of ``before'' and ``after,'' a relation that is
not in itself assumed to have any influence on the qualitative content of
time. The Maxwellian principle (73) then comes into force. Every moment is
then assumed qualitatively the same as the other, and the relation between
them conditions a progressive difference-series (66). Time has then become
``a pure form of intuition,'' or rather a typical individual-representation:
the different times are conceived as belonging to one and the same time,
which advances without being able to be concluded.

77. But this whole process of purification cannot, as experience shows, be
accomplished except under the influence of two other categories that we have
not yet discussed, namely space and number. The independence of the
representation of time is given by the fact that we have an immediate
sensation of change, and that we can recognize (estimate) very small
time-series. In longer time-series, however, we are compelled to use space as
schema or symbol. Here, then, analogy intervenes. And in the measuring-out of
time the concept of number finds application, just as in the measuring-out of
space. But the application of both these two categories to the apprehension
and valuation of time presupposes that the purging of time's primitive
qualitative nature has already advanced to a certain point. The indifference
of the moments must first stand firm.

A definition of time is, as already remarked, not possible. Mechanics too
does not define time, but only what it understands by two times being equal.
When two bodies of quite the same constitution, under quite the same
conditions, starting from the same place and in the same moment, traverse the
same length of path, the times they have needed are said to be equal. But the
presupposition here is that all moments are the same---in a qualitative and
in a quantitative respect. There must then have taken place an elaboration by
thought of the representation of time before the mathematical determinations
can come into force. But on account of the tendency to expansion and
projection that is peculiar to consciousness, this does not see that time in
its fully developed form is in a certain sense its own work, conditioned by
its own forms; it conceives time as something existing in itself, realized in
the movements of the heavenly bodies (Aristotle) or standing fast as an
absolute order of things (Newton's ``true, absolute, mathematical, equably
flowing time''). In opposition thereto one has then wished to limit time's
validity to being merely subjective or empirical, in that one kept to the
fact that the relation of time, through the analogy with the relation of
space, could be expressed as a purely mathematical relation of dependence.
Or one has regarded the relation of time as conditioned by a falling-away
from the rest of unchangeable eternity, whereby the restless striving after
the future stepped in place of the concentrated and concluded
rest\footnote{Cf. the remarkable utterance of \emph{Plotinus}: \emph{Ennead}.
III, 7, II, where time is derived from the world-soul's unrest (δύναμις οὐχ
ἥσυχος).---\emph{Augustine}, in his acute theological-psychological
discussion of the concept of time, lets time first set in at the creation and
can thereby reject the question what God did \emph{before} the creation.
\emph{Confessiones} XI, 13.}.

But there is and always remains a historical element in our cognition---despite
exact natural science and speculative metaphysics and theology. In individual
great moments existence can perhaps, insofar as our experiences and lived
experiences have reached a certain rounding-off, stand for us as a logic or
as a system of equations, or we can see ``the still world at our feet in
eternal evening light.'' But again and again we must dive down again into the
stream of time and of changes. Time can never be ``abolished'' once and for
all.

\subsubsection*{2) Number}
\addcontentsline{toc}{subsubsection}{2) Number}

78. When I go from one object to another, or from one moment to another, a
series of acts of consciousness or of thought is performed, each of which
corresponds to an object, no matter what this is, or to a moment, no matter
how it is filled---no matter whether the objects or the moments are
homogeneous or heterogeneous, single or composite. Such acts can be regarded
as identical, apart from their content or objects. Herein lies the
possibility of the concept of number.

But here too the purging of the qualitative differences takes place step by
step. Provisionally, each act of consciousness or of thought has a quality
determined by its content and by its place in the time-series. From the
content it is easiest to abstract. There then remains the peculiar
constitution that the act of thought gets by the place in the successive
series of acts of thought at which it takes place. This constitution can be
designated as a number-quality. For psychological reasons a difference makes
itself felt, e.g., between the seventh and the eighth act of thought, when I
go through a series of objects. I feel somewhat different at the one than at
the other---this is due now to fatigue or to other causes. And the difference
will be the greater, the greater the distance between two members within the
series, when they are compared.

The most primitive form of the concept of number is the serial number that
the single object gets as the object of an act of thought that follows after
or precedes other acts of thought concerning the same series of objects. The
condition for such a serial number, or for the number-quality, is that the
objects do not arise absolutely continuously, but can be distinguished in
time or in quality or in both respects. There can then be formed an
identically varying difference-series (65).---It is more easily formed by the
successive appearance of the objects than when these appear simultaneously.
In the latter case we ourselves must take the initiative, while the
succession itself contains an incitement to connect the following with the
preceding.

The proper concept of number, the concept of amount, presupposes the
discovery that one can by many different ways come to the same place in the
number-series. I can, e.g., come to No. 8 from any other number whatever,
however great the distance is, and on whichever side it lies. The difference
is only that more or fewer acts are required in the different cases, and that
the direction can be different. These acts are of the same kind as those that
led from No. 1 to No. 8. Thereby the concept 8 is defined in several different
ways, not merely as $7+1$, but also, e.g., as $100 \div 92$. It shows itself
hereby distinctly that the different acts of thought are really only
repetitions of one and the same act of thought---that they are identical
among themselves, independent of the place in the identically varying series
as well as of the constitution of the object. The concept of number can now
get its purely formal character; the quality-element is separated out.

\emph{Helmholtz} and \emph{Kronecker}, in their psychological-genetic
development of the concept of number, take their starting-point from the
ordinal number as a mark that indicates the single element's place as a
member in a series. They then seek to show how the concept of number is
gradually developed to its pure form.\footnote{\emph{Helmholtz}: \emph{Zahlen
und Messen, erkenntnis-theoretisch betrachtet}.---\emph{Kronecker}:
\emph{Ueber den Zahlbegriff}.---(Both treatises are found in the Festschrift
for E. Zeller 1887).} Against this \emph{Georg Cantor} has protested. For him
the concept of number is a general concept that holds for different sets
(whatever the constitution and ordering of the elements within these sets
are). Ordinal numbers---in the sense in which the two named authors take this
word---are for Cantor ``the last and most inessential thing in the scientific
theory of numbers''; they are marks for indicating order (notæ ordinales),
not proper numbers (not numeri ordinarii).\footnote{\emph{Cantor}:
\emph{Mitteilungen zur Lehre vom Transfiniten}. (Zeitschrift für Philosophie
und philosophische Kritik. Neue Folge. 91. Bd. 1887.) p. 86.}---It seems to
me that Cantor overlooks that the starting-point for a presentation of a
concept's course of development can very well later, when the development has
reached a certain conclusion, come to stand as something ``inessential'' and
derived. A genetic presentation follows another order than the systematic.
The concept of number as a general concept has had a long history of
development, which it can be of great epistemological interest to condense
into an overview, as Helmholtz and Kronecker have done.

It is the positive significance of repetition that separates the theory of
numbers, and mathematics as a whole, from logic. Purely logically, as
\emph{Leibniz} showed, $A + A = A$, or as \emph{Boole} expressed it: $x^2 =
x$. A concept's content or extension cannot be changed by repetition. A horse
is a horse, however many horses there may be. But mathematically $A + A = 2A$
(or $A < A^2$). The concept of number comes into being by a synthesis of
identical differences, that is, differences that are objects of one and the
same repeated act of thought---or else by a synthesis of different
identities, that is, of objects that are only different by being the object
of acts of thought that fall in different moments. Repetition presupposes at
once identity and difference. Logic keeps to the fact that repetition
presupposes identity. But mathematics points out that even identical objects
can be the object of different acts of thought. Two concepts cannot be united
by synthesis into a new concept (concept-combination) except when they are
mutually different. But two numbers can by synthesis produce a third number,
whether they are identical or different, because in both cases it is really
the acts of thought that are combined. Logic is powerless in the face of
absolute differences. (Cf. 62 and 69.) But mathematics' first beginning is
given with the thought that even in identities (A A A\ldots) different
(time-different) acts of thought are applied, and even in differences (A B
C\ldots) identical (only temporally different) acts of thought can be
applied. Mathematics becomes, by taking this point of view, a more concrete
science than logic.

The process by which the concept of number comes into being is continued
through the history of the theory of numbers. Already above there appeared
two directions in which one can count: from 1 to 8 or from 100 to 92. By
reason of these two directions in the numbering, the number-series can come
to comprise both the positive and the negative numbers; they mutually form
the continuation of each other. And now between the numbers in the
number-series (the amounts) one can continue numbering and counting---by
fractions and by the approximate number-values to which attempts to determine
irrational relations lead. If the number-formation goes in another direction
than the double direction denoted by the series of the positive and the
negative numbers---sideways, instead of forward or backward---one gets the
series of imaginary numbers (symbolized by a line perpendicular to the line
of the usual number-series). Ever greater continuity is reached within the
series---by differentials of different order---as well as beyond the
once-formed series---by infinitely growing numbers. The concept of infinity
has its basis in the fact that the same rule and law can hold, however short
the distance between a series' members is thought, or however far continued
the series as a whole is thought to be. As \emph{Leibniz}\footnote{\emph{Nouveaux
Essais}. II, 17. (Op. phil. ed. Erdmann p. 244.)} has expressed it: ``La
considération de l'infini vient de celle de la similitude ou de la même
raison.''

The very possibility of forming such series, which by virtue of their law can
constantly be continued, is of interest for the characterization of human
thought. It shows us the constant relation of breaking of identity and
difference in pure form and in the domain where it can get its sharpest
determination. Thereby a free and yet point-by-point law-governed expansion
becomes possible. It is of purely formal nature, since the question always
remains whether there are objects that can correspond to the different
members in the series. But it has its value in itself as satisfaction for an
intellectual urge that is grounded in the general nature of human thought---an
urge that gets air in the pure world of forms, just as it leads to working in
the world of real objects and in the world of values. And through the
satisfaction of this urge an important means of thought is brought about,
since, as we have already seen, the number-series by way of analogy makes
ordering and inference in other domains possible. Quantity, which immediately
is only a quality, appears, through the development of the concept of number,
as a peculiar category.

\subsubsection*{3) Degree}
\addcontentsline{toc}{subsubsection}{3) Degree}

79. The difference between qualitative differences and differences of
degree is not original, but is the fruit of an interpretation in which the
analogy with the number-series plays a decisive role.

From a purely psychological point of view no sharp difference can be drawn
between a sensation's degree of strength (intensity) and its quality. The
opposition between light and dark, between white, gray, and black, is for
immediate sensation just as much qualitative as that between yellow and blue
and red. Just as there are innumerable nuances and transitions between white
and black, so there are also successive transitions between the
color-qualities among themselves. We place gray between white and black, just
as we place, e.g., orange between red and yellow. It is not from a
psychological, but from a purely physical standpoint that one sets only
light- and dark-sensations in opposition to color-sensations; but here more
is presupposed than the immediate sensations, namely knowledge of a series of
physical differences, or rather of two physical series, the one consisting of
differences in the energy (amplitude) of the ether-vibrations, the other of
differences in their wavelength. To the first series corresponds the series
of sensations from light to dark, to the second corresponds the series of
sensations from red to violet. It is the practical or theoretical
acquaintance with physical optics that leads to distinguishing between the
strength and the quality of light-sensations. One and the same color can vary
with different physical light-strength without losing its quality, unless the
variation leads to the highest degree of strength, or to a point where no
light-stimulation at all takes place. And in a similar way it is with our
other senses. The difference between a weaker and a stronger tone, a weaker
and a stronger pressure, a lower or a higher degree of warmth, and so on, is
for our sensation of the same kind as that between two colors, but it is
interpreted as a difference of strength when we have knowledge of differences
in the strength of the outer stimulations. The strength of the outer
stimulations can be measured by the physical work that is performed---the
degree of warmth, e.g., by the height to which the mercury rises in a glass
tube. It is from such experiences that the concept of degree of strength
stems, and it is by way of analogy that it is applied to the objects'
originally qualitative differences. Energy is, physically regarded, measured
by the length of path over which a mass is moved. A number-series is then
formed, corresponding to the different lengths, and the analogizing can now
run its course, in that the number-series won is paralleled with the
difference-series of our qualitative states. If we had no standard at all
different from these states, the concept of degree could not be formed. The
concept of degree presupposes here not only the concept of number, but also a
space that can be measured.

\emph{Kant}, in his otherwise excellent presentation of the category of
degree (intensity), lays the weight on the fact that quality-series can be
formed which end with a complete cessation of the sensation.\footnote{\emph{Kritik
der reinen Vernunft}\textsuperscript{2} p. 217 f.} But no such single series
can be decisive. There is needed a comparison with another series that offers
a more perfect and clearer determination of the members and of their mutual
relation. And here the number-series and the series of places in space stand
as prototypes. Kant has here not applied the concept of analogy, although
there was here just as good ground for it as in the case of causality (cf.
75). It is only in reliance on the applicability and validity of analogy that
we can, as Kant aptly says, anticipate experience by setting up the
proposition that all experiences are intensive magnitudes.

When the concept of degree is applied to conscious activity itself, as, e.g.,
when we say that the psychic energy can have different strength, it is, as we
have already stated (see 2; 7), the analogy with the physical concept of
energy that guides us. It is, psychically regarded, the multiplicity and
heterogeneity that is combined, and the intimacy with which the combining
takes place, that can have many degrees. Every state of psychic work has its
peculiar timbre, and only by roundabout ways, through reflection, can the
synthesis that has operated in it be traced. And just as our sensation of
warmth often seems to us to have a far greater ``strength'' than the degree
of warmth that has in fact been present, so there can also sometimes (and
perhaps very often) arise a feeling of exertion and effort that in reality
does not correspond to the psychic work that reflection is able to
demonstrate. It belongs to the many possible illusions in the domain of the
inner life.---When we speak of spiritual ``progress'' or ``regress,'' the
symbolic application of number and of spatial length shows itself distinctly.
Although the measuring does not get beyond an estimate, it is nonetheless of
significance to become conscious of what the estimate is built on.

80. A quality-series stops at a definite point. Orange passes over into
yellow or red, lukewarmness passes over into cold or into warmth, water
passes over into ice or into vapor, and so on. Such experiences lie at the
basis of what has been called the law of definite number, which in more
recent times has especially been impressed by \emph{Renouvier} and
\emph{Dühring}. But there need not, as these two thinkers hold, be brought
about by it a break with the law of continuity. When a quality-series has
reached the zero-point (whether the zero-point means beginning or cessation),
the task must be to seek a new analogy. One must seek out series that extend
further than the one hitherto used for comparison, so that the beginning- and
end-members of the quality-series can correspond to members within the new
comparison-series. The whole earlier quality-series can then be regarded as a
member or stretch within a more comprehensive series. Thus, since it shows
itself that our color-series corresponds only to a fragment of the series
that the degrees of refraction of the light-rays form, we must go back to the
structure and mode of operation of the organs of sight and to their
significance in the struggle for existence---hence to a more comprehensive
series---in order to understand why color-sensations correspond only to a
limited part of the series of degrees of refraction. Even here enigmas
remain, which are a part of the general problem of the relation between the
psychic and the physical. The disappearance of light in the evening and its
return in the morning is ordered into a continuity when we take account not
merely of what is immediately accessible to our observation, but of the
earth's rotation and its position relative to the sun. It can often look as
if nature made leaps; but every such leap contains a summons to procure for
ourselves a more comprehensive field of view. Thus we conceive rest as
minimal movement, as a relation of equilibrium between forces, and thereby
abolish the leap between rest and movement.

The law of definite number is therefore not the expression of an independent
or definitive truth. It denotes a provisional ascertainment of an enigma,
points toward a new problem.

\subsubsection*{4) Place}
\addcontentsline{toc}{subsubsection}{4) Place}

81. The concept of place too, and the concept of space that belongs together
with it, has developed from being a concept of qualities to being a concept
of a pure form-relation that could be determined quantitatively, that is,
permit the application of the concept of number.

Several different senses---especially sight, the sense of touch, and the
sensation of movement---condition our apprehension of space. And in the
development of this, as \emph{Berkeley} has shown, analogy plays an essential
role. It was a triumph for Berkeley that one cannot form a representation or
an intuitive image in which only the properties ``common'' to the different
senses were contained. He rightly took this in favor of his critique of the
current doctrine of general representations. But he did not take account of
another side of the matter, namely that we can just as well refer the
sight-space, the touch-space, and the movement-space to the same concept as
we can refer horses of different form, color, and size to the same concept.
There is analogy between the different examples of a kind. One now takes the
analogy for granted, and that to such a degree that one is inclined to
conceive it as identity. But there must have taken place a great activity in
the domain of sense-intuition and of elementary thinking before those
analogies became as ingrained as they are.

Even after it had come to stand clear that it was the ``same'' space that we
apprehend by sight, by touch, and by movement, space and its places
nonetheless continued to appear with pronounced qualitative differences,
fetched now from the one, now from the other of the cooperating senses,
though mostly from the sense of movement. Up and down, right and left,
forward and backward---these representations originally denoted qualitatively
different relations, and they still do so involuntarily in all of us, when we
are not precisely thinking mathematically or epistemologically. The different
movements and positions to which those expressions practically point bring
about the qualitative differences. In the ancient representation of ``natural
places'' the qualitative peculiarity of the places in space was held fast.
Each of the ``four elements'' had thus, according to ancient physics, its
natural place. Practical motives played a role along with it, as in the
apprehension of time. Space is first and foremost interspace---the space that
separates us from what we wish to attain, and that we must therefore traverse
or leap over. The practical apprehension of space is like the jumper's, when
he estimates the distance before he takes off. Practical interest leads to
comparing the one space with the other, the relation between new places with
earlier-traversed spaces between known places, and with this the
measuring-out begins, at first in the service of the will. The purely
theoretical interest here, as elsewhere, grows out of the practical by reason
of displacement of motive. Measuring presupposes an identification, in that a
part of the new stretch, if not the whole, must coincide with the
earlier-traversed stretch, insofar as this can be repeated in representation.

The more often one can now get different space-experiences wholly or partly
to cover each other, so that they can be apprehended as identical, the more
there will take place a purging of the qualitative differences of the places.
It then shows itself that the process, the successive measuring-out or
construction, can take place in quite the same way, from wherever I fetch the
stretches that are to be measured, or the places whose location is to be
determined. The identity, or rather the uniformity, of places and stretches
begins to make itself felt. It was, as already mentioned (73), in the
Renaissance period that this purging took place, in the domain of space as in
that of time. The thought of the indifference of nature (indifferenza della
natura) was carried through. It is of great significance for research, which
thereby got a direction to seek explanations of what happens in nature in
another way than by referring to the peculiarities of places and times.

But when the places in space are only different by their location, and the
one place's location is in turn determined by its relation to the other
place, and so on, always by virtue of the same law, it follows with necessity
that no absolute locations, absolute places, can exist. Only slowly did the
recognition of the complete relativity of place-determinations penetrate.
\emph{Copernicus} still assumed the sphere of the fixed stars to be communis
universorum locus. \emph{Kepler} refused to draw Bruno's bold consequence
that the fixed stars could not be assumed to be equally far from us. And even
\emph{Newton} still spoke of loca primaria, which at once determine their own
and other places' location. The difficulties are, however, here too no
greater than in typical and general representations in general: we think in
examples, and therefore we must, every time we speak of places' location, of
movements' direction, of inertia, and so on, presuppose a definite place in
relation to which all such determinations hold. In the concept of identity
itself, too, it showed itself necessary to indicate a definite point of view
in each single case (see 73).

Only now does the concept of pure space, as constructed by combination of
homogeneous elements, become possible. During its development it has, for
practical and theoretical reasons, been the interest in measuring-out, hence
quantitative determination, that has been guiding. Metric geometry has
therefore developed before projective or descriptive geometry, which directly
investigates the likeness and unlikeness of spatial relations, and so far,
logically regarded, forms the presupposition of the former, since
quantitative comparison presupposes qualitative. Both metric and descriptive
geometry have their interest beyond orientation in the surrounding world. For
we use the intuition of space to symbolize all relations of difference,
because space, as one has rightly said, is the most exact form of
difference\footnote{\emph{B. Russell}: \emph{The Foundations of Geometry}.
(1897) p. 186.}.

82. The combination of the homogeneous that is presupposed in all metric and
descriptive investigation of space can constantly be continued further by
virtue of the same law as that according to which it takes place. So far
space is infinite. Every limit we set is arbitrary, is due to special motives
of a practical or theoretical kind. We must, as we have seen, at every
place-determination, at every determination of a direction, set a limit, take
our standpoint, and provisionally regard this as ``absolute,'' as lying on a
world-axis. But every such standpoint must in turn be determined by another,
and so on. In principle, repetition must here always be possible, since we
have to do only with our own thoughts, here acts of construction and
measurement. Another matter is whether the objects that appear, without our
ourselves producing them, will also constantly give us occasion to make
constructions and measurements, indeed whether they will make such possible
at all. It goes with space as it goes with all other categories: it is a
predicate-concept, and it depends on whether we will constantly get objects
that make the application of this predicate possible or necessary. From its
great significance for our cognition, especially as regards exactness,
nothing can in this respect be inferred with certainty.

From another side too the problem of infinity emerges, as regards space.
Between two places whose location is determined, there can always be found and
determined intermediate places. This process too can in principle constantly
be continued. While before it was the relativity of places that led over to
the concept of infinity, here it is the continuity of space. Here too we must
distinguish between the formal and the real consideration.

The ``absolute'' space, whose places are qualitatively the same and mutually
determine each other to infinity, and which not only can have no limits for
its extension, but is a continuum in all its smallest parts---this space is,
as \emph{Kant} has said, an Idea\footnote{\emph{Metaphysische Anfangsgründe
der Naturwissenschaft}. p. 149. Kant hereby in fact acknowledges that the
sharp distinction he has made between ``forms of intuition,'' ``categories,''
and ``Ideas'' cannot be maintained.}, that is, a rule according to which all
movements are to be investigated. It is a rule, then, according to which our
measurements and descriptions of space in the individual cases are again and
again to be undertaken.

Just as little as we can without further ado decide whether space is
``really'' infinite and continuous, just as little can we decide whether the
kind of space we know---the measurements and constructions we can apply to
our objects---are the only possible ones. The space that our experience seems
to force upon us has three dimensions, and within it two straight lines
cannot intersect each other when they form right angles with one and the same
third line. The question has within mathematics been discussed whether this
space (the Euclidean) is the only possible space. If one defines space as
``being-outside-one-another'' or as ``extended multiplicity,'' those
peculiarities do not follow with necessity from this definition, which could
then give a genus-concept comprising several different species. It has shown
itself possible to develop consistent doctrinal structures from the idea of
space with other properties than those named. Which of the different possible
geometries is the ``right'' one we can just as little decide as whether there
``really'' exists an infinite and continuous space. In both respects the
decisive thing must ultimately be whether there are objects that make the one
or the other answer necessary. Thought has here more possibilities than
reality can be assumed to be willing to offer.

\subsection*{c. Negation}
\addcontentsline{toc}{subsection}{c. Negation}

83. Qualitative differences can, as we have seen, by the help of analogy be
interpreted as identities. From another side, there is contained in every
difference a negation. When two objects are different, the one is in one or
more respects not what the other is. And in the identity-relation itself
negation can be traced. For even if A is identical with B, there must
nonetheless be or have been something that made it that they could need an
identification, and in that respect there must then also be something that
could be asserted of A but had to be denied of B. (Cf. 69.)

But negation is only one side of the relation of difference. Every
determination and characterization necessarily excludes many properties and
is so far a negation. If one lays, with \emph{Spinoza}, the representation of
an infinite existence comprising all realities at the basis of one's
thinking, it is clear that every special characterization must stand as a
negation: omnis determinatio est negatio. (Ep. 50.) And there is the truth in
this proposition of Spinoza's, that in every sensation and in every act of
thought we take an object out of the great stream, the comprehensive
connection, in which it belongs. (See 2--3.) But in the relation of
difference we nonetheless always have also the mutual relation between the
objects among themselves in view from the positive side, and here the real
contrast will precisely contribute to bringing out each object's peculiarity,
so that negation does not become the only point of view.

Negation means in itself only the absence of an element (property or part)
that is or has been present in another connection. Therefore negation first
comes forth when we are about to leave intuition in its immediacy. (26.) The
concept No is, as \emph{Herbart} has aptly remarked, the clearest example of
a concept that in our judging springs from experience, although it has no
object in experience. The intuitive basis of negation is the opposition
between AB and A; this opposition makes it that I can call A, when it appears
alone, A (non B). Negation is due, like every determination, to a combining.

As with time and space and other categories, here too the concept's
practical motive precedes the theoretical. Through disappointment and lack,
men have learned that things and events in the world not only each have their
peculiarities, but that the one is not what the other is. There must be
choice, and there must be resignation. But in this school it was also learned
what an important means of thought negation is. Classification and induction
become possible only by it. (22; 26--28.) Negation operates now as a
reference to a later closer investigation, now as a sign of a way that cannot
be traveled.

84. Purely logically, negation is absence, exclusion, or abolition.
Mathematically, negation means a summons to series-formation (e.g. counting)
in the opposite direction. Negative numbers mean continuation of the
number-series in the opposite direction from the series of the ``positive''
numbers. It depends on the point of view which direction one calls positive
and which negative. Likewise with directions in space and in time. If I have
my essential life in the world of memory, the continued course of the
time-series becomes for me of negative significance. Negation as a direction
in opposition to another direction gets a definiteness that purely logical
negation does not possess.

There are no negative objects; negation always presupposes two members, each
of which is in itself positive. \emph{Plato} already determines negation as a
relation of opposition between two beings (ὄντος πρὸς ὂν ἀντίθεσις. Sophist
p. 257 E). How one regards position and negation depends on the point of
view. When the zoologist in his presentation passes over from articulates to
mollusks, it is natural for him in the definition of the latter to begin with
the proposition: ``The body is unjointed, without jointed limbs.'' When the
theologian has maintained his faith as the only true view of life, he
naturally comes to describe other views as negative, perhaps even ``purely
negative.''

Often we must of course stop at a negation, without our having any content
for a positive characterization of the one member in the relation. Negation
will then, for every living thought, lead to seeking such a content. If I
have sympathy or psychological interest for people who do not agree with me,
I will try to enter into their state. Significant as a motive of thought
negation becomes especially when it is two judgments that mutually negate
each other. A problem is then set (27).

Alongside identity and analogy, negation is the most important means of
thought. \emph{Hegel}, who has so strongly emphasized its significance in
this respect, often confused it with mere opposition, in that he overlooked
that not all opposition is negation. He saw in negation the possibility of an
extension of cognition without the help of experience; in fact he of course
always filled the empty place denoted by the negation with a loan from
experience. At the same time Hegel failed to appreciate the great
significance of analogy for the development of cognition, and especially for
all metaphysical speculation. His own system rests on a great and bold
analogy and is by no means merely constructed by a dialectic spurred by
contradiction.

\subsection*{d. Rationality}
\addcontentsline{toc}{subsection}{d. Rationality}

85. When the purging of practical and qualitative elements described in the
foregoing is completed, especially by tracing-back to differences in time,
number, degree, and place, it becomes possible to form series of a transitive
kind, so that inference can take place. Even the chaotic difference-series
permits inference, since it is transitive; but thereby only a new absolute
and indefinite relation of difference is put in place of another. Of
importance, on the other hand, is especially series (6), the partial
identity-series, which affords the schema for the relation between ground and
consequence. From $A \to B \to C$ follows $A \to C$.

The relation that appears here is analogous to the one that holds between
intuition or association and judgment. The judgment is only to express what
already lies in the intuition or in the association-series. Analogous is the
relation between predicate and subject. But transition can in these cases
take place involuntarily, by an estimate or an intuition. We become, or
believe we become, immediately conscious of the identity of intuition and
judgment, subject and predicate. Closer investigation can, however,
demonstrate that several different intuitions and associations have
cooperated in the formation of judgment and the determination of the
predicate. Estimate and intuition can be pathbreaking. They can lead to
discoveries that perhaps must and can be grounded by wholly other ways than
the one that has led to them. In the grounding it is definitely decided which
presuppositions cooperate, and how far their significance reaches.

The conversion from estimate or intuition to inference (grounding) is a new
example of the purging of qualities. A conviction that has formed itself
involuntarily by the cooperation of observations, memories, moods, and
incitements can appear with the stamp of simplicity and immediacy, just like
color- or smell-sensations. One then rests securely in it, as long as no
conflicting experiences awaken unrest and doubt. When this happens, the
question becomes whether it can be converted into inference. Reflection on
the objects and elements that condition the conviction and can be traced in
it, directly or indirectly, is the presupposition for this. In the new form,
as inference, the conviction gets a somewhat different character, but there is
no psychological ground to believe that it must necessarily lose its energy,
if only the conversion into the new form fully succeeds.

It must indeed be held fast that it is only a new form that is brought about
by the transition from estimate, intuition, or conviction to inference. We
still move within the world of the formal categories, and inference is a pure
form of thought, just like time, number, degree, and place. The
presupposition is still whether the objects, which we do not produce, permit
the application of our forms, even the most compelling of them for us. We have
no right, with Aristotle and Leibniz, without further ado to make
rationality, the relation between premises and conclusion, into a metaphysical
principle, the expression of a law of existence. By emphasizing this
difference between the forms we have at our disposal and their real
significance, we approach the transition to the real categories, and through
them to the world of problems. But before we make this transition, there are
still some remarks to be made about the category of rationality.

86. Just as there will for the most part be more in intuition than judgment
can express, and in the subject more than the predicate contains, so
conclusions will not always contain all that the premises contain. Precisely
the middle-concepts that bring about the connection sought fall away. It is,
as we have seen (71), a wish, a goal, that determines which objects it is
whose content we wish determined; the determination takes place by the
pushing-out of members in a transitive series, and the conclusion thus comes
to contain fewer concepts than the premises. In the series $A
\overset{\angle}{\underset{=}{\to}} B \overset{\angle}{\underset{=}{\to}} C
\overset{\angle}{\underset{=}{\to}} D$ I must push out B and C in order to
reach $A \overset{\angle}{\underset{=}{\to}} D$. The middle members can be
forgotten, since there is no more use for them. The middle members had the
double significance in the series of connecting and separating A and D. Now
one wishes only the connection; it is this that one takes with one to the
continued thought-work, and one looks back perhaps on the pushed-out members
as ``accidental points of view'' (cf. \emph{Herbart}'s ``zufällige
Ansichten'').

The conclusion thus here forms an equivalent for the premises in a certain
definite respect, which is determined by the whole direction and task of the
thought-course. It cannot in all respects replace the premises. Perhaps these
were---seen from other points of view---more significant than the former,
just as the hunt can be better than the quarry, and just as it can be better
to seek than to find, as \emph{Lessing} has stated it in a well-known saying.
Meanwhile there will always make itself felt an urge to get beyond this
one-sided relation and attain a mutual equivalence, so that ground and
consequence, premises and conclusion, come into as intimate a relation to
each other as possible. This happens when ground and consequence can change
places, so that from the conclusion one can always find the premises on which
it rests or which it contains. The accidental thing in the way by which the
result is reached then falls away; we have a little world, a totality before
us, whose members cannot be separated from one another. We are then beyond
the mere series-formations, which could be continued into the indefinite. We
then do not forget the origin of the truth, once it has come into the world,
and truth then does not so easily become a dead treasure.

How far and in how great an extent such thought-worlds, logical systems, can
be formed is a matter in itself. And a special question becomes whether our
whole life of thought can be concluded or rounded off into such a closed
whole. The possibility of chaotic difference-series can scarcely ever be
excluded; there will, as long as new objects emerge, constantly be a work to
do in order to reach up to the higher series-formations and through them to
the development of a world of thought.

\section*{D. Real Categories}
\addcontentsline{toc}{section}{D. Real Categories}
\label{sec:reale-kategorier}

87. Even if all other qualities could forever be converted into, or rather
interpreted as, identities or quantities, the difference of time would
nonetheless always announce itself anew. One could also think of the
qualities as eternal types in an absolute relation of identity to one
another: $A=B=C=D$. Thus \emph{Spinoza} conceived the relation between
existence's infinitely many, mutually irreducible fundamental properties or
``attributes,'' of which only two, Spirit and Matter, were accessible in our
experience. Existence was thus expressed in infinitely many forms that were
mutually logically equivalent or analogous. It would be a ``still world'';
the relation of time did not hold. But in the world of experience there is a
constant wave-motion. Every intuiting gives only a momentary image, and the
moments change.

The last great opposition in our cognition, which stands fast even if all
others are abolished, is the opposition between rationality and time: there a
relation between ground and consequence, here a relation between before and
after. There an eternal system, here a surging sea. \emph{Spinoza} chose the
first point of view, and for him time became in the end an illusion.
\emph{Hume} chose the second point of view, and for him the rationality of
existence therefore became an illusion. But our life of thought stands always
before a breaking between the two points of view and has the task of finding
its way through this breaking. It is therefore that we must make a difference
between the formal and the real categories. This difference is determined
precisely by the significance of time, which was especially impressed by the
transition from ancient to modern thinking (46--50). It necessitates the
setting-up of categories that are to hold for what is distinguishable in time.
It is a demand that is made from the objects to the objects (71).

\subsection*{a. Causality}
\addcontentsline{toc}{subsection}{a. Causality}

88. The formal categories are indifferent toward the difference between the
three kinds of immediate intuiting that we call observation, memory, and
imagination. On the basis of all three kinds of intuiting, comparison and
inference can take place, identities and qualities be ascertained,
differences in time, number, degree, and place. But when the question arises
how we are to conduct ourselves when there comes conflict between the three
kinds of intuiting, we get beyond the purely formal, though we take the forms
with us. When it is a matter of distinguishing sensation from memory and
imagination, we rely provisionally on an immediate impression, the
existence-quality (36), a certain distinctness and reliability that as a rule
is characteristic of the observed in contrast to the merely remembered or
imagined, and likewise of the remembered in contrast to what is produced by
imagination. But since there can be not only many degrees in distinctness and
reliability, but also downright conflict between different objects'
existence-qualities as well as between observations and memories, a criterion
of reality is sought, and this consists, as we have seen (37--38), in the
firm and law-governed connection between the objects. The concept of reality
presupposes that on the basis of given objects an expectation can be built
that is confirmed by an ever-growing circle of new objects.

The objects must then not form a chaotic difference-series. Even then an
expectation could arise---namely that there will constantly appear something
wholly new and different. But it could not be an expectation of something
wholly definite, namely of the same thing that had constantly appeared under
the same conditions. That chaos will continue to be chaos is indeed a
knowledge, but a wholly contentless one; it denotes the lower limit of all
real operation. Could we not get beyond it, our knowledge would be merely a
knowledge of the irrationality of existence. The thought-work constantly aims
at a rationalization of the objects through series-formations that give
identity more and more application. We work our way forward through ever more
definite expectations or hypotheses---and thereby approach the ideal we call
reality. Only for thought does a reality exist. Truth consists purely
formally in the mutual agreement of our thoughts; and the more immediately
given objects, not produced by thought itself, we can bring into a connection
corresponding to the purely logical connection between our thoughts (hence
between subject and predicate, between ground and consequence), the more we
reach real truth or reality.

One became early clear that the criterion of reality could not be anything
single, but had to consist in the firm and consistent connection of all
objects. But \emph{Kant} was the first who was clear that there was only
analogy, not identity, between the logical connection of thoughts and the
real connection of the successively appearing objects. It is the relation of
time that has compelled the distinction between analogy and identity, and
thereby in turn to make a definite difference between causality and
rationality. The former contains the latter; but the latter can be present
without the former: reality is truth; but not all truth is reality.

Just as time and space first appear practically as interspaces or distances
between our purposes and their attainment, so causality too is first of a
practical character, in that it makes itself felt, indeed forces itself upon
us, through the relation between means and goal. The goal is given for our
representation, in our drive or our wish, and we now seek the conditions
without which it cannot be reached. Here we in a way experience causality
before we think about it. Here is, as I have said elsewhere, our first
encounter with necessity. The unavoidable order in the succession of objects
expresses our first concept of cause.

89. The objects can often form series without there being members in such a
series that can be connected as cause and effect. In the series A B C D the
individual members can appear at definite times and in definite places
without standing in a mutual causal relation. (Cf. 62.) We have, however,
already applied the concept of cause in setting up the series as a series of
real events, for we must then have applied the criterion of reality, which
comprises the concept of cause in itself, at every single one of the members.
When we wish to go further and find an explanation of the fact that it is
precisely these objects that arise, and in this definite order, we must go
further back and find the cause of the whole series. When lantern-slides are
shown, the spectators have the light-source and the glass photographs behind
them. What shows itself to them in each moment seems accidental, when one does
not know what takes place behind. It is a similar situation that \emph{Plato}
has imagined in his famous cave-image (at the beginning of the seventh book of
the ``Republic'').

Our experience has in many domains a sporadic character, begins from
different starting-points and is continued in leaps. This sporadic character
cannot always be overcome through the demonstration of a direct connection
between the starting-points, but only by extending the consideration to a
connection that lies further back and is of a more comprehensive kind. The
connection between a series of lightning-flashes is found by going back to the
thundercloud's electrical state and the changing tensions and discharges it
brings with it.

Historically one must often begin merely by establishing the chronology. This
can then become the basis for a further-reaching investigation. Historical
criticism has first and foremost the task of investigating the reality of the
individual events each by itself, as well as their chronological relation.
Herein it makes use of the concept of causality, as this forms itself at the
given stage of development of cognition. Whether Hannibal has lived, and when
he has lived, is essentially a task of the same kind as that whether there has
been a primal nebula out of which our solar system has developed. In both
cases one infers back from given objects. One constructs the past out of the
present and its content. It makes no difference that Hannibal has existed only
once; the primal nebula too has existed only once. In every application of the
criterion of reality, hence of the concept of cause, it is a single,
individual fact that I wish to have certainty of. I wish to know whether it is
a real person who passes my window, whether it really is a chimney I see on
the house down the street, and so on. Afterward the question then comes forth
why the person went past, and how the chimney has come to be. The one task may
perhaps be solved even if the other cannot be solved.

Historical criticism presupposes law-governedness in the world of historical
events, even if historiography should not succeed in finding the causal
connection between the events among themselves, and even if comparative
historical research should not succeed in setting up general historical
laws.---The main thing in this connection is that we are placed before the
past as before present and future. Also with respect to the past there can be
a matter of expectations that are fulfilled. \emph{Zeuthen}'s conjecture that
the Greeks must have known the infinitesimal method was confirmed by
\emph{Heiberg}'s discovery of an Archimedes manuscript. A criterion of reality
for the correctness of a childhood memory we can get in a letter that is
found.

90. The concept of cause expresses the way in which we react to a successive
difference, a change that has brought about an abolition of our equilibrium.
There is equilibrium in our involuntary striving after a goal, until we meet a
hindrance; such a hindrance is the distance between the goal and our present
state; it is equalized, and the equilibrium is restored, when we find means
and ways by which our striving can be continued in the same direction as
before, though by a roundabout way. But also where no practical motives make
themselves felt, the equilibrium can be abolished. The appearance of new
objects or changes of those at hand more or less abolish the equilibrium in
our thoughts; a previously won connection is loosened or burst, and a new one
is sought. The intellectual seeking forms an analogy to the practical and
develops out of it.

The relation of difference continues provisionally to stand in the
foreground. Two objects replace each other and are more or less qualitatively
different. What we first seek to decide is whether the second always sets in
when the first has set in, and never otherwise. In its most elementary form
causality expresses an unavoidable sequence. It is then that the thought-work
begins which we have in the foregoing described from various sides---the work
of reducing the qualitative differences to the least possible. At the same
time a work will be set going to find intermediate members, as regards time,
so that the transition, both as regards time and as regards quality, can take
place as continuously as possible. Historically it is for a great part this
interest in continuity in the causal relation that has led to the development
of the formal categories. The causal synthesis has driven forward the formal
syntheses. The history of science shows that all explanations aim at
continuity. And the more this can be brought about, the more difficult it
becomes to point to a point where the cause stops and where the effect begins.
They pass over into each other gradually in quality and successively in time,
no longer stand as two different ``things.'' We discover that by our
distinction between cause and effect we have intervened in nature's continuous
course---that the ``leap'' must be set to the account of observation and of
our thought, not of nature. It goes here as with logical inference, where we
too grasp a single member in a whole series in order to become clear about its
relation to another single member. (71; 86.)

The change that gives the impulse to our thinking, our causal thinking too,
singles out individual objects in relation to others. There is so far from the
first a leap, not only between two objects that replace each other, but also
between these objects together and other objects, which nonetheless perhaps
lie in the same series and in continuity with them. It is our interest that
isolates the cause from what goes before it, and the effect from what follows
after it. Both leaps are abolished the more the causal relation can develop
from its elementary form, in which it is a successive relation between two
qualitatively different objects, to its ideal form, in which it is a relation
of continuity both as regards time and as regards quality.

91. Is the ideal now reached when the causal relation takes on the form of
continuity? Must one not go a step further?---The analogy with the relation
between ground and consequence (88) seems to require it. For in this relation
it holds that there must be nothing in the consequence (the conclusion) that
was not in the ground (the premises), and the most perfect grounding-relation
even requires that ground and consequence can be exchanged, that the
equivalence holds in both directions (86). This analogy led the seventeenth
century's philosophers (for whom there was here more than analogy) to
formulate the causal proposition in such a way that it required that there
must be no more in the effect than in the cause, that there must be absolute
uniformity between cause and effect. In \emph{Spinoza} this presupposition
appears most sharply pronounced. It can still be traced in \emph{Kant}, and
especially in some of his successors (\emph{Hamilton} and \emph{Spencer}),
who take the step fully and find a mutual equivalence contained in the concept
of causality. It is of great epistemological interest that this conception
played an essential role in the discovery of the law of the conservation of
energy. \emph{Robert Mayer}, \emph{Colding}, and \emph{Joule} all started from
the assumption that the conservation of energy was a necessary presupposition
contained in the very concept of cause. Mayer never tired of repeating the
proposition causa æquat effectum, which for him was a consequence of ``the law
of logical ground''; it was for him a self-evident proposition. Colding found
the opposite assumption ``contrary to reason,'' and Joule called it
``absurd.'' Nor did \emph{Helmholtz} have the consciousness of setting up a
new proposition\footnote{Cf. \emph{Den nyere Filosofis
Historie}\textsuperscript{2} II. p. 497--499; 597.}. Fortunately these
inquirers did not stop at this consideration, but subjected their view to an
experimental testing.

Against this way of considering it must be maintained that the concept of
cause would not lose all significance because it did not succeed in
demonstrating equivalence between cause and effect. A causal relation could
hold between the objects even if a constant variation took place, provided
only that the future was analogous to the past in such a way that from the
fact that the event $B_1$ follows after the event $A_1$, we could infer that
the event $B_2$ will follow after $A_2$---when $A_1\,A_2\,B_1\,B_2$ denote
heterogeneous events. Even if the qualitative differences were irreducible,
the causal proposition could hold. So much the more could it hold even if no
quantitative equivalence could be demonstrated between cause and effect, so
that, e.g., A was cause of $B = A + \times$, and B cause of $C = B \div y$.
Law-governedness does not necessarily mean equivalence; this does not stand as
self-evident.

The quantitative relation between cause and effect, which for many objects it
has succeeded in demonstrating, gets its significance, epistemologically
regarded, especially from the fact that it is only by its help possible to
apply the found causal relations to future objects. For purely qualitatively
the identity of the relations cannot be determined with exactness. The degree
of likeness that must be present between $A_1$ and $A_2$ in order for it to be
justified to expect $B_2$ after $A_2$, just as $B_1$ came after $A_1$, can
only be fixed by an estimate. Only by counting, measuring, and weighing can
the identity of the relations be demonstrated, so that certain recognition
becomes possible, and even then only approximately. Although no absolute
repetition can be demonstrated, either in nature or in history, it is
nonetheless natural that thought seeks to get as far in ascertaining the
identity of the relations as is in any way possible. It is then also on this
that attention is first and foremost directed within the special branches of
research, and we must take it as a fact that the more exactly the identity of
the relations can be demonstrated, the more it shows itself that there is
equivalence between what goes before and what follows after.

But even if the equivalence-relation were carried through everywhere, there
would nonetheless continue to be a historical element in our cognition, which
is given with the direction in which the conversion from the one object to the
other takes place. There can be causal equivalence without there being
value-equivalence. It is not only human history that shows that we cannot go
back as well as forward in the series of causes; nature too has its history,
which is characterized, physically regarded, especially by the circumstance
that the conversion from heat to movement does not take place with the same
ease as the conversion from movement to heat. The purely causal consideration
here shows itself insufficient, and there is need for other real categories
than the causal category, which is indifferent to direction. This holds
whether the causal relation can be conceived as equivalence or not. And this
makes it that the standpoint of continuity in reality becomes the highest
standpoint from which the causal relation can be seen.

92. A judgment in which an unavoidable relation between an antecedent and a
consequent is expressed, so that from the former one can infer to the latter,
is called a law. But that antecedent shows itself, on closer reflection, never
to be a single object, as the popular conception of the concept of cause is
inclined to assume. What we call cause is always an aggregate of several
cooperating objects, which we can call conditions. There is not a single
causal series, but a mass of series crossing one another, and the one series
cannot be understood without the other. In practice and in everyday speech we
keep to the most decisive or to the most conspicuous condition. From a single
object, however, just as little effect can be derived as from a single premise
any conclusion can be drawn.

The contribution that a single object makes to the fulfillment of the law is
called its force. Force is, to use \emph{Leibniz}'s definition, that in a
present state which contains the ground, or a part of the ground, of a new
state. Force contains in itself no explanation; but the scientific
significance of the concept of force lies in its pointing to the fact that it
is not only the new that we understand by the old, but conversely also the old
that is understood by the new. We learn in the effects the possibilities of
the old. The new that happens casts light on the conditions in the preceding
time. Words like possibility and force (which are of kindred meaning, since
``possibility'' [Mulighed] is kindred with ``fortune'' [Formue], which in the
older language means capacity or force) have their justification by containing
a direction to seek the conditions in the preceding objects that ground the
emergence of a later object. And on the other hand these words contain a
direction to the future; for possibilities, capacities, and forces attest
their reality through the effects that proceed from them. Against
\emph{Heinrich Hertz}, who wished to abolish the concept of force in
mechanics, \emph{Ernst Mach} remarks: ``Every mechanics contains a regard to
the future, since every mechanics must apply concepts like time and
velocity''\footnote{\emph{E. Mach}: \emph{Die Mechanik in ihrer
Entwickelung}.\textsuperscript{6} p. 281.}. This accords with Leibniz's
definition mentioned above. It gets significance especially when the state
preceding a change seems to be a state of rest or a state of equilibrium. From
the fact that change can set in, we infer that the rest was a stopped
movement, the equilibrium a relation of tension between forces operating in
opposite directions. We understand the simultaneous by the help of the
successive, the static by the help of the dynamic. It is this consideration
that leads to the formation of a concept like potential energy. It is likewise
this that leads the founder of the theory of mutations, \emph{Hugo de Vries},
to state that prior to the sudden appearance of a new type there must go the
formation of an inner predisposition---prior to the mutation a premutation.
This latter is a process that takes place quite in the hidden (sich völlig im
latenten Zustande abspielt)\footnote{\emph{Die Mutationstheorie}. I. p. 353.
Cf. II. p. 637; 662.}.

The greater a contribution to the fulfillment of a law an object gives, the
more active it is said to be. Since all activity, all force is measured by the
resistance that is overcome, it is always differences of force between
objects, relative differences of activity, that we have to do with in our
thinking. ``We cannot,'' says \emph{Maxwell}, ``even say what force acts on
us, but only indicate the difference between the force that acts on one thing
and that which acts on another thing''\footnote{\emph{Matter and Motion}.
\S\ 103.}. Thus the law of inertia really contains a summons to find the
forces that change the rectilinear movement, not the forces that in themselves
give the bodies the movement they have at the moment.\footnote{\emph{Zeuthen}:
\emph{Matematikens Historie}. II. p. 340.} The same holds in the case of
psychic energy. If a certain state of mind is given, we can ask what is
required to change it---how strongly the oppositions already found in it must
be sharpened in order for a transition to a new state to set in. The mental
life's changes take place just as little ``of themselves'' as those of
physical objects; it possesses no absolute force, but we come to know its
force through the oppositions it can encompass. A mental life that was not
placed in the breaking of oppositions we could not ascribe energy to.

The concept of force presses---in the strict sense indicated---more and more
into the foreground since Leibniz's time, in opposition to such concepts as
rest, state, matter, substance, essence. According to Leibniz, only that which
acts can be called substance, and every substance is in activity. Hereby the
concept of substance is eliminated, and the law for the activity steps in its
place. The single individuality finds its expression in the law for the
changes that take place in, with, and by it. The former dominance of the
concept of substance was bound up with the old presupposition that the
unchangeable and resting was the highest (48; 77), especially with the urge
for rest for thought in an Idea that did not again set reflection in motion.
Substance is an example of a dying, if not already dead, category. \emph{Kant}'s
``Ding an sich'' was one of its last forms. As regards the mental life,
\emph{Hume} and \emph{Fichte} have demonstrated the invalidity of the concept
of substance: immediate observation does not show us the soul-substance, Hume
maintained; and the more we come to know ourselves by way of reflection, the
more we discover, Fichte maintained, that our soul consists in acting.

\subsection*{b. Totality}
\addcontentsline{toc}{subsection}{b. Totality}

93. Even where we can, in the causal series, go both forward and backward, by
mutual equivalence, the direction is, as already remarked, not indifferent.
Equivalence with respect to energy is not the only point of view. It shows
itself psychologically in the fact that a mental state is not only described
by the combining energy it has at its disposal, but is also characterized by
the qualitative constitution of the psychic objects (thoughts, interests,
purposes) that constitute the center of gravity, the real self. Thereby the
mental life's direction at any time is determined, the striving in whose
service the synthesis ultimately stands. For the combining operation does not
place all elements on an equal footing; there are elements that occupy a
central place, our constant thoughts, our innermost urge and longing---and
there are others that occupy a more peripheral place. An exchange of these
places can be one with a mental revolution\footnote{Cf. my treatise on
\emph{Begrebet Villie} in the first volume of the journal ``Psyche.''}. And an
analogy to this is found in the material domain. It is not enough to
investigate the relation between cause and effect among themselves. This is
best done when one can isolate a circle of objects from all others. The
proposition of the conservation of energy can thus only be experimentally
demonstrated when one can form isolated wholes and keep influences from
without away. But this is then always ultimately an arbitrary limitation, an
abstraction. And the question becomes, when one returns to the real
conditions, always in which direction the causal series moves during its
continued course---which members have the possibility of here anew making
themselves felt. Direction belongs just as essentially to a movement as mass
and velocity.

\emph{Leibniz} therefore set up, besides the law of the conservation of
force, a law of the conservation of direction. According to the law of
inertia, movement does not change its direction, any more than its velocity,
except by the influence of outer forces. But even when the direction is
changed, the earlier direction is nonetheless cooperative in the new
direction, which is the resultant of the earlier direction and the outer
force's direction. There is thus continuity in the direction as in the
energy. Direction is not something that first emerges at a definite point of
the movement, but is present as far back as we go, and the earlier directions
determine the following ones, just as in the psychic domain our earlier
striving and willing determine the later.

Neither in the psychic nor in the material domain is there now a single
causal series, but innumerable ones. In the mental life there is indeed
constantly a certain centralization, but all elements in it strive after a
central position and influence, and there is therefore always a mass of
crossing directions. It then depends on whether they can be harmonized,
whether they can constantly anew be united in one stream. And in likeness
with this, directions of movement go out from every material point. We need
not even keep to the causal series; every single member in such a series,
every single ``cause,'' is really a system of conditions, each of which has
its direction, and the ``cause's'' direction is the resultant of all these
directions, each of which gives its contribution to it.

Every single cause is a totality, as well as every single causal series, and
as well as several different causal series that are connected into operation
in the same direction. There are here innumerable degrees and stages of
composition, but the principle is the same. There is the possibility that
directions diverge; but it must be taken as a fact that this is not always so;
directions can be connected, so that wholes, more or less harmonious systems
of causal series or directions of activity, form themselves. In how great an
extent this is the case, experience must decide. But experience demonstrates
that it can happen. And it is therefore that we, through the concept of
direction, form the transition from the category of causality to the category
of totality. By reason of the originality of direction, the totality is never
an accidental result. That directions are united has its ground in the
different processes' earlier directions, as far as they can be traced back. It
is arbitrary here to stop at a definite point.

As examples of such totalities may be named: solar systems, planets,
organisms, personalities, societies. From a general point of view they can all
be conceived as totalities or systems of directions. Common to them is that
they have properties different from the properties that appear in the more
elementary activities in whose cooperation they consist.

Only when one forgets the necessity of direction can one think of a chaos as
an absolute initial state. In such a chaos the elements of existence would
then be present, and when at later stages of development wholes arise, this
would be due to accidental collisions and the activities of the elements.
Especially in popular presentations it often looks as if existence proceeded
like Euclid's geometry from the simplest elements, corresponding to
definitions and axioms, to totalities, corresponding to the most special
propositions. And yet already \emph{Hesiod} was clear that the chaos out of
which he imagined the dynasties of gods to have developed had itself \emph{come
to be}. Apart from the fact that an absolute beginning cannot be thought at
all, and still less proved, every attempt to establish an initial state must
ascribe to it certain dispositions, definite tendencies of direction.
Everywhere one must presuppose a totality, however different this must perhaps
be thought from the totalities that appear at later stages.

Totality is a more special category than causality. One can think the latter
without the former, since there can be thought, and perhaps are found, causal
series that do not converge, do not lead to more or less harmonious
totalities. Totality, on the other hand, presupposes causality. The relation
between the categories is, after all, that the more special, more concrete
ones have a richer content, but a smaller extension. We therefore
unfortunately have no right to generalize the fact that totalities arise. The
great question, which in this connection we do not, however, have to do with,
is whether this fact is rule or exception.

Within the formal categories there is, in rationality, an analogy to the
concept we here stand at. Also in our life of thought totalities can form
themselves, in that premises and conclusions form a kind of logical wholes or
organisms.---Psychologically regarded, consciousness itself forms a whole by
virtue of the law of synthesis. And here too it showed itself that we have no
right to ascribe to the elements our reflection can distinguish an existence
outside the psychic whole of the same kind as within it, so that the whole
would merely be their product. In the psychological antinomy (4) we have the
example lying nearest to us of the irrational relation between whole and parts
that meets us at all stages of existence.

94. The category of totality stands in the way of every attempt to explain
wholes as accidental products of elements that are presented either as
directionless or with wholly indefinite direction. It denotes a point of view
we are compelled to take, since definite relations between the objects require
it. When one often sets teleology or finality in opposition to mechanism (cf.
56: Renouvier and Hartmann), it is either an unfortunate mode of expression or
else an incorrect conception. If the formation of the whole is to be explained
by a purpose, an intention, a preceding plan, then---quite apart from the
anthropomorphism---the relation between whole and elements becomes just as
external as in the mechanical conception, which lets the whole arise from
preexisting, independent elements. Whether one lets the whole go before (as
plan) or follow after (as result), one embraces an equally external
conception. One overlooks how the whole in a certain sense preexists in the
very earlier directions of the elements. It is only for our accidental
apprehension that it comes to be all at once. It was on the way
beforehand---in the different converging directions.

And when one applies an analogy with human volitional activity, in which the
whole that is aimed at can be given beforehand in a thought of purpose and a
plan, then it must first be remembered that voluntary willing develops out of
the involuntary and in an involuntary way (see 23). And secondly it must
indeed be noted that the life of thought and will itself is an example of
totality---so that the problem here recurs. For here too there are
elements---sensations, memories, urge and drive, and so on---whose directions'
convergence and harmony is a condition for a life of thought and will to
subsist. The very thought of purpose by which one would explain the formation
of totality is itself a totality. One must then in any case be conscious that
this analogy does not solve the problem. To be sure, we see in our highest
volitional life, in a decision taken after all-sided deliberation, the most
intimate connection of manifold tendencies of direction into concentrated
operation in a definite direction. But it is nonetheless still only an example
of what was precisely to be explained. One does not get beyond the irrational
relation between whole and elements. The epistemological significance of this
is that there are constantly new tasks to solve, while both teleology and the
chaos-theory close off, each in its own way.

By the help of the concept of direction\footnote{The significance of this
concept in this connection I have already emphasized in my lecture on
\emph{Vitalisme} (1898, reprinted in ``Mindre Arbejder'' I), in my program
\emph{Filosofiske Problemer} (1902) p. 66, and in my treatise \emph{Om
Kategorier} (Vid. Selsk. Overs. 1908) p. 32.} we get the possibility of a
connection between the two categories, causality and totality, a connection
that is not constructed, but supports itself on definite characteristic marks
in our objects, to whose number here thought itself also belongs.

95. If the different totalities are compared among themselves, they will be
able to be ordered in series according to points of view that lie in the very
concept of totality. A relation between unity and multiplicity will make
itself felt in them all. It is, after all, the cooperation between different
causal series that is the condition for totality. Thereby there is given a
relation between these series' multiplicity on the one side and the degree of
unity and concentration that the totality presents on the other side. The
richer a multiplicity that is gathered into unity, and the more pronounced
this unity is, the higher the totality can be said to be. Herewith at the same
time the greater or lesser degree of activity will be given. For the richer a
system of causal series that is formed, and the more consistent it is, the
greater the totality's independence and self-activity can be---the more
independently it stands in relation to its surrounding world. An object is
said to be the more active the more it has the ground and cause of its
subsistence and operation in itself.

According to this standard we will then be able to order the totalities that
experience presents. And it is also such a standard that is applied within the
special domains, e.g. in the comparison between planets among themselves,
between organisms, between personalities, between societies. More difficult
the comparison becomes when it is to concern the different kinds of totalities
among themselves. Here the point of view of degree (intensity) comes into
opposition with the point of view of extent (extensity). According to the
first point of view the organism stands above the planet, the personality
above society; conversely from the second point of view. This consideration we
cannot here continue; it leads over to a category we shall later discuss.
\emph{Pascal}'s famous utterance about man as the ``thinking reed'' that is
nonetheless ``nobler'' than the great masses of the universe intimates the
great problem lying here. Psychic energy presents the most intimate union of
unity and multiplicity that we know, and it is from it and through it that we
know and judge all other totalities.

But the higher a totality stands in the named respects, the greater enigmas it
presents to research. Already in \emph{Aristotle} this appears, in that he
laid strong weight on the fact that all knowledge concerns general relations,
but at the same time impressed that everything really existing is a single and
individual thing, something that can always appear only as subject, never as
predicate. (Cf. 53.) The individual (total) thus stands as a barrier for
cognition. According to the conception we have asserted of the concept of
totality, we must find that there is something correct in the Aristotelian
conception, only that the opposition in him appears as a dualism, which it is
not. The general propositions that science finds or sets up are not its last
goal. As we have seen in the investigation of the formation of judgment, we
bring the judgment into its most exact form, as an equation, only for the
reason that it thus becomes easier to use it in new connections. And all our
judgments are just as many attempts to determine and exhaust our objects, the
basis for all thinking (25; 35). There can here only be a matter of an
approximation. But the more predicates and judgments we find, the more exactly
our initial representations become determined. The insight into the general
laws and the use of the universal points of view is a means to the
understanding of the objects that appear with the stamp of totality and
individuality. Through the concept of direction, as we have seen, these
concepts hang together with the other categories. Individualities and
totalities we seek to understand as a ``samfoldighed,'' an interplay of
laws\footnote{The concept of ``samfoldighed'' (con-foldedness) was set up by
H. C. \emph{Ørsted}. See my book \emph{Danske Filosofer} p. 51.}. Even if
there continues here to subsist an irrational relation, we nonetheless seek to
find more and more decimals.

In the most recent time one has on this point found a decisive opposition
between natural science and history (cultural science). The former---one
holds---seeks only the general, laws that hold for phenomena and processes
that constantly repeat themselves. Nature is the home of the general laws. But
in history there is no repetition. Every event and every personality brings
something new, exists only once\footnote{\emph{H. Rickert}: \emph{Ueber die
Grenzen der Naturwissenschaftlichen Begriffsbildung}. (1901--1902).---See also
his treatise on ``Geschichtsphilosophie'' in the Festschrift for Kuno Fischer
(1904 f.).---Cf. my critical remarks on the occasion of this work in
``Göttingische gelehrte Anzeigen'' 1906. p. 5--11, and my lecture \emph{Tænkning
og Tro i vore Dage} (Tilskueren 1907).}. But this opposition is only a
difference of degree. Nature too presents wholes whose history is limited, and
which it is the task of natural science to understand, a task in whose service
the general laws ultimately step. For \emph{Newton} it was thus not merely a
task to find the laws for the movements of the planets---and ultimately of all
material elements; the solar system as a whole too, this compages mirabilis,
was a problem to him, even though he declared it insoluble without the help of
theology. \emph{Kant} and \emph{Laplace} later took up the problem by a purely
natural-scientific way, in that they applied the general natural laws to
understand the coming-to-be of the solar system as a whole. The biological
hypothesis of development seeks in an analogous way to understand the species.
Historical science builds, as criticism, on general scientific laws in order
to establish the reality and chronology of the individual facts; but in
continuation of this it seeks to work the facts together in such a way that
individual-concepts about events and personalities can be formed. In a similar
way, in psychological characterization the general psychological laws (for
representations, feelings, and so on) are applied to the understanding of
personal life in its individual forms. Often the rounding-off must here take
place by a certain artistic capacity. But it is one thing whether the goal can
be fully reached, another whether it does not constantly stand in the
background of all research. Provisionally science is most occupied with
finding general laws and establishing individual facts; but its last goal is
to write the biography of existence in the great and in the small.

It is, moreover, not only in life, especially in human life, that we meet the
individual, that which occurs only once (das Einmalige). Every state in nature
occurs, when we make sufficiently strict demands on identity, only once (91).
And the general laws are always due to sections made from a great complexity.
Thus the law of inertia always presupposes the relation to a definite system
within which the movement takes place. The application of the law of gravity
in a single case always presupposes a system of larger and smaller bodies,
and changes in it will entail changes in the gravitational relations. It is
always a factual, historical element that gives every application of a general
law its character.

96. Not only do different wholes stand over against one another, whose mutual
relation can present great problems; but a single whole can in turn comprise
several wholes in itself and itself be part of a more comprehensive whole.
Here merely the point of view of extent is taken, and the relativity of the
concept of whole then shows itself clearly here. In \emph{Ardigò}'s doctrine
of the undivided wholes (indistinto) and the divided-out parts (distinti)\footnote{\emph{Moderne
Filosofer}. p. 35--37.} this relation is well expressed, especially since the
divided-out and determinate stands, as regards the point of view of degree,
higher than the undivided.

There can always be found more comprehensive wholes than those we stop at in
the individual case. \emph{Kant} therefore calls the concept of an absolute
totality an ``Idea,'' that is, a concept that cannot be concluded. Thus ``the
world'' is a concept that cannot be concluded, like ``reality'' (38). However
high a totality stands, and however predominantly its states and changes are
determined by inner conditions, it will nonetheless always present dependence
on conditions that do not lie in itself. It will have its open side.
Reciprocal action with something different from it then stands as a necessity.
And if we ascribe to it energy as a whole, we thereby already presuppose that
there are other objects than it. An absolute system would have to be
unchangeable; if there were changes within it, they would nonetheless have to
run their course to the end, and equilibrium set in.

One often uses the word ``world'' in such a way that one does not distinguish
between the part of existence tolerably accessible to us and existence thought
as an absolute, concluded whole. In speculations about the future of the
world---especially in connection with the question whether there is impending
such a distribution of energy that a complete state of equilibrium must set
in---the two significations ought indeed to be kept apart from each other. For
thought, ``the world'' can only denote a relative whole, can never lie
ready-made. This follows already from the inconclusibility of the synthesis.
Thought can only end with a question.

And as every whole from one point of view or another will always be able to be
regarded as part, so every part can in turn be regarded as whole. The series
of parts and wholes (series (5)) is inversely symmetric (an absolute
correlate-series, see 68). Already herein lies the fact that absolute atoms
cannot be thought. There is just as little any part that cannot be whole as
there is any whole that cannot be part. It has then also shown itself that such
an assumption leads to contradiction; already \emph{Huygens} and \emph{Leibniz}
saw that it conflicted with the conservation of energy\footnote{\emph{Den nyere
Filosofis Historie}.\textsuperscript{2} I. p. 346.}, and in the most recent
time it has become more and more clear that the atom itself stands as a whole
little world within which new problems emerge. Nor in the psychic domain can we
stop at last, wholly uncomposed elements; even the simplest sensations show
themselves, when we take account of the conditions at their arising, to
presuppose a composite process (11).

\subsection*{c. Development}
\addcontentsline{toc}{subsection}{c. Development}

97. The concept of direction led us from causality to totality. A closer
reflection on the relation between totalities and their presuppositions leads
to the concept of development, which for that matter can also be reached
immediately from the concept of cause.

The concept of cause directs us, for the understanding of an object's
arising, partly to inner, partly to outer conditions. If we now keep to the
inner conditions, we are led to a comparison of the object, as it now lies
before us, with its own earlier states. The point is then to order these in
such a way that the following states---under the constant presupposition of
the cooperation of outer conditions---can be understood from the preceding.
The more connection we can thus find, the more we can apply the concept of
development, in that the series of the different states of the same object
appears with a stamp of unity. In other words, the more we can conceive the
causal relation as continuity, the more it steps forth as development. This
is already intimated by \emph{Kant}, but was especially emphasized by
\emph{Salomon Maimon} in the following utterance: ``To ask about a
phenomenon's cause is to ask about the continuous in it and to fill out the
gaps in our observations in order thereby to make them into experiences. For
what does one understand in natural science by the word cause other than a
phenomenon's development and dissolution, so that one finds the sought
continuity (Stetigkeit) between it and the preceding?'' ``The sufficient
ground of a thing is the complete concept of the way in which it has
arisen''\footnote{\emph{Versuch über die Transscendentalphilosophie}. (1790).
p. 140; 392.}.

There showed themselves, in the investigation of the concept of cause, two
points of view that led thought in different directions: equivalence and
continuity. If we are right that this latter concept is the characteristic
thing for the concrete causal relation, while the former concept always gets
a certain abstract, methodical character, then it is also justified to
maintain the inner connection between causality and development. We can then
not make the causal series into an absolute identity-series, but each member
in it gets, the more we keep to experience, its definite place (as in the
progressive difference-series).

Neither \emph{Aristotle} nor \emph{Kant} have given the concept of
development a place in their lists of categories. But both masters have
nonetheless, each in his own way, in another connection acknowledged the
significance of this concept.---Aristotle remains, in the doctrine of
categories, standing at the ``essence,'' the object of which the nine
categories can be asserted (53). But in his metaphysics he discusses more
closely the concept of essence and finds then that it appears in two stages,
as possibility and as actuality, no matter under which of the nine categories
it is to be considered. A certain property (quality) a being can have as
possibility or predisposition; the knower, e.g., does not always have his
knowledge present, actual;---but the property can also be there in full
actuality or ``energy.'' The transition from possibility to actuality
Aristotle calls movement (κίνησις), which is thus an analogy to the modern
transition from potential to actual energy. Only as actual (ἐνεργείᾳ) does a
being reach its full, pronounced form. Aristotle seeks to describe existence
as a series of transitions from possibility (matter) to actuality (form).
What hindered Aristotle from giving a doctrine of development was his
presupposition that the fully developed form was the truly real; the stages
by which it is reached do not become the object of serious investigation, but
were dismissed with the remark that they contained the ``possibilities'' of
the completed form. The concept of possibility, which, as we have seen, has
its significance as the marking of a problem, a future task (92), did not, in
Aristotle, and still less in his successors, get this methodical
significance, but all too often became a mere word that closed off further
questioning.---As for Kant, it would have lain near for him to give the
concept of development a place among his categories, since he conceives
continuity as a chief point of view that holds for them all (54). It is only
due to the artificial thing in his arrangement that this has not happened. In
special questions the concept of development played a great role for him, as
is seen from his cosmogonic hypothesis, his anthropology, and his historical
conception.

The concept of development is a form of thought that experience has forced
through. The urge to dwell on the intuition of the objects as they appear,
and to lay weight on their mutual difference, long hindered its full
acknowledgment. The opposition between ancient and modern thinking expresses
itself quite especially on this point. The ancient tendency to hold fast the
subsistence of forms, so that the transition from possibility to actuality
was more a kind of disclosure or unwinding than a proper development (in the
sense in which we now take the word), maintained itself very long, even in a
domain where experience could seem to speak sufficiently clearly. The
so-called preformation-doctrine (Leibniz, Haller, Bonnet) assumed that the
organic individual was already present in the germ (whether one, like
\emph{Leibniz}, laid most weight on the seed, or, with \emph{Haller} and
\emph{Bonnet}, on the egg), so that growth was only partly an increase in
size, partly an uncovering of the forms\footnote{\emph{Leibniz}:
\emph{Système nouveau de la nature}. (Op. ed. Erdmann) p. 125b: La génération
apparente de l'animal n'est qu'un développement [an unwinding] et une espèce
d'augmentation.}. No organs thus really come to be during the individual
development. In opposition to the preformation-doctrine, C. F. \emph{Wolff}
set the so-called epigenesis-theory (1759), in that he demonstrated how an
organ can develop out of a predisposition that has another form and structure
than itself. The intestinal canal, e.g., appears first as a leaf, passes then
over into a groove, and becomes finally a canal with openings. The difference
between the two theories has great methodological significance. It is only
the doctrine of epigenesis, the doctrine of the real, successive arising from
a simple predisposition through changing forms to the perfect state, that
contains a penetrating summons to go into the particular and trace the
different transitions. When it is only a matter of an expansion of what
already is, the occasion for research is not so great\footnote{Cf.
\emph{Kölliker}: \emph{Entwickelungsgeschichte des Menschen und der höheren
Tiere}. (1861). Historische Einleitung.}. The concept of development here
appears as an important means of thought, a forward-driving point of view.
Thereby---and at the same time by its natural connection with the other
categories---it is understood that its application has penetrated in all
domains of research.

Remarkably enough, at about the same time as the appearance of the doctrine
of epigenesis, an analogous point of view was asserted for the human
personality. \emph{Jean Jacques Rousseau} laid, in his pedagogical theory
(1762), great weight on the characteristic difference between the different
ages of life and protested especially against the child's being regarded
merely as a little adult. The qualitative difference between the ages of life
does not exclude continuity, but necessitates different psychological and
pedagogical points of view. Just as the grown organism is more composite and
heterogeneous in structure and mode of operation than the embryo, so it is
also as regards the later stages of the mental life in relation to the
earlier. And the same opposition Rousseau found in the domain of social life
in the relation between culture, with its ever more prominent division of
labor, and the uniform and involuntary form of life of the state of nature.

Already from these first forms of the modern concept of development it is
seen that there are certain analogous features that recur wherever this
concept is applied. There is no ground to say that it is \emph{only} due to
analogy when we speak of development both in the organic, the psychological,
and the social domain. There is everywhere ``only'' analogy between examples
of one and the same concept. (See 69; 73; 75.) The analogy is a relation of
likeness that attests its significance through fruitful application. And the
concept of development affords guiding thoughts that lead to raising
questions where one would otherwise close off. It brings with it the unrest
that a category ought to bring. Already C. F. \emph{Wolff} pointed out the
analogy that exists between the different systems within the organism (nervous
system, muscle-mass, vascular system, intestinal canal), as regards
development, and it was also by way of analogy that he discovered the
metamorphosis of plants, whereby he stands as \emph{Goethe}'s predecessor.
The struggle for the application of the concept of development to the
individual and social domain, as well as to the coming-to-be of our
world-system, belongs to the eighteenth century; in the nineteenth century
the struggle was waged for its application to the arising of the organic
species, as well as to the coming-to-be of the earth's present constitution.---For
the human race as a whole the thought of development had already come forth
within great popular religions like Parsism and Christianity. In Parsism the
idea of a world-history as a development toward a conclusion seems to have
arisen, and has from there gained influence on other popular
religions\footnote{Cf. my \emph{Religionsfilosofi} \S\ 14 and 43.---\emph{Goldziher}
has especially laid weight on Parsism's influence on Mohammedanism, in
eschatological and other respects. (Die orientalischen Religionen. Kultur der
Gegenwart. I, 3, 1. p. 108 f.)}---and has thereby in turn become an element in
the general human circle of representations. It is in the religious form
decidedly the epigenesis-type, not the preformation-type, that prevails in
the representation of development; for it is under a mass of changing forms
and fates that the race, according to the religious representations, appears
in the great drama from origin to world-judgment.

98. When the concept of development can find application in so many different
domains, the question becomes which form of it one is to lay at the basis
when it is to be set up as a category. It will depend on which form is best
suited as a basis for the illumination of the other domains through analogy.

Several philosophers in the nineteenth century have formulated concepts of
development. \emph{Hegel} used the analogy (which he would not admit was
analogy) with logical thinking. One could, in his opinion, build series in
which the first member led over to the second, and in which the third member
took the two first up into itself as in a higher unity, which could then in
turn become the first member in a new trinity (55). It is now, on closer
investigation, clear that Hegel in reality combines other analogies with the
purely logical. He thinks of psychological contrast-effects, cases where the
one opposition replaces or releases the other, and he thinks of the organic
and personal development, where the earlier stages' properties, however
conflicting they are, can all give their contribution to the stage of
completion\footnote{\emph{Den nyere Filosofis Historie}\textsuperscript{2}
II. p. 175 f.}. From a biological point of view \emph{Herbert Spencer}
started, in that v. \emph{Baer}'s theory of organic development as a
progressive differentiation, a successive transition from uniformity in
structure and mode of operation to heterogeneity, formed the starting-point
for his systematic philosophy of development\footnote{\emph{Den nyere
Filosofis Historie}\textsuperscript{2} II. p. 457; 471--473.}. A psychological
starting-point \emph{Sibbern} seized, who in the different domains of
experience, the physical, the biological, and the social, found again the
type of development that psychological experience and observation had first
taught him: processes that begin from different starting-points, hence
sporadically, in order then, through breaking and fermentation, to lead to a
harmony\footnote{\emph{Danske Filosofer}. p. 99--101.}. \emph{Roberto
Ardigò} too took a psychological starting-point, namely the natural
development of human thought-life from the indefinite to the
definite\footnote{\emph{Moderne Filosofer}. p. 33--37.}.---Under different
turns of phrase certain chief features recur in these presentations, chief
features that can be formulated in the following way, most nearly in
agreement with Spencer's presentation.

When an object comes to be, it must happen by elements that were formerly in
a more scattered state being gathered into a whole that rounds itself off with
a certain independence in relation to other objects. There must take place a
concentration; the sporadic must to a certain degree be overcome, in order
for an object to appear as something apart. The convergence of different
causal series was, as we saw (93), the necessary presupposition for the
arising of a whole.

The simplest objects do not get beyond this elementary form. A cloud in the
sky, a heap of sand on the beach are examples of this. And both planets,
organisms, personalities, and societies have, at the simplest stages at which
we know them, the character of a uniform whole with greater or lesser
concentration. If development advances further, differences form themselves
within the whole; differentiation replaces mere concentration. It was this
peculiarity that C. F. \emph{Wolff} brought out in his doctrine of
epigenesis. It is the stage of differences, of breakings and crises.

As the third chief stage, which denotes the culmination of development,
appears what Hegel called the higher unity (an expression that ought not to
disappear from philosophical usage because abuse has been made of it). Here
there has been formed a rich fullness of differences in essence and mode of
operation, without, however, the concentration, the character of wholeness,
suffering under it. The properties that set the two first stages in
opposition to each other are here united in a harmonious way.

We have, by this schematic presentation, which is a definition of the concept
of development, formed a series that cannot be reversed. Each stage is
conditioned by the preceding and conditions the following; from this side it
recalls the progressive difference-series (5);---but since the relation
between the first and second stage is different from that between the second
and third, it must be called a regularly varying difference-series (3).

How the schema is to be applied in the individual case can of course only be
decided by a special investigation. But the concept of development has here
gotten a definite significance, and it contains a direction to the direction
in which the investigation of special objects is to be undertaken, when they
appear as subjected to a series of changes. It is no necessity that there
shall be development; already earlier we saw the possibility that the
directions of the causal series did not converge. Of all real categories it
holds that it is experience that forces them through. And within the real
categories the category of development is in turn more special than causality
and totality. The inverse relation between content and extension announces
itself here again in all its fateful significance. Of the schematic
development's three stages, the third is in turn richer in content and less
comprehensive than the second, and this than the first: there will be the
greatest probability of finding development-processes of the most elementary
kind.

99. One could ask whether dissolution is not just as much a category as
development. And one would not be able to deny it. It is experience that has
led to combining our general forms of thought in such a way that a
development-schema can be formed. And it is also experience that impresses the
rhythm of development and dissolution. Formally we could very well use the
word development in such a way that, besides the three stages discussed above,
the stage of dissolution, or the stages of dissolution, formed a series (of
kind (3)). Dissolution presents stages that nonetheless do not quite
correspond to those of development: the harmony of concentration and
differentiation is abolished;---there sets in either a stable concentration
or a state of scatteredness. Whether these states are in turn suited to be
initial stages for new development, so that a rhythmic series can form
itself, is the last great question that the doctrine of categories cannot
discuss. Only two considerations are here to be asserted.

In the first place, we can connect no meaning with applying the concepts
development and dissolution to existence as a whole, for the good reason that
such a concluded whole cannot be thought (96). All development and all
dissolution takes place within a more comprehensive whole. \emph{A} world can
arise and pass away, but of ``the world'' these words could not be used if
they are to have meaning.

In the second place, one will never be able to demonstrate that dissolution
ends with absolutely different elements, in a chaotic difference-series. It
would take an infinitely comprehensive and infinitely exact experience to
convince us that we stood at such a series (62). The elements in which
dissolution ends must be small worlds. One speaks then also, in the most
recent time, of intra-atomic processes, and the electrons that are now
thought of as the material world's last parts are by no means absolutely
single. And sources of energy can emerge where one least expects it. Radium
was found in a heap of refuse.

100. Theoretically regarded---and we are here, after all, moving within pure
theory---dissolution is, as said, just as justified an object for reflection
as development is. It is also, in civil life, just as honorable to be employed
at the burial service as at the maternity hospital. When most thinkers have
involuntarily occupied themselves most with the thought of development, this
need not, however, be explained by the fact that our practical interests are
attached to the ascending lines. It need not be any interest different from
the thought-work itself that hereby makes itself felt. It can quite simply lie
in the fact that dissolution presupposes development, since it presupposes
wholes that can be dissolved, while development does not necessarily
presuppose dissolution, since the objects and elements that cooperate could
belong to existence's hitherto unused funds. And thought itself, the
reflection sprung up on the basis of sensation, intuiting, and involuntary
association, is itself ``a higher unity,'' a development in a third stage; it
is then natural that it is most occupied with tracing its own analoga.

There has, to be sure, among philosophers been an inclination to conceive all
development that reaches the third stage as eo ipso valuable. In \emph{Hegel}
and \emph{Spencer} this inclination shows itself distinctly. But thereby it is
overlooked that the valueless and the value-adverse can just as well develop
according to the indicated schema as the valuable. Development is in itself a
neutral concept---and dissolution therefore really also.

It has, however, now in the most recent time been expressly maintained that
the concept of development indicates a value-standpoint; only thereby is it
supposed to be distinguishable from the concept of change. If we will speak
purely theoretically, says \emph{Rickert}\footnote{\emph{Die Grenzen der
naturwissenschaftlischen Begriffsbildung}. p. 451--463.}, we ought to speak
of a series of changes. Even in the most elementary development-processes,
e.g. the formation of a cloud, we nonetheless presuppose a goal (Ziel) that
is reached, and thus introduce a teleological moment. The concept of mere
becoming (Genesis) leads us into the unlimited, but the concept of becoming
directed toward a definite goal (Orthogenesis) makes the changes into an
articulated and unified chain, and alone deserves the name development.

To this it must be remarked that while development in the neutral sense used
above is of great methodical significance, in that it contains a direction to
trace different stages in a series of changes, each stage denoted by its
characteristic marks, it is impossible to say when a ``goal'' has been
reached. It is arbitrary where we set the boundary, where the goal is to be
reached. When a boil develops, is there a goal? Or when moral wickedness
develops, is there then a movement toward a goal?

On the other hand, we cannot think at all without provisionally fixed points.
It shows itself with time and place, with rationality and causality; also in
the application of these categories a goal is presupposed. But we then also
become conscious that we more or less arbitrarily hold fast certain members in
the series as beginning- and end-points. Already the simplest formation of
judgment must have an initial representation and a final representation. And
this is sufficient to form an articulated and unified series. When the single
act of cognition is completed, one can pass over to another---if, that is,
thought itself is not placed in dissolution.

It must, however, be conceded that the categories whole and development have a
closer relation to the concept of value than causality and than any of the
formal categories. But it is appropriate to distinguish the value-standpoint
from other standpoints. Hitherto we have only presupposed intellectual values
and interests, because they are most intimately attached to the thought-work
itself. We now extend the consideration to values in general and discuss the
basis and significance of the value-concepts.

\begin{center}---\end{center}

\section*{E. Ideal Categories}
\addcontentsline{toc}{section}{E. Ideal Categories}
\label{sec:ideale-kategorier}

100. Our thinking treats its objects according to the forms that are the
natural ones for it, and that are developed by the relation to the objects.
But during this work an interest constantly makes itself felt that can stand
in a more or less intimate relation to thought itself. It can be a purely
outer motive, so that the thought-work itself is only a means, and it can be
most intimately attached to the thought-activity itself, following it during
its course, rising and falling according as thought succeeds in mastering a
rich fullness of objects, or as this striving is hindered by resistance or by
lack of objects. Such a purely intellectual interest we have in the foregoing
for simplicity's sake presupposed (6). Now we bring out the fact that there
are other interests and values than the intellectual, and investigate how far
it is possible, on the presupposition of them, to undertake a thought-work,
and under what special forms this must take place. In advance it is probable
that the forms of valuing thought must present likeness, at least analogy,
with the previously considered forms of thought.

The three concepts we here for clarity's sake distinguish between hang in
reality closely together. Both forms and objects have their values; this is
already presupposed in the foregoing. And value is itself an object that
thought can treat and try to form.

Value has everything that satisfies an urge and thereby awakens a feeling of
pleasure or averts a feeling of displeasure. The presupposition is thus an
urge, a striving or longing in a certain direction. The concept of direction
here again shows itself in its significance, and that in a significance not
wholly different from the one brought out earlier (93), which led us to the
concepts whole and development. An urge presupposes a whole that has a
tendency to assert itself, to assert its inner connection and its
independence in relation to everything else. Value then gets, from such a
whole's point of view, everything that favors this tendency, while what
hinders it becomes of the evil. Feelings of pleasure and displeasure are,
psychologically and biologically regarded, symptoms of progress or regress
for an individual whole's striving after subsistence and development. They
contain in themselves the germ of valuing judgments, in simplest form
exclamatory judgments (30). They appear as elementary forms of psychic
energy, in that they make known how far the whole that a conscious life or an
organism constantly constitutes is threatened with dissolution at one point
or another. It is an old thought that a being that constituted an absolute
unity would not be able to suffer: dissolution would then not be
possible\footnote{Melissus of Samos (\emph{Diels}:
\emph{Vorsokratiker}\textsuperscript{1} p. 150 f.).}. There is hereby
indicated a presupposition that alone can make coherent reflection about
values possible. For insofar as it holds, an investigation can be made of the
conditions for the subsistence and development of the whole in question, and
thereby the involuntary valuation can get its confirmation or its correction.
Every immediate feeling of pleasure or displeasure can from this point of
view be regarded as a hypothesis that is to stand its test. They are
immediate qualities like the sense-qualities. Value-qualities we can call
them. There is, as with all qualities, the question how far it is possible to
form rational series, so that grounded judgments and inferences can be set
up. Such a possibility can precisely only be found if the value-quality
corresponds to the relation between a given whole (an organism, a
personality, a society) and the conditions for its subsistence. It is really
such a possibility that was already discussed in the foregoing, in the
relation between form and object. The whole that was here to be built up was a
cognition, a thought-world, and the matter was what relation must be present
between thought and its objects, if such a world was to subsist. Concepts
like truth and reality are intellectual value-qualities that we have sought to
give rational form. (See especially 36--38; 85--88.) We now seek to treat
other value-qualities in a similar way.

Our category-series gets, as remarked several times, a narrower extension and
a richer content the further we come forward within it. Of values we can,
according to the given definition, speak only where a formation of a whole has
taken place. And within the wholes it is directly only the three named---the
organism, the personality, and society---that condition valuations. There can
be a biological, an individual, and a social valuation. Yet we can very well
by analogy apply a valuing train of thought to all wholes; we can ask about
the conditions for a mountain's, a planet's, a world-system's subsistence and
development. One will, e.g., be able to ask when a star stands at the
culmination of its development---whether it is when it has developed from
nebular mass to sun, or only when its temperature has reached so low a degree
that organic life can become possible? In the latter case the presupposition
will lie at the basis that it is life that gives everything else value. By
this way one comes to a kind of cosmic valuation, so that existence gets a
different stamp for us according as it shows itself to favor the development
of life or not.

The valuation, the value-quality, becomes another according to the whole one
lays at the basis. Within the same whole it can in turn be different parts or
properties whose fate especially determines the value-quality. For every
series of values there must therefore be laid a definite value at the basis,
determined by the whole, and within it in turn by the part or property whose
conditions of life are discussed. All valuation goes back to a fundamental
value. Thereby is demonstrated the justification for setting up the
value-categories as a peculiar group. Value can only be measured by value.
Here is a circle one cannot get out of. It is the fundamental value that
determines the standard. The special values are measured by their relation to
the fundamental value.

It goes here as with time and space: every time- and place-determination must
be undertaken from a definite standpoint. Time cannot be determined in
relation to the timeless, or place in relation to the non-spatial. Thus value
cannot be determined except in relation to other value. It becomes especially
clear when we here, as with the other categories, start from the category's
primitive, practical appearance. That something has value for me I show
practically, perhaps quite involuntarily, by holding fast or acquiring it,
doing a work for it. That something has more value for me than something else
I show by preferring it to the latter. The work is released only by a
difference, and in the concept ``prefer'' the difference is expressed still
more distinctly. We can continue the series downward: a value is preferred to
the indifferent (neutral), if there is any such, and still more to the
``negative'' values that make themselves known as displeasure and pain.

From the possibility of different, perhaps mutually conflicting fundamental
values springs, as we shall see in a later section, the problem of valuation.
In particular the religious problem springs from the relation between the
human values and the vast existence that does not gather itself for us into a
whole. We feel in our value-qualities the pulse of existence as it beats
within the wholes that we ourselves are or are parts of; but have we thereby
the right to infer a heart that beats in the whole of existence? Only by the
boldest analogies can our most content-rich and therefore least comprehensive
categories be extended to hold for the vast horizons that lead beyond the
limits of our experience. About the justification of these analogies stands
the great struggle between the views of life. But when this struggle is to be
waged with the weapons of thought, it must use the means of thought that the
categories can afford.

101. Provisionally we have merely to do with the categories themselves, not
with the problems. The value-qualities can be divided from a formal and from a
real point of view, and each of the groups thus formed in turn presents
differences.

\subsection*{1) Formal Differences}
\addcontentsline{toc}{subsection}{1) Formal Differences}

a) \emph{Elementary values} presuppose a sensuous present as object for
enjoyment and appropriation, as spur and as support. \emph{Ideal values}
presuppose memory or expectation, reflection or imagination. Since they are
not supported by sensuous presence, they presuppose greater psychic energy in
order to be held fast. They have at the same time the advantage of being
independent of outer conditions, so that they can be at one's disposal as soon
as the inner conditions are present. Thereby they stand above the elementary
values.

b) \emph{Immediate values} presuppose no grounding or derivation from other
values. They make themselves felt by their own power. The fundamental values
are from the first such immediate values, and precisely therefore it is
easily overlooked that they are determinative. Immediate values can be both
ideal and elementary; their peculiarity consists in their standing as
self-evident. Of \emph{mediate values} there is a question only when an
immediate value cannot without further ado, quite involuntarily, make itself
felt or fall to our lot. They are means, ways, preparations for immediate
values, must be derived from these and borrow their light from these. When
such preparation is necessary, hence when there takes place a transition from
the involuntary to the voluntary, the immediate value becomes a
\emph{purpose}. It does not thereby lose its immediacy (in the indicated
sense); when the purpose is attained, the value makes itself immediately felt,
does not necessarily point beyond itself. The immediate value is purpose only
while the interval is being traversed. By displacement, what at first had only
mediate value can become an immediate value; the means can become purpose, and
the preparation become the best thing. The hunt can be better than the catch.
It has been mentioned in the foregoing that intellectual value often arises in
this way, in that it develops out of practical interests that needed
reflection as a means. And within the intellectual interest the same process
can then repeat itself, so that trains of thought that at first had value only
as ways to the understanding of an object, later got independent value, formed
a little thought-world for themselves. Most energetically this is expressed in
Lessing's proposition that the constant striving after truth is to be
preferred to the possession of truth (86).

We see here again the significance of the interval. Already the ideal values
presuppose it, and the difference between immediate and mediate values could
not come forth without it. The immediate values themselves could not become
conscious to us without it.---As with the categories of time, place, and
cause, the relation of distance also plays, in the category of value, an
essential role for the development of thought. And it was precisely the
distance between urge and goal that conditioned the development of concepts
like time, place, and cause. The value-categories, which on account of their
more special character form the last group, have thus constantly cooperated in
the development-history of the preceding categories.

By the relation between immediate and mediate value the concept \emph{Norm} is
conditioned, which expresses a rule for the means and ways that are necessary
for the attainment of the purpose. All different kinds of values have their
norms, conditioned by the relation between the purposes and the experiential
conditions under which they are to be reached. The value of the norm is
mediate.

c) A third formal difference between values rests on the fact that a value can
need a release, a finding-out, or a producing, before it can appear as value,
while in other cases it appears in full activity. There can then be
distinguished between \emph{potential and actual values}, if one will,
slumbering and waking values. This difference can make itself felt both as
regards elementary and ideal, immediate and mediate values. There can arise
something new in the world of values, as in the world of objects in general.
There is the possibility of purposes no one has yet thought of, because the
immediate value in question has not yet been able to make itself felt. It can
be processes of displacement that are not yet concluded, or it can be natural
processes that have not yet unfolded. Mediate values too can be potential;
there can slumber, both in us and outside us, forces that we could take into
the service of our actual values, if we could bring them forth to the light.

\subsection*{2) Real Differences}
\addcontentsline{toc}{subsection}{2) Real Differences}

The formal differences mentioned in the foregoing can make themselves felt,
of whatever kind the values are. When we call them formal, it is not thereby
excluded that they condition decisive differences in the mental states. We
feel differently when we enjoy than when we wait, long, or work, differently
when we live on mere possibilities than when we move among actual values. So
far all value-differences are real. But real differences in the narrower
sense are those conditioned by the peculiarity of the objects to which the
values are attached---by that which ``has'' value. Here two groups can be
distinguished, of which the one has most to do with \emph{the content of the
values}, the other most with \emph{their extent}.

a) It can be the purely organic subsistence, mere self-preservation, that
conditions the value. It holds here literally and elementarily: to be or not
to be. It is the organic feeling of life that here lies at the basis, but to
it can in turn attach themselves ideal, mediate, and potential values;
especially the feeling of power, conditioned by the consciousness of dominion
over the conditions of life, stands near as a source of derived values that by
displacement can become immediate. In a certain opposition to this kind of
values---the \emph{biological}---the \emph{intellectual, aesthetic, and
ethical values} can step. The \emph{religious values} develop out of these
different kinds of values' relation to existence as a whole.

b) With respect to extent there holds a difference between \emph{individual,
social, and cosmic values}. Whatever content-values man knows, it makes a
difference whether it is only a matter to him that the values fall to his own
lot, in conscious or unconscious opposition to others' lot, or whether he sees
them as goods that are common to larger circles, and are for a part determined
in their valuation thereby. Between individual and social values there will be
a constant reciprocal action, now a harmonious, now a disharmonious one. From
this arise derived categories, and from this arise special problems.---Of
cosmic values there can be a question when we by way of analogy make a
valuation of the conditions of existence of the most comprehensive wholes.---

102. The possibility of an irrational, if not an anti-rational, relation
between categories lies nearer in the value-categories than in the other
classes of categories. Throughout, human consciousness has hitherto relied on
the assumption that there will constantly be application for value-categories
as well as for the other kinds of categories. Thought's last work begins when
there shows itself doubt about the forms of thought's mutual agreement as well
as about their applicability to the objects that make themselves known to us.
Such doubts lie at the basis of the problems of philosophy, to which we now
pass over. Naturally it is also a problem to describe the development-history
of thought and give a survey of its forms. But herein we have already
presupposed the validity of the thought-work, as it happens to be
constituted. It is this presupposition whose justification is now to be
tested.

\begin{center}---\end{center}

%% ============================================================
%%  IV. THE TASKS OF THOUGHT (PROBLEMS)
%% ============================================================
\chapter*{IV. The Tasks of Thought (Problems)}
\addcontentsline{toc}{chapter}{IV. The Tasks of Thought (Problems)}
\markboth{IV. The Tasks of Thought}{IV. The Tasks of Thought}
\label{ch:opgaver}

\begin{center}---\end{center}

\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}

103. The problem has in an earlier connection been discussed from the formal
side, as a form of psychic energy (27). Energy is required both to grasp and
hold fast the conflict between objects and to feel the lack of intermediate
members between objects---both for liberation-problems and for
completion-problems. It is only through memory and comparison that problems
come to be. As long as continuity prevails, it is not discovered that the
later states are different from the preceding; for that there is required a
holding-together of members that perhaps lie far from one another in the
series. The problem then becomes how continuity is compatible with the
qualitative differences. And when consciousness, where discontinuous objects
are present, lets go the one object when it passes over to the other, no lack
of intermediate members can be felt, no urge can arise to transform
discontinuity into continuity. To these two chief forms all problems can
directly or indirectly be traced back (59). There arises in both cases a
tension that sets a work going, until a solution is won, a provisional or a
definitive, an individual or a universally valid solution. Such relations of
tension will more easily set in the more experience is extended; but on the
other hand tensions can also fall away by extended experience. Problems
therefore do not, any more than the categories, continue to be the same at
all times. In a natural history of problems, such as \emph{Richard Avenarius}
has given, the special conditions and differences in the arising of problems
are treated. There arises, as he aptly remarks, a feeling of strangeness, of
lack of security, when a spiritual equilibrium is abolished; every genuine
problem is a kind of homesickness. But the home can be sought and found by
different ways. One can seek back to the old home, let the oppositions fall,
glide again along the stream of continuity, or cling to a single one of the
members in the discontinuity-relation and forget the others. If one gives up
comparison, the problem falls away. One returns to the state before the
doubt. But one can also press forward through the doubt and form a new
spiritual home, which can then be either a permanent dwelling or a hut for
provisional shelter.

The philosophical problems hang together with the nature of thought in such a
way that they must follow thinking on its way and will constantly be set
anew, even if it happens in new forms. The type for them is given with the
psychological antinomy that springs from the very concept of psychic energy
(4). Thought itself carries its enigma with it. It cannot leap over its own
shadow. Just as little as the personality can be thought fully concluded,
just as little can thinking. It goes from problem to problem. The urge for
scientific inquiry is a special form of the urge for agreement with oneself
under all manifold and changing experiences. It expresses itself both in the
formal science and in the real---both as an urge to form trains of thought in
which the one member with inner necessity develops out of the other, and as
an urge to connect different objects in such a way that as comprehensive and
as exact a continuity as possible is attained. Personality consists first and
foremost in the inner connection between all representations, feelings, and
endeavors. It is then no sacrifice that the personality brings when it stakes
its life on inquiry. It produces in science---as in art---an objective
counter-image of itself. Only thereby is understood the intimate character
that the joy of cognition can take on, the amor intellectualis that the
inquirer feels in working his way forward to a new standpoint from which the
single object can be seen as a member in a great connection. The holy fire
that constantly anew sets thought in motion, despite everything that can
dampen and hinder it, is understood only when one has this in view. And the
personality on its side needs to be purified by bowing to thought's objective
connection, the firm order of things, within which it itself has its definite
place. Freedom is won through truth. It belongs to the nobility of the
personality that it can by its own power find the great laws that condition
its subsistence and determine the wishes that arise in it, as well as the
conditions for these wishes' being able to be realized.

104. If it is rightly said above that the philosophical problems hang
together, more than other, more special problems, with thought's own nature,
then we must in the preceding presentation of thought's growth and forms have
a guide to finding what the philosophical problems are.

In every act of thought, at whatever stage of development it takes place, we
can distinguish three different elements. Thought's psychic energy expresses
itself in certain forms, which we can by abstraction distinguish from that to
which they are applied, what in the foregoing has been called the objects. The
concept object too is due to abstraction. No object is wholly unformed;
synthesis comprises it from its moment of birth in consciousness. Only
through analogy with an artist's relation to the material he works on, and
that we can know before he works on it, can we speak of objects wholly
independent of all form of thought or all forming by thought. Here we do not
know the marble block before the hewing has begun. To the object we reckon
everything that can stand as an object of reflection. Besides form and object
there always cooperate motives, interests, values. They can vary in kind and
degree, from the bitter need that requires reflection and inventiveness in
order to keep life up, to the purely intellectual urge whose satisfaction is
immediately given with the clear and grounded connection that reflection is
able to bring about between as many objects as possible. In the doctrine of
categories this triad appeared in the difference between the formal, the real,
and the ideal categories.

It showed itself in the doctrine of categories that the forms took on ever
more special character by reason of the objects that lay before us. Here lies
the first philosophical problem, determined by the relation between form and
object. Can the forms be derived from the objects, or conversely? With what
right are the forms applied to the objects, if they cannot be derived from
these? Is it possible to reach a perfect and exhaustive appropriation of the
objects by the help of thought's forms, as these happen to be?---Such
questions belong under \emph{the problem of cognition}. The third element, the
motive or the value, does indeed here too make itself felt. There could be
raised the question whether the choice of the objects that are treated, and of
the forms that are applied, is not due to a regard for value, a motive that
can come more or less to consciousness. This question too must be discussed
under the problem of cognition, even though it can otherwise be neutralized by
our provisionally merely presupposing the intellectual interest that is
immediately directed at clarity about the relation between forms and objects.

Another problem arises with the relation between the objects among
themselves. The different sciences have each their object in a certain group
of objects. But when the question is raised which object or group of objects
is to be regarded as fundamental, as the object in whose light we regard all
other objects, \emph{the problem of existence} (the metaphysical or
cosmological problem) is set up. It is the question how far a general
world-view is possible, and what character it must take on. According to the
group of objects that is laid at the basis, the world-view will get a
different timbre.---Here too the value-element will be able to make itself
felt; but it will be appropriate, as far as possible, to eliminate it, in
order that it may be seen to what results the purely intellectual interest can
here lead. It is necessary for clarity's sake to distinguish between the
different points of view.

If we then finally bring out the value-standpoint, which, according to the
investigation of the ideal categories, hangs closely together with the concept
of whole, it gives occasion to two problems that can be summed up as \emph{the
problem of valuation}. There can, in the first place, be asked about the
different values' mutual relation and about the possibility of tracing them
back to a fundamental value, so that the relation to this determines their
mutual sequence. This is the ethical problem. There can, in the second place,
be asked about the relation between the values on the one side and the
cognition that the forms and objects make possible. It is here a matter of the
relation between the world of values and that of reality. Is there a harmony
or a disharmony between these two worlds? Hereby arises the religious
problem.---Under this falls also the question whether it is possible to
develop a well-grounded world-view out of value-concepts, so that the validity
of the forms and objects is made dependent on their relation to the values.---

If one misses \emph{the aesthetic problem} in this arrangement, I will point
to its analogy with the ethical problem. Aesthetic value rests on an object's
appearing as a characteristic whole and being intuited and held fast merely as
such, apart from intellectual and practical interest. Aesthetically we regard
reality as image and the image as reality; it is enough for us when the
object---be it now reality itself or its image---rounds itself off into a
peculiar whole. The ugly and the disharmonious too can fall under this; the
aesthetic problem would then concern the condition for such images being able
to come to be and become the object of work and appropriation. In ethics, on
the other hand, it is a matter of the work for the real whole itself, which
stands as the fundamental value. Here the difference between image and reality
becomes of decisive significance. It is the category of whole that lies at the
basis of all valuation. We can conceive the whole as image (aesthetic
problem), as basis for the judging of actions and endeavors (ethical problem),
and as that whose fate determines our feeling of the value of life and of
existence (religious problem). For a complete presentation of the
philosophical problems it would be required that the aesthetic problem got a
peculiar place. But the typical in our relation to the problems will hopefully
be able to come forth, even if I keep to the problems I feel best prepared to
treat. For that matter I must refer to my Psychology (V B, 12; VI C, 9) and my
Ethics (XXX, 1--2).

105. That so definite a distinction can be made between the different problems
is due to a spiritual division of labor that has slowly developed. The
distinction steps forth more distinctly in Platonism than in Animism, and
more distinctly in Positivism than in Platonism, on the whole more distinctly
in modern than in ancient science. Cognition's forms and principles,
existence's innermost nature, the rank and fate of values---these were three
points of view that at earlier standpoints were regarded as one. Epistemology,
cosmology (metaphysics), and ethics went involuntarily together. The
difference was in part acknowledged in post-Platonic Greek philosophy,
especially in the division of philosophy into logic, physics (= metaphysics),
and ethics. But it was only the eighteenth century's empirical and critical
philosophy that carried it through. The romantic reaction in the first half of
the nineteenth century was able only transiently to press it back.

But the division of labor must not lead to a splitting. It would conflict
with thought's character as synthesis if the treatment of the different
problems should lead to mutually contradictory results, or to results that did
not fill out each other or show a mutual kinship. It would give occasion to a
new problem. An identical solution of all problems, in the sense that a single
principle contained the answer to the problem of cognition, of existence, and
of valuation, will be an insoluble task. But there must be assumed to be an
analogy between the different problems; this follows already from the fact
that they hold for one and the same working thought, whose enigmas are
determined by its nature, even if they do not exclusively have their origin in
this.

Both the problems themselves and their solutions are determined in kind and
direction by thought itself. Common to all problems is the great opposition
between continuity and discontinuity---between the unhindered advance of the
synthesis, of the psychic work, and its stoppage at continuities within which
no distinguishing is possible, and at differences that cannot be equalized.
Both in the world of thought, in existence as the aggregate of the objects we
know, and in the world of values, the point is to determine differences
within the continuity and to demonstrate transitions between discontinuous
objects. Of course continuity and discontinuity get a different character in
the different domains, just as there is, strictly speaking, on the whole only
analogy between the different examples of one and the same concept.---

The difference of the problems makes itself known in the history of thinking
by the fact that it is different motives that lead to philosophizing, and the
analogy makes itself known by the fact that one is led from the one problem to
the other. It can be an ethical or religious interest that leads to
philosophizing, and one then begins with the problem of valuation. But it then
shows itself that for its treatment one needs a conception of the nature and
limits of cognition, as well as of the existence within which the ethical and
religious ideals are to be realized. It leads in any case to an untenable
dogmatism when one, out of one's conviction about the rank and fate of values,
will give laws for cognition and construct a world-view. The division of labor
came forth precisely because thought, experience, and the values themselves
required the overthrow of dogmatism. Another motive for philosophizing can lie
in the urge to gather one's experience and one's knowledge into a
comprehensive world-view. One then goes straight to the problem of existence,
but will soon discover that the possibility and validity of a world-view
presuppose the validity of our cognition in general, and that its nature and
limits must be investigated. Likewise it will show itself to be necessary to
investigate the influence that our interest, our life of feeling and will, has
on the shaping of our world-view, and this will lead to a special treatment of
the problem of valuation. Finally, one can begin with the urge to distinguish
between what one knows and what one does not know, hence with the problem of
cognition, and from there one will naturally come over to the two other
problems, which each presuppose the former. It is no wonder that the problem
of cognition makes itself known more and more as the fundamental
problem.---The different philosophies get a different timbre according to the
problem that in them forms the starting-point and determines the points of
view.

Yet one more problem it will be justified to reckon among the philosophical
ones. All three problems named have a relation to psychology. And it is often
psychological interest, an urge for observation and investigation of the
mental life, that forms the starting-point for a philosophizing. One then goes
from the investigation of the psychic energy to the investigation of the tasks
it sets itself, and of its capacity to treat these tasks. The question whether
psychology is a part of philosophy, or the whole of philosophy, or perhaps not
at all a philosophical discipline, must, according to what has been developed
in the first section of our presentation, be answered thus, that every
philosophy needs a historical introduction, an investigation of the psychic
energy that makes itself known in the life of thought as well as in the life
of feeling and will. So far psychology is a part of philosophy. It leads us to
the point where the three problems spring forth. Psychology describes the
place and the presuppositions from which we orient ourselves in existence. The
special branches into which psychology divides, and the special methods it
applies to the individual objects\footnote{See on this my \emph{Psykologi}.
Kap. I.}, have their independence in relation to philosophy, and the concept
of the nature of the mental life that the latter lays at the basis must
constantly be brought into as great agreement as possible with them. The
subjective, descriptive, and analytic psychology forms an intermediate member
between philosophy and special psychology. Besides, psychology itself, as
regards presuppositions and methods, is, just like all other science, itself
the object of epistemological investigation. There is a constant reciprocal
action between epistemology and psychology: the forms that the former sets up
must be psychologically possible, and the conception of the nature of the
mental life that psychology maintains itself builds on the application of such
forms. It holds of the mental world as it holds of the world in general: man
is understood by the world; but all understanding of the world is determined
by the nature of man. It is no circle, for there can take place, and there in
fact takes place, a rhythmic alternation of the two points of view.---In
accordance with this consideration, the psychological problem has been
discussed as an introduction to our whole presentation.

\begin{center}---\end{center}

\section*{A. Cognition}
\addcontentsline{toc}{section}{A. Cognition}
\label{sec:erkendelse}

\subsection*{a. Forms and Principles}
\addcontentsline{toc}{subsection}{a. Forms and Principles}

106. The categories express the ways in which thought operates in reciprocal
relation with the objects. They acquaint us with our intellectual
organization. What we understand, and whether we understand anything at all,
depends not only on the constitution of the objects, but also on the nature
of our mind, just as it depends not only on the outer things, but on the
constitution of our organ of sight, which colors we see, and whether we see
colors at all. It was the urge and the capacity for unity and continuity that
denoted the pervading character of the forms of thought. This urge and
capacity can express themselves in other ways than such as lead to what we
nowadays in science call understanding. Of all the wealth of thought that the
human mind has at its disposal, only a narrow and firmly closed series of
thoughts can be used when it is a matter of such understanding. Here
Wallenstein's words hold: Eng ist die Welt, und das Gehirn ist weit. Through
Animism and Platonism the way in the history of thought leads to Positivism.
It is not traversed once and for all; often individuals must each for
themselves go through it anew, just as in embryonic development an analogy to
the development of the species shows itself. The psychological possibilities
are more comprehensive than the epistemological ones. What the understanding
of the objects requires must of course be psychologically possible, accord
with the general laws of conscious life; but the presuppositions of
understanding are of so special a kind that they can only be known through an
investigation of the very work of gaining understanding. Science's principles
and presuppositions express the intellectual habitus that poses the questions
and criticizes the answers in a stricter and more exact way than where the
involuntary intuiting and connection of representations prevails, and where
our wishes and drives rule without being subjected to the discipline of
experience and reflection.

To the different groups of categories correspond just as many different
groups of principles.---As a fundamental principle appears the demand that
\emph{every object shall be considered in as many relations as possible},
combined with as many other objects as possible, in order that it may become
as perfectly determined as possible. It is a demand or a guiding thought that
springs from the very nature and mode of operation of psychic energy; it
formulates a condition for perfect determination.---In a certain sense it is
derived from experience: we experience that only in this way can we cognize,
and that it is also really possible to cognize by this way. But it is an ideal
we set up, an anticipation of every possible cognition that is to be worthy of
its name. Purely factually regarded, it is a hypothesis that everything stands
in relations (or, with the ambiguous expression, that everything is relative).
And it is a hypothesis that has shown itself to hold, everywhere where we can
exactly test the conditions. We meet nothing relationless on our way. So far
we can here speak of a law.---To the formal categories corresponds \emph{the
principle of consistency}, which requires strict agreement between all the
thoughts we form about the objects. A subject in a judgment cannot have
predicates that contradict it; and an inference cannot conflict with the
premises. It is again here an instinctive tendency that lies at the basis: the
urge for agreement with ourselves, for not disowning ourselves when we pass
over to new thoughts or objects. Either the new must be brought into agreement
with the old, or the old with the new. Against this urge there can in fact
arise another urge: to be absorbed in the moment's incitements and connections
of representations, without these being tested for their agreement with what
other moments bring. The principle of consistency too (which is more closely
formed in the special logical principles) can be called a hypothesis, insofar
as we start from the assumption that all objects must be subject to its
demand. But it is a hypothesis we rely on to such a degree that where strict
inference from the objects at hand leads us to self-contradictory results, we
sooner assume our experience (the aggregate of our objects) to be incomplete
than we give up the principle of consistency.---To the real categories
corresponds \emph{the causal principle}, according to which, when an object
emerges, we look around for another object to which we can attach it in
analogy with the way in which we purely logically attach consequence to ground
(conclusion to premise). This principle too sets a demand that springs from
the nature of psychic energy. Of course the demand would long ago have fallen
away had it not to a certain extent and degree been honored by the objects'
mutual relations. The causal proposition can so far be called a hypothesis or
a law. But the chief point of view is and remains that it states a thought
under whose leadership we can, in the world of experience, satisfy our urge
for unity and continuity\footnote{More fully I have developed the conception
of the causal principle here intimated in my \emph{Psykologi}, V D., 1--4,
where I have especially shown that a complete experiential proof of it is
impossible.}. The two other real categories too have their corresponding
principles: as we pursue the directions of the causal series, there arises
\emph{the question of formations of wholes and of stages of development}, and
for a perfect understanding there is required an attempt to apply these points
of view, the most special of all, and the most hypothetical. That they are
nonetheless set up as principles is justified by the fact that every object
presents itself to us, already at its first emergence, as a whole, and the
understanding of it can then only be reached when it is investigated through
what stages this whole has come to be. In reality all cognition begins with
the question about this, at least as soon as the object is recognized and
classified.---As for the ideal categories, the value-categories, the
principles corresponding to them will show themselves to belong to the named
classes: relation, consistency, causality, and totality. Valuing reflection
has no other principles of cognition than understanding reflection. When a
value-thought leads to the setting-up of a purpose, I cannot, without breach
of the principles of relation and consistency, hence without coming into
conflict with myself, refrain from seeking and grasping the means that are
necessary for the attainment of the purpose. And the relation between means
and purpose is a causal relation, hence falls under the principles
corresponding to the real categories. That peculiar value-categories and a
peculiar problem of valuation are set up comes especially from the fact that
the relations that make themselves felt in valuation present special
properties, and that the difference between the different values among
themselves too requires special investigations.

107. That a principle is true means that one can work with it, and when the
talk is, as here, of the principles of cognition, it means that by their help
one can get forward in understanding, in firm and intimate ordering and
connection of the objects. Truth can be designated as a work-value, because it
comes to be for us only through spiritual work. When therefore principles are
designated as demands, these demands are directed not only at the objects, but
are also directed at ourselves. For if the principles of cognition really for
an essential part proceed from our nature, then it is our duty---the duty of
intellectual honesty---that we follow and apply them, as far as we are able.

Already the sense-qualities are working hypotheses (cf. 12). And still more
distinctly it is so with the fundamental propositions of reflection. But it
would nonetheless be a misunderstanding if one therefore regarded them as
wholly arbitrary. Every working-tool must correspond to the hand that is to
use it, and that has from the first produced it. And in a still more intimate
relation than the mechanical tools stand to the hand, thought's principles
stand to its own essence. One will become clear about this when one goes back
from the principles that have in fact guided and guide human inquiry to the
categories, and from these to the psychic energy at its different stages: it is
this way that we have here traversed, only in the opposite direction.

Truth can then not be defined as our thoughts' agreement with existence. We
know existence, after all, only through constant work at bringing the objects
under our forms of thought. Reality, the truth of the objects, consists
already for sound common sense practically in a close connection between as
many exactly apprehended objects as possible (cf. 36--38; 68).

With this it hangs together that we understand movement better than rest. Just
as both sensation and reflection must be awakened by a change, so it is through
the laws of changes that we come to know the nature of objects. Rest we
therefore conceive as stopped movement, as an equilibrium of tendencies. If
all our objects were in constant equilibrium, cognition could consist only in
recognition and ordering. Even if the objects did not move, thought would
nonetheless move back and forth between them---if, that is, it could be the
only movable element in a resting world.

There is no absolute opposition between thought and object. Every object that
emerges in consciousness shows itself already to have undergone an elaboration
that presents analogy with the one reflection can undertake (11). And on the
other hand thought can make itself its object; a later thought can have its
object in earlier thoughts, test and compare them. \emph{Kant} committed the
fateful error of assuming ``the matter'' (what I have called the objects) as
absolutely passively received. But the relation between activity and passivity
has infinitely many degrees. An absolutely passive, a mere matter, a pure
sensation cannot be demonstrated. It would, as \emph{Maimon} aptly
remarked\footnote{\emph{Den nyere Filosofis Historie}\textsuperscript{2} II.
p. 121; 123.}, be an ``Idea'' in the Kantian sense: we can, by abstraction
from the thought-work and from the reciprocal action of the sense-elements,
approach something such, but not reach it. Thus it would also require an
inconcludable thought-work (abstraction is, after all, a work) to reach a
chaotic difference-series (62), which would hold for the pure matter.
Everywhere in our cognition we have different elements operating together. We
can therefore never come to compare either our objects in themselves, or our
forms of thought in themselves, with an existence subsisting in and for
itself. And in any case such a comparison would itself be a thought-activity,
determined for an essential part by our thought's way of operating---so that
the question of ``agreement'' would constantly have to repeat itself.

The static concept of truth, according to which truth would consist in
cognition's agreement with existence, must therefore be given up---partly
because it is self-contradictory, partly because we can in our reflection
never reach a point where we have, on the one side, our own cognition with its
forms, on the other side the absolute objects. The only possible concept of
truth becomes the dynamic one, according to which truth makes itself known in
the firm and close connection between our objects brought about by
thought-work. The dynamic concept of truth is symbolic: even if we think of an
existence subsisting in itself, cognition's relation to it can, as regards
neither objects nor forms, be congruence or qualitative likeness, but can only
be analogy.

108. So far is such a concept of truth from leading to skepticism that
precisely the opposite holds. Since we can never have ``existence itself''
before us as an object, so that we could compare it with our thoughts and with
``other'' objects, there will, according to the static concept of truth,
constantly be doubt possible about the validity of our cognition, and a doubt
that could never ever be corrected. According to the dynamic concept of truth,
on the other hand, we are at once set to work---to find new objects, and to
find closer connection between the objects we know. And this work is precisely
of the kind that psychic energy exercises. It is then no foreign norm we bow
to.

The dynamic concept of truth is a consequence of the critical philosophy, for
which the truth of our fundamental thoughts consisted in their making strict
experience possible. At the same time this philosophy found the fundamental
presupposition of cognition by analysis of such an experience. A comparison
between cognition and an absolute order of things is impossible; we can only
compare thoughts and objects among themselves. This consequence \emph{Kant}
himself did not see as clearly as some of his disciples (Maimon, Fichte,
Fries). The master himself was still caught in dogmatism's static concept of
truth, as is seen from his doctrine of the ``Ding an sich'' as absolutely
existing apart from all cognition. But when he designates the causal law as an
``analogy of experience'' and thereby means that the events in time relate
analogously as ground and consequence in our thinking, he approaches
conceiving it as a working hypothesis determined by the nature of our thought.

Natural science embraced, right up to the most recent time, a static concept
of truth. It shows itself especially in Newton's doctrine of space and time,
in the physicists' and chemists' atomic theory, and in the natural historians'
species-concepts. Just as Pythagoras projected the numbers, Plato the Ideas,
Spinoza the Substance, so space, time, and atoms were projected as the
expression of existence's absolute essence, although all these concepts are
formed during the thought-work and for the use of it. It was from such
standpoints a defect that \emph{we} could not reach the pure world of the
numbers, the Ideas, or the atoms, or the absolute Substance or the absolute
time and the absolute space. The critical philosophy, conversely, regards all
such concepts as means and forms that come to our aid during our work at
reaching the perfect connection we call reality. It is not ``reality'' that
has fallen away from the pure world of ``Ideas''; it is the Ideas that,
through the thought-work they call forth, lead us forward toward more perfect
concepts of reality.

109. There shows itself, however, at every stage a difference between
something that is taken as given and the work we more or less consciously
exercise on it. The dynamic concept of truth does not efface the difference
between object and form. On the contrary, it impresses that the way in which
the objects have hitherto been elaborated in science cannot be proved to be
the only possible one. It has been possible to set up principles and
hypotheses from which the objects can be brought into firm and close
connection; but we cannot prove that the principles and hypotheses we set up
are the only possible ones, the only ways by which such a result could be
reached. The success that has followed science is no proof of its full truth,
as long as it cannot be proved that no other presuppositions could have led to
the same result. There is thus no absolute relation of identity, only a partial
relation of identity (67; 69) present between the connection of the objects and
the presuppositions from which it is derived. Ground and consequence cannot
here change places.

Already \emph{Leibniz} remarked this, without, however, letting it get a
decisive influence on his doctrine of cognition. A hypothesis's perfect
confirmation, he states, is reached only when one can from the conclusion
return to the premise: cela donne \emph{un parfait retour}, suffisant à
démontrer la vérité de l'hypothèse. Mathematically, he remarks further, this
is often possible, so that synthesis and analysis correspond to each other
\emph{à rebours}; but in the astronomical or physical hypotheses no such
return can take place. Success does not prove the truth of the hypothesis. True
conclusions can be drawn from false premises. It is, however, probable that
the premises are correct when they are simple, and when from them one can
ground many truths.\footnote{\emph{Leibniz}: \emph{Nouveaux Essais}. IV., 12,
1; 17, 5. (Opera phil. ed. Erdmann p. 381 a; 397 b.) Cf. already
\emph{Hobbes}: \emph{De corpore} XXV, I.}

This is a consideration that is of great significance for seeing human science
in the right light. Kant brought it out in a more general form than Leibniz (in
the utterance cited), in that he emphasized that all our cognition is
determined by the human organization. Think of another organization---and
another world-picture might then be the consequence!---When Leibniz's monads
each apprehended the universe ``suivant son point de vue,'' Kant's theory did
not lie far off.---

110. What in the individual case is designated as object depends on the point
of view. A sense-, memory-, or imagination-intuition can be regarded as object
for reflection, while from another point of view it is itself formed on the
basis of still simpler objects. The difference always rests on the fact that
the moment of activity is more pronounced in the form than in the object. What
cannot be derived from the very nature of our thought is present in higher
degree in the object than in the form, at whatever stage we seek out this
opposition, which never quite vanishes. Whatever ground this underivable-from-our-own-activity
may have, it nonetheless testifies that there subsists something \emph{besides}
us. This is what we mean when we speak of existence as more comprehensive than
thought. We must here reverse Wallenstein's words and say: Gross ist die Welt,
und das Gehirn ist klein. When we distinguish between cognition and existence,
it is because our cognition is always an open system. New objects can always
emerge that can be derived neither from earlier objects nor from our forms. It
is by reason of this constant possibility that we speak of an existence of
which we ourselves are members. Both our objects and our forms point out to it
when we ask about their origin. Our forms can, after all, themselves be
regarded as objects.

Only most indirectly does existence make itself known to us, namely through
the work that the emerging objects compel us to undertake. It is not only the
objects, it is also the forms that must have their ground in a larger
connection.

\emph{Kant}, who otherwise so energetically holds fast the opposition between
our cognition and ``the thing in itself,'' and who as a rule merely lets ``the
matter'' spring from this, nonetheless intimates in several places---what for
the most part is not heeded---that our ``forms'' too have their ground in
``Ding an sich.'' If one keeps to these intimations, it shows itself, as I have
shown elsewhere\footnote{\emph{Om Kontinuiteten i Kants filosofiske
Udviklingsgang}. (Vid. Selsk. Skr. 6. Række.) p. 33--35.---To the places there
cited may still be added \emph{Kritik der reinen Vernunft}.\textsuperscript{2}
p. 145 f and letter to Reinhold, 12 May 1789.---Cf. with respect to the
question itself my \emph{Mindre Arbejder}. II. p. 13 f.}, that Kant comes to
stand nearer to the philosophy of experience than he himself thinks. For his
epistemology comes, when the forms too, the constant in our cognition, are to
spring from ``Ding an sich,'' to rest on the presupposition that the unknown
basis that contains the cause of the forms of our cognition (of that which
makes existence intelligible to us) operates in a constant and law-governed
way. The whole system of categories and fundamental propositions rests on this
presupposition, which Kant has not brought out, but which makes his whole
philosophy hypothetical and deprives it of the strict rationalistic character
Kant has given it. He flattered himself that nothing hypothetical was to be
found in his doctrine. Such a thing is found, as mentioned in an earlier
connection, in the psychological basis for his derivation of the categories;
but it is also found in his intimations about the last origin of the forms.
And when ``matter'' and ``form'' can have a common origin, there shows itself
the possibility of a greater harmony between cognition and existence than Kant
assumed---while at the same time the a priori, the absolutely
experience-independent in our cognition, is limited to a high degree. When we
apply our forms of thought to understand existence, it is, after all, in a
certain way existence that thinks itself in us---to use \emph{Spinoza}'s and
\emph{Hegel}'s mystical-speculative language. Thought is a part of existence
that, from its presuppositions, seeks to form an image of the existence in
which it is a member.

That there is an open place in our cognition---that this is not and can never
become a closed system---in this Kant was right, and at bottom far more right
than he was conscious of. He was also right that this changes not the least in
our thinking's procedure. We must follow the ways to understanding that are
pointed out to us by the nature of our mind and our position in existence. For
\emph{Copernicus}, with whom Kant liked to compare himself, it was possible to
sketch an image of how the phenomena we regard from the earth's standpoint
appear from a wholly different standpoint. But something corresponding is not
possible for the philosopher. The philosopher can become clear that our
thought's forms and principles cannot be proved to be the only possible ones;
but he cannot construct a cognition from a standpoint other than the human. Fortunately
there is also enough to do of spiritual work from this one.

111. During the constant reciprocal action and breaking between object and
form, the third element, the motive or the value, always operates along with
them. The attention to, or the discovery of, an object is often due to a
definite motive, and the same holds of the way the object is treated. Such
motives can come from the organic urge for self-preservation, or from the urge
for hope or for consolation. There must in any case be nothing in the whole
feeling-state and sphere of interest that hinders the acknowledgment of the
object and its methodical treatment. For there can be a relation of opposition
between the three elements that leads to unrest and strife, not only in the
race or society, but in the individual's own interior, so that the ``head''
leads him one way, the ``heart'' another. But objects could also emerge with so
immediate a power, and thought's unfolding through the forms that are natural
to it can take place so involuntarily, that all misgivings from the side of the
third element are blown away. In this way objects and forms react back on
motives and interests. Of significance especially is, as stated several times
above, that a purely intellectual interest is formed that is immediately
attached to the life of thought. The struggle between ``head'' and ``heart''
stands in reality between two interests. As already stated, it is to the good
fortune of all sides of the mental life that thought gets self-governance---that
it is not regarded merely as a means, still less as an enemy. Only then is
there tolerable prospect of getting all objects brought out, and all forms
applied, that are necessary for gaining understanding, e.g. for applying the
criterion of reality in an individual case. There will always be more to
discover both in the world of objects and in the pure world of thought than any
special practical motive can lead one to bring out. It is not only philosophy or
other science that could need the reminder that there is more in heaven and on
earth than it thinks; it is also, and surely to a far higher degree, sound
common sense and the faith built on tradition and inner experiences. Existence
is more comprehensive, and its connection far more intimate, than practical
motives lead one to assume. The greatest enigmas for the view of life arise
precisely at the great horizons that go far beyond the sphere of human
life-interests, although thought, guided by intellectual interest, is able to
discover them and seek to measure them out.

112. The differences between epistemological views are determined by the
different relation between form, object, and motive that they maintain. It is
the irrational in this relation that conditions the different directions in the
philosophy of cognition. Could object, form, and motive be derived from one
another, or did merely a definite character of one of these elements
correspond to a definite character of the others, the mutual divergences would
not be so great.---The different epistemological views can be traced back to
three groups, of which the one ascribes to the form, the second to the object,
the third to the value (the motive) the decisive significance.

I take, in the following discussion, account only of modern theories. The
problem of cognition is essentially a modern problem, although it of course
already made itself felt among the Greeks. For \emph{Plato} the great enigma
was how the Ideas, despite their opposition to the phenomena, could nonetheless
be called forth in us by these. \emph{Aristotle} required a special organ,
``reason'' (νοῦς), for the apprehension of the principles (ἀρχαί) that lie at
the basis of all proof and therefore cannot themselves be proved (Analyt. post.
II, 19). \emph{Grote}, with whom \emph{Gomperz} agrees, holds that ``reason''
for Aristotle only completes and concludes the induction that is to lead up to
the principles, but he concedes that Aristotle emphasizes the opposition
between induction and ``reason'' in a way that recalls Plato\footnote{\emph{Aristotle}.
II. p. 237 f; 275; 375.}. Rightly, no doubt, \emph{Teichmüller}\footnote{\emph{Studien
zur Geschichte der Begriffe} p. 409--426.} lays the chief weight on the
agreement between Plato and Aristotle on this point. Induction is, he holds,
for Aristotle only a means to grasp the principles; but it does not prove them;
it only gives occasion for reason, which stems from a wholly different source,
to draw out the general propositions. In the two great thinkers of antiquity,
then, the irrational steps forth clearly. It was this that called forth the
interesting skeptical directions in antiquity's later time.

The modern epistemological directions we can group under the three names
Formalism, Empiricism, and Pragmatism.---

α. It is \emph{Kant}'s great merit to have seen that one must first take stock
of which forms scientific inquiry has at its disposal and in fact applies,
before one can decide how they relate to the objects. It is in his footsteps
that I have developed the doctrine of categories, before I passed over to the
treatment of the proper problem of cognition. Kant is right that these forms
must be determined by the nature of our faculty of cognition, as this happens
to be. But in his attempt to give proofs for the fundamental propositions that
spring from the forms, the leaning toward dogmatizing, which despite all his
criticism he did not free himself from, comes clearly forth. He holds that the
conditions without which we cannot have strict experience can without further
ado be formulated as necessary propositions, and overlooks the great question
in what extent and to what degree we really have such strict experience. His
``proof'' for the causal proposition, e.g., presupposes that the objects are
given completely, and that their essential properties will continue to be the
same. But this is an unjustified assumption, since experience is never
concluded, and new objects can constantly emerge. Furthermore Kant presupposed
that the basis or source of all objects and forms operates constantly, a
presupposition whose necessity likewise cannot be demonstrated (see 110). It is
no accident that his nearest successor struck out in a speculative direction;
Kant's rationalism on the one side, and on the other side the danger of
empiricism that certain points in his philosophy could contain, had to lead to
attempts to let thought and its forms become sole ruler. ``Reason does not beg,
it commands''---this proud word of Kant's united with the age's religious and
poetic currents into the peculiar phenomenon in the history of philosophy that
German romantic speculation is. Romanticism's philosophy issued commands
according to which all objects had to conform. One overlooked that all
commanders are nonetheless at least compelled to direct their commands
according to the capacity to obey that can be assumed to be present.---

It is by way of analysis that the critical philosopher seeks to find thought's
forms and the fundamental propositions corresponding to them. But since one can
never know whether the analysis is complete, and since it in any case must
always stop at a definite point, there becomes something arbitrary in at which
point one sets the beginning of one's philosophizing. It is then a choice, an
act, that concludes the analysis and proceeds to the setting-up of the
principles. This point of view is with the greatest energy asserted by
\emph{Thomas Hobbes}. ``It is,'' he says, ``we ourselves who make it that the
first principles of reason are true (rationis prima principia vera esse facimus
nosmet ipsi). The first principles are not objects of science, but of art or
construction.'' Yet Hobbes concedes that the principles we thus establish we
find only by analysis, and he sets up as a special demand that they must not
contradict each other\footnote{Computatio sive Logica (1655) III, 8--9.}. A
similar position on this point was taken by \emph{Fichte}: the first and
unconditioned fundamental proposition of human knowledge---the principle of
identity---is set up by a free act. Our science rests not on a ``Thatsache,''
but on a ``Thathandlung,'' namely the thinking consciousness's resolve
constantly to remain in agreement with itself. Fichte too concedes that a work
of reflection and abstraction (analysis) precedes that act, and that when
different fundamental propositions are set up alongside one another, it must be
possible to unite them\footnote{\emph{Grundlage der gesammten
Wissenschaftslehre} \S\ 1--3.}. \emph{Kierkegaard} saw no other means of
getting reflection stopped, so that a systematic development can begin, than a
resolve, a leap. The beginning of philosophy cannot be rationalized, and
therefore there is no presuppositionless philosophy\footnote{\emph{Uvidenskabelig
Efterskrift} (Saml. Værker VII) p. 84.}. While the three named thinkers
expressly acknowledge that the act of will by which the principles are
established presupposes an analysis (reflection, abstraction) by which it first
becomes clear which principles must be established, \emph{Kroman} sets up
thinking's principles by a pure postulate, a postulate that is, however,
motivated by the fear of self-dissolution through the assumption of
contradictory assertions and by the hope of finding stability in the course of
things\footnote{\emph{Vor Naturerkendelse}. p. 270--274. By the emphasis on the
motive Kroman approaches Pragmatism.}.

What is not sufficiently emphasized by the kind of critical philosophy that one
could call the arbitrariness-theory is that there can in each individual case
be a rational ground for stopping reflection and beginning to apply what it has
found. For when our thinking has a definite task, e.g. to find the basis for a
cognition's validity, we stop reflection when we have found what is necessary
for solving such a task. It is thus no purely arbitrary act. And in Fichte and
Kroman it is moreover expressly conceded that in the first principles of
cognition our own, the human nature must lay itself bare, since it is the
endeavor to preserve the unity of our being that is decisive. It is the
fundamental law of synthesis that is acknowledged. But this fundamental law is
itself discovered by reflection on our mental nature, as this in fact shows
itself to be; and likewise it will be impossible to find the principles for the
understanding of the objects if we do not investigate these objects' factual
constitution. Nor can this form of criticism, then, free itself from a relation
to experience.

β. Empiricism, which, when it takes account not merely of the individual's but
of the whole race's experiences, becomes evolutionism, regards the objects as
sufficient ground for cognition and especially for the setting-up of the first
fundamental propositions. General principles can have arisen through the
constant action of the objects' constitution. The objects deposit their general
properties in the individual's or the race's nature, or rather in their habits.
And then it is a matter of course that no principle has validity beyond the
circle of objects that has imprinted it. Only as results do the principles have
significance, while for Kant they were grounded in the nature of consciousness,
and for Hobbes and his successor were essentially postulates. \emph{Hume} and
\emph{Stuart Mill} have developed empiricism in its most radical form, in which
all objects appear in consciousness as constituting an incoherent chaos, so
that they enter into such firm connections that can operate as principles only
through external association and habit. \emph{Herbert Spencer} extended this way
of considering from the individual to the race by virtue of the theory of
development; those firm connections could, according to him, not form themselves
in the course of the individual's life, but only in the course of generations.

Empiricism overlooks that we know the objects only in constant reciprocal
action with the forms, just as, conversely, we know these only in constant
reciprocal action with the former. Thereby it is understood that there is a
constant moment of expectation, anticipation, seeking, and questioning in our
cognition. One cannot demonstrate any absolute passivity in any living being.
There would, as we have seen in the discussion of the chaotic
difference-series, be needed an infinite analysis if its correctness were to be
demonstrated. At every stage of psychic development we know, there are always
certain inner conditions present, on which it depends both how the objects that
are accessible appear in consciousness and how they are elaborated. This holds
for each individual and each generation.

γ. Pragmatism lays the chief weight on the fact that the objects determine the
forms' application and specialization. Cognition is a constant adaptation; the
point is, as \emph{Avenarius} has expressed it, to think the objects with as
little expenditure of force as possible. Here, then, a psychic energy is
presupposed, without its being more closely investigated whether it does not by
its nature operate in certain definite ways that can in the individual cases be
formed according to the objects' peculiarity. One keeps, in pragmatism, to the
fact that we form the concepts and principles we need, and no others, and
therefore we also seek to limit the formation of concepts and principles to the
least possible. The principle of simplicity, which has played so great a role in
the setting-up of new points of view and hypotheses, is derived from the economy
of consciousness. We wish, through cognition, to make ourselves master of
things, but with as few costs as possible. \emph{Ernst Mach} has, on the basis
of the history of natural science, developed this theory. \emph{Avenarius} has
developed it in connection with an investigation of the nature and history of
problems in general. From a somewhat different side pragmatism is developed by
\emph{William James}, who especially emphasizes that we leave only for practical
reasons the involuntary continuous stream that our mental life from the first
forms. (Cf. 3.) We set up, for practical reasons, definitions and fundamental
propositions, and the weight must be laid on the motives, not on the forms.
While Mach and Avenarius keep mostly to the theoretical motives, James and his
successors emphasize mostly the practical motives, although they protest against
being supposed to overlook the theoretical interests.

The chief question in the face of pragmatism becomes whether cognition's form
and motive can be distinguished in as external a way as it holds, and in such a
way that the forms are essentially means. Whatever motives may guide us, they
can nonetheless accomplish nothing when the capacities and forces are not there.
There can never have been a stage of development where the motives operated
independently of the forms. In the involuntary and immediate life, which James
has so masterfully described, there is no difference between motive, form, and
object; they operate together in an inseparable way, as in instinct. If it is an
abstraction to isolate the object, as empiricism does, or the form, as criticism
does, so it is also an abstraction to isolate the motive, as pragmatism does. In
instinct an object at once releases an urge and an operation. Urge and capacity
are here one. And they are so much one that there can be an urge to use the
capacities or forces for their own sake, apart from every special motive. The
intellectual interest, which we have in the foregoing so often mentioned, is an
example of this kind of operation; yet it can also, by displacement of motive,
develop out of other interests.

The truth in pragmatism consists in something that the preceding presentation of
the doctrine of categories has presumably let come into its full right, namely
that the special development of the forms is determined by the tasks set to
thought. Therefore a distinction was made between fundamental categories on the
one side, formal, real, and ideal categories on the other side. But already the
fact that tasks can be set, and that they are set in a certain definite way,
presupposes that besides urge and motive there is also the possibility of
definite modes of operation present.

The economy according to which the history of thinking operates is not merely
the fruit of a peculiar urge for thrift. It is understood not merely by the
motive, but also by the form of thought's operation, as this lays itself bare
in all domains. The principle of simplicity and the urge for unity has its last
ground in the essence of consciousness as synthesis. The combined is always
simpler than the unconnected, the chaotic. All thinking is so far in its nature
economical, and biologically regarded one will only thereby be able to
understand that thinking beings have been able to subsist in the struggle for
existence. Thereby the external relation between form and motive falls away in
the last instance. The intellectual feeling of displeasure, or perhaps even
feeling of pain, that can arise from lack of connection or from
self-contradiction and doubt, is surely due not merely to a feeling of
impotence or vexation over waste of force, but is ultimately to be traced back
to the fact that the very innermost essence of consciousness is
violated\footnote{The principle of simplicity has been conceived now as a
maxim of thrift (frustra fit per plura quod fieri potest per pauciora): thus in
\emph{Will. Occam} (in the 14th century), who appeals to it,---now as a
metaphysical proposition, holding for existence; \emph{Copernicus} and
\emph{Newton} were thus convinced that nature followed the simplest ways.
\emph{Kant} found it, on the other hand, grounded in the very nature of our
cognition and found it operative every time we form the simplest judgment.
Already long before ``Kritik der reinen Vernunft'' he explained in this way
``favor ille unitatis, philosophico ingenio proprius, a quo pervulgatus iste
canon profluxit: principia non esse temere multiplicanda præter summam
necessitatem'' (\emph{De mundi sensibilis atque intelligibilis forma et
principiis}, \S\ 30.) Cf. \emph{Kritik der reinen Vernunft}.\textsuperscript{2}
p. 680.}.

Only through work do the forms of thought develop in the particular. And this
work will succeed best when the intellectual feeling, determined by the urge for
unity, continuity, and connection, is the decisive motive. Only from the moment
such an interest is awakened does thought exist as an independent mental power,
becomes more than a means. And even where it is to operate as a means for other
interests than the intellectual, it will operate best when this latter
provisionally takes the helm. Resignation is required in order to cognize, and
the straight way is not always the best. In epistemology we in any case
presuppose a stage where the intellectual interest prevails. Every task
requires, after all, not only certain forces suited to its solution, but also
motives that are as closely attached to the task itself as possible. If there
are then other motives and interests that demand satisfaction, they must try to
find this under the conditions brought about by the solution of that task. The
spiritual division of labor cannot be abolished, however many difficulties it
brings with it. There must be sought other ways to remedy these difficulties
than that of making the thought-work dependent on principles other than its own.

113. Motive, form, and object operate in every act of cognition together with a
timbre that can vary without any of the three elements ever quite vanishing.
The difference between what one calls theoretical and practical thinking is only
a difference in the timbre, respectively in the relation between the three
elements by which this is determined. And thus too the difference between the
three epistemological directions discussed is a difference in the timbre; more
than that it should in any case not be.

But there is one moment at which criticism will always retain a prominent place
in the epistemological discussion. It is the weight it lays on the analytic or
regressive method of finding cognition's forms and principles. An investigation
of this kind is, as shown in the foregoing, the presupposition both for
dogmatism, arbitrariness-theory (postulate-theory), empiricism, and pragmatism,
because it concerns \emph{what} is to be set up as dogma, as postulate, as
experience, or as good economy. Here is something every epistemology must
provide. And here is something that can be separated out from the historical
forms in which criticism has from the first appeared---both from Kant's
dogmatizing rationalism and from the postulate-theory's emphasis on the leap.
Thereby it will be able to let both motive and object come into the right they
really have.

The concept of working hypothesis can very well accord with this consideration,
when one lays weight on the fact that the way in which human thought works must
be determined by its nature. And its nature we come to know by the study of the
involuntary mental life, as psychology by the help of its different methods
investigates it, as well as by the study of the history of thinking and of
science. In the latter respect it shows itself that there is a certain type
that constantly appears in the scientific principles and hypotheses, indeed even
in mythological representations, a type that leads back to an urge for unity and
continuity. All investigation of human thought is an investigation of the nature
of this urge and its struggles to gain satisfaction.---Also in the more special
assumptions that one as a rule first thinks of at the word working hypothesis,
that stamp will be found. Such assumptions are merely more limited and adapted
forms of the general type.

\subsection*{b. Forms and Objects}
\addcontentsline{toc}{subsection}{b. Forms and Objects}

114. We now go more closely into investigating the irrational relation
between form and object that has already provisionally shown itself to be
present.

Our highest, most transparent principle of thought, the principle of
principles, the principle of identity, which makes it possible to put one
concept in place of another and thereby abolish the differences that the
objects present, meets resistance where we must stop at a plurality of
concepts that can perhaps indeed be ordered in series, but not in such as
make substitution possible. It was a romantic ideal that required all
cognition to be derived from a single principle, which thus would have to be
applicable at every single point in the series of objects and at the objects'
transition into one another. The world, the world in which we are to think,
is and remains a multiplicity, not a world of resting identities. And on this
it rests that it is a living world. Nevertheless the principle of identity
will constantly stand as a direction to work, namely to bring about as
favorable conditions for substitution as possible---to find ways by which one
can in each case conceive the differences as disguised identities.

The points of view from which different objects can thus be regarded as
identical cannot, as stated several times, exhaust their whole nature. There
will always make themselves felt oppositions that are determined by the limit
of identity. The consideration of the most important relations of opposition
of this kind will serve to illuminate the nature and conditions of our
cognition.

\subsubsection*{α. Quality and Quantity}
\addcontentsline{toc}{subsubsection}{α. Quality and Quantity}

115. As we have already seen in the doctrine of categories, science, in the
face of irreducible differences, resorts to the expedient of putting, by way
of analogy, quantitative differences in place of qualitative ones. It is the
qualities time, number, degree, and place that can most easily be reshaped in
such a way that the principle of identity can hold in the series they form, so
that substitution can become possible. And in analogy with these series the
other quality-series are ordered, so that they---as far as the analogy
holds---can come under the principle of identity. These series can then be
constructed by constant addition of identical members, just as we construct
time-series, number-series, scales of degree, and geometrical figures.---It is
first and foremost natural science that in more recent time has proceeded by
this way. In psychology a similar method can be applied only in an imperfect
degree (see 12--13).

That we can, in natural science, put quantitative changes in place of
qualitative ones comes from the fact that we have several different senses,
and that in some of these differences of magnitude are more clearly prominent
than in others. Thus touch- and movement-sensations have a more quantitative
character than sight and hearing---that is: the qualities of the former can
more easily be interpreted as having arisen through the joining-together of
identical elements. The qualitative differences in touching a larger and a
smaller surface are easily seen to hang together with the difference of
magnitude between the two surfaces; and likewise the qualitative difference
between the movement by which I touch my head and the one necessary to take my
hat from a high stand can easily be seen to hang together with the smaller or
greater extension of my arm. Movement-sensations are connected with most
other sense-sensations, and thereby measurement of these can become possible.
Where merely qualitative estimation leads to disputes, measuring-out by
movement---not only our own movement but also movements in the surrounding
world that hang together with the object the estimate concerns---can turn the
scale.

Numbers and figures can be constructed, in that they arise by repetition of
the same act of thought, the one by which a unit is represented. New
combinations can, by virtue of the repetition of the identical, be found out,
without one needing to look to experience. Thought is therefore more
self-active in the case of quantitative differences than when it stands
before qualities. These it must wait for; those it can anticipate, merely by
repeating, according to certain rules, the same acts of thought. \emph{Kant}
has clearly expressed this (in the remarkable letter to Herz of 21 Feb.
1772): ``Only thereby do things become magnitudes for us, can they be
represented as magnitudes, that we can produce the representation of them by
repeating one thing a number of times. Therefore the concepts of magnitude
can be self-active, and their fundamental propositions can be found a priori.
But in the case of qualities the question how my understanding should, wholly
a priori, be able to form concepts of things always leaves an obscurity with
respect to how our faculty of understanding should be able to agree with the
things.''

Already \emph{Aristotle} taught that movement in space (φορά) lay at the basis
of all other change (μεταβολή), whether this consisted in change of quality or
in increase and decrease (αὔξησις καὶ φθορά). But he nonetheless combated
\emph{Democritus}'s and \emph{Plato}'s attempt to trace all change back to the
movement of small parts. Partly he maintained the originality of the
qualitative differences, partly he found the teleology in nature violated by
the atomistic mechanics. Only modern natural science made earnest of laying
the quantitative properties at the basis. \emph{Galileo} stands as the great
founder of what one (in the narrowest sense) usually calls the mechanical
conception of nature. Besides the simplicity that is attained by having merely
to do with differences of magnitude---and Galileo was sure that nature
follows the simplest ways---there was also gained the advantage that one could
attain greater exactness in one's descriptions and therefore greater certainty
in one's inferences. Therefore Galileo wanted everything measured that was
measurable, and to make measurable what was not. He even considered the
quantitative properties to be the things' ``first and real properties,''
while the qualities were supposed to be only appearance. And this was in the
following time the usual conception among the representatives of the
mechanical conception of nature. The principles according to which natural
science worked were conceived as the expression of absolute realities. One
made the analogy between quantities and qualities into identity and saw in
movement and the elements that moved a world that was the true material
reality. This world was, according to \emph{Descartes} and \emph{Newton},
immediately created by God, while materialistic authors like \emph{Holbach}
held that it had subsisted from eternity as the only true reality. The
physicists were on this point metaphysicians right up to the most recent time.

116. From various sides it can be shown that a complete solution of our
problems is not reached by laying the mechanical conception of nature at the
basis in a dogmatic way.

Even if we thought that everything in material nature could be explained
according to this conception's principles, the qualities nonetheless remain
standing as immediate facts, and there is given no explanation of how the
purely quantitative differences can appear to our senses as qualitative
differences. The chemical product's properties cannot be derived from the
elements' properties, and when one kind of physical energy gets its
equivalent in another kind of physical energy, the equivalent nonetheless has
other properties than the first energy-form. The sensing subject cannot,
after all, create these qualitative differences out of nothing. In any case
one would thereby put insoluble psychological difficulties in place of the
chemical and physical difficulties one pushes aside.

It must moreover not be forgotten that there is only a difference of degree
between differences of magnitude and other differences. The former appear
immediately just as much in a qualitative way as the latter. Extension and
movement are qualities that can just as well need an explanation as the
specially so-called sense-qualities. This point was especially emphasized by
\emph{Leibniz}, \emph{Berkeley}, and \emph{Kant} against ``the mechanical
mythology.'' We have no ground to believe that we, when we have come to the
quantitative properties, are at the end, have the innermost essence of things
before us. The purely methodical significance of the mechanical conception is
very well compatible with only dynamic and symbolic truth being ascribed to
this conception. Its significance consists in its laying at the basis a circle
of symbols that can be applied with full consistency---in its having found an
analogy by whose help inferences can be drawn that are confirmed by new
experiences, which can again be expressed by help of the same symbols. But the
possibility is never excluded that another circle of symbols or models could
just as well, if not better, express the experiences and form the basis for
exact inferences. There will never be able to be given a proof that any circle
of symbols is the right one or the only possible one.

\subsubsection*{β. Succession and Causality}
\addcontentsline{toc}{subsubsection}{β. Succession and Causality}

117. The real ground of the grand attempt at conversion of quality into
quantity must be sought in the fact that the causal relation is not clear and
transparent so long as there continues to be a difference of quality between
the event that stands as cause and the one that stands as effect. When thought
seeks after a cause, there are two things it wants clarity about: why an event
occurs precisely in this definite moment, and why it appears with this definite
qualitative peculiarity. In respect both of time and of quality, the single
event is sought to be determined in relation to other events, so that it can be
inferred from these. The difficulties one has found in the concept of cause
concern precisely the differences of quality and of time. With what right do we
assume that, because the event A occurs in this moment, the event B must
therefore occur in a later moment? How can events of different character and
from different times be so connected that they cannot be separated?

The most important answer will be the counter-question: with what right does
one assume that there are absolute differences of time and of quality? Should
it not be possible to order the events in such series that from the earlier
members one can infer the following ones?---This was the way \emph{Kant} went,
in that he pointed out that there was indeed no identity, but there could yet be
analogy, between the series of events in time and the series of grounds and
consequents (premises and conclusions) in logical and mathematical thinking.
The real categories could thus be conceived in analogy with the formal ones.
And---Kant adds---only because this analogy holds can we distinguish between
dream and reality. If Kant had here said ``in so far as'' instead of
``because,'' he would have been right. He moreover proceeds dogmatically from
the assumption that this criterion of reality can be applied everywhere, in all
changes in our objects. He presupposes, in other words, the validity of the
causal principle at the very moment he wants to prove it. But the question will
pose itself anew at every new change. That was surely also what \emph{Hume}
would have answered Kant. Even if the very same situation came again, one would
not, he held, have the right to expect the same consequences; for it was not
certain that the same things were now in possession of the same power as when
the situation was present the previous time. And here no matter how many
repetitions will not bring us further: ``there will still be occasion for the
same question, and that to infinity.''\footnote{Cf. \emph{Hume}: \emph{Treatise
on Human Nature}. I, 3, 6. (ed. Selby-Bigge p. 91.)}

Kant has, without more ado, made a principle or a hypothesis into a law. His
apt connection of the causal problem with the criterion of reality (where he
moreover has all more recent philosophers, right from Descartes on, as
predecessors, see 37) does not justify going further than to maintain that we
here have a means of thought and a form of thought that we must continually try
to apply, already for the reason that we could otherwise not orient ourselves
in our surrounding world. But the concept of reality is and remains an ideal, to
which only approximation is possible. And therefore the concept of cause too
designates an ideal of thought that can never find exhaustive conditions for its
application. A complete confirmation the causal principle cannot get, and when
it is called a law, this can strictly speaking hold only for the cases in which
it has shown itself to be sufficient guidance for us. Its most important
significance is constantly to lead us on to new tasks. And where we cannot
solve these tasks, we have only the right to say that we have not found the
cause, not that there is none. The possibility was not excluded that we could
indeed apply the criterion of reality to establish the truth of a series of
events each for itself, but could not find any connection among the events
mutually (so that these form only a series of kind (2) (see 63), when the
chronological order is assumed determined). It is such a situation that Hume
seems to have imagined, in that he does not presuppose doubt about the reality
of the single events each for itself.

Even if the application of the criterion of reality is forced upon us by
practical grounds, and even if we must also apply the concept of cause in
practice by seeking means to our ends, still causality is not a means of thought
in a purely external sense. Neither is there here any purely external relation
between form, need, and object. Our thought's peculiar forms it will have a
spontaneous need to apply, without its needing to be conscious that it is
setting up a hypothesis and carrying out an experiment. To Hume's question: Why
do you apply the causal principle to new events?---the answer could be: Why not!
Let us try our powers, try how far we can get with our thought's forms and
means. We cannot behave like the man who would not go into the water before he
could swim. And when the objects show themselves capable of being ordered in
such series as the causal principle requires,---namely in series that are
ordered analogically with a series of grounds and consequents,---then we also
get courage to continue with our attempts.

Kant's discovery of the ``analogies of experience'' will be science's last word
on the significance of the causal principle, although he himself believed that
this last word contained more than it does. The experiences of daily life are
here connected with science's highest investigations under one and the same
type of thought, which hangs closely together with the inner organization of
our thought. We get through this point of view a constant summons to investigate
the objects as closely as possible, so that the analogy can be carried through
as far as possible. In the logical relation between ground and consequent,
differences of time and of quality play no role, and for the analogy to be
carried through one must therefore reduce such differences between the objects
to the least possible. Thought seeks to shape the objects into likeness with
itself---thereby is attained both the greatest intellectual satisfaction, and
there is won a more perfect means of orientation.

118. The relation between analogy and identity nevertheless brings with it still
more questions with respect to the causal problem. From the empirical side it
has been maintained that it has absolutely no significance whether the changes
that are to stand in causal relation are mutually dissimilar. \emph{Stuart
Mill}\footnote{\emph{System of Logic}. V., 3, 8. (10th ed. p. 336--344.)} thus
quite rejected the old proposition that like produces like. He regards it as an
expression of superstition and points to the many cases in which one has placed
things together as causes and effects merely because they had a certain outward
likeness to each other---as when yellow flowers were supposed to be able to cure
jaundice!---Of course the likeness must be such that the analogy with ground and
consequent really fits. The more recent science aims at approaching the analogy
to identity in as high a degree as possible. It is more than superstition, also
more than a mere effect of experience, when thought sees in every difference of
quality between events a task set for it; it is because thought is indeed
awakened by the differences, but thereafter seeks out beyond the tension and
wonder called forth by them. Wonder is only the beginning of wisdom. If one
establishes that a causal relation can be wholly clear and comprehensible even
though the members are ever so different, this will lead to one's coming to rest
too early. One has then pulled out the sting that spurs to progress.

It was \emph{Comte}'s and \emph{Mill}'s mission in the philosophy of the 19th
century to maintain the differences of the objects and their right over against
speculation's zeal to find ``the unity in all.'' \emph{William James} continues
their work in the most recent time. But in this, nevertheless, \emph{Schelling}
and \emph{Hegel} were not mistaken, that it is unity we seek. Their error was
that they too rashly believed themselves to have found it, and next, that they
proclaimed the possibility of deriving the multiplicity of objects from the
unity. But the doctrine of plurality (pluralism) contains just as much a danger
as the doctrine of unity; it too easily stops too early. The life of thought
develops only under the constant conflict between the tendency to multiplicity
and the tendency to unity, and the highest fundamental principles for cognition
must be such as contain directions for a labor under this condition.

\emph{William Hamilton}, who conceived all categories as expressions of our
imperfection, because they show that we cognize only by setting the one thing in
relation to the other, conceived it consistently also as an imperfection that we
seek after causes. It was supposed to witness that we lacked the capacity to
apprehend change as an increase of reality. It is in accordance with this
imperfection that we seek to apprehend the new as a conversion of the old into a
new form. There is this truth in it, that when such a conversion has taken
place, the difference is evened out for us; there is now needed less energy to
hold the objects together; there is, after all, won a point of view from which
they can be regarded as identical. And here it is then indeed possible to end
and to come to rest. But it is also possible that the equilibrium won and the
energy saved can be applied to new tasks, so long as new objects present
themselves. A rhythmic alternation of awakening and evening-out is
characteristic of the life of thought.

119. From quite other points of view than the one that led Hume to pose his
problem, one has in the most recent time criticized the concept of cause and
demanded its elimination. It is the time-relation that this concept, on the
usual conception, contains, that has called forth this criticism. It is indeed
especially the time-relation that makes the causal relation different from the
relation between ground and consequent and offers only an analogy with this.

From the speculative side it has been urged that so long as the time-relation is
determining for our conception of existence, our conception is imperfect, and an
absolute cognition is not reached. And our cognition works itself---one further
maintains---more and more away from the time-relation: the more clearly we
understand something, the smaller a role the differences of time play,---the
more the real knowledge merges into the formal, the cause into the ground. What
we at the beginning called cause and effect comes, with cognition's advance, to
stand as members in a whole, a whole of thought, and the relation between them
becomes purely logical, so that the time-sequence is inessential. That we first
observe the one member, then the other, has now no significance; for in our
cognition of the law-governed connection the whole process stands before us as a
unity. Not the time-relation, but the unity behind and in spite of the
time-relation, binds what we call cause to what we call effect. And there is no
interval between the end of the event we call cause and the beginning of the one
we call effect.\footnote{See especially F.\ H. \emph{Bradley}: \emph{Appearance
and Reality} (1893). Chap. 4--7; 18. --- Similar train of thought already in
Plotinus (Drews: \emph{Plotinos}. p. 181.)}

From another point of view one has come to a similar result. The interest here
goes to sorting out all the accompanying representations and auxiliary concepts
with which the objects, what is given for thought, are involuntarily or on
methodical grounds supplemented. Cognition is to approach as soon as possible a
pure description of a continuous process, within which the differences are
reduced to a minimum. This is attained the more quantitative determinations
replace qualitative ones, whereby it becomes possible to demonstrate relations
of equivalence between the objects that replace one another in experience. It is
then no longer the task of strict science to ``explain''; whoever wants
``explanations'' is referred to mythology or metaphysics. Science's goal is to
give an exact, methodical description of all relations and transitions in
experience. As now especially regards the causal relation, the objects'
relation of dependence is expressed better by help of the mathematical concept
of function than by the indefinite expressions ``cause and effect.'' Moreover
there is this unfortunate feature in every application of the concepts cause and
effect, that two objects are arbitrarily taken out of the whole continuous
connection in which they occur, and set in opposition to each other. And the
time-difference between ``cause'' and ``effect'' has no significance of
principle when the relation is formulated as a mathematical relation of
function, expressed in an equation. That time cannot be ``reversed'' really
means merely that the changes of the physical magnitudes take place in a
definite direction.\footnote{\emph{Richard Avenarius}: \emph{Kritik der reinen
Erfahrung}.\textsuperscript{1} II., p. 332--335. --- \emph{Ernst Mach}: \emph{Zur
Analyse der Empfindungen}. p. 168. \emph{Die Mechanik in ihrer
Entwickelung}.\textsuperscript{6} p. 259 f.} ---

The epistemological views that thus, with different motivation, would eliminate
the concept of cause and---in connection therewith---deny the significance of
principle of time, can be characterized by the fact that they put an ideal of
thought in place of the actual cognition we can possess at any given time. Every
investigation must begin at a certain point---in so far, with a leap. But it is a
leap that is made precisely at the point where the problem presents itself, and
where wonder is awakened. That elimination really brings with it the abolition
of all problems and of all wonder. The accidental element in the point of
departure, ``the leap,'' is really present in every act of thought. In an
inference it is arbitrary that we begin precisely with these definite premises
and seek what follows from them, and likewise it is arbitrary that we stop at a
certain proposition and seek precisely its presupposition. All thinking
presupposes limitation. Our cognition develops historically, because attention
is awakened under certain definite conditions and cannot be directed all at once
toward all objects in their successive appearance. Naturally it is of importance
that one becomes conscious of the accidental or subjective element in the point
of departure or the section, and if the labor of thought has progress, the
section too will of course become worked into a great continuous connection or
whole; only thereby is the full understanding won.

The historical character of our cognition becomes especially clear when we
reflect that we always stand between the experiences of the past and the
possibilities of the future. In each single moment we must distinguish between
what lies given and what we expect, and the expectation can never become more
than a hypothesis. The single moment, in which toward the one side a ``no
longer'' holds, toward the other side a ``not yet,'' poses the problem to us in
its whole tension, which only the power of habit often diminishes. With new
problems the tension cannot in any case fail to appear---and thought seeks and
constantly finds new problems, so long as experience is continued. The full
connection lies before us only afterward, and until it lies before us, the
assumption of it stands as a hypothesis. We shall therefore never be able to do
without concepts like force, energy, cause, possibility, which with different
nuances express one and the same relation: the relation of the earlier-appearing
objects to the later, apprehended in analogy with the relation between ground
and consequent.

The matter is not changed by the relation between ground and consequent being
expressed in the form of a mathematical equation. The equation is, after all,
only a special kind of logical judgment. It expresses, like this, a timeless
relation. For there is no time-difference between subject and predicate, or
between ground and consequent. Between the logical and mathematical relations on
the one side, the time-relations in the objects' appearance on the other side,
there can therefore always be only analogy, not identity. But this analogy
nevertheless makes the empirical science we have possible, even though it leaves
great questions standing behind.

120. One can therefore in no way say that Hume's problem has been solved by the
discovery of the conservation of energy, which makes it possible to formulate
the relation between the different forms of energy as equivalence. The relation
of equivalence excludes neither difference of time nor difference of quality,
just as little as the relation of identity needs to do so (see 69 and 73). When
A and B are equivalent, it is in a certain sense indifferent whether I pass over
from B to A or from A to B. And yet in each single actual case it is either A
that passes over to B, or B that passes over to A,---and this difference can be
a difference between life and death. In the Carnot theorem (the entropy theorem)
this comes clearly to the fore; here the direction is precisely decisive. Heat
and motion are equivalents; but it is in fact more difficult to concentrate the
heat in such a way that it can again be converted into quite the same quantum of
motion than to carry out the reverse conversion. It is a great abstraction to
disregard this difference. The Carnot theorem does not conflict with the theorem
of the conservation of energy; it is precisely one of its predecessors. But it
leads over to a more concrete problem, leads us into nature's real economy,
where precisely the directions of the causal series become of significance. An
example from the human domain, corresponding to the Carnot one from the physical
domain, is offered by the relation between genius, exemplariness, new fresh
contributions within human development, and the dominion of imitation and
repetition. Between discovery and imitation one could imagine a relation similar
to that between heat and motion, and here too there could lie a danger in the
tendency to evening-out and distribution.\footnote{Cf. G. \emph{Tarde}:
\emph{Les lois de l'imitation}. p. 412--415.}

\emph{Ernst Mach}, who otherwise does not lay great weight on time's not being
able to be ``reversed,'' because it is ``only'' the direction that becomes
different, nevertheless occasionally declares that the natural processes have a
historical character, and that in this there lies a great problem. ``When we say
that time flows in a definite direction (Sinn), this means that the physical
(and thus also the physiological) processes are completed only in a definite
direction (Sinn). All differences of temperature, electrical differences,
differences of level in general, are diminished,---not enlarged,---\emph{when
they are left to themselves}. If we consider two \emph{bodies left to
themselves}, touching each other, of different temperature, memory will offer
greater differences of temperature, observation smaller, not the reverse.
\emph{In all this there expresses itself throughout only a peculiar, deep-lying
connection in things.} But to want to demand complete clarity here already now is
the same as, in the manner of speculative philosophy, to want to anticipate all
future special research's results, thus to demand a completed natural
science.''\footnote{\emph{Die Mechanik in ihrer Entwickelung}.\textsuperscript{6}
p. 239 f. --- [The emphases are mine.] --- In his learned and acute work
\emph{Identité et Réalité} 1908 \emph{Meyerson} has set the entropy theorem in
sharp opposition to the principle of equivalence: the former witnesses to time's
reality, in that the members cannot be exchanged, while the latter tends to
abolish time and finally all difference.} Here it is thus acknowledged that not
all riddles are solved by the demonstration of equivalence. There is a
historical element that one cannot get away from. The great question is to what
degree the direction of the causal processes is left to itself. Just as the
energy theorem can only be demonstrated within connections that are thought as
closed, so that no supply or departure of energy can take place, so it goes also
with the entropy theorem. By both ways it shows itself that it is an abstraction
when we hold to the causal processes without taking account of the real whole
within which they take place, and which can contain tendencies of opposite
direction.

121. Although I have sought to show that the historical element cannot disappear
from causal research, the immediately preceding consideration shows that it is
first the concepts of whole and development that properly emphasize the
historical. When research stands before objects that lead to the application of
these categories, it stands before its most difficult tasks, and at the same time
before its last tasks. Concerning the questions that here arise, there has partly
already been discussion in the doctrine of categories (93--99), and in the section
on world-view we will come back to them. Here---under the problem of
cognition---the matter concerns especially the principles for cognition that those
categories lead to asserting, and this side of the matter we can investigate in
close connection with the causal problem.

The analogy on which the causal principle is built becomes the more difficult to
carry through, the more concrete and individual the object is that is to be
explained, and that is to say: the more it constitutes a whole and runs through a
development. There is here a whole scale with rising complications from simple
locomotion through the physical and chemical processes to the organic, psychical,
and social phenomena. The further we come forward in this series, the more the
causal principle takes on the character of an ideal demand that we can satisfy
only with a slight degree of approximation. It becomes ever more difficult to
resolve the composite processes into simple causal series,---ever more difficult
to resolve the single complicated objects into elements, each of which for itself
could be regarded as a point of departure. Ever more difficult therefore it also
becomes to establish that it is quite the same objects and the same states we have
before us when a process seems to repeat itself. The old proposition that nature
is always in agreement with itself (sibi consona, as Newton said) and makes no
leaps, seems often to be contradicted by observation. Precisely the most recent
time's research yields examples of this. In radium there was discovered a
substance whose energy-discharges could seem to conflict with the theorem of the
conservation of energy. In opposition to the older Darwinism one maintains, on
the ground of observation, the sudden arising of new types (by mutation), which
for the present cannot be established to be the result of a successive process.
(Cf. however 92.) A master in descriptive psychology like \emph{William James}
maintains that there can occur new contributions in the mental development which
could not be derived from the preceding processes, and prepares us on the whole
for the fact that we must often content ourselves with a purely external
connection of objects, a mere ``conjunction'' without rational connection, an
``and'' without a ``therefore.''

Such cases nevertheless do not lead to giving up the causal principle. They pose
problems---precisely by virtue of the causal principle. Even if, besides
differences of quality and of time, types of whole and series of development offer
ever so many riddles, still thought has after all no other means than the causal
principle with which to illuminate them. Yet there are some special fundamental
principles of a methodical kind, which hang together with the categories whole and
development, though they will in the particular show themselves at bottom to be
applications of the causal principle.

α. Where a process seems to appear isolated, ``left to itself,'' it must
nevertheless always be pursued to see whether its direction does not lead to
combination with other processes, which together with it could make possible a
development of a whole. It must in general be set up as a fundamental principle
that no element and no process is isolated. Isolation is due to an abstraction, an
``accidental point of view.'' There will naturally also be occasion to investigate
whether the apparently isolated and simple process is not due to a separating-out
from a whole that has perhaps entered into dissolution.

It has been necessary to resolve the natural connection in order to find the laws
that hold it together. These laws could not have been found except by, as far as
possible, isolating single objects, elements, or processes. But this abstraction
cannot be cognition's last word. It is not merely Romanticism, it is clear
reflection itself upon science, its presuppositions and its method, that leads to
demanding that the problem of the whole be heeded, whether it can now be solved or
not.

β. If, conversely, it is a whole that is given, the task will be to see it as a
``samfoldighed'' of laws and processes. Where in the former case a synthesis was
required, here an analysis is required. To let the whole explain itself, by
presupposing mystical unities and thoughts of purpose, gives no explanation---save
on the animistic standpoint. To explain an organism's coming-to-be, or a single
organic phenomenon, by appealing to a ``vital force,'' gives only an imagined
explanation, just as when one explains special mental utterances by the
intervention of ``the soul,'' or in sociology lets ``society'' appear as a kind of
substance different from and hovering over the single individuals.\footnote{Cf.
with respect to the organism my lecture on Vitalism (\emph{Mindre Arbejder}. I.),
--- with respect to the soul my \emph{Psykologi}. I, 1 and 7 as well as the present
work \S\ 10, --- with respect to society my \emph{Etik} VIII, 4; XIII, 5 and the
present work \S\ 40. --- The analogy that is between the three concepts comes to
light in the fact that one often uses the word organism about society, and the word
society about the organism, as well as both words about the soul, and the word
ensoulment about life and about society.} When one sets the whole in opposition to
the elements (e.g. the organism in opposition to its parts, the soul in opposition
to its thoughts, feelings, or strivings, society in opposition to the single
individuals), this can only mean that beyond the activity we in the single cases
find in the single part, there is still a world of possibilities, a capital of
potential energy, which under certain conditions can be released and alter the
single part's activity. And there is here a constant reciprocal action. For the
single part (cell, organ,---thought and striving,---individual and class) can itself
contribute to continued formation of possibilities, just as it will itself show
itself to have developed by help from a world of possibility that was more
comprehensive than itself.

γ. A third task consists in finding the different steps through which the whole has
reached its present form. Here shows itself quite especially all research's
historical character---not merely in the sense that research itself has a history, a
historical course, but also in the sense that its highest and last object is
history, finally ``the world's'' history, when by the world we understand the most
comprehensive whole of objects we are able to think. ``The universe's'' biography
comprehends and presupposes all science---and points constantly out beyond every
attainable science. Only art, not science, can finally give us pictures of a whole.
Thought ends always with an open horizon,---with a question or an exclamation, never
(when it understands itself) with a full stop. The ``meaning'' is never ``all the
way out.''

122. There is thus enough for thought to do before it gets its last word said. And
this last word itself must be said by virtue of the causal principle. It sounds
perhaps as though no more causes can be found, but can never sound as though there
are none.\footnote{On the expedient of stopping at something that is ``cause of
itself,'' cf. my \emph{Religionsfilosofi}. \S\ 10. (Second edition p. 37--38.)} As
I believe I have sufficiently shown, no proof can be given for the causal
principle's absolute validity. It is a working hypothesis---in the sense in which I
above (113; 117) have taken this word. Should there now be objects in the face of
which we must give up applying this working hypothesis?

One has thought that the human will was such an object, and one has---remarkably
enough---grounded this opinion in the interest of maintaining the human
personality's peculiarity. One did not see that one precisely violated this by
denying the possibility of a law-governed connection within the personality. For
personality precisely presupposes unity and continuity in a higher sense and in a
higher degree than any other totality we know. It therefore presupposes the causal
principle, even though its special application in the face of so intimate a whole is
bound up with special difficulties. There is no ground why we should not here too
meet with our working hypothesis and apply it as well as we can. It is in this sense
that I maintain, and have long maintained, determinism. I do not dogmatically assert
that everything, including the human volitions, is subject to the causal principle;
but I assert that neither ethical nor psychological grounds speak against it. The
ethical ones we have not here to do with;\footnote{See on this my \emph{Etik}. Chap.
V.} on the other hand there is here, under the problem of cognition, occasion to
dwell a little on the methodical side of the question, which will lead to bringing
forward some psychological considerations.

If by determinism one understands a conception that sets motive and action, motive
and personality, in a purely external relation to each other,---roughly like the
weights and the balance-pan,---then it is the psychological absurdity that has already
been rejected in an earlier consideration (10). Motives are not something one
``has'' or ``gets''; there can here be no distinction between possessor and property,
between receiving and giving. The will works throughout our whole life from the
simplest utterances of instinct and drive to the highest resolutions; it stretches
through all motives and actions as the inner binding band. The agent is himself in
his motives and in his actions, when he really is a personality. There can here be
distinguished between a central element (which I usually call the real self) and a
peripheral, with all possible intermediate degrees; but out of the circle we never
come. Here precisely the concept of the whole comes to light in its whole
peculiarity. This is the main thought of determinism, and it is in trust in it that
we seek to unravel as much of the inner world's connection as possible, which can
happen under the guidance of the causal principle. Under this endeavor we must then
often provisionally regard a single side, a single element of the personal life in
isolation; we must take single members out of the objects' continuous stream. But
this is only a means to finding a so much the deeper and firmer connection. He who
denies the causal principle's validity for the will, and thus embraces indeterminism,
must on the contrary once and for all abolish or deny the inner connection that makes
the will, the personality, into a whole.

\emph{Henri Bergson} stands in this question on a peculiar footing. He maintains, as
we know, that the immediately given in mental life is a continuous stream, which one
can only in an artificial way resolve into elements. And freedom consists precisely
in a streaming-on, through which the past's I glides over into the future. Here there
are no external differences, and no analysis is possible. Both determinism and
indeterminism analyze, separate elements and set them in external relation to each
other,---the former in a relation of cause and effect, the latter as wholly different
from and independent of each other. It is an external relation, both when one
separates elements as cause and effect, and when one sets up certain elements as
having arisen without any connection whatever with other elements. Bergson rejects
both indeterminism and determinism; but he warns against defining what ``freedom''
is, since this will always lead back to determinism, because one can only define
``freedom'' by making the act of will dependent on the personality. One can then,
according to Bergson, only maintain ``freedom'' when one does not say what it is! And
yet he has himself defined freedom, when he finds the continuity between past and
future to be its essential mark.\footnote{\emph{Les données immédiates de la
conscience}.\textsuperscript{4} p. 166--168; 174 f. --- A related but peculiar
standpoint is taken by W. \emph{James} in the following utterance: \emph{The only
``free will'' I have ever thought of defending is the character of novelty in fresh
activity-situations. The experience of activity} (The Psychological Review. January
1905. p. 15.)} He is then himself a determinist, not merely because he defines
``freedom,'' but because he defines it in the way he does. Bergson's merit is
precisely to have maintained mental life's continuity over against such analyzing as
confuses its elements, formed by the art of abstraction, with the very drops in
life's stream.

When we so often stop in our search for causes in the psychological domain, it is
especially on account of the fact that it is here far more difficult than in other
domains to demonstrate identical relations and complete equivalence. But that this is
more than a difference of degree, no one has the right to assert.

123. Often one appeals to an immediate consciousness that one could have acted
otherwise than one did. Psychologically regarded, it must be said to this that if
such a consciousness is given, the task must be to show how it has arisen. And there
present themselves here several different possibilities.

The more a mental state lays claim to our interest and attention, the less we heed
the preceding and conditioning circumstances, and the more easily we forget them. The
whole state comes to stand for us as arisen ``of itself,'' and we are disinclined to
enter upon a genetic explanation of it. Most of all does this hold in the case of a
serious decision of the will. When, after a long inner debate, the decisive
resolution thrusts itself forward and in a moment chases away all opposing
misgivings, then there seems to lie before us a break in the continuity. The new
state, in which our whole mind is concentrated in the thought of the action, stands
in such opposition to the deliberation's waste and strife that even though we retain
a memory of this, we can nevertheless not easily imagine an inner connection between
the two states. Achilles' sudden and decisive self-mastery in the face of Agamemnon
(in the Iliad's first canto) Homer could only explain to himself by Athena's coming
down from Olympus and---visible only to the wrathful hero---pulling him by his locks
and holding his sword back in the sheath. The ancient Greek consciousness appealed in
such cases to Olympian intervention; the modern indeterminism to ``the free will.''

The decision of the will itself, the resolution, moreover brings with it a peculiar
sensation of unhamperedness and force by reason of the concentration upon a single
thought. All inner resistance and constraint has fallen away. What hitherto perhaps
rested hidden in us, held down by superficial and momentary promptings, now comes to
light; there occurs a breakthrough of our real self; we have a sensation as at a
whole and full breath; it is as if we now first ``came to ourselves.'' No wonder that
we cannot in one and the same moment be in such a state and at the same time clearly
hold fast that it has developed out of preceding states.

If now nevertheless the memory of the state of deliberation can make itself felt, it
will be able to lead to assuming the objective possibility of another resolution than
the one actually made. There presented themselves during the deliberation different
possible actions before our gaze. We entered into each of them, perhaps so vividly
that each for itself almost already stood as a reality. Therefore such a rejected
possibility can stand as more than merely thought, as a way that really existed side
by side with the way we struck into in the moment of resolution, just as real as this.
The two ways seem to have crossed each other, and from the one we actually walk along
we cast a glance over to the other way, which also lies there, but which we did not go.
And thus it will especially present itself when we, in the moment of repentance,
command motives that now have a force they lacked in the moment of decision, and which
we involuntarily lay back into the representation of that moment, forgetting that it is
precisely a fruit of the repentance.

A similar illusion constantly makes itself felt when one, with strong interest and in
tension, thinks of the future. The future will indeed become different, according as I
choose the one or the other of two actions. It therefore stands for me in a double
light: partly as it will become when the one action is carried out, partly as it will
become when the other action is carried out. Neither of the two pictures of the future
fixes itself, so long as the deliberation lasts, and it then looks as if the future
itself were undetermined, wavering. Our own indeterminacy is involuntarily confused
with an uncertainty in existence itself. In antiquity this assumption of the future's
indeterminacy made itself felt not merely in the face of the part of existence that is
determined by human action, but in the face of existence itself, as regards the
sublunary world, the world of changes. \emph{Aristotle} taught that to our problematic
judgments there correspond objective possibilities in nature, just as to our apodictic
judgments there correspond objective necessities in nature. This doctrine was due to a
naive transference of the forms in which our labor of thought proceeds onto existence
itself. A remnant of it is what has held itself in the assumption that we should
immediately be able to know that we, at an earlier time, could have made a wholly
different resolution than the one we really made. ``The truth of an event,'' says
\emph{Hobbes} in his critique of the Aristotelian doctrine, ``does not depend on our
knowledge, but rests on what works prior to the event.''\footnote{\emph{Philosophia
prima}. IX, 5.} And it can be very difficult to discover all that works in us prior to
our resolution.
\addcontentsline{toc}{subsubsection}{β. Succession and Causality}


\subsubsection*{γ. Subject and Object}
\addcontentsline{toc}{subsubsection}{γ. Subject and Object}

124. In the investigation of the relation between quality and quantity and of the
relation between succession and causality we had to do with the relation between
object and form. We now direct our reflection toward both of them together, in that
we raise the question how far they are merely to be set to the account of the
cognizing subject or possess an objective character. The relation between subject and
object does not without more ado coincide with the relation between form and object.

There is an involuntary tendency to regard both forms and objects as immediate
revelations. There is ascribed to them a reality ``outside'' or ``besides''
consciousness. Thus it goes with the sense-qualities, with number, time, and space,
with the atoms, with causality, and so forth. They are fixed as absolute existences.
The question then becomes what is subjective and what is objective, and how a boundary
can be staked out.

Since our criterion of reality is the fixed and law-governed connection between the
objects, and since the differences of quality designate a limit for complete
demonstration of such a connection, there will naturally arise a tendency in the
thinking consciousness to set them to the account of the subject, while the objective
was supposed to be the quality-less atoms and their motions. In this direction went
the mechanical natural science. The speculative philosophy went still further and set
also differences of degree, of time, and of space (intensive, protensive, and
extensive differences) to the account of the subject; especially it was the difference
of quality that for the speculation of all times stood as the irrational that had to
be sorted out; had this happened, the pure logical thinking could unfold itself with
inner necessity. Both the mechanical natural science and the speculative philosophy
relied on the assumption that what was the condition for understanding and made
cognition possible must be the most objective of all.

In opposition to this the critical philosophy urges that a distinction must indeed be
made between the demands, in accordance with which by our thought's nature we arrange
the objects for ourselves in order to be able to win what for us is understanding,
---and the character of existence itself. On the ground of our thought's nature we work
unceasingly and involuntarily at reshaping the objects in such a way that there can
appear as great a continuity between them as possible. But the more our own mind's
nature and mode of working gain influence on cognition's results, the less can we
maintain these results as pure objects.---With criticism pragmatism here agrees, only
that it regards the forms as mere means, not as expressions of the subject's nature.

From this last point of view it then becomes not discontinuity, but continuity, that is
set to the account of the subject. The objects, on the contrary, become that which must
have another origin than the subjective.

125. Just as object and form are abstractions, since in every act of cognition we have
a connection of them, so it stands also with subject and object. They are two points of
view that can be taken up in different ways, but never separated. And they hold each for
itself both for the forms and for the objects.

We can make our own subject into an object for us, namely when we study it
psychologically, e.g. in order to find out the forms according to which it works in its
cognition. These forms, which are systematized in the categories, must be taken as
facts. They are made into objects when reflection is set going over them. Every
cognition takes place from a definite standpoint, which it can have its significance to
establish. We then objectify the subject. And this happens especially when we
investigate how the cognitive standpoint in question, the subject in question, has
developed to the step at which it now stands, and has gotten the forms it now works
with. It is thus not merely the objects, but also the forms, that can appear as objects.
---On the other hand no object is given or thinkable except in relation to a subject, and
as it appears for such a one. In every description of an objective order of things we
therefore presuppose, consciously or unconsciously, an apprehending subject, either
ourselves or a subject we construct in analogy with ourselves. The subject is the
Archimedean point in epistemology,---the point from which existence can be---not
moved---but apprehended. And such a point must always be presupposed. More than what
shows itself from such a point is not known, and otherwise than it shows itself from it,
it is not known. This too holds both as regards objects and as regards forms.

When we regard something as object, we must state the character of the subject in
relation to which it makes itself felt. And when we regard something as subject, we must
partly seek the objective connection that determines its character and thereby the
objects and forms that stand at its disposal, partly heed that we by this investigation
make it itself into an object for a subject (if it is ourselves, then for ourselves in a
somewhat other state, in any case in another moment, than before). We never have a pure
subject ($S$), but always an objectively determined or at least an objectified subject
($S_o$). And we never have a pure object ($O$), but always a subjectified object
($O_s$). $S$ and $O$ are mere abstractions. What we can have before us is always $S_o$
and $O_s$.

In accordance with this general consideration we can take up three different views with
respect to the relation between subject and object.

\emph{α. Psychological view.} I here seek to explain the cognizing subject's
coming-to-be, with its organization, its capacity of sensation and thought, and so
forth, by the way of natural science and history. Every subject is a product of certain
conditions, which I can seek, even if not always find. What is seen depends on the eyes
that see, and I here seek to find out how the eyes have gotten their present character.
So too with the thought-forms, which must have developed under physical and social
conditions. I then derive, as well as I can, the subject ($S_1$) from the object ($O_1$)
or from the aggregate of objects. This way go empiricism and evolutionism. But they
overlook that the inner conditions are a constant presupposition for the outer
influences. There are always presupposed earlier steps of what later appears as $S_1$.
That which $O_1$ works upon we can designate as $S_o$, since it in each case forms an
analogy to $S_1$. (Cf. 9.) Beyond the constant reciprocal action of inner and outer
conditions we do not come. There is here possibility of an infinite series:
\[ S_1 < O_1 < S_2 < O_2 \ldots, \]

The sign $<$ here means ``presupposes.'' The inner and the outer conditions presuppose
each other to infinity. In our investigations we move in each single case at a definite
place in such a series, but never come out beyond it. It is an identically varying
difference-series (65).

\emph{β. Psychological-epistemological view.} If I investigate the subject $S_1$ and find
its mode of cognition conditioned by the object $O_1$, then I must by virtue of the
constant belonging-together of $S$ and $O$ presuppose or construct a subject $S_2$, for
which $O_1$ is there. $S_2$ will not be able quite to coincide with $S_1$. The actual
character of $S_2$ is then perhaps conditioned by $O_2$, which presupposes $S_3$, and so
forth. This series too will formally always be able to be continued. It comes down to
whether we have sufficient objects. The problem poses itself ever anew, alternating in
psychological and in epistemological form. It is an interplay of causal explanation and
of interpretation. Every age, every people, every human being has its $O_s$ and its
$S_o$, which it is a matter of determining, if they will understand themselves and the
world that exists for them. But predominantly the explanation will be brought about by
later generations, who must then set themselves back to that state of things which
conditioned the earlier generations' apprehension and understanding, and learn to see
with their eyes.

If we let, as before, the sign $<$ mean ``presupposes,'' and let the sign $\{$ mean
``objectified by,'' we get the following series:
\[ S_1 < O_1 \mathrel{\{} S_2 < O_2 \mathrel{\{} S_3 < O_3 \ldots. \]

It is inconvertible and intransitive and differs from an identically varying series only
in that the relation is rhythmically changed. One could call it alternately varying.

\emph{γ. Epistemological view.} It stands in our power to form a series that formally
could constantly be continued, merely by constantly directing reflection toward our own
thinking or sensing. The thought $S_1$ can become object for the thought $S_2$, this for
$S_3$, and so forth. The act of self-consciousness will always be able to be exercised
anew, though it will of course get ever slighter content, the more it removes itself from
the original concrete and object-rich state. It is a self-producing series, always
continued according to the same law.\footnote{Cf. on such a series \emph{Dedekind}: Was
sind und was bedeuten die Zahlen? \S\ 60 (English trans. Chicago 1901. p. 60.) ---
Already \emph{Fichte} operates with series of this kind.}

If we again let the sign $\{$ designate ``objectified by,'' we get the identically
varying difference-series:
\[ S_1 \mathrel{\{} S_2 \mathrel{\{} S_3 \ldots. \]

The interest of this series is that it is produced by reflection itself, independently of
special objects or conditions. But the interest is predominantly formal, and the
alternately varying difference-series is richer in references to changing situations and
tasks. Only apparently is the concept of object pushed out in the series $S_1 \mathrel{\{}
S_2 \mathrel{\{} S_3 \ldots$. It is here $S$ itself that is constantly made into $O$. A more
special form, which offers peculiar interest, these series get when we take account of the
fact that both the psychological explanation and the epistemological objectification can
again be made into an object of objectification. The relation between $S_1$ and $O_1$ is
then in both respects object for $S_2$, and when this is again treated both
psychologically and epistemologically, the series is continued further:
\[ \left.\begin{array}{@{}c@{}} S_1 \\ O_1 \end{array}\right\}
   \left.\begin{array}{@{}c@{}} S_2 \\ O_2 \end{array}\right\}
   S_3 \ldots. \]

It is the advance through ``higher unities'' of a purely formal kind, which the
speculative philosophy tried to give a real significance. The interest of such a series is
that it expresses the law of synthesis, the constant combining of applied views and won
results. Could this series really be continued---which would require constantly new fullness
of objects and constantly wakeful psychical energy---then idealism's highest aspirations
would be satisfied. In any case the series contains a significant Excelsior!

126. From all this it follows that cognition's advance consists now in making the objective
subjective, that is, sorting out in cognition's objects what is conditioned by the character
of the cognizing subject, now in making the subjective objective, that is, making the very
nature of the cognizing subject into an object of research. We cognize the world through
the human being, but we also cognize the human being through the world. Romanticism
emphasized the first point of view and was inclined to overlook the second; positivism
emphasized the second point of view and was inclined to overlook the first. There is a
constant interplay between the two points of view; neither of them can be eliminated, and
the one cannot be derived from the other. In changing forms the relation continues to make
itself felt, so long as cognition is in development. In each single case we must determine
the $S_o$ and the $O_s$ we have to do with (whether $S_o$ is ourselves or another),---in
analogy with the fact that one can only describe a motion or apply the law of inertia from a
definite given point.\footnote{On theologizing Hegelians' attempt to cut the knot by
postulates about an ``ontological subjectivity'' see my remarks on the occasion of
\emph{Rasmus Nielsen}'s ``Grundideernes Logik'' in \emph{Danske Filosofer}. p. 189 f.}

It would be a misunderstanding to assume that subjectivity in itself is a hindrance to
cognition's constant advance. On the contrary---the more richly equipped the subject is, the
more forms and means it commands, and the more and the more many-sided its objects can
therefore be. $S$ and $O$ do not necessarily stand in an inverse relation to each other; of
this biology already witnesses through the nervous system's and the sense-organs' increasing
differentiation. Cognition is not a passive receiving or taking-in,---and even if it were
merely that, a rich equipment of capacities and organs would be an advantage. But it is a
labor, which can be carried out the better, the more many-sided the powers that are
commanded. Of course it comes down to whether precisely the capacities are present that the
definite objects presuppose. The subject can be richly equipped in its life of feeling and
fantasy; it will then perhaps not be able to perform the labor required to develop strictly
exact observations and trains of thought, but in return it can then be other sides of
existence that become accessible: \emph{Shakespeare} understood how human beings are
disposed under fate's heavy course and in severe conflicts, and in great religious
personalities there has been a feeling of life's oppositions, an inwardness and a loftiness
in the life of the spirit, which too can lay claim to bearing witness to what it is to
exist. ``Only they know the sea's secret who defy its dangers.''

Conversely, a fuller development of objectivity will also make a more perfect development of
the subject possible. Cognition is, after all, no closed system. The alternately varying
difference-series $S_1 < O_1 \mathrel{\{} S_2 < O_2 \mathrel{\{} S_3 \ldots$ expresses the
possibility that $S$ and $O$ can rise with each other. They also fall with each other; the
slighter the subjectivity is, the poorer will also the corresponding objectivity be. But
when the cognitive subject merely uses the energy it commands in its own independent way,
there will thereby nonetheless always be won the objective that can come to be at this stage.
It becomes perhaps only ``a personal truth,'' but it becomes a truth. \emph{Poul Møller}
maintained with right, in so far, that an assumption due to serious self-labor could never be
wholly false.---If $S$ wholly goes out, all truths and all riddles go away with it.

Cognition is constantly placed in this relation between $S$ and $O$. This necessary relation
\emph{Kant} overlooked, when he set up the ``Ding an sich'' as an absolute object. G. E.
\emph{Schultze}, and later \emph{Albert Lange}, have with right remarked that Kant
consistently ought to have granted that the distinction between the cognizable and the
uncognizable ``Ding an sich,'' just as well as all our cognition, could only hold for human
subjectivity. There would then have to be posited a new ``Ding an sich'' as ground for the
first, and so forth.---What Kant has had in view has nevertheless especially been precisely
the irrational, inexhaustible relation between subject and object (and from a somewhat other
point of view between form and object), and ``Ding an sich'' can in so far serve as symbol
for this inexhaustibility, which requires ever-renewed labor.

If we look at the different systems or world-pictures that human thought has formed in the
course of its historical development---let us call them $O_x$ $O_y$ $O_z$ \ldots---then they
cannot by a purely logical way be derived one from the other. The transition from $O_x$ to
$O_y$ (e.g. from Ptolemy to Copernicus or from Aristotle to Kant) presupposes the emergence
of new objects, which were not there for the $S_x$ corresponding to $O_x$. There must always
be taken up new premises, in order that we can come from the one system to the other.
Cognition, which is to present and explain existence for us, is itself always a part of
existence; therefore there can always emerge new objects for it and, consequently, new
objectivities. We have no knowledge beyond experience; but at no point have we the right to
regard experience as concluded. Cognition gives us, even where it reaches highest, only a
section of existence. Every reality we find is always itself again part of a greater reality.

127. The problem of cognition can only be posed from the human standpoint. It is thus a
special $S$ that lies at the ground. We can call it $S_m$. The human subject has certain
qualitatively marked peculiarities in its sensation and in its thinking, which we must take
as pure facts (125 α and β). Their necessity and absolute validity cannot be established. By
the way of the thought-experiment we can imagine an $S_n$ and $S_p$ and so forth---subjective,
psychical energies with other special forms, objects, and motives than $S_m$---in analogy
with the mathematicians' hypotheses about other spaces than the Euclidean. Unfortunately we
cannot construct these other subjectivities, but we use the thought-experiment to impress
upon ourselves the purely factual presuppositions that are given for a cognition with $S_m$.
One could well imagine that to the same objective differences of refraction there
corresponded other visual qualities than ours, and one could imagine that the logical
principles that stand for us as self-evident were derived from more universal principles,
just as the laws of inertia and of energy are perhaps special cases of more general
propositions.

In the face of a consideration of this kind one has declared that the truth is only one,
whether it is human beings, sub-human beings, or super-human beings that grasp it. A
proposition must, after all, either be true or false!\footnote{\emph{Husserl}: \emph{Logische
Untersuchungen}. I, 117; 151 f. --- II, 669. (The polemic is directed against Benno Erdmann.)}
---But to this it must be answered that what is really true for $S_m$ does not necessarily
become false for $S_n$, when $S_n$ from sure presuppositions (as forms and objects) can see
that the truth had to appear for $S_m$ as it did. There could perhaps be taken up a law
according to which what appears in a certain way for $S_n$ must appear in a certain definite
other way for $S_m$ and in a third way for $S_p$. This law would then be the real truth, if it
could be extended to all thinkable $S$. One of Leibniz's monads could perhaps discover (and the
monad that Leibniz himself was believed itself to have discovered) that existence is mirrored
differently in the infinitely different existing beings, without this abolishing its unity.
There is indeed not much material for such a comparative epistemology, which of course would
have to be written from the standpoint $S_m$. But $S_m$ is itself constantly placed in
development, and the history of the sciences, of philosophy, of art, and of religion now shows
us a series $S_\alpha$ $S_\beta$ $S_\gamma$ \ldots belonging under $S_m$. The most deep-going
task for the historian is not to describe mere events, doctrines, thoughts, pictures, and
dogmas, but to bring forward the subjectivities for which all these things had to stand as they
did, as regards content, validity, and value. This side of historical research is only in its
beginning. Quite predominantly we have hitherto read history from the most modern $S_m$.

Between contemporary subjectivities too the same holds. At essentially the same stage of
development different types will appear. It shows itself already as regards sensation, e.g. in
the difference between the color-blind and those who apprehend all colors, or between those in
whom one or another sense is especially developed. As regards the life of memory and
representation, it shows itself in the difference between visual, auditory, and motor
individuals. In some the forms, in others the objects, in others again the values, can within
the life of cognition play the greatest role. And the more the life of thought takes on a
personal character, the more types will appear.\footnote{Cf. on religious types my
\emph{Religionsfilosofi}. \S\ 35--43; 94--99. Cf. \emph{Mindre Arbejder}. Second series. p.
103--140.}

When truth is defined as the fixed and exact connection between our objects produced by the
labor of thought, then to the objects must also be reckoned the different standpoints from
which truth can be sought. The full truth must also contain the explanation of the errors and
of the fact that they could stand as truths. If there is conflict between different attempts to
find the truth, then the very definition of truth contains a measure for which attempt stands
highest: namely that in which the greatest multiplicity of objects is connected in the most
intimate way. That the truth is only one shows itself through its history, in which unity works
as a constantly onward-driving force, leading now to seeking out new objects, now to finding a
closer connection between those already found. Every seeker of truth has his place in this
history, of which we surely know only a slight part.
\addcontentsline{toc}{subsubsection}{γ. Subject and Object}


\section*{B. World-View}
\addcontentsline{toc}{section}{B. World-View}
\label{sec:verdensanskuelse}

\subsection*{a. Science and World-View}
\addcontentsline{toc}{subsection}{a. Science and World-View}

128. The problem of cognition arose with the question how far and how existence could
find expression in and for a single part of it. It showed itself that here one must
constantly struggle with difficulties; thought was now too big, now too small in
relation to existence, since this now demanded a single narrow way followed, now set us
before immeasurable expanses. Many thoughts must be pushed aside when the concept of
reality is to be formed; but the objects that set thought to work constantly demand new
thoughts in place of the old.

We now regard the relation between cognition and existence from another side, in that we
raise the question whether any of the objects that stand at our disposal is suited to be
laid at the ground of a concluding characterization of existence, so far as this is
accessible to us. Which object shall determine the timbre of the tone in which existence
makes itself known to us?---There is, as we have seen, an involuntary need to combine the
objects' multiplicity into unity, not merely in such a way that they are seen in their
law-governed relation to each other, but in such a way that a coherent picture forms
itself. From reflection's and analysis's going-through of the single objects we wish to
ascend to an intuiting, in which all the found and compared particulars find their place.
From the discursive we long up toward the intuitive. It is an intuiting that differs from
the one that formed the presupposition for reflection by the fact that it has gone through
reflection's purgatory. It is thus a concluding synthesis that is sought. If one will use
the old word metaphysics of an attempt in this direction, one can say that in all there is
a metaphysical need. Involuntarily the different objects that are known thrust themselves
together, and one of them will become determining for the whole's character. Even in those
for whom the word metaphysics is the expression of all evil in the world of thought, one
will, where closer investigation is possible, find such an involuntary thrusting-together,
a system-formation that has taken place without reflection having expressly been directed
toward it. Here too works what \emph{Kant} has called ``the hidden art in the interior of
the human being.'' Observation and association of representations, life of mood and
interest, work here in an interplay that can be difficult to unravel, but is not therefore
less significant. In folk-religions as well as in individual views of life this interplay
deposits its results. ``Metaphysics'' is in so far a fact, and there is then here room for
a comparative investigation of the origin, motives, and character of such involuntary
thrustings-together. Philosophical interest such metaphysics will have only when the
thrusting-together has taken place around a significant object, and when a firm bearing
rules throughout the whole train of thought. It is then a peculiar expression of psychical
energy, which can be typical of the human thought's position in existence. Otherwise where
the satisfaction of the metaphysical need leads to a sinking into a single side of
life,---and where it is merely the need for security, or perhaps even spiritual comfort,
that makes itself felt. The eyes then close to essential sides. It was this that
\emph{Goethe} reproached his pious friends (Lavater, Jacobi, Claudius) with: ``Dem
Lehrbegierigen bleibt wenig Trost bei ihnen.'' Such involuntary metaphysics gets a
dogmatic character. It does not know its own sources and often believes itself to be the
expression of ``sound common sense,'' or to be the voice of nature or of the deity.

Already on account of the involuntary tendency to metaphysical representations it will be
of significance to set up a special problem here. As a French philosopher has said: Hélas,
il est plus difficile qu'on ne l'imagine de ne point faire de métaphysique; la pire de
toutes est celle qui s'ignore.\footnote{G. \emph{Séailles}: \emph{Les affirmations de la
conscience moderne} p. 229.} Philosophy must test the spiritual powers that utter
themselves in such involuntary thought-activity, and seek to determine the possibilities
that by thought's nature and conditions must be said to be present. It is philosophy's
honor that it has distinguished between metaphysics and science, before the special
sciences, which up to the most recent time have confounded the two, have gotten an eye for
the difference. That philosophy cannot without more ado push metaphysics out comes from the
fact that the treatment of the problem of cognition establishes that the whole human
science, the most exact as the purely empirical, rests on presuppositions whose absolute
necessity cannot be grounded. Here is and remains an open place.

It will nevertheless be expedient to divide the consideration, so that we provisionally
investigate how far a world-view is possible that presupposes no other motives than the
purely intellectual interest that is all science's psychological motive, and that attaches
itself closely around thought's own laws, will only know what their consequence leads to.
In so far as a world-view is determined by the need to keep hope and courage up, or to get
air for the fundamental moods that the course of life awakens, it is most correct to regard
it under the problem of valuation.

If one will have another word than the for many frightening ``metaphysics,'' I propose the
word cosmology. This expression is used in older time partly in opposition to psychology,
partly in opposition to theology. The old metaphysics, as it was set in system by Chr.
\emph{Wolff}, comprehended, besides ontology (which corresponds to the doctrine of
categories and epistemology), the three parts: cosmology, psychology, and
theology.\footnote{Such a \emph{Metaphysics} we have in Danish by \emph{Jens Kraft}
(Copenhagen 1751--1753). See on this \emph{Danske Filosofer}. p. 19--20.} Here I will use
the word cosmology in the most comprehensive sense: world-view, which can also
etymologically be defended, since Kosmos means world and Logos thought or doctrine.

The main difficulty for such a cosmology will be that the cosmological principles, that is,
the fundamental thoughts that condition and bear the world-view, cannot without more ado be
fetched from any single domain of experience, any single group of objects, since it should
precisely be a matter of connecting all domains, all objects, into a whole and giving a
characterization of this.

An attempt to treat the problem of existence will have partly a formal, partly a real
character. Partly it must be asked what demands the concept that is to express existence
must satisfy, and whether it is possible to satisfy them. Partly it must be asked what
content one can give such a concept.

129. Formally the cosmological concepts spring out of the need for combining, in which we
have found mental life's nature, and which especially gets practical significance through
the criterion of reality. The completed combining would give us a concluded world-picture,
in which all objects find their place. But already the discussion of the problem of
cognition has taught us that an absolutely concluded world-picture will be impossible. New
objects constantly emerge, and new tasks with them. And in any case our world-picture,
if it were to be an object of complete cognition, would have to be held together and
compared with something different from itself; only then could it get its full determinacy
and its exhaustive characterization,---but \emph{if} there were something such, our
world-picture would not rightly bear its name: it would not be an absolute whole-picture.
---It is here indifferent whether one distinguishes between cosmology and theology, or not.
The contradictions become the same, whether we speak of ``the world'' or of
``God.''\footnote{Cf. my \emph{Religionsfilosofi}. \S\ 17--18.} The irrational announces
itself here just as in the problem of cognition. And perhaps there is here an inner
connection between the two problems. When cognition constantly ends in new seeking and
cannot be concluded, and when it is itself a part of existence, then already therein lies
that existence cannot be a concluded whole. And on the other side, what is irrational for
cognition, which indeed especially hangs together with the time-relation, could stem from
the fact that existence itself is not finished, is constantly placed in development, and
therefore does not offer such clear and rational forms as our research must presuppose.
The philosophical systems have been too sure that existence in itself was a closed system,
and that it was only our will and thought that constantly had to struggle to subsist and
to win through to harmony. There is moreover the possibility that existence has properties
or sides that do not reveal themselves along the lines along which alone our thought can
move. According to \emph{Spinoza} there are, of existence's infinitely many fundamental
forms (``attributes''), only two (spirit and matter) that are accessible to us. \emph{Wundt}
has here applied the comparison between the so-called imaginary numbers' [sideways] series
and the positive and negative numbers' [forward and backward] series.\footnote{Spinoza:
\emph{Ethica}, I, 11; II Ax. 3. On Wundt see my book \emph{Moderne Filosofer} p. 18.} The
formula $a+b\sqrt{-1}$ could be existence's formula, but in such a way that in our experience
$b$ was always o. Within the dimension of existence in which our thought can move, it strikes
on the irrational, the inexhaustible; but the other dimensions it would not strike on,
however rich a fullness of objects it commanded, and however consistently it worked through
these objects. In one of the more special problems we will come back to this possibility.

130. A real motive for speculation lies in the prominent role an object or a group of objects
plays or seems to play within our experience. There can be a purely intellectual interest in
getting connection into our world-view by laying an object at the ground for the understanding
of all the others, and by this way the need for formal unity can then be sought satisfied. The
different cosmological systems are just so many attempts to sound existence's depth by testing
how far it is possible to understand it from a single side in our experience. It is a series of
thought-experiments by which our most important thoughts' reach has been tested.

Every such attempt must, as \emph{Leibniz} first has seen, go the way of analogy. We are
indeed here, by help of the given objects, to form for ourselves a conception of the whole to
which they all belong. Even if we now disregard the whole-concept's irrationality, the
difficulty will arise that the whole itself is never given us as an object. It must thus be a
single object (or group of objects) that is to express the whole's character, and to which all
other objects would be referred. It is, as history shows, not merely philosophical systems, but
all religions have gone this way, whether they themselves are clear about it or not.

Analogy is now, as we have seen (74--75), a necessary and justified means of thought in science.
Science's history shows us how constantly one domain of experience is used to illuminate another.
But when an object is used to illuminate another, the analogy can always be tested through the
fruitfulness and durability of the working hypothesis it leads to. One can compare the analogy's
consequences with experience and find it confirmed or contested by the emergence of new objects.
But cosmological analogies cannot be tested in this way, since the whole that is to be expressed
by a single object being laid at the ground never itself appears as an object, as little as it
ever lies before us concluded. We are quite otherwise placed than in an analogy between parts of
our experience mutually.

A world-view's character and significance rests partly on how clearly and consistently the
underlying analogy is carried through, partly on the object that is laid at the ground. This
object we can, with an expression fetched from \emph{Goethe}, call the primal phenomenon. Yet
Goethe understood by ``primal phenomenon'' not merely an object that could serve to illuminate
other objects for us. It means for him a fact that is at the same time law, so that one merely
needs to utter it in order to have the explanation of it. It must moreover not be regarded as
composite; it is a ``limit for our beholding''; it awakens ``the great wonder,'' but designates
at the same time our cognition's limit.\footnote{\emph{Goethe}'s \emph{Farbenlehre} \S\ 124;
175--177. \emph{Gespräche mit Eckermann}, 13. and 18. Feb. 1829; 21. Dec. 1831. Goethe applied
this concept not merely in the doctrine of colors, but also in the doctrine of magnetism and in
biology. \emph{Sprüche in Prosa (über Naturwissenschaft)}. \emph{Gespräche mit Eckermann}. 27.
Jan. 1830. \emph{Unterhaltungen mit Kanzler v. Müller}. p. 39. ``Ein Urphänomen muss man nicht
weiter erklären wollen; Gott selbst weiss nicht mehr davon als ich.'' --- \emph{Schopenhauer}
applied the Goethean expression to every interpretation of ``the thing in itself,'' every attempt
to reach out beyond the world of phenomena. He himself saw in the will (this concept taken in the
widest sense) the primal phenomenon, ``the most immediate manifestation'' of the thing in itself.
Cf. \emph{Den nyere Filosofis Historie}.\textsuperscript{2} II. p. 220--222.} It is known how
fateful a role this concept played in Goethe's doctrine of colors (where it found its expression
in ``Gesetz der Trübe''); the primal phenomenon---shadows that on dark ground showed themselves
blue, on light ground yellow-red---could and must not be more closely investigated and explained!
But so dogmatically the concept primal phenomenon need not be taken. It can be used to designate
the point in our experience from which we try to orient ourselves in all directions, but which
therefore may very well itself be an object of further-going investigation and explanation. In
epistemology we have seen that although the cognizing subject stands as a presupposition that
must be understood in any cognition whatever, it can therefore nevertheless very well itself be an
object of psychological investigation. There is not any absolutely fixed point in the world of
spirit, as little as in the world of matter. Whatever significance a cosmological attempt may
otherwise have, it is the first demand that must be made of it, that it not appear dogmatically by
pushing all epistemological principles aside.

131. Every attempt to develop a world-view must rest on clarity about its moving on science's
boundary. It must not intervene in science's daily and special work, in that of the mental
sciences as in that of natural science. The primal phenomenon must itself share all phenomena's,
all objects' lot. The very fact that existence, as we know it, is to a great extent accessible to
understanding, shows itself subject to general laws, must in every world-view become of essential
significance, whether this very fact is made into primal phenomenon or not. Even if we would
regard all principles and laws as mere working hypotheses, it nevertheless serves to characterize
existence that one, as regards understanding of it and dominion over it, can only get at it along
certain definite ways. Every, even the most special branch of natural science or mental science,
thereby gets cosmological significance. It is indeed only most indirectly that existence makes
itself known in the working hypotheses; but a connection there must in any case be, and a
cosmology that would quite scorn the special sciences' views and results would contradict itself.
While \emph{Schelling} and \emph{Hegel} conceived the special sciences as imperfect approaches to
cosmology and boldly would set their systems in place of those sciences' imperfect philosophy,
their modern adherents often grasp the expedient of proclaiming the special sciences' independence,
but at the same time declaring them to be, for philosophy as such, quite
indifferent.\footnote{Thus \emph{Bradley} (see \emph{Moderne Filosofer}. p. 48,) and
\emph{Benedetto Croce}. Croce reproaches Hegel with wanting to continue Schelling's philosophy of
nature instead of proclaiming ``l'indifferenza filosofica delle discipline naturale matematiche e
la loro piena autonomia'' and referring these sciences ``al non --- teoretico, e cioè al pratico.''
(\emph{Ciò che è vivo e ciò che è morto nella filosofia del Hegel}. 1907. p. 149; 155).} But the
difficulty for every attempt lies precisely in the fact that a whole-conception can only be built
up on the ground of points of view and tendencies that already make themselves known within the
special sciences.

The work on a world-view takes place on science's boundary. But even there one must work in
science's spirit. A significant world-view is now only formed when at the ground there lies a great
enthusiasm or at least being-seized by the thought of existence regarded as a whole, especially by
a single object within it. Here it is then the task to unite being-seized and critique. It is a
matter at once of holding fast the interest for the ideas by whose help one tries to cast light
over existence, and of applying them with full consciousness of their problematic character. There
belongs great psychical energy to working on these conditions, great elasticity in mind and thought
to swing between the deep conviction a great idea can call forth, and the firm conviction that
continued discussion is constantly necessary. Fantasy, with its capacity to form whole-pictures, has
its great significance in science as in art. But in science such whole-pictures---whether they are to
hold only for single object-domains or span all---can only stand as hypotheses, as stations on a way
that can always be continued. In art, on the contrary, a definitive goal is reached when a full and
characteristic picture has been created.\footnote{Cf. my \emph{Psykologi}. V B, 12.}

When I in my time have spoken of philosophy as art,\footnote{\emph{Mindre Arbejder}. First series.
p. 184--186.} I thereby thought of that union of opposite properties which every attempt at a
scientific world-view requires: a world-picture is to be brought forth with exertion of energy and
in enthusiasm---and yet constantly stand as a hypothesis. From other points of view the artistic
element in philosophy's concluding ideas has earlier been brought forward. \emph{Schopenhauer}\footnote{\emph{Den
nyere Filosofis Historie}.\textsuperscript{2} II. p. 215 f.} thought most of the involuntary
coming-to-be of a concluding thought-construction, whereby art's ``What?'' relieves science's
``Why?'' \emph{Albert Lange}\footnote{Op. cit. p. 550.} thought nearest of the idealizing tendency,
the need to find in ideal pictures expression for the highest reality.

The transition from philosophy's empirical and critical parts (psychology and epistemology) to
cosmology offers an analogy to the transition from historical critique to history-writing. The
history-writer unites the critically treated material into a whole-picture of events and characters.
There is thereby carried out a rounding-off, in which the researcher's personality involuntarily
plays a role along. Thus the philosopher tries to work together the critically worked objects into a
whole, within which a single object determines the timbre. The point of view (the primal phenomenon)
and its carrying-through here unavoidably get a personal stamp. A great philosophical system is a
work of art, a drama, in which the single lines and scenes join together, now more with logical, now
more with aesthetic harmony. \emph{Plato}'s, \emph{Spinoza}'s, and \emph{Schopenhauer}'s systems
especially offer this stamp.

In any case an attempt at a world-view stands constantly between science and art. If such attempts
have a future, there will perhaps here develop itself a type of spirit that on account of the
hitherto ruling dogmatism has not yet been able to unfold itself properly. \emph{Georg Simmel} has
from this point of view denied that metaphysics (cosmology) should be called art: ``So poor in
independent fundamental forms our soul is not, that every picture of things that is not science
should be only art.''\footnote{\emph{Schopenhauer und Nietzsche --- Ein Vortragscyclus}. (1907) p.
128 f.} I would oppose this ``only''; it could indeed be that the artistic form was the highest, when
it rested on a great experience and a serious labor of thought. But the main thing is that the
endeavors that constantly must be made in cosmological direction get freedom to stir themselves, in
order that their significance for life's different domains can become established. Such a significance
they will moreover always have, when they really provide intellectual satisfaction for highly
developed personalities.

132. In a very sharp and interesting way the cosmological problem is posed in the now-living
philosopher who most directly has taken aim at a solution of this problem, \emph{Henri Bergson}. It
has already earlier (3) been mentioned how sharp an opposition this thinker assumes between the
immediately given and reflection or analysis. Science's work is essentially analytic. Out of the
immediately given's continuity are sorted the elements, parts, and properties, which later can only
in an external way be put together again. It is the practical need, technique, and language, that
necessitate these distinctions. And then especially language, which predominantly applies
space-pictures and therefore really can only express purely external relations. What in immediate
experience is connected and at once appears as unity and multiplicity, is now made into a mosaic
without real inner unity. If a metaphysics is to be possible, we must by an exertion of will return
from analysis to intuition, look at the objects, not with ``the understanding's'' eyes (les yeux de
la seule intelligence), which only discover the outer, but with ``the spirit'' (l'esprit), which is
one with intuition and with life itself. ``The spirit'' is the will's bending toward itself instead
of the bending outward toward action in the surrounding world. Here then existence (l'Etre,
l'Absolu) is grasped directly, not by the detour of logical and mathematical abstractions. Through
the intellectual sympathy that is one with intuition, the moving force in reality is grasped and
followed through all its turns. Only thereby is time's reality maintained over against the mechanical
crystallization of the moments. The mechanical causal relations are seen as sensuous symptoms of the
inner stream that is the true real, and it then becomes possible to follow life under its continuous
development, follow it under its constant striving, in spite of the material elements' resistance, to
break itself a way to higher forms that cannot be foreseen. The spirit or intuition, as it appears in
limited form in the human being, is a stage in the great process. It is not a product of outer causes,
but an unfolding of life's and existence's inner process (l'élan de la vie), which only at a certain
definite step becomes able to grasp itself in a beholding, and thereupon by ``intellectual sympathy''
interprets all the other currents in existence, all other striving under, around, and over us.

Bergson laments that metaphysics must use the understanding's language, because only the understanding
has a language. From this is understood the application of the many pictures in his brilliant
presentations. When the question is raised what means of thought metaphysics, on Bergson's conception,
really commands, we get no clear answer. This is understandable, since the understanding has laid claim
to language, and since philosophy (metaphysics) according to Bergson only comes to be by our going in
the opposite direction of thought's accustomed movement: Philosopher consiste à invertir la direction
habituelle de la vie de la pensée --- remonter la pente naturelle de l'intelligence!

By this definition of philosophy\footnote{\emph{Introduction à la Métaphysique}. (Revue de
Métaphysique et de Morale 1903). p. 27. --- \emph{Évolution créatrice} (1907). p. 32. --- On Carlyle
cf. \emph{Den nyere Filosofis Historie}.\textsuperscript{2} II. p. 378. (Sartor Resartus I, 10.)}
Bergson recalls \emph{Carlyle}, for whom philosophy also was a struggle against habit. It has its
justification over against dogmatic application of special sciences' methods to solve existence's
fundamental problems. But one thing is that we can see that our usual means of thought here do not
suffice; another is what means of thought we then command with the last great problem. Already above I
found the difficulty of the cosmological problem in the fact that it was here the objects' aggregate,
existence, so far as it could be regarded as whole, that was to be cognized;---but existence is not
given as object, and we must therefore use a single object to illuminate all others by the way of
analogy. If this conception is right, it will be of interest to see whether the spirited French thinker
would be able to attach himself to it.

Bergson applies so great a labor to critique of language's and technique's external and mechanical
representations and expressions, that he does not come to give us the new logic or doctrine of method
that his denial of ``the understanding'' and his appeal to ``intuition'' or ``the spirit'' really
require. It stands especially not clear what relation there is between the intuition to which we---when
we with exertion of will have worked ourselves out of the strong habit of parceling-out---are to return,
and the analysis in which the scientific work consists. And yet he tries in his last work
(\emph{Évolution créatrice}) to develop a kind of speculative vitalism, in that he will show how the
idea of life as a self-working-forward upswing (élan) is able to gather series of facts about itself.
But whence comes this idea, and on what does it ground itself, since it is not to be won by the way of
analysis?

In seeking the answer to this question in Bergson's presentation I find the hypothesis of all
metaphysics as grounded on analogy confirmed. In his \emph{Introduction à la Métaphysique} (p. 25 f.)
it runs, namely: La conscience que nous avons de notre propre personne, dans son continuel écoulement,
nous introduit à l'intérieur d'une réalité \emph{sur le modèle de laquelle} vous devons vous
représenter les autres. Toute réalité est donc tendance.---Although the word analogy is not used, the
concept is nevertheless clearly applied. Likewise there is here clearly indicated the object that lies
at the ground for analogizing. From the psychological domain a bridge is struck, through the analogy,
which leads in to all objects' interior. Only by this way can Bergson come to the result that ``the
absolute is of psychological, not of logical or mathematical nature'' (\emph{Evol. créatr.} p. 323).
Like all idealism, Bergson's too is grounded on the analogy with the human being's mental life.

The artistic trait in Bergson's presentation lies especially in the way in which he brings ``the
immediately given'' and analysis into the sharpest opposition to each other and thereby forces us to
the leap with which the metaphysical speculation, according to his theory, must begin. The whole way
that leads to the catastrophe is depicted in the most spirited way. There is only the transition or
leap itself that remains standing as a riddle. He has avoided the schematic, especially the striving
after symmetry in the arrangement of concepts, which according to Poul Møller is affectation. He points
energetically to the conditions under which we can reach out beyond all the piecemeal in the scientific
conception of existence. But even after we have followed this pointing-finger, the opposition to the
scientific analysis nevertheless remains standing for us, and there is opened no way out either to do
without it or to use it in the service of a concluding world-view. ---

After having found our hypothesis about cosmology's nature and mode of procedure confirmed by the last
attempt made in this direction, we now go over to a closer consideration of the different primal
phenomena that can be laid at the ground.

\subsection*{b. Unity and Multiplicity}
\addcontentsline{toc}{subsection}{b. Unity and Multiplicity}

133. For the philosopher it lies near to take the point of departure for his world-view in the
very fact that it is possible to orient oneself in existence by cognition's help. The primal
phenomenon then becomes intelligibility itself, as it is from thought's side anticipated in the
categories of rationality and of causality. There cannot subsist any absolute relation of
indifference between our thought's mode of working and the source of the objects that are worked
by it. When we can recognize objects,---when hypotheses formed on the ground of the past's objects
show themselves to hold for later-appearing objects,---and when with advancing experience there
shows itself an ever greater continuity in the world of qualities and of changes,---then in this
lies a witness to existence's, not merely to our thought's, nature. Our very mark of reality
consists indeed in the unity and connection that shows itself between our observations, those now
lying before us and those earlier made. It is by reason of the applicability of this mark that we
can distinguish between what are mere thoughts, and what is a reality we cannot get around, even if
we push our thoughts aside, but which we can perhaps master when we know its laws. The concept of
reality arises partly by a limitation, partly by an extension of our thoughts (110; 126).

The connection between thought and existence is, according to the dynamic concept of truth, only
indirect. But nevertheless it can, according to what has just been developed, with right be said
that existence reveals itself through thought's work. Way and goal cannot here fall quite outside
each other, especially when one remembers that detours and errors can often be more fruitful in
truth's service than the straight ways. The criterion of reality teaches that it is not merely a
matter of gathering as many objects as possible, but also of thinking them through with
consistency. Error rests on a limited thing, a part, being made into a whole; but when one then
merely consistently thinks through this limited thing, one will come to see where its limit lies,
just as one will soonest find one's way out of a forest when one follows one definite direction
instead of constantly changing direction. Intellectio fictionem terminat (Spinoza). This, that
existence's nature can be present in the midst of error, shows perhaps most clearly of all that
here is a kinship, however near or far it may be. He who seeks rightly has already found. In so far
\emph{Lessing}'s alternative (seeking of truth---and possession of truth) does not hold, and
\emph{Spinoza} and \emph{Hegel} were right in their mystically sounding proposition that in that
we think existence, existence thinks itself in us. (Cf. 110.)

The innermost nerve in the work thought has to do is the application of the relation between ground
and consequent to the changing and manifold objects. Causality becomes primal phenomenon. Its
applicability---in how great or small an extent it holds---is a fact of the greatest significance
for every world-view that will in earnest distinguish between dream and reality. It is no wonder
that the essence and validity of the concept of cause has played so great a role in the more recent
philosophy's discussions.

Mythologies and theologies too build on the concept of cause. But they apply it chiefly to make an
ascent to a ``first cause''---or to several first causes, which are then gradually ordered in a
hierarchy. One acknowledges perhaps subordinate causes (causæ secundæ), but yearns out beyond all
finite causes and seeks at any price to conclude the causal series and to hang the whole chain of
causes and effects up at an absolute point of departure or end-point, on a secure hook. The causal
principle does not permit this. It demands to know in which wall the hook sits. It requires a cause
for the first member in the causal series as well as for each single one of the others. If we are
forced to conclude, we end with an unanswered question. How far we can actually follow the causal
series back is a question that at each given time is answered by empirical science. For the
philosopher it is not this endless series that is the main thing, however much interest it may in
itself offer; the essential thing is for him the thought according to which there must constantly be
sought, and can be found, new members in the causal series. He finds here a world-order expressed,
which encloses all that exists in the most intimate way,---a unity-power that connects all objects,
however peculiar and independent they each for themselves may be, so as to stand as members in a
great, immeasurable connection. It was this power that \emph{Spinoza} in a mythological way
conceived as a being (``an eternal and infinite being'') and called substance, that which subsists.
What subsists is the great unity in existence, which makes itself known in the law-governed
connection. This idea is merely a carrying-further and carrying-through of the simple thought all
involuntarily apply in daily life, when they seek to distinguish between the apparent and the real.
The ``sound common sense's'' daily practice here meets with the highest metaphysics.

134. There are constantly members lacking within the causal series, and this can be thought
continued in both its directions. The unity-picture it is possible to form stands constantly as a
great hypothesis, as a resting-point where the synthesis has provisionally done its work, and where
there are perhaps means lacking to come further. New objects can arise or be discovered, so that new
series-formations become necessary. Ever anew thought must ascend from provisionally chaotic
difference-series to rational series. The word ``absolute'' has its scientific significance within
the single rational series; it expresses the relation between premise and conclusion, ground and
consequent, but cannot be used to designate a conclusion of all thought-connection. ``The absolute''
is a mythological remnant, in \emph{Hegel} as in \emph{Spinoza}, a sign that metaphysicians too can
become tired and feel a need to rest out in a last thought.

Even if conclusion is not possible, the world-view can nevertheless bear unity's stamp, and every
world-view must, whether it itself lays weight on it or not. Monism is therefore invincible, even
though it constantly must struggle with the advancing and manifold experience and therefore must be
designated as \emph{critical monism}. It is by virtue of synthesis's own law that every unity-view we
can form for ourselves of existence points out beyond itself: every object is set in relation to
something else, something different from itself, since only thereby can it get its full
characterization. (Cf. 58--62.) And experience then also constantly gives occasion for satisfaction
of this need, in that new objects emerge and pose their demands, which often require reshaping or
extension of the hitherto developed view. It is no imperfection in our thought that it thus
constantly seeks to set one object in relation to another. It is precisely a perfection that we
cannot stop at any point. We constantly seek anew to get out beyond the limitation (both in content
and in extent) that every finished world-picture offers. While one has hastened to form concluding
``world'' or ``God'' as secure knots on the thought-thread,\footnote{Cf. \emph{Religionsfilosofi}.
\S\ 17--18, cf. \S\ 52--55.} it can precisely be called a spark of divinity in the human thought,
that no such finished concept can satisfy it.

There shows itself here, within the problem of existence, an antinomy that forms an analogy to the
antinomies we in different forms have met in thought's psychology, in its history, in the doctrine of
categories, and in the problem of cognition. It is the conflicting relation between continuity and
discontinuity that under the different forms makes itself felt. For our world-view the consequence
becomes a constant conflict of unity and multiplicity, without a complete harmony's being here
possible. It is no use, with Hegel,\footnote{\emph{Wissenschaft der Logik}. (1812--16.) II. p. 375
f.} to distinguish between the seeking cognition and the true cognition. In the seeking cognition,
says Hegel, there is difference between subject, method, and object; but in the true cognition these
differences fall away. The question is precisely whether thought can ever cease to be seeking. Hegel's
own system has fallen to the dialectic he himself constantly demanded as truth's medium, and which
utters itself in that every thought-content is set over against another and thereby leads out beyond
itself. Thus it has gone with Hegel's system in the face of new objects and new points of view (also
in the face of certain old objects and points of view!). We should, according to the Hegelian
dialectic, by this way constantly come to ``higher unities.'' Of this we now cannot be so sure (cf.
99). We could indeed provisionally move in a downward direction; the development of truth can follow
winding ways, and we must be glad if we can hold fast the analogy with the spiral, which bends only in
order to rise: inclinata resurget!

135. But when the lack of conclusion, which every world-picture thus must offer, has its ground in
the fact that the relation between unity and multiplicity constantly makes itself felt in new
forms,---with what right is then greater weight laid on unity than on multiplicity? Why has one the
right to sympathize more with Spinoza's one ``substance'' than with Leibniz's countless ``monads''?
When there are two sides in our cognition that cannot be done without either in our science or in our
world-view, why then regard the one of them as the more significant and as the one that shall
determine the timbre in our world-view? Should it not be just as justified to embrace a pluralism as
to embrace a monism?

\emph{Charles Renouvier} and \emph{William James} have in our days with energy taken up the word for a
pluralistic world-view. Renouvier protests, on the ground of his neocriticism, against what he calls
``la superstition de l'unité''; James, on the contrary, will build his pluralism on empirical ground.

Pluralism, that is, the world-view for which the objects' multiplicity and difference is the primal
phenomenon, can seek its justification in the fact that difference, opposition, and change is the
condition for all sensation and thought. (Cf. 11; 21.) There is surely also, under philosophy's
present condition, good ground to make pluralistic points of view felt. There is still constantly too
much speculative metaphysics in the air, whether it now appears in philosophical or in theological
form. It is of significance that we not forget what positivism and criticism have taught us. The old
English school had the mission of holding philosophers' attention fixed on experience, and it awakened
the great movement against dogmatism in the two last centuries. It is no accident that the great
poser-of-problems \emph{David Hume} belonged to this school. Pluralism makes the world new for us, and
it forces us, where it appears with authority, to revise our categories, our principles, and our
methods. We are too inclined to dogmatic sleep, to believing that we possess thoughts in which all
existence can find expression.

A ``radical'' pluralism is nevertheless untenable. If multiplicity were sole ruler, no understanding
would be possible. For to understand is to find an inner connection between different objects, find a
unity-principle for them. We must indeed very often content ourselves with establishing the reality of
single, isolated objects each for itself. But in that case a problem arises by reason of the isolation
itself. That the isolation is unconditioned, observation can in any case not teach us. As several times
remarked, the assertion that a chaotic difference-series lies before us requires a long investigation.
Pluralism can just as well appear dogmatically as monism, though it does not offer such rest as the
mystical unity has yielded to many a tired thought.

The single objects are constantly determined and characterized through their connection with other
objects within a comprehensive world-order. The single object's place in time and in space is
determined by its relation to other objects within the world of space and time that is accessible to
us. Every property we ascribe to the single object expresses the way it affects other objects or is
affected by them. The force or energy the single object is said to possess is determined by a law
according to which its changes bring with them changes in other objects. The single object's
individuality is expressed in the law according to which its states and its properties change.

We do not know the objects outside a connection of one of these kinds. It is constantly due to an
analysis, a separating and distinguishing, when we speak of different objects; we have then already
taken them out of ``the immediately given's'' continuous connection. Every analysis rests on an
original, primitive, involuntary synthesis, which conditions the whole that by analysis can be
separated into its elements. Where does the one object end, and where does the other begin? Even if
the connection is only ``conjunctive'' (to use James's expression), and can only be designated by the
word ``and,'' it nevertheless encloses problems: have we drawn the boundary-line rightly? and even if
we have, what degree and kind of separation does then this ``and'' mean? Analogously it stands when
judgments are formed on the ground of intuitions, and inferences of judgments. Every act of cognition
means a section of a greater whole. Every conceptual determination is a limitation, every judgment
rests on presuppositions that point out beyond the judgment-act itself; every inference presupposes
several premises fetched from a greater, ramified connection. We move in our cognition constantly
within a whole that comprehends more than the objects we think over, and that contains the conditions
for the results to which our reflection comes.

For these grounds monism must be said to have greater right than pluralism to characterize our
world-view. It supports itself on that in existence which makes it possible for our reflection at all
to be directed in a fruitful way toward the objects it contains. As we have seen, monism subsists only
under a constant struggle with multiplicity. But our world-view rightly bears its name after that one
of the struggling parties on whose side we place ourselves. \emph{Renouvier}, whose main category was
relation (see 56), was---in his maintenance of pluralism---really only concerned to hold fast the
constant relation of unity and multiplicity (la correlation de l'un et du plusieurs), which thought
cannot do without. For \emph{James} ``the radical pluralism'' hung together from the first with ``the
philosophy of pure experience,''\footnote{\emph{A World of Pure Experience}. (Journal of Psychology,
Philosophy and Scientific Method. Oct. 1904. p. 570).} while he later has declared that pragmatism,
which is to be a radical expression of the philosophy of experience, must in equal degree reject the
absolute monism and the absolute pluralism, and he now designates his standpoint as ``noëtic
Pluralism,'' that is, a critical doctrine of plurality that lays weight on the fact that there are many
objects between which no necessary connection can be demonstrated.\footnote{\emph{Pragmatism. A new name
for some old ways of thinking}. (1907.) p. 51; 155; 166.} This conception seems not to be in principle
different from the critical monism presented above; only center of gravity or emphasis are
different---but it is indeed a difference that can become decisive for the world-view's timbre. When
James calls his pluralism a method, he must thereby mean that it especially aims at maintaining and
securing the single objects in their peculiarity and independence over against the tendency to let them
disappear in a mystical unity, which easily becomes the night in which all cats are gray. This is an
endeavor with which the critical monism can fully sympathize.

Monism and pluralism are in no way necessary oppositions, when the unity-principle on which monism lays
the weight is conceived in such a way that it precisely reveals itself through the single objects
(individuals, events, relations) as well as through the law-governed world-order that bears and
conditions them. And in this direction went \emph{Spinoza}'s, the most pronounced modern monist's,
thoughts. For him it stood thus, that the unity in existence could only be expressed through an infinity
of single beings; in the single being's capacity and need to remain in its state and maintain its
existence he saw a peculiar form of the unity-principle in existence. There is thus for him no external
opposition between the single thing and the whole: in the single individual existence's innermost
direction and striving there made itself known precisely existence's fundamental principle. Spinoza's
real fundamental thought surely appears at this point. When his system does not manage to bring this
point into full and consistent connection with other points, the main ground lies in the fact that he
believed he could get all the irrational out of the way and with right found its last source in the
time-relation. If the last principle in our world-view is timeless, the single existences, which develop
successively, cannot find a natural place within it. If, on the contrary, one makes earnest of Spinoza's
proposition that the more we understand the single beings (res singulares), the more we understand
existence's fundamental principle,\footnote{Qvo magis res singulares intelligimus eo magis Deum
intelligimus. \emph{Ethica}. V, 24.} then we must also take the consequence, which would have been to him
an offense,---that existence's innermost essence, so far as we can at all form a hypothesis about it,
must be thought to unfold itself in time. But in the same moment our monism becomes not absolute or
dogmatic, but critical.

136. Monism's strength is the fact of understanding. Its weak side rests on the fact that there are
constantly limits, hitherto unsolved tasks. Empiricists and skeptics will constantly be able to stop the
monist by pointing to these barriers. Not even the criterion of reality can we apply everywhere, thus not
carry through a strict opposition between dream and reality. The world-views will perhaps constantly
conflict at this point. But it will not be any unfruitful conflict, since it will sharpen the interest in
finding unity and connection as regards as many objects as possible. The monist's cause stands and falls
with the possibility of cognition's advance. And he will perhaps with irony leave to the pluralist the
chaos that at each time seems to remain, with the reservation that it cannot be established whether it is
a real chaos, but that its time for clearing-up will surely come---then perhaps to be relieved by new
objects' apparent chaos.

But the possibility is not excluded that the ground why full understanding is not possible is to be
sought not merely in the fact that the cognizing subjects are placed in development, but finally in the
fact that existence itself is not finished, but placed in becoming and development, just as cognition is.
Concluded cognition would only be possible if existence were concluded. The time-form, the source of all
irrationality, might be not merely one of our categories, but an existence-form that we have no right to
think away.

If now something wholly new arises in existence, so that the later real moments are not quite as the
earlier,---in Kant's language: if ``Ding an sich'' does not work constantly,---then it is no wonder that
our cognition constantly has to struggle with difficulties and perhaps finally with insoluble riddles.
And there is in any case a point in existence where something new arises, namely there where
understanding develops out of ignorance, error, or doubt. Existence indeed comprehends cognition too;
both forms, objects, and values belong to it;---it cannot then be finished so long as the worlds of forms,
of objects, and of values are concluded and brought into equilibrium mutually. There is a cosmic
development that takes place through the human thought,---perhaps also, as will later be discussed, through
human will. It was the great rationalistic systems' (Plato's, Spinoza's, Hegel's) wonderful contradiction
that even though they could explain everything else, they nevertheless could not explain the very striving
and working thought whose works the systems themselves were.

The critical monism, which holds fast the time-relation's validity and thereby existence's
unconcludability as well as cognition's, nevertheless finds a primal phenomenon in the fact that more and
more objects are brought under the causal principle,---the analogy between ground and consequent on the
one side, the series of changes on the other side, constantly anew shows itself to have real
applicability. Thereby is grounded a belief in an advancing unity-power in existence, which leads out
beyond the sporadic and conflicting. The labor of thought is seen in cosmic illumination. There occurs a
cosmic advance when thought advances---however it may then stand with other objects within the immeasurable
existence.

\subsection*{c. Spirit and Matter}
\addcontentsline{toc}{subsection}{c. Spirit and Matter}

137. The first object we investigated in the capacity of primal phenomenon was
accessibility to thought. This investigation led to seeing the possibility of a critical
monism. Should there now not be objects of a more concrete kind that could lay claim to
being regarded as primal phenomena? The world-view would thereby get a more definite
character.

Reflection could then turn toward the fact that all we experience is either of a spiritual
or of a material kind. All objects fall into two great main groups---the one comprehending
sensations, representations, feelings of pleasure and displeasure, striving and willing, the
other comprehending all that is extended and movable. It is two fundamental qualities that
here stand over against each other. Experience shows them to us at once in their difference
and in their close belonging-together. The question becomes whether one of them, and if so
which, is suited to be primal phenomenon and thus can determine our world-view's timbre.
Since our experience offers no more main kinds of objects than these two, there seems to be a
bound choice between them.

In the section on thought's psychology the relation between the spiritual and the material
was already discussed, but essentially from a psychological, empirical, and methodological
point of view. It was especially a matter of finding a view that at once expressed what we
really know about this relation, and what was suited to be laid at the ground as a working
hypothesis in continued research. What we really know is that there subsists such a relation
between the two kinds of objects that inferences can be drawn from the one to the other,---that
there thus subsists a relation of function in the mathematical sense between them. On the other
hand it was doubtful whether there also subsists a time-relation between them, with conversions
from physical energy to psychical, and conversely. But that relation of function is now also
what first and foremost is of use in continued research, so that it is suited to express the
relation between the objects psychology occupies itself with, and those physiology occupies
itself with.

When we now return to this question, it is to discuss how far our world-view is forced to stop
at this provisional result, or whether it is possible to go a step further in the one or the
other direction. Whether we methodically and empirically are dualists, materialists, or
Spinozists, it can be asked whether in either of the two groups of objects we find existence's
innermost essence expressed, in such a way that the other stands only as symptom or symbol for
it.

Every attempt to answer this question must support itself on analogy, when it is given that one
can neither derive the spiritual from the material, nor conversely. The question then becomes
whether we conceive the spiritual in analogy with the material, or the material in analogy with
the spiritual,---or whether perhaps neither of these analogies can be carried through.

138. If I lay the analogy with what is extended and movable in space at the ground, my
cosmology, not merely my method, becomes materialistic. Materialism is the human being's natural
philosophy. It recommends itself by its perspicuity: space is indeed the clearest form of
apprehension we have, wherefore there is (especially in visual individuals) a need to convert all
thoughts into space-pictures. Language witnesses to this, in that all expressions for spiritual
objects are transferred from the material, often by a very characteristic symbolism.
Communications from one human being to another about mental experiences get a clearer character
when they can make use of space-pictures, express mental life's ``inner'' suffering and working
through pictures of positions and movements in space. Even time we make perspicuous only by
space's help. Confusions of identity and analogy lie then here extraordinarily near, and it is on
such a confusion that the cosmological materialism as a rule rests.

Even after the difference between the spiritual and the material has been acknowledged, there can
be grounds that speak in favor of regarding the material as the primal phenomenon and thus
letting it occupy the fundamental place in the world-view.---The spiritual shows itself indeed,
when we strictly hold to experience, only to appear in some earthly organisms, while the material
world spreads itself in immeasurable regions. Can then so vanishing an object with right stand as
expression for existence's innermost essence---in opposition to the vast masses?---And if we agree
to regard the material as the primal phenomenon, we attain a greater continuity in our world-view
than we would otherwise be able to get. The spiritual arises sporadically---in different
individuals and often also in the same individual: there is no medium for the different
consciousnesses mutually, such as there is for the different bodies mutually,---and in the same
individual the life of consciousness is often interrupted by unconscious intervals. On the other
hand the material objects offer, the more we penetrate into them, a great continuity. Space's form
makes it possible to follow the material changes down into the vanishingly small, carry the same
laws down through what for observation stands as an interruption. The world of spirit stands as a
fragment, or as a collection of fragments, while the material world offers connections to infinity.

Is there not all ground to conceive the non-perspicuous in analogy with the perspicuous, the small
in analogy with the great, the fragmentary in analogy with the connected?

However energetically such grounds are made felt, they will nevertheless again and again be
outweighed by the thought that is all idealism's fundamental thought, that we know matter only
through spiritual activity, through the energy that utters itself in sensation and thinking, and
which therefore naturally becomes our last refuge. The very concept matter, as natural science
determines it in ever more ``ethereal'' forms,\footnote{\emph{Paul Langevin} (\emph{Physics of the
Electron}. Congress of Arts and Science. 1904. IV. p. 137.) holds that the ether must now be
regarded as ground for the concept of matter. The electronic theory of matter, in which matter is
partly synonymous with electricity in motion, explains an ever-rising number of phenomena. --- See
also \emph{Christiansen}'s Program (1905). passim.} is a thought-product, built on observations and
experiments. And even if one then again would assert that the thought that has produced this
concept is itself again a product of matter, then there follows from idealism's side as rejoinder
that this very hypothesis about thought's arising (be it now tenable or not) is also a
thought-product. We meet here the earlier (125) treated relation between subject and object in a
special form (for there are other objects than the material). If $T$ means thought, $M$ matter,
then the series $T < M \mathrel{\{} T \ldots$ is a special case of the series $S_1 < O \mathrel{\{}
S_2 \ldots$. The single combat can in a certain sense not get an end. But when we ask about the
last ground for all assumptions, it is nevertheless thought, the only object in and for which
grounds at all exist, that we come back to. It was therefore that the last proposition in my
``Psychology'' came to run thus: ``Further back than to what for the human being stands as a
thought-necessity we cannot come; the thought that explains everything becomes its last and
constant problem.'' Even if what is extended and movable in space occupies a predominant place
among our objects, both by reason of its extent, its continuity, and its perspicuity, it can
nevertheless only in and for thought make itself felt. It is an object that constantly bears the
stamp of the subject's work (an $O_s$).

But thought has other objects than the material. It works also in the world of spirit, in the
study of mental life in the thinking subject itself and in other subjects, which it conceives in
analogy with itself. The mental science makes its claim felt with the same right in the world-view's
coming-to-be as does natural science. That its objects do not fill space in the perspicuous
world-picture, as natural science's do, is not the decisive thing. There could indeed in a corner
of the universe have arisen something that was of far greater significance for the deepest
understanding of existence than all the much that moves in space. And if one lays the weight on the
fact that thought seems to appear at a very late step in the world-development, then this could
indeed be assumed to indicate that the most essential qualities need the longest time to unfold
themselves properly.

By space---says \emph{Pascal}---the universe comprehends (comprend) me as a vanishing point; but by
thought I comprehend (comprends) the universe.

139. Instead of conceiving the spiritual as a form of the material, idealism conversely conceives
the material as the form under which existence's content, which in itself is of a spiritual kind,
presents itself to the sensuous intuition. The justification or necessity of taking one's point of
departure from the spiritual or making this into primal phenomenon is grounded in different ways.

When materialism emphasizes the continuity the material world offers, it is answered to this that
this continuity is external and mechanical, as space's form brings with it. A true continuity is
only possible in thought, which connects the single elements in its unity. The most intimate unity
that can at all be thought is the one that comes to light in that thought advances from ground to
consequent. In the material world there are found only outer copies of this inner unity, by whose
help we understand everything. We cannot connect any meaning with an existence that does not bear
such a unity's stamp. If existence is to constitute a whole, we can only think it as a
thought-whole. This was \emph{Hegel}'s solution of the world-riddle. Thought and existence are for
him identical, because thought's form is the only one in which existence can be conceived as a
world.---That it is a great analogy he here operates with,---an extension of a single object's
peculiarity to being expression for all existence,---this Hegel does not admit.\footnote{Cf. also
\emph{Brøchner}'s standpoint. \emph{Danske Filosofer}. p. 202--204. --- \emph{Wundt} too stands on
Hegel's standpoint here. Neither does he admit that he analogizes. He maintains that he---led by
thought's need for unity in the world-view---merely continues and supplements what empirical
cognition gives. \emph{Moderne Filosofer}. p. 21--24.}

\emph{Leibniz}, the metaphysical idealism's real founder, was on the other hand fully conscious that
he built on analogy. It was by this's help that he conceived existence as a continuum of monads, in
that these represent infinitely many different degrees of the same thing that in the human being
appears as mental life. The fragmentary character of the world of spirit is only apparent. Nature
makes no leaps. Unconsciousness and consciousness are only gradually different, and thus it stands
also with the many different individual consciousnesses. To these psychical differences of degree
correspond the physical differences of degree between rest and motion, and between different motions
mutually.---In a similar direction as Leibniz go, on this point, Schopenhauer, Beneke, Lotze, and in
the most recent time Bergson (see 132).

The metaphysical idealism would only be able to lead to a concluding solution if it could provide
means for a positive determination of the different degrees and nuances of spiritual existence which
according to it constitute existence's innermost essence. But there are no means for this, and this
rests on the fact that the concept potential psychical energy, whose application is necessary for the
carrying-through of the idea of psychical continuity, only designates a problem or a postulate (see
8). Already for this ground the metaphysical idealism can never become anything other than a belief.
And when one tries to formulate idealism in a single idea, it unavoidably gets more the character of
the picture than of the thought.\footnote{See on this \emph{Religionsfilosofi}. \S\ 21--22.} We stand
at the boundary between thinking and poetry.

140. There is no bound choice between idealism and materialism, so that he who rejects the one
solution would necessarily have to assume the other. For it cannot be proved that what is not
spiritual must necessarily be material, and conversely. The division into spiritual and material
objects is purely empirical, and it cannot be established that existence's content should be exhausted
by this division. On this point too existence could be inexhaustible. All our knowledge develops
through a series of predicate-determinations, conclusion-representations; but whether this series is
ever complete, so that all that ``lies in'' the ground, in the indeterminate point-of-departure
representation, comes to appear as member in the series (cf. 35), cannot be decided, and still less
whether all sides of existence make themselves known in the object-ground on which our knowledge thus
builds. An Either---Or would only lie before us if it could be proved that all existence's fundamental
traits made themselves known in our experience, and that this experience was exhaustively characterized
by the twofold division into spiritual and material objects.

Spirit and matter stand (cf. 129) not to each other as Yes to No. There is a relation of opposition or
of difference between them, but there need not be a relation of contradiction between them. It is
possible that they stand to each other as two color-qualities stand to each other; the one is not what
the other is, but therefore it is not said that what does not have the one quality must have the other.
If we knew only two colors, e.g. yellow and blue, we would nevertheless obviously not be justified in
saying that what is not yellow must necessarily be blue. The necessity would in any case have to be
limited to our experience. Our color-scale has a purely empirical character, an upper and a lower limit,
which hangs together with more deep-lying physical and physiological relations. When there is more in
existence than we think of in our philosophy, the ground need not be that we have not philosophized
rightly; it can lie in the fact that this more does not make itself known as object for us. Again here
we strike on the irrational.

According to this consideration we cannot, as regards the here-lying part of the problem of existence,
conclude our thinking at a primal phenomenon, since none of the objects is suited to it. If one would
hold that one could take them both together and say that existence was ``psychophysical,'' one would
thereby only have gotten a sounding word, no solution (cf. 9).

According to \emph{Goethe} ``the highest a human being can reach is wonder,''---``and,'' he adds, ``when
the primal phenomenon sets him in wonder, let him be content; a higher it cannot yield him, and further
he shall not seek behind it; here is the limit.'' Goethe's need for poetic beholding here comes to light.
But even though we find no primal phenomenon,---and precisely because we find none,---there is room for the
great wonder. Out beyond all objects and all reflection over them there shows itself an order of things,
within which our experience and our thinking are like islands in the ocean.---It is of course a belief,
when we here speak of an \emph{order} of things; it is a belief that supports itself on the ground on which
the critical monism (134) builds.

When a feeling of loneliness can arise at the apparently vanishing and unassuming position the spiritual
occupies in relation to the material (138), then a feeling of loneliness of a deeper kind will arise when
we become clear that our whole world of spiritual and material objects, together with the knowledge and
the problems it conditions, is again only a single section of a far more comprehensive existence. Wonder
has here natural ground; it stands at our wisdom's end, not merely at its beginning. Whether it is to become
admiration is a question that must be discussed in another connection.

What speaks for seeing the matter in this way is that there are such great difficulties for thought in
reaching a satisfactory conception of the relation between the spiritual and the material. These two
objects cannot be reduced to each other, and we know no third from which we could derive them as special
cases. It was then possible that these difficulties hung together with the fact that there are sides of
existence that do not appear for us, but that contain the ground for the twofold division within our
experience. Our cognition's history shows us many examples of extended experience's being able to solve
problems or get them out of the way. We have indeed no prospect of an extension of our experience that
could here help us; but we have the right to conceive our position here in analogy with such cases where a
problem has shown itself to rest on limited experience.

Existence would offer great simplicity if there were only two fundamental forms under which it stepped
forth. But what stands for us as simple and easily surveyable can very well be due to a very intricate
connection. In our simplest sense-sensations we have examples of this, when we compare them with the
physical and physiological processes to which they correspond. It can be ``economical'' or ``pragmatic''
to operate with the principle of simplicity. But the attempt to develop a world-view has significance only
when earnest is made of investigating all possibilities. It is here not a matter of our convenience, but
of a last orientation concluding for us. Both idealists and materialists as a rule come too quickly to
their last word. It can also, in a single, relatively limited domain, be very dangerous to proceed from
the assumption that only two possibilities, an Either---Or, lie before us, because experience provisionally
offers no others. \emph{Charles Darwin} experienced this in his youth, when he in a geological treatise had
concluded that the parallel lines on the Scottish coast must be due to the sea, since no other explanation
seemed tenable. Agassiz's glacier theory showed a new possibility. And Darwin learned from this that in
empirical science one should be careful with applying the principle of the excluded middle.\footnote{\emph{Charles
Darwin}: \emph{Life and Letters}. I. p. 69. 291.}

Among thought's masters \emph{Spinoza} has here seen most clearly, when he taught that the difference
between the spiritual and the material was purely empirical, and that existence in itself unfolded itself
under infinitely many fundamental forms (attributes). In one of his earlier writings he seems to have
imagined the possibility that we could come to conceive more attributes than those two; but in his main
work (Ethica) he has given up this thought. The ground why there must be infinitely many attributes lies
for Spinoza in the fact that existence's whole fullness can only thereby find expression. It is Spinoza's
belief that here comes to light. He could simply have supported himself on the purely empirical character
of the difference between the spiritual and the material; but he would then have had to express himself
somewhat less dogmatically. Neither can we follow him in the assumption that all the different attributes
are logically coordinate or ``parallel.'' The possibility cannot be rejected that one or more of the
attributes inaccessible to us contain the ground for the peculiar difference and the peculiar ``parallel''
relation that our experience leads us to assume between spirit and matter.

Spinoza's fundamental thought preserves its significance even when the dogmatic element in his doctrinal
structure is corrected. It shows us out to a greater horizon than the one within which idealism and
materialism carry on their strife.

\subsection*{d. Subsistence and Development}
\addcontentsline{toc}{subsection}{d. Subsistence and Development}

% [text to be added: pp. 339--347]

\section*{C. Valuation}
\addcontentsline{toc}{section}{C. Valuation}
\label{sec:vurdering}

% [text to be added: pp. 347--350 — section intro]

\subsection*{a. The Ethical Problem}
\addcontentsline{toc}{subsection}{a. The Ethical Problem}

% [text to be added: p. 350 — subsection intro, if any]

\subsubsection*{α. \emph{Ethical Work}. (Formal Ethics)}
\addcontentsline{toc}{subsubsection}{α. Ethical Work. (Formal Ethics)}

% [text to be added: pp. 350--359]

\subsubsection*{β. \emph{The Rationality of Ethical Valuation}. (Real Ethics)}
\addcontentsline{toc}{subsubsection}{β. The Rationality of Ethical Valuation. (Real Ethics)}

% [text to be added: pp. 359--369]

\subsection*{b. The Religious Problem}
\addcontentsline{toc}{subsection}{b. The Religious Problem}

% [text to be added: pp. 369--374 — subsection intro]

\subsubsection*{α. \emph{The Psychological Place of Religion}}
\addcontentsline{toc}{subsubsection}{α. The Psychological Place of Religion}

% [text to be added: pp. 374--385]

\subsubsection*{β. \emph{The Historical and the Philosophical Consideration}}
\addcontentsline{toc}{subsubsection}{β. The Historical and the Philosophical Consideration}

% [text to be added: pp. 385--392]

\end{document}
