% ============================================================
%  Rasmus Nielsen, Philosophiske Grundproblemer (Philosophical Fundamental Problems)
%  Festskrift, University of Copenhagen 400th-anniversary, June 1879. 77 pp.
%  TRANSLATION (English)
% ============================================================
\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{libertinus}
\usepackage{libertinust1math}
\usepackage{textalpha}          % polytonic Greek typed directly (kept verbatim)
\usepackage[english]{babel}
\usepackage{enumitem}           % keep even if unused now — cheap insurance
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage[margin=1.4in]{geometry}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,
  pdftitle={Philosophical Fundamental Problems},
  pdfauthor={Rasmus Nielsen}]{hyperref}
\setlength{\headheight}{15pt}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\textit{Philosophical Fundamental Problems}}
\fancyhead[LO]{\textit{\rightmark}}
\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt}
\onehalfspacing

\title{\textbf{Philosophical Fundamental Problems}\\[0.5em]
  \large Festschrift for the University's Four-Hundredth Anniversary}
\author{R.\ Nielsen\\[4pt]\small Translated from the Danish}
\date{Copenhagen: Gyldendal, 1879}

\begin{document}
\maketitle


% ============================================================
%  INTRODUCTION (untitled)
% ============================================================
% Introduction (pp. 3--4)

% p. 3
\noindent
Philosophy does not, like other sciences, enjoy the advantage of
being able to build upon immediately firm presuppositions, upon
indisputable fundamental principles; for in the whole of philosophy
there is not a single proposition that stands so firm that it cannot,
when treated on its own, be assailed by criticism and dissolved in
dialectic. For that reason philosophy cannot very well parcel out its
proofs in the way that mathematics, for example, can. Although
mathematical proofs must be presented in strict order if full,
scientific insight is to be attained, the mathematician nevertheless
does not expose himself to any essential misunderstanding if, for one
purpose or another, he treats a particular problem in isolation. Those
who are competent know how the presuppositions can be proved, and those
whose prior knowledge does not suffice content themselves with the
result of the demonstration, trusting that the foundation is firm. But
upon such a trusting acquiescence the philosopher, it lies in the nature
of the case, dare not reckon. In philosophy no proof can stand alone;
the treatment of a separate part must borrow its light from the whole,
and the comprehensive system-proof must, so to speak, be transparent in
every single demonstration. But it is now well known from the history of
philosophy that the systems succeed one another, so that the preceding
ones always fall before those that follow. This is indeed to some extent
the case in the other sciences as well, but with the—admittedly very
significant—difference that certain discoveries, methods, and modes of
proof either remain unassailed or at least, under a further-reaching
elaboration, yield points of support, recognizable points of departure
for further progress. It is otherwise in philosophy. Here a new point of
view, a new thought, a new idea cannot step out into the world as a
physical or natural-historical discovery can. When a philosophical
system, with its principle and its peculiar method, falls prey to
critical dissolution and loses its standing, the whole is recast in such
a way that, in the new casting, one can hardly recognize what the abiding
element really is that the preceding system deposited. If one regards the
matter exclusively from this side, one would be tempted to believe that
philosophy was no true science. But the view is one-sided and therefore
misleading; the misleading element lies in the fact that one takes one's
models from other sciences and applies a standard that does not fit.
Philosophy must necessarily have its ideal peculiarity; what essentially
belongs to the invisible world of thoughts cannot possibly bear the same
marks as that which can be demonstrated from piecemeal-posited conditions
or exhibited in visible facts. Seen, on the contrary, from the standpoint
of the development of thought and of a consciousness growing in clarity,
it shows itself that philosophy, in steady historical continuation,
labors at the solution of certain ever-recurring problems; that every
thinker who, by placing one or more of these problems in a light
advantageous to inquiry, has made a contribution to the solution, remains
standing as an eminent figure in the history of philosophy; and that his
contribution continues to have effect, even though his school has long
since ceased to exist. For a closer understanding of this, and in order,
if possible, to contribute to a somewhat clearer conception of philosophy
as a science, attention is directed to the problems set forth below.


% ============================================================
%  I. THE PROBLEM OF KNOWLEDGE
% ============================================================
\section*{I.\ The Problem of Knowledge}
\label{sec:erkjendelse}

% p. 4 (section heading printed at the foot of p.4; text begins here)
\noindent
Among the philosophers who in more recent times have treated the
question of the conditions and limits of human knowledge, Imm.\ Kant
indisputably occupies the first place—if not by the results to which his
% p. 5
inquiries led him, then at any rate by the manner in which he has posed
the problem. Earlier thinkers—whether, like Locke, they laid the emphasis
on the senses, or, like Leibniz, on the understanding—simply presupposed
that we know from experience what experience is, that experience is
something that is understood of itself; but Kant's ``\emph{Kritik der
reinen Vernunft}'' shows that experience precisely is \emph{not} something
that is understood of itself, that in order to discover which cognitions
we can draw from the sources of experience, we must go back to the
inviolable fundamental conditions of cognition and begin with the question
of \emph{the possibility of experience}.

It is evident that the presuppositions which the knower himself must bring
with him in order to be able to make an experience cannot be borrowed from
experience, even though it is only with the help of experience that they
can come to consciousness. What, then, are these presuppositions? Without
sensation it is impossible to apprehend the sensible, and without
understanding impossible to determine what is sensed: experience is
conditioned by sensation and thought; but by what is sensation
conditioned, and how does it stand with the understanding? All sensation
rests upon an intuiting. To sense an object in space is just as impossible
without the intuition of space as it is impossible to apprehend a change
in time without the intuition of time. The pure intuitions, \emph{space
and time}, are therefore a priori necessary presuppositions for all
sensation and thereby for all experience. In order to procure a positive
content, thought must attach itself to sensation; for the understanding is
no intuiting faculty, and the old metaphysics' attempts to open up for
itself a supernatural world and, by pure understanding, to make
discoveries of supersensible things are fantastical. But that sensation
cannot bring it about that any experience whatever is made without the
assistance of the understanding is seen from this: that the one
sense-impression cannot be distinguished from the other without a
discriminating activity, and the one representation is not determined by
the other without a judging activity. The judging understanding would not
be able to exercise its activity if it did not itself contain
\textit{a priori} those root-concepts, categories, by which logical
judgments must necessarily be conditioned.
% p. 6
Which these categories are, and how many they are, can be seen from the
division of judgments in formal logic. Thus the hypothetical judgment
shows itself to be grounded in the categories of cause and effect, and the
disjunctive in the category of reciprocity. The a priori categories of the
understanding are therefore indispensable for all thought and thereby for
all experience.

The pure sense-intuitions and the a priori concepts must necessarily be
fundamentally different, indeed heterogeneous, since the senses are
heterogeneous with the understanding; the question then arises how it is
possible that forms so heterogeneous can be united in the cognizing
consciousness. We are referred to the imagination, which, on the one side
intellectual, on the other side sensible, by means of the intuition of
time, furnishes an a priori schema necessary for the possibility of
experience. It is not enough, for example, to have assured oneself of the
validity of the concept of cause as a concept alone; here a condition is
required for the concept's application to the appearances in the sensible
world, whereby the determinate order of time acquires decisive
significance.

By pointing out the subjective fundamental conditions holding for all
experience, Kant found himself able to overcome D.\ Hume's empirical
skepticism. Hume rejected the concept of cause as a kind of habit-concept
to which no experience corresponds. That things accompany and succeed one
another we can experience, but that the one is cause of the other we
cannot experience; what we by habit call cause is as unknown to us as what
we call force. Kant grants Hume this much, in that he demonstrates the
absurdity of wishing to derive the concept of cause from experience; but
inasmuch as he at the same time proves that no experience whatever is
possible without the concept of cause, he asserts victoriously and once
and for all this concept's a-priori-empirical validity so emphatically
that, according to him, experience and cause must be said to stand and
fall with one another. In particular cases of experience it can often be
doubtful what the cause is of an event that takes place. When A, B, and C
% p. 7
are accompanied by D, it may perhaps be A, perhaps B, perhaps C, perhaps A
and B, perhaps A and C, perhaps B and C, perhaps A, B, and C in
conjunction, in which the cause of D is to be sought; for the decision of
such doubtful cases one may employ induction, but induction precisely
presupposes that there must be a cause.\footnote{``Zwar scheint es, als
widerspreche dieses allen Bemerkungen, die man jederzeit über den Gang
unseres Verstandesgebrauchs gemacht hat, nach welchen wir nur allererst
durch die wahrgenommenen und verglichenen übereinstimmenden Folgen vieler
Begebenheiten auf vorhergehende Erscheinungen, eine Regel zu entdecken,
geleitet worden, der gemäss gewisse Begebenheiten auf gewisse
Erscheinungen jederzeit folgen, und dadurch zuerst veranlasst worden, uns
den Begriff von Ursache zu machen. Auf solchen Fuss würde dieser Begriff
bloss empirisch sein und die Regel, die er verschafft: dass Alles, was
geschieht, eine Ursache habe, würde eben so zufällig sein, als die
Erfahrung selbst; seine Allgemeinheit und Nothwendigkeit wären alsdenn nur
angedichtet und hätten keine wahre allgemeine Gültigkeit, weil sie nicht
\textit{a priori}, sondern nur auf Induction gegründet wären\ldots\
Freilich ist die logische Klarheit dieser Vorstellung einer, die Reihe der
Begebenheiten bestimmenden Regel, als eines Begriffs von Ursache, nur
alsdenn möglich, wenn wir davon in der Erfahrung Gebrauch gemacht haben;
aber eine Rücksicht auf dieselbe als Bedingung der synthetischen Einheit
der Erscheinungen in der Zeit, war doch der Grund der Erfahrung selbst und
ging also \textit{a priori} vor ihr vorher.'' Kritik der rein.\ Vernunft.
Leipzig 1838. S.~201.}

Here, be it noted, there is no question of the relation between the cause
and the thing itself, which falls outside our consciousness, but only of
the necessity of the concept of cause for our experience. An arbitrary
play of sensations and representations is only of subjective significance
and yields no experience. If a set of terms a, b, c, d, and so forth were
incessantly to alter their order and mutual position without any
conceivable ground whatever, we should indeed be compelled to apprehend
these changes in time, but the temporal succession would give us no
concept of cause and, as a matter of course, no experience either. If the
image of a house, for example, changed every moment in such a way that now
it turned its ridge downward, now upward, and, by the transposition of its
parts, took on the shape now of a tower, now of a whale, then we should
have to seek the ground of the changes in our own imagination and could in
no way refer such an image to any object of experience. For the sake of
experience there must be something that
% p. 8
compels us to place the one before and the other after; it is, in other
words, the necessarily determinate order of time that expresses the
relation between cause and effect. The necessity is a priori, for it is a
conceptual necessity; with this necessity the object is determined by
experience, and experience by the object.\footnote{``Der Begriff, der eine
Nothwendigkeit der synthetischen Einheit bei sich führt, kann nur ein
reiner Verstandesbegriff sein, der nicht in der Wahrnehmung liegt, und das
ist hier der Begriff des \emph{Verhältnisses der Ursache und Wirkung},
wovon die erstere die letztere in der Zeit, als die Folge, und nicht als
etwas, was blos in der Einbildung vorhergehen (oder gar überall nicht
wahrgenommen sein) könnte, bestimmt. Also ist nur dadurch, dass wir die
Folge der Erscheinungen, mithin alle Veränderung dem Gesetze der
Causalität unterwerfen, selbst Erfahrung d.\ i.\ empirisches Erkenntniss
von denselben möglich; mithin sind sie selbst als Gegenstände der
Erfahrung, nur nach eben dem Gesetze möglich.'' Anf.\ Skr.\ S.~196--97.}

From this one will understand that Kant, with inflexible rigor of
consequence, must make a divorce between the object of experience (the
phenomenon) and the thing in itself. There must be something that,
independently of us, gives our empty forms of cognition—the sensible as
well as the intellectual—a real content and determines our apprehension of
the object; but that the object, as we apprehend it, cannot possibly be
identical with the thing itself is a straightforward consequence of the
fact that, in passing through our subjective media of cognition, it must
necessarily take on a subjective tinge. Our apprehension of the thing is
an alteration of the thing: we see objects through colored spectacles.

The sharp separation between the thing in itself and its appearance finds
recognition in the exact sciences, particularly in the physiology of the
sense-organs. Helmholtz has\footnote{Handbuch der physiologischen Optik.
Leipzig 1867. S.~442--flgd.} emphatically stressed that our images and
representations are effects which the intuited and represented objects
produce upon our nervous system and thereby upon our consciousness. Here
there is a reciprocal action between an outside and an inside, and the
cause of the phenomenon is precisely this reciprocal action. To ask
whether the representation I have of a table—its form, its solidity, its
% p. 9
color, weight, and so forth—agrees with the real thing is the same as to
ask whether a tone is red, yellow, or blue. Things are determined by their
properties, and the properties by reciprocal action with our organs of
consciousness and with other things. Cinnabar is red, be it noted, for
those who can distinguish colors, but black for the red-blind. Although the
latter, according to the law of reciprocal action, have the same right in
their judgment as the former, we nevertheless say in general: cinnabar is
red; likewise we often ascribe to a thing a property as though it belonged
to the thing in itself, although this property holds only for the thing's
reciprocal action with other things, as when we say: lead is soluble, and
tacitly understand: in nitric acid, not in sulfuric acid. The result is
that our representations of things have only practical significance. Our
designations, our words and expressions, are only symbols; out of these
symbols a language can be formed, expressive of our observations, exact
investigations, and scientific cognitions, but which has nothing,
absolutely nothing, to do with the essence of things.

If this now is the final, decisive result of the inquiry, if the problem of
knowledge is herewith once and for all solved—in principle, that is—if all
further-reaching research will serve only for a further confirmation of the
Kantian theory: then not only must the works of Fichte, Schelling, and
Hegel be regarded as misguided, but every future philosophical attempt to
illuminate the essence and ground of existence must be dismissed as barren
and, for science, misleading. However the matter may stand, this much is
certain, that the Kantian doctrine of knowledge has an ambiguity that must
be drawn forth into the light. The fundamental question is: how can that
which fills our forms of cognition with \emph{real content be so absolutely
heterogeneous with these forms that the thing itself must remain entirely
unknown to us?}

Let us call the unknown X and confine ourselves to a definite object,
for example a table. That we apprehend the table in space and ascribe to
it a bounded extension, whereby it acquires its shape, is evident; but the
pure intuition of space gives
% p. 10
us no real table. The table is not a product of our own imagination; here
is a something that acts upon our senses, notably upon sight and touch—this
something is X. The table, the real phenomenon, thus comes about through
two heterogeneous contributions, a subjective one from ourselves and an
objective one from the thing in itself. We can symbolize it by an equation:
$X + Y = P_{x,y}$, where X denotes the objective, Y the subjective, and
$P_{x,y}$ a function of X and Y, the phenomenon. If we set $Y = 0$, then
the subjective vanishes, and $X = P_x$, that in the phenomenon which is due
to the objective alone, remains. That $P_x$ cannot itself be a phenomenon,
that subjectivity is required in order to make the thing into a phenomenon,
Kant has excellently shown; but how this happens, or even the possibility
that it can happen, he leaves standing in complete ambiguity. --- How,
then, does X actually stand in relation to $P_{x,y}$, the space filled by
the table? One can conceive the matter in several ways. X may be assumed,
without filling any space, to make an impression on the soul of such a
character that the determinate space-filling corresponds exactly to the
intensity of the impression. The real table is thus by no means a space
filled by X, for the X-impression brings it about only that subjectivity,
which itself fills the given space, must imagine that it is a real object
that fills it. This assumption must strike us as unreasonable, especially
when we consider that the table is not merely an object of sight, but also
an object of touch. Everywhere that we touch the table's surface, we sense
the reality. But the reality of every sensation derives, after all, from
X-impressions; there must then be just as many different X-impressions as
the table's surface has parts, and these impressions—whose difference here
is a difference of place—must be arranged and distributed in exactly the
same way as the parts of the surface, that is, arranged and distributed in
space. Yet neither does this assumption satisfy, for it would follow that
to the phenomenon of a square table there would correspond, if not indeed
a square X, then at least X-impressions that fit exactly the square, to the
phenomenon of a round table X-impressions that fit the round, and so forth.
If one separates the impressions
% p. 11
of X, which make the spatial determinations real, from X itself as that
which is outside space, then the X existing outside space has, after all,
nothing whatever to do with the real spatial phenomena; its significance
for a filling of the intuition of space is reduced to nil.\footnote{If one
assumes that X, in a certain negative way, reflects itself in its
X-impressions that refer to the intuition of space, just as the
X-impressions reflect themselves in the objects of experience, then one
gets an X-phenomenon behind every experiential phenomenon; but since the X
that stands behind the X-phenomenon must in turn reflect itself in an
X-phenomenon that removes itself still further from the form of space, one
gets X-phenomenon behind X-phenomenon ad infinitum before one reaches the
last X, separated from space, the pure ``Ding an sich''.}

Similar difficulties present themselves when we consider the phenomenal
changes that give the intuition of time real content. The
substance-phenomenon shows us what \emph{abides} in time, the
accident-phenomena what \emph{alternates} in time, and yet the substance
must be characterized by the accidents, and the accidents by the substance.
Now it is asked: is it one and the same X that determines the reality both
of the substance-phenomenon and of the accident-phenomena, or are there
here two kinds of X? If we assume the former, then we lay into X the very
same characteristic difference as into the phenomena. As the X of substance
it \emph{abides} in time, as the X of the accidents it \emph{alternates} in
time; this will then, in other words, mean that the thing in itself, the
unknown, is able to fill our intuition of time with real content only on
the inviolable condition that it, both from the side of substance and from
that of the accidents, enters into time, abides in time, and changes in
time. But this assumption stands in manifest conflict with the Kantian
presupposition. If we prefer the latter and posit one X for substance,
another for the accidents, then the multiplication of X will also be
carried further. It is then not enough to have a general X for substance in
general; we must have a particular X for each particular substance, a
sulfur-X for sulfur-substance, a
% p. 12
carbon-X for carbon-substance, an iron-X for iron-substance, and so on; as
regards the accidents, it will be doubtful how far we can make do with one
X for each group of accidents, or whether just as many X's are needed as
there are accidents in each group. The Kantian doubling—phenomenon and
thing in itself—comes, in this light, strongly to recall the Platonic: the
table and tableness, the horse and the horse-itself. However one turns and
twists it, the following question remains to be answered: Can X enter into
changes or can it not? If it cannot, then it keeps itself quite rightly
outside time, but then all phenomena of change become X-forsaken, unreal
semblances. If X can take part in the changes and give them the stamp of
experiential reality, then this reality-giving must also in one way or
another itself enter into time-determinations and take part in time. To
take up the X-impressions that determine the changes into the intuition of
time, and yet to let X be outside the intuition of time, is an absurdity,
for it is a contradiction.

The Kantian doctrine of knowledge impresses upon us that our concepts of
the understanding have validity only in the determination of the objects of
experience and find no application whatever to the thing in itself. But
when the thing in itself falls outside all conceptual determinations, how
then can it be determined? It is supposed not to be determinable by
sensation, since it is strictly held outside space and time; it must
therefore be determined by thought. Since now the understanding cannot
possibly procure for itself any positive content other than what it
receives from sensation, the consequence must surely be that the thing in
itself can be determined only negatively. One would now suppose that the
only determination that might be given of X would be this, that it excludes
all determinations. It would then be just as indeterminate that it is as
that it is not, just as indeterminate that it makes an impression on our
senses as that it makes no impression, and so on. But in this supposition
we are greatly disappointed. X becomes expressly determined by an act of
the understanding; it becomes determined by a judgment, not by a negative,
but by a positive judgment: \emph{the thing in itself is}. Being, a concept
of the understanding, is here,
% p. 13
contrary to the theory's own prescription, applied to X. Being denotes
reality, something positive; whence does the understanding get this
positive? Not from sensation, for sensation gives only what can be
apprehended in space and time. Nor from anything other than sensation, for
here there is nothing else that can give the understanding the positive. So
X must, then, be known indirectly by its effects, that is, by the
sense-impressions of which it is the cause. But this assumption too the
theory has excluded, for cause is a concept of the understanding just as
much as effect. By determining the thing in itself as an unknown cause
lying outside understanding and sensation—a cause that is to be thought as
objective and yet cannot be thought as objective, a cause that in its
effects, for the sake of reality, must be sensed and yet cannot be
sensed—Kant has, instead of showing the possibility of experience, on the
contrary proved the \emph{impossibility} of experience.

If there is no objective form at all that can belong to the thing in
itself, then this unknown must necessarily dissolve into a hazy
possibility, a ὕλη ἄμορφος, and lack all objective self-subsistence. It is
then so far from the case that the in-itself indeterminate possibility can
be said to determine the objective content of our subjective forms, that on
the contrary it borrows its apparent forms of reality from us and only
through this loan acquires the semblance of something objective. --- Here,
however, is a consideration that can be brought forward from the Kantian
side. Doubtless we must ascribe objective forms to the thing in itself,
insofar as we ascribe to it objective validity, but the point is that we do
not know these objective forms at all. To be sure, they cannot be forms
like ours; but who can prove that there are no other forms of existence
than those that underlie our sensation and thought? Not to speak of the
characteristic differences of our five senses—sight, hearing, smell, taste,
and touch—both among themselves and among different individuals, or of the
possibility that cognizing beings may have both more and sharper and finer
senses than we, there is surely no human being who can vouch that our a
priori intuitions and categories correspond even approximately to the
objective form-determinations expressed in
% p. 14
the thing itself. Is not the objectivity of space and time problematic?
Must we not, even as regards objects of experience, make a distinction
between the thing's concept and our imperfect, often faulty concept of the
thing? --- Here again we have the ambiguity.

It scarcely occurs to any reasonable person to deny the imperfection of our
intuition of space when it comes to a clear and distinct apprehension of
definite, many-sidedly combined and thereby involved relations in space. To
construct in space is not exactly an easy matter. Should anyone doubt this,
he need only make even a merely preliminary acquaintance with analytic
geometry; he will then soon learn how easily a curve or a mathematical
body, whose equation is known to him under certain conditions, can become
unrecognizable either by a displacement of the coordinates' point of
intersection or by the whole coordinate system's being altered. The more
combined the figure is, the more difficult are the special relations to see
through: here, then, manifold doubts can arise about the objectivity of
spatial relations in the particular. But it is one thing to concede the
imperfection of the representations and sensations conditioned by the
intuition of space, and another thing to deny objective validity to the a
priori form-determinations of the intuition of space. If a cognizing being
descended to us from higher regions and declared that it could distinguish
millions of parts in the space we took to be filled by a single part, we
should by no means contradict it; but if it moreover maintained that it
could distinctly see the circle as a square and the square as a circle,
should we then be convinced that its intuition of space was more objective
than ours? If the stranger gave proofs that he could not merely sense the
ether, but even count 733 trillion vibrations in one second, he would
astonish us in the very highest degree; but if he then gave as the reason
that he was able to apprehend the simultaneous as successive and the
successive as simultaneous, should we then dare to conclude from this that
his intuition of time came nearer to the objective than ours? Between space
and not-space, time and not-
% p. 15
time, no intermediate forms are possible. How many dimensions space may
appear to the sensing being to have is not here the question; the question
is of dimension and succession. By withdrawing itself from all
determinations of space and time, of dimension and succession, the whole
``objective, unknown world'' must precisely become known to us as a single
intensive point.\footnote{That the air-waves are heterogeneous with the
tones and the ether-waves with the colors proves only that a cause in the
objective can become unrecognizable by entering into reciprocal action with
the subjective; but what is incomprehensible herein proves nothing either
against the concept of cause or against the objective validity of time and
space. Of the influence of our organization there will be speech later.}

In confirmation that our concepts of reason (ideas) find just as little
application to anything objective in itself as our concepts of the
understanding, Kant sets up four so-called cosmological antinomies, namely
two mathematical and two dynamical. In this connection we confine ourselves
to the mathematical ones. First antinomy. \emph{Thesis}. The world has a
beginning in time, and is, as regards space, enclosed within limits.
\emph{Antithesis}. The world has no beginning, and no limits in space; it
is, in respect both of time and of space, infinite. Second antinomy.
\emph{Thesis}. Every composite substance consists of simple (uncomposite)
parts, and there exists nowhere anything but the simple and that which is
composed of it. \emph{Antithesis}. No composite thing in the world consists
of uncomposite parts, and there exists in general nothing simple in it.
--- Since the proof both for thesis and for antithesis in each of the
antinomies must be granted the same validity, this mutual conflict of the
transcendental ideas is supposed irrefutably to prove that we can just as
little reach the objective by means of our reason as by the use of our
understanding. Let us test the proofs. Does the thesis in the first
antinomy stand in insoluble conflict with the antithesis? We cannot see
that it does. If the world is from eternity, then there is, after all, no
contradiction whatever in the fact that up to a given point of time an
infinite series of successive world-states may have elapsed. To be sure, an
infinite series cannot by any successive synthesis
% p. 16
be completed in a finite time; it requires an infinite time in order to be
completed. But from whatever finite point of time one looks back toward the
infinitely distant beginning, one looks, after all, into an infinite time.
If there is nothing contradictory in assuming an infinite series of
represented terms, then there is nothing contradictory either in assuming
an infinite series of real terms, provided the reality, be it noted, is
presupposed to be from eternity. Why can the world not have begun in time?
Because, says the antithesis, a time must then have gone before, hence an
empty time; but in an empty time nothing can arise, since in no part of
such a time is there any condition for a distinguishing of existence from
non-being. But such an empty time is more imaginary than any imaginary
magnitude otherwise is. With an imaginary magnitude, mathematically
understood, one can nevertheless reckon; but an entirely empty time, a time
in which there is no difference between the past, the present, and the
future, is no time, and with no time one cannot reckon. As surely as the
real world really has a beginning, its beginning must necessarily fall in a
real time, and the real time begin with the real world. Since in this
respect it makes no difference whether we push the beginning's point of
time back half an eternity or a whole one, there is, as regards the world's
beginning, no conflict between thesis and antithesis. --- In considering
the limitation and non-limitation of space we come to the same result. The
real world is limited, for the outwardly real must at every moment have
definite limits and to that extent be a finite. But the world in its
finitude and limitation is also in real becoming, and can thus at every
moment be on the point of overstepping every finite limit. It is the real
world that makes space objective and sets objective limits in space; cosmic
space has no peculiar being outside the world. The antithetic is built on
the impossibility of completing the successive synthesis of the terms in an
infinite series, that is, on the impossibility of comprehending all parts
of an infinite totality. But, we ask, will then a limited
% p. 17
whole, which contains only 7 parts, be more real than a whole that contains
7 trillion parts, because the successive synthesis in the first case is
easy for us, but in the last case impossible? Surely no one will maintain
that. But if we may assume a totality with 7 trillion parts, a world-whole
with 7 trillion firmaments, to be real, notwithstanding that the successive
synthesis requires an enormously long time, how then can there be any
contradiction in presupposing the reality of an infinite totality, because
a successive synthesis (more correctly: coming-into-being) of its
individual parts must have required an infinitely long time?

For the \emph{Thesis} in the second antinomy it is adduced that nothing
substantial could be given if no indivisible parts existed; the
substantially composite must therefore consist of uncomposite parts. In
support of the \emph{Antithesis} it is adduced that no composition of
substances is possible except in space. The substance existing in space
must necessarily have substantial parts; but space has no substantial
parts, it is a \emph{totum}, no real \emph{compositum}. The contradiction
is now supposed to be: that composite substances cannot exist without
simple substances, and that simple substances cannot exist in space.
Instead of consisting of simple substances, the composite substance comes
to consist of composite substances ad infinitum, and that is absurd. ---
This contradiction, however, is more imaginary than real. If simple
substances and space are incompatible magnitudes, then not only experience
but also mathematics is impossible, for then intuition, and with it
sensation, is absolutely incompatible with thought. ``It is here,'' says
Kant, ``not enough, for the pure concept of the understanding of the
composite, to find the concept of the simple; one must find the intuition
of the simple that corresponds to the intuition of the composite, and this
is, according to the laws of sensibility, hence also in the case of objects
of sense, entirely impossible.'' --- This impossible, however, is not so
evident as it might seem at first glance. A mathematical body, a cube for
example, is an object of intuition; how does it stand here with the simple
and the
% p. 18
composite? The given cube can be intuited as a single existent, a continuum
bounded by 6 square faces; but it can also be regarded as a composite and
resolved into elements. If it is denoted by an infinite summation of
represented corporeality-elements $x^2dx$, then the third part of the
intuited cube is an integral, determined by finite limits, of these
elements. The face can be resolved into face-elements $xdx$, the line into
line-elements $dx$, and these again into elements of a higher order, and so
forth. In relation to the body the corporeality-element is the simple, just
as the face-element is the simple in relation to the face, and the
line-element the simple in relation to the line. Thought and intuition are,
as regards the mathematical body, agreed that the simple must be relative
for the sake of the composite, and the composite relative for the sake of
the simple, that therefore the fixing of an absolutely self-existent simple
is not an idea of reason, but a fixed idea. Should not the same hold for the
material body? It is supposed, according to Kant, to be an absurdity
``besides the mathematical point, which is simple, but no part, only a
limit in space, to think of physical points, which are indeed simple, but
yet have the advantage, as parts of space, of filling it by their
aggregation.'' But the alleged absurdity will not become evident to us. Let
us only put atoms and molecules instead of mathematical elements, and
aggregations instead of infinite summations: the fundamental relation
between the simple and the composite will nevertheless remain the same.
That the finite discretion is a parceling-out, a breach of continuity, must
be conceded. But even though the atoms are, in a real sense, ever so
indivisible—indeed even if, what for other reasons is an absurdity, they
are assumed to be absolute simple substances—the intuition of space can
nevertheless have not the least
% p. 19
to object against this. The atom's substantial simplicity by no means
stands in conflict with its mathematical divisibility. The extended-in-space
simple substance has, to be sure, just as many substance-elements as the
space it fills has space-elements, that is, it has infinitely many, for the
discretion goes on ad infinitum. But since the infinite discretion, as is
expressly seen from space, coincides with continuity, what can intuition
object against the simple substance—a real continuum—filling space, the
formal continuum? Since the parceling-out is presupposed, the finitude of
aggregation steps in place of the infinitude of summation; but since the
real distance between real parts precisely gives the form of space the
stamp of reality, the aggregation must surely be an eloquent proof of the
formal objectivity of space.

Kant is fully convinced of the insolubility of his antinomies, ``denn für
die Richtigkeit aller dieser Beweise verbürge ich mich\footnote{Prolegomena.
Werk.\ 3ter Band. S.~264.};'' but the insolubility presupposed, what is
thereby gained? ``That we must not think the phenomena of the sense-world
as things in themselves.'' That must be conceded. But there follows yet
something else from it, something that it is precarious to concede; for when
it is expressly added that we must not think the fundamental principles for
the connections of phenomena as valid for the things in themselves either,
then the understanding is lost, then it grows dark before the eye of reason.
If we could still stop at the preliminary result, that in all our cognizing
we must hold to our experiences and leave undecided how the things
themselves in the world unknown to us actually are constituted: that might
pass. But the antinomies drive us to an extreme that must either turn into
bottomless skepticism or awaken despair. We are to assume things in
themselves which, although they condition our experiences, nevertheless
exclude all conditions of experience, since neither space nor time, neither
unity nor plurality, neither ground nor consequence, neither cause nor laws,
neither whole nor parts, neither simplicity nor composition have the least
significance for these in-themselves unsensible and unthinkable things.
These are hard terms.

For the understanding of the conflict of the ideas of reason in the
antinomies, Kant adduces the following: ``Two mutually contradictory
propositions cannot both be false, unless the concept that lies
% p. 20
at the ground of them both is in itself contradictory; for example, the two
propositions: a square circle is round, and a square circle is not round,
are both false.'' But when the human consciousness has only a choice
between one of two: either to let the ideas of reason or the ``things in
themselves'' be ``square circles,'' then it is at least excusable if it once
more ventures to assert the right of reason by choosing the latter. This
choice is characteristic of J.\ G.\ Fichte's ``Wissenschaftslehre''.

We can briefly denote the deviation from the Kantian standpoint by
exchanging the symbol $X + Y = P_{x,y}$ for $XY = P$. That this must
necessarily give the inquiry another direction and lead to an entirely
different result is evident. When Y, the subjective, becomes nil, that is,
when the cognizing subjectivity, the Ego, vanishes, the whole becomes nil:
no indeterminable object, no self-existent X, remains. By establishing the
reciprocal determination of subjective and objective, Ego and non-Ego,
Fichte has introduced a great fundamental thought into the history of
philosophy, a fundamental principle of such incisive significance that it
can as little be excluded from the theory of knowledge as the law of
gravity from physics. Although Ego and non-Ego mutually exclude one
another, it is nevertheless self-evident that X, the non-Ego, can as little
be ascribed any self-subsistence valid outside the Ego as it can merge
together with the Ego itself. As real opposite to the Ego, the non-Ego must
therefore be an element in the Ego's essence, a barrier that is
indispensable for the Ego's self-development. Only by the thinking Ego's
becoming conscious of its opposite, its barrier, as an unconscious
presupposition that has proceeded from its own essence, does it become
conscious of itself, and overcome the barrier theoretically. The objects of
experience, the outward phenomena, present themselves immediately and with
a necessity over which the Ego has no command, and thereby take on a stamp
of outward reality; but that the Ego not only recognizes the forms and
categories in which the objects present themselves as its own forms, but
also recognizes the necessity with which
% p. 21
they present themselves as a necessity of the Ego's own nature-ground, is,
in the ``Doctrine of Science,'' the solution of the problem of knowledge.

The critical illusion in Kant consists in this, that the unknown thing
constantly appears as something self-subsistent in itself, objective,
positive—indeed as the only unconditionally positive—notwithstanding that
it shows itself from all sides that this positive can be determined only
negatively. The negative is positive, the positive is negative; this
contradiction, against which the critic defends himself with all his might,
crops up everywhere and vexes the Critique. This illusion Fichte has seen
through with clear consciousness, in that he has recognized the
significance of the negative and ascribed to X an essential value as
non-Ego.

But despite this unmistakable advance, the Fichtean solution is just as
unsatisfactory as the Kantian. What from the objective side is impossible
for Kant, namely to point out how the in-itself indeterminable X can fill
the forms of cognition with an unsurveyable manifold of real
determinations, the same is from the subjective side impossible for Fichte.
As non-Ego, X is far too uniform to yield any explanatory ground sufficient
for the diversity of the phenomena; a stone is a non-Ego, a plant a
non-Ego, an animal likewise: animals and plants have, in relation to the
Ego, the same significance as minerals and stones. The subjective
reflection of Ego and non-Ego consumes all objective relations between the
phenomena among themselves. Kant does at least get a world of experience,
even if on an ambiguous, self-dissolving foundation. Fichte, who removes
the ambiguity of the foundation, does not get even so much; he gets only a
subjectively necessary shadow-world of barriers and non-Ego, a world so
empty that the nature-content vanishes, and the philosophical interest in
the realities of experience is lost.

The Hegelian philosophy lets the problem of knowledge founder in the
absolute knowing instead of being solved. The opposition of the cognizing
subjectivity and the object of cognition, which is the bearer of the
problem, dissolves into the dialectical unity of thinking and being that
flows through the Logic
% p. 22
and issues in the Idea. In the Idea, the Absolute, subjectivity is
sublated, that is, it is transformed into a co-determination, a moment,
even if a moment of totality; and the same holds of objectivity, though
with a noticeable preponderance over the subjective. Instead of an
intuiting, thinking self-being that has the concept for its content and the
concept's reality for its object, we get in Hegel a concept that
comprehends itself, an Idea which, after having given itself subjectivity
in the form of reflection, again sublates this its subjectivity and
translates itself into the form of immediacy as objectivity in existing
objects. The concept, in absolute unity with its content \emph{the Idea},
cannot possibly maintain itself in two opposite forms at once; when the
Absolute is posited as subjective Idea in the form of subjectivity, that
is, of reflection, the objective is reduced to a moment and has vanished;
when the Absolute is posited as objective Idea in the form of objectivity,
that is, of immediacy, subjectivity has vanished: it must then comprehend
itself. But when the objective is to be able to comprehend itself just as
well as the subjective, then the self-comprehending concept is
incomprehensible. If the absolute Idea is to be the concluding unity of
subjective and objective Idea, then it must either assume a form which, in
being neither subjective nor objective, raises it above the extremes, or it
must be one of those two opposite forms in which it holds the extremes
conjoined. The former is, from Hegel's standpoint, inadmissible; for
Schelling's indifference of subjective and objective is only an abstract
presupposition, a non-ground, which stands far below the Ideas, and an
over-subjective or an over-objective has no warrant in ``the absolute
knowing.'' As regards the latter, we must then choose either the subjective
or the objective as the original form of the absolute unity; but the
Hegelian philosophy makes the choice impossible for us. If one compares the
Logic with the philosophy of nature, it appears as though the nature-Idea,
the Idea in the objective form of immediacy, were the derived, and the
Logic's Idea, the Idea in the subjectively reflected form, the original.
But by comparing the philosophy of nature with
% p. 23
the philosophy of spirit we are convinced that the matter stands the
reverse; for logic in logical reflection presupposes spirit. There can be
no logical reflection where there is no spirit; but nature is the
presupposition of spirit, just as immediacy is the presupposition of
reflection. Nature is spirit's ``An sich,'' and nature is the objective; it
is therefore the objective concept that—in an incomprehensible way!—
comprehends the subjective. That Hegel, by the immanent dialectic of the
ideal form-determinations, takes a giant step forward, that by going
further than Schelling he gets negativity developed by another measure and
in a quite different depth than either Kant or Fichte, must be conceded;
but the Absolute itself he does not get cleared up. Despite all the talk of
the self-thinking thinking, the absolutely self-knowing spirit, it is
impossible to learn what the knowing itself, the absolutely knowing,
is—whether it is absolutely subjective or absolutely objective, whether it
is a self or a self-less process or perhaps neither of the two. It is a
self-comprehending concept that is to clear up everything and bring the
last remnant of the unknown to vanish; but far from having eliminated X,
the incomprehensibly self-comprehending has transformed the Absolute itself
into an X, an impenetrable riddle that can fittingly be denoted by an
enigmatic symbol: $XY = X_{x,y}$.

If the problem of knowledge is to be solved, the following fundamental
conditions must inviolably be held fast:

\begin{enumerate}
\item[1)] \emph{There must be an insoluble opposition of subjective and
  objective, in which subjectivity itself, with its a priori forms, is the
  overreaching unity.}
\item[2)] \emph{Cognition must rest upon so thoroughgoing a reciprocal
  determination of the subjective and the objective that no object can
  either be or be thought to be without a corresponding objectivization.}
\item[3)] \emph{That imperfection of objectivization which has its ground in
  the independence of the existing objects from organically conditioned
  subjects must be capable of being remedied by subjective equivalents of
  ontological objectivization.}
\end{enumerate}

% p. 24
Against Fichte and Hegel, Kant is, as regards the first fundamental
thought, unmistakably right when he holds fast the opposition between the
cognizing subject and the object of cognition, just as he also—from the
side of finitude—is right in presupposing an unknown, a thing in itself,
that does not resolve into our cognition. But in the manner in which he
determines the thing in itself he is grossly wrong; for it is not by
excluding the a priori subjective form-determinations that the thing
withdraws itself from knowability\footnote{The more recent Critique has,
however, gone briskly forward, in that it has extended the influence of
``our organization'' to categories such as unity and plurality, substance
and accident, subject and predicate, and so on. We are, it is said,
naturally compelled to apply such categories on account of the peculiar
organization of our understanding; but it does not yet follow from this
that they are in themselves immutable. Enigmatic Critique! That the
understanding must not be conceived as a faculty by itself over against the
other faculties of the soul, that its activity must be dependent on organic
dispositions, can be seen from differing endowment; the understanding of
the individual may be disposed to one kind of intellectual activity without
being in the same degree disposed to another. Nor can it be denied that a
one-sided direction of the understanding may mislead one into fixing
fundamental concepts as exclusively a priori which yet are demonstrably not
so; one may think of the Pythagoreans' philosophical conception of the
decadic system, for example. But such missteps are, after all, to be
corrected by the understanding itself according to the understanding's
laws. Those critics, on the other hand, who give themselves the air of
being able to conceive—not merely an understanding infinitely higher than
the human, for that we can all do—no, an understanding that annuls the
difference between unity and manifold, cause and effect, subject and
predicate, must assuredly distinguish themselves by a quite peculiar
organization. ``Metaphysics'' they might now easily abolish; but how does
it go with the understanding? In order to conceive an understanding that is
absolutely heterogeneous with the human understanding, the human being
must, after all, take leave of his understanding.}, but by its apprehension
being modified according to the subjective peculiarity of the senses and of
the sensing beings. The unknown thing that is the object of sensation is a
thing with sensible properties just as well as the experiential phenomenon
that comes to our consciousness by way of sensation. In sense-systems of
different order, A, B, C, and so on, the same unknown X with its properties
must necessarily present itself in different phenomena. In the A-system the
properties $x_1, x_2, x_3, \ldots$ will present themselves
% p. 25
as $a_1, a_2, a_3\ldots$, in the B-system as $b_1, b_2, b_3\ldots$, and
thus in the first case give the experiential object $P_a$, in the last case
$P_b$. The cognizing beings have in the one system just as much right as
those in the other. But when the cognizing beings must all concede that the
thing itself falls outside their cognition, on what then does the truth of
cognition rest? The answer to this question leads us to the second
fundamental condition.

If the object in itself, with its properties, its concrete determinations,
could really stand independently outside all objectivization, and if the
objective thus standing were the really true, then the cognition of truth
would be impossible. But by working through the relation between Ego and
non-Ego, Fichte has strikingly proved that the objective, far from being
independent, is dependent on and determined by subjectivity. The
exaggeration with which he asserts the pure Ego's original self-limitation,
and the energy with which it itself posits its objects, furnishes a most
necessary counterweight against the vulgar assumption that objectivization
does not really concern the objects, but that it becomes subjectivity's own
affair whether it objectivizes truly or falsely. When we now concede to
Fichte that every real object must be a necessary product of an
objectivizing subjectivity, but cannot possibly concede to him that the
objects really are in themselves as we objectivize them to ourselves, or
that our human objectivization should be able to give the things existence,
then we resort to the third fundamental condition, in order to discover how
the imperfection of the finite objectivization is to be remedied.

The universally valid ground of cognition coincides with the ontological
ground of objectivization. When someone, for example, propounds the
hypothesis of the rotating primordial vapor out of which our planetary
system is supposed to have developed, he certainly does not assume that
that vapor was an object of experience for human sensation, any more than
his opinion is that what was then ``a thing in itself'' has since, through
our experiences, been transformed into vapor. Nor does he assume either
that the primordial vapor
% p. 26
existed in our subjective intuition of space and rotated in our subjective
intuition of time, nor that it is we who with a priori necessity have
prescribed to nature the laws according to which the whole development has
taken place. No, he rather assumes that all of it—the space, the time, the
vapor, the motion—was objective without any objectivization whatever. Since
it is now proved that the last assumption is just as meaningless as the
first, what follows from this? That here in silence there must have been
slipped in a subjective equivalent of ontological objectivization!
Hypothetically this is expressed thus: if we had been present and been able
to survey the whole, then the matter must necessarily have appeared as the
description runs. Thus: every time our limited subjectivity goes beyond its
own bounds of experience, it becomes, for the sake of the objective,
compelled to presuppose an intuiting that is not our intuiting, a seeing
that is not our seeing, a thinking that is not our thinking, an
objectivization that is not our objectivization; it becomes, in short,
compelled to presuppose an ontologically objectivizing subjectivity.

From the ontological fundamental law of objectivization, symbolized by
$XY = O$, where O denotes the object, it follows: that every objective
moment must necessarily presuppose a corresponding subjective one, since
$Y = 0$ gives $O = 0$, and that conversely every subjective moment in a
real objectivization must be determined by a corresponding objective one,
since $X = 0$ gives $O = 0$. Subjective and objective are, in the
ontological sense, correlative concepts, but not coordinate: the
objectivizing subjectivity is, as the overreaching unity of the opposition,
a \textit{dignitate prius}.

From the fundamental law it follows further that the truth of cognition
rests neither on the objective by itself nor on the subjective by itself,
but on the reciprocal determination of both. Let there be thousands,
indeed millions, of diversely constituted sense-systems in the universe,
still the a priori forms of cognition are the subjective fundamental forms
of universal reason, and universal reason is common to all cognizing
beings. The cognizing beings that, with the same sense-system, come under
the same
% p. 27
order must see that the concrete objectivization conditioned by sensation
has all the truth it can and should have, when the reciprocal action
between the subjective and the objective is precisely determined for the
order in question. The dissimilar sense-systems are just as little able to
awaken doubt about the universal validity in the peculiarity of
objectivization as the red-blind man, despite the truth of his
objectivization, is able to awaken doubt about the difference of colors
among those who have color-sense. The truth is that the reciprocal action
between subjective and objective must be different for subjectivities of
different order; but \emph{the reciprocal action itself is the truth}. To
assume an X standing outside the reciprocal action, unobjectivized, for the
really true, is a dream, an imagining; outside the reciprocal action there
is no truth, for outside the reciprocal action there is, ontologically
understood, nothing at all.


% ============================================================
%  II. THE PROBLEM OF REALITY
% ============================================================
\section*{II.\ The Problem of Reality}
\label{sec:realitet}

% p. 27 (heading printed here; section II text begins after it)
\noindent
While Hegelianism has sought to transform the world of realities into a
system of metaphysical concepts, Positivism has, for a change, thrown all
metaphysical ballast overboard, in order to hold exclusively to realities
and facts. It is, as is well known, not the realities that conform to our
concepts, but conversely our concepts that must conform to the realities;
the wisest thing philosophy can do is therefore to give up the a priori
problems, abandon its old leaky vessel, go on board the steamer of physics
as a passenger, and assume the title of Positivism. In justification of so
important a step, the school's celebrated founder, Auguste
Comte\footnote{\textit{Cours de Philosophie Positive; Tome 1mier, Paris
1830. 1miere leçon.}}, has pointed out three stages of development
(\textit{trois états}) which the human mind must necessarily pass through on
its way to the cognition of truth, \emph{the theological, the metaphysical,
and the scientific}. At the theological stage one seeks the cause of the
phenomena in a supernatural will and
% p. 28
lets the arbitrariness of the supernatural will determine the course of
things without regard to fixed laws. At the metaphysical stage the
arbitrariness disappears and gives place to such supersensible, self-valid
entities as the Absolute, Substance, the Cause, the Idea, and so on. This
intermediate stage has been necessary as a transitional step from the
theological to the scientific\footnote{\textit{La théologie et la physique
sont si profondément incompatibles, leurs conceptions ont un caractère si
radicalement opposé, qu'avant de renoncer aux unes pour employer
exclusivement les autres, l'intelligence humaine a dû se servir de
conceptions intermédiaires, d'un caractère bâtard, propres, par cela même, à
opérer graduellement la transition. Anfrt.\ Skr.\ S.\ 13.}}; but in itself
the metaphysical standpoint is barren, as can be seen particularly from the
endless, resultless disputes into which it dissolves. The positive
philosophy, which stands at the scientific stage, now characterizes itself
by this, that it regards the phenomena as subject to unchangeable laws,
seeks to reduce these laws to the smallest possible number of fundamental
laws, and considers every attempt to penetrate either to the first causes or
to the so-called final causes as utopian and
meaningless\footnote{\textit{Nous voyons, que le caractère fondamental de la
philosophie positive est de regarder tous les phénomènes comme assujétis à
des \emph{lois} naturelles invariables, dont la découverte préçise et la
réduction au moindre nombre possible sont le but de tous nos efforts, en
considérant comme absolument inaccessible et vide de sens pour nous la
recherche de ce qu'on appelle les \emph{causes}, soit premières, soit
finales. Anfr.\ Skr.\ S.\ 14. Further, S.\ 16: La preuve manifeste de
l'impossibilté d'obtenir de telles solutions, c'est que, toutes les fois
qu'on a cherché à dire à ce sujet quelque chose de vraiment rationnel, les
plus grands esprits n'ont pu que définir ces deux principes --- l'attraction
et la pesanteur --- l'un par l'autre, en disant, pour l'attraction, qu'elle
n'est autre chose qu'une pesanteur universelle, et ensuite, pour la
pesanteur, qu'elle consiste simplement dans l'attraction terrestre.}}.

To exclude metaphysics is fully justified when by the word metaphysics one
denotes a science that puts imagination in the place of experience in order
to discover the essences and \textit{qualitates occultæ} of which real
things must absolutely consist, unconcerned whether it agrees with reality
or not. But against so superficial a conception of the concept of
metaphysics one must protest in the name of science. Hegel's
``Wissenschaft der Logik'' is essentially metaphysics. Whatever imperfec-
% p. 29
tions may be pointed out in this Logic, one should nevertheless not accuse
the author of fetching his concepts from the world of imagination. The
concepts are so inherent in the things that no existing thing can escape
corresponding to its concept. That ``cultivation and discipline'' of
consciousness\footnote{``Das System der Logik ist das Reich der Schatten,
die Welt der einfachen Wesenheiten, von aller sinnlichen Konkretion befreit.
Das Studium dieser Wissenschaft, der Aufenthalt und die Arbeit in diesem
Schattenreich ist die absolute Bildung und Zucht des Bewusstseyns.'' Hegel.
Wissenschaft der Logik. Berlin 1833. Einleitung. S.\ 47.} which Hegel
demands on behalf of Logic he can with full right demand on behalf of all
the sciences. --- The relation between physics and metaphysics we may
determine thus: physics forms its concepts according to the observed facts,
metaphysics develops its concepts out of the conditions necessary for all
being and existence. To deduce the concretions of the concepts of existence
from pure thought is impossible; but to give an account of empirical
phenomena without going back to the a priori concepts is also impossible.
Gold, silver, iron are called partly different \emph{things}, partly
different \emph{substances}; what use is it now to denote reality, the given
fact, now by the word \emph{thing}, now again by the word \emph{substance},
when one is not in agreement with oneself as to wherein the designation
substance differs from the designation thing? In science one must after all
know what one is oneself saying. In order to be a \emph{definite thing}, the
phenomenon must absolutely be a \emph{thing}; in order to be a \emph{definite}
substance, it must be a \emph{substance}. On this point physics has the
advantage of consulting metaphysics. In metaphysics, the objective logic, it
is not only made evident that the thing must necessarily reflect itself in
the properties, and the properties in the thing, the substance in the
accidents and the accidents in the substance; but it is also made evident
wherein the reflection consists, and how different it is in the case of the
thing and in that of the substance.

From this standpoint it will be instructive to see two such strong thinkers
as Comte and Hegel try their strength with
% p. 30
one another. Comte maintains that there is no Absolute; and he is right
insofar as he looks only at existence and the existing. In the world of
facts the one term stands against the other, the one atom exists beside the
other. Everything is parceling-out, everything is relation. Hegel maintains
that the only thing that in truth is, is the Absolute; and he is right
insofar as he looks at the original unity of being with universal thought.
The relative is not merely relative in relation to another relative; but all
relations and relative terms have their subsistence in an encompassing
whole. If the encompassing whole is itself relative, it must necessarily
have its subsistence in relation to the relative term. But in this
thoroughgoing relation it then relates, through all relativities, to itself,
and precisely thereby comes to subsist as the Absolute. Comte maintains that
all that is logical is subjective; and he is right insofar as he thinks only
of formal logic, for formal logic with its abstractions is undeniably
subjective. Hegel maintains that the logical is essentially objective; and
he is right insofar as he starts from the essential unity of thinking and
being. The determination of thought to which no being corresponds is just
as empty as the determination of being to which no thought corresponds.

Comte assumes that thing and reality can very well be determined in
themselves and present themselves to us without any objectively determining
concept; but here he is wrong. We abstract, as it is said, our concepts from
the things; but to abstract concepts from things in which there are no
concepts is just as impossible as to pluck hairs from the bald man. Hegel
assumes that the objective concepts in their immediacy are themselves
things, and that it must therefore be possible to deduce existing facts from
logical universal concepts; but here he is wrong. How unhappily every
attempt at a construing of real things \textit{a priori} must turn out, his
philosophy of nature has more than sufficiently demonstrated.

Comte teaches that reality is reality, but not concept; Hegel teaches that
the concept as concept is reality: in the clearly
% p. 31
expressed opposition between Positivism and absolute Idealism the
\emph{problem of reality} can properly first be said to arise.

The problem is: that thing and concept must be assumed both to be the same
and yet not the same.

That thing and concept must, in an essential sense, be the same can easily
be seen. Take whatever thing or existing object you will, and you will
convince yourself that the sum of all determinations—be they material or
ideal, sensible or unsensible—must, since the determinations characterize
the object and make the thing what it is, necessarily belong to the object's
concept. In its sum of determinations the object subsists; with a change in
the determinations the object changes; with the dissolution of the
determinations the object disappears. When it nevertheless seems as though
there lay something more in the object itself than in its concept, this
derives simply from a confusion of the object's concept with our concept of
the object; for our concepts of concrete objects are far from being
exhaustive. That gold's color, malleability, specific gravity, form of
crystallization, slight affinity for oxygen, and so on, are properties that
characterize this metal and to that extent belong to the concept of gold,
there is surely no one who will deny. When we must nevertheless admit that
we have only an imperfect concept of gold itself, the reason for this is
easy to explain. We are, after all, still far from knowing all of gold's
properties, and even if we knew them from experience, we could still not
boast of knowing gold's substance so long as we are not able to derive the
properties from one another mutually or from the common unity in which they,
as accidents, have their subsistence. To be yellow, to be malleable, to
crystallize in octahedra, to have affinity for chlorine, cyanogen, and so
on, are in reality such dissimilar determinations that no mortal is able
either to point out their law-determined dependence on one another mutually
or to discover the real unity from which they all with necessity proceed.
And yet physics must concede that color, malleability, crystal-
% p. 32
form are just as much functions of the gold-atoms' quality and arrangement
as gold's unequal affinity for chlorine and oxygen; in other words: physics'
conception of the laws is not limited to certain relations between thing and
thing in outward generality; but every thing with its properties, every
whole with its parts, every substance with its accidents, every body with
its molecular structure expresses a system of law-determinations in which
term and relation reflect one another. Even the smallest parts can, in the
all-sided reciprocal action, just as little fall outside the reflection
comprehending its moments as they can fall outside the laws. Since now all
determinations, both of the particular terms and of the terms' mutual
relations, enter as real moments into the determining reflection, and the
totality of the reflected determinations is one with the concept, it follows
from this that thing and concept must be the same.

And yet concept and thing are by no means the same; they are, on the
contrary, completely opposed to one another. What a difference between the
concept of gold and the existing gold! The existing is immediately a fact,
an object of experience; but the concept as concept tolerates no object of
experience, no residues of immediacy, whether of sense-impressions or of
images or of representations: everything must be dissolved, melted, purified
in that fire of negativity which gives the concept its ideality's
unsensibleness and logical distinctiveness. From this standpoint Comte
undeniably has ground to make a divorce between concept and fact; for what
holds of the fact does not hold of the concept. One can crush a stone, but
not the stone's concept; one can boil a plant, but not the plant's concept;
one can shoot a bird, but not the bird's concept; the concept can be
thought, and one can think \emph{of} the thing, \emph{about} the thing, but
the thing as thing cannot be thought. The concept, the universal, can be
expressed in words, but the thing's individual determinateness must be
pointed at as at a This and This.

The immediacy that distinguishes thing from concept cannot be logical; for
even though the logical can point beyond itself, this always happens within
the sphere of logic: Hegel's
% p. 33
Logic is an eloquent example of this. Nor can the immediacy be illogical; if
the object-determinations were illogical and stood in contradiction with the
logical, the object would either be concept-less or stand in contradiction
with its concept. Nor can the immediate in the object be something that is
added or joined from outside to the logical, as when in certain logics a
distinction is made between the object itself and the concept expressed in
the object. What indeed should this something be? The material, matter? But
if matter with all its determinations falls outside the concept, how then
can the concept be expressed in the material object?

When now the sought immediacy can be neither logical nor illogical nor
joined to the logical in an outward manner, then it must surely be the
not-logical, which the logical needs, a necessary complement with which it
is saturated, in that it disappears; this complement is the antilogical.

If we seek the logical root from which the antilogical springs, we find it
in the category of power. Power in its immediacy is the expression of
reality, and the expression of reality's immediacy is again the antilogical.
The reason we regard the objects as realities and ascribe to them an
immediate, objective self-subsistence is precisely the impression of the
power with which they, each singly, come to meet us. The objects of
experience appear in this light like great and small potentates, who
maintain their place in proportion to the power they have; only that, for
the sake of immediacy, we must add that, without all logical reflecting,
they are what they have, and have what they are. In apprehending facts,
things, and realities from this standpoint, we have, in consequence of the
logical, come out beyond the logical. The opposition of thing and concept is
so striking that, when the thing is posited, the concept has disappeared,
and when the concept is posited, the thing has disappeared.

If we now go back to the fundamental law of ontological objectivization,
which was treated under the problem of knowledge, then it becomes possible
for us to point out the ground both of Positivism's
% p. 34
and of Hegelianism's one-sidedness. In that the Positivists immediately make
straight for facts and real objects, they entirely forget to examine the
relation between object and objectivization. The law says: as the object is,
its objectivization must be; as the objectivization is, the objectivizing
subjectivity must be. Since the adherents of the positive philosophy quite
overlook this law, they cannot, with all their interest in the exact
investigation of reality, of course get their eyes open either for
objectivization in power or for the objective real-concepts by which the
realities are \textit{a priori} determined. They are zealous for the
cognition of the unchangeable laws, without a suspicion that the
unchangeableness of the laws has its ground precisely in the system of
real-concepts. The best proof of this is the fact that Stuart Mill, whose
inductive logic has set empirical induction in the clearest light, himself
admits that the laws do not demonstrably belong to the essence of things,
that one may very well imagine regions of the heavens where ``events follow
one another by pure chance without any fixed law.'' Just as little have the
Hegelians, despite their absorption in the dialectic of the objective
concepts, been attentive to the fundamental law of objectivization. They
have categories such as ``Becoming,'' ``Appearing,'' ``Mediation,''
``Passing-over into its opposite,'' and the like; but the category of
objectivization is not to be found in the whole system. Had Hegel but once
fixed his sharp, penetrating gaze on the problem of objectivization, then we
should surely have got another explanation of the transition from the
subjective to the objective than is the case in that part of his Logic which
he calls the subjective\footnote{``Der Schluss ist \emph{Vermittelung}, der
vollständige Begriff in seinem \emph{Gesetztseyn}. Seine Bewegung ist das
Aufheben dieser Vermittelung, in welcher nichts an und für sich, sondern
jedes nur vermittelst eines Andern ist. Das Resultat ist daher eine
\emph{Unmittelbarkeit}, die durch \emph{Aufheben der Vermittelung}
hervorgegangen, ein \emph{Seyn}, das eben so sehr identisch mit der
Vermittelung und der Begriff ist, der aus und in seinem Andersseyn sich
selbst hergestellt hat. Dieses Seyn ist daher eine \emph{Sache, die an und
für sich ist}, --- die Objektivität.'' Hegel. Die subjective Logik. Berlin
1834. S.\ 170--171. When the subjective movement through mediation has
reached objectivity, and the immediate totality of being-determinations
stands before the logical consciousness as a factually given, it shows
itself what kind of subjectivity it was that comprehendingly found the
object. It is by no means the substance comprehended in the causal relation
that has transformed itself into judging, inferring, mediating subjectivity
and thereby produced the objective; it is quite simply a human reflecting
subjectivity, which has assured itself that there must be concept in the
world of objects, and now --- without in the least seeing how this concept
has become object, contenting itself with the assurance that the object is a
concept --- regards it from the outside, and seeks therein to find and
recognize the formal determinations that the subjective logic has developed.
Here there can be no question of ontological objectivization. --- ``Darum
ist das Erklären der Bestimmung eines Objekts, und das zu diesem Behufe
gemachte Fortgehen dieser Vorstellung nur ein \emph{leeres Wort}, weil in
dem andern Objekt, zu dem sie fortgeht, keine Selbstbestimmung liegt. Indem
nun die Bestimmtheit \emph{eines Objekts in einem andern liegt}, so ist
keine bestimmte Verschiedenheit zwischen ihnen vorhanden; die Bestimmtheit
ist nur \emph{doppelt}, einmal an dem einen, dann an dem andern Objekt, ein
schlechthin nur \emph{Identisches}, und die Erklärung oder das Begreifen
insofern \emph{tautologisch}. Diese Tautologie ist das äusserliche, leere
Hin- und Hergehen; da die Bestimmtheit von den dagegen gleichgültigen
Objekten keine eigenthümliche Unterschiedenheit erhält, und deswegen nur
identisch ist, ist nur \emph{Eine} Bestimmtheit vorhanden; und dass sie
doppelt sey, drückt eben diese Aeusserlichkeit und Nichtigkeit eines
Unterschiedes aus. Aber zugleich sind die Objekte \emph{selbstständig} gegen
einander; sie bleiben sich darum in jener Identität schlechthin
\emph{äusserlich}.'' Die subjektive Logik. S.\ 183--184. Consistently, the
development of the concept under this presupposition ought to have gone the
way of experience, to have penetrated the method of natural science and
thereby secured for itself the concepts' expression in facts; but instead
the illusion of the self-comprehending concept giving itself objectivity
gained the upper hand to such a degree that one, after having intoxicated
oneself in the dialectic of the immanent method, has now quite given up the
dialectical --- although dialectic is precisely a thoroughgoing critique of
ideality --- and is close to forgetting ``die absolute Bildung und Zucht des
Bewusstseyns,'' which philosophy can never dispense with, and which for so
essential a part is nevertheless owing to the great master.}. If it is
already from a logical standpoint evident that the object's concept cannot
possibly objectivize itself, that self-objectivization in the objectivization
of another is and must be subjectivity's ineffaceable characteristic mark,
then it follows irrefutably from this that a subjectivity which in thinking,
reflecting, and comprehending holds itself strictly within the boundary of
the logical will never be able to carry it further than to a logical
objectivization of logical objects.

% p. 36
In the real world the objects are not logical; they are real, immediate,
antilogical: their self-subsistence consists in immediate power, their
mutual relations are power-relations. What now lies in this? As the objects
are, the objectivization must be: power-objects must be objectivized in
power; further: as the objectivization is, the objectivizing subjectivity
must be: only an absolutely powerful subjectivity is able to let realities
proceed from their possibilities and bear a world of objects that really
stands by power. In the objects power is immediate as power in another; in
the objectivizing subjectivity power is reflected as self-power. Self-power,
the subjective power, is ideal and logical; the object-power, the objective
power, power in another, is real and antilogical. The opposition of ideal
and real power is the strongest, deepest opposition that can be thought.
Here is a real dualism; but the dualism is grounded in original unity, for
the absolute self-power has the opposition within itself and masters its
opposition.

The age-old question of the origin of the world-matter thereby comes under a
new light. Aristotle lets, as is well known, matter and form be equally
eternal, both belonging to that cycle of movements in which possibility
alternates with actuality, and actuality with possibility; matter is the
possibility, form the actuality. But this explanation is not appreciably
better than the more recent Materialists' assurance that matter and force
are equally eternal. What indeed is it, to be eternal, if it is not to be
the absolutely original? Absolutely original is, after all, only that which
has the ground of its origin in itself. But how is a being to be able to
have the ground of its origin in itself, if it does not at the same time
have a ground of origin for something other than itself? Aristotle describes
the first mover, which is not moved, the Godhead, as the being that does
nothing other than eternally to think the thinking of thinking (αὑτὸν ἄρα
νοεῖ \ldots{} καὶ ἔστιν ἡ νόησις νοήσεως νόησις)\footnote{Translator's note:
the Greek (Aristotle, \emph{Metaphysics} XII.9) reads roughly ``it thinks,
then, itself\ldots{} and its thinking is a thinking of thinking.''} and
which, in order not to stain its perfection with anything imperfect, must
exclude the possibility of being-other. But to exclude the possibility of
% p. 37
being-other is to exclude matter. That such a conception of the original
cannot withstand a critical test is easy to point out. If the eternal being
is to think itself, then this must necessarily happen by a movement of
thought in which it holds itself fast by distinguishing itself from something
other than itself; merely in order to think itself it needs an other. If,
moreover, its thinking is to be one with an active, real being, then this
active self-being must necessarily contain the possibility of being-other;
but the possibility of being-other is precisely matter. An eternal
self-being without matter-being, an eternal matter-being without self-being,
are hollow abstractions. The opposition of intellectual and real shows
itself here in its first ontological shade. One may now, with Spinoza,
attempt to lay the all-reality into the being's universal unity, conceive
the self-being as the infinite substance (\textit{causa sui}) with opposite
attributes, thought for the ideal and extension for the material, and reduce
the finite things to modifications, limitations, negations of the infinite:
it does not help. The abstraction sucks the blood of the substantial, so
that there remains neither self-being nor other-being, neither ideality nor
reality, that has any self-subsistence to show. The positivistic way out, to
draw a line through metaphysics and hold to the realities of experience, is,
as shown above, impassable; unless the decisive step is to be taken in all
thoughtlessness. For as regards the way out of thoughtlessness, there holds
of metaphysics what Hegel somewhere says of reason: the best way in which one
gets rid of it is: not to get involved with it at all.

If one is to hold fast reality without giving up thought, then the
ontological fundamental law holds, that the subjective, self-reflected ideal
must have the original power in order that the objective, the immediately
real, may have the derived. An equalization of the mighty opposition between
self-power and power in another cannot take place by subtraction and
addition. One does not increase self-power by drawing something from power
in another; nor can one strengthen the real by weakening or diminishing the
ideal. On the contrary! The greater the reality the
% p. 38
objective has, the mightier must the ideality be with which it is posited
and determined. The objective world of facts, of realities, is, in the
scientific sense, the strongest expression of the ideal power with which the
ontologically universal self-being masters and comprehends itself in all and
all in itself. All scientific truth stands and falls with the fundamental
principle that the finite subjectivity, which objectivizes in impotence,
must everywhere have its ontological equivalent in an infinite one, which
objectivizes in omnipotence.

The transition from ideal to real, from concept to thing, corresponding to
objectivization, has no intermediate determinations and can have none. How
indeed should these intermediate determinations be constituted? Logical they
cannot be, for one may enrich a concept with as many logical determinations
as one will, it still remains, as concept, within the logical. Antilogical
they cannot be either, for the antilogical enters only when the transition
has taken place. The transition must then be a power-movement, a conversion
of power's ideality into real power, a leap by which nothing is leapt over, a
\emph{logical-antilogical outburst of power}. Such an outburst of power is
originally one with the transition from \emph{possibility to
actuality}—understood, be it noted, in such a way that the possibility lies
in the concept, hence in the logical---\textit{potentia ab
posse}\footnote{Translator's note: the print here reads ``polentia af
posse'' (in roman), evidently a misprint for \textit{potentia ab posse},
``potency from being-able-to-be.''}---the actuality in the antilogical
reality, hence outside the logical. Let us see it in some examples. He who is
to walk with a light in a powder-magazine must be careful: he has an
explosion to fear. That the explosion is a transition from possibility to
actuality, an outburst of power, surely no one will deny. So long as the
powder remains in actuality, in antilogical reality—saltpetre, charcoal, and
sulfur in mechanical mixture—the explosion is only a possibility that lies
hidden in the powder's actuality; but when the explosion becomes actual, the
powder becomes unactual. The oxygen from the sulfate of potash unites with
the carbon into carbonic acid, the nitrogen develops in gaseous form, the
saltpetre's potassium forms with the sulfur potassium sulfide—in short: the
actual powder disappears; but the powder's concept, the real-concept powder,
remains behind in
% p. 39
the elements as possibility. The \emph{sudden} is here only a sensible
indication of the \emph{qualitative}. The possibility of hydrogen chloride,
a mixture of chlorine and hydrogen, can, by being placed in sunlight, show a
sudden turn into actuality. But the same transition can also take place
slowly, when the mixture stands in daylight; the transition can then be
regarded as a successive summation of manifold small, indeed infinitely
small, outbursts. The outburst rests on the qualitative: either the hydrogen
chloride's possibility or its actuality, in other words: either the ideal or
the real, either possibility or actuality, either concept or reality; an
intermediate there is not. To wish to condense the concepts into realities
or sublime the realities into concepts is a speculative bungling, in equal
degree contrary to experience and sound logic.

The whole real world rests upon an original outburst of power. If one
rejects the unity-power, the all-comprehending self-power, then the system
of real-concepts disappears and with it all the possibilities of reason. If
one assumes the self-power, then there follows with necessity, as lies in
its concept and comprehension in power, an outburst of antilogical
realities. The characteristic mark of the outburst's antilogical reality is,
seen from all sides, \emph{otherness}. The elements of actuality have a
double-sided stamp of otherness. That they are the other of self-power is
seen from this, that they constitute the real manifold, the swarm of atoms,
the parceling-out of immediate power through which the mighty unity confirms
itself. That there really is reality in this otherness is seen from this,
that there are objective relations, that the elements relate to one another
as other by other to other ad infinitum. We see it in the relation between
matter and force. Matter with its atoms is a striking expression of power in
another. It appears, on a superficial view, as though the atoms' existence
in space were only a collective coexistence with no other relation than that
which lies in the spatial relations. Consistently the Positivists must,
presumably, acquiesce in this assertion. By empirical induction one is
supposed to be able to convince oneself that the one atom-group follows
% p. 40
upon the other---usually!---according to a definite law. By taking motion as
a fact and appealing to the law, one gets free of force; that is, force is
transformed into a mathematical symbol, for example into the second
differential coefficient in accelerated motion. At this standpoint, however,
Materialism cannot maintain itself. The active relations between atoms and
atom-groups are too conspicuous for one to be able to escape the impression
of a real attraction and repulsion. The active relations now acquire a gleam
of reflection, and the reality of reflection makes itself known in force.
Then it is said that force, for example the force of gravity, is a property
of matter. But when force, the reflected-in-relation, is the active, then it
must surely also be the really actual, and what then is matter? One does not
rightly know whether it is matter that acts by virtue of its property,
force, or whether force is perhaps something particular, a kind of reflected
reality that adheres to matter, the immediate reality. The last assumption
is not satisfactory; so one goes further, draws the immediate reality under
reflection, makes matter itself into a phenomenon of forces, sublimes it
into forces and force-points. But---\textit{naturam furca expelles}---the
antilogical immediacy asserts its right; one is nevertheless compelled to
recognize a side of force that is \emph{opposite} to force, and to grant
matter, the immediate, self-less otherness, its inertia.

The unity-expression for matter and force is power in another, understood in
such a way that matter, the immediate in power, yields the existing terms,
and that force is reflected in the active relations. In matter the concept
founders, the existing stands in the concept's place; but that the
antilogical in existing matter is nevertheless logically determined by its
real-concept can be expressly known from force. That two masses existing in
space can attract one another at a distance is certainly incomprehensible
when one throws the concept away; the representation of an attractive force
adhering to matter gives no explanation. Only when one takes the
real-concept along does the
% p. 41
matter become intelligible. The existing masses are only one-sided
expressions of power in another. One supposes that one has reality's
complete expression in each of the masses by itself, and is astonished that
two such realities, which, as it seems, need nothing but the space they each
fill, can have anything to do with one another. That is an illusion. To
existing terms' actuality there belong active relations; the attractive
force, an expression of active relation, belongs inseparably in the
real-concept in which the masses' reality is originally grounded.

Let us pause a moment to hear an objection. From the standpoint of physics,
and then especially of Positivism, it must surely hold that attraction is a
fact, and that the discovery of the law of attraction, independent of any
analysis of metaphysical real-concepts, is owing solely to empirical
observation, calculation, and induction. To give the matter an appearance as
though that which depends exclusively on experience could also be
constructed \textit{a priori} is a barren shadow-boxing, a relapse into
antiquated metaphysics. This objection would be completely justified if it
did not have its ground in a misunderstanding. Here there is no question at
all of deriving real facts from our own imperfect, impotent concepts; on the
contrary! our task is to lead the facts of experience back to the
power-concepts, without which they, in their antilogical reality, would
stand before consciousness as blind facts. If it is in conflict with the
scientific method that one, with the help of experience, strives to derive
the laws of things from their concepts, how then will one defend the
procedure of mechanical physics? It is, after all, evident that the laws for
the impact of hard and elastic bodies cannot possibly be found by way of
experience alone, since no body is either perfectly hard or perfectly
elastic; they must necessarily be derived from the real-concepts: hardness
and elasticity.

From the ideal side the real-concepts are reflected in the all-comprehending
self-unity; from the real side they are inherent in the things. It is the
latter that in the present connection especially con-
% p. 42
cerns us. The actual contains possibility to another actual; the actual
gases oxygen and hydrogen contain possibility to water. In order that the
possibility may be actualized, conditions are required; the conditions have
the actual's immediacy and yet correspond in reflection to the possible.
Chemistry has examples in abundance. When the salt is decomposed, that is,
goes back into its concept, its possibility, the acid and the base step
forth in antilogical reality, and conversely; nothing but outbursts of power
in another. The real-concepts have a peculiar affinity for the antilogical;
this affinity is originally determined by the relation between power in
another and self-power. The chemist may concede that the oxygen and the
hydrogen in a certain way perish in the water, namely insofar as by the
combination they give up their particular properties, so that a new body
with new properties arises; but that this can be regarded as a return of the
realities to their real-concepts he will hardly concede. He will, in
consequence of the atomic theory, on the contrary reject the transformation
and assert the difference of the composite atoms from the simple atoms. The
water-atoms are decomposable, for each water-atom consists of one oxygen and
two hydrogen atoms; but the atoms of oxygen and hydrogen are, as simple
atoms, indecomposable. The ontological reversal from reality into concept,
from concept into reality, becomes by this mode of consideration highly
suspect. Yet wrongly. It is finitude's, otherness's, real discretion on
which the emphasis falls, but this real discretion is so far from conflicting
with the real-concept's logical essence that it on the contrary confirms it:
only by the discretion's antilogical elements entering as reflexes into
mighty comprehension does the concept acquire the character of real-concept.
The composite atom cannot possibly be regarded as a mere arithmetical sum of
two or more simple atoms; for the sake of aggregation alone a mutual
attraction must here be presupposed. Even if the attraction were—what, on
account of the peculiarity of the affinities, it cannot be—merely
mechanical, then the mechanical attraction, according to what we have seen
in the foregoing, would alone be sufficient to assert the concept's va-
% p. 43
lidity. The oxygen and the hydrogen could not possibly combine chemically if
they had no affinity for one another; but mutual affinity they could not
possibly have if their real-concepts did not correspond to one another; but
the real-concepts could not possibly correspond to one another if they were
not originally comprehended in one and the same comprehension. In the
oxygen, the hydrogen, and the water we have three real-concepts, two
co-concepts and a concretion-concept determined thereby. When the
co-concepts exist with antilogical reality, then the concretion-concept is a
possibility; when the concretion-concept exists, then we have in this
reality the co-concepts' possibility. The transition from possibility to
actuality presupposes conditions, but takes place by an outburst of power.

It has been intimated above that the positivistic assurance of the
immutability of the laws of nature stands, according to St. Mill's
admission, on rather weak feet. But even if this weakness is overlooked, even
if it is conceded that a merely empirical induction can give us a maximum of
probability that a B will invariably follow upon an A, is any essential
insight thereby gained? By no means. There is a difference between
\emph{certainty} and \emph{insight}. Suppose the relation between the
hypotenuse and the legs had been found practically by a measurement of the
triangle's sides and squares constructed thereon, then repeated measurements
of right-angled triangles could indeed give a certainty about equal to that
which a geometrical proof gives; but only the proof gives insight, the
measurement by itself gives no insight. All insight is conditioned by this,
that the representation of the one encloses the requirement of the other,
that the one cannot be posited without the other having to follow, that the
reality of the one bears within it the concept of the other. If the
hypothetical proposition ``if A occurs, then B must follow'' is held fast
merely empirically, then there is here no insight, however great the
certainty may be. One will ground the phenomenon by enumerating its
conditions, but however many conditions and intermediate determinations one
enumerates, one still does not reach the ground; for it stands with the
conditions as with the conditioned: the individual conditions are themselves
phenomena. Instead of a grounding of the
% p. 44
phenomenon one reaches only a summation of phenomena, which each in turn
must be grounded by a summation of phenomena; the grounding is a continued
demand for grounding, but the ground flees away into infinity: insight is
never reached. One can, with a maximum of probability, get certainty about
many things and about the connection of many things, without however
attaining insight, rational insight into any thing.

A rational insight into the connection of things is only possible under the
presupposition that there is reason in the things, that the things are
realized concepts. The concept realized in the thing is the thing's nature.
That every thing acts according to its nature means that every thing acts in
consequence of its concept; but what is the acting cause? For this question
the problem of concept and reality presents a new side. The concept seems
not to be able to be cause: the concept of gravity does not weigh, the
concept of fire does not burn, the concept of a stone kills no dog. It must
then be the thing that is cause. But when the thing can act only in
consequence of its concept, then the concept must surely be the cause of the
thing's being a cause. What does this lead to? Quite directly to a
distinction between two kinds of causes, ideal and real, and thereby to a
more penetrating investigation of the relation between \emph{ideal and real
causality} on the whole. Such an investigation is of importance for the sake
of insight and scientific cognition. To make matter and force into causes,
without troubling oneself further either about the concept of matter or about
the concept of force, is an easy matter; to make the real cause logically one
with the concept of the cause is not exactly difficult either; but in the
first case one excludes all insight, in the last all reality.

The logical-antilogical outburst of power, which transforms the concept into
reality, transforms at the same time the ideal causality into the real.

That the cause can act only by virtue of the unity of concept and reality is
seen from this, that acting and activity can take place only in consequence
of the same unity. The thing as thing in antilogical immediacy cannot act.
And why not? Because
% p. 45
immediacy, sharply and exactly held fast by itself, means that the concept's
content, and therewith possibility to another, is exhausted in the thing's
existence; the only thing the thing in this case is able to do is: to exist.
With exhausted concepts the things would exist side by side, excluding one
another, but for the rest have not the least to do with one another.
Herbart's absolutely positive reals can serve as an example of this. This
acute thinker, persistent in exact analyses, suffered from the fixed idea
that unconditionally independent reals should be able to be affected by one
another without the positive in them betraying anything negative. A curious
contradiction\footnote{The propensity to give empty verbal explanations
instead of a real interpretation of facts, and the weaknesses of which
metaphysicians, for the sake of the system alone, often make themselves
guilty, F. A. Lange (Geschichte des Materialismus und Kritik seiner
Bedeutung in der Gegenwart. Iserlohn. 1866. S.\ 463.) has aptly
characterized: ``Das absolut Einfache ist auch keiner inneren Veränderung
fähig, weil wir uns diese nur in der Form wechselnder Ordnung der Theile
denken können. Deshalb sagt Herbart auch nicht, dass die Realen auf einander
wirken, sondern dass sie Einwirkungen von einander erleiden \emph{würden},
wenn sie diesen nicht durch einen Akt der \emph{Selbsterhaltung} Widerstand
leisteten. Als ob damit etwas andres gesagt werden könnte, als mit der
Annahme einer einfachen Wechselwirkung! \emph{Waitz} legt in seiner
Psychologie (S.\ 81) viel Werth auf den Unterschied zwischen
\emph{Dispositionen} zu einem Zustand und wirklichen Zuständen. So geht es in
der Metaphysik. Zustände darf die Seele nicht haben; bei Leibe nicht, sonst
ginge ja ihre absolute Einheit verloren! Aber Dispositionen, das ist ganz
etwas andres; „Strebungen", warum nicht? Der Metaphysiker widerlegt mit
einem enormen Aufwand von Scharfsinn alle möglichen andern Ansichten, und wo
er seine eigne Meinung entwickelt, schiesst er einen logischen Purzelbaum von
der gewöhnlichsten Sorte. Jeder Andre sieht, dass eine Disposition zu einem
Zustand auch ein Zustand ist, dass Selbsterhaltung gegen eine
\emph{drohende} Einwirkung nicht ohne eine, wenn auch noch so feine
\emph{wirkliche} Einwirkung denkbar ist. Der Metaphysiker sieht dies nicht.
Er hat sich mit seiner Dialektik an den Rand des Abgrundes getrieben, alle
Begriffe hundertmal herumgewendet, hervorgezogen, weggeworfen, und endlich
\emph{muss durchaus und durchaus etwas gewusst werden}. Also die Augen
zugedrückt und den \emph{salto mortale} herzhaft gemacht --- von den Höhen
der schärfsten Kritik hinab in die allergewöhnlichste Verwechslung von Wort
und Begriff! Ist dies gelungen, dann geht es munter weiter. Je mehr
Widersprechendes in die erste Grundlegung aufgenommen wird, desto freier
lässt sich schliessen, wie man denn bekanntlich aus mathematischen Sätzen,
welche den Faktor Null in versteckter Weise enthalten, oft die merkwürdigsten
Dingen ableiten kann.'' It is of course self-evident that a metaphysician's
error can as little be charged to metaphysics as a mathematician's
miscalculation to mathematics. If the mathematician subtly defends his
errors, when they are clearly pointed out, instead of openly admitting them,
he does not thereby compromise mathematics, he compromises only himself. The
same holds of the metaphysician and metaphysics.}! Under the mutual
% p. 46
affecting the reals must precisely lose their absolutely positive character
and sink down to becoming relativities. That it can no more be the concepts
which, held fast by themselves in logical form, produce the things, is
self-evident. The concept of water can as little produce water as one can
drown in the concept of water. The thing's concept is its possibility, a
possibility that lies hidden in other things' actuality. Only by the actual
things having possibilities in themselves to other things are they able to
act. Of several freely existing masses each single one has in its actuality
the possibility of changing the others' place in space; it is this unity of
actuality and possibility that is expressed in the activity we call
mechanical attraction. Two pieces of wood cannot be rubbed against one
another without the possibility of the one having to correspond to the
actuality of the other; light-activity cannot be given unless the
light-giver's actuality corresponds to the ether's possibility, and
conversely. Since now actuality is reality, and possibility concept, it is
seen from this that ideal and real activity, ideal and real cause, must
presuppose one another mutually.

The unity of ideal and real causality is a unity in opposition, an
opposition-unity. From the side of reason the natural order of things makes
the impression of an infinite, in all parts coherent whole; not even the
smallest part can withdraw itself from the power of necessity, for
everything subsists in infinite reciprocal action; it is a world of inner,
essential relations, a world of concepts of ideal causality. From the
standpoint of the sensible experiences, on the other hand, the world of
realities shows itself as a world of chance, of outward conditions and real
causality. According to the connection of concepts there has, for example,
from a
% p. 47
combination between silicon (silicium) and oxygen arisen silicic acid, from
a combination between potassium and oxygen potash, from a combination
between aluminium and oxygen alumina, and from the mutual combination of
these combinations silicate of potash-alumina (feldspar)—all essentially in
consequence of ideal causality. The one-sided concepts are contained in the
more all-sided ones, and every trace of chance seems to have disappeared. By
analysis of the concretion-concept feldspar we come to the co-concepts
silicic acid, potash, and alumina; by analysis of these still concrete
co-concepts to the elementary concepts of oxygen, silicon, potassium,
aluminium—with reflected necessity on account of essential relations. If we
begin with the elementary concepts and apply synthesis, we get the same
necessity within the logical. But in the antilogical sphere of the realities
something quite different shows itself. Here we meet circumstances, chances,
and contingencies on account of outward relations. The oxygen and the
silicon can here exist each by itself, without silicic acid arising; silicic
acid and potash are found each by itself, without silicate of potash
forming, and so forth. In the circle of the outer conditions every outburst
of power is dependent on intervening chance. The outburst, antilogical in
the contingency, is in the necessity logical: the reciprocal determination
of contingent and necessary characterizes the law.

This leads us a step further. There is in the system of real-concepts a
difference between concept-values, marked by this, that some concepts
apparently lie nearest to necessity, others nearest to contingency. Concepts
such as earth, water, salt, acid, base, and the like appear as
essence-concepts, substance-concepts, concepts of necessity; while others,
such as: gravel, heap, aggregate, conglomerate, and so on, are formed in an
entirely contingent manner by the coincidence of concept-consequences that
do not essentially concern one another; such concepts are owing to
contingency to such a degree that they do not even seem to deserve the name
of concepts. This distinction between concepts of substantiality and of
accidentality has a claim to atten-
% p. 48
tion, for the solution of the problem of reality is dependent on how it is
interpreted.

If we begin from the contingent and concede that contingency makes the
concept nominal, then the concept gravel is only a denomination for a
heaping-together of parts, but no objective real-concept. The Positivist
will smile when he hears that the concept gravel too is to be included in the
eternal comprehension and have its ontological origin from an
all-comprehending self-power. But when the concept gravel is nominal, what
then is the concept granite? Also nominal! When the concept granite is
nominal, what then is the concept feldspar? Naturally also nominal; the
formation of feldspar has, after all, been conditioned by outward relations.
That this mode of consideration is false shines forth from its consequence;
for the consequence is that all objective real-concepts must disappear and
give place to nominal designations of concept-less facts. The absurdity and
superficiality herein has been sufficiently pointed out above; we are then
compelled to begin from necessity. Just as in all real necessity there is an
antilogical moment of contingency, since the really necessary is always
\emph{conditioned}, so there is in all the contingent a logical moment of
necessity; if it were not so, there would open up in existence a wide field
where the contingencies could carry on their play without regard to the
laws.

There have been carried on, in the history of the sciences, particularly of
philosophy, extensive disputes about whether, besides the physical causes,
one could also assume final causes and ascribe to these objective validity.
By distinguishing between the outer and the inner purposiveness, Kant has
held that, for the organic formations, one ought to make—not indeed a
constitutive, but nevertheless a regulative—use of the principle of
purposiveness; but even the regulative use seems more and more to sink in
credit. The obscurity has its ground in the fact that the relation between
ideal and real causality has not been sufficiently investigated.

When one without further ado separates the things from their concepts and,
with denial of the ideal causality, imagines that one
% p. 49
has enough in the physically acting causes, the final causes cannot possibly
acquire any objective significance. But that this tearing-loose of the real
from the ideal is unscientific will gradually become evident through a more
thorough treatment of the problems of knowledge and of reality. Both the law
of objectivization and the reciprocal determination of the logical and the
antilogical in the outburst of power prove that the real causality has its
ground in the ideal. When one one-sidedly clings to particulars, contents
oneself with dissection, and blindly makes the whole into a result of the
parts, then it is quite true that all talk of purpose and end must become
nominal and superficial. Every finite attempt to fix something as end and
something else as means founders on the infinite reciprocal action of
conditions and conditioned terms; the alleged higher sinks down to the level
of the supposedly lower, and conditioned necessity becomes sole ruler in the
realm of things and laws. If reason is to come into its right, if a
scientific insight is to be possible, then every approach to the solution of
the great problems must at all points point toward a development of the
parts out of the whole, toward the grounding of the real totality in the
ideal, toward a world-design with an all-comprehending subjectivity as its
author.

In the real-concept \emph{design} lies the unity of ideal and real
causality, in the concept of design purpose, and in the concept of purpose
again the concept of means. The purpose, originally seen, is that
requirement of subjectivity that the plan in the whole, the definite design,
shall be actualized through a reciprocal development of the parts among
themselves. The design itself is a system of designs: no part and no whole
can give itself design, but every part, every whole acts and must act by
virtue of its peculiar design. There must, in the name of reason, be a whole
in order that there can be parts, there must be a higher in order that there
can be a lower, there must be purpose in order that there can be means. The
unscientific arises only when the finite subjectivity presumes to survey the
plan, determine the aims, and apportion sun and wind between objective
purposes and means.

% p. 50
The fundamental question is how the subjective in reality relates to the
objective. This problem itself comes into existence with the organic
formations. What is a living organism, indeed what is life? Life is a
conditioned self-activity, and the product of this activity is the organism;
that is the shortest answer we can give here. The self-activity conditioned
by matter exhibits a self-working-out, though restless, equilibrium of ideal
and real causality. At the lower, elementary stage of development, the ideal
causality, the ontologically subjective, is—what we have often enough
emphasized—present everywhere, but so hidden in realities that the things
act as matter-subjects by their properties upon one another: masses,
molecules, atoms are real causes, physical otherness-causes on account of
the hidden self-cause. With life's organizing activity the relation is
reversed: the otherness-causes sink down to conditions and means for the
self-cause coming to light, the self-actualizing ideal subjectivity, which
in reciprocal action with the manifold of matter-subjects makes itself the
ruling center and shows what the purpose is. In an enigmatic manner the
double-sided causality separates into two relatively independent
causalities, two heterogeneous principles, the ideal and the real; this must
then lead the inquiry further.


% ============================================================
%  III. THE PROBLEM OF THE HETEROGENEOUS PRINCIPLES
% ============================================================
\section*{III.\ The Problem of the Heterogeneous Principles}
\label{sec:ueensartede}

% p. 50 (heading printed here; section III text begins after it)
\noindent
Among the most conspicuous problems of the human understanding belongs,
without doubt, the one that revolves around the relation between life and
organism, between soul and body. This problem obtrudes itself as if of its
own accord, and it is then no wonder that both in earlier and in more recent
times it has given occasion to a multitude of different hypotheses, and that
these hypotheses, each in its turn, have competed for the dignity of being
exalted into dogma. Nor has any problem lain nearer to hand when it was a
question of a striking example of the barrenness of philosophical
speculations and the imperfection of all human
% p. 51
knowledge. If one grapples with this problem, one will soon make the
discovery that it has something Protean about it and easily slips away from
him who does not rightly know where he is to grasp it in order to hold it
fast. One may start from the presupposition, which appeals to sound sense,
that life is different from the organism, that the soul is a relatively
independent being in opposition to the body, and yet, by the further course
of the investigations, be led to the recognition that soul and body, and
then of course also life and organism, cannot each have their own
subsistence, but must be assumed to develop out of one and the same
fundamental principle; the problem is then this: \emph{is the fundamental
principle originally to be set in the ideal or in the real?}

If one makes the ideal principle absolutely valid, conceives life as Idea
and the soul as the life-Idea's self-reflection in its immediate other, then
the organism becomes a phenomenon of life, the body a phenomenon of the
soul. If one, conversely, makes the real principle absolutely valid, then
life becomes a resultant of the forces acting in the organism, and ``the
soul-specter,'' a phantom formed from brain-functions, shows itself only as a
revenant from the darkness of Scholasticism. What especially makes the
matter complicated is the circumstance that, however irreconcilable the
opposition is between Idealism and Materialism, it is nevertheless
unavoidable that the Idealist must express himself materialistically about
the psychical, and the Materialist idealistically about the bodily, when it
comes to the point. ``The brain thinks, just as the stomach digests''—this
proposition is materialistic, insofar as it places the activity of digestion
on the same level as the activity of thought. The activity of thought is
ideal; the Idealist knows no purer expression for ideal activity. In wishing
to make the ideal activity material, the Materialist comes precisely to make
the material activity ideal. The Idealist, who on account of the unity of
being and thought lets matter think, must of course also let the stomach
think, conceive the activity of digestion as a dialectical conceptual
process, and thereby comes, in his efforts to make
% p. 52
a material activity ideal, to make the ideal activity material. Insofar as
the Idealist will in any way enter upon particular determinations of organs
and organic conditions, he must necessarily speak as though matter really
were material, and of such categories as bones, muscles, nerves use words
and expressions that immediately correspond to our representations of the
bodily. The same holds in the reverse sense for the Materialist when he
speaks of the psychical. However materialistically the psychical functions
are conceived, he is nevertheless compelled to use words such as sensation,
feeling, representation, consciousness, and so on; but already the immediate
and natural understanding of these words transports us with one jerk from
the material sphere over into the ideal. When the Materialist treats of the
consciousness of the brain, the spinal cord, the ganglia, he cannot avoid
distributing the lower and higher consciousnesses among organs, so that it
looks as though consciousness were after all something ideal by itself; he
lets, after all, the different consciousnesses melt together into a kind of
consciousness-unity, without being able to fuse the organs into a similar
unity.

One might on this account be tempted to the assumption that the opposition
between Idealism and Materialism was not essential, that the dispute was
only a dispute about words; but this is by no means the case: the opposition
is one of principle. From the idealist side the matter is understood in such
a way that the material activity is semblance and appearance of the ideal;
materialistically it is understood in such a way that the ideal activity
everywhere becomes semblance and appearance of the material. That there is
here semblance and appearance on both sides proves that reflection is in
movement, that there is a struggle for a monistic decision. The merely
monistic unity is, however, an ambiguous unity, and so long as one struggles
for an ambiguous unity, the struggle can be continued in all ambiguity. The
old Dualism was mechanically miraculous, but it was not ambiguous. The
miraculous science has thoroughly overcome; but the modern Monism, which has
set itself in Dualism's place, suffers from an ambigu-
% p. 53
ity that will hinder every true scientific advance, so long as it is allowed
to remain standing.

In order to remove the ambiguity it is first of all necessary to come to
clarity about how far there is at bottom any difference of principle between
the lifeless and the living, and, if such a difference indisputably takes
place, then to investigate wherein it actually consists, and what the
distinguishing marks are by which it is to be known. Without dwelling on
comparisons and descriptions, which from a natural-historical standpoint may
be orienting for practical use, we will attempt to hit the matter's core
point and direct attention to what, according to our conception, must be the
decisive thing. As the distinguishing characteristic of lifeless activity of
whatever kind—mechanical, magnetic, electrical, chemical—we hold fast what
physics has demonstrated as a law, that action and reaction under all
reciprocal actions of material terms must exactly correspond to one another,
so that no self-governing midpoint can be assumed. As the mark of
life-activity, on the other hand, it is emphasized that, under the material
points' mutual action and reaction, an ideal midpoint forms itself and
therewith a conditioned self-activity, which alters and governs the parts'
reciprocal action. The facts corresponding to this are metabolism and
form-giving, not immediately each by itself, but in active union. A
crystal-form can maintain itself when metabolism is excluded; a metabolism
can take place and the form be disturbed; herein is no trace of
life-activity. But when a metabolism takes place in such a way that a
typically determined form-development is thereby maintained, then we regard
this as an unmistakable sign of life's organizing activity. The activity's
ideal midpoint cannot be mathematically localized like a sphere's center of
gravity, which is only a resultant of the parts' gravity; nor can it be
either a proper material point or a metaphysical real, for in that case it
would have to act according to its properties like other material points or
reals. Every whole that rests solely on reciprocal action of real parts
according to properties must—what the theory of potential-functions proves
exactly—be
% p. 54
punctually subject to the law of action and reaction: here is no trace
either of self-active unity or of governing midpoint. Everywhere, on the
contrary, where cells form themselves and life manifests itself, there is an
indication of a peculiar need, peculiar requirements, which demand their own
housekeeping; life-activity has the stamp of a governing something, which
through its other, its material, will assert itself and work for its
preservation and self-development. Characteristic of lifeless things is
their complete indifference to the state in which they find themselves. No
one will assume that in water there is a hidden longing to become ice, or in
ice a quiet urge to pass over into vapor; on the contrary, a plant sprouting
in the dark, which stretches toward the sunlight, may well be said to betray
a striving and longing. It is a distinguishing mark of life that it has a
goal: self-development and self-preservation is the goal. To struggle for
one's life is to have a purpose. It must in high degree surprise everyone who
has not yet given up all hope of rational insight, that researchers of
positivistic zeal for an exact cognition of conditions and causes have
rejected every representation of final causes, and yet, as regards the
development of the organic forms, chosen the watchword: \emph{Struggle for
Life}.

With these distinguishing marks for the lifeless and the living we stand
before \emph{the problem of the heterogeneous principles}.

The reality-principle is recognizable enough. At the present stage of
science there can be no doubt that the organic formations take place in full
agreement with the general laws of nature just as well as the inorganic; for
mechanical attractions, chemical affinities, polar tensions, the relation
between work and living force, and so on, the same fundamental laws hold in
the organism as in elementary nature. The ideality-principle can, according
to our foregoing, also not be dismissed. The marks of self-activity,
self-development, self-preservation are far too conspicuous to be able to
withdraw themselves from the unprejudiced gaze. When these indications of
selfhood are nevertheless overlooked or effaced, then it is evidently not
the fault of the facts; it is only the Positivist who, from lack of insight
into the ideal,
% p. 55
supposes that he finds his advantage in denying what he cannot explain. The
ideal principle by no means calls upon us anew to introduce either a
peculiar vital force on the same level as the elementary forces or a
soul-substance, a proper real, that should sit here or there like ``a
parasite'' in the body; but it demands of us that we do not deliberately
close our eyes to what in the living and then especially in the ensouled
organisms is unmistakable, a self-activity different from matter-activity
but conditioned by matter, a peculiar articulating, self-developing,
self-preserving activity.

Perhaps one will object that the opposition of two heterogeneous principles
here set up is only a rehash of an antiquated Dualism, which the monistic
science of the present day neither can nor will recognize; we then ask
whether present-day Monism can exempt science from bowing to facts. If it is
possible for science to show how physical currents and molecular changes can
transform themselves into sensations, feelings, representations—well then,
Monism is right, and Dualism falls away of itself. But so long as the human
understanding is able to make a distinction between material conditions and
ideal causality, so long as the supposition that the same thing which from
the one side shows itself as outward and material can from the other side
show itself as an inward, self-apprehending ideal is and remains
meaningless—still more meaningless than the assumption that tone and
sound-figure are the same, only apprehended with different senses—so long
stands the opposition of two heterogeneous principles, the matter-activity's
and the self-activity's, in force as an incontestable fact. The highest that
one will ever be able to reach by way of the immediate observations is a
thoroughgoing parallelism of matter-activity and self-activity, of bodily
and psychical functions. But a plain and simple paralleling gives us no real
insight; the Dualism must have its ground in an essential unity, and the
question then becomes how far such a ground of unity can with any
probability be pointed out.

% p. 56
It is an age-old view that the whole of nature is ensouled and living, that
life is not limited to the organisms alone, but even embraces what one
otherwise calls dead matter. This way of seeing is false insofar as it is
held fast in the form of imagination and representation; but, brought under
the general law of comprehension, it may well be said to contain a moment of
truth. What is misleading in Hylozoism is: first, a one-sided propensity to
regard all stirring, all movements, especially all continued circular
movements, as unmistakable signs of life—there is then life in the solar
system, life in elementary changes, in geological formation-processes, and
so on; secondly, an ambiguous use of the word sensation: the cliff senses
the stroke of the lightning, the iron senses the magnet's nearness, nerves
and ganglia sense physical impressions, as we sense a well-being or an
ill-being; thirdly, a mystical application of the word drive, especially
formative drive. By the use of vague, indeterminable analogies, symbols, and
ingenuities one gradually breaks down every barrier between the living and
the lifeless and gets the qualitative difference reduced to a vanishing
difference of degree: the more outwardly, mechanically, and chemically the
signs of life are conceived, the more fluid it goes with hylozoistic
fantasies.

Under the standpoint of the concept, on the other hand, the representation of
a universal life spread throughout the whole of existence can be understood
by the ontological comprehension in which all that is is reflected. The
self-comprehension that objectivizes itself in the system of real-concepts
lies at the ground of all transitions from possibility to actuality in the
world of realities. Through all series of physical causes, conditions, and
effects there goes a stream of ideal cause-moments and ideal effects; this
never-stopped stream, encircling ideality's selfhood, can in logical
abstraction be conceived as a shading of a universal living, a universal
self-activity, corresponding to a universal life-concept. But further one
does not reach by this way.

If life is really to come into existence, then its existence must just as
surely have its particular real-concept in the whole design of existence as
either the oxygen or the water or the hydrocarbon or the
% p. 57
ammonia or whatever else one will name; only with the express difference that
life's concept is a selfhood-concept with a surplus of ideality, the
concepts of water, hydrocarbon, ammonia, on the other hand,
otherness-concepts, matter-concepts. The problem then concentrates itself in
the task: \emph{to make clear the fundamental relation between
selfhood-concept and matter-concept}.

In the matter-concept the self-activity is latent, just as the ideal
causality is latent in the physically acting cause. The physicist knows that
in a body there can be heat which does not act on the thermometer, because
the activity is converted into a state-form, and only comes to light again
when the state-form is changed. In like manner it holds that the ideal
causality is never immediately noticed in physical causes or in that
outburst of power by which the transition from possibility to actuality takes
place under the form of otherness. But the latent causality is nevertheless
just as little to be regarded as nothing as the latent heat. On peculiar
conditions, under peculiar circumstances, the selfhood-concept—which surely,
as fully as the otherness-concepts, must be a concept of ideal power—can set
its possibility over into actuality and come into existence. Let us test how
this will appear in fact. So long as the so-called organic substances—oxygen,
hydrogen, carbon, and nitrogen—act merely according to the laws of
otherness, the reciprocal action produces no surplus of ideality, nothing in
which a selfhood-cause can show itself; it remains at elementary formations.
If, on the other hand, the reciprocal action between atoms or molecules takes
place under such conditions that a surplus of ideal causality arises, which
cannot in the manner of otherness be equalized in action and reaction of
material parts, then there forms itself, out of the ideal surplus, a midpoint
of self-activity: life begins its wondrous work.

Although the opposition between the ideal and real principle is thereby
asserted, it will nevertheless easily be seen that this Dualism does not make
a breach in the unity of existence, but on the contrary
% p. 58
serves to ground the true, well-understood Monism. The dispute about
\textit{generatio æqvivoca} is not yet at an end, nor has one given up the
hope of being able to produce a living cell. It will be understood that the
possibility of such a production by no means stands in conflict with the
assumption of heterogeneous principles; for life-activity must just as surely
have its law-fixed conditions as any other activity. On the contrary, it must
be expressly emphasized that cell-productions by artificial means cannot be
set up as proofs that life is only a resultant of physical forces.
Selfhood's ideal midpoint can in the beginning be so weak, so hidden and lost
in the matter-activity, that it looks as though the life-principle were only
an imagination; but as the formations advance, it shows itself ever more
clearly that the material changes are subordinated to the self-activity, just
as it also shows itself how dependent the self-active is on matter and
matter-activity.

The organic functions thereby take on the stamp of doubling. No material
part, no matter-subject is able to carry out any selfhood-function whatever.
Propositions such as: ``The ganglion feels, the brain thinks'' are not
scientific; one might just as well say: ``Chlorine has an inclination for the
metals,'' ``The magnetic needle is attentive to the North Pole.'' Conversely,
the self-active in the organism cannot accomplish the least without a
corresponding matter-activity.

The \emph{organic} functions are from first to last \emph{double-functions},
each function a doubling-unity of matter-activity and self-activity. Herein
lies that the matter-activity can now be so predominant that it hides the
self-activity, and the latter again, as in pure thinking and willing, take on
the appearance of pushing the matter-activity back. Oscillations of the
matter-functions' and the selfhood-functions' alternating preponderance can
call forth manifold expressions of interferences, polarizations, of received,
given-off, and stored-up work in different measure; but whether by
physiological observations, psychophysical measurements, and psychological
interpretations it will ever
% p. 59
be able to be brought to the point that exact analyses of the oscillations
obtain a more reliable ground than the Herbartian mechanism of
representations, we must leave undecided. The first thing needful is that the
proposition about the double-sidedness of the functions be subjected to a
critical examination.

Regarded from the side of the matter-functions, the organic life-unity is a
kind of collective unity: so many cells, so many lives; so many organs of
consciousness, so many consciousnesses. The collective unity is then more
closely interpreted as a disciplining unity of connection: the smaller lives
are subordinated to the greater, the dimmer consciousnesses melt together in
the clearer; but the collective is and remains the characteristic. From the
side of the selfhood-functions, on the other hand, the unity shows itself as
an all-concentrating self-unity: one life, one soul, one consciousness, one
self in the human body has in older psychologies been made the norm for every
judgment of lower-standing organisms. But the representation of the soul as
an in-itself indivisible, unchangeable, and therefore imperishable single
being has the unprejudicedly advancing physiology given the death-blow, in
that it has by incontestable facts proved that life not only in vegetative
but even in animal organisms can be parceled out, and the psychical divided.
Only by conceiving the organic functions as double-functions do we bring it
to the point that both the one-sided ways of seeing come to their relative
right. Since the selfhood-functions are everywhere conditioned by the
matter-functions, the self-unity must necessarily be reflected in the unity
of connection, and have its validity along with the division; but since the
matter-functions again, as regards life and the psychical, are only means and
conditions for the selfhood-functions, and the psychical selfhood-functions
reflected moments in the ruling self-unity, then the old representation of
the soul as a relatively independent being nevertheless gets its
significance.

But what now is this psychical being? It cannot possibly be any immediate
substance, for an immediate substance is only a substratum. To speak of a
soul-substance, a soul-real,
% p. 60
is, when we make the matter clear to ourselves, the same as to speak of a
substratum-activity, analogous with a matter-activity, while one
nevertheless imagines that one is speaking of a self-activity essentially
different from matter-activity. On the contrary, it is in full agreement both
with the concept of self-unity and with the physiological facts to speak of a
unity of ideal causality that constantly strengthens itself under material
conditions, a self-unity which can in manifold ways be hindered and weakened,
but also, under favorable conditions, show a disposition to a concentration
whereby it raises itself to ever greater reflected independence and ideal
substantiality.

If the independence resulting from self-activity is to be more closely
investigated, then the problem takes on the character of \emph{a problem of
freedom: the real and the ideal principle set themselves against one another
as necessity against freedom}.

The conflict between freedom and necessity does not first arise with
arbitrariness and the self-conscious will's choice in the finite; its origin
coincides with the origin of life; every monstrosity, every embryonic
abnormality can be an example: the principle of selfhood is the principle of
freedom, just as the principle of otherness is that of necessity. Absolute
necessity is a concept that cannot exist; real necessity belongs to otherness
and thereby to relativity. If one supposes that one has the absolutely
necessary in an infinite totality of conditioned terms, then this is only an
imagination. Should such a necessity be absolute, then it would have to
master its terms and conditions and, in a mighty comprehension of the all,
assert its unity; but precisely thereby the mighty unity would, after all, be
a self-reflecting unity, a self-power-unity, a freedom-unity.

One has sought to hold fast the absolute unity of freedom and necessity from
the side of necessity in the category: free necessity (\textit{libera
necessitas}), from the side of freedom in the category: necessary freedom
(\textit{libertas necessaria}); that is metaphysical formalism. Necessity
does not become unconditioned in infinite
% p. 61
series of connected conditions, it becomes on the contrary conditioned to
infinity; nor does freedom become unconditioned by pushing the conditions
away from itself, but by self-actively appropriating them and making them its
own conditions.

The absolute freedom-unity can be knowing, mighty, self-determining,
all-determining without being arbitrary. When in the analysis of the
ontological concept of subjectivity we nevertheless introduce categories such
as arbitrariness, choice, decision, these by no means denote separate
actions, but determinations of finitude, antirational moments, which dissolve
in the infinite reflection and yield reflexes in the concept of absolute
self-determination.

The question of the content of self-determination leads us immediately over
into the circle of otherness, of the relativities, of the finite conditions
and causes, and here we meet necessity. That which happens with freedom must
just as surely have a ground in another as that which happens with necessity,
for no definite willing is possible without something's being willed. The
real ground of the necessary, which is not one with the sum of the
conditions, but reflected in their relations, is subject to the unity of
connection; self-determination, on the other hand, transforms the real
grounds into \emph{motives}. Necessity with the physical causes is reflected
in nothing but real-grounds; freedom's ideal causality in nothing but
motives. It is only an outward, superficial representation that the
absolutely self-determining freedom-unity should be able to treat the
connection of things, the world of laws, arbitrarily, or proceed by leaps and
leap over the real-grounds; it lies, on the contrary, in the fundamental
relation of necessity and freedom that the motives in the ideal must exactly
correspond to the grounds in the real.

At the stage of finitude the conflict between freedom and necessity shows
itself as a conflict between conditioned self-activity and the conditions to
which it is subject. The ensouled organisms suffer under hindrances, need,
and constraint; the self-conscious subjectivities apprehend the constraining
necessity as fate. Then it is said that the power of fate makes man's freedom
into
% p. 62
nothing. Wherein, then, consists man's freedom? We answer: In self-conscious,
conditioned self-activity. How far the self-determined activity reaches, and
how much it accomplishes, depends on inner and outer conditions, but it is
annihilated only when self-recollection ceases, and the light of
consciousness is extinguished. Self-consciousness can raise its determining
willing up before itself and, by a peculiar kind of abstraction, free it from
every constraint; it then looks as though the free will stood independent
over against the fateful necessity, measured it with its eyes, and challenged
it to single combat. Whether the will in outward action wins or loses is not
the decisive thing; the decisive thing is that there is willing. It is not
this or that finite thing that is chosen or willed, but the will chooses
itself, chooses unconditionally to will. From this height of abstraction
defenders of the unconditioned freedom-principle have categorically declared
that the rational, self-legislating will is able to remove all admixtures of
sensuous incitements and go against drive and inclination as the ship that
has its own motive power can go against wind and current. Experience shows,
however, that it cannot be done. And why not? Because pure self-determination
is only an abstract expression for imaginary freedom. Not even in the moments
of self-exaltation has the self-willing will been independent of impressions
from another. A mood, a feeling, a goal of thought, an ideal of imagination
has been co-operative. If we investigate such an ideal a little more closely,
we shall discover manifold influences, reminiscences, experiences, examples,
which in all concealment stand behind it, so that the will is never free of
motives. But the motives are, after all, owing to impressions from another,
regard for another; when attention fixes itself on these impressions and the
regards thereby determined, the will becomes concrete, sinks down into the
circle of conditions, and feels the weight of the motives.

The universal and yet concrete motive now becomes self-preservation, which
again can be partly of a lower, partly of a higher kind. The claims of
individuality are to be satisfied; the satisfaction is conditioned by an
order of things in which necessity rules; there is then needed no great
analytic art to discover how
% p. 63
tightly the chains of conditions wind themselves about the individual's
willing. The state of the outer world, the temperature of the surroundings,
and the juices in the blood unite with the nervous system's
physiological-psychological dispositions and make the will an organic product
of real motives. Seen from this standpoint, freedom founders in necessity.
Without waiting upon the worn-out theories—indifferentism, determinism,
predeterminism, and their various shadings—we consider it sufficient to point
to the problem itself: \emph{The problem lies in the reciprocal determination
of motive and will}. No will without motive; the strongest motive will always
dominate the will: herein lies \emph{the moment of necessity}. But the
relation between stronger and weaker motives cannot be weighed mechanically;
there is in the will's subjective energy a surplus of ideality that is
incommensurable with the objective in the motive. It is in the highest
instance the will itself that must decide which of the strong motives is to
become the strongest: the will \emph{is moved only in that it lets itself be
moved}; herein shows itself the imperishable \emph{moment of freedom}. The
freedom-functions have the unmistakable likeness to the organic ones, that
they are double-functions, functions of objectivity's real grounds and of the
will's, that is, of self-determination's subjective energy.

The freedom-life in individuals is to be regarded from two principal sides,
insofar as it partly manifests itself in single scenes and actions, and
insofar as it displays itself in a ruling striving toward a definite final
goal. From both sides the freedom-principle points toward a grounding that
lies outside the boundary of science. The objective necessity that expresses
itself in the connection of things furnishes a firm and secure ground for
scientific investigations: the grounds are convincing, and the proofs
compelling. But self-determination's subjective energy withdraws itself from
science's control to such a degree that here appeal must constantly be made to
the individuals' inner experience; but the inner experience is different in
different individuals. The grounds are not compelling, but persuasive; for
freedom-grounds must be appropriated with freedom. What use is it
% p. 64
to speak to people who have their life in nothing but finite interests, of a
freedom that obliges to self-denial and the sacrifice of the temporal in
order to win the eternal? They will not understand it. Even if they cannot
point out contradictions in the conceptual development, the ideal
freedom-enthusiasm will nevertheless in their eyes appear as an abstraction's
flight from reality, a utopian longing toward a supersensible, an
exaggeration, a fanaticism. They cannot understand it; for freedom-grounds
are ethical-subjective, and ethical-subjective grounds find entrance only
with subjectivities that are practiced in ethical self-exertion and will train
themselves in the school of freedom. The freedom-principle is not a
theoretical but a practical principle, a principle of faith; thereby it
becomes not only heterogeneous with necessity, but, as we shall see in what
follows, heterogeneous with science.

Scientific investigations of the essence of faith lead so deep into the
mysteries of subjectivity that the problem of the heterogeneous principles in
the last instance transforms itself into a decisive boundary-problem,
\emph{the problem of Religion and Science, Faith and Knowledge}.

It is in this respect remarkable to see how Comte, the founder of Positivism,
after having begun with an objectivity of facts that excludes the subjective,
ends by calling feeling and subjectivity to his aid and, in the name of
religion and morality, lamenting the modern revolt of the spirit against the
heart (\textit{l'insurrection moderne de l'esprit contre le cœur}). His
learned and clearly thinking disciple, Littré, has—not indeed without a
certain melancholy, which does honor to his devotion to the master—pointed
out this deplorable inconsistency\footnote{\textit{Auguste Comte et la
philosophie positive.} Paris 1877. S.\ 539 flgd.}. Feeling, says Comte, ought
always to rule over the intelligence; hence, objects Littré, just as in the
Middle Ages philosophy was the handmaid of theology, so should the spirit now
be the slave of the heart. Herein is something ambiguous. If Comte with his
expression means nothing more than that the spirit, the intelligence in unity
with feeling, is to work for the good and the beneficial, then there is no
one who
% p. 65
will contradict him herein; but if he thereby means that the whole governance
is to proceed from the heart, then he blinds the heart and gives morality up
as a prey to all sorts of excesses. If he means that the intelligence ought
not to work for itself or strive after pure truth and abstract theory, then he
mutilates humanity and robs it of its mightiest instrument for ennoblement and
progress in perfection. It was not the heart, but faith, the blind faith in
the authority of revelation, that held the spirit in medieval subjugation.
Should the spirit's rising against this subjugation be pernicious for the
human souls, then the Middle Ages must, historically seen, have more virtues
and fewer vices than the more recent time; but history shows that precisely
the reverse is the case. Comte sets the scientifically grounded faith
(\textit{la foi démontrée}) against the revealed faith (\textit{la foi
révélée}) and then demands that the intelligence shall submit itself to
religion; but what does this mean other than that the spirit shall submit
itself to its own work, thought to its own proofs\footnote{\textit{Anfr. Skr.
IIIme partie. Chapitre IV. Examen de l'opinion que l'esprit doit être
subordonné au cœur. S. 545. En prétendant unir les principes du régime
positif avec les conséquences du régime théologique, on tombe dans des
contradictions implicites. Subordonner l'esprit á la foi révélée est une
notion parfaitement claire; car la révélation émane de la divinité, et,
quelque inconcevables que soient les mystères qu'elle ordonne de croire, il
n'importe, une autorité devant laquelle l'homme n'est rien a parlé, et il doit
se soumettre. Mais, dans le régime positif, que peut signifier la
subordination de l'esprit à la foi démontrée, puisque cette foi démontrée
n'est pas autre qu'une œuvre de l'esprit lui-même et le résultat d'une longue
enquête qui a mis en lumière les lois immanentes du monde? C'est dire le même
par le même que de dire que l'esprit se subordonne à une foi qu'il a faite, et
énoncer une impossibilité que d'énoncer qu'il s'insurge contre cette foi?}}.

That the freedom-principle cannot find recognition in Positivism is a matter
of course. By assuming the fact to be something objective in itself and
holding to realities without real-concepts, the positivistic philosophy has
from the beginning excluded every attempt at a more deeply penetrating
cognition of freedom's subjective essence. The only scientific way from the
objective to the subjective, from necessity to freedom, goes through the
real-concepts. \emph{Things are determined
% p. 66
by their concepts, the concepts are determined in being comprehended, and they
are originally comprehended by the all-comprehending self-unity, the
ontological subjectivity.} If this assumption can be shaken, if it can really
be demonstrated that there can be things without concepts, facts without laws,
objects without objectivizing selfhood—well then, there can be talk of holding
the necessary fast without freedom and of discovering the really true, the
in-itself independent objective, by eliminating the subjective. But if every
attempt in this direction, according to what has been made clear about the
problem of knowledge and of reality, is found to be unreasonable and
meaningless, then it stands unshakably fast that the whole objective existence
is borne by objectivizing subjectivity, that the necessary order of things has
its ground in self-determining freedom. The things, the facts, which for our
empirical investigations are the first determining thing, are then not in
themselves the first; they are a derived, they exist only as outbursts of
power. The real-concepts, which for our rational analyses are logically
determining, are not the first logical thing: the system of concepts points
toward an eternal logic in an eternal self-thinking unity. From all sides,
through all determinations—be they real or ideal, logical or antilogical—the
investigations point, as in converging lines, toward the all-uniting midpoint,
the determining point of departure, the inner ideal causality for real causes
and cause-series in outward infinity, that is, toward the original freedom.

But what and how much the divine freedom is able to do, no human being can
know\footnote{If anyone objects that the same holds of an all-comprehending
self-power, we must call attention to the following. To penetrate into the
inner movements of the eternal self-comprehension is certainly just as
impossible for us as to fathom the secrets of freedom. One might for that
matter just as well settle down with the Hegelian: the concept comprehends
itself, as appeal to an ontological subjectivity, for both are in reality
equally incomprehensible to us. But precisely in the determination of the
\emph{incomprehensible} it is important to be scrupulous with respect to
\emph{possibility} and \emph{necessity}. Most aptly F.\ A.\ Lange remarks,
concerning
% (fn cont. p. 67)
``einen nicht stichhaltigen Versuch in \emph{Ueberweg's} Logik
(§§~38--44)'': ``Ueberweg bestreitet mit Recht die Art, in welcher Kant Raum
und Zeit als \emph{Form} der Wahrnehmung von dem \emph{Stoff} derselben
unterschied. Er geht sodann von dem Satze aus, dass die innere Wahrnehmung
ihre Objekte so, wie sie an sich sind, mit materieller Wahrheit aufzufassen
vermöge. Mit musterhafter Klarheit unterscheidet er das Wesen der Empfindung
von dem Wesen der Dinge, durch welche dieselbe veranlasst wird. Nur das Wesen
der psychischen Gebilde in unserm eignen Bewusstsein, glaubt Ueberweg,
vermöchten wir genau so zu erkennen, wie es ist. Da nun unsre innere Erfahrung
zeitlich verläuft, so hält er die Wirklichkeit der Zeit für erwiesen. Die
Zeitordnung setzt aber die Gesetze der Mathematik voraus und diese setzen den
Raum von den drei Dimensionen voraus, womit der Gang des Beweises abschliesst.
--- Abgesehen davon, dass der Fundamentalsatz wenigstens hinsichtlich der
Reproduction gerechten Bedenken unterliegt, scheint mir ein ganz bestimmter
Fehler darin zu liegen, dass die Realität der Zeit in uns auf die Realität der
Zeit ausser uns übertragen wird. \emph{In uns} hat nicht nur die Zeit
Realität, sondern auch der Raum, ohne dass dazu eine Vermittelung durch den
Zusammenhang der mathematischen Gesetze nöthig wäre. Nun müssen wir allerdings
aus dem Zusammenhang der Dinge in \emph{uns} mit Nothwendigkeit auf einen
correspondirenden Zusammenhang der Dinge \emph{ausser uns} schliessen; allein
dieser Zusammenhang braucht eben keineswegs Uebereinstimmung zu sein. Wie sich
die Vibrationen der berechneten Erscheinungswelt zu den Farben der unmittelbar
gesehenen verhalten, so könnte sich auch eine für uns ganz unfassbare Ordnung
der Dinge zu der räumlich-zeitlichen Ordnung verhalten, die in unsern
Wahrnehmungen herrscht.'' Geschicht.\ d.\ Materalsm.\ S.\ 497--98. --- This
can all be conceded; but from the incomprehensible there is a great leap to
the impossible, and only when the impossible is reached does doubt cease. It
is perhaps possible that a being can be so organized that it is only able to
represent space to itself in two dimensions, and that our world of experience
as a consequence must appear to it to be like a ``Ding an sich'' (anfr.\ Skr.\
S.\ 499); but in order not to make ourselves guilty of a \textit{μετάβασις εἰς
ἄλλο γένος}, we must
% (fn cont. p. 68)
guard ourselves against distracting side-considerations. The question here is
not how many dimensions space may possibly have, or what impressions different
vibration-times may cause; but the question is: Can there be forms of space
and time so essentially different from ours that for that reason they are
incomprehensible to us? If this is assumed, then the unknown space-form must,
not only on the whole, but also in each of its dimensions, be qualitatively
different from ours. There must then be dimensions that are not dimensions,
that is, dimensions which, although they are entirely incomprehensible to us,
must yet be conceived as analogous with what we call dimensions. Likewise, as
regards time, there must be successions so incomprehensible that they can only
be conceived as analogous with what we call succession; hence: incomprehensible
and yet not incomprehensible. If anyone finds this acceptable, then there must
in his consciousness be a space for the incomprehensible that is more harmful
than what is called the harmful space in the air-pump. A space that is not
space, a time that is not time, is not merely something incomprehensible to
us; it is something in itself impossible. That whose opposite is impossible is
necessary; it becomes, for the sake of criticism, necessary to give up the
barren skeptical game with incomprehensible analogies, necessary to recognize
the a priori subjective forms as objectively universally valid. In the very
same way it stands with reality, concept, and comprehending self-being. We do
not assume that every reality must necessarily have its concept, and that this
concept must lie hidden in other realities, because it is in fact less
incomprehensible to us how a real-concept can by an outburst of power become
reality than how ether-waves can be transformed into light and colors; but we
assume it because we see that the opposite is impossible. We presuppose an
all-comprehending self-unity, not because we grasp how an infinite,
simultaneous comprehension of all concepts can take place, but because we see
that concepts cannot possibly be thought without comprehension, and that
comprehension is impossible without comprehending subjectivity. Otherwise with
the particular concepts and their mutual relation; here there are countless
possibilities, logical consequences, hidden transitions, which we do not
grasp. It is in the name of science just as unreasonable to break off at an
apparently unintelligible thing and declare the incomprehensible-to-us to be
the in-itself-impossible, as it is a sign of fantastic overreach to operate
with imagined possibilities and pretend as though one grasped what nevertheless
is and remains incomprehensible to us. Strict science may by no means deny
that the divine freedom has unsearchable possibilities; but it is entitled to
make objections when a something that calls itself science wants the
unsearchable possibilities scientifically searched out, when in
pseudo-scientific form one is served with categories such as: \emph{God's
grace, God's disfavor, wrath, mercy, forgiveness, remission}, and so on. These
are, after all, possibilities that only those can know who in consequence of
their calling and dignity have free access to the Holy of Holies; let them then
communicate to us the Godhead's secret counsels and decisions—yet, be it
noted, in such a way that the communication takes place, not in science, but
--- outside science.}. Science has, as regards the content,
% p. 67
to hold to what really lies before it, what has been produced and is
demonstrably produced in consequence of the laws; we must,
% p. 68
in other words, know freedom by necessity.\footnote{Translator's note: the
print here reads ``med andre, Ord'' — the comma is misplaced, and should read
``med andre Ord,'' (``in other words,'').} There can then lie in the depth of
freedom possibilities of which science has no inkling or can have. The
self-conscious life is in the highest sense interested in everything that
concerns freedom; since now it is impossible for human knowledge to measure
out the depths of freedom, there must on this account be life-questions,
decisive freedom-questions, which science cannot answer. What does the eternal
spirit intend with this nature, what is
% p. 69
its actual meaning with the endless swarm of ensouled organisms, which, as it
seems, are born only in order to suffer, and live only in order to die? Above
all, what is its purpose with man? Are the human souls destined to disappear
with the dissolution of the present organs, or do they bear within them a
disposition to a development infinitely continued on other conditions? Upon an
answer to such questions science cannot enter, for possibility stands against
possibility.

Already the psychological reciprocal determination of \emph{real-grounds,
motives, and will} reminds us how impenetrable the actual fundamental relation
between freedom and necessity must be for finite consciousnesses. If one lays
the emphasis on the motives' real grounds, then necessity steps forth in
freedom's light, and what then will the spirit? A development of the
possibilities contained in its own inexhaustible nature-ground! The visible
world's design and this design's law-determined actualization is
objectivization's expression for the infinite satisfaction with which the
all-producing freedom enjoys its own grounded productions. Natural-necessity,
which, objectively seen, is blind and conditioned, is in the free self seen
through and willed; far from being something foreign and constraining to
self-determination, it is on the contrary that system of real grounds which
makes the will absolute by giving the motives objective validity. When it is
then said that the spirit, with all alleged economy and apparent prodigality,
cannot do otherwise than produce this real nature, this objective world with
its phenomena and laws, then this will only mean that the absolute freedom, by
determining everything according to its own nature, must absolutely satisfy
itself. This is, so to speak, the realism of the freedom-principle.

But the freedom-principle is at bottom an idealistic principle. The
freedom-idealism lies in self-determination's subjective inwardness: ideas,
ideals, and ideal concepts also belong among the will's motives. It is then
possible that the will, without withdrawing itself from the real-grounds and
becoming arbitrary, can by virtue of ideal motives determine and articulate
the possibilities in such a way that
% p. 70
real-systems of a lower order are set as conditions and means for systems of a
higher order. It is then possible that this nature, the universe in which we
live, is only a grand introduction, a step on the infinite scale by which the
eternal wisdom leads its work, its will-productions, up toward the goal of
perfection; but the opposite is, after all, also possible. Where are we now?
Is it now so that the whole development takes place within this visible
universe, or are there real-systems of a higher order in which the completion
is reached? We do not know it; possibility stands against possibility.

In the same uncertainty we are entangled when the question of the
all-determining self-being's care for the ensouled organisms, especially for
the self-conscious souls, is to be answered. It is, after all, possible that
the eternal spirit is infinitely indifferent to the weal and woe of the
ensouled individuals; it is, after all, possible that the unchangeable law of
nature, which sets a limit to enjoyment and bounds to suffering, must be the
living beings' only consolation, that the ideality which can be distilled from
self-conscious beings' anguish and pain does not benefit the individuals
themselves, but is a sublimate of the impotent ones' resignation, an odor of
sacrifice which the almighty freedom enjoys for itself alone. But it is, after
all, also possible that the spirit's essence is love, that the apparent
indifference belongs to that wisdom which solves all riddles, and will one
day, in higher regions, explain itself into an eloquent testimony of the
fatherly care with which the absolute self-being has from eternity willed to
secure for the relative self-beings a thorough, gradual, and therefore painful
attainment of the freedom to which they are destined. On which side now stands
the real truth? We do not know it: possibility stands against possibility.

Are we herewith finished? By no means. On a neutralizing equilibrium of
opposite possibilities it is impossible to live. No free rational human being
can lead his life without a rationally determined view of life; but a definite
view of life is only possible by virtue of a decisive choice between opposite
possibilities: here is the boundary-line between Religion and Science, Faith
and
% p. 71
Knowledge. The highest that scientific insight can reach is to clarify the
possibilities and show us the necessity of the choice; but for the choice
itself science cannot be responsible: the responsibility rests on the choosing
subjectivity. He who chooses to live on a view that sacrifices the individual
for the universal, bends freedom down under necessity, and lets the self
expire in infinite resignation, can just as little appeal to scientific
grounds as he who chooses to struggle for freedom's personal ideals and rely
on the omnipotence of love. Both choose in faith, and the choice of faith may
just as little efface the scientific consciousness in the latter as in the
former. If we must now concede that only those principles whose demands are
justified by scientific grounds are scientific, does it not then follow with
irrefutable consequence that the principle of faith cannot be a scientific
principle?

It is not enough to present the faith-principle and the knowledge-principle as
heterogeneous; for even when we hold to Newton's rule, \textit{principia
præter necessitatem non esse multiplicanda}, it cannot be denied that within
science's own domain there are heterogeneous principles. Some of these—such as
Materialism's, Spiritualism's, Positivism's, Idealism's—combat one another
mutually, lay bare by reciprocal criticism one another's one-sidedness, and in
this way prepare their own dissolution; but there are others that do not let
themselves be neutralized in this way. The heterogeneous special sciences have
not only each its material, its goal, and its method, but also, in consequence
of the course of the investigations, each its peculiar principle: no
encyclopedist has yet been able to derive all the sciences from one single
principle; but since the sciences nevertheless, insofar as they are sciences,
are homogeneous, the heterogeneity of the principles can only be relative. If
we now limit ourselves to calling the faith-principle and the
knowledge-principle heterogeneous principles, then we make ourselves guilty of
an ambiguity. There must, after all, be a qualitatively decisive difference
between principles which, with all their heterogeneity, fall within science,
and a principle which, on account of its unconditionally subjective, personal
peculiarity, falls outside
% p. 72
all science; in order therefore to remove the ambiguity we must expressly
impress that the principles of Faith and Knowledge are absolutely
heterogeneous.

It might well seem as though the dualistic separation of Faith and Knowledge
had significance only for those who choose to believe in a positive religion's
supernatural facts, but not for those who hold to reason and build their
religious conviction on an objective, scientific foundation. That is a great
error; Comte's remark about ``the modern revolt of the spirit against the
heart'' is, from science's standpoint, a forceful expression of subjectivity's
undeniable claim to precedence in the religious. Let one say: ``I cannot
believe in an almighty, free, and fatherly God, for science has convinced me
that such a being cannot possibly exist''; another: ``I must necessarily
believe in a personal God, for science has, in speculative dogmatics,
irrefutably proved that God is personal''—then both make themselves guilty of
one and the same misapprehension of subjectivity's share in the demonstration.

\emph{No personally coherent view of life is possible without faith, no
essential faith is possible without decisive self-determination, no decisive
self-determination is possible without a choice, no choice is possible without
a self-consciously choosing subjectivity responsible for its choice.} Could
such a preponderance of subjectivity in the religious be supposed to stand in
conflict with science and sound reason?

Science can do no more than lay before consciousness the possibilities between
which a choice must be made, and illuminate the real grounds; but transform
the grounds into motives, determine the specific weight of the motives, and
exempt subjectivity from the responsibility of the choice, it cannot. It might
well seem as though the responsible subjectivity now became at once so
independent and self-willed that it could prefer a view of life as mythically
fantastic as it liked, without owing either science or reason any account for
it. This is indeed also the case
% p. 73
with such individualities as do not see what science actually is, and in whom
reason is still a minor child; but it by no means holds of those who have made
a deeper insight into the essence of science. Even when, from inner personal
need, they choose to believe in the revealed religion and preserve the
inherited reverence for the Church and its institutions, they must constantly
render account to science, in that they render account to themselves; in no
case may man arbitrarily extinguish the light of reason, but reason's pure
light burns clearest in science.

It then becomes a life-task \emph{to unite religious faith with scientific
consciousness.} How is this possible? The positive content that is
appropriated in faith cannot, as Littré assumes, be fetched from science; and
yet the scientific consciousness must, so far as it is able, take part in
subjecting this content to a critical examination. We meet here, on the
highest stage of the spiritual life, again the double-sided functions. The
purifying criticism is at once a faith-criticism and a science-criticism. The
faith-criticism separates out all that does not originally concern the essence
of faith; the whole apparatus of tradition in outward historical form it
regards as a clothing, a husk, in which the holy content is preserved. It by
no means wishes to disparage the clothing or break apart the husk; but it will
nevertheless appropriate only what the believing subjectivity, from inner
experience, needs as an indispensable condition for the development of its
freedom-life. How the science-criticism is present everywhere can be known
from the apprehension and evaluation of what is handed down. We can illuminate
the reciprocal action under the standpoint of the two categories: possibility
and probability.

\emph{The improbable must be believed, but the impossible cannot be
believed}—that is a fundamental principle for all revelation-faith. The
revelation-faith presupposes the miracle's possibility, science stands firm on
the miracle's improbability; but Faith and Knowledge agree in distinguishing
between possibility and actuality. The miracle's possibility points toward what
we must call
% p. 74
the perennial wonder, that is, toward the general presupposition that the
eternal self-power continuously stands in a freedom-relation, an actual
will-relation, to the natural order of things, in which everything takes place
according to laws. That rational knowledge can raise no warranted objection
against this has been shown in the foregoing.

It would look more precarious if the believing subjectivity, in enthusiastic
rashness, set about reasoning thus: ``If the spirit is almighty and free, then
the absolutely free self-power must, after all, be able to change nature's
course at pleasure; the real miracles then stand as testimonies of an
unsearchable almighty-will; one must either accept them or reject them
unconditionally; here there can be no question of criticism, here only faith
or unbelief holds.'' But the reasoning that thus annihilates all criticism is
itself uncritical and false. In an abstract outward, objective sense it is
quite true that real miracles withdraw themselves from all criticism; but in
such an outward, blind facticity the real miracles have no significance
either, neither for Faith nor for Knowledge: they stand as phenomenal
monstrosities, absolutely incomprehensible \textit{τέρατα}. In the moment that
a comprehensible meaning, a religious significance, is ascribed to them, the
subjective comes into movement, and the finite relation between subjectivity
and the objective actuality entitles us to a historical use of the measure of
probability.

No real miracle has ever been demonstrated by compelling historical
testimonies. The miracle is an outward sign of an essential inward, of
freedom's power over necessity, of life's victory over death; the miracle is
sign and symbol, and becomes no more, even if its actuality were confirmed by
ten thousand sworn eyewitnesses. The given sign is at once an inner and an
outer fact; the inner is the ideal in the actuality, the essential
faith-content, the outer is the wonderful phenomenon that makes visible the
presence of the inner. The inner, the essential, is the extraordinary of
freedom and love, that which is improbable for the finite understanding, which
can be appropriated subjectively, tested subjectively, and
% p. 75
be present to consciousness at all times; the outward, the phenomenal, is the
past, it belongs to a definite time, and the age's more or less adequate
apprehension of it is to be determined by approximation according to
probability. The faith-criticism can then appropriate the improbable but yet
possible, and the science-criticism, as regards this possible's outer
appearance, hold to the probable, and both agree that the miraculous in the
sign is \emph{spirit-reality, reality for the spirit.} Spirit-reality means
that here there really is a faith-content, an unambiguous revelation of the
divine will; reality for the spirit means that the outer may not be held fast
in so sensibly immediate an outwardness that there can be talk of a breach of
nature's laws.

That one, from an exclusively objective, scientific standpoint, must regard
such an interpretation of the revelation-phenomena, indeed of the essence of
revelation, as veiled and ambiguous, is quite in order\footnote{Knowledge's
fundamental principle is that nothing may be assumed to be real which
demonstrably conflicts with the law of reality. Faith's fundamental principle
is that a real sign of the saving omnipotence may not be declared impossible
because it makes an exception to the laws known to us. Christ's resurrection is
an eloquent example. Without science-criticism the believer will, in all
credulity, take the outward in the event for something so real that the
continuity of law is broken, and that so conspicuously that the one miracle
comes to contain at least six miracles. Without faith-criticism the knower will
unconditionally reject an actuality that six times contradicts itself, and
assign it a place in the world of imaginings. If an honest reconciliation
between Faith and Knowledge is to be possible, then it must be seen that here
there is right and wrong on both sides. The believer is right insofar as he
holds fast the fact as a real sign of omnipotence; but when he then, relying on
mutually discordant accounts, apprehends the real in such a way that it becomes
a sign of the unbounded arbitrariness with which the almighty freedom can turn
all the laws of nature and spirit upside down, and makes such a sign of
arbitrariness into a sign of salvation, then he is grossly wrong. The knower is
right in that he asks, for all actuality, about laws of actuality, and guards
himself against assuming anything whatever for a real fact when the laws
contradict one another; but when he then swears by a natural-necessity whose
conditions experience has secured for him only by empirical induction, to such
a degree that he \textit{a priori} denies freedom's possibilities, then he is
wrong, even in science. And what
% (fn cont. p. 76)
then becomes the result? That there can be facts of so peculiar a nature that
the kernel, the essential content, may not be denied, and moreover of so
extraordinary an appearance that the outer form of the actuality cannot
possibly be definitively determined. If now such a fact stands in a contingent
relation to the religious faith, then it can find its place among myths and
legends; if, on the other hand, it contains a truth decisive for the
faith-life, then the sensible outer must be apprehended as a phenomenon making
visible, a sign and symbol of the real inner; in other words: the fact must be
determined as \emph{spirit-reality, reality for the spirit.} The question of
spirit-reality is, as one will see from this, not a personal faith-question,
but an objective knowledge-question. How the believer comes to terms with
himself about the fact's outer historical actuality must be his own affair;
only this stands fast: he cannot possibly have the science-criticism with him
without recognizing law-determined boundaries, boundaries that cannot be
overstepped without \emph{self-contradiction} setting in.}. Since here it is a
question of science-
% p. 76
criticism, one forgets every moment where we are, and demands the
impossible—demands that the fact itself shall be scientifically proved. That
is a thoughtlessness. The science-criticism is, so to speak, only present in
the probate court in order to secure the apprehension against warranted
objection, while the faith-criticism asserts its own peculiar principle,
absolutely heterogeneous with the objectivity of knowledge, and faith
appropriates the inheritance.

Religious faith is, in its first immediacy, a faith in something historical, an
authority-faith that stands firm on a holy, infallible tradition; the naive
faith will hear nothing either of probable and improbable or of the difference
between outer and inner, it demands certainty, historical certainty. Since now
one or another outward authority goes surety for the historical certainty
without entering upon critical investigations, it happens that faith from the
beginning gets an admixture of superstition, an admixture that is extremely
difficult to separate out when the faith-truth itself is to be preserved. If an
outward authority is to vouch for the truth, then the criterion of truth is
lost; for the authority acts not convincingly, but imposingly. Every
faith-community will then come forward with its myths and its miracles and set
about impressing the unstable soul; here there is then no question of which of
the
% p. 77
many religions has the deepest content of truth, but of which of them, by its
signs and symbols, has the best hold for impressing. It is a distinguishing
mark of the revealed—i.e. the \emph{subjectively spirit-clear}—religion, that
it, through the outward authority, everywhere in the name of truth points
toward the inward, toward the spirit's authority, the spirit's testimony.

Thus, then, the doctrine of two absolutely heterogeneous principles is by no
means any directive to tear religion loose from science, or faith from reason;
the dualistic separation is precisely the condition for attaining a reconciling
unity: of Faith and Knowledge too it holds that opposite poles attract one
another.

\end{document}
