% ============================================================
%  H. Høffding, Relation som Kategori:
%  en erkendelsesteoretisk Undersøgelse
%  Det Kgl. Danske Videnskabernes Selskab,
%  Filosofiske Meddelelser I, 3.
%  København: Bianco Lunos Bogtrykkeri, 1921.
%
%  TRANSLATION (English)
%  Translator: [HH]
%  Completed: §3d.δ The New Theory of Relativity (pp. 72--76)
% ============================================================

\documentclass[12pt]{article}
\usepackage{fontspec}
\usepackage{unicode-math}
\setmainfont{Libertinus Serif}
\setmathfont{Libertinus Math}
\usepackage[english]{babel}
\usepackage[protrusion=true]{microtype}
\usepackage[margin=1.4in]{geometry}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,
  pdftitle={Relation as Category},
  pdfauthor={Harald Høffding}]{hyperref}
\setlength{\headheight}{15pt}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\textit{Relation as Category}}
\fancyhead[LO]{\textit{\rightmark}}
\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt}
\onehalfspacing

\renewcommand{\thesection}{\arabic{section}.}
\renewcommand{\thesubsection}{\alph{subsection}.}

\title{\textbf{Relation as Category}\\[0.5em]
  \large An Epistemological Investigation}
\author{Harald Høffding}
\date{Translated from Danish by [HH]\\[0.4em]
  \textit{Det Kgl.\ Danske Videnskabernes Selskab,
  Filosofiske Meddelelser} I, 3.\\
  Copenhagen: Bianco Lunos Bogtrykkeri, 1921}

\begin{document}
\maketitle
\tableofcontents
\bigskip


% ============================================================
%  1. THE HISTORY OF THE CONCEPT OF RELATION
% ============================================================

\section{The History of the Concept of Relation}
\label{sec:1}
% pp. 3--15

% [translation to be added]


% ============================================================
%  2. THE CONCEPT OF RELATION IN PSYCHOLOGY
% ============================================================

\section{The Concept of Relation in Psychology}
\label{sec:2}
% pp. 16--25

% [translation to be added]


% ============================================================
%  3. THE CONCEPT OF RELATION IN EPISTEMOLOGY
% ============================================================

\section{The Concept of Relation in Epistemology}
\label{sec:3}
% pp. 26--76

\subsection{The Concept of Relation and the Other Fundamental Categories}
\label{sec:3a}
% pp. 26--32

% [translation to be added]

\subsection{The Concept of Relation and the Formal Categories}
\label{sec:3b}
% pp. 33--44

\subsubsection*{α.~Pure Quality-Series}
\addcontentsline{toc}{subsubsection}{α.~Pure Quality-Series}
\label{sec:3b-alpha}
% pp. 33--39

% [translation to be added]

\subsubsection*{β.~Identity and Relation}
\addcontentsline{toc}{subsubsection}{β.~Identity and Relation}
\label{sec:3b-beta}
% pp. 40--43

% [translation to be added]

\subsubsection*{γ.~The Relations of Negation and Rationality}
\addcontentsline{toc}{subsubsection}{γ.~The Relations of Negation and Rationality}
\label{sec:3b-gamma}
% p. 44

% [translation to be added]

\subsection{The Concept of Relation and the Real Categories}
\label{sec:3c}
% pp. 45--57

\subsubsection*{α.~Causality, Reality, and Relation}
\addcontentsline{toc}{subsubsection}{α.~Causality, Reality, and Relation}
\label{sec:3c-alpha}
% pp. 45--54

% [translation to be added]

\subsubsection*{β.~The Relations of Totality and Development}
\addcontentsline{toc}{subsubsection}{β.~The Relations of Totality and Development}
\label{sec:3c-beta}
% pp. 55--57

% [translation to be added]

\subsection{The Relation Subject---Object}
\label{sec:3d}
% pp. 58--76

\subsubsection*{α.~Introduction}
\addcontentsline{toc}{subsubsection}{α.~Introduction}
\label{sec:3d-alpha}
% pp. 58--61

% [translation to be added]

\subsubsection*{β.~On Copernicanism}
\addcontentsline{toc}{subsubsection}{β.~On Copernicanism}
\label{sec:3d-beta}
% pp. 62--66

% [translation to be added]

\subsubsection*{γ.~Philosophical Standpoints}
\addcontentsline{toc}{subsubsection}{γ.~Philosophical Standpoints}
\label{sec:3d-gamma}
% pp. 67--71

% [translation to be added]

\subsubsection*{δ.~The New Theory of Relativity}
\addcontentsline{toc}{subsubsection}{δ.~The New Theory of Relativity}
\label{sec:3d-delta}
% pp. 72--76

\noindent Despite the great expansion of the horizon that Copernicanism
gave impetus to, natural science had held fast to the concepts of absolute
place and absolute motion, which had their justification only within the
old worldview with its bounded and fixed frame. Galileo's and Newton's
absolute space replaced the space bounded by the eighth sphere. But there
came a time when the concept of relation here too made its consequences
felt, and not only with respect to the relations among objects, but also
with respect to the relation between objects and the knowing subject.

It was the first proposition of kinematics, and thereby of natural
science, that was first examined from the standpoint of the concept of
relation. Neumann (\emph{Ueber die Prinzipien der Galilei-Newton'schen
Theorie}, 1870) drew attention to the fact that the law of inertia,
according to which the velocity and direction of a body change only through
external influence, necessarily presupposes in each individual case the
specification of the particular place in relation to which velocity and
direction are preserved when no external influence is exerted. Neumann
could find no place that could lay claim to being taken as the starting
point for an absolute coordinate system, but held that one must
hypothetically assume a center in order for the concept of absolute motion
to retain its meaning. Mach, who at roughly the same time had arrived by
his own path at the insight that the law of inertia rested on the
presupposition of a determinate coordinate system, did not find Neumann's
solution satisfying. He demanded that the concept of relation be carried
through. The dogmatic conception of the law of inertia, in his view, only
falls away when one becomes clear that in itself any body in the world can
be used as a starting point, and that in practice in each individual case
we use one body or another as the presupposed fixed starting point for the
coordinates in relation to which the velocity and direction of a motion are
determined (\emph{Die Erhaltung der Arbeit}, 1872). This is the path that
has in fact always been taken, only within the old worldview, which rested
precisely on the presupposition of an absolute center and an absolutely
fixed boundary-sphere of the world, no choice was required. Mach lays
great weight throughout on the historical presuppositions on which the
research of each period rests, but then shows how these presuppositions
vary from age to age, and that none of them can lay claim to absolute
necessity and intelligibility. What matters here always comes down to what
French physicists call \emph{points de repère}.\footnote{Cf.\ e.g.\
  Painlevé (\emph{Mécanique}, p.~399) [in the collective work \emph{De la
  Méthode dans les Sciences}, 1909]: ``\ldots\ les mouvements absolus,
  c'est-à-dire au fond les mouvements convenablement repérés.'' [i.e.,
  absolute motions are, at bottom, motions suitably referred to reference
  points.]}

An even more radical application of the concept of relation appeared at
the beginning of the twentieth century. What is of interest in this
connection is that it has proved necessary to take special account of the
apprehending subject and its standpoint. Copernicus had already pressed
this point as regards the apprehension of the motions of celestial bodies,
and Neumann and Mach as regards the velocity and direction of motion.
Einstein\footnote{\emph{Über die spezielle und die allgemeine
  Relativitätstheorie} (1916).} has now sought to show that what holds
for space and motion holds also for time. That this was not discovered
earlier is, according to Einstein, to be explained by the fact that
attention was paid only to velocities that were very small in relation to
the velocity of light. The final temporal measure becomes, for modern
physics, the velocity of light, whereas for Aristotle, and even for the
Copernicans, it was drawn from the motions of celestial bodies. (Bruno's
doctrine of the relativity of the concept of time rests on this measure.
See above.) On the basis of the electromagnetic theory of physical
phenomena, for which the mechanical conception of nature stands as only a
special case, Einstein seeks to show that there can be demonstrated
circumstances under which two observers, of whom one is attached to fixed
and the other to moving axes, but both at rest relative to their own axes,
will not arrive at the same apprehension of time: what for one observer
stands as a relation of simultaneity between two events, will for the other
stand as a relation of succession. A temporal specification acquires
meaning only when it is stated in relation to which coordinate system it
holds, and in particular whether that system is itself relatively at rest
or in motion.

If one were to hold that an illusion must be present in one or the other
observer, the answer from the Einsteinian standpoint is that each observer
has his own system of measurement, which is just as legitimate as
another's, though it is of course perfectly possible for one observer to
attempt to place himself at another's standpoint and reason from
it.\footnote{Langevin, in \emph{Bulletin de la Société française de
  Philosophie}, 11 October 1911. --- It must presumably be assumed that
  the observer who can place himself at another observer's standpoint knows
  sufficient physics.}

The facts and inferences on which the Einsteinian theory rests are still
a matter of dispute, and views stand sharply opposed to one another. In
such a dispute the philosopher as such cannot intervene, and philosophy
has time to wait. Whether physics can really demonstrate situations in
which what stands as simultaneity for one observer stands for another
observer as successive, physicists must determine among themselves. If
this must be answered with yes, we stand before the significant result
that natural science has once again, as with the Copernican hypothesis,
by its own path come to see the necessity of not merely holding to the
content of observations, but of taking account of the observer's
presuppositions. In an interesting discussion in the \emph{Aristotelian
Society} (1919), Whitehead impressed upon us the significance of what he
aptly called ``the percipient event,''\footnote{See also his
  \emph{Principles of Natural Knowledge} (1919), p.~68. --- Whitehead's
  conception differs from Einstein's, however, in that ``the percipient
  event'' for him is only one factor alongside many others, and not, as in
  Einstein, of essential significance for all measurement of time and
  space. Cf.\ on this point Haldane: \emph{The Reign of Relativity}
  (1921), pp.~69\,f.; 107\,ff.} and Nicholson declared that among the
data of ultimate character there must also be counted ``the essential
framework of our modes of perception.''

Against the Einsteinian hypothesis, as against earlier demonstrations of
the epistemological significance of the concept of
relation,\footnote{Cf.\ \emph{Den menneskelige Tanke} [The Human Mind],
  p.~304.} it has been urged that there can only be one truth in each
individual case, and that it would contradict the principle of
non-contradiction if two observers, of whom one declares simultaneous what
the other declares successive, were both to be right.\footnote{Kroman, in
  \emph{Tidsskrift for Fysik}, XIV, p.~15.} But there is no reason to fear
for the unity of truth, even if Einstein should be right. The real truth
lies in this case in the necessity, physically demonstrated, that different
observers under certain determinate conditions arrive at different
apprehensions of time. Each individual observer apprehends time as he must
under his conditions. But if an observer is also a physicist, he will
(assuming the correctness of Einstein's theory) be able to explain to us
why another observer arrives at a different apprehension than he himself
does. (Cf.\ Newton's and Leibniz's emphasis on the lawfulness of motion as
the expression of its objectivity. Above, p.~66.) Naturally, the physical
explanation of the different observers' different apprehensions of time
itself rests on certain presuppositions, presupposes a coordinate system
from which the coordinate systems of the various observers are determined,
and here there is the possibility of a new discussion. When this discussion
comes to an end, it is because no standpoints can be demonstrated that are
still more fundamental. Kant long since maintained that all necessity is
hypothetical. But it is necessity nonetheless --- or rather: precisely
because of this it is necessity. A necessity without presuppositions we do
not know.

If the new theory of relativity wins continued confirmation, what comes to
hand will not merely be a result of physical research that is interesting
in and of itself. This theory also contains, as Einstein himself has
declared, a spur to continued inquiry. If it is the case that there is the
possibility of different systems of relations in the apprehension of an
object, this constitutes precisely an invitation to discover all the
particular relations that lie at the basis in each individual case, and
inquiry is thus led ever further rather than closing off in dogmatic
fashion, as if the presuppositions one had hitherto worked with were the
only possible ones. There are, from an epistemological point of view, two
interests that attach to the concept of relation. One is that light is
cast on the nature and conditions of our knowledge; the other is that new
tasks are indicated, a \emph{plus ultra} appears, in that every thought
points beyond itself. ---

Campanella said in his apology for Galileo that if Galileo were right,
one would have to philosophize in a new way. But if Einstein is right,
one need not therefore philosophize in a new
way.\footnote{Einstein himself agrees with this. In his preface to Lucien
  Fabre's book \emph{Les théories d'Einstein} (1921) he says: ``The theory
  of relativity neither can nor will provide a world-system, but only
  indicate a limitation to which the laws of nature are subject.'' That it
  is not merely a matter of a limit, we have seen above.}
Even from a purely philosophical point of view, as the foregoing
presentation has shown, attention was long since drawn to the significance
of the concept of relation for knowledge and worldview. But if Einstein is
right, there now lies before us a new, unexpected, but significant
experience in that respect. ---

Through the necessity of taking account of the knowing subject, and
through the possibility of a conflict between the each-for-itself
necessary apprehensions of different knowing subjects, the newer natural
science --- in Einstein as in Copernicus --- offers an analogy to what
asserts itself powerfully in the domain of the human sciences, as I shall
now attempt to show.


% ============================================================
%  4. THE CONCEPT OF RELATION IN ETHICS
% ============================================================

\section{The Concept of Relation in Ethics}
\label{sec:4}
% pp. 77--98

\subsection{Historical Research and the Concept of Relation}
\label{sec:4a}
% pp. 77--80

% [translation to be added]

\subsection{Ethics and the Psychological Relations}
\label{sec:4b}
% pp. 81--86

% [translation to be added]

\subsection{Ethics and the Historical Relations}
\label{sec:4c}
% pp. 87--89

% [translation to be added]

\subsection{Ethics and the Individual Relations}
\label{sec:4d}
% pp. 90--91

% [translation to be added]

\subsection{Kant's Ethics and the Concept of Relation}
\label{sec:4e}
% pp. 92--93

% [translation to be added]

\subsection{Philosophy of Religion and the Concept of Relation}
\label{sec:4f}
% pp. 94--98

% [translation to be added]


% ============================================================
%  5. THE CONCEPT OF RELATION IN COSMOLOGY
% ============================================================

\section{The Concept of Relation in Cosmology}
\label{sec:5}
% pp. 99--[end]

% [translation to be added]

\end{document}
