% ============================================================
%  F. C. Sibbern, On the Concept, Nature, and Essence of Philosophy.
%    A Presentation of the Propaedeutic of Philosophy.
%  Copenhagen: At the Author's Own Expense (Printed by Director
%    Jens Hostrup Schultz), 1843.
%  TRANSLATION (English) — SCAFFOLD.
%  Source of truth for translation: transcription.tex (Danish, COMPLETE + proofed,
%    §§1--21 + Fortale + Rettelser; verified compile 59 pp., 0/0).
%  Method: ../../../TRANSLATION-PLAYBOOK.md. Fill each "% [text to be added: ...]"
%    marker batch by batch (~10 printed pp.), mirroring transcription 1:1.
%  Same policy as the companion volume *Om Erkjendelse og Granskning*:
%    - Register: lightly modernized (faithful meaning, Sibbern's long periods untangled).
%    - Key terms: on first use in a section, gloss the Danish in brackets, e.g.
%      "fundamental cognition [Grunderkjendelse]". Maintain GLOSSARY.md.
%    - Layout: English-only, section-keyed; mirrors the transcription structure 1:1.
%  Page-offset (Danish scan, for reference): printed page = PDF page - 11.
%  Book-specific: emphasis in this book is BOLD Fraktur (Fett), already encoded as
%    \emph{} in the transcription -> keep as \emph{} (renders italic). Latin already in
%    \textit{}; German block-quotes are plain (Fraktur in the source); no real Greek.
%  See TRANSLATION-NOTES.md for state, decisions, and the section->page map.
% ============================================================
\documentclass[12pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{libertinus}
\usepackage{libertinust1math}
\usepackage{textalpha}   % real Greek glyphs if any arise (only transliterated names here)
\usepackage[english]{babel}
\usepackage{enumitem}    % for \arabic*)-style lists (playbook: easy to forget)
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{setspace}
\usepackage[margin=1.2in]{geometry}
\usepackage{fancyhdr}
\setlength{\headheight}{28pt}
\addtolength{\topmargin}{-13.5pt}
\usepackage{hyperref}

\hypersetup{
  pdftitle={On the Concept, Nature, and Essence of Philosophy},
  pdfauthor={F. C. Sibbern},
  hidelinks
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[CE]{\textit{\leftmark}}
\fancyhead[CO]{\textit{\rightmark}}

\setstretch{1.3}

% Group-heading helper (the centered bold rubrics of the Indhold / body)
\newcommand{\grouphead}[1]{%
  \bigskip\begin{center}\large\textbf{#1}\end{center}\smallskip}

% TITLE — rendering to confirm with Hans (see TRANSLATION-NOTES).
\title{\textbf{On}\\[6pt]
  \Huge\textbf{the Concept, Nature,}\\[6pt]
  \LARGE\textbf{and Essence of Philosophy.}\\[14pt]
  \normalsize A Presentation of\\[4pt]
  \Large\textbf{the Propaedeutic of Philosophy.}\\[16pt]
  \normalsize By\\[6pt]
  \large Dr.\ Frederik Christian Sibbern,\\[4pt]
  \normalsize Professor of Philosophy.}
\author{}
\date{Copenhagen, 1843.\\
  At the Author's Own Expense\\
  printed by Director Jens Hostrup Schultz,\\
  Royal and University Printer.}

\begin{document}
\maketitle
\thispagestyle{empty}
\cleardoublepage

%% ============================================================
%%  PREFACE  (Fortale, PDF 8--9). Transcription L78--111.
%% ============================================================
\chapter*{Preface.}
\markboth{Preface}{Preface}

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

Had I not, up to now, let every writing of mine that has come forward
independently in the literature be accompanied by a preface---not exactly
because I have expressly reckoned such a thing among the proper appointments of
a book, but because each time it has felt natural to me---I should this time
have been inclined to dispense with any preface at all, since I had nothing to
say that could be its subject. For the truth is that I would rather have
refrained from expressly declaring that what several of my readers will perhaps
expect to find stated here I mean to leave to them to settle for themselves. I
mean this: in what relation the present writing stands to my book
\textit{Om Erkjendelse og Granskning} [On Cognition and Inquiry], published at
the close of the year 1821. If I am to declare myself on the matter---for how I
intend to use both writings in my lectures on the propaedeutic of philosophy
[Philosophiens Propædeutik] does not belong here to explain---then, so far as
the public is concerned, I would say nothing more than this: I could wish that,
in its relation to the present book, the reader would regard that earlier one
as though it were by another author, one whom I had drawn upon and to whom I had
therefore often referred, while yet having a different object before me than he.
For in that earlier work the object is science [Videnskab] and scientific study
in general, and it is offered as an introduction to academic study at large.
Here, by contrast, I have had philosophy in particular before me, and have
therefore had more to enter into with respect to \emph{this} than had come
forward in that other work. ---Finally, I ask that among the errata [Rettelser],
those marked \textbf{NB} in particular may be heeded.

\bigskip
\noindent\textbf{Copenhagen,} the 28th of October 1843.
\nopagebreak
\begin{flushright}\textbf{Sibbern.}\end{flushright}

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

\mainmatter

%% ============================================================
%%  PRELIMINARY REMARKS  (§§ 1--2, printed pp. 1--6)
%% ============================================================
\grouphead{Preliminary Remarks.}

\begin{center}\large\S~1.\end{center}
\markboth{\S~1}{\S~1}
\begin{center}What is to be undertaken here, and where we set out from.\end{center}

It is the concept, nature, and essence of philosophy that we shall here seek to
clarify for ourselves. We cannot, then, confine ourselves to the question: What
is meant by philosophy? but have to do with the far more telling question: What
is philosophy? Yet we must of course, straightaway and from the very
first, take account of the word's meaning, attending to what is called
philosophy---and has through the course of history been so called, and has
asserted itself under that name. For although the concept we are to seek to give
is to be developed and determined as reflection on the matter itself demands,
still the question---whether what is thereby set up corresponds to what one
calls philosophy, or whether perhaps, with all the validity it may have in and
for itself, it is something quite other than what one has in view by the word
philosophy---can only be settled by taking into consideration the meaning with
which this name passes and holds good. By the name, the object to be spoken of
is pointed out, and one must from the first take care not to end up speaking of
some other object instead of this one. It is most correct, then, to set out from
a given representation [Forestilling] of what philosophy is---a given meaning of
the word.

But the matter itself, thus pointed to by the name and thereby lifted out for
us, we must then surely consider entirely as it wills to be seen, and must
pursue it through its whole consequence, so that the concept, further on, cannot
be bound to, nor in its whole carrying-through be conditioned by, that given
representation.

In this way we do indeed come to consider philosophy in philosophy's own light,
and thus come to lay our own philosophy at the foundation; for we are now to
philosophize our way to the cognition [Erkjendelse] we seek, and in all
philosophizing philosophy at once asserts itself. But this circularity is a
consequence of the whole position of the matter; for philosophy is but one
member within the world-totality [Verdenstotalitet] which philosophy itself---%
according to the concept of it that we must presuppose as given---is to clarify
for us.

{\small\emph{Note~1.} There naturally arises here the question whether the
concept of philosophy ought not to be treated in two quite different places:
first, in the considerations introductory to philosophy, where the philosophical
undertaking cannot of course be dealt with without laying at its foundation a
concept of what philosophy is; ---and second, in the very midst of philosophy's
system, once the speculative fundamental cognition [Grunderkjendelse] has been so
far attained and developed that philosophy---stepping forward as a co-engaging
member in the spiritual life unfolding in consciousness---comes to be considered
from the speculative---more precisely a psycho-biological---standpoint. One
readily sees that, although philosophy would lead us back to the first
fundamental moments and beginning (\textit{initia}) of all things, it
nevertheless itself belongs in the very midst of the world's course, where it
can first arise in the higher life of consciousness, or in the midst of the
sphere of spirituality, and must then show itself as an essential principal
moment in the totality of spiritual life. The question of philosophy's proper
essence [Væsen] belongs, then, there where poetry, morality, and religiosity too
are to constitute themselves for us in their essential belonging-together with
the totality of spiritual life. ---\par}

{\small The question then naturally arises whether we ought not to distinguish,
from this spirit-philosophical and properly speculative consideration of
philosophy's essence, another, preliminary one belonging to a mere introduction
to philosophy---which we might then call the merely propaedeutic, and which would
at the same time be called the merely explicative. This one would merely pursue
the given concept of philosophy through all its consequence and thereafter
determine and expound it, whereas the other would let philosophy itself
constitute itself for us and thereby show itself in its proper essence. This last
would then be the one which, if the very expressive designation the philosophy of
philosophy is to be adopted, would have to be furnished with that name.\par}

{\small Yet to such a merely propaedeutic exposition of the concept of philosophy
we neither can nor will here confine ourselves, nor keep it separated from the
speculative; for it is philosophy itself---not the concept that has formed about
it---that we have before us here, and the clarification of philosophy, if it is
to be complete, cannot rest at that merely propaedeutic level.\par}

{\small In the meantime we must surely seek constantly to remain aware how far
our concept, as it more and more determines and develops itself for us, remains
within the domain [Gebeet] of the merely propaedeutic---so that it can be set up
as valid with respect to the various philosophical parties (Platonists,
Skeptics, Kantians, Lockeans, and so forth), and can thus be laid at the
foundation in the history of philosophy---and how far, on the other hand, it goes
further, and so is already developed according to the dictate of a definite
philosophical fundamental view [Grundanskuelse].\par}

{\small\emph{Note~2.} One will readily see that it would be altogether
inappropriate to give the first and fundamental part of philosophy itself---which
Hegel has called Logic---the name of the philosophy of philosophy, as some have
done. Hegel, too, is very far from having been able to give any occasion for
this, since with him the philosophy of philosophy---though this name is not found
in him---appears at the close of the third and last part of his whole
system.\par}

{\small\emph{Note~3.} Concerning the expression \emph{nature and essence} and its
relation to the expression \emph{concept}, be it noted that what is here most
immediately and first sought is the concept of philosophy and its exact
determination, but that this concept is then to unfold itself for us, or to step
out into the whole sum-total [Indbegreb] of its consequence, or into its whole
content---which is what we have in view by the expression \emph{nature and
essence}. (I refer here to my \textit{Logik}, \S~35.)\par}

\begin{center}\large\S~2.\end{center}
\markboth{\S~2}{\S~2}
\begin{center}A survey of the whole course of what follows.\end{center}

If, then, we ask first of all: What does one understand by philosophy?---whereby
we are only to indicate provisionally just so much that we may know with
certainty what it is that is to be treated, but must also at once come to
indicate the first points to which we have to direct our attention---then we note
that the received concept implies that philosophy is to aim at a certain
fundamental cognition [Grunderkjendelse] serving for a deeper clarification of
the whole of existence [Tilværelse], but that it is to form itself through
thinking and reflection, and thus to become a cognition that is to be aware of
its own inner groundedness and able to give an account of it. In this way, then,
we come to let our consideration take the following course:

First and foremost we take into consideration that 1) philosophy is a whole body
of cognizing [Indbegreb af Erkjenden], and we therefore consider the nature and
essence of cognition in general.

Next we observe that, since philosophy aims at a certain deeper fundamental
cognition that looks toward the last or innermost fundamental relations
[Grundforhold]---yet such a cognition shows itself as philosophy only when it
comes forward as sprung from rational cognition [Fornufterkjendelse], and
especially only when investigation [Efterforskning] or inquiry [Randsagelse] has
thereby been set in motion---we are hereby led to consider the endeavor of
cognition in general in its transition from its first principal stage, the stage
of first, immediate cognizing, to its second principal stage, the mediated but at
the same time higher stage of cognizing called forth by the spirit of inquiry
[Randsagelsesaand]---the stage at which cognition becomes a cognizing according
to insight, and at the same time---since here testing must take place, an account
must be rendered, grounds come under consideration, and groundedness is brought
to light and made good---a cognition in force of grounds [i Kraft af Grunde];
which then lets us see philosophy as that which 2) is a whole body of intelligent
and rational cognizing.

To this attaches, further, the observation that, since this whole body must form
a well-connected totality [Heelhed], philosophy in its carrying-through and
execution must become 3) a scientific cognition---whereby we are brought to
consider science [Videnskab] and scientific character [Videnskabelighed] in
general.

But since---according to the received concept---not all scientific cognizing is
philosophy, then, by asking ourselves what sort of science it is in which
philosophy consists, we come to become aware that the endeavor to press through
to the grounds cannot but lead to the forming of 4) a \emph{science} that in a
certain \emph{all-comprehending manner} aims at ultimate grounds and an ultimate
fundamental cognition, and that it is this above all that one has in view when
one speaks of philosophy; and here, when the question is now raised whether it is
the ultimate grounds and fundamental relations of all cognition, or of the whole
of existence, that are thereby meant, it will appear that---although a twofold
principal task [Hovedopgave] for philosophy has in this respect formed, and had
to form---it is nevertheless in the end only 5) the ultimate grounds and
fundamental relations of \emph{existence}, or as one is wont to say, of \emph{all
things}, in the finding-out and cognition of which the all-comprehending
fundamental cognition is to form, wherein philosophy consists.

This determination of the concept of philosophy, handed down from ancient times,
from which we are thus brought to set out, shows us philosophy as that which
comes forward over against, and in a certain opposition to, the sum-total of all
that cognizing whereby everything is given us and presents itself to us---%
everything whose ultimate grounds and fundamental relations are now to be in
question, and which we are accordingly prompted and called upon to gather for
ourselves into a single overview.

In its pursuit through the various principal respects or principal directions, we
shall next see the concept of philosophy lead to a whole cycle of conceptual
determinations, which can be developed into a whole cycle of definitions.

Thereupon we shall proceed to consider philosophy in its execution. But when we
now first take into consideration what ultimate grounds are, and how matters
stand with our pressing toward them, the question of philosophy's foundation
[Grundlag] and starting-point will prove to have to precede the question of the
execution of philosophy's work---to which, then, once we have thereupon fully
answered for ourselves the question what philosophy is, a consideration of
philosophy's significance in the totality of existence will naturally be able to
succeed.

With this, the course we shall take is fully indicated.

{\small\emph{Note.} It must be noted here at once---what will be more closely
treated further on in its place---that alongside philosophy proper, of which alone
there is question here, there is a whole body of sciences which, without being
philosophy proper, are nevertheless also called philosophical, but which I, to
distinguish them from the former (see further on, \S~11), give the name of
explicative philosophy, and to which I here pay no regard whatever.\par}

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  INTRODUCTORY CONSIDERATIONS. ON COGNITION IN GENERAL. (§§ 3--7, pp. 7--24)
%% ============================================================
\grouphead{Introductory Considerations.\\[2pt]\normalsize On Cognition in General.}

\begin{center}\large\S~3.\end{center}
\markboth{\S~3}{\S~3}
\begin{center}Philosophy is a) cognition. ---The fundamental moments of all cognition.\end{center}

All cognizing aims at an \emph{is}, whereby something, in asserting itself in us,
is posited by us as an objective [det Objective] to which we must ascribe
validity, and which, insofar as it comes forward for us in this its validity, we
call the true---which is then to be taken entirely as it gives itself to us, that
is: shows itself to be.

Here, then, we see an opposition [Modsætning] between a self-conscious
subjectivity and an objectivity subsisting independently of it---one which,
moreover, comes forward for us not only from without but also from within. But
cognition sets in only when this opposition is so overcome that what is comes
forward for the subjectivity too exactly as it is, so that it comes to be for
us---that is, is posited by us to be---just as it is in itself; whereby we come,
in the very act of living in our own individual subsistence, at the same time to
live in the fullness of objectivity, indeed of the true.

The mediating and unifying element between the true and the subjectivity---which
latter shows itself at the same time as an individuality---is, on the side of the
first, the determining necessity with which it asserts itself, and on the side of
the subjectivity it is \emph{partly} the latter's becoming-determined and the
coming-to-consciousness that thereby sets in, which may be called a receiving
(\textit{scimus, qvia accepimus}), \emph{partly} the individual life's own doing
that answers to that becoming-determined. ---But this last is again partly one
that aims merely at a directing of the mind [Sindshenvendelse], partly one that
aims at inner subjective re-presentation and objectification. In the first we see
a subjective reaction and response, called forth by the awakening power of the
determining influence, that aims merely to make the reception deeper and
fuller---both through an accompanying devotion and through a supervening
recollection---as a care that the true may, in all its points and relations, come
to its rightful full power in us, so that in each and all we may be determined
wholly in accordance with the true's own bearing. In the subjective
re-presentation and objectification, by contrast, we see a productivity springing
from the individual life itself, which has its inner ground and mode of activity
within itself, and now, as a producing doing, brings forth and constitutes within
the subjective self-consciousness the inner presentation, in and through which the
true comes to be taken up into the subjective life. (And here, properly, it is
first said: \textit{scimus, qvia facimus}. ---For the rest I refer here to my work
\textit{Om Erkjend.\ og Granskn.}, \S\S~1--3, as well as to my \textit{Psychologie}.)
But what makes it that in our own productivity we nonetheless have, not something
merely subjective, but---by means of and through this merely subjective---the true
itself, is the power with which the true so determines and governs the subjective
life that creates and forms the representation, that it thereby at the same time
presents itself as the valid for the self-conscious beholding.

{\small\emph{Note~1.} When we here speak of an ``is,'' we see here not only an
opposition between an ``is'' and an ``is-not,'' but also an opposition between an
``is'' and a ``shows itself'' or ``seems.'' Hereupon one may at once ask whether
this ``seems'' too must not be said to be, and conversely, whether an ``is'' can
ever be anything other than a ``shows itself.'' Here it would be a matter of
asking what that ``is'' really means; and when we then observe that we can in
general have to do only with a being insofar as it asserts itself, but that now
every seeming too asserts itself, then the question is whether there is here a
differing self-assertion, and on what such difference rests---whereby a difference
between a merely momentary and subjective and a more constant and objective
self-assertion may perhaps be assumed, and within the latter there may again be
distinguished a) a merely sporadic, b) a merely microcosmic, c) a macrocosmic
validity, and only the last is to be acknowledged as the properly true one in the
last instance. But here the consideration would arise that all apparent, all
sporadic, all microcosmic self-assertion must in the end be grounded in the
macrocosmic order of things, so that here it is to be noted that all that other
self-assertion is seen in its true light---that is, in its proper being---only by
being seen in its emergence from, and reducibility to, the macrocosmic order of
things. ---But all the considerations that might thus be called forth by
reflection on that ``is'' do not belong here, but to the fundamental
considerations of speculative philosophy itself. ---This much is certain: that
throughout all our striving for cognition, the opposition between an ``is'' and a
``seems''---an ``in itself'' and a ``not in itself''---and along with it the
opposition between an objective and a subjective, so asserts itself that in the
whole course of our cognizing, as in the whole of the rest of our life bound up
with it, we cannot withdraw ourselves from it. And to this we hold ourselves
here, where we have merely to expound the nature of cognition as it gives itself
to us, and where we take without further ado the existence of a cognizing, with
all the moments therein, as given---just as we take the existence of a philosophy
as given---without entering here into the question of the proper reality of all
cognition and all philosophy, which it belongs to speculative philosophy itself
to decide. ---The concept of cognition we have here given has plainly---this is
still to be noted---validity even for one who would deny all our cognizing any
true reality; for such a one too must nonetheless have a concept of cognition that
he lays at the foundation, and it can only be the one given here; and one who
would ascribe to all our cognizing only a certain subjective validity must still
admit that cognition rests on the agreement of our representing with something
that takes on the character of an objective.\par}

{\small\emph{Note~2.} As an ``is'' forms itself, intelligence at once intervenes,
and a so-called category---or better, an a priori fundamental determination---%
asserts itself. Herewith cognition too at once comes forward as that which issues
in a judging, whereby a something (as subject) comes forward as that which is
something definite (the fundamental determination of whatness), and this
what-it-is comes forward as an other, which is nonetheless, by the judging itself,
made one with the former (as predicate). This fundamental nature of judging is
expressed in the so-called principle of identity.\par}

{\small\emph{Note~3.} The question arising here---how that productivity, having
its inner ground and mode of activity, and hence also the regulative principle for
it, within itself, nonetheless comes to present a real objective---is of
speculative nature, and finds for us its answer in the Idea of a
\textit{harmonia originaria} between individuality and totality. The determining
influence of the objective is indeed only an eliciting one, since it is the
individual life itself that, by its own activity within the individuality's circle
of consciousness, brings about that whereby the objective is to become the
cognized. But just as the whole individuality first comes forth at all through the
total course of things, and becomes, in its whole existence, as that course brings
it about, so too its whole organization---both in inner, psychic, and in outer
respects---is constituted in such a way, answering to the totality, that every
influence coming from the totality, whether it come from without or from the inner
universal, calls forth its own determinate corresponding activity in the
individual organization, or in the microcosm, and this so that it now, according
to the latter's own fundamental nature, leads precisely to such a certain
determinate production that just that determinate objective is thereby represented
by us. So that, although we constantly behold objects in our own soul, we
nonetheless behold the actual objects as they are in themselves, because our inner
representations become images of them, or counter-images to them, inasmuch as our
inner life of consciousness---according to that inner harmony of its with the
totality, when it (be it noted) is in its normal course---must each time bring
forth, out of its own inner working, precisely that nerve-action, with
corresponding representation-action, which produces just that representation which
answers to the object. The individual organism, which contains within itself the
real possibility of all possible representation-productions, is as it were
calculated to work, at every point---that is, in every respect---in such accord or
concordance with every influence coming from the side of the total, that the
former each time produces precisely what the latter aims to call forth (as when an
instrument, according to its own entire construction, renders---though in its own
individual way---precisely the tone that is struck on another). We feel ourselves,
however, in this not as related to mere counter-images, but as related to actual
objects, because we here constantly feel our organism subjected to an affection
which, coming from something other than ourselves, is what calls forth and compels
that action and production of our own soul---so that the affection of our
individual organism, or microcosm, distinguishes for us sensation from mere
representation.\par}

\begin{center}\large\S~4.\end{center}
\markboth{\S~4}{\S~4}
\begin{center}Philosophy is b) intelligent and rational cognizing. ---The fundamental
moments of cognition in their appearance as such. ---Cognition as intelligent.\end{center}

The individual life of consciousness, determined by the power of the objective,
shows itself, insofar as it comes forward over against the objective, as that
which is in itself free, carrying with it an inexhaustible possibility of acting
in other ways than the way which the determining necessity each time brings with
it. The necessity thus refers here to a freedom, and has such a freedom for its
object, since it is precisely the in-itself-free (the in-itself-free imagination
and reflection) that is bound by the necessity; and only thereby does cognition
acquire the character of a cognizing, since only thereby does it show itself as an
ascription of validity [Gyldighedstillæggelse]. [The true, too, comes forward as
the valid and as that which is cognized in its validity only to the same degree as
this free element, coming forward as such, then shows itself to be that which is
determined and bound.]

But in the first natural cognizing, which forms itself of its own accord, and in
which we on the whole comport ourselves as natural beings (cf.\ my
\textit{Psych., ny Udarb.}, \S~26), the opposition between the objective and the
subjectivity, though it may show itself in particular cases, does not come forward
in such a way that it wholly asserts itself as such. Cognition finds itself then,
despite all the mediating element lying within it, at a first stage of immediacy,
and the moments of cognition, though certainly present, are so lost in the
cognizing as a whole that they do not come forward as such. But then neither does
the true come forward for us in \emph{its truth}, nor does it come to assert
itself by its formally acknowledged validity.

In order for the true to attain such a higher governance, the individual life of
consciousness lying on the side of the subjectivity must first come wholly forth
into its freedom. And to this, too, the striving for cognition leads in its
advance, in that we---whether by some other interest, or by mere desire to know
[Videdrivt] and the higher life working itself forth through it---are moved to go
beyond the immediately given, to seek other and more: either merely a) by looking
further about us in the sphere opened to us, whereby not only the directing of the
mind, that is, attention, becomes free, but also, as expectation sets in, the
representation begins to find itself unbound; or else b) by asking after other and
more than what is given in that sphere, because through it we find ourselves
awakened to seek something such as that to which the given points---so that a
spirit of inquiry [Randsagelsesaand] asserts itself, and scrutiny [Granskning] or
investigation [Efterforskning] comes forth, and now not only the representation,
the imagination, but also the pursuit of relations and connections lying within
it, or reflection, comes forward as such. ---But the new that is now in this way
formally sought is sought at the same time as the valid, and as answering to the
investigation in such a way that that questioning can find itself satisfied;
whereby one strives after a result by which the representing and pursuing that has
become free---or has stepped forth into its freedom---can again be sufficiently
determined and bound, with a necessity that is now also sought, and which, when
found, will come to come forward as such.

But the necessity---together with the being-governed by, and being-at-home in, the
power of the objective---that sets in in this way has then a different and freer
character than that first immediate necessity. It now comes forward as necessity
of understanding and reason [Forstands- og Fornuftnødvendighed], and we ourselves
now show ourselves as beings of reason who ascribe validity because we acknowledge
according to an express admission, assent, and approval grounded in insight;
whereby the acknowledgment, or ascription of validity, too comes---stepping forth
into its freedom---to show itself as such. For although at first it may be only new
objects, of the same kind as those given in an immediate way, that are thus sought
under a merely tracking observation, they must now not only show themselves to us
as the ones we sought, but we must now also be assured that they are the right
ones, and that we now objectify them to ourselves rightly, so that nothing else
presents itself to us and is posited by us than the in-itself-valid, or that which
\emph{ought} to be posited. ---But this then naturally has such a retroactive
effect on what was earlier immediately assumed that it too can be drawn into the
same consideration, so that \emph{all} cognizing can become the object of a
question as to its validity.

Cognition will in this way become one that we ourselves confirm in ourselves, in
that we confirm ourselves in it, so that now the assurance too comes to show itself
as such, and indeed as a potentiated assurance, or as an assurance of whose
validity we are assured. ---But certainly the uncertainty lying in the questioning
may here too come forward in such a way that something doubting, dubitative, may
set in, which is however overcome when a result is attained that becomes
sufficiently satisfying.

Here, then, there emerges a properly mediated cognizing---mediated, namely, by
thinking, and resting on thinking and the satisfaction of thinking---an intelligent
cognizing, in the formation of which the fundamental demands originally lying in
the pursuit of connection and relation that strives toward cognition, that is, in
thinking itself (in the intelligence in the subject), are so in motion that it is
in accordance with their satisfaction that insight, approval, and acknowledgment
set in.

\begin{center}\large\S~5.\end{center}
\markboth{\S~5}{\S~5}
\begin{center}Continuation. What the intelligent cognizing aims at.\end{center}

Such an intelligent cognizing forms itself, in all inquiry that reaches its goal,
at least to this extent, that it is seen a) \emph{that} something definite is to
be assumed and ought to be posited. But if we ask whether also that b)
\emph{which} we cognize---the content of cognition---constitutes itself for us
through thinking, that is, through necessity of understanding and reason, we note
that this too is something to which thinking leads. We thus come to distinguish
between 1) the affection-necessity that confirms itself by itself---though through
the medium of thinking or intelligence---together with the cognizing that rests
upon it, and 2) the necessity constituted by the consideration of understanding
and reason \emph{with respect to the assumed content itself}, together with the
cognizing that rests upon this. In the first case the necessity is of that first
immediate kind that preceded inquiry, and it asserts itself in the same way, save
that we have now, through reflection, found ourselves entitled to grant it room
and power. The validity of cognition here comes forward for us on the ground of a
pursuit of relation and connection, and the acknowledgment, or ascription of
validity, has its foundation in such a pursuit; but it does not yet assert itself
in the constitution of cognition with respect to the cognized itself. [As when,
by means of tests, one assures oneself that one has observed a phenomenon
correctly.] In the second case, by contrast, that which is to be cognized is
fixed and determined by the intelligence, so that cognition even takes on the
appearance of being one that we ourselves constitute by our own thinking and
determining---although here too we are determined, since here too that which must
guide and determine this our constituting becomes the foundation. Only the last
kind of cognizing can we properly call the intelligent cognizing, and it is then
one that partly a) takes place in force of the pursuit of connection and relation,
and partly b) consists in a cognition of the connections and relations
themselves, together with what lies at the ground in or for them, which in them
makes up the inner constitutive element.

{\small\emph{Note~1.} It will easily be seen that, while under the first case we
reckon in the reflection that determines us to ascribe validity to a felt
affection-necessity, the reflection on the groundedness of this ascription of
validity---when this is now regarded as the object of our questioning---will, by
contrast, carry us over into the second of the two cases here distinguished.\par}

{\small\emph{Note~2.} We note further that there may be cases here in which a
cognition of a ``what'' can, together with its ``that,'' constitute itself for us
in an intelligent way, without our having penetrated into the object's ``how.''
Thus the validity of our faith in a governance [Styrelse], and the validity of
our living in this faith, can prove themselves to us in an intelligent way---and
this so that the very cognition of the existence of such a governance constitutes
itself for us through thinking, without an insight into that governance itself
being thereby formed.\par}

{\small\emph{Note~3.} Concerning the expression necessity of understanding and
reason, by which the necessity we may also call the intellectual is here
designated, be it noted that the double expression has come in because here one
is to think most immediately of the understanding (\textit{intellectus}), as that
which enters into the pursuit of connection and relation---the both objective and
subjective pursuit of thought that here takes place, and which leads to insight,
and so also to understanding (\textit{intelligere}); but that then, when the
acknowledgment takes place as such, so that there is posited what we see we must,
and indeed \emph{ought} to, posit, the sense for the all-valid in its all-validity
is in motion in the manner of recollection---that is, reason.\par}

But now the objective can also show itself to us through---and address us
through---the beneficently satisfying that it can have for us, and can now assert
itself in us, in its validity, by this way too. We then come to be moved by it in
such a way that we attach ourselves to it, give ourselves over to it, and love it;
whereby, once an acknowledgment of it also develops, a quite different kind of
cognition sets in than cognition in force of insight and acknowledged
groundedness---namely the one I have elsewhere called the sympathetic
[sympathetisk], and have presented as the third principal kind of cognizing, in
the threefold division: a) the immediately beholding, b) the intelligent, c) the
sympathetic cognizing---to which last belong partly the aesthetic, partly faith
and what may be placed together with it. See my \textit{Psych., ny Udarb.}, \S~33,
p.~130; cf.\ \textit{Maanedskr.\ f.\ Liter.}\ 1838, no.~6.

{\small\emph{Note~5.} But this threefold division we must here take into
consideration in yet another respect. We have come to see how cognition, from its
first principal stage---that of immediacy, or the stage of first immediate
cognizing---passes over to assume a new character, by coming forward as the
mediated, but at the same time intelligent and thereby higher, cognizing called
forth by the spirit of inquiry. But now the question must arise how this stands to
that sympathetic cognizing, which likewise shows itself as mediated in such a way
that in it too the subjectivity comes forward as such over against the objective.
We note, then, that this last is here to be reckoned with that first immediate
cognizing; for although, in comparison with the immediately beholding cognizing,
it shows itself to be more mediated---and so far to be set alongside the
intelligent---it nonetheless, just as much as the immediately beholding, forms
itself prior to all inquiry and testing, and yields something that is only by the
intelligent cognizing to be made the object of a question as to its validity and
its content. ---Over against the intelligent cognizing---which, as we shall see
more closely further on, can also be called the rational---we find another, at
which we do not arrive through a rational striving, but which forms itself prior
to it, and has its own foundation, independent of the rational (by which we mean
the rational as it comes forward in its rationality). This other cognizing we may
then call the extrarational or preterrational (because we do not have it by way of
inquiry and rendering of account, or by grounds that have presented themselves to
us as such, but obtain it in another way). To such extrarational cognizing now
belongs both that which owes its existence solely to outer sensation, together
with what is to be placed alongside it (to which the intuition of space and time
belongs), and also the whole sympathetic cognizing, which likewise forms and
maintains itself outside that rationality. All such extrarational cognizing is
also, as regards its arising, by its nature more dependent on immediate influence,
affection, than the intelligent. For the rest, we also find in the sphere of the
intelligent much that forms itself in the manner of immediacy; and it must be
noted generally that humanity, especially among the cultivated nations, now
appropriates in the manner of immediacy, or of itself---that is, without express
reflection and deliberation---much toward which one had at first only slowly
worked one's way. This, especially as regards the music of the ancients, has been
elucidated by the Reverend Pastor Bindesbøl in his review of \emph{Dr.} Bojesen's
doctoral dissertation \textit{de harmonica scientia Græcorum} (1833) in
\textit{Maanedskr.\ f.\ Liter.}\ 1837, 8th number; cf.\ the introduction to
\emph{Dr.} Bojesen's school program 1843, \textit{de tonis s.\ harmoniis
Græcorum}, p.~6.\par}

{\small\emph{Note~6.} The question that here naturally arises, when we look to
human striving as striving and to its course and fate in humanity---why, then, not
all cognizing, including the deeper and higher, gives itself to us in the simpler,
more direct, and more immediate way, as the immediacy of genius nonetheless shows
to happen in individual cases in the higher spheres---has been treated by me in my
work \textit{Om Erkj.\ og Gr.} in the Note to \S~4, pp.~32--36, with which what is
set forth in my \textit{Pathologie}, \S~23, pp.~55--56, may be compared. But I
will here return to this question, in order to give a more comprehensive answer to
it than was there given.\par}

Three things are here to be looked to: We note that we must not only a) work our
way slowly, through much investigation and exertion, toward the higher cognizing,
but that the way even b) passes through much doubt and uncertainty, indeed c)
opens such a possibility of error that this has in many ways become actual; and we
now ask why the whole state of affairs is such. To this the answer is: a) that it
thus belongs to the individual spiritual course of development on the one side,
and to the full objectification of the objective on the other, so that both the
full advance of personality on its side and the advance of the realm of the
objective on its side must bring it about so; and that, as regards b) doubt in
particular, the full self-confirmation and confirmation of the matter can take
place only in this way, point by point; and that c) the possibility of error could
be nothing other than a direct consequence of the whole state of affairs, since
this entailed that reflection and imagination had to become free and to assert
themselves in their freedom, in order that a life in intelligence and freedom
might develop and form itself---but that such a life was in turn the necessary
foundation for the development of a realm in spirit and love.

But finally it remains to be noted that, when one looks to the condition of
humanity as it actually shows itself to be, it must surely be taken into account
that the general human sinfulness, and the abnormality or aberration and
disturbance bound up with it in the whole of human life, could not but intervene
inhibitingly, disturbingly, and distortingly in the human striving for cognition,
and indeed in the whole course and manner of cognition---and that this must be
taken into account when we set actuality before our eyes and find it full of
uncertainty, doubt, and error.

\begin{center}\large\S~6.\end{center}
\markboth{\S~6}{\S~6}
\begin{center}Cognition as rational, working toward an ultimate foundation.\end{center}

Insofar as the questioning and inquiry that set in lead to the testing under which
we render account of the validity of the new becoming-determined that is to set
in, the determining shows itself as a ground, or as that whereby we, in rendering
account, find ourselves determined to assume what we assume (\textit{ratio}); and
insofar as such a determining takes place point by point in the whole of the new
cognizing now forming itself, there set in here, in a continuous way, grounds
without cease (\textit{rationes}), and cognition then shows itself as a cognition
in force of grounds---and indeed as one that asserts itself through the
consciousness, now coming forward as such, of its groundedness. The intelligent
cognizing then shows itself, with respect both to the rendering of account and to
the consideration of grounds that here takes place, as a rational one. The
in-itself grounded and valid is then assumed only insofar as it has shown itself
to us in its groundedness, and thereby in its truth.

But when grounds are sought and pursued, and a full assurance is to be sought in
groundedness, then one must go back to ultimate grounds and an ultimate
foundation---which is then thereafter to be regarded as the first foundation, and
the grounds as the first grounds for the whole cognizing, which only thereby
becomes thoroughly rational.

{\small\emph{Note.} The in-itself-valid true thereby comes---as it now steps
forward in its acknowledged validity and truth---to rule with a higher power, in
that it comes to rule as that to which we ourselves render our homage. ---But the
resisting subjective too is thereby lifted to a higher stage; for since cognition
here proceeds from inquiry and testing, and the agreement and assent that follow
from them, the subjectivity comes to move with freedom and with harmony in that
sphere of higher necessity and being-governed. ---The governed is now ensouled,
inspirited.\par}

\begin{center}\large\S~7.\end{center}
\markboth{\S~7}{\S~7}
\begin{center}Philosophy is c) scientific cognition. On science and scientific
character in general.\end{center}

If we consider more closely what grounds are, we see that---since they consist in
that which brings it about that, when something definite is given, some definite
other must also be, or be posited to be---they point back to a certain connection,
and this again to a certain fundamental connection, according to which it is that
with or from the one the other follows; so that the cognition of grounds must
become a cognition of connection, and cognition in force of grounds must become a
cognizing in force of the pursuit of connection. But thus the rational cognizing
must lead to the whole body of cognitions one has becoming a well-connected one,
in which each particular, as an account is given of it, is at the same time seen
in its being taken up into a forming totality of cognition. Hereby cognition
becomes scientific, and the whole body of cognition becomes science in the proper
sense. ---Hereby a plurality of cognitions---which, insofar as they come forward
each on its own with its determinateness, make up the detail [Detail] of the
science---orders itself together in such a way that a well-connected whole, or a
system, forms itself; in which, however, the cognizing itself, the very insight
and rationality according to which every particular is determined and the whole
system is ordered and formed, is so much the main thing that everything depends on
such a thing---which we now call sometimes the spirit of the science, sometimes
its Idea---coming to stir and live in that totality of cognitions; in which,
however, every particular must now come forth each on its own, so that the detail
of the science, by an intelligent and rational pursuit of the matter's own detail,
unfolds itself into a whole body of concepts and conceptual determinations, and
the system shows itself as a whole of conceptual cognizing answering to the
objectively given whole.

{\small\emph{Note.} To the threefold division here named as lying in science as
such---a) spirit or Idea, b) system, and c) detail---there will further be added,
when one looks to the so-called treatment of a science (both in the study that is
to lead to it and in the communication of it to others), method [Methode], which
may then be regarded partly in a subjective-logical, partly in a didactic respect.
But herewith there is no longer question of the matter itself, but only of our
endeavor of cognition and presentation. Concerning both that objective threefold
division and this subjective element, I refer for the rest to my book
\textit{Om Erkj.\ og Granskn.}, where there is also (in \S~17) discussion of the
inner organic relation of reciprocal action among Idea, system, and detail. Only
on this relation of reciprocal action will I dwell somewhat here, and I note the
following about it:\par}

We find ourselves in the midst of a great totality of objects and relations, and
this not only in the outer respect, but also with respect to the realm of
intelligence and of the higher manifestations and workings within us. Our life
forms itself under these conditions in a sporadic way, from innumerable points
that assert themselves, each point in its own way; and to this there corresponds a
whole body of observations or remarks, each of which then comes forward with its
own---actual, or at least presumed---validity. Such observations can further be
made deliberately at innumerable points, since each time that is observed and
apprehended which each time comes forward. Only when a whole body of such
observations has been gathered can that spirit of pursuit begin to work which is
to lead to the well-connected total cognition, to system and systematic insight.
Here the isolated observation of every particular in its particularity will
constantly assert its right, so that the systematic cognizing must become such as
the detailed observing brings it about; but conversely the systematic insight will
in turn have a determining, clarifying, purifying influence on the detailed
observing. The whole will thus, in and for our cognizing, be determined by the
detail, and the cognition of detail, together with the way of seeing and setting
forth the detail, be in turn determined by the cognition of the whole.

But in the whole, as the gathering-together or coming-together of the manifold
into unity, there must further reveal itself that in force of which both the
particular is taken as it is taken---being in a determinate way taken up into the
whole and ordered within it---and the whole is organized as it is thereupon
carried out. And thus we come to something more deeply underlying: the, as it
were, omnipresent fundamental cognition, which rises up beneath the whole
observing and organizing, and now constitutes both the fundamentally determining
and the fundamentally guiding, and yet at the same time itself only develops and
takes on definite shape in the course of that striving. This we now call sometimes
the Idea, sometimes the spirit in the science.

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  THE PROPAEDEUTIC CONCEPT OF PHILOSOPHY AND ITS EXPOSITION. (§§ 8--14, pp. 25--58)
%% ============================================================
\grouphead{The Propaedeutic Concept of Philosophy and Its Exposition.}

\begin{center}\large\S~8.\end{center}
\markboth{\S~8}{\S~8}
\begin{center}Philosophy aims d) at an all-comprehending fundamental cognition.
The subjective principal task.\end{center}

A whole body of cognitions can be intelligent and rational, and moreover so
entirely scientific that it leads to the development of a proper science, without
therefore being philosophy---indeed without so much as coming into contact with
philosophy, according to the prevailing concept of it, to which we here (in
accordance with what was noted in \S~1) have to hold ourselves. Of this we find
telling proofs both in the natural sciences and in mathematics. The question then
is what is peculiar to the science that is called philosophy.

We note first of all that philosophy is to lead us back to the ultimate grounds.
But here the question now arises what is meant by this, and the following is then
to be taken into consideration:

It was cognition that, under the rational striving, first and most immediately
showed itself as that which passed over into becoming one that came to take place
in force of grounds; it was cognition that was to show itself to us in its
groundedness and validity, and thereby in its truth. So far, then, the grounds in
question were grounds of knowing (\textit{rationes cognoscendi}). Here the
constant question is: what are we to assume and posit, and by what right do we
each time assume what we assume and, according to this assuming, cognize; but the
endeavor here presses toward an ultimate foundation.

By this we are now first and most immediately led to look to the first source or
origin of cognitions---by which is meant that whereby they are given us, or which
exercises the determining influence upon us. But here, in different spheres of
cognition, a different kind of immediate determinateness comes forward, which in
each sphere by itself yields something first, from which we, by pursuing the
relations that show themselves therein, are led to more; and this more, when it
now comes forward for us, either a) asserts itself by the same kind of immediate
determining as that first, so that the mediating becomes merely a leading-on, or
else b) shows itself as that which is assumed in an intelligent way on the ground
of that first, or is assumed as something seen to have to be a consequence of it,
even though this does not stand before us in the same immediate way as that first.
In the first case the testing can aim only at whether we really are determined
(affected) as it presents itself to us (whether we really do see, hear, and so on,
this and that so and thus); in the second case the testing aims at that which is
to be assumed in such a way that it must be seen to correspond to and agree with
the grounding referred to it, so that here it is a question of intelligent
satisfaction (for example, in determining the representation one is to form of the
planetary system). See above, \S~5.

But of a quite different nature is the pursuit and the question concerning the
validity of our cognizing, when the dubitative that may here arise, and the
problematic (especially under great collisions, where from one side one thing,
from another another, seems bent on asserting itself), lead to asking after the
validity of all human cognizing---indeed after the groundedness of all assurance
whatsoever---so that the reliability of all those relatively first sources, and
the validity of their authority, indeed the objective truth of everything given
through them, are made the object of a question and a consideration that then
becomes so all-comprehending that an ultimate foundation for all cognizing as
cognizing, or for all assuming and ascription of validity whatsoever, and an
all-comprehending fundamental cognition in this respect, are demanded. But one
easily sees that then there is no longer question of grounds of knowing, but of
all cognition's own groundedness and validity, so that here there is even question
of something other and more than what we may call the \textit{universam
cognoscendi rationem}, when by this, namely, we mean the fundamental nature and
constitution of all cognizing. Hereby we pass over into the domain of philosophy;
indeed both the nature of the matter itself and a glance at the whole history of
philosophy must lead us to set up the question here arising as one of philosophy's
principal tasks, which we will then call philosophy's subjective principal task.
But of this we shall speak further on, where it will appear that this task is
taken up into a larger one.

\begin{center}\large\S~9.\end{center}
\markboth{\S~9}{\S~9}
\begin{center}The objective principal task. ---Philosophy aims e) at the ultimate
grounds of existence.\end{center}

For the whole rationality in our cognizing aims at more than merely the
groundedness and validity of cognitions. It aims at the fundamental relations of
the objects themselves, and thus also at these objects' own grounds, in that we
strive that what each point in existence itself is grounded in may show itself to
us. For our whole endeavor of cognition aims at our being able, in and through
cognition, fully to appropriate and to live our way into objectivity; but then we
must enter into the connection and relations in all that is given, so that
everything therein may show itself to us, each in its significance within the
totality. And thus we come not only to call ourselves to account with respect to
our apprehension of the objective, but come also to call the objective itself to
account at every point---to call the things themselves to account---that they may,
as it were, give us a reckoning of what they themselves are grounded in, and how
it stands with their fundamental relations. ---But since we now also here strive to
go to ultimate grounds and fundamental relations, we come---since the whole of
existence asserts itself to us as forming a single whole---to strive after an
all-comprehending fundamental cognition corresponding to it: not as an aggregate
of several, but with inner unity and wholeness in itself.

And hereby we again pass into the domain of philosophy and meet its second
principal task, which we will call the objective.

\begin{center}\large\S~10.\end{center}
\markboth{\S~10}{\S~10}
\begin{center}A propaedeutic definition of philosophy.\end{center}

Of philosophy in general we may then say that it arises when the endeavor a) to go
to the ultimate grounds also aims b) at becoming all-comprehending, and this in
such a way that it would lead everything back to c) one universal foundation, so
that one all-comprehending ultimate fundamental cognition may form itself. And
herewith the first propaedeutic definition of philosophy is given.

But when we now add that philosophy, as we have seen, and as the history of
philosophy too teaches, has from the first a twofold principal task---since both
the question of the ultimate foundation and groundedness, reality and objectivity,
truth and certainty of all our cognizing, and the question of the ultimate grounds
and fundamental relations of the things themselves, come forward as chief and
fundamental questions---we come, by considering the relation of the two principal
tasks to one another, to see that the subjective task is to be subsumed under the
objective. For the ultimate ground and constitution of all our cognizing
(\textit{universa ratio cognoscendi}) must after all rest on our whole standing
with respect to the objective, our whole incorporation into and fundamental
relation to the totality that presents itself to us; so that, just as we find
ourselves only as members and parts of a great whole, so too we must find our
consciousness, and thus also our cognizing, so entering into this whole that our
whole life of consciousness, with all our cognition in its wholeness, can become
only one among the chief objects whose ultimate fundamental relations and
fundamental essence are to be treated.

Thus our definition will determine itself more closely as follows: that we regard
philosophy in general as that whose task it is, in an all-comprehending way, to
press through to the ultimate grounds and fundamental relations of the whole of
existence, or of all things, in order, in force of the fundamental cognition
thereby won, to clarify everything. ---Philosophizing is then the thinking
consideration, or pursuit of thought, in which partly a striving to reach such an
ultimate---or, as it may in another respect also be called, first---fundamental
cognition works itself forward, and partly such a fundamental cognition, insofar as
it has formed itself, further seeks to assert itself.

{\small\emph{Note~1.} When we cognize all things in their ultimate grounds, we must
also come to cognize our very cognizing---which belongs among these all things---in
its ultimate groundedness. In the total order of things in which we move, we find,
among other things, cognizing beings; the existence of these, with the whole
striving after cognition that appears in them, must then be grounded in this order
of things, and must, as regards its nature and mode of existence and as regards its
reality, rest upon it. ---This too will be seen when the subjective principal task
is pursued. Even the subjective-idealist solution of it---according to which it is
assumed that the existence presenting itself to us has no proper objectivity, but
has reality, validity, and existence only in and for our consciousness, or our
knowing of and with ourselves---nonetheless turns, in seeking to determine the
reality of our representing, on the question what we are to take the whole
objectivity given to us to be; and when it teaches us that we live only in a world
that presents itself to us---of whose reality, independently of this its
self-presentation to us, we can never speak, since we can never have to do with
anything other than what presents itself to us---it thereby nonetheless determines
both our existence and the mode of existence of what we call the objective, or call
the things over against us; and it does so in such a way that we nonetheless also
find ourselves over against them, and so must reckon ourselves too among the
totality of things; for ``thing'' here means in general only something that is
spoken of and that can be the object of an assuming and judging. But cognition we
nonetheless find again as something definite in the midst of that totality; and if
this totality is only a system of necessary representations, then our endeavor of
cognition finds itself nonetheless inserted as a member within this system of
necessary representations. But the subjective principal task's inclusion under the
objective is seen especially when we come to that solution of it which was set
forth above by way of remark (\S~3, Note~3), or even when we merely follow the path
taken there; and this principal task has therefore also, in another form, been able
and bound to arise in the historical course of philosophy's development in such a
way that the matter was entirely taken up under the pursuit of the objective
tasks---namely as the question of the relation between soul and body, and the
connected question of the relation between individuality and surrounding world;
whereby both question and solution have taken shape in such a way that the task
could and had to extend also to plants---indeed to everything that, as a member of
the world, stands under the determining influences of a surrounding world, and yet
acts out of its own peculiar nature.\par}

{\small\emph{Note~2.} Here the question arises how far a doctrine such as
Locke's---which aims only at elucidating the fundamental constitution and the
source or origin of all cognizing, as it is found to be under exact psychical
observation---can be reckoned, both in and for itself and with respect to its
effect, not to explicative philosophy alone (under which it straightforwardly
belongs), but also to philosophy proper, or speculative philosophy---irrespective,
moreover, of what worth and validity we may grant it, on which it here does not
depend. ---At bottom it aims only at elucidating what I have called the
\textit{universam cognoscendi rationem}; but it has come forward as---or at least
been made to hold as---that in which not only philosophy's whole subjective
principal task was to find its settlement and quieting, but also the foundation for
the question of the ultimate grounds and fundamental relations of all things was to
be so elucidated that in this respect one could refer to nothing but empirical
inquiry [Empirien]---so that speculative philosophy in the objective respect would
thereby be dismissed, and thus the objective principal task, as a speculative task,
decided. So far, then, such a philosophy must be regarded as one that has had to do
with both philosophy's subjective and its objective task. It has sought, in an
all-comprehending way, to set up a fundamental cognition concerning the ultimate
grounds of all cognizing, and concerning the way to the finding-out and
determination of the ultimate fundamental relations of all things.\par}

{\small\emph{Note~3.} Concerning a skepticism such as Hume's, something similar is
to be noted; but Greek skepticism especially---as that which sprang from the whole
course of the fundamental-philosophical movements and bears their stamp---must now
be acknowledged as philosophy in the proper sense, insofar, namely, as here, as
there, it is only a question of the propaedeutic concept of philosophy that holds
for the history of speculative philosophy, and insofar too as a totality of
philosophizing that neither leads nor is meant to lead to a philosophical science,
but only to the development of a philosophical way of viewing and of taking life,
can be called philosophy. ---But it is now also seen that the concept of philosophy
here given fits that skepticism entirely; for the skeptical doctrine is not that on
the empirical standpoint nothing is given that can be assumed with reliability, but
that, when one goes to the ultimate grounds, nothing can be set up as incontestably
certain---save that skepticism could not, of course, itself come forward as a
doctrine, since it had of course to admit that it too must become the object of its
own doubt, and so in the last instance dared not come forward as a proper
fundamental cognition, wherefore it was to hold only as an ultimate fundamental
quieting, aimed at bringing about the ἀταραξία of the mind.\par}

{\small\emph{Note~4.} When Kant, in order to work toward a similar quieting, sought
to establish that there can be question only of a merely subjective reality and
validity of all our cognizing, but that such a validity now also stands as firm as
can be needful for us---whereas all properly speculative striving would be an
attempt at a transcendence beyond all the capacity of our reason---he has thereby
not only treated the subjective principal task, but has also sought to settle the
objective, namely by cutting it off or dismissing it. ---Fichte, on the other hand,
in the period when he declared his Doctrine of Science [Videnskabslære] to be only
a carrying-through of Kantianism, and could declare it such, went nevertheless,
precisely by this carrying-through, so far beyond Kant that he at bottom allowed
himself that transcendence, in taking the matter speculatively; for he sought
speculatively to clarify and justify the Kantian doctrine by leading it back to a
deeper foundation.\par}

{\small\emph{Note~5.} It must here be noted generally that our determination of the
concept of philosophy leaves it undecided whether there are such ultimate grounds
of things as can be cognized by us in a rational way. Our conceptual determination
therefore holds also for those who, on religious grounds and with regard to
positive revelation, reject all philosophy that would venture into the region of
the divine; for they do so because they deny that human reason, at least in
humanity's fallen state, is capable of pressing through to the highest truths---%
which is the same as pressing through to the ultimate fundamental relations of
existence. ---It must moreover be noted generally that we have not set up
philosophy as a doctrine of the ultimate grounds of things, but only as a
fundamental cognition that forms itself under a striving to reach the ultimate
grounds---that is, a doctrine with respect to them, or concerning them.\par}

\begin{center}\large\S~11.\end{center}
\markboth{\S~11}{\S~11}
\begin{center}Philosophy steps up to what is otherwise given. ---A survey of the
sum-total of all this.\end{center}

As that which is to press through to the ultimate grounds and fundamental
relations of existence, or of all things, philosophy presupposes that existence,
with all it contains, has asserted itself in us, and has been apprehended and is
known, and now presents itself as that which philosophy steps up to. Philosophy
thus forms itself over against, and in a certain opposition to, both the whole
given existence, with what it contains, and the whole body of all that cognizing
through which everything that is now to find its ultimate clarification in
philosophy antecedently asserts itself in us. Everything that philosophy is thus
thought to step up to we call---without, however, meaning to deny that philosophy
too can form itself only by being given to us---the sum-total of the Given
[Indbegrebet af det Givne], which, moreover, is given us partly in an empirical,
partly in an a priori way, and in the latter case again partly in a sympathetic,
partly in an intelligent way.

A survey of all this Given will teach us that we have to distinguish therein the
following spheres:

1) The domain of outer experience or empirical inquiry [Empirie] (of natural
science and human history), in which cognitions form themselves along the way of
explicative pursuit, and the first foundation of cognition is observation by means
of the outer senses---whether under an immediate becoming-aware or experiencing,
or under the apprehension of what others have thus observed and let come to us
through so-called transmission (tradition); but the main thing becomes the
understanding's penetration into what is thus given, since properly empirical
science can consist only in that which is drawn from, or stands out from, the
immediately sensed or experienced and the transmitted, with abiding validity and
for abiding use.

2) The domain of independent a priori construction or operation, in which we
ourselves construct, constitute, produce the objects by an activity that is not
reproductive, but aims at realizing thoughts conceived in advance, which then form
themselves \textit{a priori}. It is then here this a priori element (the
understanding's own inner dictates) that becomes the object of study. But the a
priori element we have to do with here aims as yet only at constituting finite
objects and at reflecting on the relations that thereby come forward. ---In the
mathematical sciences, both pure mathematics and the mathematical-mechanical
sciences, we find the greatest and most telling example of such a domain; but
examples can also be found elsewhere, for instance in versification and in several
spheres of the poetic-aesthetic art, likewise in certain parts of such sciences as
do not otherwise belong here---of which syllogistics, grammatical syntax in
certain respects, and others, can furnish examples.

3) The domain of inner, psychical observation, which we shall come to call the
domain of explicative philosophy.

For the history of philosophy must bring us to distinguish two kinds of
philosophy, in that from a) the philosophy proper treated above---which we call
the speculative, but may also call the constitutive---we distinguish b) a great
body of sciences that have also been called philosophy, and go under that name,
although they are only explicative, or each pursue its own certain determinate
given in its givenness, without going back to the ultimate grounds of all things,
and thus belong under the domain of the explicative sciences, like those named
above under No.~1.

To this domain of inner psychical observation, or of explicative philosophy,
belong:

a) The domain of inner experience, psychology, in which the objects of our
observation and pursuit are our inner states of soul, with all that stirs in them,
insofar, namely, as it \textit{de facto} asserts itself (so that there is no
question of any ought, anything to be heeded as the all-valid---though certainly
the sense for the all-valid, and the human being's relation to it, as it is
\textit{de facto} given, must here be taken into account). That part of the
considerations belonging here which concerns the way in which all cognition forms
itself has especially been called ideology.

b) The domain of mere reflection-philosophy, in which the a priori element that
underlies our whole thinking and the pursuit, thereby called forth, of the
relations in the world---and so underlies empirical inquiry, as governing and
guiding the whole empirical apprehension and pursuit---is observed and explicated
in its universal validity or universality (thus also as the valid-for-us). Here,
then, in the factual, the universally valid that comes forward therein is studied.
To this belong explicative logic and ontology.

c) The domain of explicative philosophy of the Idea, in which the inner
manifestations of the infinite and eternal---insofar as they come forward as such,
and also show themselves as the higher on which it depends both what we ourselves
are to be ensouled and governed by, and by which the goods of life are rightly and
thoroughly to be assessed---become objects of explication and explicative
clarification. To this belong explicative philosophy of religion, moral
philosophy, and aesthetics.

But if we now look back to speculative philosophy, which is now to step up to all
these domains, we note that, since it too has after all asserted itself in the
world, and is thus to be found in the world, our speculative philosophizing must
also enter into relation to what is thereby given; and so we come to add still:

4) the domain of speculative philosophizing itself, insofar as it is given us in a
historical way---that is, the history of philosophy.

{\small\emph{Note~1.} It has its difficulty to distinguish sufficiently the whole
domain of mathematics from the rest of the a priori, so that an essential
difference may come to light between mathematics and explicative philosophy. As
for the difference of mathematics from speculative philosophy, that is easily seen.
Mathematics goes back only to its own ultimate grounds, its own ultimate
foundation, and is at every point able to lead everything back to it. But it does
not go back to, or seek to go back to, the ultimate grounds of all things, nor to
the question of the ultimate ground and groundedness of all cognizing. It has, in
general, nothing whatever to do with the tasks of existence itself; it moves only
in a realm of possibilities (something it has, moreover, in common with a merely
explicative formal logic), and is, of all the sciences, the one that least of all
comes into contact with the speculative. ---But why is it not reckoned under
explicative philosophy? If we consider especially---not geometry, which Kant, as
Schmidten notes in his work \textit{Om Mathematikens Væsen}, p.~30, alone seems to
have had in view where he would determine the difference between philosophy and
mathematics---but arithmetic and algebra, as the theory of functions, then we must
acknowledge that it may be asked why mathematics, insofar as it elucidates the
nature and essence of calculation, is not just as much to be called philosophical
as the explicative logic that elucidates the nature and essence of thinking;
especially since calculation, subjectively, both underlies and can come to
underlie every exact apprehension of the Given and penetrating cognition therein;
and is also, objectively, to be regarded as underlying the whole of existence
itself, just as the logical has such an objective reality and existence as
foundation for existence; so that, whether explicative logic be counted as
philosophy because what it sets out belongs to the general fundamental, or because
it explicates the inner a priori in us, there would seem to be ground for
reckoning mathematics too as belonging thereto.\par}

{\small Here, then, the following is to be noted. It is given us that mathematics
is not reckoned as philosophy, whereas simple explicative logic and ontology,
provided they are taken in the right, thorough way, are reckoned as such. This
usage, and the view expressed through it, is what we are to clarify by finding out
what it is that has made possible its formation. Thus: we are to seek out the
difference that may here show itself. And then we might first note that logic and
ontology always aim at the more fundamental, the ultimate fundamental; for what
they explicate underlies mathematics as well as every other science. It is the
direct coming-forward of an ultimate underlying that is here explicated. This
difference, however, could not be enough, since psychology is reckoned wholly to
explicative philosophy, although it does not aim wholly at anything generally
fundamental. What must become the main point is that the apprehension and
observation, and the pursuit of the observed, that take place in the cited domains
of explicative philosophy, are to be called psychical, in a sense in which that
which occurs in mathematics cannot be so called. For although in mathematics too
there is observed what the a priori in us has to tell us with respect to the
relations that occur and are to be pursued, and the tasks that are to be solved,
yet these relations themselves---indeed everything that is observed, that is, the
objects of the consideration---cannot be said to be of psychical nature (for the
possible application of mathematics in psychology is not here in question). In
logic and ontology, by contrast, there are observed partly
\emph{consciousness's own facts}, partly the \emph{a priori that constitutes the
life of consciousness itself}; whereas, when in mathematics the formula of the
parabola is found out, or any function whatever is determined and pursued, this may
indeed be said to be something we learn from the a priori in us alone, but not
something in which this a priori is itself observed, nor something that constitutes
our own existence of soul. If, on the other hand, someone were to demonstrate the
mathematical as a foundation in our reflection on many relations in the world, and
even in our life of soul, then this would indeed be explicative philosophy, but
would then belong not to mathematics but to psychology.\par}

{\small\emph{Note~2.} When the all-valid is to be studied in and through the
factual, the apprehension of the factual becomes of essential importance in and
for the doctrine of the all-valid. This holds especially with respect to
explicative philosophy of the Idea, where therefore---although it must throughout
come about through one's own observation, so that others' observations are only to
be taken as those that are to serve this end---the history of the sciences becomes
of a far different and greater importance for the sciences themselves than that
which, for example, the history of mathematics can have for mathematicians. But to
go so far as to make the whole philosophy of religion consist in the doctrine of
religion's historical course of development, as has happened in our time, is not to
take the matter rightly.\par}

{\small\emph{Note~3.} It has, furthermore, its difficulty to distinguish
sufficiently between reflection-philosophy and explicative philosophy of the Idea,
since in both the all-valid comes forward as such, so that an ought asserts itself
just as much in logic as in morals---wherefore Treschow too reckoned logic to
practical philosophy (which, in his work \textit{Om Philos.\ Natur og Dele}, he
divided into logic, moral philosophy, aesthetics, and teleology, according to the
ends: the true, the good, the beautiful, and the useful). But when one notes that
the admonition to respect the logical demands for truth's sake---that is, out of
love of truth---is of moral nature, one will see that the ought that arises in
logic is only a ``so it ought to be,'' not a ``so \emph{we} ought;'' and that
former ought is therefore rather a must. ---The proper ought expresses a mastery
as such, or aims at what is to be ruling in the one governed, when this one shows
itself as such---indeed as desiring and willing.\par}

{\small\emph{Note~4.} That sciences of mixed nature arise, we readily become aware.
We have examples of this in physics and astronomy, insofar as in both the
empirical and the mathematical interlock. In psychology, alongside what belongs to
the domain of inner experience, that which belongs to the rest of explicative
philosophy of the Idea will also come to speech. In aesthetics, under the
Idea-philosophical observation and development, one must constantly look to the
fundamental aesthetic concepts that have historically asserted themselves. Moral
philosophy and philosophy of religion cannot be carried through without
empirically given conditions---to which indeed the whole of sinfulness
belongs---coming into consideration. ---But everywhere the thorough consideration
and orientation requires that one be able to become aware of what belongs to each
domain.\par}

{\small\emph{Note~5.} We saw here that psychology takes on a mixed nature by
entering upon the domain of the Idea-philosophical. But in yet another respect it
shows itself as mixed, since outer observation---namely of others' life and
conduct, that is, something belonging to the domain of outer empirical
inquiry---must here be taken into account, together with what the inner psychical
observation, which becomes the foundation, teaches us. ---The philosophy of
language, too, must develop itself through a union of empirical study (of the
historically given languages) and psychical observation.\par}

{\small\emph{Note~6.} The philosophy of history is still to be named here as mixed,
since in it historical apprehension and pursuit must go in one with an observation
that must be of psychical nature. For the philosophy of history is to clarify the
whole life-course of humanity with respect to what is to be noted therein in the
respect of spirituality, and the development it has taken in that respect. The
fundamental moments of the life of humanity, and the course of their movements, are
then to be observed; but thus the consideration becomes psycho-biological.\par}

{\small\emph{Note~7.} We must still direct a special attention to certain sciences
that form genuine intermediate links, in that they present genuine synthetic
unities of something given in an outer, historical way, and something belonging to
the domain of explicative philosophy of the Idea. ---What is meant hereby is a)
so-called positive revelation and positive theology, and b) positive jurisprudence
and politics. In the domain of the first, something that rests essentially on
historical transmission is to be brought to live and rule in us, and to ensoul and
fill us, in the same way as that in which what belongs to the domain of philosophy
of the Idea makes itself both Idea-moving and practically effective in us; and the
former must therefore be expounded in a corresponding way. In politics, as well as
in the domain of the positive jurisprudence intrinsically belonging to it, there
constantly meet---at least as regards so-called administration and the endeavors
belonging to the improvement of legislation---the regard for what natural right and
the care of the state demand, with the regard for what, in a given state, has
partly been positively laid down and thereby set going, partly historically
developed; and both what the one and what the other demands in its consequence must
be heeded, and a solution of the tasks sought that can at once satisfy in both
respects---that is, both in the Idea-philosophical and in the empirical-historical
respect.\par}

{\small\emph{Note~8.} That, finally, explicative and speculative consideration too
can be united in one scientific development and presentation must still be
noted---of which the present philosophical propaedeutic of ours may be an
example.\par}

{\small\emph{Note~9.} Of a different nature than the sciences of mixed nature
discussed so far are the more general sciences, in which that which shows itself as
generally underlying in several different domains is separated out and considered
by itself in its inner fundamental nature, universally valid for these
domains.\par}

{\small Such a science is general biology, insofar as in it the life in plants and
animals, the inner life of soul in the human being as individual, and the life in
humanity are had in view at once and equally.\par}

\begin{center}\large\S~12.\end{center}
\markboth{\S~12}{\S~12}
\begin{center}Exposition of the given concept of philosophy: three chief points of
view herein.\end{center}

To the sum-total of all that we have called the Given, and have now set out before
us in a survey, philosophy steps up---by which, now as hitherto, we constantly mean
only the speculative, not the explicative---in order to press through to its
ultimate grounds and fundamental relations, and, in force of the cognition of
these, to clarify it from the ground up. But since it is thus called forth by the
all-comprehending spirit of inquiry that comes into motion in the scrutinizing and
grounding subjectivity---which, in order to attain a philosophy, must first
philosophize its way to it---philosophy, with respect to the fundamental relation
between the true and the subjectivity in which the true is to assert itself,
presents itself to us under three chief points of view, namely: I.~with respect to
the cognizing subjectivity and philosophy's significance for it; II.~with respect
to the coming-together into unity of the subjective and the objective; III.~purely
in and for itself, without regard to what stirs in the cognizing subjectivity.

\begin{center}\large\S~13, a.\end{center}
\markboth{\S~13a}{\S~13a}
\begin{center}Exposition of the concept of philosophy in the first chief respect.\end{center}

With respect to I.~the subjectivity that cognizes and strives after thorough
cognition, the spirit of inquiry in motion lets us see philosophy first of all as
a) the fundamental cognition that is to proceed from a calling-to-account that goes
to the ground, through which everything, each thing severally, is to establish
itself for us in the validity it is to have for us, by, as it were, exhibiting its
title to assert itself. ---But hereby b) everything, in its whole compass, shows
itself as that which presents a problem to be solved; and philosophy has therefore
been able to be represented as that which is to solve the riddle of existence,
after itself first having let such a riddle arise and take shape for us. ---But
since to this attaches the remark that the philosophical questions often take the
turn of asking after the possibility of that whose actuality no one doubts---indeed
whose actuality is perhaps formally established in advance---we see philosophy as
that which may be said c) to aim at elucidating the possibility of the actual,
which then means explicability, intelligibility, comprehensibility, where actuality
and thinking are seen to meet as those that seek a coming-together into unity and a
so-called reconciliation---which, however, becomes at the same time a leveling and
uniting of the many mutually opposing moments of existence itself, even for the
cognizing consideration. ---But since, under all such thinking, the spirit of
investigation comes into motion as a spirit of discussion or debate, d) philosophy
comes forward as that which forms itself in an all-comprehending way in force of
overcome dubitation---that is, according to an affirmation in force of overcome
negation. (Cf.\ \textit{Om Erkj.\ og Gr.}, \S~20, p.~162.)

But thus philosophy will present itself to us, on the one side, as that which is to
bring the subjectivity to e) a full coming-to-clarity with itself
(Selbstverständigung, as Jacobi called it), and on the other side as that which is
to lead to f) a full coming-to-clarity with the totality in its totality, and
thereby to full orientedness within it; from which g) an all-comprehending
satisfaction in the domain of cognition must follow.

{\small\emph{Note~1.} Kant began the whole work of his philosophy (at the beginning
of his \textit{Kritik der reinen Vernunft}) with the question: how are synthetic
propositions \textit{a priori} possible?---after he had first established the actual
existence and reality of such propositions. Later this led to the question: how is
a valid cognition possible at all? But this question, with the one lying within
it---how is an objectively valid sensation possible?---had already been in full
motion in the Cartesian school, in Malebranche and Leibniz, just as it had already
fully come forward in antiquity. ---How is speculative philosophy possible? we shall
come to ask further on. ---How is human freedom possible? how is evil possible? are
questions that naturally take precisely this form.\par}

{\small\emph{Note~2.} The nominal concept of philosophy that Fichte gives in the
philosophical journal published by him together with Niethammer (year 1797, number
1), as well as the one he set forth in his latest lectures in Berlin, may still be
brought to remembrance here. In the former he sets up philosophy as that which is to
render the whole of experience explicable by deducing it \textit{a priori}, where
then, once philosophy had completed its work, one would have to see whether what it
had constructed \textit{a priori} corresponded to experience in such a way that the
task could be called solved. Since we possess the whole of experience only in our
knowing, he did not thereby deviate from his earlier concept of philosophy,
according to which he had given philosophy the name of \textit{Wissenschaftslehre},
as that which was to make the whole of human knowing in its wholeness---as knowing
and the sum-total of knowing in general---the object of explanation. In his latest
lectures he stated that every science had its phenomenon---he used precisely this
expression---to explain. The phenomenon that philosophy had to explain was now
human knowing itself in general, or as a whole, als ein Wißthum. This phenomenon
had now first to be expounded, then explained. The former he did in the doctrine
\textit{von den Thatsachen des Bewußtseyns}, which came in the first part of his
course; ---the latter he did in the \textit{Wissenschaftslehre} proper. One easily
sees that, in similarity to the way in which Kant began his \textit{Critique}, he
could have begun with the question: How is knowing possible at all?\par}

{\small\emph{Note~3.} We here come to recall the often-mentioned but, as it seems,
in our time commonly misunderstood or unknown definition of philosophy that Wolff
gives in his \textit{Discursus præliminaris de philosophia in genere} (prefixed to
his \textit{Logik}). It runs, in \S~29, as follows: \textit{Philosophia est scientia
possibilium, qvatenus esse possunt.} Wolff says of it in the note that he first
found it in the year 1703, that in the year 1705 he communicated it to the Breslau
church and school inspector Caspar Neumann, against whose objections he defended it
in a private letter, but that he first made it public in the preface to his
\textit{Elementa Aërometriæ}. This, however, does not give the information I had
expected to find in it. The meaning of that definition, on the other hand, is
sufficiently elucidated in the cited \textit{Discursus} itself---not directly,
indeed, but in several ways indirectly. The words on which it especially depends are
the last: \textit{qvatenus esse possunt}. In a preceding paragraph (\S~6)
\textit{cognitio philosophica} is defined as being \textit{cognitio rationis eorum,
qvæ sunt vel fiunt}; and one sees that he regards all rational cognition of
that---or all insight into that---which is or happens as philosophical, whereas the
mere knowledge of the factual is set up as \textit{cognitio historica}; and the
third of the three kinds of cognition he distinguishes, the mathematical, is to aim
at the exact determination of finite things by determining their quantitative
relations. When he now, in treating more closely this difference between historical,
philosophical, and mathematical cognition, says among other things in \S~17:
\textit{Aliud est nosse factum, aliud perspicere rationem facti, aliud denique
determinare qvantitatem rerum}; when in \S~7 it reads: \textit{Differt cognitio
philosophica ab historica, haec enim in nuda facti notitia subsistit, illa vero
ulterius progressa rationem facti palam facit, ut intelligatur, cur istiusmodi qvid
fieri possit}; and when one now, in the examples he adduces, meets all sorts of
phenomena that philosophy is to explain, one sees that his meaning must be that
philosophy is to render actuality explicable, by showing how everything in it is
able to be. ---Later, after that definition of \textit{cognitio philosophica} has
been followed by the above-cited definition of philosophy itself, it reads in \S~31:
\textit{In philosophia reddende est ratio, cur possibilia actum conseqvi possint},
and immediately after: \textit{In philosophia itaqve reddenda est ratio, qvomodo ea,
qvæ fieri possunt, actu fiant}; furthermore it reads in \S~32: \textit{qvodsi ergo
plura fuerint, qvorum unum perinde possibile est ac cetera, philosophia docere
debet, cur illud potius fiat aut fieri debeat, qvam cetera}; on which occasion
philosophy is spoken of as \textit{rationes eorum, qvæ sunt vel fiunt, reddens}, and
as that which therefore \textit{docere debet, qvænam sit ratio, ut in dato aliqvo
casu unum potius fiat, qvam alterum}. In \S~37 philosophy is named as that
\textit{qvæ rationem exponit, cur ea, qvæ sunt et fiunt, esse ac fieri possint}. In
\S~46, finally, we find the following definition of a philosopher: \textit{Philosophus
est, qvi rationem reddere potest eorum, qvæ sunt, vel esse possunt}; and on this
occasion philosophy is named immediately after as that \textit{cujus est rationem
reddere, cur possibilia actum conseqvi qveant, et unum potius fiat in dato casu qvam
alterum, qvod perinde possibile erat}. ---If one now compares all this, and a good
deal more, with that definition---\textit{Philosophia est scientia possibilium,
qvatenus esse possunt}---one sees that the weight must fall on the last words, and
we see that their meaning must be to indicate what has already been mentioned: that
philosophy is to elucidate how everything that \textit{de facto} is found to be, or
to happen, in nature (to which, according to \S~21, \textit{natura rationalis}, that
is, the life of consciousness, also belongs) is able to be---or what its real ground
of possibility is. With the actual facts and factually existing things philosophy
has nothing to do, insofar as it is a question of which they are; this belongs to
historical knowledge; but to elucidate the grounds of them---that is, to show what
calls forth or brings about their realization---is philosophy's task. But that which
brings about realization is not actuality itself, but is the real ground of
possibility of it. Thus we must without doubt interpret that definition.
---Philosophy moves in the domain of the thinkable---there is thinking back and
forth---so far we have to do with \textit{possibilia}; but now it is a matter of
determining what, out of all this thinkability, rises up as that which has reality
and constitutes the underlying element in actuality; which is then not the concrete
actual itself, but is that in which the actual has its ground, and from which,
accordingly, it is to be discerned how this actual is able to be (\textit{actum
conseqvi possit}). It is thus the realm of the thinkable that meets with that of the
actual; the former is to elucidate the latter, the latter is led back to the former
and taken up into it. Actuality is here without doubt thought as that which has an
inexhaustible realm of possibilities for its foundation---a representation one often
meets in the eighteenth century.\par}

\begin{center}\large\S~13, b.\end{center}
\markboth{\S~13b}{\S~13b}
\begin{center}Exposition of the concept in the second chief respect.\end{center}

But since the demands that here come to meet all the Given out of the inner of
subjectivity, and require satisfaction, are of purely objective nature---or assert
themselves as those that come forward indeed in and through subjectivity, and so
that this subjectivity is, by their satisfaction, to come to clarity and quieting,
yet are the all-valid's own demands, in which the all-valid speaks in its
all-validity---philosophy, II.~with respect to this coming-together into unity of
the objective and the subjective, comes to show itself to us as that which is to
let us see everything a) in a complete agreement with the all-valid that stirs in
the subjectivity itself throughout the whole striving for cognition, this all-valid
then coming forward as such---or, with reason; and this in such a way that
therefore b) all cognizing, insofar as it is of intelligent nature, must come to
show itself as that which is to be deduced from reason itself, so that philosophy
cannot but aim at an all-comprehending construction of intelligent cognition
\textit{a priori}, and so far shows itself as constitutive in force of full reason.
---But since the grounding and questioning consideration thus comes to cognize
everything in its harmony with that which constitutes the innermost essence of this
considering itself, it comes at the same time to c) so full an acknowledgment of
the valid and true, and thereby also to so full an assimilation or taking-up of the
totality of what is given in existence, that this thereby d) loses its character of
givenness, so that we become at home in it, and thus that opposition is overcome
which obtained so long as the Given still stood before us merely as something given.

{\small\emph{Note~1.} Of a quite different kind from the appropriation and taking-up
we here discuss is a) that which occurs through the kind of ideal beholding that
lets us see the inner moments of the all-valid and all-constitutive step forth and
stir as the innermost underlying element in the wholly individual shapes and
life-movements of real actuality---which occurs partly in the domain of an ideal
empirical inquiry, partly in that of poetry and the rest of ideal art, wherein the
underlying element shows itself in its powerful presence in the concrete, in a
wholly individual way; furthermore also b) that which occurs in a sympathetic way
through our own interested and willing, loving and religious attachment to, and
living in and for, the totality of existence. In all this too we have something
whereby, or through which, the true can verify itself to us just as fully as
through philosophizing, only in a quite different way---so that the remark that it
can, indeed must, becomes a philosophical one, but not this verification itself. It
is to such a verification that Christ points with the words (John 7:17): ``My
doctrine is not mine, but his that sent me. If anyone will do his will, he shall
know whether the doctrine is of God, or whether I speak of myself.''\par}

{\small\emph{Note~2.} We have noted above how all our cognizing not only proceeds
from, but also throughout essentially and necessarily rests on, a \textit{Scimus,
qvia accepimus} (see \S~3). It is also of importance that we throughout feel this
givenness, and feel the corresponding sense-filled quality in all its strength, and
therein feel the power of objectivity (the Lord's power) as such. But now there
lies in this at once---and becomes inseparable from it---a \textit{Scimus, qvia
facimus}, in which the personal moment shows itself in its unity with, but also in
its relative opposition to, that universal moment. But when the moments---which,
under this united receiving and self-own determining, at the first stage still work
in immediate unity and are indeed lost in the result (the whole assuming)---come to
step forth as such, under the reflection that turns partly toward the connection and
relations of the Given, partly toward the validity of our own assuming, then we come
to a cognizing in which we find ourselves because we have ourselves confirmed
ourselves in it, and have moreover acknowledged. And now it may be said:
\textit{Scimus, qvia cognoscimus (nobiscum noscimus)}. Hereby there shows itself
most immediately a \textit{Scimus, qvia intelligimus}. But now we have noted that
it, both in this respect and with respect to the \textit{Scimus, qvia credimus} that
belongs under that \textit{Scimus, qvia accepimus}, may also be said: \textit{Scimus,
qvia experti sumus; scimus, qvia coram vidimus}---so that everywhere a consented
cognizing may show itself; indeed that the self-confirming faith especially is to be
called not only a consented but also a willed cognizing. But now the relation, both
in this respect and through philosophy, may go so far that it may be said:
\textit{Scimus, qvia sumus}---if it be otherwise permitted to use the Latin
\textit{sumus} in the same way as the corresponding expression is used and the
corresponding thought is thought in Danish and French, when the word is construed
with an object of the matter in the expressions: det er mig, \textit{c'est
moi}.\par}

{\small\emph{Note~3.} With respect to philosophy's being here acknowledged as
science in force of full reason, Hegel's definition in the first edition of his
\textit{Encyklopädie}, \S~5, may be recalled: ``Die Philosophie wird hiemit für die
Wissenschaft der Vernunft ausgegeben und zwar insofern die Vernunft ihrer selbst als
alles Seyns bewußt wird.'' But hereby it is overlooked that all being rests
essentially on a unity of the rational and the extrarational, the intellectual and
the non-intellectual, since all being on the one side presupposes the power or force
that realizes or actualizes the conceptual---or at least brings it about that it
asserts itself---and on the other side presupposes an ultimate substrate in which it
is made to assert itself, as in an other than itself.\par}

{\small What in our time has been so much discussed concerning the unity of thinking
and being there is here occasion and call to take into consideration. First, it
seems hereby, when the matter is taken subjectively, to be meant that our thinking,
as soon as it has full validity in itself, also has full objective validity (as, for
example, the astronomer's calculation, when it is entirely perfect, must also
necessarily be confirmed by actuality)---so that our thinking is to be regarded as a
real and valid sense for the real objective intellectuality. So far, then, the
subjective and the objective intelligence are meant to be one in such a way that
whoever fully has the one, \textit{eo ipso} has the other. This is entirely well
grounded. ---But now there is next also talk of a coming-together into unity of both,
namely thinking and being, in an objective respect. Such a unity can only be
understood to mean that at every point where we meet the one, the other is too, but
that yet at every point, besides the intellectual, there is also something other
than it, so that the unity is a so-called synthetic unity of two. The unity is then
not of the kind that one could say: whoever has or sees the intellectual has and
sees \textit{eo ipso} the bearing element; but one can only say: where the one is,
the other is too---so that the unity is not to be called identity, but rather, if
you will, espousal. ---He who fully has the subjective intelligence lives
\textit{eo ipso} in the fullness of the objective intelligence; but he who has the
full intelligence, or he who has the full fullness of reason, does not \textit{eo
ipso} live in the regions of sight, hearing, smell, poetry, love, religiosity, and
so on; one can only say that he lives at the same time in these regions insofar,
namely, as the fullness of reason cannot fully be had without the fullness of life
being fully had, and conversely.\par}

{\small\emph{Note~4.} Concerning the all-comprehending construction \textit{a priori}
and the constitutive character of philosophy, be it noted: It must from the first not
be overlooked that here it is only a question of this, that in philosophy, when it is
thought completed---that is, when one looks to the goal that is aimed at---the
totality of intelligent world-cognition is to constitute itself for us in a purely a
priori way, but that it cannot here be the meaning that the totality of the
world-apprehension, or of the world-picture we bear in us, should ever be able to
constitute itself purely \textit{a priori}. When we consider this, and consider that
all thorough insight must be a cognizing and discerning from the ground up, but that
the ultimate ground is the all-constitutive, which is the absolute \textit{prius},
and that the manifestation of this in us is reason in us, then the matter will show
itself clear and simple. We may say: all insight whatever, and thus also all insight
in the whole domain of experience, must both come from, and show itself as coming
from, the fundamental source of insight, reason.\par}

{\small Considering this matter more closely---which has found and finds so much
objection, but which, once all misunderstanding in this respect is removed, will
easily be seen to be only a direct consequence of the concept of philosophy---we note
the following: Philosophy is, after all, to aim at fathoming and cognizing the true
in its truth; if now this is to be found only in the inner essence of things, which
can be cognized only by understanding and reason, then philosophy must become
rational cognition. But reason does not stand to what it cognizes as, for example,
sight to the illuminated objects, so that it should merely straightforwardly take up
and present it to itself as it happens to be found. Reason can cognize nothing except
by owning it as originally agreeing and harmonizing with itself---which follows from
the nature of that cognizing with which insight and proper concept, and so also
testing, are to be bound up. But if now that true essence of things, cognizable only
by reason in this way, is at the same time that which brings it about that these
things are as they are, and thus constitutes the ultimate ground of these
things---and if therefore neither it itself nor they can be cognized in their truth
without its being seen as their ground, and they as grounded in it, it as necessarily
bringing them about, and they as following from it---then the true cognition requires
not only to cognize the essence of things, which is the object of the highest
fundamental Idea, but also to see how everything else that is grounded therein is to
be derived from it, whereby all of it is first seen in its true significance and in
its relative essence. But to see how it can be derived from the fundamental Idea is
the same as that we ourselves in representation derive it therefrom, and, insofar as
this happens in an all-comprehending way, derive from it, and accordingly set up, a
whole system of cognitions in which everything is seen---but at the same time also
set---in its place. But this is precisely to construct \textit{a priori}. And if now
that which is thus to be cognized in its essence by being cognized in its connection
with the highest true is precisely experience and, in general, all the Given, then it
is precisely the whole content of this experience, the inner underlying element of
this Given in its whole extent, that is to be seen as a consequence of the
fundamental Idea---thus derived from it, and thus also re-constructed \textit{a
priori}.\par}

{\small If we must in general concede that cognition is that through which the whole
is fully to come to the individual's consciousness, by, as it were, being reborn in
it as in an image, whereby the whole then comes to live again in the individual---and
if we must now further acknowledge that the true first, or \textit{prius}, in things
is that which is the object of the highest fundamental Idea, the eternal---how could
we then do otherwise than acknowledge that this must in the same way show itself as
the all-grounding first in the whole of our cognition, as it is in itself, if our
cognizing is to re-express that whole of things; so that, just as by that objective
first all things are constituted, so too by the fundamental working in which this
first presents itself in our consciousness, that inner image of all things must at
every point be constituted. But if we are now, in and by this, to come to see and
cognize truth in its truth, then the image too must be formed in us under and
according to our own discerning, and a determining guided by discerning---and so we
must ourselves, in force of the fundamental Idea, execute that image of the whole in
our consciousness; but that is to construct it in force of the fundamental Idea,
which must then show itself as the all-determining \textit{prius} in our whole of
cognition. Only insofar as this happens can we say that this whole of our cognition
really and in truth corresponds to that whole of things, in which the things do not
stand naked and bare, each for itself, but all are only insofar as they hang together
with the all-determining first, and this is present in them.\par}

{\small If, however, we now hear much opposition to the proposition that philosophy,
according to its concept, aims at a reproductive a priori construction of the given
whole of experience---indeed if we even read in Steffens: ``Naturwissenschaft
\textit{a priori} ist der Tod aller Naturphilosophie''---we are naturally brought to
the thought that here, by the expression \textit{a priori}, something other is meant
than what we think of by it. And to this thought we come all the more since we see
that it is philosophy of nature whose death all natural science \textit{a priori} is
declared to be. We may then either suppose that by that natural science \textit{a
priori} is meant merely one that would set out from certain so-called facts of
consciousness alone, or from certain general concepts of the understanding and
general propositions about the subjective constitution of all cognitive activity,
abstracted merely from partly logical, partly psychological observation, or in other
such ways would set out from something subjective that the human being brings with
him and would assert as necessary imaginations---something of which we find eminent
examples not only in Fichte, but from which we cannot even acquit Kant's
\textit{Metaphysische Anfangsgründe}. Or else we may suppose that that proposition,
and other such things heard from the most excellent philosophers, is merely the
expression of the most well-grounded recognition that we, in our time, are still far
too far from having penetrated so deeply into the study of nature that an endeavor to
construct for oneself a system of nature \textit{a priori}, merely according to the
fundamental Ideas to which we have already come, could be anything but highly
precarious.\par}

{\small As certain as it is that in morals---indeed even in psychology---one should at
once go straight to the highest Idea and bring everything into connection with it (for
it is the one whose power is to ensoul the human being, and with which he is
everywhere to seek to feel and bring his life into concord), just as certain is it
that in the study of nature, by contrast, one should hold oneself close to the living
contemplation of given nature itself, and only, guided by it, gradually prepare the
possibility of that attachment to, and firm and definite descent from, and deduction
from, the highest Idea. But a possible construction \textit{a priori} remains the
goal. It is here entirely as in astronomy, in which we have a telling example of what
construction \textit{a priori} means, and how a thorough striving after cognition can
and must lead to such a construction. As regards our planetary system, astronomy has
come so far that, out of its \textit{prius}, its fundamental Idea, supported on certain
fundamental data, it can \textit{a priori} construct the totality of all the planetary
movements and of the relations that thereby set in. The same the physician strives
after in his spheres, and precisely the same is it, too, that philosophy strives after
with respect to the totality of all world-relations.\par}

{\small But when we now thus acknowledge this to be the highest aim of the
philosophical striving, the question naturally arises how we now come to the
all-comprehending fundamental Idea that is to be the foundation for the construction
\textit{a priori}, through which everything that is to have validity for us is to
constitute itself for us in its validity. And here it is to be noted that this
fundamental Idea, this foundation, certainly cannot constitute itself for us in the
purely a priori way that we find, for example, in mathematics. But of this, as well
as of the question whether it is finally to be able to become evident to us in its
validity by itself alone, we shall speak further on in \S\S~15 and 16.\par}

\begin{center}\large\S~14.\end{center}
\markboth{\S~14}{\S~14}
\begin{center}Exposition of the concept in the third chief respect.\end{center}

Finally, we consider philosophy III.~purely objectively, looking to its aim in and
for itself, without regard to the cognizing subjectivity and what stirs in it. Then
we see it aim at pursuing everything in such a way that this everything must show
itself a) in its ultimate grounds and fundamental relations, indeed in its ultimate
groundedness; but thus it shows itself b) in its truth or in its true essence, so
that there is seen c) what reality and significance are properly to be ascribed to
it, and thus d) what it is in its innermost essence. But this brings it about that
e) every particular is seen in its being taken up into the whole, so that the
manifold is cognized in its inner unity, and the whole comes forward for us in its
wholeness---whereby the fundamental philosophical striving shows itself as an
all-comprehending striving for unity. But under such striving to see every
particular ordered within the totality of existence, philosophy comes to strive
after f) an all-comprehending world-view [Verdensanskuelse], in which all things
can show themselves to us in their ultimate world-relation and in their
world-significance; and philosophy thus aims at becoming a speculative cosmology.

But this striving must necessarily lead further. For the whole world itself can
show itself to us in its right light only through this: that everything finite is
seen g) in its being led back to, and taken up into, the realm of the
all-constitutive, the infinite and eternal---whereby, conversely, this realm must
be seen in its stepping-out into the finite, as the real present therein. ---But
this must lead us further to h) the central point in the whole totality, and thereby
to the Godhead; and philosophy must then seek its ultimate foundation in a
speculative theology, and must then become a unity of speculative theology and
speculative cosmology.

{\small\emph{Note~1.} The question naturally arising here---how a cosmology, and thus
the whole of philosophy itself, is possible for us---we shall come to answer further
on (\S~15, Note~2).\par}

{\small\emph{Note~2.} Insofar as the all-constitutive shows itself to us as the
absolutely beginningless and timeless, in whose whole ideal fullness no becoming,
change, or transformation is even so much as thinkable---since all such already
presupposes the ideal, and consists in certain realizations of something
ideal---we call the all-constitutive the eternal. Insofar, by contrast, as it is
seen partly in its inner inexhaustibility as entailing a countless multitude of
determinations, partly in its relation to that which is subject to time, as that
which rules and lives therein, we call it the infinite. ---Finite, by contrast, we
call everything that, coming forward, each as something on its own, has its
determinate being with its determinate determinations, so that of it it can be
said: ``So and so it is; so and so, on the other hand, it is not''---whereas the
infinite is the inexhaustible fullness of all possibilities. But now every
particular, and indeed every finite thing, is only a moment within this fullness.
If we behold it rightly, we behold in it the fullness of the infinite in one of the
points or moments in which it is and stirs, as in one of the particulars belonging
to its total organization, or to its realm---which particular is, however, then
seen in its emergence from, and in its being taken up into, the totality of this
realm, whose countless multitude of particulars, in their organic coming-together
into unity, is precisely the fullness in which the infinite is and stirs. ---From
this one will understand the following words of Steffens in his \textit{Grundzüge
der philos.\ Naturwiss.}, p.~3: ``Es giebt für das wahre Erkennen kein endliches
Ding; das Endliche wird vielmehr nothwendig in die ewige Form und sofort in das
ewige Wesen gesetzt, in welcher es aber mit diesem eins ist.''\par}

{\small\emph{Note~3.} If we now look back, we readily become aware how the many
moments in the concept of philosophy here adduced, and the sides from which to view
it, teach us that philosophy can be defined by means of several different concepts
made the foundation, and that thus a whole \emph{cycle of definitions of philosophy}
can open itself for us---among which some will straightforwardly be reckoned to the
propaedeutic ones, while others will come forward as those into which our own
philosophical fundamental view enters. Cf.\ above, \S~1, Note~1, p.~3.\par}

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  THE FOUNDATION OF PHILOSOPHY AND ITS WORK. (§§ 15--19, pp. 59--80)
%% ============================================================
\grouphead{The Foundation of Philosophy and Its Work.}

\begin{center}\large\S~15.\end{center}
\markboth{\S~15}{\S~15}
\begin{center}The philosophical fundamental Idea.\end{center}

We are now to speak of how philosophy comes about. But here the question first
naturally arises whether, in investigating the ultimate grounds and fundamental
relations of things, there is any other way to go than a simple, continued, and
all-comprehending explication and pursuit of all the Given on all sides and in all
directions, until one thus arrives at the ultimate underlying element; or whether
the matter stands otherwise with those first or ultimate grounds of things that we
have named---and whether they are to be scrutinized by a quite different way---than
that which obtains with all that kind of grounds and fundamental relations which we
will now here, to distinguish them from the former, call the nearer grounds, which
in every science's domain are to be pursued up to their own peculiar first or
ultimate foundation. To this the answer is that, just as speculative philosophy has
come forward and asserted itself in humanity as a thing on its own, alongside all
other scientific striving, so too it has its own foundation and origin in the human
spirit; and the question of the ultimate grounds of things thus stands otherwise
than the investigation of what we have here, by contrast, called the nearer grounds.
---In the consideration of the many given things, there comes forward, out of the
thinking spirit's own innermost (out of reason), something that steps up to all that
Given, and now constitutes the a priori foundation for the whole of speculative
philosophizing, and constitutes that which inmostly stirs within it, and lets us
see the whole Given in the light in which it shows itself to us when the question of
the proper essence and reality of things comes into motion. And it is well to note
that this a priori speculative element, out of its foundation and its fundamental
movement, steps up to the rest of the a priori (the moral, the religious, and more)
just as much as it steps up to what is given in the domain of sensibility. This a
priori speculative element, which calls forth and drives the whole of speculative
philosophizing, and which stirs within it at every point as the immanent fundamental
mover, we call the philosophical Idea, the philosophical fundamental Idea, which is
now to enter into all the Given, mediate it with itself, assimilate it to itself,
and so take it up into itself. Without something of this kind, the whole of thinking
would not take the speculative turn at all---which presupposes that we go beyond the
whole chain of particulars, and must then proceed from its own peculiar impulse.

We see from this that, although in speculative philosophy a pure rational cognizing
is striven after, it can nevertheless not form itself except under the
Idea-movement's constant conferring [Conferents] with all the Given in its
givenness---just as it is, after all, constantly this Given that is to be cognized in
its innermost essence, so that the eye must be constantly directed to it, and the
sought-for all-clarification must essentially rest on a satisfaction of the demands
coming from this side.

{\small\emph{Note~1.} The speculative Idea is as the moral fundamental Idea, and we
can clarify for ourselves the former Idea's relation to the manifold by looking to
the latter. By the latter we mean nothing definite that could be uttered in a
definite proposition, or pointed out in some definite particular, but mean by it
that which brings it about that in all individual moral cases we judge as we judge,
and discern what we discern, and are moved by that by which we are moved. This moral
Idea therefore exists only in the totality of all moral insights and judgments, as
the foundation present at every point for the whole of moral cognition---which,
however, can attain its fullness of content only under the Idea's constant conferring
with the sum-total of that part of human life and human relations with which morals
have to do. For it is at first only the still wholly indeterminate and so far
contentless Idea of that which so ought to rule in us that we must strive to become
governed and ensouled, penetrated and filled by it. But when it is now asked what it
is by which we are thus to be governed, and how in all respects, then it will depend
on our---next after having investigated and pursued what lies in what has now been
stated---entering with it into the whole of our given life.\par}

{\small In precisely this same way it now stands also with the philosophical
fundamental Idea, which is at first nothing other than the still indeterminate and so
far contentless Idea of an all-constitutive All-being.\par}

{\small\emph{Note~2.} We have here come to the point where we can answer for
ourselves the question raised above (Note~1 to \S~14) concerning the possibility of
philosophy, with respect to its aiming at an all-comprehending world-view, while we
can yet survey only so small a part of the objectively given world. ---That which
constitutes the total existence in its innermost also constitutes our rational
existence in its innermost, and now, in the fullness of its totality, asserts itself
in us from within through the working of the philosophical fundamental Idea; and
hereby speculative cosmology becomes possible.\par}

{\small\emph{Note~3.} We must here look to the sporadic in existence, to which I have
elsewhere drawn attention (first in \textit{Philos.\ Arch.\ og Repert.}, p.~256 ff.,
later in my \textit{Psychologie, ny Udarbeidelse}, \S~28, and in several places). It
is a consequence of the fact that the organizing working and creating in
existence---regarded as a total organization---at every stage in the course of
development steps out in all its moments at once, from countless substrate- and
subsistence-points, so that thus, as all moments assert themselves each for itself,
the whole manifold that is to form the organization's fullness can step out in the
relative separateness and difference of its content, in order thereupon to come
together into the unity that becomes fullness only as the unity of what is thus
separated and relatively self-subsisting. The sporadic is thus the organization on
the way to constituting itself, on which way it essentially belongs to it to pass
through the stage of the sporadic---that is, through the debate of the moments of
existence. ---Existence forms itself through a life-dialectic, like the dialectic
through which (see further on) the life in our consciousness, and finally philosophy
especially, in turn forms itself.\par}

{\small In the life of consciousness, then, the philosophical tendency thus comes to
step forth as a tendency on its own within the whole system of tendencies; and it is
not at first, and cannot be, the one that stirs inmostly in all our cognizing, as it
strives by its essence to be in a scientific respect, but must now enter into
conferring with all the rest. ---The world does not at first open itself to us solely
from its innermost foundation, so as now, out of this, to build itself up in us by
the power of this its central point alone; on the contrary, it opens itself to us
from the many points of its fullness at once, and just as the religious, the moral,
the poetic tendency now in its own way, so too the fundamental-philosophical must in
its own way first win for itself its all-comprehending governance. The central point
thus itself becomes only a moment within the totality during the latter's working
toward constituting itself in us; indeed it is itself in itself divided into a system
of central tendencies, which also have among themselves to enter into a conferring
with one another. [What we call forces in nature are in general the world-tendencies,
or rather the tendencies of the all-constitutive, in their separateness, since
\emph{They} too form an organization with fullness, in which each moment has its
relative subsistence on its own; ---an infinity of wills here comes together, or is
to come together, into one fully organized fundamental willing and fundamental
governance, stirring thus, in the manner of the manifold, in its unity, in whose
realm, however, all those more particular fundamental tendencies continue to come
forward as such in their separateness; for only thus is the realm an organization
with life-fullness.]\par}

\begin{center}\large\S~16.\end{center}
\markboth{\S~16}{\S~16}
\begin{center}Wherein the philosophical fundamental Idea has its foundation.\end{center}

But now the question naturally arises whether the philosophical fundamental Idea is
itself to be taken as something given, from which one has to set out, or whether it
too is to prove itself to us in its validity, like everything else, and thereby
constitute itself as fundamental Idea. To this we must answer that it certainly is
to do so, and that it must therefore itself be subjected to the all-comprehending
discussion and testing that it itself sets in motion, in that, in the
carrying-through of this discussion, it must itself affirm itself to us in its
validity. But this it can do only by showing itself as that which, so to speak,
holds good, or is able to accomplish what it is to accomplish, by really showing us
the totality of all the Given in such a light that this everything thereby becomes
clear and intelligible to us in its ultimate grounds, so that the Idea both
satisfies or contents itself for itself, and at the same time contents us in the
wholeness of our existence by contenting us with respect to all the moments of our
existence and of the whole world-existence.

{\small\emph{Note~1.} But---it may be said---since this whole discussion and debate
thus setting in can be conducted only under that fundamental Idea's own influence
and presidency, and moreover under that of a fundamental view gradually given along
with it, does not this fundamental Idea then come to step forth in force of an
originally given, and with the character of such? To this the answer is: certainly
it is so, and therefore all philosophizing rests on an original philosophical sense
and impulse, from whose side all the rest is met; but in the same way every moment
in life and existence gets its voice in the all-comprehending discussion, and
nothing can be turned away that would assert itself as a co-belonging moment. But
now everything depends on the unity-Idea's being here so satisfied that thereby that
in force of which this unity is constituted---or the fundamental cognition stirring
in this unity, as the appearance of the fundamental being subsisting and stirring
therein, in and for our consciousness---really comes to show itself as taking up all
these moments into itself. Thus the philosophical fundamental Idea itself (the ruler
itself, or that which is to work its way forward to becoming it) comes into a
reciprocal struggle and debate with the whole manifold, or with the whole so-called
detail.\par}

{\small\emph{Note~2.} By what is here adduced we are led back to the relation
between, on the one side, philosophy, and on the other side---partly empirical
inquiry, together with its transfigured return in the ideal counter-images of it in
poetry and art---partly faith, that is, religion, together with what further
presents itself from the side of sympathetic cognition. It is thus the opposition
between the rational and the extrarational---or rather between the wholly rational
cognizing and that which rests on an extrarational foundation (though also coming
forward in its partial rationality)---that we here return to (cf.\ \S~5, Note~5,
p.~18).\par}

We note, namely, that, although speculative philosophizing must not only look to all
that is thereby given, but must also heed it and respect it and know itself in
agreement with it, it is nevertheless not to \emph{build} on experience or on
faith---indeed it cannot even do so, since it is only philosophy itself that sets
both the domains of experience and of faith in the light in which they would have to
be seen if they were to be able to form such a foundation.

For the closer clarification of this matter---especially as regards the question of
philosophy's relation to faith and religion, and the opinion that philosophy must
ultimately \emph{build} on faith---the following is here noted:

For religion itself, and for what most essentially matters therein, religiosity,
faith lies at the foundation, just as at the foundation of poetry there lies
something ingeniously poetic, and at the foundation of music there lie the ear and
the rest of the musical sense, together with the feeling for beauty and poetic
feeling. Concepts of religion, finally, can, by intellectual observation and
explicative pursuit of what is given in and by religion and religiosity, in its
connection and consequence, be developed and formed. An explicative philosophy of
religion can in this way come about, as one that has faith for its foundation.

As regards, by contrast, the relation between faith and speculative philosophy, on
the one side faith is not to seek its foundation in speculative-philosophical
consideration, as if philosophy were its basis and its whole reality thus depended
on the outcome of the philosophical discussions, but is to hold fast to its own
foundation, independent of this. But on the other side philosophy neither is to nor
can \emph{build} on faith, as if faith were its proper basis. On the contrary, it
belongs to the work of philosophy proper that it, alongside much else, is precisely
to make the question of faith's reality, and of the reality of faith's whole
content, the object of its consideration and discussion; and it therefore---though
it must respect faith in its givenness, like everything else given, and must let it
have its incisive co-working and voice in the total discussion---cannot \emph{build}
on faith, since it must \emph{build} on that which is to set both faith and all the
rest in the right light. In philosophy, faith with its whole content is to show
itself to us in its innermost and ultimate rationality or intelligence, or in its
innermost Idea-truth; but the source of all such rationality in our cognizing that
goes to the last is speculation itself in its wholeness. [For the rest, it goes
without saying that our beholding of faith in its Idea-truth is not the actual faith
itself; so far, then, faith comes to form the abiding opposition to philosophy, and
that in a collateral way; and already in the domain of cognition the full attainment
in this respect comes therefore to rest on a synthetic unity of faith and
philosophy. In a similar way it stands with philosophy's relation to the moral, the
poetic, and the artistic, as well as with philosophy's relation to empirical
inquiry, in whose domain the full cognizing must rest on a synthetic unity of
philosophy and empirical inquiry, with equal right and validity from both sides. But
of this we shall speak more closely further on, in \S~21.]

\begin{center}\large\S~17.\end{center}
\markboth{\S~17}{\S~17}
\begin{center}How philosophy comes about. ---a) Explication in speculative tendency.\end{center}

If we now ask how philosophy comes about, the answer will be that the most natural
way will be this: that A) one sets out from considerations of all the Given, as it
can be known by way of explication, but so that the chief and fundamental points
therein are made objects of a speculative consideration that aims at leading them
back to one innermost all-comprehending fundamental cognition, by heeding the
intimations they may offer---which intimations, however, can of course present
themselves as such only to the very philosophical fundamental Idea that comes to the
fore under such consideration, but as yet emerges only sporadically or point by
point. That which thus forms itself we may call an explication in speculative
tendency.

But now, under such considerations, we find something that we meet again everywhere
in all the sciences as the common foundation for them all. This is partly the
logical, partly the ontological. ---Explicative logic and explicative ontology must
therefore form the most fundamental part of that explication in speculative
tendency, since they aim at elucidating and pursuing the most fundamental and
moreover most general fundamental concepts in human knowing, which we may also
regard as the inner indices in us of the first or most primitive world-relations.
But the deepest fundamental relations open themselves to us only when, in speculative
tendency, we explicate to ourselves what belongs to the domain of philosophy of the
Idea.

\begin{center}\large\S~18.\end{center}
\markboth{\S~18}{\S~18}
\begin{center}b) Speculation and dialectic in essential unity.\end{center}

The transition to speculative philosophy proper, however, first occurs when B),
under the all-comprehending conferring of all with all, an all-comprehending
discussion and debate of all against all sets in---under which, as every particular
shows itself to us in the relative validity to be ascribed to it, and with the
significance and corresponding place it comes to occupy in the totality of our
cognizing, the all-comprehending fundamental Idea, and moreover the cognition of God
and world-view, so clears itself for us that it can now become the foundation for
the sought-for all-comprehending clarification. ---Philosophy shows itself here,
then, as that which, both in its formation and in its final execution, rests on an
\emph{essential unity of speculation and dialectic}.

By speculation we understand the deeper grounding and spiritual beholding by which
the philosophical fundamental Idea that comes to the fore keeps itself constantly
present to the soul's eye, so that it constantly and at every point asserts itself,
and therein develops and forms itself, as was elucidated above (\S~15, Note~4,
pp.~60--61) by the example of the moral fundamental Idea. Speculation thus corresponds
to what we otherwise call the spirit of the science. ---By dialectic, by contrast, we
understand the philosophical pursuit of thought by which, under the influence and
dominion of that speculative fundamental cognition, every object, every point and
moment, is discussed and debated so that it may show itself what each object, each
point and moment, is to count as being---what significance and essentiality, what
reality and non-reality, are to be ascribed to it. Dialectic may here be said to be
speculation itself, setting itself in relation to the manifold and determining it and
forming it after itself; (thus I have also called dialectic in my work \textit{Om
Erkj.\ og Gr.}, \S~14, to which, as well as to \S~13 of the same, reference is here
made throughout). Every particular thereby comes, both in the sphere of actuality and
in that of concepts, to step forth in its relativity, pointing out from itself to
another on which it rests, on which it depends, and in which it is only a
moment---constantly at the same time itself stepping forth with different validity,
and thus also non-validity, according to the relations in which it comes to have or to
want to assert itself.

{\small\emph{Note~1.} I refer further, as regards speculation especially, to my
treatise on the intellectual intuition in \textit{Philos.\ Arch.\ og Repert.},
p.~125 ff.; concerning dialectic see the same, p.~387 ff., as well as my
\textit{Logik}, \S~90.\par}

{\small\emph{Note~2.} Here the following words of Steffens in his \textit{Grundzüge
der philos.\ Naturwiss.}, p.~36, deserve to be recalled: ``Wie das Wissen speculativ
wird durch die Anschauung der Universalität, so wird die Speculation wissenschaftlich
durch die Anschauung der Individualität.''\par}

{\small\emph{Note~3.} The speculative purely alone, separated from the
dialectical---indeed from the rational in general, even in its explicative form---we
have in a peculiar way in the mystical (see the just-mentioned treatise, p.~235).
---The dialectical moment alone, without speculative content, we have, pure and bare,
in sophistry and scholasticism; with a definite tendency, on the other hand---and thus
with a certain content, only no scientific one, but merely a practical---we have it in
the skepsis of the ancients. ---In the case of Parmenides one might perhaps also have
occasion to ask whether in him the speculative does not come forward without the
dialectical.\par}

{\small\emph{Note~4.} It is still to be noted that philosophical content, in rational
appearance, as deep-seeing cognition with respect to the innermost world-relations in
the respect of spirituality, can come forward without scientific form in poetry, in
intuition-filled utterances of the speculative fundamental Idea; for this too can, in
a certain immediate, ingenious way, break through in such utterances, according to the
awakened sense for it, without having gone the way of the pursuit of thought and
discussion---though not without a certain eye for the dialectic of life. Already Plato
noted these and other ingenious workings of the higher life working itself forth as
those that led to a certain rational cognition, without having formed themselves in
the proper manner of rationality. A certain original sympathy between the macrocosm
and the microcosm, and a sympathetic being-moved springing from it by certain
outbreaks of the fullness of the Idea, according to reason's immediate glance, here
comes to light in the domain of intelligent cognition---that is, outside the domain I
have called (cf.\ above, \S~5, Note~4) the domain of sympathetic or devotion-filled
cognition.\par}

{\small\emph{Note~5.} Through the all-comprehending discussion---both in the positive
direction, insofar as it determines for each concept its relative significance and
moreover its being and, so to speak, its role in the whole system, and in the negative
direction, insofar as relativity, and thus also the transitions, are therein seen and
pursued---dialectic shows itself as that in which stringency and fluidity come forward
in synthetic union. Every concept, while it holds fast in its own peculiar meaning and
in the consequence of the same, is nevertheless kept constantly open for a certain
determinability, so that it can constantly turn differently, and, in its application,
constantly---while other things are determined in force of it---itself be subjected
partly to testing, partly to modification; since nothing may be set up as fixed prior
to the rest, but only the going-through and determination of every particular gives
each particular its fixity and significance. ---An opposite way of taking philosophy,
more resembling the mathematical, especially the geometrical, has been called
dogmatism. Against such a way Kant set up his criticism, demanding an all-comprehending
testing of the cognitive activity itself in advance. ---Dogmatists are otherwise
opposed to skeptics. But dogmatism finds its proper opposite in dialectic.\par}

{\small\emph{Note~6.} A dialectic of life was named above (in Note~4). By this is meant
the debate of the moments of life, under the struggle of life, through which the valid
gradually comes to show itself in its validity, the relative in its relativity, the
unreal in its unreality, and whereby a school of life thus forms itself for us as well,
in that it gradually clears itself out for the thoughtful and level-minded where the
genuine hold is in all respects to be found, and where not. Such a dialectic of life
can then also show itself as a Dikaiosyne of life, a Nemesis of life.\par}

\begin{center}\large\S~19.\end{center}
\markboth{\S~19}{\S~19}
\begin{center}Philosophy's stepping-out into c) a philosophical system.\end{center}

Through the carrying-through of speculative dialectic---under which philosophy shows
itself as that which rests equally on content and on form---the way is opened C) to
develop philosophical cognizing into a systematic cognizing, which can then proceed
to present itself in a philosophical system, but must find its complement in its
opposite, the free judicious [element], and must become of organic nature, in that
therein, throughout, every particular must equally be confirmed by, and become
confirming for, the rest, so that the cognition of each particular rests in the
total intuition, while this in turn must rest in the merging of every particular
into it.

{\small\emph{Note~1.} Of such organic nature is also all explicative science in its
domain, insofar, namely, as it will not stop at being a merely ordered knowledge of
the factually given (under which the natural laws too, insofar as they are merely
found out, not explained, are to be reckoned), but will proceed to a certain
fundamental cognition that is to give insight into what is thus given. Mere
explication, by contrast, is not of such organic nature; nor is mathematics, since
the particular in it can stand perfectly fixed before the rest has merged for
cognition, so that the cognition of the particular does not rest in the total
intuition. But something dialectical may indeed come to light in mathematics, when
the mutual dependence of the determinations on one another shows itself, and each
moment is seen to play its relative role in a total function; and in this way
something organic may also show itself within the sphere of a certain principal
function. Schmidten, indeed, in his interesting work \textit{Om Mathematikens
Væsen}, p.~31, would have mathematics wholly possess that organic character; but I
do not find that he elucidates this claim of his, let alone justifies it.\par}

{\small\emph{Note~2.} From philosophy's organic nature it is now also seen why it
cannot come forward with the stringency and absolute certainty of mathematics---which
is connected with philosophy's higher character. And here it will at the same time
be seen how the manifold disputes among philosophers---concerning even the very first
foundation---are to be regarded: disputes such as one has had, moreover, in just as
great a number in the domain of religion. The all-comprehending debate of all against
all---which is to take place as that through which the fundamental view is first to
clear itself for us, and which therefore, in every individual in whom philosophy is
to form itself from the ground up, must take place and be fought through anew---we
meet on a larger scale in those disputes, and thus see many great thinkers among the
philosophers as representatives of one or another principal moment, which now comes
forward through them with one-sided force and in a one-sided direction. Nowhere is it
more to be seen than in the sphere of the philosophical debates and of the religious
movements that our spiritual life as a whole is still in its first process of
formation and constitution, and therefore still only seeks that firm, incontestable
foundation and that sure way and direction in its course which Kant, in the preface
to his \textit{Kritik der reinen Vernunft}---the work in which he hoped to establish
such a foundation and such a way for philosophy---notes that mathematics had already
attained when Thales, or whoever else it may have been, found out the law for the
isosceles triangle; just as Kant also holds that formal logic must be assumed to have
attained its fixity already before Aristotle, and that empirical natural science,
especially by taking the way of experiment, has come to reach the sure course of a
firm science. ---It must of course be otherwise where the full validity of the
individual cognition essentially rests on the all-comprehending total's standing
fixed for our intuition with respect to all its principal points or principal
moments, and where therefore the rule \textit{Totum est parte sua prius} becomes a
rule for insight as well. ---Many have believed that fixity with respect to the
meaning ascribed to words must contribute much to agreement; but it is really a
fixity with respect to the concepts that matters; yet behind every conceptual
determination---when it is not to be merely nominal and thus leave the matter itself
unclarified---there lies, in the domain of philosophizing, in the deeper thinker, the
whole fundamental view of the world-relations; there lies even a fundamental
tendency, as a fundamental willing. ---It may further be noted here that the
philosophical disputes have given a natural occasion for a certain eclecticism to be
able to form itself alongside a certain skepticism.\par}

{\small\emph{Note~3.} Concerning the difference between systematic cognizing and its
presentation in a definite systematic form, as well as the relation between what is
thus systematized and the free judicious [element], see: \textit{Om Erkj.\ og Gr.},
the Notes to \S~15, p.~135 ff.\par}

{\small\emph{Note~4.} As regards philosophy's so-called division, nothing can be
directly taught about it in this place, since it comes to depend on the whole
philosophical course of the Idea, as this---as a kind of existence's own course of
manifestation---will form itself for the philosophizing person during the execution
of philosophy's work itself. Therefore it shall here be spoken of only in a note.\par}

{\small A consideration and pursuit of philosophy's subjective principal task one
may---as Kant at bottom did with his \textit{Critique}, and as has frequently been
done by introducing one's philosophy with a so-called ``Erkenntnißlehre''---very well
let step forth as the first member, although we are thereby already led into the very
midst of the whole of existence, so that philosophy's cyclical character, grounded in
its organic character, hereby at once fully comes to light. We thereby give an account
from the first, insofar as we regard the subjective principal task as taken up into
the objective; but we must then, later, in its place, let what we have thus taken up
by way of introduction order itself within the totality of that course of existence's
manifestation which we afterward pursue.\par}

{\small Next, as a more properly real introduction to philosophy (distinct from its
propaedeutic), there would have to come a carrying-through of the explication in
speculative tendency named earlier (\S~17), so that the fundamental view which was
afterward to form the foundation could, through the conferring of all with all and
debate of all against all connected with it---that is, through dialectic---so far
constitute itself for us that it could then be laid at the foundation, and could
moreover further, in its execution and carrying-through, stand its test. In this way
we would obtain a carrying-through of the philosophizing by which one is to
philosophize one's way to the ultimate all-comprehending fundamental cognition, and
would thereby already obtain philosophy itself in an indirect way.\par}

{\small Philosophy itself would then presumably begin with a fundamental philosophy
that gave a formal presentation of the whole foundation; thereafter existence would
have to be gone through in such a way that the fundamental view was carried through
the same spheres of existence that one had provisionally considered in that real
introduction---whereby there would first arise the doctrine of the general
fundamental relations of existence (a so-called ontology), then a philosophy of
nature (insofar as one were capable of it), and thereupon a philosophy of spirit. In
the domain of the latter one would have to pass, from the natural foundation for the
objective spiritual that opens itself in consciousness, over to considering this
spiritual itself. And here would presumably first be the place for logic
proper---insofar as this is, after all, the doctrine of the logical in its appearance
in cognizing beings---as well as for the philosophy of philosophy, which would then
set itself on a level with the doctrine of the logical. To this would then be added
the speculative consideration of the aesthetic-poetic, the moral, and the religious.
By the two last we would then be led to present the human life thereby governed and
ordered, both in the individual respect and with respect to individuals'
being-together to form the larger organic wholes, state and church---the former as
holding for the earthly standpoint, the latter as pointing out beyond the sphere of
existence in which we here on earth find ourselves, so that here the doctrine of
immortality would find its place.\par}

{\small\emph{Note~5.} Philosophy's so-called method in the systematic respect---or the
whole kind and manner of the course of the Idea---will of course so depend on what
will show itself during the execution of philosophy's work itself, that nothing can
be taught about this in advance either, other than that in every domain one must let
it take the course that the matter itself will each time demand, without laying it
out throughout to carry through a certain fundamental type.\par}

{\small\emph{Note~6.} It is especially by coming forward as systematic that
speculative philosophy shows itself as constitutive. Cf.\ above, \S~13,b, Note~4,
p.~51 ff. To what was there noted the following is here to be added:\par}

{\small The philosophical fundamental Idea, and the fundamental view proceeding from
it, is to form itself for us and confirm itself for us as that which is valid with
respect to the totality of the Given; and thus it can neither form and constitute
itself at first, nor lead to the carrying-through and constitution of a whole
philosophy, except through and by means of the all-comprehending conferring with all
the Given---or with all that which it, according to its concept, is to serve to clarify
for us from the ground up. All the Given is, after all, to be taken up into the
fundamental view in such a way that all of it can thereby be seen in its innermost
rationality or reasonableness; but then it must also at every point show itself that
it is now really the Given given to us in the world---that in whose midst we are, live,
and move---which is now thus seen in its rationality, seen in the light of the
fundamental Idea, and seen as taken up into the realm of the Idea; but thus philosophy
must, not only in its working toward its foundation, and while it seeks this and
moreover seeks the starting-point for its deduction and constitution, but also in its
whole execution and carrying-through in and through the systematic whole, at every
point be assured of its agreement with actuality, as that on which its own real
validity rests. Just as it can form itself only under an all-comprehending conferring
of all with all and debate of all against all, so too, in the execution and
carrying-through of the whole systematic cognizing, it can come forward and unfold
itself in its all-comprehending validity only by showing itself as that which has
taken up into itself the results of the whole conferring and debate, as its own
moments. Just as it becomes what it becomes through this its content thus won, so too
it must further bear this within itself, as the content valid for the given actuality,
or the valid sum-total of moments; but then it must also, in the system, at every
point come forward with this its validity---that is, in and with its agreement with
the realm of actuality. Otherwise there would not set in that reconciliation and
coming-together into unity according to which we should move in the realm of the Idea
as in that of actuality, and in the realm of actuality as in that of the Idea. (Namely
as regards actuality's rationality; for only of this is there question here. Of the
extrarational, which from a realm of rationality makes the realm of the Idea into a
realm of actuality, there is no talk here; here it is only a question of the unity of
the realm of the Idea and actuality's rationality.)\par}

{\small And here two questions may answer themselves for us. The one is the
question---oddly enough---that has arisen in our time: in what relation philosophy
stands to life and actuality. The other is that which concerns the assertion, arisen
in our time, that philosophy is to begin wholly without presupposition.\par}

{\small As regards this assertion especially, one must on the contrary set up the
proposition: that philosophy, as it begins to form itself, must have the totality of
all presuppositions before it, and must not only develop itself under the resulting
reciprocal relation, but also, when it is now to constitute itself in the
philosophical system and begin its deduction and construction \textit{a priori}, must
come forward with the foundation formed and developed under that reciprocal
relation---with the fundamental Idea that has, through this reciprocal relation, come
to its fullness of content, which has taken up and dissolved all those presuppositions
into itself, and now shows itself as that which bears a corresponding fullness within
itself, and thus at every point comes forward in its agreement with actuality. (For
the rest I refer here to my work \textit{Bemærkn.\ og Undersøg.\ betreffende Hegels
Philosophie}, which is a separate offprint of some articles inserted in
\textit{Maanedskr.\ f.\ Literatur}, 10th year (1838).)\par}

{\small Even if one went up to the highest point and thought philosophy to be like
the divine all-determining, which would surely seem to be without any presupposition,
one would come to no other result. For, first, it is to be noted that the divine
all-determining and all-constituting can by no means be thought as one that, were it
to begin in time, would have to begin without presupposition; for prior to all such
beginning the whole divine fullness would have to go as the absolutely beginningless,
in which every particular is and rests in the totality. Were philosophy's beginning
thought in likeness to this, then its appearance too would already, at the first step,
have the total Idea with its whole content for its presupposition. And if one will go
more closely into the divine essence, and here look especially to wisdom, as that
whose counter-image philosophy is especially to be, and ask oneself whether the divine
wisdom does not, out of itself alone, constitute the total order of things, and
whether in a similar way reason come to fullness would not, purely out of itself
alone, constitute philosophy---then the answer must be that the divine wisdom cannot
be thought without presupposition; for it has the whole fullness of the Godhead for
its presupposition, and, more closely, love especially must show itself as its
presupposition, or as that which we must at once think in one with it. But in a
similar way philosophy must, at once from the first, go in one with real
life-apprehension and real life-view, and have the whole full living for its
presupposition.\par}

{\small An example of a science that, before all others, can be said in a certain
relative sense to begin without presupposition, we have in pure mathematics, which
begins with certain first observations of certain possibilities, and now, according
to these, itself constitutes for itself what it further pursues. If one will object
that it nevertheless presupposes ordinary human thinking, with the general
fundamental determinations, or so-called categories, underlying it, and general
fundamental demands, which without further ado come to hold and to assert themselves
in mathematical thinking as thinking---whereas philosophy, by contrast, makes these
very first presuppositions objects of an investigation and critique---then the answer
is that the very science that is to investigate these first presuppositions of all
thinking can itself come about only by means of thinking, and thus, just as fully as
mathematical thinking, must let the fundamental presuppositions lying in all thinking,
merely as thinking, hold and assert themselves precisely in the thinking by which they
themselves are to be investigated. ---So that such a science (ontology and logic)
cannot, then, come forward as more free of presupposition than mathematics. Such a
science can, however, as merely explicative---or as setting out from a simple
spiritual observation of the fundamental determinations coming forward in all
thinking---come forward with a certain freedom from presupposition, not unlike that
which can be ascribed to mathematics. In the meantime, a regard for actuality will
nevertheless constantly have to be taken therein, such as pure mathematics---which
moves only in a purely formal domain of possibilities---does not have to take. ---Such
a merely explicative ontology and logic is also what we have before us in Hegel's
so-called objective and subjective Logic, which, however, will not only come forward
as wholly free of presupposition, but even seems to be meant to hold for the proper
foundation---which it can nevertheless never be, when the all-comprehending foundation
is meant, which only the fundamental features of an all-comprehending world-view can
give.\par}

{\small\emph{Note~7.} At this place we can now also, with particular regard to
philosophy's rights, note what is to be called the domain of speculation---namely,
that this is one with the whole domain of life and existence, when, that is, one
looks only to the latter's rational foundation; and that there thus, on the one side,
cannot be any particular sum-total of objects that could be said to belong to
speculation alone, but that also, on the other side, there can be nothing from which
philosophy could be excluded. I hereby bring to remembrance what I have set forth,
especially with respect to philosophy's right and call to intervene in the positive
Christian religion---indeed in the whole cause of Christianity and its whole
work---partly in my \textit{Philosoph.\ Archiv}, both in the first number, especially
p.~71 ff., and in the fourth number, especially p.~364 ff., partly in a German
treatise, printed in the third number of the \textit{Theologische Zeitschrift}
published by Schleiermacher, de Wette, and Lücke in the years 1819--1822, and
concerning the relation between Christian faith and philosophical cognition. In these
places, moreover, the matter is also considered with respect to the interest of
Christianity and of the Christian himself, so to speak, so that there has also been
treatment of philosophy's significance for Christianity---whereby philosophy's
relation to revelation, as to a holy authority, has also been touched upon. For the
rest, I shall come to treat all this again in my \textit{Christendomsphilosophie},
when I have brought it so far to completion that I shall be able to come forward
publicly with it.\par}

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  THE FULL DETERMINATION OF THE CONCEPT OF PHILOSOPHY. (§ 20, p. 81)
%% ============================================================
\grouphead{The Full Determination of the Concept of Philosophy.}

\begin{center}\large\S~20.\end{center}
\markboth{\S~20}{\S~20}
\begin{center}Final definition of philosophy.\end{center}

The principal moments in the concept of philosophy we can now, insofar as philosophy
is considered purely objectively, gather for ourselves in the following definition:

\emph{Philosophy}---namely, speculative philosophy proper---is the science of the
true essence of all things, developing itself, under an all-comprehending conferring
of all with all and discussion and debate of all against all, in force of reason,
that is, out of an a priori ground---the philosophical fundamental Idea---by a
going-back to the ultimate grounds and fundamental relations of all things (or, as
they show themselves to be from another side, the first grounds of all things), and a
pursuit of all that to which the cognition of these must further lead, so that an
all-comprehending scientific---that is, systematic---cognition of God and world-view
may form itself for us.

{\small\emph{Note.} By the expression ``speculative philosophy proper'' it is
indicated that philosophy is here taken \textit{sensu eminenti}, whereby it is
pointed out that its concept too has its dialectical side. ---How far all philosophers
would be able to subscribe to this our final definition will here also have to be
considered.\par}

\begin{center}\rule{0.18\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  PHILOSOPHY AS A MEMBER AND MOMENT IN THE WHOLE SPIRITUAL LIFE. (§ 21, pp. 82--86)
%% ============================================================
\grouphead{Philosophy as a Member and Moment in the Whole Spiritual Life.}

\begin{center}\large\S~21.\end{center}
\markboth{\S~21}{\S~21}
\begin{center}Philosophy on a level with the empirical and the suprarational,
---with poetry, morals, and religion.\end{center}

That the whole of philosophy, although it is an all-comprehending rational cognition,
is nevertheless not a cognition all-comprehending in every respect---let alone that it
could be an all-comprehending foundation for a full living---must already be clear from
the foregoing, since it is seen from it that what we cognize philosophically must at
the same time verify itself to us partly empirically, partly sympathetically, and also
through the a priori rationality that forms itself independently of speculation, and
indeed more immediately. Already hereby the philosophical shows itself to be only a
co-moving fundamental moment in the totality, which not only has its all-comprehending
debate within itself, but also, with its whole content, is drawn---as regards the
validity of its whole striving, indeed its whole existence---into a larger debate that
is no longer conducted on philosophy's field alone, but on the field of the whole of
scientific endeavor, where philosophy can then come forward only as one power
[Potents] beside several, indeed must even seek to maintain itself over against, on the
one side, empiricism, on the other side, suprarationalism, while the empirical and the
suprarational in turn have to maintain themselves in their subsistence independent of
all philosophy.

But further, the whole of cognizing itself, with the whole of scientific endeavor in
general, again comes forth and subsists only in a debate that is conducted on the
domain of the whole of living. Here the whole of scientific endeavor, with all the
cognizing lying within it, becomes first and most immediately a power over against, and
in cooperation with, art---especially ideal art and the poetry lying within it, in
which the ideal content of the philosophical world-view, that is, the fullness of
reason, shows itself as stepping out into, and realizing itself in (incarnating itself
in), an ideal counter-image of actuality.

But next, both these powers stand finally in relation to that in us which makes us more
than cognizing and beholding beings---namely, to feeling and willing beings; and now
both philosophy and poetry, both science and art, must go into unity with all this,
under a debate in which everything depends on mind and will being filled and
strengthened by them, and in turn reacting back upon them; under which religiosity,
with faith and love, trust and hope, in the domain of the mind, and morality, with
reverence and holiness, in the domain of the will, acquire a significance like that
which philosophy has in the domain of scientific endeavor and poetry in that of art.

{\small\emph{Note~1.} We here return again to the opposition between the rational and
the extrarational, and to the remark that life and existence subsist by an essential
unity of both (\S~13,b, Note~3, p.~50), to which we may attach the remark that life in
consciousness first and most immediately arises in the domain of the extrarational,
although it cannot stir in this without at once coming into the domain of the cognition
of connection and relation---that is, into the domain of rationality. But now we note
that to the rationality which thus at once asserts itself we come either from the side
of the nearest, through the ideal glance at the immediately given and the pursuit of
what most immediately attaches to it, while the Given in its givenness becomes the
foundation---be it now in the domain of empirical inquiry or in that of suprarational
movedness; or else by going to the deeper grounds and setting out from these, in order
to interpret this Given to ourselves, so that its explanation, and the theory thus
formed, becomes for us the foundation. On this it is that the opposition between
rationalism and empiricism in the one respect, and the opposition between rationalism
and suprarationalism in the other respect, in general rests. As suprarational we here
regard everything that, by a power other than that of rationality, and beyond what we
see a way to account for, so asserts itself to us as something higher that we cannot
but give it room. Yet especially only that which, in a religious respect---or with
regard to the governance of life to which we must give ourselves over---exercises this
power over us is called the suprarational; which then also, insofar as this its power
springs from an origin that may be regarded as exceeding the whole ordinary earthly
order of things (as an outbreak of a higher order coming forward point by point and in
flashes), is called the supernatural. ---That we constantly live and move both in the
region of the empirical and in that of the suprarational (in that we, for example,
constantly live in trust in a governance) will readily be seen.\par}

{\small\emph{Note~2.} Of the relation between philosophy and Christianity, of which
there was talk above in Note~7 to \S~19, pp.~79--80, there shall be no further treatment
here---only that I will here bring to remembrance the following words, which are read in
my \textit{Philos.\ Arch.}, p.~62: ``although philosophy is the spirit in Christian
cognizing, yet the trustful faith---going out beyond all seeing and cognizing, enclosing
within itself, as in a life-kernel, all that is cognized and much more than that---which,
as a surrender to the power of conviction in life and deed, is a practical living in and
with the fullness of this conviction, is the soul in the whole Christian existence, in
which that philosophical element is but one moment---though certainly an essential
fundamental moment.''\par}

{\small\emph{Note~3.} Under the philosophical striving's appearance in the larger human
life, it becomes a natural consequence of the sporadic in existence that several
so-called philosophies form themselves, which lead to the formation of different
philosophical schools, partly in a theoretical, partly in a practical respect.\par}

{\small\emph{Note~4.} We may here, finally, look to the supremacy that Hegel and his
school have wished to claim for philosophy. For the endeavor of cognition, in its
advance to the thorough cognition of the full truth, there stands as highest aim the
cognition of truth in its truth, and moreover in the unfolding of all its moments, as
such, and their taking-up into unity; and here, then, completed philosophy, as fully
carried-through cognition of truth, comes forward as the highest at which we aim.\par}

But if we now think ourselves to have come so far, we must nevertheless see and
acknowledge that the realm of truth must now also show itself to us in its real power in
concrete life itself, as the realm of the Idea realized therein and constantly realizing
itself. But then we see that, in order to come to the full, we cannot stop at philosophy,
but must go beyond it. To this will now first and foremost belong philosophy's going into
unity with our living in the whole sphere of actuality, insofar as we are and move in it
in a wholly concrete way---that is, with our living in nature, human life, and history.
But this goes further; for now art forms itself---namely, ideal art---as the
counter-image of life, in which we see the fullness of the infinite in powerful presence
stirring and living in the whole play of concrete life, whereby it at the same time so
addresses us that it is beheld with delight, so that we now take it up personally, and
live our way into it, in a quite different manner than we could so long as it stood
before us only in its universal validity. And thus we see that, if beauty were enough to
designate the whole aesthetically-poetically appealing, we could and would have to say:
the realm of truth is to be for us more than a realm of truth---it is to become for us a
realm of beauty; but better we say: a realm of poetry.

But the human being is not only to study the all-valid as such, and rejoice in it; he
must also set himself in the right rapport or fundamental relation to it with respect to
his personal self---or he must make the all-valid all-valid for himself, in his own
innermost living, as this is to proceed from the personal central point in him. This
central point we find in the human will, whose most essential fundamental utterance must
be the personal desire's and striving's surrender to an all-valid, all-comprehending
will, which then constitutes itself for us, partly as the ruler and governor in all our
striving, willing, and doing, partly as our deepest hold and the object of our deepest
attachment and surrender, insofar as we are subject to the total course of life's
dispensations. Thus we come to acknowledge that it is not enough that the realm of truth
rises for us as that which also comes forward as a realm of beauty, or better a realm of
poetry, but that it must also show itself to us as the realm of the good; yet this
consideration completes itself for us only when we now, going further, see that we must
live in the realm of truth and of the fullness of life as in that which is at the same
time for us a realm of love and of holiness---and our thereby-designated being-taken-up
into the fullness of life will then stand for us as the highest aim.

\begin{center}\rule{0.28\linewidth}{0.4pt}\end{center}

%% ============================================================
%%  RETTELSER (errata leaf). The Danish errata correct only the ORIGINAL's
%%  orthography, so they are meaningless in English. Per the erkjendelse policy,
%%  render a one-line translator's note here at finish (or omit). Decide at finish.
%% ============================================================
{\footnotesize\noindent\emph{Translator's note.} The Danish first edition (1843)
closes with a leaf of \emph{Rettelser} (errata) correcting a number of
page-and-line orthographic slips in the Danish text. As these corrections concern
only the spelling of the Danish original, they have no bearing on the English and
are not reproduced here; the substantive (NB) corrections have in any case been
incorporated into the Danish transcription from which this translation was made. The
same leaf also notes that, by a printer's oversight, two consecutive sections were
both numbered 13; following the original's table of contents, they are given here as
\S~13,a and \S~13,b.\par}

\end{document}
